### 2021-04-15 00-author.Rmd
#First upload of book

---
output:
  pdf_document: default
  html_document: default
  word_document: default
---
# About the Author {- #author}

John Measey is Associate Professor of Biological Sciences at Stellenbosch University. He has authored or co-authored more than 200 peer reviewed scientific papers and book chapters, and five books. He has been the Editor-in-Chief of an ISI journal for 9 years, and currently serves as Associate Editor for 4 other journals. He has graduated more than 20 postgraduate students, and his blog on writing and publishing in biological sciences is read by thousands globally. British born and educated, he lives and works in the beautiful Western Cape, South Africa.

# Do you have something to contribute? {- #contribute}

This book is written in bookdown [@xie2016bookdown] specifically to make it a 'live project' that will be open to anyone who wants to contribute, improve, or use as the basis for your own book in a different language, or for a different region on the world. It would be impossible to cover every aspect of writing a PhD in Biological Sciences, and it may be that I have missed out ones that are very important to you. Equally, what is currently written will become obsolete very quickly as new initiatives are begun, and old problems are resolved. For this reason, this guide needs to be a 'living document', and anyone who wants to provide feedback or contribute new sections is more than welcome. Please feel free to open an issue, or make a Pull Request if you spot a typo.

### 2021-04-15 1.01-beforeyoustart.Rmd
#First upload of book

---
output:
  html_document: default
  pdf_document: default
---
```{r iris}
knitr::kable(
  head(iris), caption = 'We must have the iris data and some R code, otherwise it would not seem right',
  booktabs = TRUE
)
```

\mainmatter

# Introduction  {#intro}

Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you are going to do afterwards. There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make the informed decisions from the outset. For example, reconciling yourself to using the statistical package R will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on. Getting, and staying in, the right mental framework will also be very important for your ability to work under pressure. Again, getting into good habits at the beginning will put you in a commanding position later.

## So you are doing a PhD? {#beforeyoustart}

In this first chapter of the book I intend to make you familiar with what a PhD is, how to approach thinking about it, and what to look for in an institution and advisor. You may already have some or all of these things fixed, but there is still some use in going over these basic points. Try to keep an open mind. Doing a PhD will take several years of your life, and may well define your future career prospects. Therefore, to say that this is a big decision is an understatement. It is not a substitute for getting a job or a way of remaining a student for longer. If after reading this section you are in any doubt about whether or not a PhD is right for you, then I would suggest that you do need to think about it further. Don't take my word for it, go and talk to people who have done a PhD (whether or not they now have a job), or those who are doing a PhD now. Especially if their experience is in your focal research area, they are likely to know a lot more about it.

### What is a PhD?

PhD comes from the Latin _Philosophiae Doctor_, and means "Doctor of Philosophy". All doctorates derive from philosophy which is essentially the mother of all academic subjects.

Obtaining a PhD allows you to (legitimately) call yourself a Doctor, and/or put the letters PhD after your name. You likely already have other letters (BSc, MSc, etc.) after your name, so this won't be anything new). Becoming a Doctor is (in most parts of the world) the last step in your formal training as a researcher. However, in some countries (see list [here](https://en.wikipedia.org/wiki/Habilitation)) a further process (Habilitation) is required to lead research groups and/or teach at universities (leading the way to a professorship).

PhDs are granted by universities (for the most part) and those are the institutions that set the standards. In many countries, these universities are required to meet the criteria of national bodies. Essentially, the top-down control is to ensure that the PhD qualification adheres to national and international standards. You may be aware that there are mutterings of whether qualifications are what they used to be. This has also been said of PhDs. However, I'm of the opinion that the standards are much as they were. You will really set the level on the bar of your own PhD. You either pass or fail. If your intention were to do the 'bare minimum' required to get a PhD, then you would have to look carefully for an advisor that would be happy to preside over such a thesis. Suffice it to say that this would be a very poor place to start, and there may be lots of other reasons why doing a PhD really is not appropriate for you.

### Is a PhD more than a thesis?
Yes. The thesis is really only the final evidence of all of the work and experience that you undertake while doing your PhD. The time that you spend doing your PhD will likely include participating in conferences, giving talks, participating in journal clubs, seminars, workshops and working groups. In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular area. You will also learn and develop your skills in presentation, analysis and treatment of data. It is an intense learning experience. It should also be a fun and positive experience.

While your PhD thesis will be a crucial and focal product of your PhD, and producing it will require a large amount of work, the thesis cannot be considered equivalent to the PhD.

Having said all of this, we have not really answered the question:

__What is a PhD?__

Because every PhD is different, it's quite a hard question to answer. Instead, I think that a better approach is to ask for the qualities that examiners are asked to assess when examining your PhD thesis. Again, this will be different in different institutions and countries. Some countries require an oral defence where a specialist jury is presented with the PhD dissertation and can grill the candidate for hours in public. Others, like my defense at Bristol University, happened behind closed doors with just two specialist examiners: an internal and an external.

### Criteria used to judge PhD studies
In order to provide you with some general idea of what is required, I list below the (edited) questions asked of examiners for a PhD thesis by my university:

- Has the motivation for the objectives of the study been formulated satisfactorily?
- Do the research results constitute a meaningful contribution to the knowledge of and insight into the relevant field of study?
- Does the dissertation distinguish clearly between own, new contributions to, and known results in the relevant field of study?
- Is the candidate capable of evaluating the scientific meaning of his / her results and of placing this in context within existing knowledge in the field of study?
- Does the candidate show signs of independent, critical thinking or other signs of originality?
- Does the candidate show that they are sufficiently capable of doing independent research?
- Does the dissertation show that the candidate is sufficiently familiar with the relevant research techniques and methods?
- Does the dissertation show conversance with, and a critical attitude towards the pertinent literature?
- Is the material presented in a clear, systematic and logical manner?
- Is the linguistic, stylistic and technical editing of the dissertation acceptable?
- Are the research results acceptable for publication?

If you compare these PhD requirements with those of an MSc, the key difference is that for a PhD you are required to produce a "meaningful contribution to the knowledge of and insight into the relevant field of study", and show that you are "...capable of doing independent research". Thus, a PhD degree prepares you to be an independent researcher, and so you are expected to have a critical attitude toward other research.

### Who are the examiners?
The next question to ask is who the examiners are? Again, different institutions will have different guidelines about who can be an examiner for a PhD. It is likely that most will agree that they should be holders of PhDs, in tenured positions (sometimes universities insist on full professorial appointments for examiners), and with experience of advising PhD students as well as having a proven record of their own publications. Hence, most PhD examiners will likely be senior academic figures. Their views are likely to have been informed by their own experience in academia, their own institutions and systems and prior experience of examining PhDs at other institutions.

I can tell you that many examiners will treat the exercise the same irrespective of the questions asked of them. They will treat the examination as an extra-large task in peer review. This is not unfair. Personally, 
I think that there are some special extra hurdles required for a PhD (over and above 5 chapters that are publishable):

- First is the clear ability to conceptualise and carry out a body of scientific work that is novel. Here I mean novel in that it has not simply been a repeat of some previous work on a different taxon or system. This is easier to judge from an oral defense, over reading the thesis. 
- Second is the clear knowledge of how the work that is conducted fits into the bigger picture. That is that the work demonstrates that the person is a scientist at a broader level than the narrow focus of their research.
- Third is whether the student has been able to bring the study together in a way that shows that it is more than the sum of its parts. In essence, together the 5 data chapters should provide a bigger overview of a broader topic, and show the direction of future work.

I prefer a thesis that fits together within a stated framework. I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the introduction of their proposal. A useful concept to be aware of is the 'Hierarchy of Hypotheses' [@heger2018hierarchy]. In this approach, you are encouraged to consider the 'bigger question' in your research area, identifying both what studies have produced evidence and identifying gaps from a theory driven approach. Generally, it is only possible to consider these types of concepts once you are familiar with the literature (both theoretical and experimental).

### What will you do after your PhD?
It sounds silly to think about what comes after your PhD even as you are just starting, but the reality is that you always need to be planning ahead in order to make the most from what you are doing now. During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas. Knowing what you want to do when you finish will help guide your decisions so that you can maximise opportunities as they arise. Both inside and outside of academia, it is likely to be the contacts that you make that will help in your next career move.

#### Doing a PhD actually reduces your chance of employment and lowers your pay
The reality is that most employers don’t want someone with a PhD, because it is a very academic qualification. It will take you so far down an academic rabbit hole, that for an employer there’s very little to use. Most employers want people who understand the nuts and bolts of doing research, but who can apply this knowledge outside of the academic system. Employers who want people with PhDs (and there are a few) generally want professional researchers who are essentially academics. Good work if you can get it, but it tends not to pay well. You might end up only being fit to be an academic, and that life is getting increasingly harder without any end to the expectations of the employers… like the equivalent of doing a PhD every year. Something only very few people will enjoy.

At the same time, governments the world over, and especially developing countries, are providing bursaries for people to do PhDs. Why? The number of PhDs in a country is considered  (by the [World Bank](https://datacatalog.worldbank.org/dataset/world-development-indicators)) to be one of many measures of the development of that country. As the percentage gets larger, so countries can convince the world bank that they are developing (and then get granted its favours, see Figure \@ref(fig:PhDMap)). This is all well and good, but strategically, governments should consider why they want these researchers in their populations, and more particularly for what kinds of jobs. It seems somewhat unfair to produce lots of young PhD holders that have no careers to join, or prospects to use their PhD just so that your government can tick a check box on their road to development, and so borrow more money from the World Bank. If you are from a developing country, you cannot assume that just because you are given a bursary by your government it follows that they will give you a job or a career. This will likely always be your own task.

(ref:PhDMap) **Where in the world are there most PhDs?** In this map of the world, countries shaded are those that have data at the declaring the proportion of PhD (or equivalent) in their population (Max percentage between 2010 and 2020 (Data from the World Bank). This ranges from 2.97% in Switzerland (dark red) to 0.02% in DR Congo (light yellow). Notably there is no data held in these years for many countries in Asia, Africa and South America.

```{r PhDMap, fig.show='hold', out.width = '95%', echo=FALSE, fig.align='left', fig.cap="(ref:PhDMap)"}
knitr::include_graphics('figures/PhDMap.png')
```

In a recent report from the UK, it was found that by age 35, the earnings gap for those with PhDs was lower than the same cohort that left academia after their BSc or MSc [@britton2020earnings]. Women who obtain PhDs tend to increase their earnings by ~8%, while men _reduce_ their earnings by ~9%. However, for those in Biological Sciences women seem likely to earn no more than those with undergraduate degrees by age 35. Men earn around 10% less. The point here is that you should not be doing a PhD in Biological Sciences because you think you will earn more money in your following career. Having made this point, I also feel that it is important to point out that how much you earn may not be directly related to job satisfaction or lifetime happiness. To my knowledge, this is not (yet) reported, but could be a lot higher for those with PhDs, even though they earn less.

It is important to realise that many academics are looking for good students to do PhD work. This is because they need someone to do research within their groups, and students or postdocs do most of this work. For this reason, an academic you know while you're doing your MSc studies may start to try to persuade you to do a PhD. This could be fantastic for you, it could be just what you're looking for. But the reality is that you will never do a PhD for somebody else, you are always going to be doing it for yourself. if the only reason you are doing a PhD is because someone else told you that you should then you are likely to have a really hard time. By the time you graduate, you may have a PhD but without any idea of what you're going to do with it.

### Think about your career before you start your PhD
The last thing a PhD should be is a stop gap for you to fill your time because you can’t think of anything else to do. Although this does sound ludicrous, there are many people in academia who did PhDs for this reason, or simply because they were offered an opportunity to do a PhD and felt that they couldn’t turn it down.

You should be aware by now that there are lots of people in academia who are ‘institutionalised’. By this I mean that they have never had a job (or a life) outside of academia. This is not something that I can recommend for anyone. Such people are quite peculiar in many ways, and they tend to operate outside of the norms and standards of the rest of society. In particular, they make academia a more difficult place to work, because they tend to accentuate and maintain a lot of the silliness within these institutions. If you are worried about becoming such a person, then I think it would be advisable to take some time outside of academia. For example, if you have never worked but are thinking of doing a PhD, consider taking a year (or two) to do a job before starting a PhD. If it is in a related field it will likely help you greatly. I suggest that it will certainly help keep your feet on the ground, and in touch with some level of reality or normality, during your PhD. You’ll also have some experience and connections to go back to once you’ve finished.

### Don't rely on becoming an academic
Academia is getting more competitive. Whereas once it was almost certain that you would be employed as an academic after finishing your PhD, this is no longer the case. Indeed, the number of people who get PhDs is greatly increasing the world over, and most are in direct competition for a decreasing amount of academic posts. In 2019, 39% of respondents to a survey said that the most important reason they decided to enrol in a PhD programme was to pursue an academic career [@woolston2019phds]. Further, 56% wanted academia as their first choice for a career. Yet in Belgium, for example, the numbers of PhD graduates entering the market nearly doubled (in the first 15 years of the 21st century) while academic positions remained static [@levecque2017work]. Universities appear to be exploiting this excess of qualified people to hire them on contracts, reducing the number of full time positions. Essentially, this means that in most places your time as a PhD student, postdoc and junior lecturer will keep you outside the employers core staff. Joining those on the inside (with well funded posts) becomes increasingly more difficult. Some have suggested that the world of academia is becoming more like a gang producing drugs [@afonso2014academia].

### Doing a PhD is really hard work
It’s not a MSc with an extra year. It really is much harder and tougher on you mentally and physically. The chances that you’ll end up jaded and malcontent with the entire academic system are very high (maybe even obligatory). You are probably not going to be very happy while doing it (there are some exceptions of course), and it will likely have negative impacts on all the good things that you currently have in your life (think family, friends and relationships, as well as free time and fun).

### Your mental health will likely suffer
Doing a PhD will be the cause of stress not only in your working life, but will also impact other areas of your life. Being aware of this at the outset will allow you to alert close members of your non-working life (think partner, family and friends) that they may well need to act as a support network. It is worth tracking your mental health during your PhD to check that you are not getting into difficulties.

### Any PhD has to be yours
It’s entirely useless to do a PhD for someone else. You must take complete ownership in order to do it. In fact, you need to be obsessed with the PhD subject and really want to do nothing else at all. A PhD is an obsession where you can be unwavering and manic in your fascination with a subject. Very few people really understand what a great privilege this is. If you don’t feel like a total maniac for learning, a PhD is not for you.

### Does it matter where you do your PhD?
Not all universities in the world have a PhD program. This means that you cannot do a PhD in some institutions. Larger institutions are more likely to hold PhD programs, have more potential advisors, and may even have funds for bursaries to conduct PhD studies. It is likely that you are going to need to do some work at the university in order to justify any bursary that they may want to give you. Typically this kind of work is the teaching of undergraduates which is a generally good and positive experience in the world of academia.

The country where you study a PhD does not determine the quality of institution (see Figure \@ref(fig:THEreseasrchcountry). Rather look at the merit of the particular institution that you are interested in. Surveys like that from the [Times Higher Education Supplement](https://www.timeshighereducation.com/world-university-rankings/2021/world-ranking) can help give you an overview.

(ref:THEreseasrchcountry) **It’s not the country that determines the quality of your university.** In this plot you see the Times Higher Education (THE) World Ranking (2021) score for Research plotted by country (with ≥ 10 universities). Note that most universities in most countries have mean THE Research scores in the lowest quartile. Almost all countries that have institutions in the top quartile, have most of the institutions in that country in the bottom quartile. The point here is that you should look at the quality of the institution that you want to attend, rather than assume that all universities from certain countries have high or low research profiles.

```{r THEreseasrchcountry, out.width = '95%', echo=FALSE, fig.cap="(ref:THEreseasrchcountry)"}
 knitr::include_graphics("figures/THEreseasrchcountry.png")
```

Depending on where you study might mean that you have different amounts of course time. There are some institutions that will carry on funding you for as long as you want to carry on studying. This is unusual. Most countries now have a restriction on the amount of time that you can spend studying a PhD.

It does make a difference where in the world you do your PhD. There are some beautiful places to work. Some places are more expensive to live and therefore survive. If you are a biologist then you might want to consider where your study animals or plants are in order to decide your place to study. But most of all your choice of where to study is likely to be dictated by the presence of an adviser that you want to work with, and who wants you as their PhD student. Advisors are tied to their institutions, and any principal advisor will need to register you at their own institution, meaning that this will be where you spend your desk time. Some advisors might be more flexible than others, but it will be most likely that you will need to move to the institution where your advisor is based.

### What to look for in an advisor
Throughout this book I will call the person who is the academic in charge of your PhD studies your ‘advisor’. I will not call them your supervisor. The reason for this is because I think it is much better for you to think of this person as giving you advice and not the person who tells you what to do. This comes back to the idea that the PhD is yours. It is not your advisor’s PhD. They will not write it. They will not do all the work that's required for it. That will be yours alone.

Just as I've attempted to do in this book, that person will give you advice. It will be up to you whether or not you follow that advice. Many of my students have not followed my advice. For some this has been the right thing to do. For others it has not. No one can know what the best thing to do is in different situations, because they will almost always be unique. You will be the expert on your project. That means that as the person closest to your project, you are the person who is best placed to make a decision. But your advisor can, and should, help you to make those decisions.

Your advisor can help because your advisor has experience of doing research themselves. They hopefully have some experience of advising other students. And they should have some experience of having done research in your chosen study area before, including having published papers in appropriate journals.

The advisor that you're looking for should be offering or doing projects in which you are interested. There may only be a few people in the world that are doing this type of research, and you may end up having to choose between them. Hence, while you're reading about how to pick a good advisor, your shortlist may only be of two or three people in the world. The project and subject area are going to be strong areas driving your choices moving forwards.

One of the best ways of picking a good advisor is by talking to current or past students. You should ask them how happy they are with their advisor and whether they are receiving the support that they need to do their projects. A good adviser should have an open door so that you can talk to them whenever you need to. Thus, once you have talked to the students in their group, you should also make an appointment to talk to them about your project. It might not be one meeting, there may be several. The more contact you have with them, the more likely you are to be sure that this is the person whose help that you want when conducting research over the next few years.

Good advisors tend to have good strong research groups that work well together. You can look through the publications of a research group to see how well they work together. You should see the names of many of the graduate students appearing on publications demonstrating the level of collaborative engagement within the research group.

A good adviser should have a good network. Both during your PhD studies and after them, you will rely primarily on the network of your advisor as a basis to building your own network. For example, you will go to meetings that they go to, you will be introduced to people that they know, and you may have opportunities to collaborate with people within their networks [see @adams2012rise]. Doing a PhD with an advisor that has no network and that does not link up with other academics (or worse is disliked by other academics) might mean that your own work is not recognised, or even sidelined. You can easily check on the network of a potential advisor by looking at the collaborators in their publications, or on their website.

Having said this, the best academics also tend to be very busy, and that may mean that they do not make good advisors. The balance is definitely very tricky, and one of the best ways to make sure that you have enough time for all of your students is to limit the number of students you have. Clearly, this works against prospective students.

### What to avoid in an advisor
Remember that relationships occur on a one-to-one basis, and it may be that not every advisor gets on with every student. There are some advisors who have many problems with many students, and these are best avoided no matter how high their profile is [see @malaga2004meyer;@abbott2004junior; but also see @vences2004meyer for alternative viewpoints]. There are other advisers who are so busy that they are never able to meet with their students, and pass this task on to somebody else in their group. Again no matter how high profile these people are, if they can't advise their own students they are best avoided.

At the end of the day there is no such thing as the perfect advisor. You will likely have to make a trade-off on the above points with someone who has a project that they are willing to advise you on, and your chosen area of research.

### What makes a good project?
After having made lots of statements about how a project should be your own project the following advice might seem a bit strange. Clearly the work that you're going to be doing will need to fit into the group of the adviser that you've chosen.

The most obvious point to make here is that a project should be something that you want to do and that you're prepared to dedicate yourself to for at least 3 years.

Projects that are cross-disciplinary are likely to lead to more opportunities in more disciplines. You may want to make sure that you do not narrow your focus entirely into a single technique or method. If you view your PhD as a training experience, then it is better not to be doing exactly the same techniques that you have mastered previously, for example during your MSc. It is worth bearing in mind that academia has trends and fashions and that becoming an expert in a brand new technique could land you a job and career. It could equally mean that your entire knowledge of a subject area is replaced by a better approach within one or two years.

### Funding your PhD studies
Funding may be one of the biggest obstacles that you come across when wanting to do a PhD. It's important that you consider the funding of your PhD in two separate ways:

#### The funding that you need to live 
This money is typically called a bursary or stipend depending on where you are in the world. It seems to me that many countries offer bursaries that are not actually enough to live on (e.g. buy food, rent a house etc.). If you're offered a PhD bursary, especially if it's in another country, then do check on the living cost at that location and whether or not the bursary that you are being offered will be sufficient money in order for you to survive. Also, be very careful to check how long the bursary is for. For example, at my university a PhD bursary is given only for 3 years. After this three-year funding period, you will be expected to have submitted your PhD for examination. If you need further funding to live while your PhD is being examined, and before you graduate then you will need to get a job. There is no more money!

It's a really good idea to talk to people who have already done a PhD to find out more about the different available bursary opportunities, and how people manage with and without them. There is no one single model, and the more people you talk to the more likely you are to discover one that might suit you. Of course conferring with other people can include reading the accounts of many different people who share their experiences on social media. There are blogs and Twitter accounts dedicated to this stuff. Try and be as open-minded as you can.

#### The cost associated with your studies. 
The money associated with the project that you were doing also needs to come from somewhere. Depending on where in the world you can do these studies they will be different expectations of where this money comes from. This is a very important topic for you and your potential advisor to discuss. During the proposal phase of your PhD you'll be expected to put together a budget to determine the total costs of your studies. In reality you or your advisor should have a ballpark figure of how much this Is going to cost and have some idea about where the money is going to come from.

Again if you are moving to a new location to do a PhD then beware that costs are not the same in different countries. For example, you might be used to paying very little money for accommodation and travelling in the field, essentially meaning that you can do as much fieldwork as you like. But in some countries fieldwork will make up such a substantial proportion of the funding that you may be limited in the amount of field work that you can afford to do.

It is possible that your potential advisor will tell you that they don't have any funding for a project that you may want to do. This is both a curse and an opportunity. There are many opportunities to raise money to fund PhD projects. Raising the funding yourself is one way of making sure it really remains your project. But the availability of this kind of funding does vary from country to country and you may want to embark on the fundraising side of your PhD before you register. Especially if you have a limited amount of time once you have registered in order to complete your PhD studies. 
Some projects are much cheaper than others. For example, if your intention is to sequence genomes of every animal or plant that you work with, the finance required will likely be far greater than your bursary. If however your intention is to do a PhD that models distributions of populations of animals and plants, then the entire costs may only run to a computer remote access to a computing cluster and data storage.

Funding, and the availability of funding, may well be the tipping point that makes you decide where you will do your PhD studies, and with whom. If you haven't already got the message (see above) then these are important decisions and you may be better off postponing the start of your PhD in order to raise money, and do the project you want, in the place you want, and with the person you want to advise you.

### 2021-04-15 1.02-useR.Rmd
#First upload of book

## Reconciling yourself to doing things you've been avoiding {#useR}
Some 30 odd years ago as an undergraduate I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. Each time the cursor got to the end of the screen I hit the carriage return button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC. This is practically where we are now with R [@rcoreteam2021r], and indeed a wealth of other alternative programming platforms.

(ref:Rlogo) **The ubiquitous R logo.** R (and the invaluable GUI RStudio) has become the go to platform for many statistics and figures in biological sciences

```{r Rlogo, out.width = '33%', echo=FALSE, fig.align='centre', fig.cap="(ref:Rlogo)"}
 knitr::include_graphics("figures/Rlogo.png")
```

If you are not already aware of the prominence that R now has in conducting statistical analyses in biological sciences, then you’ve missed out on a great deal. The real point here is not that you can only use R, but that R represents a platform that can be easily reproduced (because it’s free unlike many statistical software packages), and by saving and making the code available, you are making your data analyses reproducible. We will see elsewhere the importance of reproducibility in sciences [@baker20161500], and the critical role this has in transparency and therefore best practice.

What is perhaps most remarkable about learning R is the freedom it gives you away from large amounts of commercial software. As a basic programming language it can be used for statistical analysis, drawing figures, conducting geographic information system (GIS) level work, and even as a word processor on which to write your thesis [and indeed this book @xie2016bookdown].

If you already know all of the above, then you won’t face any major barriers, and you will fit into the new and more transparent world of science with great ease. If the thought of using R fills you with dread, in the same way that using a computer did some 30 years ago, then you should reconcile yourself now before you go any further. Learning R will be a steep learning curve, but it will be made easier by the large amount of excellent video tutorials and online courses that are available.

### And learning lots of other stuff...
The same principle that I’ve written about above can be extended to all sorts of other aspects of your PhD project. You will have to learn new skills, and some of these will be things that you have previously resisted having to do. Allow yourself to be open to all aspects of learning, and all skills. The act of learning opens new pathways in your brain, and using this ‘muscle’ will facilitate future learning exercises.

### 2021-04-15 1.03-lifescientific.Rmd
#First upload of book

## The scientific project and scientific living {#lifescientific}

Throughout the book, I make reference to the ‘scientific project’ as a greater plan to which our studies in biological sciences is simply a part. I consider the scientific project as a path that humanity should continue to travel in order to reach a higher and more just way of living. Our societies have evolved from places where superstition and belief were the basis for societal rules that favoured the privileged classes. Science is a process in which we reach conclusions based on egalitarian reasoning and evidence, something that we can aspire for the larger community around us. Already, most societies are making the majority of their collective evidence based decisions through the guidance of science. Therefore, if we consider our knowledge based societies to be democratic, then all citizens of that society should understand the processes under which the foundation of that knowledge is acquired [@sagan2011demon]. Yet globally, public skepticism in science has been rising, and has reached as high as 35%. To consolidate this growing position of science in society, and the decline in public acceptance, scientists will need to become more open, communicative and seen to be conducting their own profession in a way that reflects these ideals. There is evidence that increasing communication from scientists to the public results in a decrease of skepticism ([see here](https://multimedia.3m.com/mws/media/1898512O/3m-sosi-2020-pandemic-pulse-global-report-pdf.pdf)).

Imagine a world in which all of the principle decisions are made as a result of advances in science, but vanishingly small numbers of people actually understand how science works. This is not a world that we would want, but as Carl Sagan [@sagan2011demon] pointed out, it’s one that could all too easily happen. Therefore, to prevent the workings of science falling into the hands of a minority, we have a duty to communicate our science to the majority.

Ultimately, what we want from the scientific project is to draw everyone in. That is not to say that there is no place in the world for other subject areas. One of the greatest misadventures in my own education was the indoctrination that I received that there were two streams in education, one for STEM and one for arts. I hope that you managed to escape this narrow minded and factually incorrect view, because as you will discover there are high levels of creativity demanded within the scientific framework, and those who have more talent in communicating their science will likely be more successful in their careers as a result. Moving forward with the scientific project does not mean that everyone must become a scientist, but instead that the process by which scientists establish knowledge is understood. Indeed, as we shall see (below) sceptical thinking, as exemplified in science, is indeed a healthy way for everyone in society to avoid time wasting and potentially being deceived or conned. It is also vital that the scientific process be transparent in a way that results can be understood and therefore respected.

Here I provide an example of the way in which there is a need for efficient communication between biological sciences and civic society:

### Example: Invasive species 
Alien species are those that have been moved by human agents across biogeographic barriers into parts of the world where they did not evolve. Once these aliens arrive in their new environment, some breed and become established, and after a lag a fraction of these spread and become invasive, often with substantial impacts on the environment and the socio-economic efforts of people in those invaded areas. Invasive species disproportionately affect poorer people in our society, often completely disassociated with the pathways responsible for the introduction and spread. Removing invasive species requires a considerable amount of effort, and unless this is supplied by tax-payers, via governmental policy, falls on those directly impacted. Invasion science, the expanded transdisciplinary area stemming from invasion biology, is required to engage with all stakeholders in invasions, and most importantly to prevent new invasions. Because introductions are made by humans there is a critical need to communicate with those stakeholders along introduction pathways.

For existing invasions, social scientists and economists are needed to determine their impacts on people in invaded areas and determine how to best alleviate the invasion or mitigate its effects without it moving further and impacting more communities. Biological scientists need to assess the impact on the environment, and investigate ways in which the species can be removed, their spread curtailed, or their impacts neutralised. Importantly, the results need to be discussed with all stakeholders to achieve a desirable outcome for those impacted most. Invasions at larger scales will normally require governmental resources to ensure that communities can continue to function, especially to best deliver their services to the society. Governments require evidence based decisions that are reached from well researched studies conducted in the field. Effective measures may necessitate large sums of money, and governments require that this process be transparent so that the taxpayers’ money they are investing is spent efficiently.

We know enough about invasive species to be aware that there are greater threats to society from species that have not yet arrived than those already present. This means that all societies need to guard against the threat of potential future invasions, and still maintain the trade that brings needed goods into their areas. Most societies have no checks in place, but the immediate need to prevent new invasions, and halt the effects of existing ones, is recognised on a global scale. Biologists have an important role to play as key agents in informing society on the impacts (especially when they disproportionately impact those least privileged in society), pathways, removal and prevention of invasive species. Failure to effectively communicate with any of the stakeholders along the invasion pathway will result in more propagules arriving, and with them the increased chance of more species becoming established and eventually invasive.

In the case of biological invasions, communication needs first to flow from the various stakeholders to direct the research needed. The solutions required are context specific, and the lessons learned need to be communicated back to the stakeholders, including governmental policy makers, in the form they require to action the findings. Ineffective communication from scientists to the stakeholders may result in solutions being ignored or misinterpreted. Documentation disseminated to the global scientific community, through scientific publications, is needed in case it may help the response of other societies with similar problems.

### Should society lead the sciences or sciences lead society? 
Of course, the answer is simply that there must be feedback between science and society. As science advances, society learns of these advances and views are advanced. To become incorporated as a societal norm takes time. Over time, views from science that appear radical to society become mainstream and accepted. Think of the views on the earth being round, or more recently the general acceptance that the universe started with the ‘big bang’. The counterculture, also known as post truth, faction in society that espouses views accepted by the mainstream and proved by science beyond reasonable doubt is not a novel phenomenon. Indeed, in the scientific project, we need to accept that these views will persist, but empower the general populace to understand why and how they can be shown to be false. Hence, your view that the earth is not flat is bolstered if you can provide three ways to demonstrate this as fact (pick the ones you like from 7 listed [here](http://crosstalk.cell.com/blog/seven-ways-to-prove-earth-is-round) or [here](https://www.unlv.edu/news/release/round-earth-clues-how-science-proves-our-home-globe)). Otherwise, your assertion that it is not flat will appear as easily denied as a ‘flat-earther’s’ assertion that it is. The power of the scientific argument in society is that it is not simply contradiction or gainsay, but that there is evidence.

The scientific project is currently not the open virtuous eutopia that we would like. Indeed, as scientists are humans and part of the current society, they are prone to the same trappings of wielding power and influence in order to get ahead. As you get toward the end of this book, you will see me calling out increasing parts of the scientific project that are currently riddled with bias, thereby weakening the entire process and disenfranchising the most vulnerable. Reforming science is no easy task. Reforms are needed throughout the structures that fund, employ and disseminate scientific research. Sweeping this all away will take a concerted effort, but might well be easier in science than in other sectors of society. Nearly all scientists recognise that science needs to be objective and impartial in order to succeed. Their objectivity makes them an ideal group to kickstart a fairer, more transparent and open system. As you will see, there has already been a move to make science more open, and by embracing this transparency the next generation of scientists will reap the benefits, and be on a sounder footing to improve a better scientific society.

### Carl Sagan’s “Baloney detection kit”

or __"How to know when what you smell is bullshit"__

In his book, Demon-Haunted World, Carl Sagan [-@sagan2011demon] provides his take on how science provides scientists with what he terms ‘Baloney Detection’, commonly referred to today as ‘bullshit’. These are well worth recording here as they provide a guiding light for all scientists whether they be novices, emerging or established. They are also easily taught, and the foundation of what we should spread in the Scientific Project.

As humans, we are all susceptible to bullshit, hoaxes, charlatans and contrived deceptions as they are often more appealing than the truth. However, science is about truth seeking, and by applying a set of simple rules, Sagan argues, we can avoid being misled or distracted by ideas that may waste our time and energy by using ‘skeptical thinking’. These rules are useful in your scientific life when considering hypotheses or when building arguments to explain your results. However, as Sagan argues, these simple rules are also useful in your day to day life, and as a simplified example of the scientific method, we can pass them onto all our fellow citizens.

1. You should always seek independent confirmation of results. 
1. Seek out different opinions on the results and consider all possibilities
1. Don’t believe facts simply because they are spoken by an authority on the subject. We know that humans are prone to [confirmation bias](type1error.html), and often exhibit motivated reasoning, especially if they have some vested interest. Even the most experienced have made mistakes in the past and will again in the future. My preferred way of thinking of this rule is to acknowledge that experts are made of ex- (a has been) and -spurt (a drip under pressure).
1. [Use a hypothetical deductive framework](howhypothesis.html). Instead of accepting the first (hypothesis) explanation, think of all the possible reasons why a result might have occurred and then consider how you could (or perhaps you already have in your experiment) disproved each one. You may find that an untested hypothesis to explain the results represents a major caveat that could be tested, or can’t yet be tested. Either way, you need to remain open to the possibility that this explains your results. This leads neatly onto:
1. Don’t get too involved with keeping your chosen hypothesis [opening you to potential cofirmation bias](type1error.html). Remain critical to the possibility that it might be wrong, and provide ideas on how to substantiate or disprove competing hypotheses. If you don’t, others will. 
1. If at all possible use a quantitative method to test your hypothesis. You should already be well on the way to this one inside your PhD, but it is worth bearing in mind especially when it comes to considering testing other competing hypotheses. How could they be quantitatively tested?
1. Arguments ([as we see later](Argument.html)) consist of a chain of statements that follow one from the other. This chain is only as strong as its weakest link, and the strength of each link in the chain (your argumentative statements) must be critically considered by you. 
1. [Occam’s Razor](Argument.html) dictates that given two hypotheses that equally explain the results, the simpler one is likely more plausible. Or put another way, the more links you have in your chain, the more likely that one of them is not as strong as you would like it to be.
1. Your [hypothesis](howhypothesis.html) (or argument) must always be falsifiable. That is, that your test should be able to find that it is incorrect. If your test cannot show that your hypothesis is incorrect, then it isn’t worth much.

Sagan provides plenty of real world examples of each of these arguments, and I’d suggest that you consult his book [@sagan2011demon] if you are interested in following these up. Similar approaches, but written in different styles, are provided by Richard Dawkins in his [-@dawkins2004devil] book: A Devil’s Chaplain, and Julia Galef in her [-@galef2021scout] book: The Scout Mindset.

### Live your life scientific
>__“Science is more than a body of knowledge, it is a way of thinking”__
>Carl Sagan [-@sagan2011demon]

I would encourage you not to stop your objective and evidence based thinking when you leave the office. By applying the rules of science to other sectors of society, you will find that there are many societal norms that could benefit from similar reform. Certainly, a lot of the technology developed in science can be used to help society reform, but scientific thinking will help everyone. As well as bringing the world to scientific thinking, we can take scientific thinking to the world.

### Citizen science
For a long time I struggled with the term 'Citizen science' as I reasoned that many such ventures did not involve citizens doing science. To me, most seemed focussed on outsourcing large tasks to willing public participants: i.e. having citizens do some grunt work. Collecting data, or doing a fraction of the analysis is not science. However, in light of the scientific project, I have changed my mind. All of these projects offer citizens a way to participate in scientific projects in a way that was previously unavailable to them. They promote direct communication between scientists and citizens. In this way, they serve to promote the understanding of science to the general public.

One aspect of this will be to bring more of the population into conducting scientific studies. This can be at the level of funding scientific studies. Crowdsourced scientific studies promote communication between the scientists and the funders to a level that is meaningful to both groups, such that they make up a large scale collaborative research interaction [@uhlmann2019scientific]. Citizen science contributions mean that the collection of data can be shared on a much broader scale, than would be traditionally possible. Of particular interest in the biological sciences is the identification and collection of locality data for biological species (on platforms such as iNaturalist). But increasing numbers of projects are demonstrating novel ways for the public to increasingly participate meaningfully in scientific endeavour [@bonney2014next; @silvertown2009new]. Science is a joyful and spiritual experience for most of us that participate in it. There’s no reason not to spread that joy as widely as possible.

### 2021-04-15 1.04-healthy.Rmd
#First upload of book

## A healthy body to go with your PhD mind {#healthy}

While I’m busy handing out advice, here’s some of the best that I have to offer. If you don’t do it already, then when you start your PhD start regular exercise. There’s an old adage that ‘a healthy body leads to a healthy mind’. I never gave it much attention until I started exercising regularly some 7 years ago. Getting fit was the best thing I ever did to improve my work.

### Being physically active improves concentration

The first thing I found was that I could sit and concentrate on work for longer. Before I got fit, I was constantly distracted and needed to take lots of breaks. Now I find that my concentration is much better and that as my fitness stamina increases, so too does my ability to concentrate.

### Time to think

I used to think that the time I spent exercising was unproductive with respect to work. Now I have begun to realise that this represents some of my best thinking time. I use the time I spend trail running to turn ideas over in my head and especially to think through the logic in arguments, and potential flaws in experimental design. This isn’t to say that I don’t need to put time in at the computer writing it all down, or talking through ideas with colleagues. But I find the meditative time I spend exercising to be especially productive.

### Balancing work with life

Having already told you that doing a PhD might unbalance you mentally, regular exercise is a really good way to make sure that you retain that balance with life [see @hotaling2018publishing]. By life, I'm referring to everything away from your academic work. This might include, family, friends, sport, hobbies (even fishing). If you don't have friends outside your academic life, then it would be a good idea to find some so that you can keep in contact with the 'real world'.

If you already exercise, then you likely know how important this is. But if you don’t then my best advice (above anything else that you might read in this book) is to start as soon as you can. Do whatever is right for you. If you don’t already know, then try out different clubs or activities at your institution. There are normally lots to choose from. Make sure that you schedule regular time for your activity, even if that means that you run in the mornings before field work starts, or you spend your evenings doing tai-chi on the beach after a day at sea doing research on dolphins. Take your exercise with you wherever you go for your studies. Take it as seriously as you do your studies and you’ll find that both will benefit.

Once you have a healthy body, you’ll find that there are all sorts of things that go along with this. The desire to fill it with healthy food resulted, and this later tied in with my interest in the gut microbiome. I now have many fermented foods in my kitchen (much to the horror of my kids), and the improvement of my digestive system. I could go on, but I’ll spare you. Instead, I just want to emphasise that I consider that not only will your quality of life improve, but so will your work if you include regular exercise as part of your PhD regimen.

### 2021-04-15 2.01-Hyp.Rmd
#First upload of book

# How to get started with writing  {#part2}

__Before you start writing__

The writing part of the PhD is often the most difficult for many students. I think that there are several reasons for this. Writing well comes with practice, and most PhD students have had little experience with academic writing - even if they have already completed a MSc. The same could be said of most early career scientists (even after their PhDs), because writing takes practice and we shouldn't expect that it'll come automatically or immediately. In order to get this practice you need to get started, and (again) many students find starting to write challenging. Here is where this guide will help.

However, before you start writing, it's a good idea to understand some of the ways in which academic writing differs from other writing styles. Good planning is the key here, along with knowing exactly what it is that you'll be writing about. That sounds facile, but at the foundation of each scientific study lies the hypothesis, which is central to being able to build the rest of the chapter.

In this part of the book, I start with an explanation of the hypothesis. As you will see in [Part III](hypothesis.html), the hypothesis is central to all of the writing in each chapter or paper, and you need to have a very good idea of how to formulate your research question into a hypothesis before you begin. Then I build on helping you think through lots of different aspects of writing in science: style, citations, writing concisely, scientific names, and lots more. Finally, I introduce you to the idea of using a formula to structure your writing.

Before you start writing in earnest, it’d be good to read this part of the book and make sure that you are familiar with the ideas therein. Later, you may need to refer back to sections once you’ve started writing.

## How to write a hypothesis  {#hypothesis}

When you read about the formula of writing a chapter (see end of part 2) you will see that everything in your chapter or paper revolves around the hypothesis or questions that you ask, normally at the end of your introduction. In this chapter, we will take a look at how to write a hypothesis. This is a sticking point for many students. We are used to using and writing questions and statements in day to day communications, as well as reading popular media. But hypotheses (the plural of hypothesis) only rarely float across our desks. So how do we write one, and how do we know if our hypothesis is good?

In addition to this chapter, there is some good information out there on the web, and it's worth looking at this too: (e.g. [Wikihow](https://www.wikihow.com/Write-a-Hypothesis), [Wikipedia](https://en.wikipedia.org/wiki/Hypothesis), etc.). There's also some less good stuff out there, so read critically.

(ref:Hypothesis) **Generating a good hypothesis for your PhD chapter isn't easy.** A good hypothesis will be invaluable in helping you write the chapter. It is important to start working with hypotheses

```{r Hypothesis, echo=FALSE, fig.align='center', out.width='90%', fig.cap="(ref:Hypothesis)"}
knitr::include_graphics('figures/hypothesis.jpg')
```

### What is a hypothesis?  {#whathypothesis}

A hypothesis is a statement of your research intent. It tells the reader (because just like all of your other written work, it is intended for an audience who will read it), what you plan to do in your research. But there's a little more to it than this. The hypothesis becomes a part of the scientific method if it is testable and (importantly) falsifiable, as well as being informed from previous published work on the subject.

Yes, your hypothesis must be informed by the literature, which is why you spent so much time and effort crafting your introduction to inform your reader of the same. This is also why your hypothesis usually comes at the end of your introduction, because you spend all of the introduction telling your reader about it. There's not much point in writing more after the hypothesis, because once your reader has read that, they are ready to learn about how you went about testing it (in the Materials & Methods). The other important point to make is that the literature should dictate how you write your hypothesis, and the variables that you include. If, for example, you know that temperature is the most important variable, but all of the literature suggests that it is oxygen, you can't ignore oxygen and you should also frame your hypothesis using this variable (you can have more than one hypothesis after all!). In this case, you will also need to provide a sufficient introduction to temperature as a variable to justify its inclusion in your hypothesis. Perversely, your aim is not to prove that your idea is right, but to show that the hypothesis can be refuted.

We try to write a hypothesis that is falsifiable: i.e. you can prove (usually using statistical tests) that it is not correct (or at least show that the likelihood that it is correct is very low). That's why it is conventional to provide the 'null hypothesis' that is the falsified version of the statement, suggesting that there is no relationship between the variables you have proposed to measure. The convention is to label this 'null hypothesis' H0, while the 'alternative hypothesis' (the one that says your variables are related as you suggested) is written as H1. When you formulate your hypothesis, it is traditional to write your alternative hypothesis to indicate the directionality of your tested variables. This way the reader can simply imply that the null hypothesis is when there is a relationship, but this will need to be stated if the null hypothesis is more complex.

Karl Popper [-@popper2005logic] was the philosopher who proposed that without being able to refute or falsify a scientific problem, it ceases to be scientific. This is the reason for our null hypothesis. If the null is not available as a possible outcome, then logically, there is no science.

> Karl Popper [-@popper2005logic]:  
> **"...it must be possible for an empirical scientific system to be refuted by experience". **

Most importantly, your hypothesis must come first, before you do the experiment or study. Hence the reason why this chapter comes at the start of part 2. Setting the hypothesis after the work is already done is fraudulent, and goes against the scientific method. Obviously, it isn't fair to pose the hypothesis once you already know the answer. This is why there is so much emphasis put on formulating your hypothesis during your research proposal. Getting it right will determine what you do and how you test it. If you think of an extra hypothesis that would be really useful to test once you've already done your study, you can conduct a post hoc test, but this should have more stringent levels of statistical assessment.

Writing a hypothesis isn't easy, but it is essential and once you've understood what to do most of the rest of what you are writing for should make sense.

### What a hypothesis isn't

It is not a question and so should never have a question mark after it.

It isn't really a simple prediction: if this then that. You will see on the internet that hypotheses are explained in this simple predictive framework. I say that a hypothesis is not a simple prediction because it lacks the mechanistic and scholarly aspect of a good hypothesis, which is what we want to achieve.

### A formulaic way to start writing your hypothesis:

__"If. then. because."__

Later in the book, I will emphasise that you must have introduced all the variables that you plan to use to test your hypothesis in your introduction. This usually comes in the second paragraph of your introduction, where you emphasise the utility of the dependent variable/s (what you are planning to measure) and your independent variable (what you will manipulate). Both of these variables should then feature in your hypothesis. Next, by paragraph four you will have identified the problem that you are interested in tackling. In addition, your introduction will provide all of the pertinent literature that has relevance to this hypothesis, giving the all important context.

A simple way to consider making your hypothesis is to adopt an **"If. then. because."** construction where you add in your problem statement using your independent variable after 'if', and your prediction using your dependent variable after 'then', and finally the expected mechanism after 'because'. Using our example above with the "If. then. because." construction, we would say: "If environmental temperatures in which tadpoles develop are increased then tadpole development rate is faster because they follow the classic metabolism of ectotherms". Both independent variable (temperature) and dependent variable (tadpole development rate) are present in this hypothesis, and the predicted relationship between them is clear. In addition, the causal mechanism is stated. You can watch a video about using the "If. then. because." construction here, or read more here. I say that this is a formulaic way to start writing your hypothesis, because it usually ends up as an inelegant statement, which can be better refined for a reader. A citation for your stated mechanism might also help clarify exactly where the justification for this comes from.

Mechanisms (or causal explanations) fall into three main areas: endogenous, exogenous and evolutionary [@allen2017scientific].

#### Endogenous causal explanations

Endogenous causal explanations focus on the mechanisms happening inside an organism, such as physiological processes, hormones, reproductive state, etc.

#### Exogenous causal explanations

Exogenous causal explanations concern mechanisms that are outside the body of individuals. Common exogenous mechanisms are climatic factors (temperature, humidity, precipitation, etc.), or may relate to the availability of food, predators or mates.

#### Evolutionary causal explanations

These mechanisms have evolved through time, and often relate to exogenous mechanisms triggering endogenous processes over multiple generations.

Note then that the above mechanisms are not mutually exclusive in their nature, and it may be useful to combine different approaches within biology to ask hypotheses across all of these levels. Mechanisms in biological sciences are rarely simple or act on multiple organismal levels, so designing a controlled experiment in order to test a specific mechanism thoroughly can be very demanding. In other words, can you be sure that the cause is really responsible for the effect that you are measuring?

A good hypothesis will often take an existing hypothesis further, to try to better refine the knowledge on a subject. Hence, it is perfectly acceptable to state that you are building on existing hypotheses (and giving the appropriate statement) when making your own.

### Teleological versus Causal Hypotheses

A teleological argument refers to the reason or a purpose of a particular process. For example, you may measure vertical migration of water fleas and suggest that diurnal migrations are made because the water fleas want to avoid predation. This is a teleological hypothesis because you are suggesting that the reason behind a process is the desire by water fleas to avoid predators. Although a reduction of predation may be a consequence of vertical migration in water fleas, each water flea does not think about predation and then starts it’s upward movement as a result. A common mistake made in biology is to apply teleological arguments to processes that have no purpose or region. Evolution is often mistakenly suggested to have a purpose (e.g. to evolve to a more advanced state), but in fact evolution is not a goal-orientated process. There is no end-point to evolution, and evolution did not start in order to meet some predetermined form or function. On the other hand, a causal hypothesis focuses on the factors about A that cause B.

You should have realised that biologists are principally interested in causal hypotheses, because most mechanisms that are studied in biological sciences have no predetermined goal. If you are a behavioural ecologist, then you will need to be particularly aware of these two types of hypotheses, and when teleological explanations may be appropriate: many types of behaviour are goal orientated.

### How to evaluate your hypothesis

Once you've written your hypothesis, how do you decide whether or not it is good? To do this, you might think that you need plenty of experience (and yes, that does help). But really, you just need to look for the elements that are discussed above. So once you've written your hypothesis, try to objectively answer the questions below (for more see [here](https://youtu.be/9KoSGQlIFkM)):

- Is there a clear prediction (if. then. statement)?
- Does the prediction use independent and dependent variables correctly?
- Is the mechanism supported by the literature?
- Is the hypothesis testable/falsifiable?
- Does the hypothesis use concise wording and precise terminology?

If your hypothesis meets all of the criteria above, then you've done a good job!

Probably one of the hardest issues that you will face in biological sciences is to determine whether your dependent variable is reliant only on your independent variable of choice. For example, a lot of variance in biological sciences will relate to the climate (especially with global change studies), but if your independent variable is temperature, this means that you will need to keep all other climatic variables the same. That is, if temperature is your independent variable, it is the only variable that can change in your experiment. This type of experiment is challenging as temperature often affects other variables (especially that they may vary in an unpredictable way). As soon as you have more than one independent variables, you can no longer test your dependent variable because you don't know which independent variable it is reacting to.  Isolating variables is notoriously difficult, especially when we move from the laboratory to the field. You will need to think very carefully about what variables other than those of interest are potentially impacted by your experimental design. If you cannot control for them, this will likely mean that you need to change your hypothesis, or change your experimental design. If you are unsure, then I would encourage you to look carefully at the experiments that others have conducted in the literature.

### 2021-04-15 2.02-TypeI.Rmd
#First upload of book

## Being aware that you can get it wrong {#TypeI}

**Type I and Type II errors **

With all the will in the world, when you are testing your hypothesis using statistics there is a chance that you will accept your alternative hypothesis when it is not valid. This is known as a ‘false positive’ or a Type I error (see Figure \@ref(fig:TypeITypeII)). It is also possible that you will accept your null hypothesis when you should have accepted your alternative hypothesis, also known as a Type II error, or a ‘false negative’. While it won't be any fun to get a Type II error, as scientists we should be more worried about Type I errors and the way in which they occur. This is because following a positive outcome, there is more chance that the work will be published, and that others may then pursue the same line of investigation mistakenly believing that their outcome is likely to be positive. Indeed, there are then lots of ways in which researchers may inadvertently or deliberately influence their outcomes towards Type I errors. This can even become a cultural bias that then permeates the literature.

Humans have a bias towards getting positive results [@trivers2011folly]. If you’ve put a lot of effort towards an experiment, then when you are interpreting your result you might feel motivated towards your reasoning making you more likely to accept your initial hypothesis. This is called [‘motivated reasoning’](https://youtu.be/w4RLfVxTGH4), and is a rational explanation why so many scientists get caught up in Type I errors. This is also known as a confirmation bias. Another manifestation of this is publication bias which also tends to be biased towards positive results, with as many as 84% of studies supporting their initial hypothesis [@fanelli2010positive]. Put together this means that scientists being human are more vulnerable to making Type I errors than Type II errors when evaluating their hypotheses. Although there are deliberate or inadvertent ways of making of Type I errors, here it is important to understand that simply by chance you can make a Type I error: accepting your alternative hypothesis even when it is not correct. To minimise this possibility to a reasonable level, most biologists set α (alpha) to 0.05 as an acceptable risk that they will encounter a false positive (Type I error).

(ref:TypeITypeII) **Type I and Type II errors.** In this figure, the columns refer to the truth regarding a hypothesis, even though the truth is unknown to the researcher. The rows are what the researcher finds when testing their hypothesis. The blue squares is what we are hoping to achieve when we set and test our hypothesis. The grey squares may happen if the hypothesis we set is indeed false. The other two possibilities are the false positive Type I error (red), and the false negative Type II error (black). In the table it seems that the chances of getting one of the four outcomes is equal, but in fact this is far from reality. There are several factors that can change the size of each potential outcome to testing a hypothesis.

```{r TypeITypeII, echo=FALSE, out.width='90%', fig.cap="(ref:TypeITypeII)"}
knitr::include_graphics('figures/TypeITypeII.png')
```

In the following figures (Figures \@ref(fig:HighlyUnlikely), \@ref(fig:Unlikely) and \@ref(fig:Likely)), you see the outcome of 1000 hypotheses. You could think of these as outcomes of 1000 attempts at testing similar hypotheses in a global scientific effort by individual scientists. The power of the analysis is set at 40%, and α is set at 0.05. It is important to realise that α applies to all the tests that are made (whether or not the hypothesis is actually correct). This means that when α is set at 0.05, 5% of the time scientists will find that their hypothesis is correct whether or not this is actually true is, as we shall see, an interaction of several variables.

(ref:HighlyUnlikely) **The likelihood that highly unlikely hypotheses you test are accepted.** This figure has been redrawn after figure 1 in Forstmeier et al [-@forstmeier2017detecting]. This is a graphical representation of an argument first made by Ioannidis [-@ioannidis2005why].

```{r HighlyUnlikely, echo=FALSE, out.width='90%', fig.cap="(ref:HighlyUnlikely)"}
knitr::include_graphics('figures/HighlyUnlikely.png')
```

### Changing the likelihood of a hypothesis

The difference between Figure \@ref(fig:HighlyUnlikely) and Figures \@ref(fig:Unlikely) and \@ref(fig:Likely) is the likelihood that the hypotheses are correct changes. In the first one (Figure \@ref(fig:HighlyUnlikely)) we see highly unlikely hypotheses that will only be correct one in a hundred times. The blue squares denote the proportion of times in which the hypotheses are tested and found to be true. The black squares denote false negative findings, i.e a Type II error. The red squares denote false positive findings, i.e. a Type I error. Because the hypothesis is highly unlikely to be correct the majority of squares are light grey denoting that it was correctly found to be untrue. Because α is set at 0.05, we can expect that 5% of the 990 incorrect hypotheses will give us a false positive (49 or 50 times). Although it might seem unlikely that anyone would test such highly unlikely hypotheses, there are increasing numbers of governments around the world that create incentives for researchers to investigate what they term blue skies research, which might be better termed high risk research or investigations into highly unlikely hypotheses. The real problem with such hypotheses is that you are more likely to get a Type I error than actually find that your hypothesis is truly correct.

(ref:Unlikely) **The likelihood that unlikely hypotheses you test are accepted.** In this figure we see a scenario of unlikely hypotheses that are found to be correct approximately one in 10 times. Now we see that the possibility of committing a Type I error is roughly equivalent to a Type II error and to the probability of finding that the hypothesis is truly correct. Thus, if your result comes out positive, you are unlikely to know why.

```{r Unlikely, echo=FALSE, out.width='90%',  fig.cap="(ref:Unlikely)"}
knitr::include_graphics('figures/Unlikely.png')
```

(ref:Likely) **The likelihood that likely hypotheses you test are accepted.** In this figure we see a scenario of likely hypotheses that are found to be correct approximately every other time.

```{r Likely, echo=FALSE, out.width='90%',  fig.cap="(ref:Likely)"}
knitr::include_graphics('figures/Likely.png')
```

Now imagine some groundbreaking research (probably published in a high impact factor journal) that finds a highly unlikely (but presumably desirable) hypothesis is correct (our first scenario of 1 in 100). Researchers from other labs try to repeat the experiment, and some of them also find that they get positive results and get these published. None of the negative results get published, or the investigators don’t bother to submit them and move onto other areas of science. The positive results are written into text books and future generations of scientists test these types of hypotheses expecting them to be correct (our last scenario of 1 in 2). When the negative result is produced, a reasonable assumption is that a Type II error has been found, and that these scientists need to increase the power of their tests, inevitably increasing the cost of the experiment. But there are plenty of other ways that researchers could end up finding that their negative result is in fact positive, especially if this is what they are expecting. Simmons et al [-@simmons2011false] refer to this as the ‘researcher degrees of freedom’ in which researchers inadvertently or deliberately end up getting positive results because of the way in which they treat either their hypothesis or analyse their data. It is easy to see how a false scientific culture can be erected in these mistaken beliefs, and it is important to be aware of this possibility.

### Increasing the statistical power

In all of these figures the power of the analysis is set at 40%. The statistical power of any analysis depends on the sample size (or number of replicates) you're able to use. Some research has suggested that in most ecological and evolutionary studies this is actually more like 20% [see references in @forstmeier2017detecting]. What is important to notice in the next figure \@ref(fig:20PowerHyp) is that when we change the power of the analysis (in this case from 40% to 20%) we influence the proportion of Type II errors over finding that the hypothesis is correct. While the overall numbers of Type I errors does not change, if your analysis tells you to accept your alternative hypothesis, there is now a 1 in 5 chance (20%) that it will be a false positive (Type I error).

(ref:20PowerHyp) **The likelihood that likely hypotheses you test are accepted when you increase the power.** Here we see the scenario in which the hypotheses are quite likely to be correct one in two times. Now we can see that the possibility of creating a Type II error is highest. Next the blue squares show us the chances that we find the hypothesis is truly correct. Lastly, there's only 11% chance of a Type I error.

```{r 20PowerHyp, echo=FALSE, out.width='90%', fig.cap="(ref:20PowerHyp)"}
knitr::include_graphics('figures/20PowerHyp.png')
```

What you should see when you look at these figures is that there is quite a high chance that we don't in fact correctly assign a true positive hypothesis. There's actually much more chance that we will commit a Type II error. Worse the more unlikely a hypothesis is, we are increasingly likely to commit the dreaded Type I error.

From the outset we should try to make sure that the hypotheses you are testing in your PhD are very likely to find positive results. In reality, this means that they are then iterations of hypotheses that are built on previous work. When you choose highly unlikely hypotheses, you need to be aware that this dramatically increases your chances of making a Type I error. The best way to overcome this is to look for multiple lines of evidence in your research. Once you have the most likely hypothesis that you can produce, you need to crank up your sampling so that you increase the power of your analysis, avoiding Type II errors.

In biological sciences we have more options than many scientific disciplines to investigate potentially false culture beliefs by using different model species, and looking at problems through different mechanisms. Presenting complimentary results from multiple lines of evidence within your thesis is an extremely powerful way to build evidence for your hypothesis. However, this requires creativity of experimental design as well as an open mind. It may mean that you have to draw upon the most fundamental of scientific abilities: to be comfortable with being wrong. It may also require dogged determination in terms of publishing negative results against the currently prevailing culture. As we shall see (in parts 4 and 5), publishing has another set of challenges.

If your appetite is piqued by this subject, be sure to read the references cited, and read around this subject.

### 2021-04-15 2.03-NoHyp.Rmd
#First upload of book

## What happens if you don’t have a hypothesis? {#nohypothesis}

So far we have concentrated on thesis chapters or manuscripts that have hypotheses to test. I have emphasised that there is a need to focus all of the sections of your writing to helping frame and understand the hypothesis (in the Introduction), explain your approach in collecting and testing the variables (in the Materials and Methods), provide the results of your tests (in the Results), and respond to the hypothesis (in the Discussion) together with placing it in context. But what, I hear you cry, if you don’t have a hypothesis to test?

There are lots of reasons why manuscripts don’t have hypotheses, and the chances are that in a PhD thesis (for example) you will have at least one chapter where there isn’t a hypothesis tested. So, given all the emphasis above, what can be done?

### Central problematic
If you have a particular problem that your research is trying to solve, this may result in a new methodology or perspective rather than testing a hypothesis. This work is still suitable for a journal manuscript (and there are specific journals dedicated to new methodologies). Instead of introducing your hypothesis, you can simply introduce the problem that your methodology applies to. A new methodology will probably best fill an existing gap, and so your introduction will likely point out what this gap is, provide evidence for why existing methodologies don’t fill the gap, and outline the variables in your novel approach.

Similarly, if you have a question but not a hypothesis, that your research tries to answer, you can take exactly the same approach to introduce your question.

Hence, no matter what your central problem is, you still need to identify this and put it at the heart of your manuscript so that all of your sections address this point. Of course, there may be more than one, and if so you will need to clearly articulate each one and explain how they are related to each other. If you are unsure about how to move forward, the best thing to do is to look to the literature for other examples of what you’re trying to do. Someone has very likely done something similar before. Don’t forget to chat about it with your advisor, as they may have ideas about where to look.

### Does it matter that you don’t have a hypothesis?
Having a hypothesis makes your work stronger than if you only have a question or prediction. In each case, while your results might help answer your question or confirm your prediction, they won’t test the mechanism (so that you can’t explain why you have the results that you do), and they may not be falsifiable/refutable. Of course, Popper would argue that if you don’t have a null hypothesis, then your methodology isn’t scientific.

To me, it is important for you to know exactly what your study is before you start writing about it, or even before you do the experiments. It is essential for you to frame the context of your study. The central reason why you are writing the study will lie at the heart of the manuscript. In the formulaic approach, it will be verbalised at the end of the introduction. All parts of the methodology will explain how you approached this central aim. So if you can’t pin down what this is, then make sure you discuss it with your advisor.

### Avoid HARKing 
Hypothesising After Results Known (HARKing) is the practice of creating your hypothesis once the results are in and you’ve done some preliminary data analyses to see what is significant [@forstmeier2017detecting]. HARKing has become prevalent in science because of the confirmation bias by journals (only accepting papers that can statistically accept the alternative hypothesis), especially those with higher impact factor.

### 2021-04-15 2.04-BigIdea.Rmd
#First upload of book

## What's the big idea? {#bigidea}
In the previous chapters, I've talked about the importance of having a hypothesis, and building that hypothesis in a logical framework within the introduction (see part 3). The introduction serves to inform the reader about why this particular hypothesis was chosen, introducing both the response and determinate variables, as well as the presumed mechanism by which the hypothesis can be falsified (or upheld).

### The funnel introduction 
But so much of what you read doesn't have any clear hypothesis, and there doesn't seem to be any thought about building on bigger concepts or ideas. In this chapter, I suggest that this might not be true only of papers on herpetological invasions (my specialist area), but also of herpetology in general (although I concede that some areas, such as herpetological physiology are actually quite good). Further, I contend that using these wider hypotheses or theories would actually be good for the authors concerned, as it would likely garner them a wider audience. Moreover, a greater number of biologists might come to realise how valuable reptiles and amphibians are as models in biology.

(ref:Introfunnel) **A framework for writing your introduction starts with the big idea and ends with the hypothesis.** The funnel introduction is so called because it resembles a funnel in which you channel the reader from their own interest in a wider subject area into the relevance of your particular study.

```{r Introfunnel, echo=FALSE, fig.align='left', out.width='90%', fig.cap="(ref:Introfunnel)"}
knitr::include_graphics('figures/Introduction.png')
```

### So where would we find all of these big ideas?
There are quite a few papers that synthesise hypotheses in various areas of biology. Here I provide two to get you started. Ask your advisor about relevant reviews of hypotheses in your area of biological sciences. The first is by Mark Velland [-@vellend2010conceptual] on theories in community ecology. The next is by Jane Catford and colleagues [-@catford2009reducing] on hypotheses in invasion biology, but I encourage you to look for more up to date versions. Each of these papers will give you a list of big ideas, together with the citations for seminal papers that have built them. You will note that many of these theories are very old with some dating back to Darwin. You can think of the way that such ideas are structured as a hierarchy of hypotheses (Figure \@ref(fig:JeschkeHagar61)).

(ref:JeschkeHagar61) **A hierarchy of hypothses can be used to help determine your Big Idea.** A useful structure for thinking about how hypotheses are structured was presented by @heger2018hierarchy in what they termed the _Hierarchy of Hypotheses_

```{r JeschkeHagar61, echo=FALSE, fig.align='left', out.width='90%',  fig.cap="(ref:JeschkeHagar61)"}
knitr::include_graphics('figures/JeschkeHagar61.png')
```

Of course, there are many ways to approach and test these theories, but if you don't know about them, your work may actually make a considerable contribution to upholding or refuting them, but go totally unrecognised. When the significance of your work isn't realised, it's unlikely that it'll be widely read and cited.

Let's face it, if all the effort of the work that we put into papers is just going to get buried, then is it really worth it? The work that we do is also expensive, so making it as relevant as we can to as wide an audience as possible is something that we should be concerned about.

So, I encourage you to **stand on the shoulders of giants** (Figures \@ref(fig:Googlescholar) by using big ideas in your introduction. Make sure that the data that you collect can actually be used to respond to some of these big ideas. Then make sure that you cite them, giving them the importance that they deserve (yes, even as key words) so that others can find your work, and you might even find that one day, your work has shoulders that are broad enough for others to stand on!

**The take home message**

Reading the literature can really expand your mind and broaden your horizons. When undertaking a literature review [or when reviewing a paper], take the time to think about not only what has been tested, but what could have been. Make a list of theories and hypotheses in your own field and try to rank them in a hierarchy.

### 2021-04-15 2.05-paragraph.Rmd
#First upload of book

## Writing a paragraph {#paragraph}

In the following chapters, I will explain how to write a scientific paper in a formulaic style, as well as the advantages that this has over starting from scratch each time. The suggestion is that once, you sketch out your paper in a 'bullet-point' outline. This outline will give you an idea of the subject in each paragraph, as well as how the paragraphs fit in with each other. The other suggestion was to flesh out this outline by assigning relevant citations to help you remember those main ideas.

Writing paragraphs has a skill all of its own, and the aim of this chapter is to go over the basics of how to put a paragraph together. Other people have also written about this, and you can read an example here.

**This paragraph starts with a topic sentence which contains the principle idea that citations are needed for examples.** The following sentences put this idea into the context of your study. This should be based on examples from existing literature with the corresponding in-text citations provided (Smith et al 2017; Jones 2018). You can use examples in your paragraph to make your point. For examples, Smith and Jones (2017) carried out a study that showed that examples without citations have no credibility for readers. And this was backed up by another study by Jones and Smith (2018), which provided further evidence for the need to provide examples. Once you’ve provided your evidence, and introduced any covariates that you need to, like the use of literature for examples, then you are ready to sum up and link. *After reading this paragraph, the reader will know why citations are needed for examples, and they’ll be ready to read more about the topic of writing a paragraph in more detail.*

In the paragraph above, the topic sentence is bold, the supporting sentences are normal font, and the clincher is in italics.

### Topic sentence
A topic sentence allows the reader to understand quickly the idea/topic you are putting forward in the paragraph. It must be in the context in which you are going to develop the same topic. There’s no point in just mentioning a topic in passing or using it in a different way than you will later.

Make your topic sentence relatively simple. Don’t be tempted to add multiple clauses. If the topic sentence is too complex, you’ll lose your reader right at the beginning of the paragraph.

### Supporting sentences
Supporting sentences convey all the relevant information to the reader. They are going to be statements that are well cited, showing readers where the original ideas came from. Be sure to keep these sentences on topic, and regularly refer back to your outline to make sure that you keep to the original objective of the paragraph. These are the meat of the paragraph and it’s really important to get them right.

If you are writing about differences, then state which way the difference is. For example, if bullfrogs are larger than leaf litter frogs, then say this. Telling your audience that something is different than something else only ends up leaving them guessing about the way in which the things are different. There’s no point in drip dripping this information through. Set it down in as little space as possible so that your reader doesn’t get bored.

Sentences within the meat of the paragraph interact, and there’s a great example to show how this is done here. These interactions usually dwell around pertinent variables (such as those that you are going to deal with in your paper). By using the same or similar words within the paragraph, you are able to demonstrate to the reader how those different points interact.

While repeating the names of the variables, or their abbreviations, can be helpful, repeating descriptive words becomes quite tedious to the reader. Repeatedly reading the same ideas repeated over and over quickly bores the reader. Bored readers are less likely to take in what you write, even if you are repeating the same information (yes, I'm repeating myself - boring isn't it). It will also give the reader the impression that your vocabulary is very limited. These days you can do a quick right click on a highlighted word to get a drop down list of synonyms (Figure \@ref(fig:synonym)). This can allow you to go back and replace your repeated word with: recurrent, frequent, recurring, repetitive, constant or continual. That’s more than enough to spice up the paragraph. However, if you’re not sure whether a word is correctly replacing another, ask a friend to read it.

(ref:synonym) **How to quickly get a synonym to make your paragraph more diverse.** An example of the synonym function in Microsoft Word (other wordprocessors are available) - right click on the highlighted word.

```{r synonym, echo=FALSE, fig.align='left', out.width='50%',  fig.cap="(ref:synonym)"}
knitr::include_graphics('figures/synonym.png')
```

Your supporting sentences are best built in the formula of an [argument](argument.html).

### The magic of threes
We can learn a lot about writing convincingly in science by borrowing from storytelling, and one of the greatest lessons in storytelling involves the magic of threes [@angler2020telling].

In putting together your paragraph you can use the magic of threes to break your dialogue into three parts: thesis, antithesis and synthesis (a, b and a+b). Think of it this way: there’s a standard explanation of how a phenomenon that you want to investigate interacts with the variable of your choice (thesis). Then there’s some contradictory evidence that suggests that the standard explanation is not necessarily correct (antithesis). Lastly, there is your approach to investigate exactly how the variable and phenomenon interact which should explain (or at least advance our understanding) of how the variable and the phenomenon are related (synthesis). It may be that this construct should be fleshed out into three paragraphs instead of three sentences within a paragraph.

You won’t be able to construct every paragraph with the magic of threes, but once you have learned this rule, you’ll be able to use it to help emphasise important points in your manuscript that you want the reader to remember as they move forwards. What if your list is two instead of three? Well then you’ll lose the magic, but maybe you can think it through it again? If you have two, then you normally have three by means of the combination of the two (a, b and a+b).

What follows is more advice about telling a good story: using an example and avoiding being boring: i.e. replacing those long lists with a maximum of three items.

### Use an example
Examples are a very powerful way of conveying ideas in a short amount of space. Don’t replace your paragraph with an example, but do use an example if it shows the reader just what you want. You should be able to do this in a sentence (or two), but if you’re tempted to go on, it’s probably not a good example. Because the first paragraph sets out ideas, it's unlikely that an example there will be a good idea.

### Avoid lists
I’m not a big fan of paragraphs which are simply a long list with little or no thought offered. The worst ones are where there are so many citations along the way that it’s really hard to pick out what is a sentence and what is citation. I understand that it’s important to show precedent and that there is merit in showing how widespread an idea is over taxa or in different disciplines. You never find these lists in journals where words and/or citations are limited, which suggests that you can dispense with them.

Instead, you can avoid long lists by pulling them into themes. Once you’ve managed to get three themes, you’ll find that you can work them into your paper time and again. For example, each theme might well be worth a separate paragraph or section in the discussion.

### The last sentence of the paragraph: the clincher
Once you’ve conveyed all of the information that you planned to impart in your outline, it’s time for the last sentence. This should conclude the evidence that you’ve provided on your topic. Try not to make it lame. For example: “This shows that little work has been done.” One of my pet hates is authors who insist on telling us that “little work has been done”, as if this was a justification for doing anything. Instead, make it a real clincher about why the topic is important, or how and why you will tackle it. Instead (or as well if you can), you may want this sentence to link on to another paragraph (topic), especially if flow is important at that part of your outline (see below). If you have just constructed a paragraph with thesis, antithesis and synthesis, then your clincher might be simply to underline that your approach in the paper is to investigate the synthesis. Either way, make sure that your last sentence is on topic, and one that sticks in the readers’ minds.

### Above all - read it!
Your paragraph is not finished until you've read it. Reading is an essential part to writing that cannot be emphasised enough. Does your paragraph say what you planned in your outline? If your paragraph and any other text doesn't make sense to you, it sure won't impress anyone else. If you can't bear to read it through immediately, then do it after you've written two or three paragraphs. I suggest that you don't wait until you've finished the manuscript. Rather get the text right as you go along.

### How does the paragraph fit into the flow?
So now we’ve gone over the formula, it’s time to take a step back and look again at the paragraph in the context of your outline. Remember that the paragraph represents a single subject, but that it is still just part of the manuscript as a whole and you need that to flow from beginning to end. This means that it’s not enough to write each paragraph in isolation, but to think of the way in which they link together as a whole, especially when you are trying to make a logical argument.

You might get a comment from me about missing logic in the argument or a lack of flow in a paragraph. This usually comes from not having a good outline (see below). You should be using the outline to provide the logic in the flow of your argument, flesh it out with the relevant references. If you already know this well, then writing the paragraph will just be a question of styling the outline. Remember that there is a purpose to your paragraph. You have an argument to make the reader understand. Don't be tempted to include information that is distracting or moving away from the flow of that argument.

I am very fond of ending a paragraph on a linking sentence. Essentially, this shows how two ideas are connected in the last sentence. This really helps with getting the flow of an introduction or discussion, but linking sentences are not always the best way to end a paragraph. Sometimes there’s no option but to change the subject completely, and then you should go for the clincher idea (see above). For example, you may want to end the paragraph by seeding a new twist on the paragraph’s idea.

### Seeding ideas
The introduction sets out the established literature in order to put your study in context, but your discussion provides you with an opportunity to present new ideas, or to turn and twist existing ideas in a new way. Once you’ve got a good idea what these are, I like to seed the introduction with hints as to what these might be. Sowing seeds early in a manuscript will provide the reader with hints as to where you are going. Writing these seeds as questions is a really good way of sowing them into an introduction. You can then go on to answer them (if only partly) in the discussion. Beware though, there’s no point in asking a major question in the introduction to which your data has no relevance!

### Breaking the rules
Just as in the other instructions on formulaic paper writing, when writing a paragraph you shouldn’t feel totally constrained so that you can’t break the rules. Breaking the rules can set you free, and much of what you read that really stands out will do this. However, it’s much easier to break the rules and get it wrong, than break them and get it right. The idea of this chapter was to help you get started, not to communicate with those who are already writing great stuff. So if you're already great, don't break it by doing any of this!

### Some common paragraph mistakes
- Don't start your paragraph by linking back to the previous point. Start with your new topic sentence. If you aren't ready to change the topic, you aren't ready for a new paragraph.
- Don't start the paragraph with an unsubstantiated or over-inflated claim (keep it calm and within the bounds of what you are citing). Similarly, don't start with a trivial caveat. And don't start with a list or long string of citations.
- Don't let your paragraph get too long. It is said that 250 words is the threshold. If you haven't covered what you wanted to in your outline within 250 words, then you need to rethink your outline for that paragraph! Definitely don't split a paragraph that's too long into two!
- If you started without an outline, it's never too late to go back and write an outline. Simply write down what you are trying to say (as a set of bullet points), and why (see below for an example).
- Your paragraph should have more than one citation in it. If you find yourself citing the same paper several times in the same paragraph, then there's something wrong with your outline, and the way that you are reading and citing papers (see below). Remember that what you write is about making your own argument from the literature, and not copying others arguments.

The English grammar rules apply when writing science. I can’t go into these in detail here as they are legion and complex. However, here are a few mistakes that are often made by those writing in science:

- You should not start a sentence with an abbreviated genus. Instead write the genus name in full.
- You should not start a sentence with a numeral. If you have to use the number as the first word, write it out. Otherwise, rearrange your sentence. 
- Make sure that acronyms are spelt out on first use (this may be needed for each section)

### 2021-04-15 2.06-argument.Rmd
#First upload of book

## Construct a logical argument in your writing {#argument}

Writing is not straightforward. Your objective is to communicate with a reader, someone who you’ve likely never met and will never meet. You need to communicate highly complex information. But more than that you need them to see things as you do. You need to provide them with your reasoning and your argument, and have it make sense to them; preferably to the point where they agree with you.

To communicate, you need to start from common ground. The beginning of your introduction starts with the most general concepts in the context in which you are writing. The context depends on your audience, and this in turn relates to the particular journal that you are writing for. Even if you are writing a thesis, you should pick out a target journal for each of the chapters. Once you’ve established the common ground, you need to carry the reader towards the hypothesis or question that you propose. The easiest way to do this is to make use of a logical argument:

A series of statements that introduce a starting premise, provide evidence for and against that premise (perhaps adding in an example that makes your point), point out what missing information would allow reaching a better understanding of said premise, and logically conclude that what you are doing is going to fill this gap.

This logical argument style is most prominent in the early introduction of your manuscript, although the entire introduction could be seen as one long logical argument, with a few smaller more precise arguments being thrown in along the way. You might also use a logical argument in your discussion to explain how you deduce certain inferences from your results, or provide a logical extension for a future study.

The following science argumentation model (Table \@ref(tab:argument)) is modified from Cope et al [-@cope2013science]. Try using this model to guide your writing of a paragraph of your introduction (or discussion) where you attempt a logical argument.

Table: (\#tab:argument) **A framework for a scientific argument.** Modified from Cope et al (2013), this model argument can be followed to provide as a way of getting started.

| *Structure*  | *Explanation*  |
| ------------ | -------------- |
| A position statement / question / hypothesis / theory / problematic | This could be the big idea in your manuscript, or one of several competing concepts that you are introducing. The context in which you are writing might mean that this idea theory needs no introduction (e.g. the theory of evolution in the journal Evolution), but you need to be confident that your audience will understand what you are proposing. Remember to cite the person who came up with the idea. |
| Claim 1 | One potential explanation or interpretation of the original idea. Evidence: Literature that agrees with this interpretation (could include an example) Reasoning: Your justification that links the evidence to the claim or interpretation |
| Claim 2 | Another potential explanation, of the previous interpretation of the original idea | *Evidence*: Literature that agrees with this interpretation (could include an example) *Reasoning*: Your justification that links the evidence to the claim or interpretation |
| Counter - claims | Other possible interpretations or counter claims. |
| Conclusions | Evaluation: your judgement on weighing up the evidence for the idea. |

If your logical argument doesn’t arrive immediately, don’t force it. Give it some thinking time to let it settle in. I find that logical arguments develop best when I’m not sitting at my desk writing. They come when I’m reading other papers (especially those with well written arguments), taking a deep, hot bath, or when I’m out running in the mountains. If you are more social, talk through your logical argument with others, perhaps over a cup of coffee. It really helps to keep the ideas turning in your head. Don’t cut yourself off from work thoughts when you are away from work.

There are other shorter forms that might suit you better. Consider another classic form of the scientific argument: “compare and contrast” which allows you as a writer to quickly familiarise your reader with some key examples. Another sentence structure to consider is explaining cause and effect. For practical wording examplesof both of these classic arguments, [see part 2](phrasebank.html).

### Arguments in Latin
From time to time you may come across comments about arguments that you’ve made in your work made in Latin. Why do people use Latin and not simply say it in English? What do these phrases actually mean? While you may think that these people are being pompous or trying to demonstrate some scholarly superiority by quoting Latin at you (and you might be right), there is some important reasoning to consider both in the meanings of these argumentum (as they are known in Latin), and the fact that they are as old as Roman scholars, and often much older. This then is the point, that some arguments are faulty, and that this has been known for thousands of years.

Here I have provided some of these arguments, and the reasons why they may appear in your thesis in writing biological sciences.

_Ad hominem_ - “To the man” means that the argument attacks the person and not the argument.

_Non sequitur_ - “It does not follow”: suggests that one statement in your argument is not linked to the next. This is also known as a logical fallacy, a formal fallacy or a deductive fallacy (for a full list of fallacy statements and their explanations, see here).

_Post hoc, ergo propter hoc_ - “it happened after, so it was caused by”: mistaking cause and effect simply because one thing happened after another. The two events may have been completely unrelated.

_Reductio ad absurdum_ - “reduction to absurdity”: attempts to establish a claim by comparing it to an absurd opposite.

_Novacula Occami_ - "Occam’s razor": The idea that the simplest hypothesis, or argument is most likely to be correct. Another way to think of this is if you have two models that you test are equally likely (e.g. within 2 dAICs), the one with the least number of variables should be considered superior. This is also known as the priciple or law of parsimony: _lex parsimoniae_.

Likewise, there are some arguments in English that you will hear time and again. I take the opportunity to also list these here:

**Correlation does not equal causation** - That two events or variables are related when in fact they are co-incidental.

**Absence of evidence is not evidence of absence** - The idea that positive evidence of the existence of a relationship is superior to a lack of the same evidence. This is also known as a Type II error (false negative), where the lack of finding a significant relationship does not lead to a conclusion that the relationship does not exist. Later we discuss how a [lack of finding of a significant result](transparency.html) leads to confirmation bias in science.

**Observational selection** -"The root of all superstition is that men observe when a thing hits, but not when it misses." Francis Bacon [-@bacon1620novum]. This refers to the cherry picking of studies that support your hypothesis while omitting those that do not support it. Later we see how this can be abused when writing a review or even the introduction to your chapter.

**Straw man** - Erecting a false or simplified statement in order to make it easier to attack.

**False dichotomy** - Considering that two extremes of a continuum are the only possible alternatives.

We have seen elsewhere how science is built on the works of those that have worked before us. When we construct a logical argument in science, we do so using this scholarly accumulation of knowledge as presented in the literature: citations. That is to say that your claims must be backed up by the literature. To do this, you will need to read that literature and make sure that it can back up your claims. Beware of making a baseless claim.

It may be that when researching your question, you come across the same argument using the same literature rehashed time and again in different papers (it happens). Does this give you a green light to do it again? I’d like to think that you already know that there are likely lots of other untapped and better examples out there, and it’d be well worth your while constructing your argument yourself.

Beware of copying an argument wholesale. You might well end up getting muddled, or worse perpetuating an error. Better go back and make sure that you understand the original premise and the works that promote or oppose this. Having told you to beware, I’m now going to encourage you to read, because [reading is one of the best ways in which you can learn about writing a logical argument](criticalreading.html).

Reading critically will make you aware of when you come across an argument. Ask yourself:

* Did you understand it?
* Was it written in a conventional style (as in the table (\#tab:argument))
   + If not, how was the style broken with and did this improve or detract from the understanding?

In this writing book, I concentrate on providing formulaic approaches to writing, because these are by far the easiest ways for inexperienced writers. However, I encourage you to learn and experiment with writing styles as you become more experienced. [Critical reading](criticalreading.html) is one of the best ways to learn about alternative writing styles (see here): reading is probably the best way of improving your writing. Alternatively, and especially if the above hasn’t clicked with you, you can read more about writing a scientific argument (see for example [here](https://msu.edu/course/iah/231b/gifford/sciaargs.htm)). And there’s plenty more out there.

### 2021-04-15 2.07-storytelling.Rmd
#First upload of book

## Storytelling in science? {#storytelling}

I started thinking about this topic some years back as I often need to write popular articles that make some of the science that we do more generally accessible. But then I came across an article by Nick Enfield [-@enfield2018our] that made me think again.

“No storytelling” is a comment that I sometimes make when reading drafts of manuscripts, chapters, and even when editing for journals. What do I mean by this? Stories are deterministic. That is to say that the storyteller has an end in mind when they start telling the story, and the telling is a way to get to their goal. A story that’s ‘pointless’ will frustrate the audience and won’t engender them to listen to that storyteller again. In a good story, reaching that goal will often result in lots of twists and turns with the goal shrouded in mystery until it is revealed. In a teaching story (like a parable), the goal may be overt, such that the audience relates to the narrative and buys into the same conclusion.

If we did science like we tell stories, we would decide on the way the system works before we studied it, and then design the experiment in order to reach our desired goal. You should have recognised by now that this is not the way we do science. This is clearly an undesirable way to go about doing science because we should never prejudice the result that we’ll get from a study before we do it. Hence, writing stories and writing science are different processes.

(ref:Storytelling) **It's story time.** Are you sitting comfortably? Then I'll begin.... but it won't be science

```{r Storytelling, echo=FALSE, out.width='95%', fig.cap="(ref:Storytelling)"}
 knitr::include_graphics("figures/Storytelling.jpg")
```

### The danger of a teleological argument

Teleological arguments are those which have their end point as the aim. While it is generally acknowledged that this is a good way to tell a story, it can be a trap when it comes to conducting scientific inquiry. Here, we need to approach science in a very different way to storytelling. When I was doing my PhD I was very frustrated as I had the impression that my advisor knew what result he wanted and designed the study to show it. What probably happened is that he was displaying “motivated reasoning”, but the result is known as “confirmation bias”. When an experiment failed to meet the expected result, he declared that it had failed. As Frances Bacon so elegantly put it:

>__“Once a man’s understanding has settled on something (either because it is an accepted belief or because it pleases him), it draws everything else also to support and agree with it. And if it encounters a larger number of more powerful countervailing examples, it either fails to notice them, or disregards them, or makes fine distinctions to dismiss and reject them, and all this with much dangerous prejudice, to preserve the authority of its first conceptions.”__ 
>Francis Bacon [-@bacon1620novum]

In our work, we read other studies and observations to formulate a question that we frame as a hypothesis. We then devise an experiment that will test this hypothesis in the most objective way, so that we can fairly accept or reject our null hypothesis.

### SCience is the very opposite of storytelling. Or is it?
I often tend to think of the answer to the question as the goal of the study. That I don’t know what the answer is, doesn’t spoil the story for me. The important thing is asking the question. An unexpected answer might send us back to thinking more about the system that we are studying and result in a greater revelation.

A good example of this is the study published by Becker et al [-@becker2018extreme]. Francois found a (spookily) strong relationship between probability that Rose’s dwarf toadlets survive and local rainfall in the breeding season. More rainfall equals greater survival is what one would automatically think (because frogs like rain), but that wasn’t the result obtained. Francois found that survival increased with less rainfall. It wasn’t until we made the connection between the fact that increased rainfall during the breeding season meant that the toadlets spent more time in puddles, that we understood why their survival was decreasing. As the toadlets aren’t feeding during this time they lose weight, and also expose themselves to increased predation pressure. In a dry year, the toadlets will head back to their subterranean refuges much earlier and continue to pursue ‘safer’ feeding and hiding habits. While we might intuitively feel that a ‘safe toadlet’ is a better life-strategy, reduced rainfall means reduced reproduction, and so results in a failure for toads that don’t manage to pass on their genes that year. The result is a variable life-history with the weather, something that was previously unknown. To me, that’s a great story!

But the scientific paper that was written about this ‘story’ [@becker2018extreme] doesn’t have a narrative style, and doesn’t fit the description of a narrative that we discussed above. Instead, it follows the formula that we set out below: Introduction, Materials and Methods, Results and Discussion. This style does not treat the experiment as a complication on the way to the story’s goal. The structure introduces the rationale behind conducting the experiment, then objectively explains the findings, and lastly discusses their meaning in relation to what is already known. From this study, I can spin the story (above), because I know what the endpoint is. But this storyis just one interpretation of the correlation that was found. [Correlation is not causation](argument.html), and more studies would be needed to test the hypothesis concerning the underlying reasons suggested for the relationship observed. The key differences in the storytelling style from the scientific formula are the absence of a known goal at the start of the process (see a more detailed discussion of this by Yarden Katz, [-@katz2013storytelling] here).

There is a role for stories when communicating science as this increases interest, facilitates understanding, and enhances memory. This is particularly true when communicating science to the public and making it more accessible, but it also applies to interactions between scientists, for example at conferences. The [presentations](#proposal-presentation.html) that tell a story and entertain are those we tend to remember. Not easy, but if we do want to communicate well with each other, then we need to learn the art of storytelling, without compromising our scientific objectivity. In addition, there are lots of lessons from storytelling that we can use within scientific papers, and to which I subscribe wholeheartedly. For example, don’t be boring, don’t [bamboozle](style.html), write an active narrative (not in the passive tense), use the magic of threes, and many more.

I've taken the deterministic interpretation of storytelling (a story with a known ending), and I've used this to argue why it is not a good tool for science. But it is a great tool for communication. And storytelling can be used to improve the way that we write science, and especially the way in which we communicate science.

There are other fundamentals in stories that are particularly useful and should not be ignored. One notable feature of stories is that they make facts easier to understand and remember. Indeed, there is even the idea that this is why storytelling evolved in human societies. For example, most religions are based on stories that are geared to understanding of societal morals and rules. Martin Angler’s [-@angler2020telling] excellent book about how to turn science into stories should be consulted to give you some deeper insight into why stories are more powerful than science. Or try the exercise described by Torres and Pruim [-@torres2019scientific].

### Can storytelling help memory of science?
Some argue that it can, and should be an integral part of scientific writing. In her [blog post](https://www.annaclemens.com/blog/elements-of-a-story), Anna Clemens argues that stories should be used within the scientific context. I'll let you read this for yourselves, and ask you to make up your own minds. If you like this way of thinking, you can get deeper into the depths of story thinking with Angler [-@angler2020telling].

### 2021-04-15 2.08-citations.Rmd
#First upload of book

## Why do you need to cite? {#citations}

Citations are a means of recognising that your work is built on the work of others. They are part of the fundamental tenets of science that we don't start from scratch each time we conduct a study.

(ref:Googlescholar) **Google Scholar home page.** Have you ever wondered what the blurb on the front of Google Scholar means? Who is standing on whose shoulders? Google and the Google logo are registered trademarks of Google LLC, used with permission.

```{r Googlescholar, echo=FALSE, out.width='90%', fig.cap="(ref:Googlescholar)"}
 knitr::include_graphics("figures/Googlescholar.png")
```

> Isaac Newton (in a letter to Robert Hooke in 1675) wrote:      
> **"If I have seen further it is by standing on the shoulders of Giants."**

Although the phrase is often attributed to Isaac Newton, it turns out that it was well accepted in Newton's time and has been traced back to the 12th century.

Essentially, standing on the shoulders of giants (Figures \@ref(fig:Googlescholar) is a recognition that all research is built upon research that has gone before it, and this is the basis for citations in the text of scientific papers. Patrick Dunleavy [@dunleavy2017citations] argues that citations are required to meet 7 criteria with respect to academic writing. Others have suggested that there are more criteria [see @harwood2009interview], and that these are a fundamental premise of academic work. But it was not always thus. If we dwell a moment longer, we can look back at the World's longest running scientific journal: *Philosophical Transactions of the Royal Society.*

You can download the first of those papers of 1665 and note that there are no citations (other than to books or letters) because there were no previously published articles from which to draw. However, even then authors noted that ideas came from previous authors and we can regularly find acknowledgements that refer back to Aristotle (~350 BC).

(ref:Boyle) **A page from the first volume of the Philosophical Transactions of the Royal Society.** Robert Boyle An Account of a Very Odd Monstrous Calf, _Phil. Trans. R. Soc._ 1665 1 10; https://doi.org/10.1098/rstl.1665.0007

```{r Boyle, echo=FALSE, out.width='90%', fig.cap="(ref:Boyle)"}
 knitr::include_graphics("figures/Boyle.png")
```

### Research is built on existing work and ideas

It would therefore be very unlikely that your idea/ideas has no basis in existing literature. If you can’t find it, the chances are that you haven’t looked in the right way. Try Google Scholar, Scopus, Web of Science and vary the search terms or try searching for articles citing something similar.

Citations demonstrate to readers where your ideas have come from. Citations can also be used to reduce what you need to write – especially with respect to methodology. If you (or others) have already provided the methodology in full then you can give a much simpler description and the citation.

Citations need to be used to back up any statements of fact that you make in your PhD chapter. Any examples need citations for where the examples came from, and when you make [arguments](argument.html), you add credibility to your arguments from each side by citing that these are from published works.

### Your citations say a lot about you

When you choose to cite one paper over another, you are making a statement about what you find credible in the literature and what you don't. For example, if you choose to cite a poorly conducted study as the cornerstone for one of your arguments, then others may interpret that your arguments are built on shaky foundations. Experienced (often senior) readers who know the literature well will be able to judge the quality of your work not only by what you've written, but on what you've based your ideas through your citations. As we will see later on, there's a lot to think about when deciding what to cite, and what not to cite.

### Cite while you write (not afterwards)

Automatically, I'd suggest that you cite whenever you write. Using citations becomes habit forming, and you’ll end up wanting to use citations everywhere. Popular scientific writings tend not to have citations, but they can still be there but subtly different.

Within a manuscript for a scientific journal or in your thesis, you can expect that your [introduction](introduction.html) and [discussion](discussion.html) sections are going to be full of citations. The methodology will likely also be citation rich. By using a lot of citations in your methods, you can often save your needing to write a lot of sections which would otherwise be very detailed.

Later in the book ([Writing with a Formula](formula.html)), I explain how to start writing with an outline, then flesh out the outline with citations from your reading, and plan which citations you are going to use carefully before you start. This won't stop you from adding more later, but this ensures that your work is built on your scholarly endeavour.

I've seen many students try to write their text first and then only later fill in citations. I've had students hand in thesis chapters and manuscripts that have [ref] written strategically at the end of many sentences. I've even had the experience of doing this myself when I was a early career researcher (I probably still have those manuscripts somewhere because it's a sure way of never getting anything finished). I suggest that you **do not do this.** The reason is that if you write first and try to insert citations later, you are going to spend lots of time looking for the citation that fits your text. This is clearly not the right way around. Your text should be based on what's present in the literature.

I was planning to write that no-one would ever write an entire manuscript without references, and then put them all in at the end. Then I came across a video by [Pat Schloss](https://youtu.be/yk54_j3-4RM), and realised that I'm wrong [@schloss2018riffomonas]. Pat states early on (in the video) that he prefers to write first and cite later. He's clearly a very experienced researcher and knows his literature very well. Indeed, he knows it all so well that he already knows what citations fit what statements without needing to look them up. However, later on in the video Pat admits that even he sometimes gets it wrong, and that means that he ends up having to rewrite text when he finds the publication that he was thinking of and notes that what they actually found did not fit what he wrote. Although I can identify with what Pat is saying, I would find this re-writing far too time consuming to take this approach myself. I am familiar with most of the pertinent literature in my specialist area, but I'd rather check things before I start writing than leave it until later. I find that my memory is not only wrong about what other people have found, but also about what's in my own work. Therefore, I'd suggest that while you may aspire to write like Pat later on in your career, right now it'd be better for you to plan your writing with citations in mind, and cite while you write (not afterwards).

When you submit a draft of your chapter to your advisor, they may suggest additional citations that are known to them. Don't simply add blindly, but look them up and check their relevance to the statement concerned. They may require recrafting of the statement, or a query back to your advisor.

### Citation Styles

There are a number of different styles, and this is likely to depend on the journal that you are writing for [see @pandey2020methods]. The two most prevalent styles in biological sciences are often referred to as Harvard (name, year) and Vancouver^1^ (superscript numbers that are listed in the reference section). Most universities and journals require Harvard style. The intricacies of how exactly this will get carried out will change from institution to institution and journal to journal. You will have to find out what is relevant to you.

#### Vancouver Style

Vancouver style dates back to medical journals in the 1970s and refer to conventions to avoid fraud in medical sciences. In biological sciences, they are not used as widely as Harvard style, but are chosen as they don’t impede the flow of the text for the reader.

Each style has its pros and cons. Vancouver style is more equitable and anonymous, as all cited people are represented only in the references and do not have their names plastered throughout the text. The diminutive numbers take up less space, allowing the reader and writer to concentrate more on their prose. On the other hand, if you are familiar with the literature, or you are interested in learning more about the literature, you’ll find that the little numbers take much more time to cross reference with the list at the end, than does Harvard style. Reference strings in Vancouver style can either be separate numbers, or numbers listed as a series. Because citations are numbered in the order in which they first mentioned, citation strings later in documents often end up as a list of non-sequential numbers separated by commas.

Here is an example of text [from @measey2016global] with citations in Vancouver style:

>Amphibian populations are currently declining across the globe^1-3^ and alien amphibians are at least partially driving these declines through competition^4^, hybridization^5^ and introduction of novel pathogens^6-9^.

^1^Berger L, Speare R, Hyatt A. _Chytrid fungi and amphibian declines: overview, implications and future directions. Declines and disappearances of Australian frogs_ Environment Australia, Canberra. 1999;1999:23–33.

^2^Collins JP, Crump ML, Lovejoy III TE. _Extinction in our times: global amphibian decline_. Oxford University Press; 2009.

^3^Daszak P, Cunningham AA, Hyatt AD. Infectious disease and amphibian population declines. _Diversity and Distributions_. 2003;9(2):141–50.

^4^Dufresnes C, Dubey S, Ghali K, Canestrelli D, Perrin N. Introgressive hybridization of threatened European tree frogs (*Hyla arborea*) by introduced _H. intermedia_ in Western Switzerland. _Conservation Genetics_. 2015;16(6):1507–13.

^5^Kupferberg SJ. Bullfrog (*Rana catesbeiana*) invasion of a California river: the role of larval competition. _Ecology_. 1997;78(6):1736–51.

^6^La Marca E, Lips KR, Lotters S, Puschendorf R, Ibanez R, Rueda-Almonacid JV, et al. Catastrophic population declines and extinctions in Neotropical harlequin frogs (Bufonidae: *Atelopus*) _Biotropica: The Journal of Biology and Conservation_. 2005;37(2):190–201.

^7^Martel A, Spitzen-van der Sluijs A, Blooi M, Bert W, Ducatelle R, Fisher MC, et al. *Batrachochytrium salamandrivorans* sp. nov. causes lethal chytridiomycosis in amphibians. _Proceedings of the National Academy of Sciences_. 2013;110(38):15325–9.

^8^Pimm SL, Jenkins CN, Abell R, Brooks TM, Gittleman JL, Joppa LN, et al. The biodiversity of species and their rates of extinction, distribution, and protection. _Science_. 2014;344(6187):1246752.

^9^Wake DB, Vredenburg VT. Are we in the midst of the sixth mass extinction? A view from the world of amphibians. _Proceedings of the National Academy of Sciences_. 2008;105(Supplement 1):11466–73.

#### Harvard Style

The standard way is to use the name and year in parentheses at the end of the statement to which the citation is relevant. Names of three or more authors are frequently reduced to “et al.” (often written in italics) which is short for the Latin et alia, meaning “and others”. References strings in Harvard style can go on for an entire line of text or more. Some journals have rules that mean on first mention the citation should have all authors (up to a certain number). This can become tedious when citations start taking up more space than text.

A repeat of the above example, except with Harvard style shows how much more space these same citations take up.

>Amphibian populations are currently declining across the globe (Wake and Vredenburg, 2008; Collins et al., 2009; Pimm et al., 2014) and alien amphibians are at least partially driving these declines through competition (Kupferberg, 1997), hybridization (Dufresnes et al., 2015) and introduction of novel pathogens (Berger et al., 1999; Daszak et al., 2003; La Marca et al., 2005; Martel et al., 2013).

Berger, L., Speare, R., Hyatt, A., 1999. _Chytrid fungi and amphibian declines: overview, implications and future directions. Declines and disappearances of Australian frogs_. Environment Australia, Canberra 1999, 23–33.

Collins, J.P., Crump, M.L., Lovejoy III, T.E., 2009. _Extinction in our times: global amphibian decline_. Oxford University Press.

Daszak, P., Cunningham, A.A., Hyatt, A.D., 2003. Infectious disease and amphibian population declines. _Diversity and Distributions_ 9, 141–150.

Dufresnes, C., Dubey, S., Ghali, K., Canestrelli, D., Perrin, N., 2015. Introgressive hybridization of threatened European tree frogs (*Hyla arborea*) by introduced _H. intermedia_ in Western Switzerland. _Conservation Genetics_ 16, 1507–1513.

Kupferberg, S.J., 1997. Bullfrog (*Rana catesbeiana*) invasion of a California river: the role of larval competition. _Ecology_ 78, 1736–1751.

La Marca, E., Lips, K.R., Lotters, S., Puschendorf, R., Ibanez, R., Rueda-Almonacid, J.V., Schulte, R., Marty, C., Castro, F., Manzanilla-Puppo, J., others, 2005. Catastrophic population declines and extinctions in Neotropical harlequin frogs (Bufonidae: *Atelopus*) _Biotropica: The Journal of Biology and Conservation_ 37, 190–201.

Martel, A., Spitzen-van der Sluijs, A., Blooi, M., Bert, W., Ducatelle, R., Fisher, M.C., Woeltjes, A., Bosman, W., Chiers, K., Bossuyt, F., others, 2013. *Batrachochytrium salamandrivorans* sp. nov. causes lethal chytridiomycosis in amphibians. _Proceedings of the National Academy of Sciences_ 110, 15325–15329.

Pimm, S.L., Jenkins, C.N., Abell, R., Brooks, T.M., Gittleman, J.L., Joppa, L.N., Raven, P.H., Roberts, C.M., Sexton, J.O., 2014. The biodiversity of species and their rates of extinction, distribution, and protection. _Science_ 344, 1246752

Wake, D.B., Vredenburg, V.T., 2008. Are we in the midst of the sixth mass extinction? A view from the world of amphibians. _Proceedings of the National Academy of Sciences_ 105, 11466–11473.

From here on (and throughout the book - because the book uses a form of Harvard style), I concentrate on Harvard style as this provides more freedom for how citations can differ.

### Moving from Harvard to Vancouver

Essentially, every time you would have had the year and date, you will replace this with a superscript denoting the number for that particular citation. However, Vancouver style does not allow for much variation in how you cite. You can’t use (e.g. ^1,2^) or (see ^3^ for a review). This means that when you originally wrote for a Harvard style journal, but then change the manuscript to submit it to a journal that uses Vancouver style, you’ll need to remove any explanations of citations. That doesn’t mean removing the citations themselves, but it may mean re-writing some sentences. Simply changing your reference manager to use the different referencing style will only be the start of your work.

### Where within a sentence should the citation come?

There are a number of different styles, and this is likely to depend on the journal that you are writing for. The standard way is to use the name and year in parentheses at the end of the statement to which the citation is relevant.

The impact of all invasive amphibians is similar to that of invasive birds and mammals [@measey2016global].

You’ll see that sometimes the names are brought into the sentence and become the central agents of the text. The phrase et alii (often abbreviated to et al.), literally means "and others" referring to the other authors (typically used when there are more than two - but check the journal format!).

Sometimes, instead of "et al." you can write "and colleagues" or "and others". This is something to do occasionally when you are looking to diversify some text. Don't over do it though.

>Measey et al. [-@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals.

This technique is very useful when you then want to add another sentence or two about this same study.

>Measey et al. [-@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals. They did this by constructing GISS scores for all individuals in all groups.

Because the authors are the subjects of the first sentence, the citation becomes implicit in the second sentence. Then you don’t need to use the same citation again within the paragraph.

What about page numbers?
Sometimes you'll see a citation with a colon and page number after. This really only needs to be used if you are quoting specific text on a particular page:

>Measey et al. [-@measey2016global: 976] proposed that using GISS scores could show that "some amphibians can have devastating impacts to the environment".

### What about the order of the citations in a string?

A citation string is a list of two or more citations that all assert the statement given. When you provide a string of citations you will need to decide which one comes first. Normally, this is given in the order of precedence: the oldest citation comes first, the youngest last. However, journals have their own styles and these may dictate how citation strings are ordered. For example, they could be alphabetical. If your referencing software has a style then you can relax. Otherwise, you'll need to look it up!

### What about citations as taxonomic authorities?

Taxonomists have special rules for this, and I'll do it in another chapter (see section on [Scientific names and taxonomic authorities](taxonomy.html)). These are not the same as regular citations (because they don't appear in the literature cited and you don't have to have read the descriptions), and only some journals ask for them.

### Is it possible to mis-cite?

Yes. One of the most common ways in which students mis-cite a paper is to use statements made in the introduction (or discussion) which were not the subject of the study. For example, in the introduction of their paper, Measey et al. [-@measey2016global] make comments on amphibian decline.

>Amphibian populations are currently declining across the globe [@wake2008are; @collins2009extinction; @pimm2014biodiversity] and alien amphibians are at least partially driving these declines through competition [@kupferberg1997bullfrog], hybridization [@dufresnes2015introgressive] and introduction of novel pathogens [@berger1999chytrid; @daszak2003infectious; @lamarca2005catastrophic; @martel2013batrachochytrium].

However, it would be wrong to give a statement on amphibian decline and cite Measey et al. [-@measey2016global]. They did not study amphibian decline. Instead, you should read the papers that they cite [e.g. @wake2008are; @collins2009extinction; @pimm2014biodiversity], and read around those to find studies on amphibian decline that are appropriate for your context. This underlines one important aspect of choosing citations where the statement that you make relates directly to the study carried out in the citation.

Another common mistake is to forget which paper has which information. You can try to make sure that you don’t do this by taking better notes or a more accurate plan.

### Should I cite without reading the paper?

No. When you are citing a study, you should be sufficiently familiar with the publication that you are endorsing the study in relation with the statement that you make (but see below). If you are not convinced by the nature of the study that you are tempted to cite, then rather don’t cite it and use another one. If you can't get hold of the paper, this is another reason why you might not cite it. This is a regular reason given for why Open Access journals attract more citations.

### What should I not cite?

This does depend on the journal you are writing for. Some journals don’t permit citations to unpublished data or web sites. My suggestion would be to avoid such sources anyway (with particular exceptions – see below), unless it is really important that you include it. Other examples of texts to avoid are: text books (use the original paper instead), newspapers or magazines, blog posts, facebook pages (or other social media sites), predatory journals (or any non-peer reviewed article). These are all examples of __grey literature__.

There are of course exceptions. When you are writing about social media sites, or newspaper coverage, you will want to cite those sources.

I don't like citations to guide-books or general text books. This has become very common, but really smacks of laziness. Most of what's written in a guidebook has already been documented in the scientific literature, but guide-books generally don't provide sources for the information that they provide. Thus, it's easy to look up and find something in a guide-book, but suggests (to me) that you haven't spent enough time or effort reviewing the literature. Some guide-books are excellent and the authors have gone to a lot of trouble to incorporate original data and observations. But this is unusual, and most data can be found in publications.

One exception to the not citing websites relevant to our work is the IUCN Red List. Note that all entries on this site now have DOIs, and this might be a good guide for what is available to cite. The Digital Object Identifier ([DOI](doi.html)) is very useful as it means that there is a consistent record of that version. Otherwise, you could cite any website (even this one) and then the owner can go and change the site and it no longer says what you thought it did. The DOI removes this problem; there will always be an archived version with that particular DOI.

You should not cite papers that have been retracted.

### Do I cite the review or the primary literature?

The primary literature are studies or experiments that are done in order to test a hypothesis. Secondary literature includes reviews, syntheses or meta-analyses. Primacy (see below) is important, but this depends on the space you have and whether the review contains all the information you need to cite. Sometimes reviews (especially meta-analyses) provide more information. It’s preferable to use primary literature, but sometimes reviews (or meta-analyses) are actually more expedient to use, especially if they are not the focus of your study. You can even cite both when relevant.

### What is primacy and why does it matter?

Primacy in the literature is the paper/s where the authors provide the original idea or concept that you are citing. It is important in that you should give credit to original ideas over those who copy or simply repeat them, or even those who review them. Sometimes primacy is less important, especially if the concept is well known and has changed substantially. Then you should cite the most recent work that shows the ideas that you are wanting to show. Other times, primacy is important - especially when few have built on ideas or concepts since they were put forward. If in doubt place a citation to the original study and the most recent study, then flag this using a comment for your advisor to consider. The important point here is for you to give credit where it's due and not to overlook those who put in the hard work to publish original content.

### How many citations is enough?

Some journals have a word limit, or even a limit to the number of references that they allow. Others do not, and you should probably use what is recently published as a guide to what is acceptable. For the chapter of a thesis, you should err on the upper end; from 50 to 100 references. Note that citations may well be more as you may cite a paper more than once.

Obviously, everything you cite needs to be in the References (or [Literature Cited](references.html)) section, and you may well need to spend time deleting extra stuff. You can get around this common issue by using a citation manager like [Mendeley, Zotero or EndNote](references.html). I’d always suggest that you use one of these tools as they can really help with your reading too. These days they are busy turning into a kind of scientific social network. They regularly make suggestions of what you could be reading based on what you read. This can be useful.

If you are looking for multiple examples of a statement that you’ve made and there are many possible, I’d suggest that you aim to produce three. Make sure that you use a suffix (like “e.g.”) to show that you are aware that these are examples of a widely reported phenomenon. You can choose these as you like, but may want to consider using what you consider to be ‘the best’ examples, and/or references that you are planning to use elsewhere in your paper/chapter. This can drive the total number down considerably, and helps to keep the citations more relevant to your work.

### Should I cite myself?

If you are publishing relevant and appropriate papers, then there is no reason not to cite yourself. In many cases (such as with your thesis work), your own publications are likely to be more relevant to methodology and subject matter than much of the other work that is out there. However, if you’ve previously published on termite fungus and now you’re publishing on frog toes, it’s unlikely to be relevant. Choosing citations should include your comfort with being transparent. If you know when citing yourself that citing another article would be more correct in showing where an idea came from, but citing your own is more relevant to the application, then you should feel comfortable in citing both.

I would avoid citing your thesis if possible, rather put it all into papers. There are times though when citing your thesis is unavoidable. Within your thesis, I would suggest that you do provide citations to different chapters as this will help the examiners see how the chapters relate to each other.

Take a look at this article on self-citation. They claim that self-citations in the Natural Sciences run at 33% [@vanraan2007scoping], which provides you with a nice idea of what is acceptable. Remember that this would include not only your citations to your articles, but all citations to your co-authors articles too.

### Should I cite my friends?

It may be easier to cite your friends if you already know their work well. You may have heard them talk and know that the subject is relevant. They may be encouraging you to cite them, but should you?

In these days of scientometrics, we do need to acknowledge that citations act as a kind of currency. They count towards your H-index [@hirsch2005index] and this can reflect on your prospects as a postdoc or employee. What’s also clear is that cited papers get cited more, so it could really help your friends if you cite them. Obviously, the inverse is also true, so beware of the politics of citing. However, the most important points have already been raised, but will be tackled [in detail later](impactfactor.html). The study must be relevant and appropriate before it gets included as a citation in your work.

### Does the impact factor of the cited article matter?

Papers in journals with high impact factors are more likely to be cited because their contents are already thought to be of interest to a wide range of people. Indeed, there is evidence that identical statements published in many journals, those with higher Impact Factors are cited more [@perneger2010citation]. That is something that the editors of that journal will have considered even before sending the manuscript to review. Sometimes (but not always) the impact factor of the journal can be an indication of the quality of the study. But you should judge this for yourself when you critically read the paper.

I find that the first paragraph of a paper is more likely to contain citations of higher impact journals. This is in part as these are likely to be more cross cutting (as is often the case for the first paragraph). In the end, if you need to make a choice, choose the paper that is most relevant, irrespective of impact factor.

### 2021-04-15 2.09-databases.Rmd
#First upload of book

## Literature databases {#databases}

Databases of scientific literature have been around since the 1960s when the number of journals and papers started increasing so much that scholars simply couldn’t keep up. Quite simply, you should use one or more of these databases to look for relevant literature to read. The first database, and what continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of journals (titles, abstracts, authors and other metadata), but also the citations that each paper used. As we have seen in the last chapter, citations record the way in which people find the literature useful, and are therefore a highly effective way of finding related and pertinent literature in your field.

Before you start using a literature database, you will need to have some key literature with which you are familiar and that relates directly to your project. If you don’t have this, then check with your advisor. Often, this key literature is cited in the advert for the PhD, the grant application that funded it, or if this is a topic that your advisor has already conducted some research on, then their own publications may include part of the key literature. If you are not familiar with your advisor’s (including co-advisor’s where appropriate) papers, then a great first step is to search for them on a database and read through titles and abstracts that are relevant to your research topic. Once you have a set of key literature (this might only be 5 to 10 papers), then it is time to go to the literature databases to do your search.

### Searching the literature using a database

There are two typical ways in which you can search for literature about your area of interest. The first is to use keywords and the second is to use citations of key references. I suggest that you use both of these tactics when doing your initial literature search. Further, I’d suggest that it’s worth your time keeping a record of what you have searched for, how and when and then download the results. This is called a systematic literature review. Using this tactic from the outset of your interactions with literature databases will give you a good idea of what proportion of the literature you are reading, and you might then be able to convert your finding into an objective literature review - making a great first introductory chapter for your thesis.

### Keywords

While searching for keywords sounds easy, picking the keywords that you should use is more difficult and requires a lot of practice with the databases and your key references. Remember that keywords are often not single words, but also include short phrases. Look at all of the key references and the keywords that they contain. Any keywords that they share are going to be important for you. Most literature databases have the option of using keywords or keywords and words in the title and abstract. I would start with the broader search term and see how well they work. However, these searches alone may return such a large amount of literature that you can’t possibly read it all. In this case, you will need to use more than a single keyword. The trick is to find the correct combination of keywords that will deliver all (or most) of your key references, while not delivering returns of thousands of papers that are not relevant to your study. Ideally, you want a search term that will deliver around one hundred papers at most, where the majority have relevance to your study area. Luckily, literature databases have some tricks for you to add to combine your keywords in ways that you can include or exclude certain topics.

#### Boolean terms or operators

There are three principle Boolean operators: AND, OR and NOT.

You should use these to combine a string of keywords. They are relatively straightforward to use:

- Use AND when you want both terms included. For example, *performance AND lizards*. In this example, all papers that mention performance AND lizards in the title, abstract or keywords are returned. 
- Use OR when you want to include both terms. For example, *performance AND lizards OR geckos*. This example should return more literature than the previous search as it will include, all papers that mention *performance AND lizards OR geckos* in the title, abstract or keywords are returned. You will note that geckos are lizards, but people who write about geckos might not mention that they are lizards in the title, abstract or keywords, so you may need to use these extra search terms in order to pick up all of the relevant literature on lizards. If your research requires you to cover all lizards, then you might need a lot of search terms in the OR string in order to pick up all of the literature that covers lizard performance.
- Use NOT when you specifically want to exclude a keyword. Following our example above, the search term *performance AND reptiles NOT snakes* should give you all performance papers on lizards, but you may want to exclude turtles and crocodilians as well if they have a lot of hits.

#### Proximity operators and parentheses

You can also use the term NEAR to indicate that two terms need to occur within a specified number of words. For example, performance NEAR/5 reptile will capture items where these occur within 5 words of each other, but no greater. This might be useful to exclude some papers that mention performance and reptiles in the abstract, but are not about reptile performance.
Once you start putting together search strings you will need to use brackets to make sure that the correct terms are grouped. For example, performance AND (lizards OR geckos) will get what we described above, whereas (performance AND lizards) OR geckos will give you all papers on lizards and performance plus all papers on geckos (not what we wanted).

Explore the advanced search options of the specific database that you are using to see the correct semantics needed for a search.

#### Wildcards

Most literature databases allow you to search using different wildcards. These are *, $ and ?. These can be placed at any point within a search term and help overcome differences in spelling, or families of words that you may want to capture or exclude in your search.

- Use * when you want to replace a set of letters at the end of a word. For example, _reptil*_ will return all terms including Reptilia, reptile and reptiles.
- Use ? when you want to search for any single letter at that place in a word. For example, _sterili?e_ will capture words with sterilize and sterilise.
- Use $ when there is more than a single letter that might change within a word, or the omission of a letter. For example, _colo$r_ will return both US and UK spelling, color and colour, respectively.

#### Combining searches

Lastly, there is the option of combining searches, so that you can add search strings together or keep them apart. You will need to use the advanced search options in your database to do this.

You will need to carefully check your database for their precise semantics when using these Boolean operators in a search string. They are all slightly different, often using quotes around the keywords, and your search won’t work as you intend unless you are using the Boolean operators in the way the database dictates.

Your specific search term may end up having a long list of terms. Remember to save this whenever you make a search, especially if you refine it during your studies. Most literature databases will provide the search string that you used together with the results. Check your search string carefully for possible spelling errors that could influence your search.

#### Citations

In addition to using keywords, you should search citations of your key references. Typically, when you start reviewing the literature in your specific area you will have a few key references that have been given to you by your advisor. Having read these, you will already know that they cite other works of interest, and you should follow up to look at all that seem relevant. But that’s not all, these key publications are likely to have also been cited themselves by newer papers. Tracking citations of key papers can help you keep abreast of what is happening in your field. Indeed, given the prolific ever increasing nature of scientific publications, it might be the only way to keep on top of what is happening. Some literature databases allow you to receive alerts if a paper is cited, and you should set this up for all of your key citations, and perhaps for some of the most prominent authors in your field.

You can’t search every paper for all the references that it cites, and its own citations. Although this would give you a solid systematic basis for your literature search, it is likely to be too much work. Getting good search terms for your keywords, and finding key references to search for citing articles is a better strategy as it makes your literature searches manageable. If you refine and save your search terms it also makes it repeatable. This will provide you with a solid basis for moving forward with your PhD, and with appropriate alerts make sure that you keep up with the latest literature in your field.

The above section is a brief guide, and your library is likely to have in depth guides on how to use the specific databases that they house. They may also provide training sessions, or online videos to watch. It is worth becoming familiar with exactly how the databases that you will use work, so this is time well spent at the beginning of your studies.

What else are literature databases used for?
The Science Citation Index used to be published as a single volume that emerged each year and was available in the reference section of the library. By the time that I started at university as an undergraduate in the 1980s, it was several volumes large and my first task for my tutor (one Dr Dave Thompson) was to look up all of his citations. Within a couple of years, the database was digitised and available on CD, and eventually became accessible through the internet.

Clearly, the Science Citation Index cost some money to put together, and universities paid annually to receive the books, and then the CDs and finally to access the online versions. They were not for profit organisations, but eventually they were bought out by the evil profit making parasites of science. Yes, publishers saw the dependency that scientists have on a resource and decided to own it for themselves and then radically increase the price to access it.

### Administrators want metrics

But it wasn’t just scientists that were interested in the databases of literature. Administrators became interested in the metrics that the databases produced on academics so that they could use them as a measure of performance, or potential in the case of hiring academics. The obsession with simplifying metrics by administrators who don’t understand the work that their academics produce, has driven the importance and proliferations of literature databases. However, it is widely acknowledged that the way in which journal metrics are used is unhealthy, and this has led to the [San Francisco Declaration on Research Assessment](https://sfdora.org/read/] (known as DORA)).

In the life sciences, there are three prominent databases used: Google Scholar, Web of Science (owned by Clarivate Analytics aka Thomson Reuters) and Scopus (owned by Elsevier). But new players are entering the market all the time, and two of the newest are Dimensions from Digital Science and Microsoft Academic.

### Google Scholar, Web of Science or Scopus?

Have you ever wondered why Google Scholar (GS) scores are so inflated compared to other citation databases like Web of Science (WoS) or Scopus? I've always noticed that Scopus has better coverage than WoS, and that GS is bigger than both (and a lot messier with lots of weird duplicates and poorly entered stuff), but is there anything more to it than that?

Well it seems that there are some people who have already thought about this, and come up with a good idea of exactly what's different. Martín-Martín et al [-@martin-martin2018google] have done a great job of analysing all this stuff from some 2.5 million citations. What they found inspired me to write this chapter, in which I've focussed on the results from the Life-Sciences. But I encourage you to go read the article for yourself.

I have been known to take the odd peak at my Google Scholar profile over the year, and see how it's coming along. I rarely check on WoS or Scopus, 'cos it's a bit of a faff getting signed in and doing the search. Plus it looks so much smaller when one is habituated to seeing those double digits in GS! However, I've always been a bit uneasy about citing my GS citation rate, H-index or i10 (among others that they give) as I've never really known what all that extra represents. Something grey and unseemly? Well, it turns out that it's all good stuff, and perhaps GS is the better one to cite as it's a more inclusive index: more inclusive of different document types and different languages.

It is worth noting that of the three indices, Google Scholar seems to be the easiest to manipulate. Delgado López‐Cózar et al [-@delgadolopez-cozar2014google] show how it is possible to manipulate citations and therefore metrics of individual researchers by depositing false documents into institutional repositories (note that they did this as an experiment, and you may find that restrictions on doing this at your own institution are prohibitive or lax). The authors also note that a consistent problem with Google Scholar is the lack of transparency, and therefore verification, of how the bibliometrics are calculated. This will make a difference to what papers are found, as Google Scholar tends to report those with the greatest citations first on the first page of outputs. Hence once a paper is initially boosted by manipulation, it will be more likely to be found and increasingly cited.

To directly compare your citation metrics from different platforms, and also to get more unbiased metrics than these platforms provide, consider using the software Publish or Perish [@harzing2007publish]. This also provides a great platform on which to harvest data from Google Scholar into a database, or put it directly into another database, like EndNote, Zotero or Mendelay.

### What’s the difference between indices?

Martín-Martín et al [-@martin-martin2018google] allows you to interpret your GS results as a more inclusive citation index. While WoS and Scopus aren't exclusively English or journal publications, they are mostly. But that extra third that GS gives you allows you to show the extra scope that your work is getting outside that English journal mainstream. Is your GS score more than a third higher than your WoS or Scopus score? If yes, then your work is having a greater impact elsewhere in the world, and there's nothing wrong with that.

The excerpts from the two tables above show how well GS correlates with both WoS and Scopus in our area (Biological Sciences). It also tells you by how much the GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus. Again, if you deviate from this with a higher score, you can give yourself a pat on the back for having work that's reaching more people in more parts of the world.

### 2021-04-15 2.10-sortware.Rmd
#First upload of book

## What software should I use to write my PhD? {#software}

Thirty years ago, when I started to use computers as an undergraduate, the word processing standard was a piece of software called WordPerfect. Back then it seemed inconceivable that anyone would want to use anything else, or that WordPerfect could lose its grip on the market. WordPerfect became a victim of the Microsoft revolution that standardised most PC operating systems (think Windows) and software using Microsoft Office in the mid-1990s. The chances are that today your institution will still have subscriptions to Microsoft Office making it free for you to use. But given that changes in software can happen so fast and appear to change with increasing frequency, should you use Microsoft's Word to write your PhD?

The simple answer is that you should use whatever you, and your advisor, feel most comfortable with. You will face a lot of challenges during your PhD including learning to write with lots of different kinds of software. Each one you learn to use will require a steep learning curve and that requires time. If there's any good time to start with a new word processing package it will be before you start your PhD. If you do decide to go with a new writing package then my suggestion would be to choose open source software like latex or R markdown. More on these later.

If you feel comfortable with a word processing package with a nice GUI, like Microsoft Word, then go ahead and use it. My best suggestion is that you do not save your files in their default *.docx format. Instead make sure that all of your files are saved as *.rtf or html. This should make little difference to your opening and editing but will mean that all of your texts are forward compatible with other kinds of software. It will reduce your file size and reduce the chances of file corruption: i.e. when part of the file malfunctions and you lose the contents. That’s right, in addition to loss of data by back up deficiencies [see Part III](data.html), you also have the risk of having corrupt files that will also mean you lose all of the contents. Even if you always use Word and never use any other file format, do yourself a favour and save each version as a *.rtf or *.txt file as well. Just like a back up, this means that you won’t have lost all of the text if a file goes corrupt. And yes, I do know people who have multiple backup versions of corrupt files: it’s happened to me.

The real problem with word processing packages like Microsoft Word is when files get very large as they tend to do during a PhD. Once you start putting chapters together especially if they include embedded images tables and the like you will find that many word processors start misbehaving. This can be especially frustrating in the closing phases of your PhD when you're trying to submit it. Especially if there's a tight deadline. However this problem is easily overcome by chunking your thesis into documents equivalent to chapters and then merging PDFs from each chapter.

### Alternatives to conventional word processors

Many academics are turning to latex in order to write their manuscripts as this gives a lot of possibility to insert links, formulae, references, images, etc in a very simple way. Once you are ready to print a latex document it can be shared very easily as a PDF.

Be aware that the problems may come not with what you want but with how your advisor and or co-authors are comfortable with. I find that commenting on a PDF takes more time than doing the same on a word document. I don't mind working with latex but I know some advisors are very adverse to having to use this platform. Obviously in order to get the feedback from your advisor you are going to need to use a platform that they are also comfortable with. So please consult them on what they are happy to use. You may also need to instruct them about how to make comments. For example track changes is a function that a lot of people like to use but it doesn't translate back into other platforms seamlessly.

As with many things in this book I urge you to keep an open mind with respect to software. For example if your lab is progressive and already experimenting with new and flexible software platforms then I suggest that you learn with them. The future of computing is certainly likely to be open-source software and so the sooner that you become more familiar with this the easier you will find it to adapt in future.

Another consideration is how your word processing package is going to work with referencing and citations. Again your lab may already use some software that you want to integrate with your word processor.

If you are feeling adventurous, try writing in R Markdown, using RStudio as a text editor GUI (Graphical User Interface) and the [Bookdown](https://bookdown.org/) platform to publish your chapters into a single thesis file in virtually any format. There is also great support for both of these formats with free guidebooks about how to use them [@xie2016bookdown; @xie2018r]. Given that I wrote this book in R Markdown with bookdown, I feel that I should point out (albeit briefly) the advantages to this platform:

 - Reproducibility
 - Dynamic Content (include statistical code and graphics inside your thesis)
 - Focus on content (not formatting)
 - Numerous output types
 - Independent reference manager (via .bib file) and style files (via .csl file)
 - Free software and community support

Importantly, what you want to try to avoid is conducting a major switch between software right at the end of your PhD. Try making such changes when you have more time, like when you’ve just finished.

### Naming versions of files

Good file management is an important aspect of a PhD. I like my students to give each chapter of their thesis a nickname or short running title that we both agree on to name the files. An example might be “Swimming_performance_V1.rft”. The reason why I don’t like “Chapter_1.rtf” as a filename is that every student has a chapter 1, so I can’t distinguish between them. In addition, I’m not good at remembering which chapter is chapter 1 and which is chapter 2. Sometimes, towards the end of a thesis, the chapters get switched around anyway, causing even more confusion. Notice that I don’t like having spaces in filenames. This is a hangover from old days of MS-DOS, but is consistently useful today when using R and other coding software. However, some software platforms react oddly with underscore (_), and so you might need to work with a hyphen instead of a space (-).

Once I’ve commented on a file and sent it back in an email, or uploaded it to their folder, I’ll add my initials Swimming_performance_V1_JM.rtf. Each time I comment on this version, I will add a number to my initials and the student adds one back, so eventually we might come to: Swimming_performance_V1_JM3_MM2.rtf. After three rounds, we would normally send a clean V2 out to co-authors and once their comments are back in we should be all set.

While you and your advisor can adopt any naming convention you like, you and they need to instantly be able to distinguish between a new version and one that they have already commented on by the file name.

### How many versions of each file should you keep?

I would suggest keeping at least one older version of each file to minimise the possibility of losing the content through corrupt files, but you should also be keeping those *.rtf or *.txt files as well, just in case. Because of my early experience with file corruption, I started to keep all versions of files. This is unnecessary, and I’ve never needed to go back to anything but the last version. Nevertheless, text files take up very little space and it’s not a problem to keep more versions if you want to.

### 2021-04-15 2.11-paywall.Rmd
#First upload of book

## How to get around the paywall {#paywall}

Once you've understand where the money for science and publishing comes from, you will understand that there is no real shame in trying to access papers that are behind a paywall. Essentially, I’m going to recommend that you move from legitimate to illegitimate, and then if all else fails to get an interlibrary loan. This is last on the list as someone will have to pay for it. Your library may have an allocation that they can use before the bill comes to your advisor’s desk.

### Go to your library

Unsurprisingly, this should be your first port of call. It’s probable that your library has an electronic portal that you can use to access all of their subscriptions. But don’t forget that they have stacks with old copies and you might need to physically enter the building. Try it. They are great places!

### Look for links from Google Scholar

Google scholar has lots of links (on the right) labelled with “[pdf]”, and these are often integrated with your library. If there’s nothing at first, click on the “All 9 versions” link (below and to the right). This may reveal that there is more than one version with a pdf, and you might have access to one of the others. Check for a link to the author on google scholar, and then look for a link to their website.

### Google the title in quotes
Sometimes, Google scholar won't show a pdf, but you can Google the title "in quotes" and get back a direct link to the pdf. Although rare, it's worth a try before moving on.

### Go to the author websites and look for links

Authors often have official university websites, and sometimes their own home curated sites that may have pdfs of all their papers. Use the clues at your disposal, including a general search engine, and follow the trail.

### Go to an academic social media site

Many academics also have accounts on academic social media sites (e.g. Academia.edu; ResearchGate.net). You will need to join the network and have your own account, but these house a lot of pdfs that you can’t get elsewhere.

### Go into the black open access

There are a few of these, and some have plugins to web browsers. You may need to be away from campus and using your own machine, or in an internet cafe to access them (a lot are blocked by universities through pressure from publishers). For some, like [Sci-hub](https://sci-hub.se/), you will need the [DOI](DOI.html) (you can get this from the publisher’s website). At the time of writing, Sci-hub has >85,483,812 papers. [Unpaywall.org](https://unpaywall.org/) has >28,112,995 papers. [Open Access Button](https://openaccessbutton.org/) don’t say how many they can reach, but “millions” of papers. And [freefullpdf](http://www.freefullpdf.com/) is a website that searches the internet and indicates where you can download pdfs (but sometimes you might hit a paywall).

### Use your contacts at other universities

The chances are that you have friends at other universities where the subscription you are looking for is not behind a paywall. If not then ask your advisor or labmates, or post it on your lab’s (electronic) notice board.

### Go back to you librarian and ask them for help

Librarians have contacts at other libraries. Some have had careers at other institutions or have met people at conferences who’ll be prepared to help them. Librarians are great people and well worth getting to know. Their job is to help you get stuff, so it’s always worth engaging with them.

### Get an interlibrary loan

This comes last on my list as it costs money. If what you are after is a book, then you might end up going here very quickly as many older books are not available online. If it is a paper, it will likely be copied or scanned into a pdf where it occurs. If it is a physical loan, the interlibrary loan will be time limited. When filling out the form, make sure you know where the loan will be billed to before you submit it. And don’t come here first, the chances are that you can get it somewhere in the list above much faster and cheaper.

### 2021-04-15 2.12-taxonomy.Rmd
#First upload of book

## Scientific names and taxonomic authorities {#taxonomy}

There are quite a few conventions that it's worth being aware of when it comes to using names. When talking about common names, it is a bit of a quagmire as there are few standards that are followed by all journals. Scientific names can also be in great (taxonomic) flux, but at least there is definitive help. Here I will outline my interpretation of what's what in using names.

### Scientific names

There is only one valid scientific name for a species. This is expressed in a binomial: two names, first the genus (with an initial capital letter) and then the species name (lower case). This name is italicised by convention (or in some cases underlined). Note that no other taxonomic names (e.g. Family, Order, Class, Phylum) are written with italics (although they all get capital first letters), only the genus and species are italicised. The single name is a hard rule as taxonomy can't survive in any other way. The most recently published version is the one to use, but happily with many vertebrates [mammals](https://www.mammaldiversity.org/), [birds](https://avibase.bsc-eoc.org/avibase.jsp?lang=EN), [fish](https://www.fishbase.se/search.php), [reptiles](http://www.reptile-database.org/) and [amphibians](http://research.amnh.org/vz/herpetology/amphibia/) there are well curated databases to which you can refer for the latest taxonomic treatments. Also see [WoRMS](http://www.marinespecies.org/) for marine species. My suggestion is that you defer to these (with citation) and you won't go far wrong. If you do this early on in your thesis, remember that you may need to update it later.

#### Consider this example:

When my former student, Mohlamatsane Mokhatla, started his PhD in 2013 he worked on three species, and over time these changed so that at various points in his thesis. You can see the combinations he had to contend with in Table \@ref(tab:taxonomy).

Table: (\#tab:taxonomy) **Taxonomy is dynamic which means that species names change over time.** The way in which genus and species names can change in only a period of 5 years.

| ![](figures/S-capensis.png) | ![](figures/A-quecketti.png) | ![](figures/X-laevis.png) |
| ---- | ---- | ---- |
| *Amietophrynus rangeri* | *Amietia angolensis* | *Xenopus laevis* |
| *Amietophrynus rangeri* | *Amietia quecketti* | *Xenopus laevis* |
| [*Sclerophrys capensis*](https://amphibiansoftheworld.amnh.org/Amphibia/Anura/Bufonidae/Sclerophrys/Sclerophrys-capensis) | [*Amietia delalandii*](https://amphibiansoftheworld.amnh.org/Amphibia/Anura/Pyxicephalidae/Cacosterninae/Amietia/Amietia-delalandii) | [*Xenopus laevis*](https://amphibiansoftheworld.amnh.org/Amphibia/Anura/Pipidae/Xenopus/Xenopus-laevis) |

The taxonomy of two species changed such that he had to constantly revise the names in his thesis. Luckily, he could simply conduct a "replace all", but it did become quite confusing. There's a chance that they've changed again, so I've provided hyperlinks for you to check out what the current name is. The lesson is that if you want to avoid this, work on *Xenopus laevis* or conduct your PhD in as short a time as possible!

In taxonomic papers, and certain journals with a taxonomic focus, there is an insistence that you cite the "taxonomic authority" after the species name when used for the first time. This is essentially a citation in which you acknowledge the original description that accompanies the name. Let's take Mohlamatsane's examples again:

> *Sclerophrys capensis* Tschudi, 1838             
> *Amietia delalandii* (Duméril and Bibron, 1841)     
> *Xenopus laevis* (Daudin, 1802)

It might seem that there's a mistake above, because Tschudi, 1838 isn't in brackets. This is actually deliberate. Tschudi, 1838 isn't in brackets because the genus name hasn't changed since the description by Tschudi [1838](tschudi1838classification) placed *S. capensis* in *Sclerophrys* in 1838. But it is clear in the table above that the genus name did change to *Amietophrynus*, and it was recognised under a different species name (*A. rangeri*). Unlikely as it may seem, the original description of the species was by Tschudi in 1838, but this was long forgotten (or ignored because the type specimen was so poorly preserved), and the species was regularly referred to as *Bufo rangeri*. When the genus *Bufo* was broken up into lots of newly named genera by Frost et al [-@frost2006amphibian], Tschudi's description was once again disregarded [@ohler2016identity]. However, in 2016 new work on the type specimen revealed that it was the same as *Amietophrynus rangeri,* but because Tschudi's description was older, the taxonomy deferred to the rule of precedence. If you want to read more about it, you can look at all the changes on the AMNH amphibian database.

To clarify, the use of brackets around the taxonomic authority only happens when the species is no longer in the genus in which it was originally placed by the person who described the species. Daudin [-@daudin1802histoire] originally described *Bufo laevis*, and Duméril and Bibron [-@dumeril1841erpetologie] originally described *Rana delalandii*. Hence you will see that both of these authors are within brackets after the names of their respective species.

The generic name of a Latin binomial is often abbreviated to the first letter (e.g. *X. laevis*). This should only be done once it is well established in your writing what the full name is. Or, in other words, you should refer to the species in full on the first mention, and then switch to the abbreviation. Because sentences shouldn't start with an abbreviation, you might need to relax this from time to time. You also need to treat the abstract as a stand-alone document in this respect. You might find that you have two species, which belong to different genera that share the same first letter. In this case, most authors use a different abbreviation: for example *Ap. rangeri* and *At. angolensis*. This is exasperated when both species names are the same. Keen readers would have noticed that Mohlamatsane doesn't have this problem any more!

I have seen some comments that suggest that once a generic name is abbreviated in a paper, it shouldn't be written in full again. In my opinion, this is rather a silly idea. It is up to you (the writer) to avoid ambiguity. If there is no ambiguity, then you may write the genus and species in full to increase the readability of your piece.

You should not start a sentence with an abbreviated genus. *X. laevis* would not be good, because your first word of the sentence is abbreviated (*X.* ). *Xenopus laevis* is a perfectly admissible way to start a sentence.

Lastly, there is a convention that when referring to the generic name alone you should always mention that it is a genus. So you shouldn't write about those *Bufo* that hop around the house, but instead about those members **of the genus** *Bufo* that hop around the house.

### Different codes for the nomenclature of different taxa

Nomenclatural codes exist to ensure stability in names. At first glance, this might seem like quite a trivial (academic) matter. But think about what happens when a particular species is named in law. That name could (in theory) be changed by some unscrupulous biologist to open up a legal loophole, or at least some ambiguity. Once you are aware what can go wrong with taxonomy and the implications that it can have on many lives, you will understand the need for stability. Happily, there are associations of very serious people who guard the stability of taxonomy through the codes. It is important to realise that there are codes governing the naming of different levels of taxa.

 - Viruses – [International Code of Virus Classification and Nomenclature](https://talk.ictvonline.org/information/w/ictv-information/383/ictv-code)
 - Bacteria & Archaea [International Code of Nomenclature of Prokaryotes](https://www.the-icsp.org/bacterial-code)
 - Algae, Fungi & Plants [International Code of Nomenclature for algae, fungi, and plants](https://www.iapt-taxon.org/nomen/main.php)
 - Animals [International Code of Zoological Nomenclature](https://www.iczn.org/the-code/the-international-code-of-zoological-nomenclature/)

There are many peculiarities for each taxon, and if your PhD involves taxonomy explicitly you should consult the specific code. Otherwise, it's unlikely that you'll need to be aware of the contents of each or any code.

#### Peculiarities of plant taxonomy

Plant taxonomy is considered the taxonomic mother discipline as the work of Linneaus in 1753 was the first botanical work that consistently used Linneus' binomial (two names: Genus species) nomeclatural system [@linnaeus1753species] which pre-dated the more inclusive (and better known) landmark publication in 1758 [@linnaeus1758systema]. In addition to the Latin binomial name (Genus and species), plants (and some other taxa) often require further descriptors or **ranks** that denote subspecies, varieties, subvareties, etc. I have included a (non-exhaustive) list of these in Table \@ref(tab:plantrank), together with some examples.

Because Linneaus named so many plant species is his landmark publications, instead of writing out the full taxonomic authority after the species names, botanists abbreviate this to "L.". Like Linneaus, other prolific plant taxonomists can have their names abbreviated, but plant taxonomists became so numerous that there needed to be some standardisation of names and their abbreviations. A database of these **Standard Forms** can be found at [the International Plant Names Index database](https://www.kew.org/data/authors.html). Note that Standard Forms in plant taxonomy can include initials.

Table: (\#tab:plantrank) **Plants often need more information than the simple Latin biomial.** This table provides the required abbreviations that you will need to add in order to properly describe the rank of your plant taxa.

| Rank | Abbreviation	| Example |
|-----|-----|-----|
| Subfamily | subf. | Poaceae subf. Pooideae |
| Subgenus | subg. | _Metrosideros_ subg. _Metrosideros_ (Myrtaceae) |
| Section | sect. | _Senecio_ sect. _Senecio_ (Asteraceae) |
| Subsection | subsect. | _Stachys_ subsect. _Swainsonianeae_ (Labiatae) |
| Series | ser. | _Mimosa_ ser. _Paucifoliolatae_ |
| Subseries | subser. | _Palicourea_ subser. _subcymosae_ |
| Subspecies | subsp. | _Trichophorum cespitosum_ subsp. _germanicum_ |
| Variety | var. | _Closterium acerosum_ var. _elongatum_ |
| Subvariety | subvar. | _Hieracium vulgatum_ subvar. _oblongatum_ |
| Form | f. | _Prunus cerasifera_ f. _atropurpurea_ |

Some plants exist in associations (such as lichens which are an association between an algae or cyanobacteria and one or more fungi. For these associations there is a specific code for their nomenclature: International Code of Phytosociological Nomenclature [@theurillat2021international].

There is also a code for cultivated plants, and if you study any kind of agronomical system, you will be familiar with not just the normal binomial name, but also a specific cultivar or variety (*var.*) that follows the binomial: [The International Code of Nomenclature for Cultivated Plants](https://www.ishs.org/scripta-horticulturae/international-code-nomenclature-cultivated-plants-ninth-edition).

### Common names

Common names have a different set of problems to those that we've seen for scientific names. These are largely focussed on: what name to use, and whether or not to capitalise the common name.

What name to use - This is tricky. My suggestion would be to take the most commonly used common name, as that's likely to be the one that most people will recognise. However, it's not always that simple and because there's no "correct" common name, you have some leeway to choose the one you prefer.

Let's take an example: The African Clawed Frog is a very widely used name for *Xenopus laevis* which is used all over the world as the model amphibian. However, African Clawed Frogs is also the commonly accepted name for the entire genus, *Xenopus*. Thus, shouldn't we avoid using this for the species *X. laevis*? So, let's look at the alternatives. Frost [-@frost2020amphibian] lists 13 common names for this species. Names such as Platanna and Common Platanna are difficult to use in scientific papers as they are only regularly used in South Africa. Clawed Toad is not appropriate, as it is not a toad (reserved for members of the family Bufonidae), a confusion that dates right back to Daudin's 1802 description. Clawed Frog has the same problem as African Clawed Frog, and other variants such as Upland Clawed Frog and Smooth Clawed Frog also seem inappropriate as this species is neither confined to uplands, nor the only member of the genus which is smooth (although 'laevis', the species name, does mean 'smooth').

There has been a recent movement to formulate common names into genera (or even families) such that the specific common name takes on a specific epithet of the generic or familial common name. In a world of stable taxonomy (such as that enjoyed by ornithologists), I can understand that this is possible. But imagine the requirements in common name changes for Mohlamatsane in his thesis over only a few years. Surely the useful thing about common names is that they are used commonly by normal people who don't worry about taxonomy? If we insist on common names constantly changing, don't we risk alienating the public?

I eventually decided to use 'African Clawed Frog' as this is the most widely recognised common name for this species. It does become an issue when writing about *X. laevis* and other *Xenopus* species (as we have), but we've managed to resolve this without too much difficulty.

### Should common names have capitals?

**Warning: lots of people get very upset about this point and there are two views:**

#### View 1.

Common names are proper nouns and proper nouns start with capitals. The Western Leopard Toad is not any toad, and the African Clawed Frog isn't any frog. The Cape platanna is any platanna living in the Cape, but the Cape Platanna is *Xenopus gilli*! In this last example, you can see how capitalisation reduces ambiguity and differentiates between adjectives and nouns.

#### View 2.

Common names are not formal names (like scientific names) so why does it matter? Many journals do not capitalise common names and insist on lower case unless they are proper nouns, like African in African clawed frogs. This seems to have arisen from editors who dislike superfluous capitalisation.

On balance, I'd say that it's likely that your usage will be dictated by the policies of your faculty, advisor or journal.

### 2021-04-15 2.13-style.Rmd
#First upload of book

## Writing style {#style}

**Did you ever feel that reading a paper was like trying to solve a puzzle?**

During my PhD I read a lot. My study species was the subject of tens of thousands of papers, and I was convinced that I'd find what I'd need somewhere in those musty old reprints and heavy volumes that came from the library shelves. It was easy to believe that the authors of those papers had set great puzzles for me to try to understand their content. The realisation that this is not the best form of communication came as a personal epiphany some years ago.

> “The goal of good writing is straightforward: to make your reader's job as easy as possible” 
>      Kevin W. Plaxco [-@plaxco2010art].

This great quote from a paper by Plaxco [-@plaxco2010art] cuts to the heart of why it’s essential to write well. As an editor, I once experienced an author who thought the opposite, and for me this was an epiphany. The submitted manuscript was an impenetrable mess. I could tell that there was good work in there, but as an editor I felt that the information had been made so obscure by the authors that my readers were unlikely to get much out of it. The author was a colleague, and so I decided to phone him and chat about the need for much greater clarity in his manuscript. Experience with email has taught me that they can often be taken in the wrong way; usually in the worst possible way.

The response I got surprised me. The author recognised that his text was dense and was unapologetic. “Let the reader work to make sense of my data,” he said. My epiphany came not because this was a totally alien concept, but because during my PhD days this was exactly how I had thought scientific writing should be done. I had spent so many hours slogging through dense and dreary papers by well-respected figures. The reward when I finally understood what it was that they were trying to communicate felt so great that I believed the puzzle they had set me was what I should set my own readers. Happily, my advisor didn’t feel the same way, but for years I continued to believe that a paper should be a decent puzzle for my readers to crack. This might seem a crazy concept, but it’s not an uncommon problem. For a great take on this in Chemistry, see this great article by Murray [-@murray2011skillful].

This has led to a general perception that "academics stink at writing" - as Steven Pinker put it. But, I think that Pinker is outdated. Certainly, as Plaxco [-@plaxco2010art] has it, there are a lot of ways that academics can improve their writing, for each other. But the belief that all academic writing is going to be accessible to the general public is far fetched. This is because there are a lot of terms when writing journal articles that are there for precision and brevity. These terms will exclude a larger audience, especially when they get so dense that looking them up is simply too arduous. What Plaxco [-@plaxco2010art] urges us to do is to make the rest of the text as accessible and easy as possible.

Any good writing is an artform, and that includes good academic writing. You can't expect to be the best until you've spent some time at it.

There are still authors out there who attempt to set puzzles for their readers, but they aren’t in the mainstream any more. Instead, biological sciences has some inspiring writers, and many more are taught how to inspire future generations towards communicating great science. That’s not to say that there are no puzzles left. Much of what we do requires great puzzle solving abilities. However, let’s keep the puzzles away from communicating with our audience, make their job as reader easy or even pleasurable, and they’ll keep coming back for more.

### Don’t bamboozle

It’s easy to use jargon. The whole point of jargon is to convey a (usually complex) idea in a short amount of space. Using a word (or two) instead of using several sentences clearly has some advantages. However, there is such a thing as too much jargon. Simply put, it’s unnecessary to use jargon when you can use plain English in the same amount of space. My old tutor at Liverpool University, the late, great Brian Moss, shared the following example of too much jargon [@bump2009large], when plain English would have been much shorter. Do yourself a favor and read the abstract of Bump et al [-@bump2009large]. The fact that it gave Moss a chance to write about moose wasn’t lost on anyone (see Figure \@ref(fig:moss))!

(ref:moss) **Bamboozled Moss takes on the moose.** Brian was so unhappy after reading Bump et al (2009) the above that he felt compelled to write a letter to the British Ecological Society Bulletin.

```{r moss, echo=FALSE, out.width = '95%', fig.cap="(ref:moss)"}
knitr::include_graphics('figures/moss.png')
```

### 2021-04-15 2.14-voice.Rmd
#First upload of book

## When is it possible to retain your own voice when writing? {#voice}

Should one use “I” or “we” when writing a scientific paper? Although there was a tradition not to use first-person pronouns when writing scientific articles, this has fallen by the wayside in recent years as the use of I or we makes writing simpler for writers and readers. Use “I” if you are the sole author (or in your thesis) and “we” for two or more authors.

You will regularly see “I” or “we” being used in the last paragraph of the introduction and the first paragraph of the discussion. It’s also quite common to use this in the methods section. The reason why these sections have “I” or “we” is quite obvious, as it allows you to place your own voice on actions and decisions. After all, it is your aim and hypothesis for your study, so you should own it. Using first-person pronouns also allow you to be more concise, and there are some great examples here where you can see what happens when you don’t use them.

The first use of “I” or “we” at the end of the introduction is powerful. Up to that point, the reader has read four or five paragraphs of logical arguments outlining the background information that makes up the reasoning for your study. Using the first-person pronoun now makes the reader sit up as they have you communicating directly with them what you aimed to do. But that power is lost with repetition. So if you were to continuously use it throughout the methodology, the reader may get tired of reading a string of statements stating that you did this, then you did that, and so on. The power of using “I” or “we” is to use it sparingly to boost clarity.

### Does writing “I” or “we” help people to understand who you are?

Actually, all of your writing will do this. As long as you are using your own words (and not plagiarising), your writing is likely to be unique to you and hence recognisable as you. It never ceases to amaze me that it is possible to convey the same information in so many different ways. Your way will likely change with time and experience, but remain yours throughout your writing life. If you keep writing, people will be able to recognise your distinct style.

### The Vonnegut advice

Kurt Vonnegut [-@vonnegut1980how], infamous novelist, provided some advice for writers. First, he made an impassioned plea for technical writers (yes, you) to use your own writing style. The easiest way, Vonnegut argues, is to write about a subject that you care about. He's totally correct on this, and I'm hoping that after all your research you care passionately about your subject. Next, not to ramble, but to keep your writing simple.

What is most impressive about Vonnegut's advice is that he argues from the readers' standpoint. His parting advice:

> "Our audience requires us to be sympathetic and patient teachers, ever willing to simplify and clarify - whereas we would rather soar high above the crowd, singing like nightingales."
> Kurt Vonnegut

Having said that you should write about subjects that you feel passionately about, this does not mean that you should write with unsubstantiated passion. For example, it makes no sense to write: "The threats to this spectacular species surpass any mechanism known to humankind" ...unless you have substantive citations or results to back this up. Yes, it's passionate, but that should be balanced by the need to keep what you write factual.

### 2021-04-15 2.15-concise.Rmd
#First upload of book

## Writing concisely {#concise}

So far, I have put an emphasis on a writing formula in order to get started writing. Many students have a problem in knowing where and how to get started with what seems to be a daunting prospect: writing a proposal or writing the entire thesis. I still feel that getting started quickly and efficiently with some confidence is one of your most important steps. But once you've made it through your first draft, you'll begin to find that you need to start refining your writing. That is saying what you need to say in less words and space.

In an ideal world, what goes into your thesis will be of the same quality as the accepted manuscript. The reality is that the thesis is often more wordy, with the result that reviewers and/or editors will ask you to be more concise. How can we achieve this?

Someone else has written a great essay with 10 top rules to help you be concise. The essay is Open Access, and there's no need for me to repeat the words of Scott Hotaling [-@hotaling2020simple]. However, I am going to write some comments of my own on his rules. There are some other posts that are also worth a second look along with this one: see here and here.

Here, the rules are Scott's, the comments are mine:

### Rule 1. Take writing seriously

It is likely, that like me, you chose to do science at school instead of useful subjects such as English and/or other languages. It might have also occurred to you that it was a really bad idea not to have done more earnest work in languages now that you are expected to write like a professional. Certainly, if you did pay more attention in languages in school, you may not be suffering with the rest of us now. Personally, I never received tuition in the grammatical construction of English. Thus I'm left in the dark most of the time about exactly what is wrong. Instead, I simply try to rework a sentence into what I know is correct [and I often don't achieve that]. The lesson then is that should you come across an opportunity to learn more about writing, do take it. 
Although we talk about writing science here, the grammatical rules of writing (in English) are fairly universal (with some local exceptions), so do practice your writing. Maybe write a letter (or long email) instead of making a call sometimes. It'll also help you to read (see here). If you are a keen reader, then you could do worse than reading some Kurt Vonnegut. Scott provided this link to an article I hadn't seen before. Go on, give it a read!

### Rule 2: Identify and stick to your message

It's important to avoid distracting your reader (see [Part II](part2.html)). Your aim is to be thorough, which will mean including all relevant information, but don't allow your writing to sideline them into taking the wrong direction. You need to stick to your narrative like a highway. The highway analogy is useful as although you will see signs to other places that you can point out (to the reader), you shouldn't turn off the highway. The best of highways will also be free of traffic jams (think long, complex sentences here). and toll free ([no paywall](paywall.html)).

This will mean that you may need to delete some of what you've written, and that can be hard (especially when they are the best bits). You can always keep a file with all the best bits that you've never used. Maybe one day you'll use them. Or maybe one day you'll see that they weren't quite as good as you thought they were.

### Rule 3: Get to the point

If the highway is 'your message' then getting to the point is the big sign that states the destination.

Some of the best papers I've read manage to encapsulate the whole point in the first sentence, or better still in the title. Our formula has you getting to the main question in the last paragraph of the introduction, but you should have already 'got to the point' in the first paragraph - i.e. the point that is the bigger picture.

### Rule 4: Keep your Methods and Results contained

You can find a guide to writing your methods in [Part III](part3.html). This rule is about not allowing the methods to creep out of this section and into the results, or even the discussion. Similarly with the Results. There are some specific times when this is permitted (such as a post-hoc test), but generally you shouldn't expect to do this.

### Rule 5: Do not repeat yourself (too often)

Redundancy is often rife in proposals, theses and manuscripts. If you've produced a table with all of the results, then they don't need to be in the text. The same with a figure, especially Figure 1 which is often a descriptive map or diagram of apparatus. Have it once, but you don't need it twice.

Copy and pasting is very easy, and a good way of suddenly producing vast quantities of text. But repeated text is quickly recognised by the reader and appears very boring and cumbersome. This is likely to happen in the Methods and Results sections. If you find yourself deciding that you can simply cut and paste this paragraph while changing the variable names and the numbers, then you are wrong. Don't ever do it.

Also, please don't cut and paste sections of methods from one part of your thesis to another. Just don't do it.

People also have a tendency to build the abstract by copying and pasting text from the main sections. Don't do it. The reader will quickly see the repetitions and become bored. Similarly, the conclusion/summary section is also often copied from lines above. If it's not worth writing again, then it wasn't worth writing before. Make it fresh, and keep it interesting!

### Rule 6: Avoid unnecessary or inefficient “lead‐ins”

Interesting. I'd say that you need some practice and a critical eye to spot these kinds of errors. Having your advisor help or someone who has edited a lot of text. Scott's got some great examples, and so I'll let you peruse them here.

### Rule 7: Use first‐person, active voice

For some reason, most students avoid this at all cost. There appears to be the idea that saying "I" or "we" isn't correct for scientific writing. In fact, it can be the easiest way to avoid the passive voice. I've written more on this elsewhere... see here.

### Rule 8: Remove unnecessary words

This is really getting down to the nitty gritty. It's hard to do this yourself. It's much easier for someone else to show you. Typically I only do this level of word-smithing for abstracts or when the imposed word limit means that you really need to remove excess words. However, Scott is correct that if you can learn to do this yourself, it will improve your writing.

Personally, I like to slip in the odd 'as well as' instead of 'and', just to ring the changes. Scott would remove them, and you can see where he'd edit out other examples. Do we really have to be so hardline? I'd say that there are times when it helps.

### Rule 9: Simplify your language

This is always a good idea. Making three short sentences instead of one very long one is much better. It also helps you avoid complicated grammatical clauses.

### Rule 10: Seek and embrace feedback

I spend a lot of my time reading and commenting on the writing of my students. Every comment is made to improve the document. I do get mad when I find that this help is ignored without an indication why. My suggestions or comments are simply deleted. Certain students become very hard work when they don't even explain why they refuse to change something. You should have noted by now that it's all about being flexible, and being stubborn about anything you've written isn't going to work. Reading the work of these students quickly becomes a chore.

### Rule 11: Read it yourself

Here's a rule that Scott didn't have. Perhaps it should have been inserted higher up the list. As I've said elsewhere (here for example), it would help you to have your work read through by one of your colleagues before you give it to your advisor. But above all, you must be prepared to read __all of your work__ yourself. Not just the first time you write it, but for **every** version. Please don't ever expect your advisor to read something through when you can't be bothered to read it yourself.

Being concise is a great way to write well, and the abstract is a case in point.

Please don't forget to read Scott's 10 rules as written by Scott [@hotaling2020simple].

### 2021-04-15 2.16-foreign.Rmd
#First upload of book

## Writing a PhD if English is not your first language {#foreign}

Reading this book you will learn of the many biases in biological sciences and how you need to be aware of them in order to properly interpret your work. Be aware also that biological sciences is biased in where studies take place, and who publishes them [@culumber2019widespread; @nunez2021monolingual]. What we really need then are more studies from less studied parts of the world, and especially we need these to be researched and published by people from those countries. Why then do we insist that science be written in English?

English may be one of the most difficult languages to learn how to write as a foreign language because rules are so inconsistent, and the vocabulary is so large. At 171476 words, English has an enormous vocabulary, followed closely by Russian (150000), Spanish (93000) and Chinese (85568: although you should note that counting words isn’t very accurate and as for counting dictionary entries, Korean wins). I cannot defend the use of English as the language of science, but I can assert that having a universal language for communicating science is massively helpful, and makes a large swathe of the literature available to those scientists who are prepared to become proficient in a second language. However, none of this really helps you to prepare writing in English as a second language.

One of my 'pet hates' is that many non-English speaking editors seem to insist that you have your work checked by a 'native English speaker' (see Figure \@ref(fig:editor)). This suggests to me that some of the worst people to comment on whether or not something is written in good English, is another foreign language speaker. The lesson here is that you should be careful who you trust. Moreover, in my experience, having advice from a 'native English speaker' is no guarantee to getting a well written chapter or manuscript. Most of the world’s scientists did not grow up speaking English. Yet, rightly or wrongly, English is the language in which science is currently written. So, if English isn’t your mother tongue, should you expect to receive help during peer review?

### Message from an editor

(ref:editor) **The message from the editor: "your work should be checked by a native English speaker"** Should you expect to receive help with your English when you submit a manuscript?

```{r editor, echo=FALSE, out.width = '95%', fig.cap="(ref:editor)"}
knitr::include_graphics('figures/editor.png')
```

Perhaps I should start from the outset by stating that English is my mother tongue, and that I have spent many hours correcting the language of colleagues for whom it wasn’t. However, most of these hours were spent when I was a postgraduate student or postdoc. I no longer think that correcting English is my role either as an editor, reviewer or as an advisor.

As I’ve stated above, I did not study English, and would be the first to admit that my grasp of grammar and syntax is by no means perfect (those of you with better English will have picked this up by now!). I have read enough correct English to know when something is incorrect, but that doesn’t mean that I know how to correct it. I have spent many long hours trying to decode what others have written, and in some cases this has involved me rewriting entire manuscripts. I still do this as a co-author, although I do remember asking one colleague to please send any further drafts in their native Spanish as it would be easier to translate than it would be to re-write.

The time component is at the crux of my reasoning why, as an editor, reviewer and advisor, I will not provide an English language service. It is both a time consuming and an unsatisfying experience. We all have our own voice, and correcting while maintaining other people’s voices is a painstaking task. There’s an entire profession that specialises in this (think translator). These days there are also services available from publishers to non-English authors to help them correct their English.

In addition, there turns out to be research suggesting why peer review is not the best way to improve writing [@shashok2008content]. Peer review works much better at screening technical content than it does at improving the communication of that content. And this agrees with the opinions of some non-English speakers who ask for editors and reviewers to concentrate on the science, a sentiment with which I agree.

Given that I’m not going to correct your English, what remarks would you expect me to make when reviewing a poorly written manuscript?

No matter how you might hope that it isn’t true, a poorly written manuscript will not get a good review. If your reviewer is struggling to understand what you have written, this becomes the overall impression that they will likely pass onto the handling editor. I try to separate my difficulties with English from my review of the science. But this isn’t always possible. Frequently, a poorly written manuscript will mean that I won’t be able to understand why the research was undertaken, what was done or what it means. This is bound to impact the review negatively.

The more I struggle to read, the more negative the review becomes. I see this as inevitable. What shocks me is that some senior researchers consider it to be their right to submit poorly written manuscripts and have reviewers or editors correct them [if you don’t believe me, see @statzner2010negative]. Worse, I’ve received manuscripts that are co-authored by people I know are native English speakers, but are full of glaring mistakes that appear never to have been checked. For me, this violates the terms that all authors have approved the final manuscript.

But I've also got to admit that it's not easy. There are other places where you might be better to go to get advice on writing a paper or thesis as a second (or even more!) language [see @firth2017writing]. I'm full of admiration for those of you that have to do this.

### Must a ‘native English speaker’ check your work?

Being a native English speaker, I am not above having people complain about my English. I do get it wrong, and my English can often be improved. I’m always happy to receive help, and see it as a sign of how I can improve the clarity of a manuscript. However, I hope that my manuscripts are never so poorly written that a reviewer or editor cannot make sense of them. So it is a matter of degrees.

There are many people who are not native English speakers who write far better than I do. My ‘pet hate’ is that reviewers and editors insist that a manuscript must be corrected by a ‘native English speaker’. I’ve seen so many very poorly written essays, theses and manuscripts written by native English speakers that I know that having one correct your manuscript is unlikely to be of much help, unless they themselves are good writers.

Being a ‘native English speaker’ doesn’t automatically qualify you to write well, edit well or do any of the things that non-English speaking editors think that it does.

That non-English speaking editors often comment that a ‘native English speaker’ should read my text, simply underlines the problem that many editors themselves are incapable of knowing whether or not something is well written.

### So if not a native English speaker, who should do it?

Anyone you know who writes good English and is willing to help you. One option is the service offered by **AuthorAID** (http://www.authoraid.info/en/). Using AuthorAID, you can find a long term mentor who will help you with your English. You can read more about this approach here [@freeman2006publishing].

Failing that, I’m afraid that the best route will be to pay for help.

### But surely my native tongue can be used for something?

Yes. It is essential that you use other languages to communicate your science with the wider community. Most of the world doesn't speak English, and so there is a real need for this to be done in your language. See a nice blog from the LSE on this subject here.

Lastly, consider the words of

>Kurt Vonnegut [-@vonnegut1982how]:
>"No matter what your first language, you should treasure it all your life. If it happens not to be standard English, and if it shows itself when you write standard English, the result is usually delightful, like a very pretty girl with one eye that is green and one that is blue."

### 2021-04-15 2.17-plagiarise.Rmd
#First upload of book

## Making sure that you don’t plagiarise {#plagiarise}

Essentially plagiarism is when you copy somebody else's work. We're most commonly familiar with plagiarism of writing and these days this is especially easy with the copy and paste function. Most people are not even aware that they have plagiarised because sometime in the past they copied and pasted the work of somebody else into a document and this later became incorporated into their text without them being aware of it.

Plagiarism is a problem because essentially you are taking somebody else's work without attributing it to them.

### How to know if you have plagiarised

Today there are several pieces of software that are used to scan text that's written and available on the internet to discover plagiarism. One such example is Turnitin. One of the outputs of TurnItIn is to highlight text that matches other text already on the internet. As almost all journals published on the internet TurnItIn can accurately determine if text has been copied from another article or website.
I usually set TurnItIn to determine plagiarism with five or more consecutive words.

Remarkably it is very difficult to come up with exactly the same words that someone else used to describe a phenomenon. Most people when they think about it feel that it wouldn't be that surprising if they came up with exactly the same words as somebody else.

When caught plagiarising most undergraduates claim that they simply read an article and then later happen to write the same words that were in the article. They categorically deny that they ever copied or pasted text from the article into their work.

__Try it. When you try it you will learn what plagiarism is all about.__

Read an article, and then try to write text that is exactly the same as that in the article without looking back at the article itself.

Unless you have an eidetic memory you will fail at this task.

(ref:plagiarism) **It's always tempting to copy and paste, but it is likely to lead to plagiarism.** Copy (Ctrl + c) and paste (Ctrl + v) have become so easy that it is tempting to pick up portions of appropriate text directly from papers and then slot them into our own work (Figure \@ref(fig:plagiarism)). However, this is plagiarism and can easily be found by using software like TurnItIn. Most institutions will require checks for plagiarism on your thesis after submission, with dire consequences if your text fails.

```{r plagiarism, echo=FALSE, out.width = '95%',  fig.cap="(ref:plagiarism)"}
knitr::include_graphics('figures/plagiarism.png')
```

This is not to say that no 5 or more words can ever be the same as someone else's. There are situations in which this happens. Think of addresses, quotes, certain laboratory equipment or protocols, and certainly references at the end of your paper. So there are many times when TurnItIn will come back with matching text. This is not what we're looking for in plagiarism.

###  What to do if plagiarism is detected in your work

It's remarkably easy to remove plagiarism from your work.

__Here's what you do:__
- Read the sentence that has been plagiarised several times to yourself. 
- Now without looking at that sentence, write another sentence that has the same meaning. Because it's very hard to replicate somebody else's words without copying them, what you should find is that you've written a fresh plagiarised sentence. 
- This can now be added to your text, changed as appropriately to fit your existing text. And that should be the end of your plagiarism worries.

__Here's what not to do:__
- Take the sentence swap out some of the words for synonyms and pass it off as your own. 
- The sentence will still have the same structure that you copied and essentially this is still somebody else's work. Moreover TurnItIn will still recognise this as plagiarism.

### How can you make sure that you never plagiarise?
Quite simply if you never cut and paste, you won’t plagiarise. It’s that easy. I understand why people copy and paste as a way to get started, or because someone else has written something so well, it’s hard to believe that you could ever write it any better. But actually, you can write it just as well, and writing it in your own words is worth so much more.

Don’t forget that the penalty for plagiarism in your thesis might well be that you fail.

It still seems amazing to me when I submit a student manuscript to TurnItIn and see that it is completely free of any plagiarism. There are so many words in English (>643 000 according to the Oxford English Dictionary) and so many different ways of putting them that you really can have your own writing style. Your writing style will be as unique to you as a fingerprint, and it will be entirely free of plagiarism. It's something you can celebrate.

Want to read more about plagiarism to understand what it is: Louw [-@louw2017defining]

### What is grey plagiarism?

I would maintain that there is not really any grey area in plagiarism. So what are people talking about when they refer to a grey zone? The nominal grey zone within plagiarism comes when you copy your own text, this is sometimes referred to as self-plagiarism, but this also refers to the associated practice of ‘manuscript recycling’ [@geraldi2021self-plagiarism]. Essentially this is really just the same as plagiarism but obviously if you are the original author then if you use your own text it's not the same as using somebody else's, because the work is still yours. However this is frequently complicated by multiple authors on a paper. For example, if I wrote a paper with five other people, even though the words were mine, if I use them again they would actually be attributed to all six authors and not just to me. Thus, if I use them again with five different authors it would still be plagiarism.

This is a particular problem in biological sciences (and some other sciences) when it comes to the methodology section ([see part 3](methods.html)). Many students believe after having written the [Materials and Methods](methods.html) once, they can’t possibly write it again in another way. Actually, you can and you should. It’s good practice, but remember that you can also cite the methods in a previous chapter. For example, chapter 2 has one method written out in great detail. Chapters 3 and 4 then reference chapter 2 while giving only a brief section on that methodology. You should already have seen some examples of how people do this in the literature, especially when citing their own previous work.

This is further complicated by having different publishers for your papers. Let's say I wrote a paper, I was the sole author and it was published by a journal that retained the copyright. If I use the same text again in another paper where again I was the soul author, but it was published in a different journal and they retained the copyright, that would be a breach of copyright to the first journal. It would be regarded as plagiarism.

Obviously, this situation really shouldn't occur because if you are the author of your own work then you should retain the copyright it should never be given to a publisher or a journal. The fact that many journals want to obtain copyright on work that they don't pay for is a complete disgrace but it's something that we're not going to go into right now (you can see rants here and here about science publishing scandals).

### Manuscript recycling

Manuscript recycling is the practice of re-publishing the same manuscript (or with minimal changes) in different journals. Manuscript recycling is a bigger problem in other subject areas, notably economics [@geraldi2021self-plagiarism], publishing issues in economics appear to need a lot more professionalism.

### Overall - just don’t plagiarise

The simple way around this problem is not to plagiarise any of your own text (self-plagiarism) or any of anybody else's. This is, as I've explained above, relatively easy because as long as you don't copy and paste any text you'll find it very hard to actually use somebody else's words. However it can get a little bit tricky when you are writing the materials and methods section of a paper, and especially when that paper uses exactly the same materials and methods as a previous paper. The temptation is great to go and copy the text that you've written before. But be aware that, however grey, this is plagiarism and you should not do it.

The likelihood is that because you're using the same names and their relationship is staying fixed as shown in the example above in materials and methods you are more likely to come up with text that a piece of software like turn it in will show is plagiarised even when it isn't.

### 2021-04-15 2.18-phrasebank.Rmd
#First upload of book

## Academic phrasebank {#phrasebank}

It can be really tough to get started writing especially if you are writing in English as a foreign language. It could be that the academic phrasebank is just what you're looking for.

When reading papers, I often come across the same phrases time and again. They seem so well honed to the situation that I’m envious about how the writer dreamt them up. Stealing them isn’t quite [plagiarism](plagiarism.html), but then they don’t regularly fit the exact situation that I’m looking for.

Imagine cutting and pasting them into a file, or even better a database that’s ready for any eventuality. Banking them over time so that they are always available.

Then imagine that you did this and made it available to everyone…
http://www.phrasebank.manchester.ac.uk/

Well, John Morley has already done it, and he’s made it available to everyone. The academic phrasebank is a fantastic resource:

the academic phrasebank

### How to find the phrase that you need:

- Look across the top for different sections of text that you are currently writing.
- Then look down the left hand side to see the type of language function that you need. Click on something that looks applicable and watch the database open up.

### Getting started

It strikes me that this is going to be a fantastic resource for those of you who are not writing in English as your home language. And for those of you who just need a few words to get started. It’ll also be useful when you are wanting to wordsmith your introduction or abstract.

Is it plagiarism? Given that the chapter before is all about not copying and pasting, am I now telling you that it’s ok to copy and paste? Most of these phrases are less than five words or you will find that they are non-consecutive parts of a sentence. I don't think that most of these would constitute plagiarism if you pasted them into your document. However, I would suggest that you spend some time refining what you've written and that when you do that, some of these phrases may then come out.

### 2021-04-15 2.19-critical.Rmd
#First upload of book

## Why critical reading is crucial for improving your writing {#critical}

It is hard to emphasise exactly how important your reading will be, when it comes to your writing. If you are sitting with a blank page in front of you and feel that you have nothing to draw on, then think again. All of the reading that you have done to get to this point has already helped you more than you think. As I’ve pointed out elsewhere, your chapter or manuscript is likely to resemble the formula that many others already follow, and standing on the shoulders of those giants will help you again. But it’s possible that your reading isn’t helpful if you aren’t being sufficiently ‘critical’ when you read: and by critical I am talking about the methodical approach that you take to reading that improves your comprehension of a paper (Figure \@ref(fig:Criticalreading)).

(ref:Criticalreading) **Critical reading is seen as being made up of five important aspects**. This diagram shows how each of the five aspects feeds into the next.

```{r Criticalreading, echo=FALSE, out.width = '95%',  fig.cap="(ref:Criticalreading)"}
knitr::include_graphics('figures/Criticalreading.png')
```

### Making notes on good and bad styles

When you read, it’s worth making all sorts of comments about what you’re reading. It’s tempting to think that the only aspects of a paper that you should take notes on are the science (questions, approach and findings), but this is only part of your mandate as a critical reader. Of course you should be taking notes on the science, but you should also be using it as an opportunity to learn tips and tricks of writing. Once you are actively writing, and I’d suggest that this should have been right from the very beginning of your studies, you should be actively thinking about how the authors are writing. Are they writing well? Are parts written poorly? Then the critical part is to cogitate, albeit briefly: what works, what doesn’t and why.

Some papers will only offer the science that they contain. Others will stand out because you’ll realise that you enjoy reading them. As a critical reader, your task is to decide why you enjoy reading a paper:

- Is it the style? 
- Is it the way in which they formulate the logic of their argument? 
- Could it be that they’ve broken the standard formula? 
- Is it just the title, or the first sentence that you enjoy?

Once you’ve decided what it is, make these notes on the paper and file them on your referencing system (however you do it) under a style heading. I’d suggest that this be completely separate from your notes on the science. This is as simple as highlighting or making a quick note each time you see something that you like, or dislike (use different colours consistently). I used to underline text in pencil, and then make an illegible comment in the margin. It worked for me, not because I could read the comment, but because I could remember writing the comment.

You can also learn from badly written papers. Again, ask yourself exactly what it is about the writing style that doesn’t work for you? If you don’t know then comment in your database and make a group of papers that don’t read well. Do they have anything in common?

It won’t be every paper you read that you make notes about the style. It will probably only be a minority of papers in your area. However, this critical reading where you make notes on the ones that do stand out will help you greatly when it comes to your own writing.

#### Some examples

- Instead of using a taxonomic list (I don’t like lists) and long citation string to describe lots of different studies that had previously made the same point, a co-author (in a recent review study) organised them all into the same categories we were presenting in a review. The result was that the reader was reminded of our categories at the same time as seeing how well others had already covered these same topics.

- Being able to sum up a paragraph with a clincher sentence that also has a segway into the next paragraphs is a great skill to have

- Some authors are able to generate a theme that they develop over several paragraphs or pages

If you’ve spotted what it is that you like, write it down and commit it to memory. The very fact that you start to see things in other people’s writing that you like and admire means that you’ve managed to start reading critically. Keep this up, and it will help your writing no end as you open up your mind to critical reading.

You don’t have to restrict yourself in what you can learn while reading other people’s papers. Think about the stages in writing a paragraph. Look at the paragraphs you are reading and see if you can spot the subject and summary sentences. Maybe they are absent from the entire paper, or maybe they are present in every paragraph. Does it help enhance the readability of the paper? What about the times when authors break the rules? Can you see in your reading that the majority of authors follow rules? Do they follow the formula? What is the power in shaking it all up? If it works, try and analyse why it works. Then you will be reading critically.

### Make handwritten notes

When I started reading papers, I made photocopies (1 sided was as sophisticated as it got) from volumes of bound journal issues, and made notes in pencil on the back and front of those pages as I read. I understand that most of you will be working on pdfs which offer possibilities to add notes in text boxes that go alongside. But there still may be more advantages to writing notes using your own handwriting. A study by Mueller and Oppenheimer [-@mueller2014pen] showed that students who typed retained less of the content from a lecture than those that made notes by hand. They showed that this was because those who used a keyboard tended to take notes that were more shallow, while those that wrote their notes had a deeper thinking. My take on this is that the brain is making extra connections when you are taking notes by hand. Those extra connections will benefit you when it comes to trying to recall the study.

Another advantage that making notes by hand brings you is that you can deliberately associate the location in which you read, and practice your handwritten notes, with the place where you want to eventually write and recall what you’ve learned [see @bouton1993context]. Thus you can manipulate this facet of human memory to do all your learning, for example, at your desk in the department, or in the university library. Eccentrics may want to link it to reading in the bath, or another place of great comfort. But be careful not to make your ‘place’ too specific. You could move house or have someone else allocated to your desk, and that wouldn’t work out well. However you decide to make your notes, make sure that it works for you both in the present, and for the later stages of your PhD studies.

Similarly, if you make notes on style, you are likely to remember them in the same way that you remember the paper and its contents. An alternative to writing everything onto a printed copy is to make notes in a notebook. That way you get the advantage of building your memory with your own writing, and you can easily find everything in the same place when you forget where you read the fascinating insight. Alternatively, you may want to print the most important papers in your area, and make notes in pencil all over them.

It’s important to emphasise that you should not copy and paste parts of someone’s paper that you like. It’s easy to do, but any copying could be your undoing if you later forget and paste your ‘notes’ into your manuscript. It’s so easy to plagiarise like this that the only way to really avoid it is by being very strict with yourself. Making notes by hand will force you to look away from the written text while you write, and you’ll be unlikely to inadvertently copy anything. The advantages of writing notes by hand then make it greatly advantageous to restricting yourself to your computer (tablet or cell!).

### Recognising good or bad in what you read

It’s going to be hard to read and think at the same time. A little like walking and chewing gum, it comes with experience. And if you do fall over, stop chewing for a while and wait until you’ve got into your stride before having another go. Thinking is all part of the reading process, and while I agree that it might be hard to think on multiple levels while you read, it is possible. For example, consider what other worthy things you’ve thought about while reading this book. If you’ve not stopped to check your phone, or think about eating or drinking something, well done. It might be worth creating an atmosphere with less other stimulants while you read. If you only have your reading material and notebook, you’ll be able to concentrate all of your thinking on them instead of getting distracted.

If you can’t decide what it is that you enjoy from one paper, take it further. Take a look at the first author (likely to be the person responsible for the writing style). Have they written other papers, and do these have those same qualities? Or maybe it’s the lab that they came from. Try to find the lab’s pages on the internet and see if there are other inspiring papers there.

There is also a possibility that you are struggling to get insight with your reading because English isn’t your first language. The subtleties of some of the usage of English may be passing you by. Because of the large number of words, it is possible to write something in English in many different ways. The overall aim is to convey the largest amount of information possible in the smallest number of words, while enhancing the context, meaning and readability. It’s not easy, but the only way you’ll ever get good is by starting, and reading and writing as much as possible.

 If you are still struggling, a really good way forward is to start a journal club with your colleagues. It is probable that you aren’t the only one who is having difficulties. There may already be a journal club in your institution which may (or may not) be a good place to go. What you do need is a safe space in which you aren’t afraid to say when you don’t understand. Think of it as a book club for scientific papers. After all, it may be that no-one understands and the paper is badly written. Alternatively, it can be that someone is able to help you quickly and easily. They can then give you extra insights into how and why they think the author is crafting their paper. This will then help you with your reading as well as your writing.

If you are looking for good fodder for your journal club, consider reviewing a preprint and publishing your review with a doi [@teixeiradasilva2017fortifying]. This will make the whole exercise a lot more worthwhile and produce a net output for your club.

### 2021-04-15 2.20-proposal.Rmd
#First upload of book

## What is needed for your research proposal? {#proposal}

In my institution, you need to have your PhD or MSc approved by a group of 3 independent staff members (i.e. not including advisors) in your department. The department requires that your written proposal is publicly available to everyone in the department to look at (normally deposited in the secretary’s office) at least 7 days prior to the oral presentation of your proposal. This is not unique to my university, many institutions require that you formally propose the work that you plan to do in a PhD. They may all have very different requirements. It is well worth checking out what your institution requires and by when. Talk to your adviser about this.

Your proposal is a collection of documents, where each document is a chapter of the thesis, and is roughly equivalent to a published paper. Ideally, these should be bounded by an introduction, which gives the bigger picture of how the thesis is placed, and a time-line at the end that demonstrates how you can complete the thesis within the two or three years for your MSc or PhD, respectively. Note that your project proposal is placed on your permanent file within the department.

### Structure

We expect that your final thesis will be a set of papers (data chapters sandwiched between an introduction and discussion), and so we ask you to prepare the proposal in the same way. That is, you will normally have two (MSc) or five (PhD) data chapters. In the proposal, each chapter will consist of three parts:

### Proposal introduction

I’m not going to go into what’s inside the introduction here, as it’s already been done in other chapters on formulaic writing (see later in this [part](formula.html)) and specifically about the introduction (see [part 3](introduction.html)). Although you may feel that there's no point putting a lot of effort into this introduction it serves as a literature review for your upcoming PhD study, and is well worth the time and effort.

### Methods and Materials

What’s in the [Methods and Materials](methods.html) section is also covered here. Essentially you need to:

- Introduce important aspects of your study organism and/or study site
- Describe exactly how you’ll set up your experiment and/or collect your samples
- You must explicitly state how each of the variables (introduced in the [introduction](introduction.html)) is collected
- (If necessary) Describe how you’ll turn your collected data into the variables needed for your analyses
- Data analyses (how you’ll test your [hypotheses](hypothesis.html)). This should include a description of the statistical or analytical techniques. It’s useful to structure this section by each of the questions/hypotheses that you are posing

### Hypothetical Results

It’s a very useful exercise to imagine how results will show that you have (or have not) accepted your hypothesis. You can’t show results that would cover every eventuality, but give some typical scenarios that you think may happen if you accept and reject each hypothesis. This could be a bar chart, a map, a scatter-graph, etc.

### Your Proposal - what is it good for?

Once you’re done with your proposal, you might feel that you’ve done an awful lot of work without having added anything towards achieving your PhD or MSc. However, it’s actually a really useful document that you can use in a number of ways:

#### Copy and paste directly into your thesis

One PhD student estimated that she pasted 60-70% of her proposal text directly into her thesis. This included nearly all of the methods and materials section verbatim. Note that this **is not** plagiarism. You wrote the original proposal as a single author and it was not published, so you are at liberty to use the text again.

#### Use it to raise money to do your studies.

Writing grant proposals is essentially the same as your thesis proposal, and so you can use this document (with some modifications to tailor it to the grant objectives) to [raise money](funding.html) for your work.

#### Deposit the proposal to conform to Design and Analysis Transparency

This is the subject of another chapter (see [part 3](transparency.html)).

### 2021-04-15 2.21-proposal-presentation.Rmd
#First upload of book

## Making a presentation from your chapter, paper or proposal {#proposal-presentation}

In theory, it couldn’t be easier to take your written work, be it a research proposal, paper or chapter, and turn it into a presentation. Many people find presenting ideas easier than writing about them, as writing is inherently difficult. On the other hand, standing up in front of a room of strangers, or worse those you know, is also a bewildering task. In this section, the aim is to break down the task of making a presentation into a number of smaller steps so that you can tackle each one in turn and make sure that you are ready.

1. **Give yourself plenty of time to prepare**
Although your presentation might be weeks away, you are going to need ~4 weeks to get the presentation ready, run it past your advisor (and potentially other co-authors), and practice your talk. As usual, my advice is not to leave it until the last minute to prepare as you'll get out what you put in. 
2. **Know who your audience is and what they expect**
If you have no idea of who your audience will be and what they will be expecting, then attend some other talks in the same seminar series, or ask your advisor and/or other lab colleagues for input. The more cross-disciplinary your audience is, the more general you should make your talk. If you've still no idea, pitch it for a very broad audience at last year undergraduate level.
3. **Be sure of exactly how much time you will have to talk**
Be aware that your time slot might include time for questions, and so find out exactly how much talking time you have. If there is a long time set aside for discussion, you might want to throw some extra slides in that can help facilitate answering expected questions.
4. **Make sure you know the format of the presentation**
If you don't know, then you can't make any assumptions about what might be present in the room where you give your presentation. I turned up to give my talk with a presentation on a USB drive only to find that there is no computer, no projector and a room of people expecting me to entertain them for an hour. Even if you know that a projector is present, it's good to have some extra context: 
  + for example is the computer Mac, PC or Linux; 
  + should you bring your own laptop; 
  + is there a pointer; 
  + can it advance slides
  + If in doubt then bring your own accessories.
5. **Know exactly where you are giving your talk and plan for plenty of travel time**
Hopefully, you will be very familiar with the venue, and travelling to the venue will only involve a brief walk. I've made some rookie errors in the past, including deciding to cycle to the venue, getting caught in a cloudburst and arriving literally dripping wet (it was in the UK). The more comfortable you are with the whole setup, and the more organised you can be about getting yourself there without incidents, the better the talk will go for you.

Essentially, you have a story to tell, but does not mean you are story telling. It means that your presentation will require you to talk continuously for your allotted period of time, and that the sentences must follow on from each other in a logical narrative; i.e. a story.

### So where do you start?

#### Here are some simple rules to help guide you to build your presentation:

- One slide per minute: However many minutes you have to present, that’s your total number of slides. Don’t be tempted to slip in more.
- Keep the format clear: There are lots of templates available to use, but you’d do best to keep your presentation very clean and simple.
- Be careful with animations: You can build your slide with animations (by adding images, words or graphics). But do not flash, bounce, rotate or roll. No animated little clipart characters. - No goofy cartoons – they’ll be too small for the audience to read. No sounds (unless you are talking about sounds). Your audience has seen it all before, and that’s not what they’ve come for. They have come to hear about your research proposal.
- Don’t be a comedian: Everyone appreciates that occasional light-hearted comment, but it is not stand-up. If you feel that you must make a joke, make only one and be ready to push on when no-one reacts. Sarcasm simply won’t be understood by the majority of your audience, so don’t bother: unless you’re a witless Brit who can’t string three or more sentences together without [in which case you have my empathy].

#### A formula for presenting a proposal

1. You need a title slide (with your name, that of your advisor and institution)
1. Several slides of introduction
  + that put your study into the big picture
  + explain variables in the context of existing literature
  + explain the relevance of your study organisms
  + give the context of your own study
1. Your aims and hypotheses
1. Methods and Materials
  + Images of apparatus or diagrams of how apparatus are supposed to work. If you can’t find anything, draw it simply yourself.
  + Your methods can be abbreviated but the methods are important for a proposal, especially the numbers of replicates and the variables that you plan to measure from them. 
  + Analyses are important. Make sure that you understand how they work, otherwise you won’t be able to present them to others. Importantly, explain where each of the variables that you introduced, and explained how to measure, fit into the analyses. There shouldn’t be anything new or unexpected that pops up here.
1. Expected results
I like to see what the results might look like, even if you have to draw graphs with your own lines on it. Use arrows to show predictions under different assumptions.
1. Budget
1. Timeline

#### A formula for presenting a chapter or paper

1. You need a title slide (with your name, that of your advisor and/or co-authors and institution). Most institutions have logos, and they like you to use them.
1. Several slides of introduction
  + that put your study into the big picture
  + explain variables in the context of existing literature
  + explain the relevance of your study organisms
  + give the context of your own study
1. Your aims and hypotheses
1. Methods and Materials
  + You are going to have to skip details of the methods
  + Be prepared to pare this down to the important stuff
  ++ variables
  ++ replicates
  ++ analyses
  + Use diagrams and images if you can
1.Results & Discussion
  + If you have a lot of results, then it makes sense to discuss them while they are still fresh in the minds of your audience. 
  + If you had several hypotheses, then place reminders of what they are when you respond to them with the results
1. Summary 
  + state initial hypotheses and whether or not you accept them
  + give some ideas for next steps

### Know your story first, and then make your presentation

Despite previous sections [of this book](storytelling.html) cautioning you against storytelling, you really need to have turn your study into a story in order to present it well. This means that you know what the message you want to reveal to the audience is.  When you develop your story, be imaginative and creative. There are plenty of different styles to telling stories, and you should feel free to borrow styles that best fit your scenario. For example, you can reveal the punch line right at the beginning, and then work back through the presentation to show how you arrive there (but always try to keep something to reveal on the way there). Props are also excellent to produce in talks, something that you can bring with you, and show to an audience, that reveals more about your study. You want your audience to understand your story, but you also want them to remember it. Anything that will help them remember is likely to be a good prop.

In order to keep your audience with you, you need to produce reminders throughout the presentation. This could be themed images or colours that correspond to particular concepts, questions or ideas. If you get confused and can't work out how to present all of the information, then the chances are that your audience won't be able to keep track either. Slim down your study to something that will be easier to put to a story. Just like a real story, some things can't be left out, like characters (i.e. variables) that you rely on to answer your questions. Don't take time to talk about anything off the prepared storyline, as you'll likely lose the audience or confuse them. Treat your main variables like characters that appear with some regularity and consist characteristics throughout the story.

Remember that you don't need to include every detail in your slides or your talk. If the audience can follow your story, but wants to know the details at the end, then they can ask questions about the details. Indeed, you can tell your audience precisely this as a way of side-stepping chunks of (especially) methodologies.

The audience is there to listen to you tell your story, and your slides (if you decide to use them), should be props that help you tell your story. Talk and interact with the audience to tell your story. Point to the slides to help you illustrate the points that you can't otherwise make, or to underline important points in your story. Above all, you need to make sure that the audience is listening to you tell your story.

__Fewer words are better. Tell the story yourself using the props on the slide__

### Slide layout

1. Your aim is to have your audience listen to you, and only look at the slides when you indicate their relevance. 
1. You’d be better off having slides without words, then your audience will listen instead of trying to read. As long as they are reading, they aren't listening. Really try to limit the words you have on any single slide (<30). Don’t have full sentences, but write just enough to remind you of what to say and so that your audience can follow when you are moving from point to point.
1. Use bullet pointed lists if you have several points to make (Font 28 pt)
1. If you only have words on a slide, then add a picture that will help illustrate your point. This is especially useful to illustrate your organism. At the same time, don’t have anything on a slide that has no meaning or relevance. Make sure that any illustration is large enough for your audience to see and understand what it is that you are trying to show.
1. Everything on your slide must be mentioned in your presentation, so remove anything that becomes irrelevant to your story when you practice.
1. Tables: you are unlikely to have large complex tables in a presentation, but presenting raw data or small words in a table is a way to lose your audience. Make your point in another way.
1. Use citations (these can go in a smaller font: 20 pt). I like to cut out the title and authors of the paper from the pdf and show it on the slide.
1. If you can, have some banner that states where you are in your presentation (e.g. Methods, or 5 of 13). It helps members of the audience who might have been daydreaming.

__Large font, symbols that are consistent, familiar themes and photos all make for a good slide__

### Practice, practice, practice

1. It can’t be said enough that you must practice your presentation. Do it in front of a mirror in your bathroom. In front of your friends. It's the best way of making sure you'll do a good job.
1. If you can't remember what you need to say, write flash cards with prompts. Include the text on your slide and expand. When you learn what’s on the cards, relate it to what’s on the slide so that you can look at the slides and get enough hints on what to say. Don’t bring flashcards with you to your talk. Instead be confident enough that you know them front to back and back to front.
1. Practice with a pointer and slide advancer (or whatever you will use in the presentation). You should be pointing out to your audience what you have on your slides; use the pointer to do this.
1. Avoid taking anything with you that you might fiddle with.

### 2021-04-15 2.22-transparency.Rmd
#First upload of book

## Starting out transparent {#transparency}

The scientific method requires us to pose a falsifiable hypothesis, design a controlled, rigorous and repeatable methodology, and report and interpret our results honestly. In theory, it’s all pretty straight forward. I’d like to think that in a world where science is adequately funded, we’d have already reached a situation where transparency was endemic in the scientific system. However, science has not been so lucky, and where there are shortcuts, some people will take them in order to get ahead of others. One of the objectives of this book is to demystify what happens in biological sciences. In other words, the demystification in this book is only really needed because so much of what happens does so behind closed doors, and out of the gaze of less privileged scientists. As you read through this book, you will see countless examples of where things are not as they should be. I have always tried to insert solutions when I highlight problems, but these are admittedly piecemeal, rarely proven, and even in a best case scenario might be best thought of as temporary solutions. The logical antidote to all these shady scientific dealings is to turn to transparency.

We would do better moving to open and transparent science. The open science movement is gaining traction [@sullivan2019open], and I hope that it will become the norm in the very near future. Those of you who are reading this now are likely to be students during an interim period prior to open science becoming mainstream. Therefore, you will be instrumental in adopting this process as soon as possible to ensure that science becomes a fairer place for scientists from all backgrounds, and with all results. The idea is that this will impact research all the way from planning to publishing in a set of [Top Factors](https://topfactor.org/), which are likely to be increasingly adopted by funders (see [here](https://youtu.be/MLbgSGA581Q)).

In this book, I have tried to consistently encourage you toward transparent and open practices in science. Here my aim is to introduce this topic and provide an overview of the main areas in which you can currently make a difference by opening up your research. As the Centre for Open Science explains (COS: [see here](https://youtu.be/9YuNGB3vNOw)), this will take a shift in the culture of the scientific community, which is why you need to understand and adopt a transparent approach from the outset of your research career. There are much better and more comprehensive guides and online resources so I encourage you to look through the literature cited here. Although I advocate hard for the COS, please be aware that this is not the only set of guidelines (see also [recent moves by the Public Library of Science](https://plos.org/open-science/preregistration/)).

### Preregistration

You will spend a considerable amount of time at the beginning of your PhD studies planning what you want to do. In many institutions, this will take the form of a formal proposal. It will be agreed with your advisor, and may well pass the inspection of a committee. You will put a lot of effort into reading the literature in order to ask the best hypotheses possible. You will design your experiments with rigor and control, and potentially redesign them after your advisor and committee comment. The effort put into your proposal is totally disproportionate if it’s never looked at again. Yet, this document is of historical importance, because it says what you think before ever doing the experiment and collecting or analysing the data. Thus, it can in future prove that you were not conforming to confirmation bias, for example by [HARKing](TypeI.html): rewriting your hypothesis after getting your results in order to get a significant result [@kerr1998harking; @forstmeier2017detecting].

Confirmation bias is a problem in science because of the way that science is published. In brief, articles for journals with higher impact are regularly selected based on significant results. If you don’t have that significance, they are unlikely to want to take your manuscript. For this reason, many scientists have sought to have significant results to report, and this is called confirmation bias. Confirmation bias is bad because it violates the assumption that we are answering the hypothesis that we started with, or it could cause scientists to manipulate their data until a significant result is achieved. For example another form of confirmation bias is p hacking: repeating tests with different approaches in order to a significant result. Rubin [-@rubin2020does] gives a good list and set of explanations. As there are a lot of known ways in which researchers have been thought to deliberately or accidentally report false-positive findings, one solution can be archiving your intentions, referred to as preregistration [@forstmeier2017detecting] see Figure \@ref(fig:preregistration). If you register your proposal (or any research plan), you can present this historical document to a journal (probably 4 to 5 years later) to show that you have done what you planned to do (see [webinar on preregistration here](https://youtu.be/8QK2-udwoK8)).

What have you got to lose? With a good proposal that you feel confident with, you have nothing to lose. By registering your proposal, you are most likely to gain, especially in the future if more journals require preregistered hypotheses and methods in order to submit to them. What this does mean is that you will need to do a good job of your proposal. This can be daunting when starting a new PhD, especially with respect to the analyses if you are not from a strong analytical background. Your proposal period should be time to make sure that your knowledge of analyses that you will do is sufficient. I would suggest that the best way to make sure that you are proficient is to obtain a working knowledge of how to handle a dataset such as the one you are going to generate in your experiment during your proposal period. If your lab already has similar datasets, borrow one. If not, generate some data that you can use.

(ref:preregistration) **Preregistration of your research proposal is a step towards transparency in research.** Preregistering your proposal formalises the first review process, while publishing your manuscript formalises the second [after @scheel2020registered].

```{r preregistration, echo=FALSE, out.width = '95%', fig.cap="(ref:preregistration)"}
knitr::include_graphics('figures/preregistration.png')
```

What if new analyses appear during the period, or you are forced to change your experimental protocol. Certainly, most analytical software will be updated over the course of your studies, and some might be superseded. None of this is a real problem to your preregistered content. Logging the software that you intend to use will demonstrate the approach and type of analyses that you intend to take. While it is unlikely that you will ever be held to account for minor changes to equipment or protocol, things do change along the way. It doesn’t mean that you won’t be able to report your results, or that your research won’t be viable. But it does mean that you will need to be transparent about why it changed. For this reason, it is a good idea to document the changes to your proposed research plan and why they happened. It is surprisingly easy to forget! There are some great tools on different platforms for adding this information, together with a timestamp so that it’s clear when it was done.

#### Does this mean that you can’t perform any data exploration?

Although at first sight, preregistration appears to confine you to only testing ideas that you had before you started a study, actually the opposite is true. When you preregister your proposal (or research plan) it places a time-stamp and therefore a historical trail on your ideas. If as you go along with your study you gain new insight and ideas, you can register these with their own time-stamp. When it comes to analysing your data, you use a separate section labelled “exploratory analyses”. You can then apply post-hoc rules to these findings, which will not stop you from reporting them, but gets around the awkwardness of [HARKing](TypeI.html). This serves to reinforce the distinction between prediction and postdiction.

#### Will preregistration of research eliminate the bias from science?

It is probably too early to tell [@rubin2020does], but it is certainly a good place to start [@nosek2018preregistration]. The more researchers know and subscribe to [transparency](transparency2.html) in their research, the more it will shift the culture in science for the better [@forstmeier2017detecting].However, there are more biases existing in science, and the system needs buy-in from stakeholders to promote equity by eliminaion of bias.

#### What Platform should you use?

When choosing where to archive your proposal (or any of your data, analyses, etc.) there are lots of platforms to choose from: [Bitbucket](https://bitbucket.org/product), [Figshare](https://figshare.com/), [Github](https://github.com/), [OSF](https://osf.io/), [Zenodo](https://zenodo.org/) and the list will undoubtedly grow. Making a decision about what you are going to use now may not require that you stick with this same platform for you entire career, but there are some things that you should consider:

__Here are some of the considerations that you should take on board:__

- What do people in your lab and institution use? It’ll be easier to use the same platform as your advisor and other lab members
- Some platforms require a subscription, check whether your institution is a member.
- Avoid using any platform that is tied to a publisher. 
  - Although it’s not possible to future proof this (as publishers have been shown to buy up anything they think will help them control the academic market), you can check how the platform is set up and opt for those that are non-profit organisations. 
  - Other important aspects are “open source”, free to use and access. 
- Can the platform function for more than one aspect of transparency?
- As the need for transparency in science grows, it will become important to have more aspects of projects archived. Does the platform that you’ve chosen cover all the stages from conception to publication?
- Is it easy to use?
  - Some of the platforms will be more intuitive to use, while others require a steep learning curve. Consider how friendly they may be to other collaborators, older advisors, etc.
- Are the archives easily compatible with other platforms?
Working across platforms might be important for your project, especially if you start with collaborators that are already using different platforms.

The platform you choose should work for you (rather than having you work for it). If you are someone who loves to have everything ordered and organised, then you’ll love seeing this all laid out. If you aren’t, then these platforms are going to be a massive help to getting all of your plans sorted. Make updating your platform a habit. For example, you could make sure that notes taken during meetings with your advisor on different projects are logged onto the platform. This way you both have a record of what decision was made when. Remember that you can choose what you make accessible to the outside world.

### Transparency as you move forwards

There are a whole lot more transparency criteria that you will need to be aware of later on in your PhD when it comes to publishing. Becoming familiar with the entire process now will be to your advantage, so I encourage you to read more about this topic. Sullivan et al [-@sullivan2019open] provide a nice overview about how to get started, but be sure to consult documentation at [OSF](https://osf.io/).

### 2021-04-15 2.23-funding.Rmd
#First upload of book

## Generating funding for your PhD research {#funding}

To be honest, writing a funding proposal is an art in itself and should be the subject of an entirely different book written by somebody else. What is written here is only really scratching the surface so I suggest that if generating funding for your PhD research is going to be a large part of your work then you should look for another book.

### Funding databases

Funding is one of the most important aspects of doing science, and something that different emphasis is placed on in different parts of the world. In North America, in particular, graduate students are expected to be able to demonstrate that they are able to raise their own funds, as this is expected of them in their jobs as academics. No matter where you are, your CV will be greatly improved by showing prospective employees (or labs) that you can generate your own funding. North America has a lot of opportunities to apply for funding for all sorts of reasons. There are less opportunities elsewhere, but the pool of people applying is also smaller. There are now so many opportunities, that several databases exist to help you find appropriate funding for your particular situation.

I am only going to address some of these portals, and to quickly consider some of the major reasons why you might want to apply for funding.

### Reasons you might want to apply for funding:

- To improve your CV. Even if it’s only a small amount of funding that you apply for, it will make your CV look better. You might also want to search for prizes. Academic prizes carry both money and that feel good, look good feeling that CVs need.
- Your project is unfunded. If you are completely unfunded, or you have a bursary but no running expenses, then you are going to need money to do your work. The more work (particularly field work) that you want to do, the more money you’ll need to raise.
- Your project is funded, but you want to do more. If you want to do a lot extra (like side projects or paying assistants to collect extra data) then all of this is possible if you can raise some more money to do it. 
- You want to attend a conference or workshop. There are great opportunities for travelling around the world, but they cost money and international conferences are often very expensive.
Publishing in Open Access journals. I find the idea of putting cash into the pockets of publishers abhorrent, but you may have little choice. There are some opportunities to get funding for OA publishing. 
- It is clear that without making an application, you won’t get any funding. But where do you start?
Databases. There are many databases, but here we’ll mention 3 that are easy to use.

#### Mendeley:https://www.mendeley.com/ 
This is probably my favorite. It’s easy to use and apply or remove filters. You can use the same login that you have for the referencing software and SCOPUS.

#### Open 4 Research: https://www.open4research.eu 
For this you’ll need an academic email address to register, but you’ll get access to their research database. It is nice and logical to use, and you’ll find local and international funding opportunities.

#### Research Professional:https://www.researchprofessional.com/ 
You’ll need to create a login for this site, although you might get an in with your university address. This provides you access to what could well be the same database as those above. This is a slick database and relatively easy to use.

#### Opportunities in your department or societies:
Probably the most likely place to get funding is where you already have an “in” (where you are already known). This will include professional societies where you are a member, or your department or university. These places also have prizes, so it’s good that you know what is on offer in order to benefit the most.

### 2021-04-15 2.24-fear.Rmd
#First upload of book

## Fear of submitting written work {#fear}

Many people feel nervous of handing in their written work, and feeling a little nervous is good. It shows that you care deeply about your work. However, when it gets so bad that you can't hand work in any more, then something has gone wrong. It is possible that you are suffering from **impostor syndrome**. This chapter is to help you overcome your fear of handing in your written work.

### What is impostor syndrome? {#impostorsyndrome}

Impostor syndrome is an experience you have when you believe that you are not as competent as you think others perceive you to be. It is not uncommon in many professions, and especially prevalent in academia [@clance1978imposter]. This is now widely recognised and there are lots of useful shared experience out there to read [e.g. @dickerson2019how].

(ref:conflicted) **It is common to worry about handing in your work.** Many people feel conflicted about whether or not they should hand in their work. Is this part of imposter syndrome?

```{r conflicted, echo=FALSE, out.width='95%', fig.cap="(ref:conflicted)"}
 knitr::include_graphics("figures/conflicted.png")
```

- Are you fearful that your work won’t be well received? 
- Is it really up to the standards that are required?
- Could you do better if you just spent some more time?

These are all really common thoughts, and they go along with ‘imposter syndrome’ for a lot of early career (and even older) academics [@woolston2016faking, @hutchins2017what]. There are some important points to think about here:

1. Everybody has these ideas and you aren’t alone
1. Handing in work and getting feedback is part of the learning experience
1. The fact that you care so much about your work and how it is perceived is a good thing. If you didn’t care, then this would be a problem. 
1. If you never hand anything in, you won’t get your post-graduate degree

### Think of it this way

The process of writing is part of your learning process, and you aren’t learning alone. That’s why you have an advisor. While you think that there may be a great intellectual gap between you and your advisor, I can promise you that there isn’t. But your advisor is an experienced reader. Hopefully, there have been many students who have benefited from learning with your advisor (talk to them about their experience). They already work with a lot of students and help those people to bring their writing to a level where it can be accepted by an academic community.

Like other members of the academic community, your advisor also has experience of receiving critical feedback about their written work. Sometimes it is painful and sometimes it feels personal. But getting feedback (or peer review) is a fundamental part of the publishing process. Talk to your advisor and other lab members about this process. Ask people how they overcame their fear of submitting written work.

### What can you do to help yourself overcome the fear of submitting written work?

1. Give it to a friend or colleague to read
1. Remember that most of what you write should be understandable to the majority of people who can read (think of articles in The Conversation as a reasonable reading/understanding level). This means that you can give your written work to fellow postgraduates and postdocs and ask for feedback. 
1. Annotate your text with places where you are particularly unsure.
If there’s a certain part that you are struggling with, annotate it (add comment in word) and point out that you are struggling with this particular section
1. Ask for a meeting with your advisor
You can sit down together with your advisor and discuss the points where you are uncertain before handing in the work. Or, after you’ve written it you can ask for a session when you are given verbal feedback
1. Produce a checklist
If you specifically worry that there may be errors in what you write (grammatical, spelling, plagiarism), then make a checklist that you can tick off prior to submitting. Once your check list is done, don’t mess with the written work again (or you could add more errors), just hand it in!
1. Set a deadline for yourself
If you don’t already have one, having a set of deadlines that you give yourself to give your writing to work to colleagues and hand in to your advisor. If you know that you aren’t good with deadlines, share them with as many people as you can!
1. Provide yourself with a reward for submitting your work. Rewards are a really simple way of helping you to do things. They help simplify otherwise apparently complex fears. 
1. Tell your advisor about your fears
There’s nothing like honesty. Your advisor may be able to cut you some slack, or might sit down with you and look at how the two of you can overcome this difficulty.
If you have more than one advisor, there may be one who you are more confident to read your work, and you can suggest that reading is done in series (instead of in parallel - actually I’d advise this).
1. Talk to your advisor about what really gets you the most upset. If you can’t do this face to face, then you could annotate it in a reply to their comment. The chance is that your advisor doesn’t know how upsetting it is and you’ll be helping them in their future interactions with students. 
1. Ask your advisor about their fear of handing in and get them (and others in your lab) to share their stories. You might find that you have common ground to start sharing how these problems can be overcome.

At the end of the day, this is teamwork. Either you and your advisor are a team, or there are a bunch of advisors helping you. It is in everybody’s interest that the job gets done. Getting a line of dialogue moving with your advisor is essential, even if you have to arrange a meeting about a different subject and then introduce the problem later in the meeting. There are deadlines that need to be kept. You need to find a way of getting through the fear, and if it’s going to be a persistent problem, you’ll need to work out what works best for you.

### 2021-04-15 2.25-formula.Rmd
#First upload of book

## Why use a formula to structure each chapter or paper? {#formula}

Very few people can sit down at a desk and write an academic paper from beginning to end. For most of us it takes a lot of hard work over many weeks or even months to write something that can be submitted to peer review. In part two of this book I approach the writing of a chapter or a paper for peer review in the same formulaic way, and will refer to papers or chapters interchangeably.

Sometimes it seems that every paper I read is just repeating the same formula over and over again. I even tend to forget that it's there until someone breaks the mould. Breaking the mould produces papers that really stand out, captivating the reader straight away. So why aren’t we writing those manuscripts every time we write a paper? Moreover, why don’t we teach students to write standout papers that will captivate everyone. Should we really be teaching formulaic writing?

Here I argue that not only is writing to a formula good practice, but it’s the best way to learn scientific writing. When you look to see who is writing those knock-out articles (i.e. "broken mould"), the answer is that they are really experienced researchers who have written hundreds of papers, and (unlike some of us) in doing so have learnt the essence of great writing. Such greatness is not innate and comes rarely to any researchers (even those who have written hundreds of papers). That's why I suggest to my students that they follow the formulas that are by now well recorded in many blogs and websites on scientific writing.

There is another reason why writing to a formula is recommended. Not only is it easier, but because it is so common and widespread, it also gives the reader a familiarity with reading your work. This means that editors are able to skim through and find exactly what makes your work worth publishing, and it means that those who are only semi-interested are more likely to cite your work.

### So what is the formula?

If we were in my office, at this point in the meeting I usually go to my white board and start drawing funnels (as a metaphor) and blocks. However, others have much better diagrams and here I’ve adapted an idea from Brian McGill’s blog [@mcgill20165]. But you can also find other useful material in a paper by Sheela Turbek and colleagues [-@turbek2016scientific], and in a paper by Kevin Plaxco [-@plaxco2010art]. If you are of the Twitter persuasion, you get lots of useful tips and links from WriteThatPhD and Write4Research.

The diagram below (Figure \@ref(fig:formula)) is supposed to show the movement of a liquid medium from one vessel to another. This is to make you think of your readers' understanding moving fluidly from one section to another. The introduction concentrates the ideas into your particular approach, while the discussion allows the reader to understand how your results fit into the larger body of knowledge.

(ref:formula) **Using a formulaic approach to writing your PhD chapter.** The formula of writing a paper or chapter that we will concentrate on in this book.

```{r formula, echo=FALSE, out.width = '95%', fig.cap="(ref:formula)"}
 knitr::include_graphics("figures/formula.png")
```

#### My take on how to write a formula paper

But what goes into the top, and perhaps more importantly, what comes out of the bottom? I'd like for the discussion to end in a greater understanding of the system for the reader. But maybe if we could also inspire the reader to read more, that would be a great result.

My suggestion is that you now go and read 5 relevant papers in your field and annotate them to see how well they fit into the formula described above. This will give you a practical idea of how the formula is put into practice.

#### Start with an outline
__So now that you've got the idea that writing to a formula is a good idea, what next? __
Next, I suggest that you begin to plan out your formula as a bullet-point outline. Start by writing a general sentence for the subject of each paragraph. Then use sub-points to plan each sentence within the paragraph. Lastly, annotate these with citations that you want to use. Personally, I'd prefer to see the bare outline, and then a fleshed outline before you start the writing proper. This would allow the advisor and student to get on the same page to be sure that what is planned is thought out.

Example of an outline draft of the [introduction](introduction.html):
This outline is taken from Measey et al [-@measey2020cost]:

__Introduction__

* It is important to assess the impacts of invasive species from published studies in order to compare them
  + Not all studies are equally rigorous
  + Inferences from less robust studies should be treated with caution
* More complex study designs tend to result in greater power to detect effect direction
  + Increasingly complex designs are likely to cost more money to conduct
  + Researchers are likely to conduct cheaper (small scale) experiments before large landscape scale ones
* EICAT and SEICAT rely on studies carried out at scales that are appropriate to the invasive population
  + Cheaper studies will result in impacts of lower magnitude, and/or confidence and not reflect the species’ true impact
  + Poorer regions of the world might be expected to produce impacts of lower magnitude
  + But EICAT is used to inform policy
* There is evidence of inequality in existing impact assessments of birds and amphibians
  + geographic disparity for birds
  + This could come from funding disparities
  + how can we achieve a representative impact level for all alien species, without enough peer reviewed studies?
  + will the impact score of species be independent of research costs?
* Aims
  + Whether studies that result in higher impact scores cost more to conduct
  + Whether impact is related to study complexity or time taken to conduct the study
  + Whether EICAT and SEICAT scores demonstrate changes through time
  + Whether different EICAT and SEICAT scores occur with equivalent confidence

__Materials and Methods__

Paragraphs are indicated by the solid dots, and points to make inside the paragraphs hollow dots.

In the above example, you may not know what EICAT and SEICAT are (they are scoring systems used for environmental - EICAT - or socio-economic - SEICAT - impacts of alien species), but I would hope that you can follow the broad argument of the introduction explaining why the study should be done.

#### Next flesh out the outline with citations

In the next stage, fleshing out, I ask that citations for statements above be added along with any examples. Below is an example of the citations added to paragraph 4 above.

- There is evidence of inequality in existing impact assessments of birds, amphibians and plants with lots of DD species (30%; Evans et al. 2016; 38%; Kumschick et al. 2017; 79%; Rumlerová et al. 2016)
  + Geographic disparity for birds (Evans and Blackburn 2020)
  + Examples of birds from Evans et al (2020)
  + This could come from funding disparities
  + How can we achieve a representative impact level for all alien species, without enough peer reviewed studies?
  + Will the impact score of species be independent of research costs?

#### Lastly it’s time to write the text

And once you have the fleshed out outline, it’s time to start writing your first draft. Remember that the best way to start writing is to do just that. It's unlikely that your first effort will be the one that you will finally submit. But start writing. Then read, go back and polish. And keep polishing until you achieve your goal.

### 2021-04-15 2.26-goodenough.Rmd
#First upload of book

## When is it good enough to submit to my advisor? {#goodenough}

Once you've read and revised the draft text three times yourself, and you feel that you can't really improve it any more, hand it over. Personally, I need to wait a day or more between drafts when I work on something else, or do some reading. I find that it really helps to think about writing while not writing. Run through the arguments in your head when you are doing something else. That could be time in the gym, out running, cycling or swimming. Or talk through the logic of your argument with a friend or colleague while having coffee. You’ll find that spending time thinking away from sitting in front of your word-processor will be really valuable in promoting productivity once you sit down again.

If possible, hand it to an office mate or other friend first. Your reader shouldn't have to be an expert in your field, but only needs to have a loose grasp on scientific writing to follow any paper. As a guide, any recent graduate of a biological sciences degree (BSc) should be able to understand. This makes it easy to find lots of potential readers for your manuscript before handing it in.

__In summary__

Writing a paper as a formula is not the only way of writing a scientific manuscript, but it has two really important qualities which I recommend:

1. it makes writing easier, and 
1. readers will be familiar with the style and this will help them understand your work.

### 2021-04-15 2.27-data.Rmd
#First upload of book

## Data Management {#data}

Almost everyone has had the experience of having written some really excellent work, and then had some terrible technology malfunction meaning that they lose everything. Some failure to backup stories are truly horrific. However, they all represent a failure on the part of the person involved to properly backup their work.

At the time of writing, it is very easy to keep copies of documents on the cloud, or even write straight onto the cloud. Talk to your advisor about what options are available in your institution for backing up your work. Make a plan early on, before you start with your study, and then stick with that backup plan as you move through your PhD. A really good option is to start working in one of the Open Science platforms. For example, OSF allows you to create a project where you can continue to use lots of different cloud based resources, and provides an integrated home for them (for an introductory video see here). OSF is also a great platform as it is free to use, and as long as individual files are less than 5 Gb, OSF will host them. This means that for most projects, raw data files can be stored on the cloud at no cost to the project.

### Should the cloud be the only copy of your data?

No. I suggest that you keep a hard drive with a copy of your data that you regularly deposit on a monthly basis, or after significant bouts of data collection such as field trips or experiments. A hard drive has the advantage that in the case of an internet outage, you always have access to your data.

Don't try to store data on a laptop or desktop computer. These are for working files only. Backup is a separate operation where you store files. While computers are good for computation, they are frail and much more likely to suffer from catastrophic failures. They are also objects that are prone to being stolen.

### Backing up in the field

If you are going to be doing lots of field work, then you might want to make a specific plan about backing up data to the cloud during the field time. It may seem expensive, but I'm sure that it won't be as expensive as losing the data. Alternatively, or in addition, you may want to take a portable storage option.

I still prefer field notes and data to be written into a book (preferably with waterproof paper such as [rite in the rain](https://www.riteintherain.com/)). This allows you to make a photographic backup of the hard copy. Similarly, you can design datasheets that you use in the field with a pencil, and then photograph them to make a hard copy. I don't like entering data directly into a laptop or other device. For me, the advantage is that it is too easy to mis-type a data point into a device, and then there is no record about what it should have been. These days there are multiple different ways to enter data into devices (including dictation), and so you should make your own plan about what you feel most comfortable with. If you do make a hard copy, you will then have to transcribe it later. If you make mistakes, you can go back to the hard copy.

### Make a data management plan, and stick to it

However you decide to manage your data, make sure that you have a plan and that you stick to it. Don't become a victim of some unexpected event that will lose your data. These events will still happen, but at least your existing data will be safe. Creating a secure data management plan (Figure \@ref(fig:Data-management)) is something that you should design and then talk through with your advisor. There is much more to data management than simply backing up your data. A great start is to read through the [British Ecological Society's guide to Data Management](https://www.britishecologicalsociety.org/wp-content/uploads/2019/06/BES-Guide-Data-Management-2019.pdf). [See part 3](SuppInfo.html) for information about how to archive and distribute your data.

(ref:Data-management) **The data management cycle.** It is possible to see your data as a cycle between creation and distribution. Your plan should aim to cover each of these steps (redrawn from the British Ecological Society's guide to Data Management).

```{r Data-management, echo=FALSE, out.width='95%', fig.cap="(ref:Data-management)"}
 knitr::include_graphics("figures/Data-management.png")
```

You probably won't be able to think about all the potential pitfalls that may happen during your PhD, but if there are some obvious ones, then do consider them and see whether you can mitigate the circumstances should they occur.

### 2021-04-15 3.01-Title.Rmd
#First upload of book

# Writing the sections that make up your chapter  {#part3}

Getting over the cognitive inertia of starting to write is top of the list for most of us. The approach provided here is to present you with a formula that breaks down each section into smaller more reasonably sized chunks. However, you still need to make a start. Procrastinating starting to write is normal. It seems a massive task, even if you’ve done it lots of times before. You want to write the best manuscript that you can, but you aren’t feeling on top form today, so doesn’t it seem a better idea to wait until tomorrow? It’s not. You must start today and start now. Don’t set the bar so high for your first draft. In fact, don’t set any bar other than a personal goal of getting the first draft done.

For the first draft, type away confident in the knowledge that 90% of the first draft will get junked or changed so much that you won’t recognise it. That’s ok, and it’s not a waste of your time. In fact, the first draft is invaluable in getting your ideas and perspective down on paper, allowing you to better organise and sift through your thoughts to bring clarity to your manuscript. Allow yourself a reward for finishing your first draft, something that you won’t do until it’s done. Now make life easy for yourself by changing the order of what you write.

Although I have placed the contents of this chapter as you’d expect to see them in a typical paper, this is not the order I would suggest that you write them in. The most important part of writing up your chapter or paper is to have your hypothesis or question clearly stated. Once you have this, it will help you to know the literature that is already there and so writing the introduction next seems logical. Of course, you may already have [your proposal](proposal.html) to work with, in which case you can start by reviewing what you already have and deciding whether or not you need to tweak this, or if you need a fresh start.

The methods and materials are the easiest section to write first, especially if you've already done the practical work. This section will also help you focus on the analysis that you are doing for the results. Again you should review your proposal and see whether there were any significant changes to your planned methodology and note them down. This will not be an issue even if you have [preregistered your proposal](transparency.html), you simply need to provide a rational explanation for why things changed.

Next write the results. Remember that you are only looking to respond to your hypotheses, preferably using your (preregistered) proposal plans. Tables and figures are compiled along with the results. Now you have the materials and methods together with the results in full. I suggest that you make a list of all the relevant discussion points you want to make and then turn your attention to the introduction. Begin with an outline, before you flesh it out.

Lastly, write the abstract. Populate your article with references as you write. While you can leave the formatting of references until the end, it’s dangerous to not put citations in as you write as you are likely to forget which paper is which by the end.

## Title page {#title}

Most of the following information in this section on the title pertain to getting a manuscript ready for submission to a journal. For your thesis, I would recommend that you have a title page for each chapter that lists any collaborators and their contributions so that this is transparent for the examiners. It also helps break up the document and can be a good spot for a non-academic illustration (if your institution permits).

Nearly all journals will require you to have a title page for your manuscript. This may or may not include authors. Check to see whether the journal that you are submitting to conducts double-blind review. If they do, they will require a title page without any indication of authors or institutions (they will also likely ask you to remove the acknowledgements section). This should be clearly specified in the instructions to authors of your chosen journal.

### Names and addresses are important {#ORCID}

Getting authors' names correct, their correct addresses (many people have more than one affiliation), can be tricky. Check that you have all of the information required of a title page before you submit. While you need the names of your co-authors to be displayed correctly, you do not need their titles (no Dr. or Prof., etc.).

Indicate clearly who is the corresponding author. Normally, if you have done all of the rest of the work, the corresponding author should be you. You need to learn how to start this role at some point, so it might as well be now. Having the correct name and address will be important to each author. Along with the author names and addresses you should record the ORCID [see part 3](ORCID.html) of each author. The ORCID is simply a unique identification code for individual researchers. ORCID is a non-profit organisation and there’s nothing sinister in signing up. If they (or you) don’t have one, then you should ask them to create one before the submission. It takes less than 5 minutes. Go to www.orcid.org

Note that the same information given in the title page will also be needed when you enter the information into the journal submission site.

Work closely with your advisor, and make sure that your co-authors are aware of the correspondence that arrives from the journal. Use an address for correspondence that you know you check regularly. It;s also worth checking the spam of your email regularly so that you catch any correspondence that might have been misclassified. The golden rule for the corresponding author is the timely communication with the other authors when decisions have to be made.

### The Title

The title of your paper, chapter or book is the first thing that any reader will read, and so should be well considered. If your reader cannot quickly understand your title, the chances are that they won’t bother reading any further. Your title will be your selling point, and your aim is to use it to draw your readership in. Once you’ve managed to invegle your readers to download your paper, your title is also their hook for remembering your paper in their database of thousands of others. Having their key word in your title will help here, and as ever with writing your challenge is to think like your reader. The best titles are those that sum up the entire study in five to seven words. This is best done in a narrative that tells the story ([see part 2](storytelling.html)) of your manuscript in its entirety. This may sound daunting, but you should get into the habit of summing up your story quickly (for friends, relatives and work colleagues). Then it’s a question of refining this story into the short single sentence that makes up the title. While the narrative approach may not work for you, you do want the title to provide enough information so that the potential reader knows what they will find before they open it. Your title doesn’t just have to work for you, it needs to work for a wide audience.

Some people are excellent at writing titles that contain puns of well known phrases or sayings. These can be brilliant, working both to inform what’s in the paper as well as providing some familiar input that helps retain them in memory. However, many fail to do either and are simply a waste of space. If you are tempted to use a pun as your title, make sure that it is widely appreciated, and not just among your co-authors and lab.

You don’t have to come up with a killer title from day one. Most of my manuscripts have a working title that gets revised as I write, and is always open for change before submission. If you have great ideas for a title, do note them down. I find that the more options I have, the more likely I am to come up with something that works for everyone. It also helps to mix and match from a set of candidate titles. Once you have come up with something that looks good to you and your colleagues, test it by entering it into your database of choice (with the default being Google Scholar - see Part 2). Your first 10 results should include a set of papers that you have likely cited in the upper area of your introduction. If you don’t recognise any out of the top 10, it’s time to look at another of your candidate titles.

#### Some title ideas to start you off
- Don’t start by looking for the best title
  + Write a number of candidate titles and ask your co-authors to vote for their favorites.
- The shorter and catchier your title can be the better: 6 words (±1) is an ideal.
  + Allow yourself a longer subtitle if needed, but don’t go over 20 words total (some journals may limit your total to less). 
  + Consider the (former) 120 character limit of a Tweet as an upper limit
- Do include your principal finding if possible
- Include as many key words as you can

#### Things to avoid in your title
 - Don’t feel obliged to include [taxonomic terms](taxonomy.html) unless it is relevant or compulsory
   + Some (taxonomically minded) journals will insist on the species name followed by the taxonomic authority, and/or the family and order, in your title
 - Avoid obscure or specialist words that won’t be understood by your readership
   + There are times when key words are necessarily specialised and your readership will expect this, but simpler words in your title will open up your readership which will otherwise remain narrow
 - Don’t simply define the scope of your work without including your content.
 - Don’t have your title as a question. Rather provide the answer!

### Key Words
The key words are a way for readers to find your content with searches. Typically, the advice is to use words that are not in the title or abstract. This is because many databases have combined searching facilities for title, abstract and key words. I struggle to think of appropriate key words. Some of the big idea words and short phrases from the introduction. Then I tend to look at articles in the same genre, and see what key words they have used. As with the title, I suggest that you enter your chosen key words into your literature database of choice [see part 2](database.html) and see what comes back. You should see a group of papers that look wholly familiar and preferably those that are already cited in your manuscript. If not, it’s time to look again.

### 2021-04-15 3.02-Abstract.Rmd
#First upload of book

## The Abstract {#abstract}

A good abstract is very important as, like a good title, it advertises the content of the paper and draws readers in. In a world where the quantity of scientific literature is increasing, it is more likely that someone will read your abstract but not your paper. Actually, it’s far more likely that someone will read your title and use that to decide whether or not to look at the abstract. Moreover, when you submit your work to the journal, your editor may decide whether or not to immediately reject your manuscript based on the content of the abstract. Therefore, it had better be good!

### What is the abstract?
The abstract is a concise paragraph that sums up the major points of your manuscript so that the potential reader will be able to assess whether or not they want to read the entire paper. It is an abstract of the entire document. It should not be abstract!

### So what would a good abstract contain?
A good abstract is a summary of the highlights of the paper. You can’t hope to include all of the results, but you should include sample sizes, variables and relevant statistics when they support your major hypotheses. You must include the broader subject area that your study fits into, and show how your results are relevant to this.

None of this is easy, and you should not expect to write your abstract in a single sitting. It will likely require multiple iterations, and some intense word-smithing to make it as good as it can be. Abstracts almost always have a word limit, and that makes it challenging and means that you have to be concise [see part 2](concise.html).

Increasingly, you’ll hear that a good abstract is citable. This means that it contains enough information that someone knows that it can be used to cite for a specific fact. Of course, these people should download and read the entire study, but there is no guarentee that they will.

### Where do you start?
Just like planning your writing in general, I’d suggest starting your abstract with an outline. Use bullet points to make a list of things that you feel that you should include. Rearrange your list until you have all the introduction points at the start, results in the middle and discussion at the end.

There’s no need for detailed methodology, but it is useful to know the approach. For example, ‘we used a common garden experimental approach’ or ‘we sampled 85 animals from three invasive populations’. 
Although many abstracts are provided as a single paragraph, some are structured into the sections of the paper, or as numbered points. Using this formula is a good way to get disciplined about boiling your paper down into a small amount of concise words. It’s worth keeping a copy of any abstract that you compose like this in case you decide to submit to a journal that requires it.

Starting with this framework will ensure that the abstract is well balanced.

At this point, I’d circulate it to your advisor to ask whether there are other key points that should be included. As a rule, it’s easier to start with everything present, and only then cut the words down to something within the abstract word limit. If you wordsmith your abstract and then try and add a key point later, it’ll never come out so well.

### When do you write your abstract?
Although the abstract comes on page 1 of your manuscript, only try to write it once you’ve got to the end of the process of writing your manuscript so that you are aware of all the major points that you need to include. Maybe once you've [submitted your second draft](goodenough.html) of the manuscript to your advisor (given that you are both happy with the content), you can formulate the abstract.

### Do abstracts for conferences differ?
Yes. It’s likely that an abstract for a talk will not be the same, unless you have already published the study: but even then it’s probably worth re-writing it. I’d suggest that your conference abstract be more descriptive and thought provoking, such that it is aimed toward the specific audience of the conference. Pose questions that you will answer in your talk.

### Where do people go wrong in writing the abstract?
- The most common mistake is getting the balance wrong. I often see an abstract that gets to the results, runs out of space and simply stops. It is important to have a statement about your interpretation of the results which relates back to your hypothesis.
- Another common mistake is to have a very simply worded abstract that conveys very little information. There is definitely going to be a lot to put in, so expect to write something too long with too much content, and then cut it down by discussing with your advisor what can be left out.
- Text from the main document is copied into the abstract. I’ve written about this [elsewhere](plagiarism.html), but you should never copy parts of your writing. You could start like this, but wordsmithing must change the words and their order substantially. Otherwise, when the first sentence of the abstract is the same as the first sentence of the manuscript, it will really turn off your readers.
- Too many statistics or no numbers are also common mistakes. Adding numbers and statistics are important, especially when they take the place of more words. 
- There is only time for one or two comments on the discussion, so pick out something important. Look through your written discussion and rank the points that you make in terms of interest to others. Can you combine two or more points into a single sentence?
- The introduction of the abstract must frame the bigger question, and the discussion must show how your data responds to this. 
- Don’t be tempted to overreach with your claims in the abstract (or the main text). Include major caveats when they are relevant.
- Finish with something conclusive and strong, preferably how this study changes the understanding, and not a caveat or suggested further study.
- The abstract does not contain citations (or very rarely and only if unavoidable), so don’t put any in!
  + If you have to have a citation in your abstract, then it must include sufficient information that it can be found without opening up the paper: _first author, abbreviated journal name, volume, first page_

### 2021-04-15 3.03-Introduction.Rmd
#First upload of book

## The Introduction {#introduction}

### Why do we need an introduction?
The introduction and discussion appear to be two common stumbling points for students writing chapters or manuscripts. First, what to put in and what to leave out. And second how to construct it. They are complimentary, and if written well can be read without the Materials and Methods or Results, with the reader still gaining all of the major information of the manuscript.

The introduction is going to be the first part of your manuscript that anyone reads. Yes, they’ve already taken in the title and the abstract. These are almost like bait to draw readers in. The meat starts with the introduction. And it’s not only important to get all the correct content in there, it’s also important not to give misleading information that might distract the reader.

When you write the introduction to a paper you must not be tempted to stray away from what we are aiming to introduce. The aim of our introduction is to explain to the reader the hypothesis you are testing, and the approach you have taken to test it.

### So what do we put into the introduction?

As I’ve said elsewhere, the [hypothesis](hypothesis.html) itself is made up of different parts, and each of these must be explained in the introduction. We also need to understand the approach that you’ve decided to take in your study (e.g. experimental, lab or field approach, observations or natural history). All of these decisions that you made were informed by the literature, as is your general understanding of the subject that you are studying. So making sure that you cite ([read about citing](citations.html)) the relevant literature is a key ingredient of the introduction.

However, this is where I think many people get side tracked. Researchers love reading, and it’s super easy to get sucked into all the amazing things that people have done in an ever expanding and increasingly interesting literature. We are often tempted to show exactly how well read we are. Or put in that fascinating tit-bit that we stumbled on by accident. However, you must keep focussed on the goal, to introduce the hypothesis to the reader, and try not to allow yourself (and consequently your reader) to get distracted.

Consider this analogy: You can think of the introduction as being a highway to your hypothesis. As you drive the highway, it's fine to see signposts that lead to other places, but don't be tempted to turn off the highway to visit them. You do need to take your reader to your question as efficiently as you can. Point out the relevant sites on the way, but don't stop until you get where you're going.

### And the construction…?

Previously, I’ve described the introduction like a funnel (Figure \@ref(fig:introduction)), where we channel the reader into our hypothesis by starting broad and ending up narrow ([see the formula](formula.html)). This can draw criticism that journal articles are expected to fit a certain style, that this style is boring, and that what we really need is for people to write in a more exciting and varied way. Although I do not disagree with this, my objective here, is to try to demystify writing and enable you to get started. The easiest way to get started is, I believe, through a formula. But it is important to say that the funnel isn’t the only way, and I’ve read some great papers where the first sentence of the introduction is the hypothesis. However you do it, the hypothesis is at the heart of your introduction because it is the reason for your work.

(ref:introduction) **The introduction funnel.** My suggestion is to keep to the funnel if you want to make life easier for yourself. Start by writing an outline of where you want your text to go. Then add in the references that are pertinent to each paragraph of the outline. Make sure that there aren’t any paragraphs with a single citation repeated over and again; it’s more likely that there is a lot more relevant information out there.

```{r introduction, echo=FALSE, out.width = '95%', fig.cap="(ref:introduction)"}
knitr::include_graphics('figures/introduction.png')
```

Table: (\#tab:introparagraphs) **Here is a guide to the contents of each paragraph of your introduction.** Although you should not feel constrained by this suggestion, it might help you get started when planning your bullet pointed structure of the introduction before you start writing.

| **Paragraph** | **Description of contents** |
|--------|---------|
| Paragraph 1 | Overview of the major theme. |
| Paragraph 2 | Identify knowledge gaps in the field. |
| Paragraph 3 | Identify the problem and the gaps you intend to fill. Introduce the important variables that your hypothesis includes. It’s not impossible to mention others, just don’t get distracted. |
| Paragraph 4 | Introduce the approach that you are using, and the organism of choice. |
| Paragraph 5 | Clearly state the hypotheses that are to be tested. |

Note that you can shuffle your paragraphs in Table \@ref(tab:introparagraphs) to the point where it still makes sense to the reader. Don’t be overly strict or dogmatic with this (or any) advice. For example, there may be more than 5 paragraphs, but use the framework to get started. Do what works for you in your situation. But beware of making any introduction too long.

Now that you’ve fleshed out your outline with relevant citations, it’ll be time to pass it by your advisor to check that you are moving in the right direction before you start writing. My suggestion is always to use your advisor to get the advice that you need – that’s what they’re there for!

Remember that while this section on writing the introduction is short, much of [Part II](part2.html) concerns writing style that is applicable to your introduction. In particular, sections on composing your [hypothesis](hypothesis.html), writing a [paragraph](paragraph.html), and an [argument](argument.html) should be referred to when you are writing your introduction.

### 2021-04-15 3.04-MaterialsMethods.Rmd
#First upload of book

## The Materials and Methods {#methods}

You should start writing this section in your proposal, and you may have to alter parts of it when turning it into your thesis later (if you change your methods). For most people, it’s a relatively simple exercise, but here are some pointers to get you started.

The Materials and Methods section should be citation dense, especially if you have used standard methods that have been written up elsewhere. If there are different methods, you should also explain why you used one over the other, as reviewers or examiners may require this.

If you are introducing new methodologies, then you can expect to write a lot in this section about exactly how this was done, as well as the background to the reasons why the new approach to the methodology was taken.

Your aim is to produce a coherent methodology that anyone in the next 200 years (or more) can pick up and follow in order to replicate your study.

(ref:methods) **There's nothing to writing the methods except remaining methodical.** Your methods are likely to have some generic aspects to other study types in your field of biological sciences.
```{r methods, echo=FALSE, out.width = '95%', fig.cap="(ref:methods)"}
knitr::include_graphics('figures/methods.png')
```

### Sections or subheadings

As ever, I’m going to recommend inserting section headers into your Materials and Methods to help you break it up into bite sized chunks. These are principally for the reader, but will also make your job easier.

An important point here is to make sure that you keep your variables (what you are measuring to test your question) out of the sections like study organism and animal husbandry. Don’t be tempted to scatter variables and how you measured them throughout the Materials and Methods section. Keep them clear and maintain a logical flow, using a time-line if that makes sense.

### Study organism or study system

Having already written the [introduction](introduction.html), you will know that there isn’t too much space to write a lot of information about your study organism there. However, often there is important background information that is needed on the species, which would be distracting or lose the flow of the [introduction](introduction.html). By having a ‘Study Organism’ section at the start of your Materials and Methods, you can add in all relevant information in a paragraph or two.

For example, if you are working on African clawed frogs, _Xenopus laevis_, as a model organism you will want to use this space to describe why they are a good system for this particular investigation. Alternatively, your study could be detailing their invasion biology, in which case this would be the section where you can explain their distribution, that they are principally aquatic, or that they are invasive on four continents. You probably won’t need all of these points, but just the aspects of this species’ biology that is relevant to your study. You should make sure you know whether or not the journal requires the taxonomic authority. If you are not sure, include it now ([see here for an explanation](taxonomy.html)).

Similarly, if you are working on invasive zebra fish, _Danio rerio_, you can write in this section the known dates when they were introduced to Japan, and how we know that those animals came from. You can also give details of their natural range in South Asia.

If you are working on a community of animals, or it is the geographical region that is more important than any single species, you should explain the study system here. For example, if your study is about the rainforest, then you should provide some background information about this kind of vegetation here.

If relevant, you may need to provide a map with the location of samples that you used in this section. My preference is to try to provide a composite figure that will also include an image of your study organism (if relevant).

### Animal husbandry / Growing Conditions / Culturing Techniques
If you have kept animals in the lab, grown plants or cultured bacteria in order to do your experiments, then you need to provide details of how you did this. For plants this will include all relevant information on their source, propogation, growth medium and climatic regimes. For animals, you need to include their source, welfare, housing, feeding, light:dark cycles, temperature, etc. If you bred animals to produce your study life-history stage, all relevant information is needed here. All this information should be available from your ethics application. Some journals will want you to include the details of your ethics permission here.

### Repetitive methodology
If you’ve used the same methodology for each part of your data collection (like DNA extraction, sequencing, etc.), you should have a section that explains all of this prior to any experimental manipulations done. I would suggest that this is detailed in the first relevant chapter of your thesis, and then referred to in later chapters. It’s best to go to look to see what other people have reported on in order to know what level of detail to include. For example, 20 years ago it would have been important to include detailed information of how you extracted DNA from tissues. Today, you can probably say that you used a standard extraction kit or salt extraction without saying more.

### Do you need to include formulae?
Some of your methodologies will include calculations that rest on mathematical formula which can be written in mathematical notation. Certainly, there are plenty of examples in the biological sciences, especially in modelling, where these formulas are needed because they are developed within the methodology in order to produce the [results](results.html). If your thesis involves this type of mathematical progression, then you will already know this is expected because all of the literature that you read will contain equations and formulas. However, if you are calculating (for example) the surface areas of a leaf, through a set of measurements, do you need to provide the formula you used in your materials and methods?

I would move ahead on the premise that formulas that are well established in the literature need only be referenced, and not quoted. If you are building on any formula with new variables introduced by you in your study for the first time, then you will need to show the progression of these formulas. Remember that equations, formulas and theorems are all numbered in the text in a way that you can make reference to them later.

### Experimental manipulations and data collection
If you have done more than one kind of experiment, it’s probably a good idea to give each a separate subheading. Be consistent with these subheadings between the Materials and Methods and the  [results](results.html).

One very important point here is to fully explain the collection of all of the variables that are used in your analyses. Remember to use exactly the same names for your variables here, as you have in the [introduction](introduction.html) (and will use in your [figures](figures.html) and  [results](results.html) sections). It’s very important that the reader gets this consistency across sections. Please also remember to include all units that you collect data in (there is a space between numbers and units!), and the accuracy of the measuring equipment used.

Your variables may be sufficiently complicated that you need to add extra subsections in order to accommodate and explain them properly. This is fine. It’s also a good idea to use a table to explain a complicated sampling regime, or even a flow diagram. If you find yourself having trouble getting all of the information across in words, then go back to drawing a diagram that explains your data collection and ask around to see whether others find this an easier way to understand your setup. Diagrams and/or tables are there to replace words, so once you’ve decided which you want to include, you can delete the rest.

#### Equipment

Many journals require that you name the company that you bought equipment from as well as the town and country where it was made. This is to help others that might want to buy the same equipment (although it’s a bit outdated in these days of multinational companies). You should try to get all of the model numbers of equipment as you do the work, so that you are not scrambling later on to find out what they were. It’s really quick and easy to take a picture of this at the time.

If you have built or designed new equipment for your Materials and Methods, you can expect to make a diagram of this here, if it can’t be adequately described in a paragraph.

### Data analysis
This section is getting increasingly longer as people do ever more fancy statistical manipulations. As above, it’s very important that you use the same names for the variables as you have introduced and collected them. In addition to explaining the specific tests that you performed, you should explain what roles the variables have as dependent, independent, or random variables in your model.

__The independent variable__ has variation that does not depend on that of any other measure. These are usually measured by you during the experiment.

__The dependent variable__ is usually the one that you are testing to see whether any of the independent variables explain it.

__Random variables__ are subject to variations outside of your experimental control but that you want to make sure do not influence your interpretation of the dependent variable. Examples of random variables might be the order in which you did the experiment, or the position in your experimental setup).

You should be familiar with the terms above from when you formulated your hypothesis (see [Part II](hypothesis.html) for a refresher).

Any transformation of any of the data that you collected (e.g. log, ln, cosine) should be mentioned here, along with tests that you performed to ensure that they adhere to the conditions required for the statistical test that you conducted.

Give the name for each package in R that you use, together with its citation. Some journals will have particular software conventions that you need to respect and adhere to. Also give a citation to R as well as the version that you used in your analysis. If you don’t use the latest version of R, then your reviewers may want to know why, so always remember to update your version of R when you do your analysis.

### Did you leave anything out?
By following your Materials and Methods, anyone else should be able to repeat your work. If there isn’t enough detail for someone else to conduct the same experiment or survey, then you will need to add this information somewhere. For example, a list of sampling points or a database of sequences might be needed in Supplementary Information.

The Materials and Methods section can often get overly long, and is relatively easy to edit to make it much shorter. In journals where there is a word count limit, you may want to move some of the Materials and Methods into a supplementation information section. But in your thesis, try to keep the Materials and Methods section as detailed as you can.

### 2021-04-15 3.05-Results.Rmd
#First upload of book

## The Results {#results}

This is perhaps the easiest section of your chapter or manuscript to write. However, there are still many mistakes that are made. Let’s start with the approach that you should take, and then go through some of the common mistakes.

### The correct approach - responding to your hypotheses
Yes, the easiest way to write the results is simply to have the questions or hypotheses that you posed at the end of the introduction and have them as subheadings for your results section. Once you have these subheadings, you simply assemble all of the results with respect to that hypothesis. You should already have all the tests that you used in the methods section.

The results section often includes a lot of statistical testing, resulting in lots of numbers inside brackets within the text. This is all expected and fine, but consider consolidating multiple iterations into a table (if appropriate). And if your tests are already in a table, then they should not be repeated in your text. Instead, simply make your statement asserting the results and add “(see Table 1)”.

When describing the relationship between two variables, make sure that you state the direction of the effect. This means not simply stating that they have a significant relationship, but what that relationship is. This means stating that one increases with the other (a direct relationship) or one increases when the other decreases (an inverse relationship). Once you have described the result, either make reference to the Table or Figure where this can be seen, or describe the results of the test in brackets.

If you do have to insert the results of a statistical test into the text, include the test statistic, degrees of freedom (which incorporates N - the sample size) and the value of P. Only if P is less than alpha (usually 0.05) can you use the word ‘significant’ when describing the effect.

(ref:results) **You may well find that your results section is the shortest section of your manuscript.** Only report what you need to in this section. Keep results that don't respond to your hypotheses for the Supplementary Information.
```{r results, echo=FALSE, out.width = '95%', fig.cap="(ref:results)"}
knitr::include_graphics('figures/results.png')
```

### When to have a table or graph?
Most journals will allow you to have as many tables or graphs as you need to give your results. There are a few important things to consider. First, that data shouldn’t be repeated. So don’t have a table of results where the same results are also in a graph (or even two graphs that show the same data). Also, don’t repeat data from a figure or table in your text. You need to decide how best to present your data. If you can’t decide then discuss it with your advisor by presenting them with the alternatives. Good results should convey your findings clearly, and consulting with others should quickly help you decide which approach is effective.

See sections on [tables](tables.html) and [figures](figures.html) below.

### Descriptive results
Some studies have a set of descriptive results that, by convention, are placed without a section header at the start of the results. These normally convey the size of the dataset that entered into the rest of the analyses. If the size of the dataset varies by design in the methods, then you don’t need it here. But you can’t always control the sample size, so it needs to be established before moving on to the rest of the results. In genetic studies, for example, this might include the number of samples that were collected and successfully sequenced, and thus entered into the rest of the analyses. Your text can then be broken into sections that correspond with your questions, unless you only have one in which case further sections are not warranted.

### Responding to hypotheses
While I encourage you to present data that corresponds to hypotheses here in the results section, you should not formally accept or reject your hypothesis here. This is done in the discussion, and allows you to present caveats and other important issues.

### Common Mistakes
- Having extra analyses in the results that you didn’t include in your methods. If you really didn’t anticipate conducting the extra tests, then these are actually _post hoc_ (after the event) tests (not anticipated before the study) and you should think carefully about how to treat these. Otherwise, you’ll need to r-ejig your methods to make sure that all tests are included.
- Meaningless precision. This can happen throughout a thesis or manuscript (e.g. in the Materials and Methods), but it is most common in the results when too many significant figures are reported. For example, the output of your analysis might tell you that flight feathers in one population are 2.6484332 mm longer than those of another (± 0.0654562 mm), but in your methods you only use callipers that have a precision of 0.01 mm. Reporting results beyond the precision whereby they can be measured quickly becomes meaningless. Whenever you obtain the output of a statistical package, you need to make sure that you use your practical head to re-interpret these results and think about what they actually mean. 
- Repeating numbers in the text that are already in tables or graphs. The text is likely to be packed with numbers already, so don’t repeat any that are already represented elsewhere.
- Repeating text in different sections. As you’ve seen elsewhere in this book, repeating text simply isn’t acceptable. It doesn’t matter if you did the same thing many times, you can always write something in a different way. Having repeated text is the fastest way to bore the pants off your reader, and therefore must be avoided at all costs.

### 2021-04-15 3.06-Discussion.Rmd
#First upload of book

## The Discussion {#discussion}

The discussion should be the final section of your manuscript or chapter that you are going to write (given that the [abstract](abstract.html) will come last). This is because for the discussion you must already know all of the rest of the manuscript, starting with the hypothesis, which dictates what will go into all of the other sections. While writing the other sections, I often make notes under the heading ‘discussion’ so that it acts as an _aide memoire_ to ideas that I’ve had during the study.

Before you start writing your discussion, make a plan and then discuss this plan with your advisor. I’d make this suggestion for all of the sections of your chapter or paper. It doesn’t take that long to do, and it provides an opportunity for you and your advisor to talk about the [results](results.html) of your work and discuss them together. Such discussions should be eye opening for both of you, and they provide a great opportunity for you and your advisor to get excited about the work you’ve done, your [results](results.html) and what they mean. I get a lot of enjoyment during these discussions, especially when sharing the excitement of the [results](results.html). Sharing thoughts before you start writing is important because by talking about it, you and your advisor are more likely to come to a consensus about the best way in which to interpret your [results](results.html). Conversely, presenting them with the discussion finished might not be the best way of pitching ideas to them. Once you've had time to discuss it, you might have a lot more work to do in rewriting what you've already produced.

There is a convention that you do not refer to the [tables](tables.html) or [figures](figures.html) in the discussion. This is because they have already been referred to in the [results](results.html), and a reader should have already consulted them there. Occasionally, you may want to have a new conceptual diagram in the discussion, but this is rare. Most often you do not refer to figures or tables and no new ones appear in the discussion.

In general, the first and last paragraphs of the discussion are key to the reader, but the discussion must also consider caveats and limitations in the experimental design and interpretation of your [results](results.html), as well as providing a concise discussion of the results in the context of existing literature. This is also your opportunity to suggest new hypotheses and how they could be tested.

(ref:discussion) **Your discussion needs structure too.** Let's remember that if you are struggling to write, there is the potential to follow a formula, such as the one I outlined previously.

```{r discussion, echo=FALSE, out.width = '95%', fig.cap="(ref:discussion)"}
knitr::include_graphics('figures/discussion.png')
```

Again, I'd like to repeat that the convention described above is a formula that is intended to help you getting started with writing. It is not the only way to write, and if you feel confident that another way is better then I encourage you to explore this with the support of your advisor. Be imaginative and don't feel that you are constrained or compelled to do anything in a formulaic way.

### First paragraph of the discussion
Your discussion begins by you responding to your hypothesis, clearly stating whether or not it has been accepted, and putting this into the wider context of the study (i.e. paragraph one or two of the [introduction](introduction.html) with relevant literature). You can then follow these statements by emphasising what you consider to be the most important finding, and explain how it adds to existing literature. However, don’t be tempted to over-interpret your results, or claim that they mean more than they do ([see section on speculation](speculation.html) below).

This first paragraph of the discussion doesn’t have to be very long (three to four sentences), but you should make sure that you end by providing a link to the following paragraph or explaining how you will move the discussion on in sections.

### To sub-section or not to sub-section the discussion?
My preference is to plan the discussion before you write it, just as you did for the [introduction](introduction.html) using a bullet pointed list. This will provide you with logical sub-section headers for the discussion under which you can write the first draft. When your chapter has a simple aim that is easily communicated, I’d suggest deleting these sub-section headings before you finish. However, many studies are more complex and contain multiple experiments or evidential approaches. It is then sometimes wise to leave sub-sections in your discussion so that your reader can more easily follow the text. Where possible, these should be the same sub-sections that you have broken your [methods](methods.html) and [results](results.html) into, especially where these relate to specific hypotheses or aims. Or it may be more appropriate to discuss the different approaches separately, specifically when the literature that you refer to falls into different groups.

If you are stuck and can’t decide which way is most appropriate for your work, spend more time on fleshing the outline specifically to include the literature that you want to cite. Try it one way, and then the other, and you should quickly be able to tell which makes more sense. Of course, you should also ask your advisor for their opinion – that’s what they are there for, after all.

When considering what sub-section to write first, go back to the order that you’ve presented the questions or approaches in the rest of your chapter or manuscript. Keeping the order consistent throughout is a really good way of helping your reader follow what you want to communicate. Shuffling the order in each section is almost guaranteed to get them lost and wishing that they hadn’t started reading.

**Next you need to discuss!**

The discussion is about explaining the meaning of your [results](results.html) to the reader. I often find that people write a lot of inappropriate information in the discussion. Remember that this section is not going to provide background information, and is unlikely to bring up new topics that need introducing. It may be that your [results](results.html) prompt you to introduce a new area of research that wasn’t covered in your [introduction](introduction.html), and this is fine. But for the main part, you should discuss your [results](results.html) in the context of existing literature. You can expect that the literature that you use in your discussion will only partially overlap with your [introduction](introduction.html), with plenty of new citations. Similarly, it can be that discussing your [results](results.html) will mean that you end up with paragraphs that have no citations.

When providing different sides of an [argument](argument.html), try to use your [results](results.html) to conclude that one side is supported more than the other. If your results don't help with this particular point, then it could be that you are trying to discuss something that isn't directly related to the work. This is a very common problem in discussions, and a good test is asking yourself how your [results](results.html) add to the point you are trying to discuss. If they don't leave it out and move on.

### Caveats and limitations
An important aspect of the discussion is to consider how the interpretation of your [results](results.html) may be incorrect. For example, if you have done an experiment, how well controlled was it and how well could it be considered to scale up to real world interactions? Could you have measured other variables? Almost every study will have caveats and limitations, and it is very important that you report them in a considered approach. You can also add ideas on how to improve your approach to eliminate important caveats that you have identified, even if this means suggesting something that adds evidence from a completely different field of biology.

My preference is not to provide all the caveats and limitations as a separate paragraph. Instead, mention them when you are discussing relevant aspects.

### Should you speculate in the discussion? {#speculation}
Reviewers will often be unhappy with speculation in the discussion section. Speculation isn’t that hard to spot, as it occurs when you make claims for which your [results](results.html) have no foundation. I think that it is healthy to have one or two statements that are speculative, but clearly labelled as such. After all, after writing this paper, you are going to be one of the world experts in the topic, and thus your deeper understanding is often worth relating to the reader. However, I suggest that you speculate in combination with suggesting what work could be done in future. If you really feel that the point has to be made, you must clearly label it as speculative.

Perhaps an easier trap to fall into is over-interpretation. This is when you suggest that your [results](results.html) mean more than they do. It’s an easy trap to fall into, especially after setting up the study in relation to key topics in the discipline (presented in paragraph one or two of the [introduction](introduction.html)). You will probably find it hard to see where you have over-interpreted, and this is something that having your work read by your advisor, or another colleague, will really help. You may then be asked to ‘tone down’ your claim, or to place it into the direct findings of your [results](results.html).

Again, my preference is not to place all speculations or future hypotheses in the same paragraph of the discussion. These aspects should appear as the topics they relate to are discussed.

### Don’t beat up on others
Your [results](results.html) may show that other researchers were wrong with their interpretation or findings. Whatever you may think of them, never use your discussion to be disrespectful to other researchers or their work. This has been referred to as the “bully pulpit”. As with all aspects of professional interactions, consider how you would like to be treated, and act accordingly. This is not to say that you shouldn’t point out mistakes that were made before, but be sure not to become emotive or insulting.

Generally, such comments won’t get through the peer review process, and remember that you might be insulting the examiner of your thesis (or the reviewer of your paper) – which is not likely to go down well!

### Where next?
The ‘where next’ aspect of your discussion is important as it may provide the reader with ideas for their own work. Of course, these are questions that you may wish to pursue in your own career, or they may be corroborating evidence from other disciplines that you will never undertake. Either way, making pointers for continuing aspects of the research is an important component of the discussion. Providing new lines of research may also allow you to speculate about what you consider to be the most important angle of this topic now that you have presented your [results](results.html). This should be justifyable and not gratuitous. There's no point in suggesting what other studies could be done just for the sake of it, or just because you happen to have already done them in your thesis. But if there are legitate parts of your thesis that link to your discussion of your chapter (and this is quite likely), then you should definitely point to them.

### Last paragraph
The last paragraph of the discussion is your take home message. It’s a summary paragraph that sets out what you aimed to achieve, and what the new state of understanding of the topic is now that your [results](results.html) are out. This should include the key literature that can now be reconsidered.

#### Never repeat text
Please remember that while this might sound similar to your first paragraph, it is not the same. This final paragraph should not replicate any text that appears elsewhere in your chapter or manuscript (not even the abstract). Never repeat or copy text generally, even within your own chapter (or between chapters). For the reader, it’s very easy to spot and it gives the impression that you have nothing to say and are simply filling space. This is not the kind of impression that you want to give your reader, especially if they are examining your work!

As always, there are a number of other places to look for more advice to write your discussion, and I’d encourage you to read as widely as possible. For example, try Hess [-@hess2004how] and Sanli [-@sanli2013how].

### 2021-04-15 3.07-Acknowledgements.Rmd
#First upload of book

## The Acknowledgements {#acknowledgements}

This is the part of your manuscript where you get to thank all of the people who have helped you along the way. It is easy to forget people and so it is worth considering making notes about who has helped you along the entire time with your PhD. Then all you have to do is remember where you wrote it down!

### Thank your funders

Any people or organisations that funded you or any component of your work. Check with them they may have specific wording that they want you to use. Remember that your institution may not have given you money but may have supported you in other ways you should probably thank them. There are registers of funders, and you should use to see whether your funder is present and exactly how they should be acknowledged. One example is the [Crossref Funder Registry](https://www.crossref.org/services/funder-registry/). Funders rely on being able to search [databases](databases.html) for their name and find all of the work that they have funded. This means that you should use the particular wording that they supplied, especially if they specify this in their funding letter.

If your funders aren't in the database, you might suggest that they get added.

### People and organisations to thank

 - Try to remember all of those people who helped you doing fieldwork, or in the laboratory. Especially laboratory technicians without whom a lot of the work that we do isn’t possible.
 - People that may have helped administer your funds or other office administration. 
 - Organisations that gave you permission to do your work. In some journals the acknowledgements is the correct place to put your permit numbers along with the names of the organisations who provided them. 
 - Increasingly, permission from ethics committees is placed in the materials and methods but do check with the journal instructions to authors.
 - People who supplied you with samples, images or other elements of your work or presentation of your manuscript.

Many journals have specific ways in which they want people's names to be written. You will need to follow the guidelines. For example, authors are often referred to by their initials in the acknowledgements. This is useful when some people in your paper need to thank particular funding bodies who didn't fund others.
Thank anyone who gave you a photograph that you used in the paper.

### Who not to thank (in your manuscript acknowledgements)
 - You do not need to thank anyone who is an author. The exception is in your thesis acknowledgements, where you should mention them and their support.
 - Personally, I tend not to thank people who were paid to do a specific job and only did that job. The acknowledgements is a section to name people who have gone out of their way to help you.
  - Your family for getting you to grad school (unless they really helped you do the work). However, in your thesis acknowledgements, you should absolutely include your family.

If in doubt it's probably a good idea to be as inclusive as you can in the acknowledgements.

If in doubt with the acknowledgements refer to other acknowledgement sections in the same journal and see how they are formatted and written. Remember that if your journal is conducting double-blind review, they may ask for the acknowledgements (and the title page) to be removed from the main manuscript and uploaded separately.

### 2021-04-15 3.08-References.Rmd
#First upload of book

## The References {#references}

Putting together the reference section of your chapter or paper can take a very long time. There are quite a lot of different software packages out there to help you do this (e.g. [Endnote](https://endnote.com/), [Mendelay](https://www.mendeley.com/guides/desktop), [Zotero](https://www.zotero.org/) - see more about these [here](https://subjectguides.library.american.edu/c.php?g=479020&p=3323781)). These are a mixed blessing. Firstly, most of them are not free, so unless your university subscribes to their providers you would have to pay for the privilege of using them. Next, all of the ones that I've ever used are very demanding and do not necessarily work well with your word processor of choice. Given all the time I have invested in reference formatting software, I think I might have been better off just copying and pasting references I have at the end of a document and then formatting them by hand. This tends to be how I do it now. Having said this, I used Zotero as the referencing software for this book. It took very little time to learn how to use, it was painless to enter (most) of the references into the database, and it is free and works with [Bookdown](bookdown.org) and Google Docs.

Having the references at the end of your chapter or publication is actually very important because in-text citations are not sufficient to determine exactly which paper you are referring to. Once you become very familiar with your own specific field you will find that you do glance at the references to make sure that the authors are citing the papers you think they are. It's also a great place to learn about literature that you don't know about already, and for this you need the full citation in order to look up the paper.

The right time to make sure that references are in the document is when you are writing the document. Do not leave this until the end. It has happened to me many times that I have written an in-text citation, did not paste the citation of the end of the document, and by the time I had finished writing, had forgotten what I had read and where I had read it. This then leads to a lot more time wasting trying to find that or another relevant citation.

Lots of journals are very fussy about the way that citations are given in their publications. I think it's really very cheeky especially as many of them also force you to pay for the privilege of doing all the work. There are some exceptions to this (e.g. [PeerJ](www.peerj.com)).

There are lots of better things that you could be doing with your time instead of formatting references. And so I'm not going to spend an awful long time here telling you that you must do it right. I really believe that on first submission of your manuscript you should be able to submit your references in any format. it really only makes sense for you to do all the work formatting them if you aren't going to pay to have published the paper.

There are lots of anal academics out there who will delight in looking for every missing full-stop, comma and capital-letter missing from your references. Perhaps it's because they feel secure pointing out these errors, but are incapable of knowing whether or not you've written good or bad science. Please don't join them and if you can look for a better world where you spend your valuable time doing more important things. For me, at this moment, it does not include any referencing software that I know of.

### Digital Object Identifiers (DOIs) {#DOI}
A [DOI](https://www.doi.org/) is an [ISO](https://www.iso.org/) unique character string to identify physical, digital or abstract objects. Their beauty is that they are persistent over time, so once issued they will always be a way to reach a particular object.

DOIs are extremely useful as you can usually click on an active link DOI and go straight to the article in question. Therefore it is well worth adding the DOI to your references if you can. Some journals will allow this (and even demand it), while others have yet to come around to how useful they are.

However, DOIs cannot replace references, otherwise we'd need to be able to click on every link all the time, and couldn't read any paper without a connection to the internet. It's still really useful to be able to read a formatted reference at the end of a paper.

#### DOI Tools

If you are writing your references by hand and need a DOI for every reference then there is this very useful online software that will provide the DOI if there is one for every reference you enter:
[doi.crossref.org](https://doi.crossref.org/simpleTextQuery)

There's also an equally useful database that provides BibTex for DOIs that you enter: [doi2bib.org](https://doi2bib.org/)

### 2021-04-15 3.09-Tables.Rmd
#First upload of book

## Tables {#tables}

If you think that tables are just about displaying a series of boring numbers, then it’s time to free your mind. Tables are an amazing vehicle to display information that can be numeric, written or pictorial. From the simplest of two by two tables with several levels of organisation, the reason for putting your information in a table is that you can provide it in a simpler and easier way to the reader than if you were to put all the information in full into a paragraph of text. There are few rules on Tables, but you'll find some guidelines in this Table \@ref(tab:tables).

Most journals want tables formatted so that there are only horizontal lines. This makes them easier to print, but it can make following text across a large table difficult. Shading every other row really helps with this, and if you have the option of using this in your thesis then I'd recommend it.

As a biologist, I first started to use tables to provide large quantities of numbers. Tables of morphological measurements or meristics (counts). These were always tricky to put together in word processors, until they became better integrated with spreadsheets, and now it’s really easy to put these tables together and even format them in a spreadsheet and then cut and paste them into your word processor.

Table: (\#tab:tables) **Some general dos and don'ts for tables**. There are exceptions for all of these, but if you can stick to them you are unlikely to go far wrong.

| **Do** | **Don't** |
|------|------|
| Use a table as a means of reducing the length of your text. |  Merge cells, especially in irregular ways, as this makes formatting and copy/paste of your table prone to errors. |
| Make your headings clear and concise. | Omit units of measurement |
| Summarise extensive data. | Try to include every data point (that's what [Supp Info](SuppInfo.html) is for). |
| Use tables to provide repeated statistical information on tests or models. | Feel the need to include all of these in the ms  (that's what [Supp Info](SuppInfo.html) is for). |
| Provide a full table legend that explains exactly what is in it (see [figures](figures.html)). | Make tables so large that they have to run over several pages. If this can go into [Supp Info](SuppInfo.html) then it should. |
| Be adventurous by using tables to produces ideas, even with entire paragraphs of text or photographs. | Feel obliged to stick to using tables only for numbers. |
| Use shading for every other row. | Have rows or columns so close together that you can't tell them apart. |

### 2021-04-15 3.10-Figures.Rmd
#First upload of book

## Figures

I consider myself to be a visual person, somebody who understands more by looking at pictures than by reading words. Not everyone is the same but there are a lot of people who can derive more information from a good figure. Therefore for me, a figure is an opportunity to convey maximum amounts of information. It can be a relatively simple figure like a graph or bar chart, a map, a conceptual diagram, or my favourite, a composite figure.

The way in which to visualise your data requires another book on the topic, as there are a myriad of decisions to make on all sorts of topics. What I provide below is a very brief start to thinking about these topics. If you need to explore more of the world of data visualisation, I suggest that you read books on this topic. Happily, people like Kieran Healy [-@healy2018data], Jack Dougherty and Ilya Ilyankou [-@dougherty2021hands-on] and Claus Wilke [-@wilke2019fundamentals] have each written such books. Claus points out that while there are ways in which you can go wrong when plotting data (such as drawing lines to infer relationships that are not there), the skills in data visualisation are more in the aesthetic and interpretive environment of what tools to use that make the data easiest to interpret. How will the figure enhance understanding of the data?

Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’ books is R ([see part 1](useR.html)), and the many packages that have been developed for use within the R environment [@rcoreteam2021r]. The distinct advantages that this produces are the almost limitless additions and details that you can make to these figures to produce exactly what you want, in a [vector format](vector.html). The drawback is that you might need to invest some time to learn how the graphics package works, and how to manipulate it to do what you want.

I agree with the approach of Dougherty and Ilyankou [-@dougherty2021hands-on] who suggest that you start any figure by mapping out what you want to convey and how. Sketch this out and discuss it with your advisor.

### Graphs

You should be aware of all the things that make up a good graph already. here I'm going to provide some extra pointers on what I think is important and should be considered when drawing a good graph:

- __Labels__. This seems completely obvious and yet it is a common problem with submitted papers. Axes should be labelled with both an explanation and the units in which they were measured. 
  + __Capitals__. Be consistent in capitalising labels and other words in your figure. Avoid using just capitals (e.g. ALL CAPITALS ARE DIFFICULT TO READ AND UNSIGHTLY)
- __The scale__. make sure that the numbers of the scale are large enough to read even when the figure is reduced the size that it will be printed on the page. This includes the text that crossed the axes.
  + If you have used a log scale, make sure that this is prominently labelled.
- __Data points__. I am a great fan of including all the data points that go to make up any summary statistic that is displayed on the graph. Thus, if you provide a mean or something like a box plot for your x-axis category, then please add all data points as well. This can be done using the jitter function in R. 
- Don’t draw relationships that aren’t there. The ease of plotting graphs makes it possible to display relationships between data points that don’t exist. This is most commonly done across a set of discrete measurements when a continuous line links them without any data to suggest this relationship. Be especially careful when plotting lines that you can infer a continuous value between points plotted.
- Compiling graphs. if you have a series of graphs that share the same x or y axis then consider compiling them together. This essentially turns them into a composite figure, but it often means that readers can compare data easily from one graph to another. remember that you will have to make sure that the scales on the shared access are all the same. 
- Use consistent symbols and colours throughout your thesis. Your thesis will likely plot data on similar variables in many ways throughout. By using consistent symbols (e.g. circles for females and triangles for males) and colours throughout the thesis for the variables that chapters have in common, you will help your reader understand a figure much more quickly. The consistency of the graphical environment of a thesis will facilitate understanding.
  + Consistency is especially important across panels within a composite figure.
- Colour. The reason for all the fuss about not having colour in figures is mostly to do with the extra cross cost involved in printing them. Typically these days it is possible to use colour in figures and for these to be reproduced without any extra charge online. Thus if the data that you have is best shown with colour then use colour. There are some things to be mindful of when using colour.
  + Colour blindness: This affects quite a large portion of the population, particularly men. To avoid using colours that can be easily confused simply look them up for example here
  + Printing colour in monochrome: some people still like to print out papers in order to read them, annotate them or share them at a journal club. using a monochrome printer many colours are indistinguishable. you can therefore change the hue or intensity of a colour to make sure that it can be distinguished when it is printed. 
  + Aesthetic quality: some colours don't go well together. a lot of people have already determined this and you can find a good guide [here](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3).

### Maps
There are many reasons why you might need to include a map in your paper when writing for biological sciences. I find that the easiest way of creating Maps is to use GIS software like QGIS or ArcGIS. Both of these require a good degree of learning in order to master. To have a map with sampling points is relatively straight forward. Especially with QGIS you can watch ‘how to’ videos on the internet to learn most of what you need to know. things to remember when you're drawing a map:
If your map is a detailed view of an area then consider a small context map in the corner showing where your area is
- Include the scale
- Provide a North arrow especially if you have rotated the view
- Some journals require a specific mention of the geographic projection [see here](https://www.gislounge.com/map-projection/), or to provide a citation for each layer of the map (which makes sense when you think about it). Thus it’s a good policy to make sure you know the provenance of the layers that you are using.
- When using colour see [these](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) handy sets of colours to use at Colorbrewer.
- Make sure that any points you have included are big enough so that you can see the point and the shape of a point when the figure is reduced
- Strictly speaking maps in figures should not have embedded legends (keys). The correct way to explain symbols is to do it in the figure legend. However, sometimes if there are many different symbols it is more expedient to use an embedded legend.

#### Maximising the content of your figure
Sometimes you will have the option of including more information in a figure than the original plan that you started with. This is perhaps especially true when it comes to producing maps with GIS, because you have the option of adding extra data layers. But remember that this information should not be gratuitous. In figure \@ref(fig:not-fancy), we can see a map that shows a number of sampling sites. The map has a scale bar and north arrow as needed.

(ref:not-fancy) **A plain map can give all the information needed.** This map adequately shows the positions of sample sites (or anything else) as red points within the southern African region.

```{r not-fancy, echo=FALSE, fig.show='hold', fig.align='center', out.width='90%', fig.cap="(ref:not-fancy)"}
knitr::include_graphics('figures/not-fancy.png')
```

However, it is relatively simple to add more information to this map in the form of layers. In figure \@ref(fig:fancy-map), a digital elevation model (DEM) allows us to see the relative elevation at the sampling points, while transparent colours provide information about the different biomes in which the samples were made. Lastly, I've used a piece of sea to insert an image of the species sampled. There is also a reference map (top left) showing the area of southern Africa where the sampling points occur. This figure is now much more informative, and has greater aesthetic appeal to me, but all of this added information is only of use if relevant to the manuscript or chapter.

(ref:fancy-map) **Maps tend to have a lot of space, and GIS provides the opportunity to add layers to give rich information, as well as improving aesthetics.** This map shows lots more information about the biomes in which each of the sample sites (red dots) falls, the elevation and an image of the species sampled. Inset of Africa (top left) shows the area sampled with a red square. I've also managed to fit in an image of the species sampled into the sea (bottom right).

```{r fancy-map, echo=FALSE, fig.show='hold', fig.align='center', out.width='90%', fig.cap="(ref:fancy-map)"}
knitr::include_graphics('figures/fancy-map.png')
```

### Conceptual diagram
Conceptual diagrams or figures are very useful in the introduction or methods when you have undertaken a complex study design. There are many different ways of doing this, but essentially the information needs to be conveyed in a way that is more simple in the figure than writing it out. Boxes and arrows are sufficient for most purposes. If the figure gets too complicated to understand, it's probably of little use.

I encourage my students to draw conceptual diagrams of how the thesis chapters relate to one another, and then to include this diagram in the introduction to the thesis to help examiners and any other readers to understand how the chapters are interrelated.

I suggest that it is also very useful when you're undertaking a review to draw a conceptual diagram for yourself of what you were trying to achieve. This may or may not then be included in the review when it comes to publishing.

In order to make your conceptual diagram clear the following points should be followed:

- Keep any text in the diagram large enough so that when it is reduced to printing size it can still be read
- Try not to write full sentences but just notes
- Make use of arrows to show how ideas are related
- Look at other conceptual diagrams especially those that resonate with you as a reader

### The composite figure
For me, composite figures started when I noticed other papers were including small pictures of species in phylogenies. As a reader I really liked this way of understanding more about the animals that were being studied. The composite figure also comes about by making use of blank spaces that naturally occur within figures. Many graphs have large white spaces toward the top (left or right), and small images of study animals, cartoons of study designs or systems can be inserted to improve the immediate recognition of what the figure is portraying.

#### Ideas for composite figures:
include a picture of the subject species in a small corner of blank space in one of your figures. For example, maps with a large area of sea (If your subject is a terrestrial species) are especially good for this (see Figure \@ref(fig:fancy-map)).

### Sending your graphics file to the publisher
One consistent problem is the lack of understanding that authors have for different file types. I can be rather confusing as these are stated in lots of different ways, and often in units that are not intuitive. Journals usually have long and detailed explanations of how to do this in their instructions to authors, yet this is one of the most common difficulties that authors have. I will provide a brief overview of the most common problems, but better overviews exist and I encourage you to read them [e.g. @wilke2019fundamentals].

Be aware that the publisher won't be able to change anything within your figure, like the fonts. This means that you will need to use the same font in your figures as stipulated by the journal (in the instructions to authors). You also need to be critical of the size of any fonts that you use so that they can be seen even if the figure is reduced to fit the page width, or half page width. Be especially careful of numbers on scales as well as labels.

Most publishers will require you to name the files of your figures to correspond with their instructions to authors. If they don't have specific requirements, then use clear file names (e.g. Figure_1.png), and not cryptic ones. Remember that each figure needs a single file, so if you have produced a composite figure, then you will need to output this as a single graphics file.

#### Vectors vs bitmaps {#vector}
The biggest difference among image types are the difference between vectors (a line drawn between two points) and bitmaps (using pixels to create images). Vector diagrams will produce superior images on screen that will allow you to zoom in without loss of resolution (e.g. svg, eps, pdf). With a bitmap, the more you zoom, the more you will be able to see the pixels that make up the figure. If at all possible, create your figures using vector graphics (such as in R or drawing software such as Adobe Illustrator), and store them in this format. The only exceptions will be images (e.g. png, jpg, tif, raw, gif, bmp) that you add to composite figures, insert these with the highest resolution available.

If your figure is a photograph (or another tonal image type), then it will have to be a bitmap. However, if you have a line drawing, try to get it output as a vector rather than a bitmap as not only will the quality be better, but your options for editing and improving will be greater (should the editor require this).

#### Sizing your figure
When you export your graphic from the graphics package that you are using to draw it, specify the resolution (most often in dpi: dots per inch) given by the publisher. Minimum resolution for photos and tone images is 400 dpi, and 600 dpi for line drawings. Remember that the dpi will change if the size of your figure changes. Hence you must know the printed size of your figure (in cm or inches) on the page. This is usually either one or two columns. Rarely, a journal might allow you to use landscape instead of portrait for a figure. As a rule, you should use the page setup in your package to specify the output size of your figure (page layout function) before you start, and then set the resolution before you export the figure. It’s always possible to draw at a higher resolution and convert to a lower resolution. Low resolution images need to be redrawn to make them higher resolution.

If you are using R, you can specify your graphical output in your coding. If you are using a GUI [(like R Studio)](https://www.rstudio.com/), you can choose to export as an image (i.e. bitmap) or pdf (i.e. vector). R Studio will allow you to set a generic or custom page size before you save.

### The figure legend
The most important point with the figure Legend is that it should be a stand-alone explanation of what is in the figure. The reader should not have to refer to the text in order to understand what is in the figure. For example, if the figure is about a species then species name should be included in full. The explanation of any variables or categories needs to be provided. But it should not be taken to absurd levels, is not a reason to repeat everything that's in the materials and methods. Take a look at some figure legends that are already published in the journal that you are hoping to submit to and you should have a good idea of what to write. Figure legends need to be full, grammatically correct sentences.

Remember that the figure legend is placed underneath the embedded figure in your thesis. This means that the itself graphic should not have any title. Traditional submissions to journals require figure legends to be at the end of the manuscript, before the figures themselves. Many journals now allow you to embed a lower resolution image of the figure on first submission together with the legend underneath in order to facilitate peer review. Check on the requirements for you institution or journal before finalising this.

A trend with online journals is that each figure (and table) needs a topic or title sentence, followed by a description. This is a good way forward with any figure legend. Both need to be uploaded along with the image file when you submit the manuscript.

### 2021-04-15 3.11-Credit.Rmd
#First upload of book

## Who did what? {#credit}

More and more journals are asking authors to supply information about who did what. This is an attempt to increase transparency and reduce the incidence of ghost authorship.

There are a lot of politics around authorship. How do you decide whether or not someone did enough to be included on an author line. This is a very important point that we cover elsewhere. All that I will say here is that this should have been decided early on in the study. Everyone should know whether or not they are going to be an author at the time that they participate. No one should ever find out but they were not an author by reading their name in the acknowledgements.

### CRediT where it’s due

After many iterations, there is now a widely recognised list of 14 different ways in which authors can contribute towards a study (Table \@ref(tab:CRediT)). Even when the study is a chapter in your PhD thesis, it's unlikely that you can put your name next to every role in the CRediT taxonomy. This does give you some realistic perspective about the collaborative nature of science.

Table: (\#tab:CRediT) **There is now a recognised list of different ways in which authors contribute to publications.** The CRediT taxonomy allows you to ascribe contributions from each of the authors into one or more of 14 different roles.

| **Ways in which authors contribute to studies** |
|------|
| Conceptualisation | 
| Data curation |
| Formal Analysis |
| Funding acquisition |
| Investigation |
| Methodology |
| Project administration |
| Resources |
| Software |
| Supervision |
| Validation |
| Visualisation |
| Writing – original draft |
| Writing – review and editing |

If you want to know more about each of these roles and how they Are defined in the CRediT taxonomy then please visit the site. You can also ‘claim’ these roles under your [Rescognito profile](https://rescognito.com/), linked to your ORCID account.

Even if the journal that you are submitting to does not use the credit taxonomy I would suggest that you use this as the basis for your saying who did what in the publication. See a blog article on this [here](http://crosstalk.cell.com/blog/giving-authors-the-credit-they-deserve). Otherwise you could end up with what is known as hyperauthorship, a phenomenon evidenced by massive co-authorship levels [@cronin2001hyperauthorship].

### Ordering the author line up 
In Biological Sciences it is traditional for the author who did most of the work to take the first position. For you and your thesis you are most likely to be the first author, certainly for all of your thesis chapters.

The last position is normally for the head of the laboratory in which the work was undertaken. This would usually be your advisor, but could be another collaborator, if for example, you went to do some work in their laboratory. Funding also plays a role here, and if you are unsure about who should be listed last, then ask your advisor.

In between the first and last authors is a bit of a no-man's land. Some people order it by who did most (closest to first), but then there is also a push to be closest to last place (closest to being more senior). Alphabetical also works, but that often benefits or penalises those who have names that regularly appear higher up or lower down in the alphabet. One colleague is always first to suggest alphabetical order as he regularly ends up as last author. If it isn’t part of your thesis, should you still be an author?

There is such a thing as shared first (and even last) authorship. You see these appearing with increasing regularity, marked with an asterisk or other symbol. Hence, there can be several top spots, and these should work just fine for your CV.

### 2021-04-15 3.12-SuppInfo.Rmd
#First upload of book

## Supplementary Material {#suppinfo}

Twenty years ago, the idea of supplementary material for a journal article was practically unheard of. Today, it seems that few papers don’t have some supplementary material associated with them. Why this big change?

### What can go into Supplementary Material?
Practically anything!

- If you have lots of extra analyses that are too large to go into an article. 
- If your methods are extensive and there’s a word limit in the journal
- If your analytical tools throw out a bunch of super interesting graphics, but you can only use one or two
- Most of all, you can put in the raw data that makes up the basis for your work, together with the analytical script that you used to analyse it.

### There has been a big movement toward transparency and repeatability of scientific studies
This is a really positive movement in science and it should receive all of our support as it is part of the greater scientific project [see part 2](transparency.html). Being proud of your work is being proud of the entire product, including the data and the analysis. If you felt they were good enough for peer review, then you should be happy to make them all freely available. Not sure if you haven’t made some mistakes? We all make mistakes and you shouldn’t think that you will have an entirely fault free career. Whenever and wherever possible conduct your professional due diligence and be open and honest with your research. Science is a career when we should welcome errors and mistakes that we have made in the past. There’s no problem with this. Indeed, the problems only start when you refuse to acknowledge past errors and mistakes, making you prone to motivated reasoning and confirmation bias.

There are lots of data repositories out there. Some specialise in particular kinds of data, like genetic data in [GenBank](https://www.ncbi.nlm.nih.gov/genbank/), or sound data in [The Macaulay Library](https://www.macaulaylibrary.org/) at the Cornell Lab of Ornithology. Others are general repositories of data and you can post anything there, including the scripts that you used to analyse the data. By doing this you ensure that any people that want to use your work in the future have full access to it. Your data deposit will be given a DOI. You can use this DOI in your publication so that readers know where to find the data.

Once I happily allowed the publishers to store my data. These days, I simply don’t trust them to curate and keep my data for posterity. The way that most publishers of academic journals behave, there is no reason for them to make sure that my data is safe and secure. Therefore, I would rather go to an independent curator of data that specialises in not-for-profit data curation. One example is [Zenodo](https://zenodo.org/) which is built and operated by CERN and OpenAIRE. This site has the advantage that it is integrated with GitHub (if you are already a user, but check it out if you aren’t). You can choose to open your data immediately, or embargo it until your paper is published. It’s a great platform, and so much more secure than using a publisher. Another example is the [Open Science Framework (OSF)](https://osf.io/) that aims to be a one stop shop for all of your storage needs. OSF best practices can be found here. You will need to make sure that your data is correctly sorted with sufficient metadata for someone else to understand [see @roche2015public].

Once we did statistical analysis with a GUI and a point and click approach. Forget to click a checkbox and the results were not repeatable. As the analyses have become increasingly sophisticated with more and differing pieces of software, it is becoming increasingly important to record the code that you write that gets your results together with your data. This is the difference between making your results and therefore your study repeatable.

Talk to your advisor. Your lab or your institution may have a repository that they prefer.

### Why stop with just depositing your data and analyses online?
Let’s face it, you probably put a lot of effort into proposing your thesis hypotheses. Why not deposit this too? Remember that it does not have to be accessible to the world, but it would improve the integrity of your study if your original hypotheses are the same as the ones that you submit for publication. Some journals are now asking whether you have stored hypotheses ahead of conducting the experiments. They don’t stop you from submitting if you haven’t. But why not? It’s a great step forward in terms of transparency and provides more credibility when you eventually write that manuscript with the same aims. And it shouldn’t stop you from publishing another study that happened along the way, even though you hadn’t planned it. I suspect that in future funders will require us to deposit our successful funding proposals in an online repository before they release funds - hopefully they’ll also be willing to support these facilities financially.

### 2021-04-15 4.01-formatting.Rmd
#First upload of book

# Formatting your thesis  {#formatting}

Let’s face it, after spending so much time writing your thesis, having it formatted so that it looks very smart for submission is gratifying.

If you chose [Bookdown](bookdown.org) or [thesisdown](thesisdown.org) to write your thesis in, then you'll simply be able to download the correct style file (cls) and bibliography formatting file (bib) and press the build book button in [RStudio](www.rstudio.com) and you'll have your formatted thesis in the time that it takes your computer to render the pdf. You'll still need to look through it and make sure that the output is what you want.

For other, mere mortals, using word-processing packages, you will likely need to invest a lot more time trying to pull together the chapters of your thesis into the correct format. (I'd suggest that right now is not the time to learn Bookdown - unless you are way ahead of schedule!) Try to make it a fun event by listening to your favourite music (very loudly) or having some of your banned tipple (do both if you dare!). It doesn’t require the same level of creative concentration required when writing your thesis, but it does need great attention to detail.

Be patient and don't try to build mega-big files in word-processing documents, they tend to crash. Instead, copy the template file and compile one chapter at a time, placing them together only at the end for a final build.

Your university will have their own stipulation for exactly how they want the thesis formatted. Here, I provide a check-list for the last things to do before you submit. The chances are that your university will have a bunch of helpful links. Make sure that you follow the guidelines of your own institution well in advance of submitting your thesis.

### Must do check-list before submitting your thesis:
The following points are based on what may irritate or annoy your examiner (and you really don’t want to do that!):

- Spell check - yes, it sounds obvious but doing a final spell check is a good idea. Not only this, but take the time to have your word processor ignore or add all of the special words (e.g. species or site names) that it doesn’t otherwise recognise. This will ensure their consistency throughout (within and between chapters).
- Make sure that your language settings are set to ‘UK English’ (or the English setting for where you are based. 
- Look out especially for words that have different accepted spellings like those ending in -ise or -ize. Decide which you want and be consistent. Consistency is king!
- Capitalisation of [common names](taxonomy.html), place names and not adjectives. For example, ‘South Africa’ has two capitals, but ‘southern Africa’ only has one.
- Grammar check - always good to take a final look, especially for chapters that you wrote some time ago. 
 + Use the word processor automated options to help you. 
- Have your computer or another device read the text so that you can hear anything obviously wrong.
- Pay attention if you have used “we” or “I” and make sure they are consistent in your thesis. As a rule I encourage “I” in a thesis, unless the chapter is also a manuscript, in which case “we” is usually correct.
- Page layout. Really important to get this right in your template. Make sure that your template has:
  + Correct paper size (A4 and not US letter - or visa versa!)
  + Margins
  + Line spacing
  + Page numbers
  + Line numbers (really helps your examiners)
- Headers and Footers. If you can manage a chapter specific header, it’s useful to show your name and a short chapter title.
- Sections and subheadings. I’ve encouraged you to use subheadings throughout your thesis. Here you have a chance to number them sequentially. This is very useful for your examiners and may be a requirement for the university. Using the [word processor’s](software.html) built in functions will make this task consistent and easy.
- I dislike writing within a formatted document (as word processors can start getting weird), so my preference is to cut and paste written text into a template at the end. 
- Remember to give them a check through before handing it in. If you’ve done the sections correctly, then the contents page will come out right.
- [Title page](title:html) - prescribed very strictly by the university. The librarians place the SU watermark after final submission to the library. It may be tempting to change your title now that you know what's in the thesis, but many universities have strict policies about this. If you want to change the title, make sure that you are able to do so.
- Content page - word processors can do this automatically if your thesis is formatted correctly throughout (see sections and subheadings above). 
  + You can check this to make sure that you’ve done all of your sections and subheadings correctly.
- [Acknowledgements](acknowledgements.html) - this is your time to say thank you to all the special people that have helped during your study. There are probably more than you realise, but in addition to your friends and family (who most people don’t forget), think about the people who administered the work, lab mates (past and present) who were always there to help, and people who gave permission at study sites. 
- [References](references.html) - probably one of the most dreaded sections of any thesis preparation, but they do have to be done. If you’re one of these people that has everything in a [database](database.html), then you’ll be laughing or cursing your database throughout. While it might be tempting to only look through the data within the database, spend some time to see how it’s displayed in the thesis. A mistake in the reference database will be multiplied many times in the thesis. Remember that examiners love to take a random look through the reference section to make sure that it’s all good. After years of painfully entering references themselves, they know just what to look for.

### Mistakes people make:
Other than the obvious things, all mentioned above, here are some of the mistakes I’ve seen.

- Submitting the wrong version (yes this does happen!). Probably worse if having a mixture of right and wrong versions for different chapters (worse because it takes longer to sort out)
- Last minute additions to text with incorrect spelling and or grammar
- Two correctly spelled synonyms sitting next to each other when only one is desired (probably came about when editing)
- Forgetting to check for plagiarism (see here)
- Comments and or edited text (especially when it’s marked as being by someone other than the student).
- Page numbers that start again and again at different sections
- Lots of blank pages or spaces (avoid blank pages if you can, and try to limit the amount of blank space (never >half a page).
- Leaving important people out of the acknowledgements (e.g. advisor, administrators, funders, etc.)

### 2021-04-15 4.02-part4.Rmd
#First upload of book

# The obligatoin to publish your work {.unnumbered #publish}

Now that you've submitted your PhD, you need still need to do something with it. For most of you this will be publishing it in a journal of your choice. It will then take you through what happens to the manuscript and give you insight about where you might want to publish, how to choose the journal. Publication is important. Here is a list of reasons why you have some obligation to publish the work in your thesis:

- If your work is unpublished, then it will not be used. Without use, all the work that you put into it falls away.
- Your published work will become the foundation for future students and academics who are active in your field. This might well be in another country or on another continent.
- Your work was most likely paid for by tax-payers in the country where you studied. Publishing this work is a way of handing back the value of what you found. If it is published Open Access, the public can also read it for themselves. 
- Other funders of your thesis work may have made publication a prerequisite of the funding criteria.
- It is hoped that by publishing your work it will become more accessible to the scientific community.
- By publishing your work you will find that both you, and your work, become known by an international community that may well invite you to participate in the academic process further [@marks2013misuse].
- Both inside and outside of science as a career, peer reviewed published papers are seen as an important accomplishment in addition to the production of your thesis.
- The work that you have already put into your thesis (especially if you have followed the advice in this book!), will mean that you are close to having chapters that can be submitted as publications. 
- Your advisor and lab colleagues who helped with your thesis work may depend on your publications. 
- There is a lot of satisfaction to be had by seeing your work published. It has been argued that this depends on who you are and where you come from [@husemann2017publicationism]. 
- The satisfaction of having someone say how enjoyable or inspiring your work is to read will certainly not be as frequent as if you managed to publish a novel or newspaper column. But when it does happen, it will bring a smile to your face.

This is by no means an exhaustive list, but I hope it will give you some insight into the importance of publishing your results. To counter this assertion, please be aware that there is also research that suggests that publication driven science is not healthy, and that we need a new way to motivate science [e.g. @stergiou2014impact]. Indeed, it has been suggested that the ‘publish or perish’ mentality of academia has resulted in the retention (and even deterioration) of poor experimental design, and hence higher false positives in the behavioural sciences [@smaldino2016natural]. However, before you can join the debate about the future, I’d argue that you need to provide your credentials for the present.

Taking your work to the next level is not a trivial step, even though you already have your chapters written in the style of manuscripts. The challenges of publishing work, like other challenges in academia, are much easier if you possess inside knowledge of the system. The demystification of the next steps along the road in academia are the subject of another book:

### 2021-04-15 4.03-lastnote.Rmd
#First upload of book

# Last note {- #lastnote}

I really hope that this book has been helpful and that it has achieved what I set out to do: provide you with the guide on how to write a PhD in biological sciences. If you feel that this book has important items missing, out-of-date or simply wrong, then please help. Any good guide relies upon the people that use it to keep it viable.

You can contribute to the project at the Github pages for this book, using bookdown.

If I could emphasise one really important aspect of all of your PhD studies; that is to enjoy them!

### 2021-04-15 index.Rmd
#First upload of book

--- 
title: "How to write a PhD in Biological Sciences: a guide for the uninitiated"
author: "John Measey"
date: "`r Sys.Date()`"
documentclass: krantz
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
site: bookdown::bookdown_site
description: "You don’t have to be a genius to write a PhD. It will always involve a lot of hard work and dedication, but the process of writing up is a whole lot easier if you understand the basic ground rules. This book is a guide specifically for writing a PhD in the *biological sciences*. It will guide you from deciding whether or not you should actually do a PhD, provide practical, up-to-the-minute guides to getting started with planning and writing, and consolidate this with the nuts and bolts of writing for the biological sciences. The book guides you through the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication in peer-reviewed journals. Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
github-repo: yihui/bookdown-crc
graphics: yes
#cover-image: images/cover.png
---

```{r setup, include=FALSE}
options(
  htmltools.dir.version = FALSE, formatR.indent = 2, width = 55, digits = 4
)

# install the packages needed by this book; you fill out c(), e.g. c('ggplot2', 'dplyr')
lapply(c('xfun'), function(pkg) {
  if (system.file(package = pkg) == '') install.packages(pkg)
})
```

# Welcome {-}
Welcome to this guide on how to write a PhD in biological sciences. I will not tell you what to write. My intention is to tell you how to approach writing with respect to style, format and formula (planning).

Writing a PhD is a daunting prospect. One does not put pen to paper, or, rather, finger to keyboard, only after finishing all of your experiments and/or labwork. Writing is actually something you should start from day one of your PhD, or even before day one, when you make the first initial pitch of big ideas to your advisor on what you want to do. If you can find the discipline to start writing from day one, then you are much more likely to make a much better go of it than waiting until the final, pressured, stretch. But to do this, you need to understand how to write a PhD thesis. You need to understand what you are aiming to achieve, and, crucially, you need to understand why you are doing your PhD.

This guide intends to demystify the writing process so that the eventual goal is clear, and provide the necessary information on how to get there. It takes you from those initial ideas, all the way through proposal writing, posing hypotheses, testing them, and then how to write it all up.

This book started as a series of blog posts written by me for my students (john.measey.com/blog). I wrote them because I felt that I was repeating the same advice year after year. Instead of rewriting comments on manuscripts time after time, I thought I would consolidate them and put them all in one place. Later, I started looking for particular problems that students were having, and then writing blog posts to try to preempt those problems, and to provide solutions before tasks were completed. Finally, I have tried to fill in the missing gaps in the blog posts. But there are still plenty of gaps, and I regard this book as a work-in-progress. It is my intention that this book will always be added to, and improved by a community of practice, and I hope that this will include you.

## Why read this book {-}

Embarking on a PhD is intimidating as, for most students, it will be their first experience working within the academic system. The voyage of discovery is often made very frustrating as much of what goes on in academia is assumed knowledge. Academics accumulate knowledge throughout their careers, but what can be done for those who are uninitiated? What is needed is a guide. In the same way that we had guides to foreign countries and cities, postgraduate students need a guide that newcomers can refer to before, during and when making decisions about their time within academia. Note that this is not a rulebook. There are times when the guide will be accurate and others when it will be vague, but providing some insight to point you in directions where you can explore more.

## Structure of the book {-}

This book is intended to be a guide. Don't sit down and read it from cover to cover. If you do, then I think you have likely misunderstood what this guide is for. Instead, I want you to think about this book as a guide much in the same way that you might pick up a guide to a foreign city, country or even like Douglas Adams' fabled electronic guide _The Hitchhiker's Guide to the Galaxy_ @adams2017ultimate (if you are not familiar [see here](https://en.wikipedia.org/wiki/The_Hitchhiker%27s_Guide_to_the_Galaxy)). However, instead of having hostels or restaurants recommended by name with a description of what makes them particularly praiseworthy, this guide provides a generic idea of what such places are for, what they tend to be like, the best ways to get to them, and pitfalls to avoid.

None of the topics is explored in any great depth. This is deliberate as this guide is here to provide you with a starting point for you to explore if you think it is necessary. I have provided literature (where possible), and you should use this in the standard way to research the topic yourself should you need or want to know any more. In the same way, none of the advice provided in this book should override what you decide for yourself or with your advisor. Take your decisions from an informed position.

**The book is divided into three main parts:**

[**Part 1**](part1.html)
Part one is written as something that you should read early on in the process, once you’ve committed to doing a PhD. This shortest section of the book deals with what a PhD is, how it’s examined, with whom, where and when to do it. At the start of your PhD is the time to be thinking of the bigger questions both in your study area, and your life, and this section provides some pointers and things to consider.

[**Part 2**](part2.html)  
Part two of the book sets out a practical guide to getting started with writing. The approach I've taken here is describing the production of a PhD in biological sciences which is made up of data chapters, and where each chapter is essentially a potential publication in a science journal. If you didn't want to do a PhD but just wanted to write a publication, then most of what is here (and in Part 3) will be of use as you could take the same formulaic approach as given in this book. Similarly, if you were writing a MSc project, this section would also be useful. There is a lot of technical stuff involved in writing science, and in part 2 I cover as much as I can while still remaining general to the subject area.

[**Part 3**](part3.html)
A typical PhD will consist of five data chapters where each data chapter is a publication. There is a shortish introductory chapter (sometimes a literature review) that comes before these data chapters, and a concluding synthesis chapter at the end. That's not to say that this is the only way of writing a PhD. There are many different ways and it is a good idea for you and your advisor to talk about this together before you start.

This part covers how to write the core sections of a paper: abstract, introduction, methods and materials, results, discussion and references.

Throughout this guide you will find links and references of things to read elsewhere. The reality is that there is a lot out there for you to read. I am not the only person to write a guide, and I can't pretend that all the answers are here in this book. You will find that in your particular situation you need to look around. What I aim to do in this book is provide you with enough background information to help in your onward search, and provide you with enough inside knowledge to do a PhD.

## Why 'A guide for the uninitiated'? {-}

I think that most people with doctorates would agree that a PhD is not awarded to people because they are particularly bright or smart. If you had to be a genius then I wouldn't have a PhD. Indeed, I don't consider myself to be particularly clever, but I worked very hard to get my PhD. I was hampered by the fact that I didn't know anything about the goals and aims of the academic process of working towards a PhD, so it took a lot more work, wasted time, and (let's not mince our words) real pain. The end product was a fraction of the potential that I could have achieved, if I had understood more about the process. If I had only had a guide to tell me what it was all about, I could have saved myself so much time and energy. In short, I feel that I was uninitiated, and this is the guide I wish that I had had.

So, this guide is my practical attempt to help you to get the most that you can out of your PhD time. After all, it should be an amazing journey that culminates with you achieving academic maturity: the ability to conceive and carry out scientific investigations. When you have a PhD, you become a scientist. Too often, however, it's a nightmare journey of cul-de-sacs, and groping in the dark.

You should be aware from the outset that doing a PhD is a major undertaking. It will take several years of your life and will impact heavily on your day to day life, to the extent that your mental health may suffer. A PhD represents the culmination of an incredible amount of work, not to mention that you've already had to complete one to three previous degrees beforehand. It may not be something that you ever finish to your own satisfaction. But finish it you must.

## Software information and conventions {-}

I used the **knitr**\index{knitr} package [@xie2015dynamic] and the **bookdown**\index{bookdown} package [@R-bookdown] to compile my book.

## Acknowledgments {-}

Before I go on, there are a great many people that I need to thank. First and foremost are my students, past and present, who have inspired me to put together first the blog posts and then the book. It is because you wanted more that I put this together. I have also been a student, and have been inspired by colleagues around the world who have been exemplary advisors. This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help. I thank this greater academic community for sharing and inspiring. Thanks go to the many reviewers and editors who have taken their time to improve my writing. I am still learning. Lots of the text in this book has been improved by feedback from my students and postdocs. A special mention must go to my brother, Richard, who has hosted my lab website for more than a decade, and especially for saving blog posts from hacking attacks. Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning. James Baxter-Gilbert, Anthony Herrel,  Lisa O'Connell, Mark O'Connell, James Vonesh all read parts and commented. Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and Stellenbosch University. A special thanks to the librarians who have supported many of my more extreme rantings about publishers.

```{block2, type='flushright', html.tag='p'}
John Measey  
Cape Town
```

### 2021-04-15 1.01-beforeyoustart.Rmd
# Changing sections to chapters

@@
 
-### What is a PhD?
+## What is a PhD?
 
@@
 
-### Is a PhD more than a thesis?
+## Is a PhD more than a thesis?
 Yes.
@@
 
-### Criteria used to judge PhD studies
+## Criteria used to judge PhD studies
 In order to provide you with some general idea of what is required, I list below the (edited) questions asked of examiners for a PhD thesis by my university:
@@
 
-### Who are the examiners?
+## Who are the examiners?
 The next question to ask is who the examiners are?
@@
 
-### What will you do after your PhD?
+## What will you do after your PhD?
 It sounds silly to think about what comes after your PhD even as you are just starting, but the reality is that you always need to be planning ahead in order to make the most from what you are doing now.
@@
 
-#### Doing a PhD actually reduces your chance of employment and lowers your pay
+### Doing a PhD actually reduces your chance of employment and lowers your pay
 The reality is that most employers don’t want someone with a PhD, because it is a very academic qualification.
@@
 
-### Think about your career before you start your PhD
+## Think about your career before you start your PhD
 The last thing a PhD should be is a stop gap for you to fill your time because you can’t think of anything else to do.
@@
 
-### Don't rely on becoming an academic
+## Don't rely on becoming an academic
 Academia is getting more competitive.
@@
 
-### Doing a PhD is really hard work
+## Doing a PhD is really hard work
 It’s not a MSc with an extra year.
@@
 
-### Your mental health will likely suffer
+## Your mental health will likely suffer
 Doing a PhD will be the cause of stress not only in your working life, but will also impact other areas of your life.
@@
 
-### Any PhD has to be yours
+## Any PhD has to be yours
 It’s entirely useless to do a PhD for someone else.
@@
 
-### Does it matter where you do your PhD?
+## Does it matter where you do your PhD?
 Not all universities in the world have a PhD program.
@@
 
-### What to look for in an advisor
+## What to look for in an advisor
 Throughout this book I will call the person who is the academic in charge of your PhD studies your ‘advisor’.
@@
 
-### What to avoid in an advisor
+## What to avoid in an advisor
 Remember that relationships occur on a one-to-one basis, and it may be that not every advisor gets on with every student.
@@
 
-### What makes a good project?
+## What makes a good project?
 After having made lots of statements about how a project should be your own project the following advice might seem a bit strange.
@@
 
-### Funding your PhD studies
+## Funding your PhD studies
 Funding may be one of the biggest obstacles that you come across when wanting to do a PhD. It's important that you consider the funding of your PhD in two separate ways:
 
-#### The funding that you need to live 
+### The funding that you need to live 
 This money is typically called a bursary or stipend depending on where you are in the world.
@@
 
-#### The cost associated with your studies. 
+### The cost associated with your studies. 
 The money associated with the project that you were doing also needs to come from somewhere.

### 2021-04-15 1.02-useR.Rmd
# Updating sections to chapters

@@
-## Reconciling yourself to doing things you've been avoiding {#useR}
+# Reconciling yourself to doing things you've been avoiding {#useR}
 Some 30 odd years ago as an undergraduate I was told that I would have to use computers in one of my 2nd year courses.
@@
 
-### And learning lots of other stuff...
+## And learning lots of other stuff...
 The same principle that I’ve written about above can be extended to all sorts of other aspects of your PhD project.

### 2021-04-15 1.03-lifescientific.Rmd
# Updating sections to chapters

@@
-## The scientific project and scientific living {#lifescientific}
+# The scientific project and scientific living {#lifescientific}
 
@@
 
-### Example: Invasive species 
+## Example: Invasive species 
 Alien species are those that have been moved by human agents across biogeographic barriers into parts of the world where they did not evolve.
@@
 
-### Should society lead the sciences or sciences lead society? 
+## Should society lead the sciences or sciences lead society? 
 Of course, the answer is simply that there must be feedback between science and society.
@@
 
-### Carl Sagan’s “Baloney detection kit”
+## Carl Sagan’s “Baloney detection kit”
 
@@
 
-### Live your life scientific
+## Live your life scientific
 >__“Science is more than a body of knowledge, it is a way of thinking”__
@@
 
-### Citizen science
+## Citizen science
 For a long time I struggled with the term 'Citizen science' as I reasoned that many such ventures did not involve citizens doing science.

### 2021-04-17 2.04-BigIdea.Rmd
Add files via upload

@@
-## What's the big idea? {#bigidea}
+# What's the big idea? {#bigidea}
 In the previous chapters, I've talked about the importance of having a hypothesis, and building that hypothesis in a logical framework within the introduction (see part 3).
@@
 
-### The funnel introduction 
+## The funnel introduction 
 But so much of what you read doesn't have any clear hypothesis, and there doesn't seem to be any thought about building on bigger concepts or ideas.
@@
 
-### So where would we find all of these big ideas?
+## So where would we find these big ideas?
 There are quite a few papers that synthesise hypotheses in various areas of biology.
-Here I provide two to get you started.
+Here I provide a few to get you started.
 Ask your advisor about relevant reviews of hypotheses in your area of biological sciences.
@@
 The next is by Jane Catford and colleagues [-@catford2009reducing] on hypotheses in invasion biology, but I encourage you to look for more up to date versions.
+Lastly, consider these wide ranging papers about so-called 'laws' in biology and ecology [@dhar2010laws; @lawton1999are].
 Each of these papers will give you a list of big ideas, together with the citations for seminal papers that have built them.
@@
 You can think of the way that such ideas are structured as a hierarchy of hypotheses (Figure \@ref(fig:JeschkeHagar61)).
+Building such a hierarchy of ideas in your own specialist subject area is good reference material to prompt [critical reading](critical.html) and reflection.
+You may want to take this structure forward when thinking how to [introduce your PhD chapters](PhDreview.html).
 
-(ref:JeschkeHagar61) **A hierarchy of hypothses can be used to help determine your Big Idea.** A useful structure for thinking about how hypotheses are structured was presented by @heger2018hierarchy in what they termed the _Hierarchy of Hypotheses_
+(ref:JeschkeHagar61) **A hierarchy of hypothses can be used to help determine your Big Idea.** A useful structure for thinking about how hypotheses are structured was presented by @heger2018hierarchy in what they termed the _Hierarchy of Hypotheses_.
 

### 2021-04-17 2.06-argument.Rmd
Add files via upload

@@
-## Construct a logical argument in your writing {#argument}
+# Construct a logical argument in your writing {#argument}
 
@@
 
-### Arguments in Latin
+## Arguments in Latin
 From time to time you may come across comments about arguments that you’ve made in your work made in Latin.
@@
 Another way to think of this is if you have two models that you test are equally likely (e.g.
-within 2 dAICs), the one with the least number of variables should be considered superior.
+within 2 δAICs), the one with the least number of variables should be considered superior.
 This is also known as the priciple or law of parsimony: _lex parsimoniae_.
+
+## Other arguments in English
 

### 2021-04-17 2.20-proposal.Rmd
Add files via upload

@@
-## What is needed for your research proposal? {#proposal}
+# What is needed for your research proposal? {#proposal}
 
@@
 
-### Structure
+## Structure
 
-We expect that your final thesis will be a set of papers (data chapters sandwiched between an introduction and discussion), and so we ask you to prepare the proposal in the same way.
+We expect that your final thesis will be a set of papers (data chapters sandwiched between an [introduction and discussion](PhDintro.html)), and so we ask you to prepare the proposal in the same way.
 That is, you will normally have two (MSc) or five (PhD) data chapters.
@@
 
-### Proposal introduction
+## Proposal introduction
 
@@
 
-### Methods and Materials
+## Methods and Materials
 
@@
 
-### Hypothetical Results
+## Hypothetical Results
 
@@
 
-### Your Proposal - what is it good for?
+## Your Proposal - what is it good for?
 
@@
 
-#### Copy and paste directly into your thesis
+### Copy and paste directly into your thesis
 
@@
 
-#### Use it to raise money to do your studies.
+### Use it to raise money to do your studies.
 
@@
 
-#### Deposit the proposal to conform to Design and Analysis Transparency
+### Deposit the proposal to conform to Design and Analysis Transparency
 

### 2021-04-17 3.01-Title.Rmd
Add files via upload

@@
-# Writing the sections that make up your chapter  {#part3}
+# Writing the sections that make up your data chapter  {#part3}
 
@@
 
-### The Title
+## The Title
 
@@
 
-#### Some title ideas to start you off
+### Some title ideas to start you off
 - Don’t start by looking for the best title
@@
 
-#### Things to avoid in your title
+### Things to avoid in your title
  - Don’t feel obliged to include [taxonomic terms](taxonomy.html) unless it is relevant or compulsory
@@
 
-### Key Words
+## Key Words
 The key words are a way for readers to find your content with searches.

### 2021-04-17 3.05-Results.Rmd
Add files via upload

@@
-## The Results {#results}
+# The Results {#results}
 
@@
 
-### The correct approach - responding to your hypotheses
+## The correct approach - responding to your hypotheses
 Yes, the easiest way to write the results is simply to have the questions or hypotheses that you posed at the end of the introduction and have them as subheadings for your results section.
@@
 If you do have to insert the results of a statistical test into the text, include the test statistic, degrees of freedom (which incorporates N - the sample size) and the value of P.
-Only if P is less than alpha (usually 0.05) can you use the word ‘significant’ when describing the effect.
+Only if P is less than α (alpha is usually 0.05) can you use the word ‘significant’ when describing the effect.
 
@@
 
-### When to have a table or graph?
+## When to have a table or graph?
 Most journals will allow you to have as many tables or graphs as you need to give your results.
@@
 
-### Descriptive results
+## Descriptive results
 Some studies have a set of descriptive results that, by convention, are placed without a section header at the start of the results.
@@
 
-### Responding to hypotheses
+## Responding to hypotheses
 While I encourage you to present data that corresponds to hypotheses here in the results section, you should not formally accept or reject your hypothesis here.
@@
 
-### Common Mistakes
+## Common Mistakes
 - Having extra analyses in the results that you didn’t include in your methods.

### 2021-04-17 4.00-part4.Rmd
Add files via upload

# Now that you have finished your data chapters {#part4}

In this last part of the book, I am going to presume that you have done all of the work for the data chapters, and written them up as distinct chapters of your thesis. 
When you look at PhDs that are stored in your institutions libraries, or those that you've referred to in the course of your study, you will notice that they have two additional chapters in addition to data chapters. These are often quite short (although lengths vary), and serve to introduce and conclude the PhD thesis (Figure \@ref(fig:thesis-construction)). While these are not usually considered to be the 'meat' of the PhD itself, they are important sections to reflect on and carefully consider, as they attempt to pull together data chapters that are written as stand-alone documents into a cohesive thesis. To this end, they require the same careful planning that any other section in your thesis deserves, with one important condition: that they should not impede the timely submission of your thesis.

(ref:thesis-construction) **A typical construction of the data chapters in your thesis together with a introduction and conclusion.** The exact construction of your thesis may vary, especially with the number of data chapters that you have. The relationship between thesis chapters may also not be linear as presented here.

```{r thesis-construction, echo=FALSE, fig.align='left', out.width='90%',  fig.cap="(ref:thesis-construction)"}
knitr::include_graphics('figures/thesis-construction.png')
```

At this point, we have to acknowledge that different institutions require different deadlines of their students in order to finish their PhD studies and submit their theses for examination. In most institutions (to my knowledge) the time to submission is the most important, and can vary between 3 and 5 years (most institutions stipulate a minimum period of 2 years and some have no maximum duration). This period is critical with respect to funding. If funding ceases and your institution allows you to continue your studies, you may have to raise the funding to pay fees as well as to maintain your living costs. After submission there will be a period of time while your thesis is in examination, and another period before the final graduation. Each of these milestones comes with its own set of deadlines and conditions. It is well worth being aware of all these issues well in advance of any submission deadline (ask your advisor). Important aspects for you to note are whether or not you need to remain registered at your institution after you have submitted your thesis, and whether or not your institution will require payment for your registration.

Whatever your deadlines are will dictate how much time (and energy) you have to devote to the opening and closing chapters of your thesis. They can become publishable in their own right, or at the bare minimum, they can serve to allow you to submit a completed thesis. Here I will focus on the latter, minimum requirement (in my eyes), with the understanding that you should dedicate as much time as you have available to produce a quality product that you can feel proud of.

The next chapters deal with writing the [introductory chapter](PhDintro.html), the [concluding chapter](PhDconc.html), and then finally [formatting your thesis for submission](formatting.html). These final hurdles require some time and thought. Given that not all of you will have all of the time that you need, this is often the time when the demand on your writing skills reaches a peak. If you are in this position, I urge you to use your time wisely and to plan carefully what you will write.

### 2021-04-17 4.01-PhDintro.Rmd
Add files via upload

# How to introduce your PhD chapters {#PhDintro}

For each of your data chapters, you will have already written an [introduction section](introduction.html), and the same principles that you used to construct this will apply here. Instead of aiming to introduce any one of the hypotheses in the chapters, however, your introduction will aim to introduce the ideas and concepts explored inside the data chapters of the thesis.

My suggestion would be for you to start with the conceptual diagram of the thesis that you produced in your [proposal](proposal.html). If you didn't construct one then, you may be reluctant to do it now, but having a good conceptual idea of how the components of your thesis fit together will really aid you in putting together both this introduction and the final conclusion. For your [proposal](proposal.html), you may also have written a general introduction, and it would be worth revisiting this now, even if you don't actually use any of it.

Once you have a conceptual diagram of your thesis, my suggestion is that you formalise this into a diagram that you can present in the introductory chapter. Even if it's a very simple concept (like figure \@ref(fig:thesis-construction)), it will help your examiners (and any other readers) understand what to expect in your chapters and how they are interrelated.

Next produce a hierarchy of hypotheses for all of the chapters and concepts in your thesis (see figure \@ref(fig:JeschkeHagar61)). This will help you to make sure that you cover all of the [big ideas](bigidea.html) and concepts in the outline. Your outline should follow the standard funnel [formula](formula.html) (see Figure \@ref(fig:introduction)) this time with the highest level ideas in your hierarchy of hypotheses given full prominance in the first paragraph.

Although it's tempting, I'd urge you not to copy any text from any of the introductions in any of your chapters. By all means, draw upon their contents and references therein to give you inspiration, but this introductory chapter can't be a cut and paste job (Figure \@ref(fig:plagiarism)). The reason being that your examiners will spot this straight away, and it won't put your work in the best light. As before, if you can produce an outline of what you need to say in your introductory chapter, you should be well practised enough (by now) to write original text that will do the job of introducing all of the chapters.

### Citing your chapters

Your aim in this introduction is to introduce all of the chapters in your thesis, thus you will to cite them. My suggestion would be that you leave these citations to the final paragraphs (or subsection) of the introduction, and not have a structure that centres around introducing each chapter as a paragraph. Remember that you may well have this (in the form of abstracts) throughout the thesis, so it would be redundant to do that here. Instead, try to introduce chapters by concentrating on their themes and/or contributions to the bigger ideas that you address in your work.

The way in which to cite your chapters is most straightfoward here by using their chapter number (i.e. **Chapter n**). This will make it obvious to the examiner (or any other reader) that it is within the content of the thesis. If chapters are published, you could add a citation to the paper in addition, but not instead of.

### Formulate the objectives of your study?

At this point in your thesis writing, it is worth looking back to the [criteria used to judge a PhD at your institution](beforeyoustart.html). Yours may differ from those stated in this book, but it is clear that you should be able to formulate the objectives of your study, and that this includes not only the objectives of the individual data chapters, but of the study overall (here in the introductory chapter).

By the end of the introductory chapter, anyone reading should know what the thesis is about, and what it is that you aim to find out. This doesn't mean that they know what you found in each of the data chapters, but that they know the topics explored within the thesis and what was done to address them. Specifically

### Review consistent thesis themes

If your thesis has a study organism, experimental approach, or environmental system, then it would be appropriate to provide a reasonably comprehensive review of this in the introduction. The aim of this would be to explain why this theme was chosen. The extent of the review would depend entirely on how much time and energy you have to devote to this.

Ideally, you would aim for your first chapter to be a review of the literature in your study area, with the last section of the review rationalising the need for the particular approach that you have taken to address unexplored areas. However, as stated above, it may well be that you don't have the time to write a comprehensive review for your introductory chapter. If you do, then the most important aspect will be what to cover in the review (exactly what portion of the literature), and whether and how you want to publish it. Publications of reviews broadly fall into traditional reviews and systematic reviews (including meta-analyses). The latter require considerable amount of time and effort, and systematic reviews (i.e. those in which you assess an unbiased slice of literature on a topic) might be the only way to publish a review in some journals or subdisciplines.

Although the review of consistent themes could be anything from two pages to an entire manuscript, you should make it clear to the examiner (or any other reader) why, out of all potential options, you have taken the particular approach you have.

### Removing the contents of chapters to place in the introduction

You may decide that some of the contents of (particularly data chapter 1) might be more appropriate if it was moved to the introductory chapter. Before you do this I'd suggest that you discuss it with your advisor. Each of the data chapters should have already been signed off (by your advisor) and so it would be a mistake to start chopping them now without consultation. It is also perfectly legitimate to refer in the introduction to a section of the thesis where a particular topic is dealt with in detail. If you are unsure, you'd you best advised by your advisor.

## How long should the introductory chapter be?

This is probably the most common qestion I get asked, most often by students that have only days left in which to write it. Obviously, there is no simple answer (just as there is no particular length required for a PhD). Typically, I would suggest that the introductory chapter be longer than any of the introductions to any of the data chapters within your thesis, but not longer than the data chapters themselves - unless you've managed to pull off a stand-alone review.

### 2021-04-17 4.02-PhDconc.Rmd
Add files via upload

# How to conclude your PhD {#PhDconc}

By the time you come to write your concluding chapter of your PhD, you should have written all of the [data chapters](part3.html) and the [introductory chapter](PhDintro.html). Firstly, I should congratulate you for getting this far. You've almost finished, and there's just one final push required to complete the document. You are now the best placed person to tell yourself what should be in the concluding chapter.

## Using the funnel structure
In principle, you can use the same funnel structure that we described for the [discussion](discussion.html) as a way of structuring your concluding chapter (see figure \@ref(fig:discussion)). This would include starting with a synopsis of what your thesis is trying to achieve. Clearly, this starting section may be more than a paragraph (as it was in the [data chapters](part3.html)), and the precise structure will be dictated by the complexity of your thesis structure (Figure \@ref(fig:thesis-construction)).

## What to include in the concluding chapter?

I suggest that you take your conceptual outline (from the [introductory chapter](PhDintro.html)), and use this to being thinking about how best to construct the outline of your concluding chapter. If your data chapters are used to feed into a final output, then consider using the approach of a [logical argument](argument.html) as the outline.

The main body should review the concluding findings of each of the data chapters, and especially the way in whcih they interact - as these are the aspects that are least likely to have been touched on already in the individual data chapters. It is hoped that the thesis represents a body of work, and not a series of unconnected works. Thus your task in this final chapter is to present to the examiner (and any other reader) exactly how the chapters fit together into a body of work.

Your concluding chapter should provide all of the highlights and insights that have been gained from your PhD thesis. It is unlikely to you need to dwell on individuals caveats from the chapters, unless there are general caveats from your approach that span chapters - in which case these probably deserve attention.

Your concluding chapter will cite the data chapters of your thesis quite heavily, as it essentially pulls them all together, along with the objectives that they aimed to achieve, in the overarching concepts of your study area.

You also have an opportunity to point to where, in your opionion, the direction of your particular field should move next. **Do provide your opinion**. Remember that you are now one of the most qualified people in this particular area to be pronouncing on what should be done. Your opinion is therefore valid, and your concluding chapter is a good place to provide a reasoned argument about your own ideas.

## Concluding section

Your PhD may not have one overriding simple conclusion, indeed it will likely not have, so don't try to force one if there isn't one. Instead, attempt to sum up your findings in a way that explains how they have added to and changed the existing body of knowledge in your subject area.

### 2021-04-17 4.04-lastnote.Rmd
Add files via upload

# Last note {- #lastnote}

I really hope that this book has been helpful and that it has achieved what I set out to do: provide you with the guide on how to write a PhD in biological sciences. If you feel that this book has important items missing, out-of-date or simply wrong, then please help. Any good guide relies upon the people that use it to keep it viable.

If you have found this book useful, then please remember that you can contribute to the project at the Github pages for this book, using bookdown. If your contributions, no matter how small, improve the book, then you will be improving the experience of doing a PhD in Biological Sciences for future PhD students, and that has to be worth something.

Remember that there is another book that starts off where this book ends: from thesis chapters to published papers. It also tackles some of the broader issues within academia, which are likely to be relevant if you are planning your career in this area.

**If I can emphasise one really important aspect of all of your PhD studies (including the writing); that is to enjoy them! **

### 2021-04-17 index.Rmd
Add files via upload

@@
 --- 
-title: "How to write a PhD in Biological Sciences: a guide for the uninitiated"
+title: 'How to write a PhD in Biological Sciences: a guide for the uninitiated'
 author: "John Measey"
@@
 documentclass: krantz
-bibliography: [book.bib, packages.bib]
+bibliography:
+- book.bib
+- packages.bib
 biblio-style: apalike
@@
 site: bookdown::bookdown_site
-description: "You don’t have to be a genius to write a PhD.
-It will always involve a lot of hard work and dedication, but the process of writing up is a whole lot easier if you understand the basic ground rules.
-This book is a guide specifically for writing a PhD in the *biological sciences*.
-It will guide you from deciding whether or not you should actually do a PhD, provide practical, up-to-the-minute guides to getting started with planning and writing, and consolidate this with the nuts and bolts of writing for the biological sciences.
-The book guides you through the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication in peer-reviewed journals.
-Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
+description: You don’t have to be a genius to write a PhD. It will always involve
+  a lot of hard work and dedication, but the process of writing up is a whole lot
+  easier if you understand the basic ground rules. This book is a guide specifically
+  for writing a PhD in the *biological sciences*. It will guide you from deciding
+  whether or not you should actually do a PhD, provide practical, up-to-the-minute
+  guides to getting started with planning and writing, and consolidate this with the
+  nuts and bolts of writing for the biological sciences. The book guides you through
+  the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication
+  in peer-reviewed journals. Whether you are a student just starting your studies,
+  or a supervisor struggling to cope, this book will provide the insider information
+  you need to get ahead.
 github-repo: yihui/bookdown-crc
 graphics: yes
-#cover-image: images/cover.png
 ---
@@
 
-**The book is divided into three main parts:**
+**The book is divided into four main parts:**
 
@@
 The approach I've taken here is describing the production of a PhD in biological sciences which is made up of data chapters, and where each chapter is essentially a potential publication in a science journal.
-If you didn't want to do a PhD but just wanted to write a publication, then most of what is here (and in Part 3) will be of use as you could take the same formulaic approach as given in this book.
+If you didn't want to do a PhD but just wanted to write a publication, then most of what is here (and in [Part 3](part3.html)) will be of use as you could take the same formulaic approach as given in this book.
 Similarly, if you were writing a MSc project, this section would also be useful.
@@
 This part covers how to write the core sections of a paper: abstract, introduction, methods and materials, results, discussion and references.
+
+[**Part 4**](part4.html)
+The concluding part of this book is written to cover the things that remain after you have finished writing all of your data chapters.
+Most importantly, these include the introductory and concluding chapters that sit as bookends to the data chapters that you have already produced.
+These chapters make your thesis into a 'body of work', and allow you space to reflect on what you have achieved in your PhD studies.
 
@@
 
-## Software information and conventions {-}
-
-I used the **knitr**\index{knitr} package [@xie2015dynamic] and the **bookdown**\index{bookdown} package [@R-bookdown] to compile my book.
-
 ## Acknowledgments {-}

### 2021-04-18 1.01-beforeyoustart.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
 
-Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you are going to do afterwards.
-There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make the informed decisions from the outset.
-For example, reconciling yourself to using the statistical package R will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on.
-Getting, and staying in, the right mental framework will also be very important for your ability to work under pressure.
-Again, getting into good habits at the beginning will put you in a commanding position later.
+Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you plan to do afterwards.
+There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make informed decisions from the outset.
+For example, reconciling yourself to using the [statistical package R](useR.html) will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on.
+Preparing yourself into the right mental framework will also be very important for your ability to work under pressure.
+Getting into good habits at the beginning will put you in a commanding position later.
+The aim of this first part of the book is to help you assess your current position, and realign yourself toward an optimum mindset.
 
 ## So you are doing a PhD? {#beforeyoustart}
+
+Universities are worried about two major aspects of those who start PhD studies:
+
+ - The median time to completion in many countries worldwide is much longer than it should be [e.g. 7.5 years in the USA @geven2018how].
+ - Completion rates are shockingly low [as low as 50% in the USA @litalien2015improving].
+
+This has spurred many responsible for postgraduate students to try to ensure that they receive more support for their studies.
+But there is also the need for students to comprehend right from the start that they are undertaken a major commitment that will likely impact many different parts of their lives.
 
@@
 If your intention were to do the 'bare minimum' required to get a PhD, then you would have to look carefully for an advisor that would be happy to preside over such a thesis.
-Suffice it to say that this would be a very poor place to start, and there may be lots of other reasons why doing a PhD really is not appropriate for you.
+Suffice it to say that this would be a very poor place to start, and if this were your attitude there may be lots of other reasons why doing a PhD really is not appropriate.
 
@@
 The time that you spend doing your PhD will likely include participating in conferences, giving talks, participating in journal clubs, seminars, workshops and working groups.
-In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular area.
+In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular area, and learn to communicate this to different types of audience.
 You will also learn and develop your skills in presentation, analysis and treatment of data.
@@
 This is not unfair.
-Personally,
-I think that there are some special extra hurdles required for a PhD (over and above 5 chapters that are publishable):
+Personally, I think that there are some special extra hurdles required for a PhD (over and above 5 data chapters that are publishable):
 
@@
 
-I prefer a thesis that fits together within a stated framework.
-I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the introduction of their proposal.
+I prefer a thesis that fits together within an explicit [conceptual framework](PhDintro.html).
+I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the introduction of their [proposal](proposal.html).
 A useful concept to be aware of is the 'Hierarchy of Hypotheses' [@heger2018hierarchy].
-In this approach, you are encouraged to consider the 'bigger question' in your research area, identifying both what studies have produced evidence and identifying gaps from a theory driven approach.
+In this approach, you are encouraged to consider the '[bigger question](bigidea.html)' in your research area, identifying both what studies have produced compelling evidence, and identifying knowledge gaps from a theory driven approach.
 Generally, it is only possible to consider these types of concepts once you are familiar with the literature (both theoretical and experimental).
@@
 ## What will you do after your PhD?
-It sounds silly to think about what comes after your PhD even as you are just starting, but the reality is that you always need to be planning ahead in order to make the most from what you are doing now.
+It sounds silly to think about what comes after your PhD just as you are just starting, but the reality is that you always need to be planning ahead in order to make the most from what you are doing now.
 During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas.
@@
 It is important to realise that many academics are looking for good students to do PhD work.
-This is because they need someone to do research within their groups, and students or postdocs do most of this work.
+This is because they need someone to do research within their groups, and students or postdocs do most of this work [see @chamberlain2016ten].
 For this reason, an academic you know while you're doing your MSc studies may start to try to persuade you to do a PhD.
@@
 This is unusual.
-Most countries now have a restriction on the amount of time that you can spend studying a PhD.
+Most countries now have a restriction on the amount of time that you can spend studying a PhD, and given that completion times are so long [@litalien2015improving], there will be added pressure to finish.
+It may be important to consider the institutional support given to postgraduate students, outside of their research sphere.
 
@@
 If you are a biologist then you might want to consider where your study animals or plants are in order to decide your place to study.
-But most of all your choice of where to study is likely to be dictated by the presence of an adviser that you want to work with, and who wants you as their PhD student.
+But most of all your choice of where to study is likely to be dictated by the presence of an advisor that you want to work with, and who wants you as their PhD student.
 Advisors are tied to their institutions, and any principal advisor will need to register you at their own institution, meaning that this will be where you spend your desk time.
@@
 Hence, while you're reading about how to pick a good advisor, your shortlist may only be of two or three people in the world.
-The project and subject area are going to be strong areas driving your choices moving forwards.
+The project and subject area are going to be strong reasons driving your choices moving forwards.
 
@@
 You should ask them how happy they are with their advisor and whether they are receiving the support that they need to do their projects.
-A good adviser should have an open door so that you can talk to them whenever you need to.
+A good advisor should have an open door so that you can talk to them whenever you need to.
 Thus, once you have talked to the students in their group, you should also make an appointment to talk to them about your project.
 It might not be one meeting, there may be several.
-The more contact you have with them, the more likely you are to be sure that this is the person whose help that you want when conducting research over the next few years.
+The more contact you have with them, the more likely you are to be sure that this is the person whose help and advice that you want when conducting research over the next few years.
 
@@
 
-A good adviser should have a good network.
+A good advisor should have a good network.
 Both during your PhD studies and after them, you will rely primarily on the network of your advisor as a basis to building your own network.
@@
 Having said this, the best academics also tend to be very busy, and that may mean that they do not make good advisors.
-The balance is definitely very tricky, and one of the best ways to make sure that you have enough time for all of your students is to limit the number of students you have.
+The balance is definitely very tricky, and one of the best ways for advisors to make sure that you have enough time for all of your students is to limit the number of students you have.
 Clearly, this works against prospective students.
@@
 There are some advisors who have many problems with many students, and these are best avoided no matter how high their profile is [see @malaga2004meyer;@abbott2004junior; but also see @vences2004meyer for alternative viewpoints].
-There are other advisers who are so busy that they are never able to meet with their students, and pass this task on to somebody else in their group.
+There are other advisors who are so busy that they are never able to meet with their students, and pass this task on to somebody else in their group.
 Again no matter how high profile these people are, if they can't advise their own students they are best avoided.
+Chamberlain [-@chamberlain2016ten] suggests that there are 10 types of academic advisors, implying that 9 of these are rather disfunctional.
 
@@
 After having made lots of statements about how a project should be your own project the following advice might seem a bit strange.
-Clearly the work that you're going to be doing will need to fit into the group of the adviser that you've chosen.
+Clearly the work that you're going to be doing will need to fit into the group of the advisor that you've chosen.
 
-The most obvious point to make here is that a project should be something that you want to do and that you're prepared to dedicate yourself to for at least 3 years.
+The most obvious point to make here is that a project should be something that you want to do and that you're prepared to dedicate yourself to for ~3 years.
 

### 2021-04-18 1.02-useR.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
 It was a steep learning curve, and I struggled.
-Each time the cursor got to the end of the screen I hit the carriage return button.
+As my only previous experience had been with a typewriter, each time the cursor got to the end of the screen I hit the carriage return button.
 And each machine had an odd white device attached called a mouse, and I had no idea what that did.
@@
 If you are not already aware of the prominence that R now has in conducting statistical analyses in biological sciences, then you’ve missed out on a great deal.
-The real point here is not that you can only use R, but that R represents a platform that can be easily reproduced (because it’s free unlike many statistical software packages), and by saving and making the code available, you are making your data analyses reproducible.
+The real point here is not that you can use R, but that R represents a platform on which analyses can be easily repeated, and by saving and making the code available, you are making your data analyses reproducible.
 We will see elsewhere the importance of reproducibility in sciences [@baker20161500], and the critical role this has in transparency and therefore best practice.

### 2021-04-18 2.01-Hyp.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
 Writing well comes with practice, and most PhD students have had little experience with academic writing - even if they have already completed a MSc.
-The same could be said of most early career scientists (even after their PhDs), because writing takes practice and we shouldn't expect that it'll come automatically or immediately.
+The same could be said of most early career scientists (even after their PhDs), because practice is measured in completed pieces, it usually takes many years rather than weeks or months.
+YOu shouldn't expect that scientific writing will come automatically or immediately.
 In order to get this practice you need to get started, and (again) many students find starting to write challenging.
-Here is where this guide will help.
+This guide was written to help you get started.
 
@@
 
-Before you start writing in earnest, it’d be good to read this part of the book and make sure that you are familiar with the ideas therein.
+Before you start writing in earnest, it’d be good to read this part of the book and make sure that you are familiar with the ideas herein.
 Later, you may need to refer back to sections once you’ve started writing.

### 2021-04-18 2.08-citations.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
-## Why do you need to cite? {#citations}
+# Why do you need to cite? {#citations}
 
@@
 
-### Research is built on existing work and ideas
+## Research is built on existing work and ideas
 
@@
 
-### Your citations say a lot about you
+## Your citations say a lot about you
 
@@
 
-### Cite while you write (not afterwards)
+## Cite while you write (not afterwards)
 
@@
 
-### Citation Styles
+## Citation Styles
 
@@
 
-#### Vancouver Style
+### Vancouver Style
 
 Vancouver style dates back to medical journals in the 1970s and refer to conventions to avoid fraud in medical sciences.
-In biological sciences, they are not used as widely as Harvard style, but are chosen as they don’t impede the flow of the text for the reader.
+You are probably familiar with it even if you don't know the name.
+It is the convention that uses small superscript numbers to denote the place of citations in the main text, and the full reference given at the end of the document is provided in order of the citations (and their numbers), instead of alphabetical order.
+In biological sciences, Vancouver style is not used as widely as Harvard style, but is chosen as the superscript numbers tend not to impede the flow of the text for the reader.
 
@@
 
-#### Harvard Style
+### Harvard Style
 
@@
 
-### Moving from Harvard to Vancouver
+## Moving from Harvard to Vancouver
 
-Essentially, every time you would have had the year and date, you will replace this with a superscript denoting the number for that particular citation.
+Moving within styles is always a hassle, especially as formatting the references can take a long time, but moving between these styles is especially tricky as it may require you to re-evaluate the way in which you use the citations.
+
+Moving from Harvard to Vancouver style means that every time you would have had the year and date, you will replace this with a superscript denoting the number for that particular citation.
 However, Vancouver style does not allow for much variation in how you cite.
@@
 
-### Where within a sentence should the citation come?
+## Where within a sentence should the citation come?
 
@@
 
-### What about the order of the citations in a string?
+## What about the order of the citations in a string?
 
@@
 
-### What about citations as taxonomic authorities?
+## What about citations as taxonomic authorities?
 
@@
 
-### Is it possible to mis-cite?
+## Is it possible to mis-cite?
 
@@
 
-### Should I cite without reading the paper?
+## Should I cite without reading the paper?
 
@@
 
-### What should I not cite?
+## What should I not cite?
 
@@
 
-### Do I cite the review or the primary literature?
+## Do I cite the review or the primary literature?
 
@@
 
-### What is primacy and why does it matter?
+## What is primacy and why does it matter?
 
@@
 
-### How many citations is enough?
+## How many citations is enough?
 
@@
 
-### Should I cite myself?
+## Should I cite myself?
 
@@
 
-### Should I cite my friends?
+## Should I cite my friends?
 
@@
 
-### Does the impact factor of the cited article matter?
+## Does the impact factor of the cited article matter?
 

### 2021-04-18 2.09-databases.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
-## Literature databases {#databases}
+# Literature databases {#databases}
 
@@
 Quite simply, you should use one or more of these databases to look for relevant literature to read.
-The first database, and what continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of journals (titles, abstracts, authors and other metadata), but also the citations that each paper used.
+The first database, and what continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of selected journals (titles, abstracts, authors and other metadata), but also the citations that each paper used.
 As we have seen in the last chapter, citations record the way in which people find the literature useful, and are therefore a highly effective way of finding related and pertinent literature in your field.
@@
 
-### Searching the literature using a database
+## Searching the literature using a database
 
@@
 This is called a systematic literature review.
-Using this tactic from the outset of your interactions with literature databases will give you a good idea of what proportion of the literature you are reading, and you might then be able to convert your finding into an objective literature review - making a great first introductory chapter for your thesis.
+Using this tactic from the outset of your interactions with literature databases will give you a good idea of what proportion of the literature you are reading, and you might then be able to convert your finding into an objective literature review - making a great first [introductory chapter](PhDintro.html) for your thesis.
 
-### Keywords
+## Keywords
 
@@
 Any keywords that they share are going to be important for you.
-Most literature databases have the option of using keywords or keywords and words in the title and abstract.
+Make a note of all relevant keywords for your searches, as ultimately this will help you when you need to determine the keywords for [your own chapters](title.html).
+
+Most literature databases have the option of using just keywords, or keywords and words in the title and abstract.
 I would start with the broader search term and see how well they work.
@@
 
-#### Boolean terms or operators
+### Boolean terms or operators
 
@@
 - Use AND when you want both terms included. For example, *performance AND lizards*. In this example, all papers that mention performance AND lizards in the title, abstract or keywords are returned. 
-- Use OR when you want to include both terms.
+- Use OR when you want to include either term.
 For example, *performance AND lizards OR geckos*.
@@
 
-#### Proximity operators and parentheses
+### Proximity operators and parentheses
 
 You can also use the term NEAR to indicate that two terms need to occur within a specified number of words.
-For example, performance NEAR/5 reptile will capture items where these occur within 5 words of each other, but no greater.
+For example, performance NEAR/5 reptile will capture items where these terms occur within 5 words of each other, but no greater.
 This might be useful to exclude some papers that mention performance and reptiles in the abstract, but are not about reptile performance.
+
 Once you start putting together search strings you will need to use brackets to make sure that the correct terms are grouped.
@@
 
-#### Wildcards
+### Wildcards
 
@@
 
-#### Combining searches
+### Combining searches
 
@@
 
-#### Citations
+### Citations
 
@@
 
-What else are literature databases used for?
+#### What else are literature databases used for?
+
 The Science Citation Index used to be published as a single volume that emerged each year and was available in the reference section of the library.
-By the time that I started at university as an undergraduate in the 1980s, it was several volumes large and my first task for my tutor (one Dr Dave Thompson) was to look up all of his citations.
+By the time that I started at university as an undergraduate in the 1980s, it was several volumes large and my first task for my tutor (one Dr Dave J Thompson) was to look up all of his citations (thankfully not too many at that time).
 Within a couple of years, the database was digitised and available on CD, and eventually became accessible through the internet.
+
+Citations provide an idea of how the literature is used.
+This can be used to assess not simply the productivity of a scientist (i.e.
+number of publications), but their usefulness by others (i.e.
+number of citations).
 
@@
 
-### Administrators want metrics
+## Administrators want metrics
 
@@
 
-### Google Scholar, Web of Science or Scopus?
+## Searching by scientist's name
+
+Once you have found a scientist who works within your field and produces influential work that you are keen to read, then you may want to search for all of their other papers to determine whether there is more literature that you have missed.
+The databases make such searches easy, but you will need the (correct) spelling of your scientist's name, and also to be aware that they may have used different iterations of their name throughout their career.
+
+### Overcoming the difficulties of common names
+
+While Dave Thompson could always add his middle initial to specify his work, as time went by there were more and more scientists publishing as D J Thompson.
+This made it more difficult for administrators and scientits alike.
+This has begun to be a bigger global problem, such that the databases have begun to use profiles of particular scientists that they can 'own' and curate.
+Each database has its own way of doing this, and it can become rather a chore so that not many scientists have their own profiles, which makes it a problem particularly for those with common names.
+
+The [ORCID](orcid.html) number has been errected to provide a common platform for [authors to curate](title.html), and now some journals and funding authorities won't allow you to submit or apply without one.
+
+## Google Scholar, Web of Science or Scopus?
 
@@
 
-I have been known to take the odd peak at my Google Scholar profile over the year, and see how it's coming along.
-I rarely check on WoS or Scopus, 'cos it's a bit of a faff getting signed in and doing the search.
+I have been known to take the odd peak at my Google Scholar profile over the year, and see how the citations are accumulating.
+I rarely check on WoS or Scopus, because it's a bit of a faff getting signed in and doing the search.
 Plus it looks so much smaller when one is habituated to seeing those double digits in GS!
@@
 
-It is worth noting that of the three indices, Google Scholar seems to be the easiest to manipulate.
+However, it is worth noting that of the three indices, Google Scholar seems to be the easiest to manipulate.
 Delgado López‐Cózar et al [-@delgadolopez-cozar2014google] show how it is possible to manipulate citations and therefore metrics of individual researchers by depositing false documents into institutional repositories (note that they did this as an experiment, and you may find that restrictions on doing this at your own institution are prohibitive or lax).
@@
 
-### What’s the difference between indices?
+## What’s the difference between indices?
 
-Martín-Martín et al [-@martin-martin2018google] allows you to interpret your GS results as a more inclusive citation index.
+According to Martín-Martín et al [-@martin-martin2018google] The GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus.
+If you deviate from this with a higher score, you can give yourself a pat on the back for having work that's reaching more people in more parts of the world.
+Google Scholar results reflect a more inclusive citation index [@martin-martin2018google].
 While WoS and Scopus aren't exclusively English or journal publications, they are mostly.
@@
 If yes, then your work is having a greater impact elsewhere in the world, and there's nothing wrong with that.
-
-The excerpts from the two tables above show how well GS correlates with both WoS and Scopus in our area (Biological Sciences).
-It also tells you by how much the GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus.
-Again, if you deviate from this with a higher score, you can give yourself a pat on the back for having work that's reaching more people in more parts of the world.

### 2021-04-18 2.10-sortware.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
-## What software should I use to write my PhD? {#software}
+# What software should I use to write my PhD? {#software}
 
@@
 The chances are that today your institution will still have subscriptions to Microsoft Office making it free for you to use.
-But given that changes in software can happen so fast and appear to change with increasing frequency, should you use Microsoft's Word to write your PhD?
+But given that changes in software can happen so fast and appear to change with increasing frequency, should you use Microsoft's Word, or another piece of software to write your PhD?
 
@@
 You will face a lot of challenges during your PhD including learning to write with lots of different kinds of software.
-Each one you learn to use will require a steep learning curve and that requires time.
+Each one you learn to use will require a steep learning curve, and that will require your time.
 If there's any good time to start with a new word processing package it will be before you start your PhD.
-If you do decide to go with a new writing package then my suggestion would be to choose open source software like latex or R markdown.
+If you do decide to go with a new writing package then my suggestion would be to choose open source software like [Rmarkdown @xie2018r].
 More on these later.
@@
 If you feel comfortable with a word processing package with a nice GUI, like Microsoft Word, then go ahead and use it.
-My best suggestion is that you do not save your files in their default *.docx format.
-Instead make sure that all of your files are saved as *.rtf or html.
+My best suggestion is that you do not save your files in their default .docx format.
+Instead make sure that all of your files are saved as .rtf or html.
 This should make little difference to your opening and editing but will mean that all of your texts are forward compatible with other kinds of software.
@@
 That’s right, in addition to loss of data by back up deficiencies [see Part III](data.html), you also have the risk of having corrupt files that will also mean you lose all of the contents.
-Even if you always use Word and never use any other file format, do yourself a favour and save each version as a *.rtf or *.txt file as well.
+Even if you always use Word and never use any other file format, do yourself a favour and save each version as a .rtf or .txt file as well.
 Just like a back up, this means that you won’t have lost all of the text if a file goes corrupt.
@@
 Once you start putting chapters together especially if they include embedded images tables and the like you will find that many word processors start misbehaving.
-This can be especially frustrating in the closing phases of your PhD when you're trying to submit it.
+This can be especially frustrating in the [closing phases of your PhD](part4.html) when you're trying to submit it.
 Especially if there's a tight deadline.
@@
 
-### Alternatives to conventional word processors
+## Alternatives to conventional word processors
 
-Many academics are turning to latex in order to write their manuscripts as this gives a lot of possibility to insert links, formulae, references, images, etc in a very simple way.
+Many academics are turning to latex in order to write their manuscripts as this gives a lot of possibility to insert links, formulae, references, images, etc.
+in a very simple way.
 Once you are ready to print a latex document it can be shared very easily as a PDF.
@@
 Be aware that the problems may come not with what you want but with how your advisor and or co-authors are comfortable with.
-I find that commenting on a PDF takes more time than doing the same on a word document.
+I find that commenting on a PDF takes more time than doing the same on a Word document.
 I don't mind working with latex but I know some advisors are very adverse to having to use this platform.
@@
 For example if your lab is progressive and already experimenting with new and flexible software platforms then I suggest that you learn with them.
-The future of computing is certainly likely to be open-source software and so the sooner that you become more familiar with this the easier you will find it to adapt in future.
+The future of computing is certainly likely to be open-source software and so the sooner that you become more familiar with this, the easier you will find it to adapt in future.
 
 Another consideration is how your word processing package is going to work with referencing and citations.
-Again your lab may already use some software that you want to integrate with your word processor.
+Again, your lab may already use some software that you want to integrate with your word processor.
 
@@
  - Focus on content (not formatting)
- - Numerous output types
+ - Numerous output types (including .docx, .html and .pdf)
  - Independent reference manager (via .bib file) and style files (via .csl file)
@@
 
-Importantly, what you want to try to avoid is conducting a major switch between software right at the end of your PhD. Try making such changes when you have more time, like when you’ve just finished.
+Importantly, what you want to try to avoid is conducting a major switch between software right at the end of your PhD. Try making such changes when you have more time, like after you’ve finished!
 
-### Naming versions of files
+## Naming versions of files
 
@@
 In addition, I’m not good at remembering which chapter is chapter 1 and which is chapter 2.
-Sometimes, towards the end of a thesis, the chapters get switched around anyway, causing even more confusion.
+Sometimes, towards the end of a thesis, the chapters get their orders switched around anyway, causing even more confusion.
 Notice that I don’t like having spaces in filenames.
@@
 While you and your advisor can adopt any naming convention you like, you and they need to instantly be able to distinguish between a new version and one that they have already commented on by the file name.
+No advisor wants to spend time commenting on a chapter and then be told that it was an old version.
 
-### How many versions of each file should you keep?
+## How many versions of each file should you keep?
 
-I would suggest keeping at least one older version of each file to minimise the possibility of losing the content through corrupt files, but you should also be keeping those *.rtf or *.txt files as well, just in case.
+I would suggest keeping at least one older version of each file to minimise the possibility of losing the content through corrupt files, but you should also be keeping those .rtf or .txt files as well, just in case.
 Because of my early experience with file corruption, I started to keep all versions of files.
@@
 Nevertheless, text files take up very little space and it’s not a problem to keep more versions if you want to.
+More on this in the chapter on [data management](data.html).

### 2021-04-18 2.11-paywall.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
-## How to get around the paywall {#paywall}
+# How to get around the paywall {#paywall}
 
@@
 
+### Go to the intitutional repository of the author
+
+Most university libraries curate an institutional repository of pdfs of published final manuscripts accepted for publication, so-called Green Open Access copies.
+These are identical to the printed versions save for the typesetting and proof reading that is done after this accepted copy.
+
 ### Go to an academic social media site
@@
 Academia.edu; ResearchGate.net).
-You will need to join the network and have your own account, but these house a lot of pdfs that you can’t get elsewhere.
+You will need to join the network (using an email address from an academic institution) and have your own account, but these house a lot of pdfs that you can’t get elsewhere.
 

### 2021-04-18 2.13-style.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
-## Writing style {#style}
+# Writing style {#style}
 
@@
 During my PhD I read a lot.
-My study species was the subject of tens of thousands of papers, and I was convinced that I'd find what I'd need somewhere in those musty old reprints and heavy volumes that came from the library shelves.
+My study species was the subject of tens of thousands of papers, and I was convinced that I'd find what I'd needed somewhere in those musty old reprints and heavy volumes that came from the library shelves.
 It was easy to believe that the authors of those papers had set great puzzles for me to try to understand their content.
@@
 > “The goal of good writing is straightforward: to make your reader's job as easy as possible” 
->      Kevin W. Plaxco [-@plaxco2010art].
+> ---  Kevin W. Plaxco [-@plaxco2010art].
 
-This great quote from a paper by Plaxco [-@plaxco2010art] cuts to the heart of why it’s essential to write well.
-As an editor, I once experienced an author who thought the opposite, and for me this was an epiphany.
+This great quote from a paper by Plaxco [-@plaxco2010art], who cuts to the heart of why it’s essential to write well.
+As an editor, I once experienced an author who thought the opposite, and for me this was the epiphany.
 The submitted manuscript was an impenetrable mess.
@@
 I had spent so many hours slogging through dense and dreary papers by well-respected figures.
-The reward when I finally understood what it was that they were trying to communicate felt so great that I believed the puzzle they had set me was what I should set my own readers.
-Happily, my advisor didn’t feel the same way, but for years I continued to believe that a paper should be a decent puzzle for my readers to crack.
+The reward came when I finally understood what it was that they were trying to communicate.
+This felt so great that I believed the puzzle they had set me was what I should set my own readers.
+Happily, my advisor didn’t feel the same way, but for years I continued to believe that a paper should be a decent puzzle for readers to crack.
 This might seem a crazy concept, but it’s not an uncommon problem.
-For a great take on this in Chemistry, see this great article by Murray [-@murray2011skillful].
+For a great take on this in Chemistry, see this article by Murray [-@murray2011skillful].
 
-This has led to a general perception that "academics stink at writing" - as Steven Pinker put it.
-But, I think that Pinker is outdated.
+The preponderance of inpenetrable guff has led to a general perception that "academics stink at writing" - as Steven Pinker put it [@pinker2014why].
+But, I think that Pinker [-@pinker2014why] is outdated.
 Certainly, as Plaxco [-@plaxco2010art] has it, there are a lot of ways that academics can improve their writing, for each other.
@@
 
-Any good writing is an artform, and that includes good academic writing. You can't expect to be the best until you've spent some time at it.
+Any good writing is an artform, and that includes good academic writing.
+You can't expect to be the best until you've spent some time at it.
+If you consider a portfolio of your completed written work as 'your experience', you will get some insight into exactly how small this is, even once you've finished your PhD.
 
@@
 
-### Don’t bamboozle
+## Don’t bamboozle
 
@@
 Simply put, it’s unnecessary to use jargon when you can use plain English in the same amount of space.
-My old tutor at Liverpool University, the late, great Brian Moss, shared the following example of too much jargon [@bump2009large], when plain English would have been much shorter.
+My old advisor at Liverpool University, the late, great Brian Moss, shared the following example of too much jargon by Bump et al [@bump2009large], when plain English would have been much shorter.
 Do yourself a favor and read the abstract of Bump et al [-@bump2009large].
@@
 ```
+
+Whatever your writing style, you need to be conciious of your ultimate aim in writing: to communicate with your reader.
+Making your text as easy to read as possible [as @plaxco2010art put it] isn't easy, but it will be rewarding.
+The reward you will get is in willing readers to the message that you want to convey.
+This is something that will stand you in good sted no matter what career you have after your PhD.

### 2021-04-18 2.14-voice.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
-## When is it possible to retain your own voice when writing? {#voice}
+# When is it possible to retain your own voice when writing? {#voice}
 
 Should one use “I” or “we” when writing a scientific paper?
-Although there was a tradition not to use first-person pronouns when writing scientific articles, this has fallen by the wayside in recent years as the use of I or we makes writing simpler for writers and readers.
+Although there was a tradition not to use first-person pronouns when writing scientific articles, this has fallen by the wayside in recent years, because the use of I or we makes writing simpler for writers and easier to understand for readers.
 Use “I” if you are the sole author (or in your thesis) and “we” for two or more authors.
@@
 It’s also quite common to use this in the methods section.
-The reason why these sections have “I” or “we” is quite obvious, as it allows you to place your own voice on actions and decisions.
+The reason why these sections have “I” or “we” is quite obvious, as it allows you to place your own voice on your actions and decisions.
 After all, it is your aim and hypothesis for your study, so you should own it.
-Using first-person pronouns also allow you to be more concise, and there are some great examples here where you can see what happens when you don’t use them.
+Using first-person pronouns also allows you to be more concise, and there are some great examples given by Kurt Vonnegut [-@vonnegut1980how] where you can see what happens when you don’t use them.
 
@@
 
-### Does writing “I” or “we” help people to understand who you are?
+## Does writing “I” or “we” help people to understand who you are?
 
@@
 
-### The Vonnegut advice
+## The Vonnegut advice
 
@@
 > "Our audience requires us to be sympathetic and patient teachers, ever willing to simplify and clarify - whereas we would rather soar high above the crowd, singing like nightingales."
-> Kurt Vonnegut
+> 
+> --- Kurt Vonnegut [-@vonnegut1980how]
 

### 2021-04-18 2.15-concise.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
-## Writing concisely {#concise}
+# Writing concisely {#concise}
 
-So far, I have put an emphasis on a writing formula in order to get started writing.
+In this book, I have put an emphasis on a [writing formula](formula.html) in order to get started writing.
 Many students have a problem in knowing where and how to get started with what seems to be a daunting prospect: writing a proposal or writing the entire thesis.
-I still feel that getting started quickly and efficiently with some confidence is one of your most important steps.
+I still feel that getting started quickly and efficiently with some confidence is one of your most important first steps.
 But once you've made it through your first draft, you'll begin to find that you need to start refining your writing.
-That is saying what you need to say in less words and space.
+That is saying what you need to say as much in less words and space.
+A bit like constantly rewriting your Tweets to get the same information into 140 characters (yes, it's been increased to 280 but that's not as concise).
 
@@
 The reality is that the thesis is often more wordy, with the result that reviewers and/or editors will ask you to be more concise.
-How can we achieve this?
+How can you achieve this?
 
-Someone else has written a great essay with 10 top rules to help you be concise.
-The essay is Open Access, and there's no need for me to repeat the words of Scott Hotaling [-@hotaling2020simple].
+Scott Hotaling has written a great essay with 10 top rules to help you be concise with your own writing.
+The essay is Open Access, and there's no need for me to repeat Scott's [-@hotaling2020simple] words.
 However, I am going to write some comments of my own on his rules.
-There are some other posts that are also worth a second look along with this one: see here and here.
 
-Here, the rules are Scott's, the comments are mine:
-
-### Rule 1. Take writing seriously
+## Rule 1. Take writing seriously
 
@@
 Certainly, if you did pay more attention in languages in school, you may not be suffering with the rest of us now.
-Personally, I never received tuition in the grammatical construction of English.
+Personally, I never received tuition in the grammatical construction of English (you may know this already).
 Thus I'm left in the dark most of the time about exactly what is wrong.
-Instead, I simply try to rework a sentence into what I know is correct [and I often don't achieve that].
+Instead, I simply try to rework a sentence into what I know is correct (and I often don't achieve that).
 The lesson then is that should you come across an opportunity to learn more about writing, do take it.
+
+Use writing professional emails (to your advisor, co-authors or collaborators) as an opportunity to write concise messages that communicate efficiently and precisely the aim of your message.
+As a PhD student, you should consider yourself a professional and these communications to colleagues will be received with more import if you can write concisely.
+A personal plea from me would be to ask that you don't start your email by enquiring after the health of your recipients.
+The fashion to to this has turned into a surge in recent years (the pandemic notwithstanding), and the reality is that this are simply superfluous and gratuitous words.
+You certainly don't expect them to write back starting with a paragraph about their health.
+Save this kind of correspondence for your friends and relatives, who may well appreciate you enquiring about their health.
+When your advisor, co-authors or collaborators become friends, you can move to a less formal style, but I'd argue that you always want to keep your email communication concise.
+
 Although we talk about writing science here, the grammatical rules of writing (in English) are fairly universal (with some local exceptions), so do practice your writing.
 Maybe write a letter (or long email) instead of making a call sometimes.
-It'll also help you to read (see here).
-If you are a keen reader, then you could do worse than reading some Kurt Vonnegut.
-Scott provided this link to an article I hadn't seen before.
-Go on, give it a read!
+It'll also help you to [read critically](critical.html).
+If you are a keen reader, then you could do worse than reading some Kurt Vonnegut [-@vonnegut1980how].
 
-### Rule 2: Identify and stick to your message
+## Rule 2: Identify and stick to your message
 
@@
 
-### Rule 3: Get to the point
+## Rule 3: Get to the point
 
@@
 
-### Rule 4: Keep your Methods and Results contained
+## Rule 4: Keep your Methods and Results contained
 
@@
 
-### Rule 5: Do not repeat yourself (too often)
+## Rule 5: Do not repeat yourself (too often)
 
@@
 
-Copy and pasting is very easy, and a good way of suddenly producing vast quantities of text.
+Copy and pasting is very easy, and a good way of suddenly producing vast quantities of text (Figure \@ref(fig:plagiarism)).
 But repeated text is quickly recognised by the reader and appears very boring and cumbersome.
@@
 
-Also, please don't cut and paste sections of methods from one part of your thesis to another. Just don't do it.
+Also, please don't cut and paste sections of [methods](methods.html) from one part of your thesis to another. Just don't do it.
 
-People also have a tendency to build the abstract by copying and pasting text from the main sections.
+People also have a tendency to build the [abstract](abstract.html) by copying and pasting text from the main sections.
 Don't do it.
 The reader will quickly see the repetitions and become bored.
-Similarly, the conclusion/summary section is also often copied from lines above.
+Similarly, the [conclusion/summary](discussion.html) section is also often copied from lines above.
 If it's not worth writing again, then it wasn't worth writing before.
@@
 
-### Rule 6: Avoid unnecessary or inefficient “lead‐ins”
+## Rule 6: Avoid unnecessary or inefficient “lead‐ins”
 
@@
 Having your advisor help or someone who has edited a lot of text.
-Scott's got some great examples, and so I'll let you peruse them here.
+Scott's got some great examples [@hotaling2020simple].
 
-### Rule 7: Use first‐person, active voice
+## Rule 7: Use first‐person, active voice
 
@@
 In fact, it can be the easiest way to avoid the passive voice.
-I've written more on this elsewhere...
-see here.
+There are other reasons why [using the first person](voice.html) is helpful.
 
-### Rule 8: Remove unnecessary words
+## Rule 8: Remove unnecessary words
 
@@
 
-### Rule 9: Simplify your language
+## Rule 9: Simplify your language
 
-This is always a good idea. Making three short sentences instead of one very long one is much better. It also helps you avoid complicated grammatical clauses.
+This is always a good idea (see Figure \@ref(fig:moss)). Making three short sentences instead of one very long one is much better. It also helps you avoid complicated grammatical clauses.
 
-### Rule 10: Seek and embrace feedback
+## Rule 10: Seek and embrace feedback
 
@@
 Certain students become very hard work when they don't even explain why they refuse to change something.
+Reading the work of these students quickly becomes a chore.
 You should have noted by now that it's all about being flexible, and being stubborn about anything you've written isn't going to work.
-Reading the work of these students quickly becomes a chore.
 
-### Rule 11: Read it yourself
+## Rule 11: Read it yourself
 
@@
 Perhaps it should have been inserted higher up the list.
-As I've said elsewhere (here for example), it would help you to have your work read through by one of your colleagues before you give it to your advisor.
-But above all, you must be prepared to read __all of your work__ yourself.
+It would help you to have your work read through by one of your colleagues before you [give it to your advisor](goodenough.html).
+But above all, you must be prepared to read __all of your work__ yourself, and be prepared to work on it when you read.
+Edit and improve it on every read.
 Not just the first time you write it, but for **every** version.
@@
 
-Being concise is a great way to write well, and the abstract is a case in point.
-
-Please don't forget to read Scott's 10 rules as written by Scott [@hotaling2020simple].
+Being concise is a great way to write well, and the [abstract](abstract.html) is a case in point. Please don't forget to read Scott's 10 rules as written by Scott [@hotaling2020simple].

### 2021-04-18 2.16-foreign.Rmd
Add files via upload
# some changes after reading Parts I and II

@@
-## Writing a PhD if English is not your first language {#foreign}
+# Writing a PhD if English is not your first language {#foreign}
 
-Reading this book you will learn of the many biases in biological sciences and how you need to be aware of them in order to properly interpret your work.
+Reading this book you will learn of some biases in biological sciences and how you need to be aware of them in order to properly interpret your work.
 Be aware also that biological sciences is biased in where studies take place, and who publishes them [@culumber2019widespread; @nunez2021monolingual].
@@
 
-### Message from an editor
+## Message from an editor
 
@@
 
-### Must a ‘native English speaker’ check your work?
+## Must a ‘native English speaker’ check your work?
 
@@
 
-### So if not a native English speaker, who should do it?
+## So if not a native English speaker, who should do it?
 
@@
 
-### But surely my native tongue can be used for something?
+## But surely my native tongue can be used for something?
 

### 2021-04-19 2.25-formula.Rmd
Add files via upload
# updated references with DOI and discussion table

@@
-## Why use a formula to structure each chapter or paper? {#formula}
+# Why use a formula to structure each chapter or paper? {#formula}
 
@@
 
-### So what is the formula?
+## So what is the formula?
 
@@
 However, others have much better diagrams and here I’ve adapted an idea from Brian McGill’s blog [@mcgill20165].
-But you can also find other useful material in a paper by Sheela Turbek and colleagues [-@turbek2016scientific], and in a paper by Kevin Plaxco [-@plaxco2010art].
+But you can also find other useful material in other papers [-@turbek2016scientific; @plaxco2010art @tomaska2007teaching].
 If you are of the Twitter persuasion, you get lots of useful tips and links from WriteThatPhD and Write4Research.
@@
 
-#### My take on how to write a formula paper
+### My take on how to write a formula paper
 
@@
 
-#### Start with an outline
+### Start with an outline
 __So now that you've got the idea that writing to a formula is a good idea, what next? __
@@
 
-#### Next flesh out the outline with citations
+### Next flesh out the outline with citations
 
@@
 
-#### Lastly it’s time to write the text
+### Lastly it’s time to write the text
 

### 2021-04-19 3.06-Discussion.Rmd
Add files via upload
# updated references with DOI and discussion table

@@
-## The Discussion {#discussion}
+# The Discussion {#discussion}
 
@@
 
-Again, I'd like to repeat that the convention described above is a formula that is intended to help you getting started with writing.
+I'd like to repeat that the convention described above is a formula that is intended to help you getting started with writing.
 It is not the only way to write, and if you feel confident that another way is better then I encourage you to explore this with the support of your advisor.
@@
 
-### First paragraph of the discussion
+## First paragraph of the discussion
 Your discussion begins by you responding to your hypothesis, clearly stating whether or not it has been accepted, and putting this into the wider context of the study (i.e.
@@
 
-### To sub-section or not to sub-section the discussion?
+## To sub-section or not to sub-section the discussion?
 My preference is to plan the discussion before you write it, just as you did for the [introduction](introduction.html) using a bullet pointed list.
@@
 
+In order to provide a critical evaluation of your research findings, you really need to present a reasoned argument [@jenicek2006how].
+In some respects, when writing the discussion your walk a tightrope between inferring the meaning of your results for a wider audience, and making claims that are too broad (Table \@ref(tab:discuss)).
+Overreaching in the discussion (and consequently [abstract](abstract.html)) is a common problem that examiners and reviewers regularly ask authors pull back on their claims.
+
 When providing different sides of an [argument](argument.html), try to use your [results](results.html) to conclude that one side is supported more than the other.
@@
 
-### Caveats and limitations
+Table: (\#tab:discuss) **Some discussion dos and don'ts (adapted from Jenicek 2006).** Balance your discussion so that you provide a reasoned argument that reflects what you found.
+
+| **Do** | **Don't** |
+|----|----|
+| Provide a summary of your major findings | Use statements that are too broad |
+| Interpret your results with respect to mechanisms | Over-represent your results |
+| Give problems with methods and techniques | Provide unrealistic solutions to deficiencies |
+| Compare similarities and contrasts with other studies | Ignore conflicting literature |
+| Provide practical implications of your results | Be unrealistic about what could be done |
+| Give directions for future research | Simply state what you have done elsewhere in your thesis |
+| Restrain your conclusions to evidence from your data | Make unsubstantiated claims that overreach your evidence |
+| Provide insight into unresolved questions, controversies and biases | [_Reductio ad absurdum_](argument.html)  |
+| Explain the study design's limitations and uncertainties | Inflate the importance of your findings |
+| Highlight the strengths of your study | Forget to mention the limitations |
+| Provide alternative interpretations and hypotheses | Go into tangential issues |
+| Give unexpected results and anomalies | Avoid being critical of how your study could have produced unexpected results |
+| Mention literature that contradicts your findings | Use your conclusions as a 'bully pulpit' |
+| Provide arguments that are logical and coherent | Erect a ['straw man' or 'false dichotomy'](argument.html) |
+| Make conclusions based on your logic | Make emotional appeals to the reader |
+| Make sure that you've discussed all results | Suddenly introduce new data or results not previously presented |
+| Boast about your findings | Apologise or undermine your own work |
+| Be concise | Waffle on |
+| Take a look at target journal requirements and suggestions for a  discussion | Feel that journal advice is not relevant to you |
+
+## Caveats and limitations
 An important aspect of the discussion is to consider how the interpretation of your [results](results.html) may be incorrect.
@@
 
-### Should you speculate in the discussion? {#speculation}
+## Should you speculate in the discussion? {#speculation}
 Reviewers will often be unhappy with speculation in the discussion section.
@@
 
-### Don’t beat up on others
+## Don’t beat up on others
 Your [results](results.html) may show that other researchers were wrong with their interpretation or findings.
@@
 
+Things brings to mind Freud's narcissism of minor differences, in which he pointed out that bitter fights could start between groups when they in fact agree on 90% on issues.
+So it is likely to be true that someone you might disagree with one one point, is likely to be allied with you on nearly all of their other views and opinions.
+Making and maintaining allegiances is more likely to get you in a position where you can constructively discuss any minor disagreements.
+Consider this before you are tempted to launch a negative sounding social media post.
+
 Generally, such comments won’t get through the peer review process, and remember that you might be insulting the examiner of your thesis (or the reviewer of your paper) – which is not likely to go down well!
 
-### Where next?
+## Where next?
 The ‘where next’ aspect of your discussion is important as it may provide the reader with ideas for their own work.
@@
 Providing new lines of research may also allow you to speculate about what you consider to be the most important angle of this topic now that you have presented your [results](results.html).
-This should be justifyable and not gratuitous.
+This should be justifiable and not gratuitous.
 There's no point in suggesting what other studies could be done just for the sake of it, or just because you happen to have already done them in your thesis.
-But if there are legitate parts of your thesis that link to your discussion of your chapter (and this is quite likely), then you should definitely point to them.
+But if there are legitimate parts of your thesis that link to your discussion of your chapter (and this is quite likely), then you should definitely point to them.
 
-### Last paragraph
+## Last paragraph
 The last paragraph of the discussion is your take home message.
@@
 
-#### Never repeat text
+### Never repeat text
 Please remember that while this might sound similar to your first paragraph, it is not the same.
@@
 As always, there are a number of other places to look for more advice to write your discussion, and I’d encourage you to read as widely as possible.
-For example, try Hess [-@hess2004how] and Sanli [-@sanli2013how].
+For example, try Hess [-@hess2004how], Jenicek [-@jenicek2006how] and Şanlı [-@sanli2013how].

### 2021-04-19 3.08-References.Rmd
Add files via upload
# updated references with DOI and discussion table

@@
-## The References {#references}
+# The References {#references}
 
@@
 It took very little time to learn how to use, it was painless to enter (most) of the references into the database, and it is free and works with [Bookdown](bookdown.org) and Google Docs.
+
+Earlier in this book, I provided a guide to the differences between Vancouver and
 
@@
 
-### Digital Object Identifiers (DOIs) {#DOI}
-A [DOI](https://www.doi.org/) is an [ISO](https://www.iso.org/) unique character string to identify physical, digital or abstract objects.
+## Digital Object Identifiers (DOIs) {#DOI}
+A [DOI](https://www.doi.org/) is an international standard ([ISO](https://www.iso.org/)) unique character string to identify physical, digital or abstract objects.
 Their beauty is that they are persistent over time, so once issued they will always be a way to reach a particular object.
@@
 
-#### DOI Tools
+### DOI Tools
 
@@
 
-There's also an equally useful database that provides BibTex for DOIs that you enter: [doi2bib.org](https://doi2bib.org/)
+There's an equally useful database that provides BibTex for DOIs that you enter: [doi2bib.org](https://doi2bib.org/)

### 2021-04-19 2.12-taxonomy.Rmd
Add files via upload
# Plant taxonomy update on advice of Al Ellis

@@
-## Scientific names and taxonomic authorities {#taxonomy}
+# Scientific names and taxonomic authorities {#taxonomy}
 
@@
 
-### Scientific names
+## Scientific names
 
@@
 
-#### Consider this example:
+### Consider this example:
 
@@
 
-### Different codes for the nomenclature of different taxa
+## Different codes for the nomenclature of different taxa
 
@@
 
-#### Peculiarities of plant taxonomy
+### Peculiarities of plant taxonomy
 
 Plant taxonomy is considered the taxonomic mother discipline as the work of Linneaus in 1753 was the first botanical work that consistently used Linneus' binomial (two names: Genus species) nomeclatural system [@linnaeus1753species] which pre-dated the more inclusive (and better known) landmark publication in 1758 [@linnaeus1758systema].
+Until the Botanical Code was revised in 2012, all new botanical diagnoses had to be written in Latin.
 In addition to the Latin binomial name (Genus and species), plants (and some other taxa) often require further descriptors or **ranks** that denote subspecies, varieties, subvareties, etc.
@@
 
+Unlike animal authorities, for plants if the combination of Genus and species has changed, both the authority for the original genus placement, and the author responsible for the new combination are given after the species name (with the former in brackets).
+In addition, for plants no year is given, and the name of the authority is usually abbreviated.
 Because Linneaus named so many plant species is his landmark publications, instead of writing out the full taxonomic authority after the species names, botanists abbreviate this to "L.".
@@
 
-Some plants exist in associations (such as lichens which are an association between an algae or cyanobacteria and one or more fungi.
+Algae sometimes exist in association with other taxa (such as lichens which are an association between an algae or cyanobacteria and one or more fungi.
 For these associations there is a specific code for their nomenclature: International Code of Phytosociological Nomenclature [@theurillat2021international].
@@
 
-### Common names
+## Common names
 
@@
 
-### Should common names have capitals?
+## Should common names have capitals?
 
@@
 
-#### View 1.
+### View 1.
 
@@
 
-#### View 2.
+### View 2.
 

### 2021-04-24 00-author.Rmd
Add files via upload
# After some small edits over the week

@@
 output:
+  html_document: default
   pdf_document: default
-  html_document: default
   word_document: default
@@
 
+```{r fig.align='right', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/About-Me'}
+knitr::include_graphics('figures/john.jpg', dpi = NA)
+```
+
 # Do you have something to contribute? {- #contribute}
 
-This book is written in bookdown [@xie2016bookdown] specifically to make it a 'live project' that will be open to anyone who wants to contribute, improve, or use as the basis for your own book in a different language, or for a different region on the world.
+This book is written in bookdown [@xie2016bookdown] specifically to make it a 'live project' that will be open to anyone who wants to contribute, improve, or use as the basis for your own book.
+One of the amazing potentials for bookdown books is that all the files for this book are hosted on [Github](https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences/tree/main).
+You have the opportunity to fork this repository and write your own version for a different discipline, a different language or for a different region of the world.
 It would be impossible to cover every aspect of writing a PhD in Biological Sciences, and it may be that I have missed out ones that are very important to you.
-Equally, what is currently written will become obsolete very quickly as new initiatives are begun, and old problems are resolved.
+Equally, parts of what is currently written will become obsolete very quickly as new initiatives are begun, and old problems are resolved.
 For this reason, this guide needs to be a 'living document', and anyone who wants to provide feedback or contribute new sections is more than welcome.

### 2021-04-24 2.08-citations.Rmd
Add files via upload
# After some small edits over the week

@@
 
-^1^Berger L, Speare R, Hyatt A.
-_Chytrid fungi and amphibian declines: overview, implications and future directions.
-Declines and disappearances of Australian frogs_ Environment Australia, Canberra.
-1999;1999:23–33.
-
-^2^Collins JP, Crump ML, Lovejoy III TE. _Extinction in our times: global amphibian decline_. Oxford University Press; 2009.
-
-^3^Daszak P, Cunningham AA, Hyatt AD. Infectious disease and amphibian population declines. _Diversity and Distributions_. 2003;9(2):141–50.
-
-^4^Dufresnes C, Dubey S, Ghali K, Canestrelli D, Perrin N.
-Introgressive hybridization of threatened European tree frogs (*Hyla arborea*) by introduced _H.
-intermedia_ in Western Switzerland.
-_Conservation Genetics_.
-2015;16(6):1507–13.
-
-^5^Kupferberg SJ. Bullfrog (*Rana catesbeiana*) invasion of a California river: the role of larval competition. _Ecology_. 1997;78(6):1736–51.
-
-^6^La Marca E, Lips KR, Lotters S, Puschendorf R, Ibanez R, Rueda-Almonacid JV, et al.
-Catastrophic population declines and extinctions in Neotropical harlequin frogs (Bufonidae: *Atelopus*) _Biotropica: The Journal of Biology and Conservation_.
-2005;37(2):190–201.
-
-^7^Martel A, Spitzen-van der Sluijs A, Blooi M, Bert W, Ducatelle R, Fisher MC, et al.
-*Batrachochytrium salamandrivorans* sp.
-nov.
-causes lethal chytridiomycosis in amphibians.
-_Proceedings of the National Academy of Sciences_.
-2013;110(38):15325–9.
-
-^8^Pimm SL, Jenkins CN, Abell R, Brooks TM, Gittleman JL, Joppa LN, et al. The biodiversity of species and their rates of extinction, distribution, and protection. _Science_. 2014;344(6187):1246752.
-
-^9^Wake DB, Vredenburg VT. Are we in the midst of the sixth mass extinction? A view from the world of amphibians. _Proceedings of the National Academy of Sciences_. 2008;105(Supplement 1):11466–73.
+The corresponding references in [Vancouver](vancouver.html) format are in the section on [references](references.html).
 
@@
 
-Berger, L., Speare, R., Hyatt, A., 1999.
-_Chytrid fungi and amphibian declines: overview, implications and future directions.
-Declines and disappearances of Australian frogs_.
-Environment Australia, Canberra 1999, 23–33.
-
-Collins, J.P., Crump, M.L., Lovejoy III, T.E., 2009. _Extinction in our times: global amphibian decline_. Oxford University Press.
-
-Daszak, P., Cunningham, A.A., Hyatt, A.D., 2003. Infectious disease and amphibian population declines. _Diversity and Distributions_ 9, 141–150.
-
-Dufresnes, C., Dubey, S., Ghali, K., Canestrelli, D., Perrin, N., 2015.
-Introgressive hybridization of threatened European tree frogs (*Hyla arborea*) by introduced _H.
-intermedia_ in Western Switzerland.
-_Conservation Genetics_ 16, 1507–1513.
-
-Kupferberg, S.J., 1997. Bullfrog (*Rana catesbeiana*) invasion of a California river: the role of larval competition. _Ecology_ 78, 1736–1751.
-
-La Marca, E., Lips, K.R., Lotters, S., Puschendorf, R., Ibanez, R., Rueda-Almonacid, J.V., Schulte, R., Marty, C., Castro, F., Manzanilla-Puppo, J., others, 2005.
-Catastrophic population declines and extinctions in Neotropical harlequin frogs (Bufonidae: *Atelopus*) _Biotropica: The Journal of Biology and Conservation_ 37, 190–201.
-
-Martel, A., Spitzen-van der Sluijs, A., Blooi, M., Bert, W., Ducatelle, R., Fisher, M.C., Woeltjes, A., Bosman, W., Chiers, K., Bossuyt, F., others, 2013.
-*Batrachochytrium salamandrivorans* sp.
-nov.
-causes lethal chytridiomycosis in amphibians.
-_Proceedings of the National Academy of Sciences_ 110, 15325–15329.
-
-Pimm, S.L., Jenkins, C.N., Abell, R., Brooks, T.M., Gittleman, J.L., Joppa, L.N., Raven, P.H., Roberts, C.M., Sexton, J.O., 2014.
-The biodiversity of species and their rates of extinction, distribution, and protection.
-_Science_ 344, 1246752
-
-Wake, D.B., Vredenburg, V.T., 2008. Are we in the midst of the sixth mass extinction? A view from the world of amphibians. _Proceedings of the National Academy of Sciences_ 105, 11466–11473.
+You can find the corresponding references in [Harvard](harvard.html) format in the [references](references.html) section.
 
@@
 
-Another common mistake is to forget which paper has which information. You can try to make sure that you don’t do this by taking better notes or a more accurate plan.
+Another common mistake is to forget which paper has which information.
+You can try to make sure that you don’t do this by taking better notes or a more accurate plan.
+And do the citations before or as you write, not afterwards!
 

### 2021-04-24 2.09-databases.Rmd
Add files via upload
# After some small edits over the week

@@
 
-### Citations
+## Moving items into your reference manager
+
+An invaluable way in which databases have become extremely useful is the ability to move items that you have found during your search directly into your [reference manager](referencesoftware.html).
+Beware that these regularly come with errors, and that the amount of errors is related to your database source (Scopus and WoS are much better than Google Scholar).
+This means that after importing the refrences, you will need to go back to your reference manager and edit them or they will come out incorrectly when you use them.
+Of particular note to biologists is the need to have all species names in [italics](taxonomy.html), and you'll need to do this yourself.
+
+## Citations
 
@@
 
-#### What else are literature databases used for?
+### What else are literature databases used for?
 

### 2021-04-24 2.10-referencesortware.Rmd
Add files via upload
# After some small edits over the week

# Reference managers {#referencesoftware}

Because published papers play such a fundamental role in your [reading and comprehension](critical.html) of a subject, and [citations](citations.html) are pivotal in chapter or paper writing, putting these elements together in a good reference manager makes a lot of sense. This is a piece of software in which you can store and view all of your papers (in pdf format or via links), make notes on them, and use them to make citations and build bibliographies. Unsurprisingly, they become something that you may well use a lot during the course of your PhD, and it is worth thinking carefully about which one you want to use.

There are quite a lot of different software packages out there to help you do this (e.g. [Endnote](https://endnote.com/), [Mendelay](https://www.mendeley.com/guides/desktop), [Zotero](https://www.zotero.org/) - see more about these [here](https://subjectguides.library.american.edu/c.php?g=479020&p=3323781)). These are a mixed blessing. Firstly, most of them are not free, so unless your university subscribes to their providers you would have to pay for the privilege of using them. Next, all of the ones that I've ever used are very demanding and do not necessarily work well with your word processor of choice. Given all the time I have invested in reference formatting software, I think I might have been better off just copying and pasting references I have at the end of a document and then formatting them by hand. This tends to be how I do it now. Having said this, I used Zotero as the referencing software for this book. It took very little time to learn how to use, it was painless to enter (most) of the references into the database, and it is free and works with [Bookdown](bookdown.org) and Google Docs.

My advice here is (like software for [writing your PhD](software.html)) to pick the format that your references are going to be stored in carefully. For example, the BibTex format (see below) is a generic format that many of the software packages are capable of importing and exporting. Using a generic referencing format means that you will (in future) be able to change packages with minimum loss. Note that there is a difference between being able to output in BibTex and storage as BibTex files. In the case of the former, you will require a working piece of software in order to output from that format. If your software stores references as BibTex files, you are always able to take this file and use it elsewhere.

**An example of the way in which the generic BibTex format stores references** 
```
@article{measey2017counting,
	title = {Counting chirps: acoustic monitoring of cryptic frogs},
	volume = {54},
	copyright = {© 2016 The Authors. Journal of Applied Ecology published by John Wiley \& Sons Ltd on behalf of British Ecological Society.},
	issn = {1365-2664},
	shorttitle = {Counting chirps},
	url = {https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2664.12810},
	doi = {https://doi.org/10.1111/1365-2664.12810},
	abstract = {Global amphibian declines have resulted in a vital need for monitoring programmes that follow population trends. Monitoring using advertisement calls is ideal as choruses are undisturbed during data collection. However, methods currently employed by managers frequently rely on trained observers and/or do not provide density data on which to base trends. This study explores the utility of monitoring using acoustic spatially explicit capture–recapture (aSCR) with time of arrival (ToA) and signal strength (SS) as a quantitative monitoring technique to measure call density of a threatened but visually cryptic anuran, the Cape peninsula moss frog Arthroleptella lightfooti. The relationships between temporal and climatic variables (date, rainfall, temperature) and A. lightfooti call density at three study sites on the Cape peninsula, South Africa, were examined. Acoustic data, collected from an array of six microphones over 4 months during the winter breeding season, provided a time series of call density estimates. Model selection indicated that call density was primarily associated with seasonality fitted as a quadratic function. Call density peaked mid-breeding season. At the main study site, the lowest recorded mean call density (0·160 calls m−2 min−1) occurred in May and reached its peak mid-July (1·259 calls m−2 min−1). The sites differed in call density, but also the effective sampling area. Synthesis and applications. The monitoring technique, acoustic spatially explicit capture–recapture (aSCR), quantitatively estimates call density of calling animals without disturbing them or their environment. In addition, time of arrival (ToA) and signal strength (SS) data significantly add to the accuracy of call localization, which in turn increases precision of call density estimates without the need for specialist field staff. This technique appears ideally suited to aid the monitoring of visually cryptic, acoustically active species.},
	language = {en},
	number = {3},
	urldate = {2021-03-28},
	journal = {Journal of Applied Ecology},
	author = {Measey, G. John and Stevenson, Ben C. and Scott, Tanya and Altwegg, Res and Borchers, David L.},
	year = {2017},
	note = {\_eprint: https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/1365-2664.12810},
	keywords = {acoustic array, acoustic spatially explicit capture–recapture, anurans, call density, non-invasive sampling, population monitoring, sensor networks, signal strength, time of arrival, triangulation},
	pages = {894--902},
	file = {Full Text PDF:C\:\\Users\\Zotero\\storage\\****\\Measey et al. - 2017 - Counting chirps acoustic monitoring of cryptic fr.pdf:application/pdf;Snapshot:C\:\\Users\\****\\1365-2664.html:text/html},
}
```
## Other points to consider

When choosing a reference manager, here are some points that you may want to consider:

 - Does it integrate with your [word processing package](software.html)?
 - Does your university have a licence?
 - Do other members of your lab use this software?
   + is there a communal repository?
 - Can you use it to read files and make notes on them?
 - Does it output into referencing formats that will be useful to you?
   + Or will it allow you to make a custom output format?
 - How much time will it take you to learn to use it?
 - Is there good support available?

### 2021-04-24 2.10a-sortware.Rmd
Add files via upload
# After some small edits over the week

--- 2021-04-18 2.10-sortware.Rmd
@@
 
+Another important consideration is whether or not your chosen word processing package will allow seemless integration with your [reference manager](referencesoftware.html).
+Your lab may have a communal repository of references, and so this may then constrain what you are comfortable using.
+
 If you feel comfortable with a word processing package with a nice GUI, like Microsoft Word, then go ahead and use it.
-My best suggestion is that you do not save your files in their default .docx format.
+These certainly have advantages like being able to get a synonym at the click of a mouse (Figure \@ref(fig:synonym)).
+My best suggestion is that you do not save your files in their default ".docx" format.
 Instead make sure that all of your files are saved as .rtf or html.
@@
 And yes, I do know people who have multiple backup versions of corrupt files: it’s happened to me.
+
+Keeping files in a generic (rather than software manufacturer's) format will allow easy sharing of these files, and access to them in future.
+It might seem inconceivable now that anyone would not be able to read a .docx file, but in 10 years, you may be so used to the latest format that after another 5 years it becomes very difficult to retrieve information from a .docx file without paying someone to retrieve your data.
 

### 2021-04-24 2.22-transparency.Rmd
Add files via upload
# After some small edits over the week

@@
-## Starting out transparent {#transparency}
+# Starting out transparent {#transparency}
 
@@
 There are much better and more comprehensive guides and online resources so I encourage you to look through the literature cited here.
-Although I advocate hard for the COS, please be aware that this is not the only set of guidelines (see also [recent moves by the Public Library of Science](https://plos.org/open-science/preregistration/)).
+Although I advocate hard for the COS, please be aware that this is not the only set of guidelines (see also [recent moves by the Public Library of Science](https://plos.org/open-science/preregistration/)) and [Peer Community In](https://peercommunityin.org/).
 
-### Preregistration
+## Preregistration
 
@@
 
-#### Does this mean that you can’t perform any data exploration?
+### Does this mean that you can’t perform any data exploration?
 
@@
 
-#### Will preregistration of research eliminate the bias from science?
+### Will preregistration of research eliminate the bias from science?
 
 It is probably too early to tell [@rubin2020does], but it is certainly a good place to start [@nosek2018preregistration].
-The more researchers know and subscribe to [transparency](transparency2.html) in their research, the more it will shift the culture in science for the better [@forstmeier2017detecting].However, there are more biases existing in science, and the system needs buy-in from stakeholders to promote equity by eliminaion of bias.
+The more researchers know and subscribe to [transparency](transparency2.html) in their research, the more it will shift the culture in science for the better [@forstmeier2017detecting].However, there are more biases existing in science, and the system needs buy-in from stakeholders to promote equity by elimination of bias.
 
-#### What Platform should you use?
+A new initiative from [Peer Community In](https://peercommunityin.org/) is the possibility of submitting to their [Registered Reports](https://rr.peercommunityin.org/), which goes much further towards removing the confirmation bias.
+The Registered Report (RR) is in effect the registration of a proposal (i.e.
+preregistration) with review.
+If the RR is approved by reviewers then the study is, in principle, given the green light for publication whether or not the hypothesis posed is accepted or rejected.
+I say "in principle" because those same reviewers are shown the ms again once the results are in.
+They need to check that the methods proposed were followed, the analyses were conducted in the same way they were proposed, and that the conclusions are justified by the results.
+In terms of figure \@ref(fig:preregistration), Peer Community In are offering to organise the two yellow review arrows.
+In addition, there are a bunch of journals that have already signed up to accept RRs that are signed off after completion (notable among these is [PeerJ](www.peerj.com)).
+To me, this represents an important step in the right direction toward transparency, and the elimination of confirmation bias.
+What would be great to see is the number of conventional journals sign up with RRs based on the quality of the study design and execution, and the concomitant abandonment of Impact Factor as a driving force in publishing.
+
+### What Platform should you use?
 
@@
 
-### Transparency as you move forwards
+## Transparency as you move forwards
 

### 2021-04-24 3.08-References.Rmd
Add files via upload
# After some small edits over the week

@@
 # The References {#references}
-
-Putting together the reference section of your chapter or paper can take a very long time.
-There are quite a lot of different software packages out there to help you do this (e.g.
-[Endnote](https://endnote.com/), [Mendelay](https://www.mendeley.com/guides/desktop), [Zotero](https://www.zotero.org/) - see more about these [here](https://subjectguides.library.american.edu/c.php?g=479020&p=3323781)).
-These are a mixed blessing.
-Firstly, most of them are not free, so unless your university subscribes to their providers you would have to pay for the privilege of using them.
-Next, all of the ones that I've ever used are very demanding and do not necessarily work well with your word processor of choice.
-Given all the time I have invested in reference formatting software, I think I might have been better off just copying and pasting references I have at the end of a document and then formatting them by hand.
-This tends to be how I do it now.
-Having said this, I used Zotero as the referencing software for this book.
-It took very little time to learn how to use, it was painless to enter (most) of the references into the database, and it is free and works with [Bookdown](bookdown.org) and Google Docs.
-
-Earlier in this book, I provided a guide to the differences between Vancouver and
 
@@
 
-Lots of journals are very fussy about the way that citations are given in their publications.
-I think it's really very cheeky especially as many of them also force you to pay for the privilege of doing all the work.
-There are some exceptions to this (e.g.
-[PeerJ](www.peerj.com)).
+[Earlier in this book](citations.html), I provided a guide to the differences between Vancouver and Harvard referencing styles.
+Here I provide the output for the references that were given in that section to demonstrate the way in which these different referencing styles take up very different amounts of space.
+In that section, I provided some quotes from a published paper that had 9 citations.
+Below, you can see what these 9 citations look like in the two different generic outputs: Vancouver and Harvard.
+
+## Vancouver style {#vancouver}
+The following references follow the text written in [Vancouver style](citations.html) in the section on citations.
+Note that the order of citations here corresponds to the order in which they are cited.
+
+>^1^Wake DB, Vredenburg VT. Are we in the midst of the sixth mass extinction? A view from the world of amphibians. _Proceedings of the National Academy of Sciences_. 2008;105(Supplement 1):11466–73.
+>
+>^2^Collins JP, Crump ML, Lovejoy III TE. _Extinction in our times: global amphibian decline_. Oxford University Press; 2009.
+>
+>^3^Pimm SL, Jenkins CN, Abell R, Brooks TM, Gittleman JL, Joppa LN, et al. The biodiversity of species and their rates of extinction, distribution, and protection. _Science_. 2014;344(6187):1246752.
+>
+>^4^Kupferberg SJ. Bullfrog (*Rana catesbeiana*) invasion of a California river: the role of larval competition. _Ecology_. 1997;78(6):1736–51.
+>
+>^5^Dufresnes C, Dubey S, Ghali K, Canestrelli D, Perrin N.
+Introgressive hybridization of threatened European tree frogs (*Hyla arborea*) by introduced _H.
+intermedia_ in Western Switzerland.
+_Conservation Genetics_.
+2015;16(6):1507–13.
+>
+>^6^Berger L, Speare R, Hyatt A.
+_Chytrid fungi and amphibian declines: overview, implications and future directions.
+Declines and disappearances of Australian frogs_ Environment Australia, Canberra.
+1999;1999:23–33.
+>
+>^7^Daszak P, Cunningham AA, Hyatt AD. Infectious disease and amphibian population declines. _Diversity and Distributions_. 2003;9(2):141–50.
+>
+>^8^La Marca E, Lips KR, Lotters S, Puschendorf R, Ibanez R, Rueda-Almonacid JV, et al.
+Catastrophic population declines and extinctions in Neotropical harlequin frogs (Bufonidae: *Atelopus*) _Biotropica: The Journal of Biology and Conservation_.
+2005;37(2):190–201.
+>
+>^9^Martel A, Spitzen-van der Sluijs A, Blooi M, Bert W, Ducatelle R, Fisher MC, et al.
+*Batrachochytrium salamandrivorans* sp.
+nov.
+causes lethal chytridiomycosis in amphibians.
+_Proceedings of the National Academy of Sciences_.
+2013;110(38):15325–9.
+>
+>
+
+## Harvard style {#harvard}
+The references below are exactly the same as those above, but follow the [Harvard](citations.html) referencing format.
+The order of these references is alphabetical.
+If a paper has the same authors the oldest one usually comes before subsequent papers.
+
+>Berger, L., Speare, R., Hyatt, A., 1999.
+_Chytrid fungi and amphibian declines: overview, implications and future directions.
+Declines and disappearances of Australian frogs_.
+Environment Australia, Canberra 1999, 23–33.
+>
+>Collins, J.P., Crump, M.L., Lovejoy III, T.E., 2009. _Extinction in our times: global amphibian decline_. Oxford University Press.
+>
+>Daszak, P., Cunningham, A.A., Hyatt, A.D., 2003. Infectious disease and amphibian population declines. _Diversity and Distributions_ 9, 141–150.
+>
+>Dufresnes, C., Dubey, S., Ghali, K., Canestrelli, D., Perrin, N., 2015.
+Introgressive hybridization of threatened European tree frogs (*Hyla arborea*) by introduced _H.
+intermedia_ in Western Switzerland.
+_Conservation Genetics_ 16, 1507–1513.
+>
+>Kupferberg, S.J., 1997. Bullfrog (*Rana catesbeiana*) invasion of a California river: the role of larval competition. _Ecology_ 78, 1736–1751.
+>
+>La Marca, E., Lips, K.R., Lotters, S., Puschendorf, R., Ibanez, R., Rueda-Almonacid, J.V., Schulte, R., Marty, C., Castro, F., Manzanilla-Puppo, J., others, 2005.
+Catastrophic population declines and extinctions in Neotropical harlequin frogs (Bufonidae: *Atelopus*) _Biotropica: The Journal of Biology and Conservation_ 37, 190–201.
+>
+>Martel, A., Spitzen-van der Sluijs, A., Blooi, M., Bert, W., Ducatelle, R., Fisher, M.C., Woeltjes, A., Bosman, W., Chiers, K., Bossuyt, F., others, 2013.
+*Batrachochytrium salamandrivorans* sp.
+nov.
+causes lethal chytridiomycosis in amphibians.
+_Proceedings of the National Academy of Sciences_ 110, 15325–15329.
+>
+>Pimm, S.L., Jenkins, C.N., Abell, R., Brooks, T.M., Gittleman, J.L., Joppa, L.N., Raven, P.H., Roberts, C.M., Sexton, J.O., 2014.
+The biodiversity of species and their rates of extinction, distribution, and protection.
+_Science_ 344, 1246752
+>
+>Wake, D.B., Vredenburg, V.T., 2008. Are we in the midst of the sixth mass extinction? A view from the world of amphibians. _Proceedings of the National Academy of Sciences_ 105, 11466–11473.
+>
+>
+
+Be aware that these styles (Vancouver and Harvard) represent two broad approaches, every publisher has their own variation on these, and that they can vary substantially between different journals.
+Full details are always provided by the journal (in their **Instructions to Authors**), and you can refer to these to see how your references need to be formatted.
+I also provide some pointers towards using [reference managers](referencesoftware.html).
+These can save you a lot of time, if you've already invested time in setting them up.
+Alternatively, they can be very frustrating.
+
+Lots of journals (or perhap's it's their editors) are very fussy about the way that citations are given in their publications.
+If you have used a decent reference manager then you won't care, and it'll just be a case of finding the appropriate format from their repository (or making it yourself).
 
 There are lots of better things that you could be doing with your time instead of formatting references.
-And so I'm not going to spend an awful long time here telling you that you must do it right.
+And so I'm not going to spend a long time here telling you that you must do it right.
 I really believe that on first submission of your manuscript you should be able to submit your references in any format.
 it really only makes sense for you to do all the work formatting them if you aren't going to pay to have published the paper.
+An increasing number of journals do allow you to use any referencing style (within the generic Vancouver or Harvard format) when you first submit.
+You are only then required to format them if your ms is likely to be accepted.
 
@@
 Please don't join them and if you can look for a better world where you spend your valuable time doing more important things.
-For me, at this moment, it does not include any referencing software that I know of.
 

### 2021-04-24 index.Rmd
Add files via upload
# After some small edits over the week

@@
   easier if you understand the basic ground rules. This book is a guide specifically
-  for writing a PhD in the *biological sciences*. It will guide you from deciding
+  for writing a PhD in the biological sciences. It will guide you from deciding
   whether or not you should actually do a PhD, provide practical, up-to-the-minute
@@
 
+# Under Contract {-}
+This book is currently under contract with CRC Press. 
+```{r fig.align='center', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/Writing_Publishing'}
+knitr::include_graphics('figures/Cover.png', dpi = NA)
+```
+
 # Welcome {-}
@@
 
-This book started as a series of blog posts written by me for my students (john.measey.com/blog).
+This book started as a series of blog posts written by me for my students ([john.measey.com/blog](john.measey.com/blog)).
 I wrote them because I felt that I was repeating the same advice year after year.
@@
 Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning.
-James Baxter-Gilbert, Anthony Herrel,  Lisa O'Connell, Mark O'Connell, James Vonesh all read parts and commented.
+James Baxter-Gilbert, Anthony Herrel,  Allan Ellis, James Vonesh all read parts and commented.
 Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and Stellenbosch University.

### 2021-07-20 00-author.Rmd
Add files via upload
# After copy edits

@@
 
-```{r fig.align='right', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/About-Me'}
+```{r fig.align='right', out.width = '15%', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/About-Me'}
 knitr::include_graphics('figures/john.jpg', dpi = NA)
@@
 
-# Do you have something to contribute? {- #contribute}
+## Do you have something to contribute? {- #contribute}
 
 This book is written in bookdown [@xie2016bookdown] specifically to make it a 'live project' that will be open to anyone who wants to contribute, improve, or use as the basis for your own book.
-One of the amazing potentials for bookdown books is that all the files for this book are hosted on [Github](https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences/tree/main).
+The easiest way for readers to contribute content directly is through a [GitHub pull requests](https://help.github.com/articles/about-pull-requests/).
+At the repository for this book, you will find Rmd files for each chapter, and as a GitHub user, you can simply edit the Rmd file and submit the changes.
+If I am happy with the changes proposed, I will merge your content with that of the book and add your name to the [Acknowledgements](#acknowledge).
+
+One of the amazing potentials for [bookdown books](https://bookdown.org/) is that all the files for this book are hosted in a repository on [Github](https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences/tree/main).
 You have the opportunity to fork this repository and write your own version for a different discipline, a different language or for a different region of the world.
-It would be impossible to cover every aspect of writing a PhD in Biological Sciences, and it may be that I have missed out ones that are very important to you.
-Equally, parts of what is currently written will become obsolete very quickly as new initiatives are begun, and old problems are resolved.
+It is also my hope that this guide can grow to become a community of practice for those conducting PhDs in Biological Sciences.
+It will not be possible to cover every aspect of writing a PhD in Biological Sciences, but it may be that I have missed out ones that are very important to you.
+Equally, parts of what is currently written will become obsolete as new initiatives begin, and old problems are resolved.
 For this reason, this guide needs to be a 'living document', and anyone who wants to provide feedback or contribute new sections is more than welcome.
 Please feel free to open an issue, or make a Pull Request if you spot a typo.
+
+## Creative Commons License {-}
+How to write a PhD in Biological Sciences: a guide for the uninitiated is copyright to John Measey and distributed under a Creative Commons BY-NC-ND 4.0 International License.
+You may freely share this content for non-commercial purposes, with a source credit to [www.HowToWriteAPhD.org](www.HowToWriteAPhD.org).
+
+## Disclaimer {-}
+
+Although I have tried to make the information is this book as accurate as possible, it is provided without any warranty.
+The author and publisher have neither liability nor responsibility to any person or entity related to any loss or damages arising from the information contained in this book.
+
+```{r iris}
+knitr::kable(
+  head(iris), caption = 'We must have the iris data and some R code, otherwise it would not seem right',
+  booktabs = TRUE
+)
+```

### 2021-07-20 1.01-beforeyoustart.Rmd
Add files via upload
# After copy edits

@@
 ---
-```{r iris}
-knitr::kable(
-  head(iris), caption = 'We must have the iris data and some R code, otherwise it would not seem right',
-  booktabs = TRUE
-)
-```
 
@@
 
- - The median time to completion in many countries worldwide is much longer than it should be [e.g. 7.5 years in the USA @geven2018how].
- - Completion rates are shockingly low [as low as 50% in the USA @litalien2015improving].
+ - The time to completion in many countries worldwide is much longer than it should be.
+For example, 7.5 years is the median time to completion in the USA [@geven2018how], Europe universities have largely adopted the Salzburg Principles that doctoral programmes should be three to four years for full-time candidates  [@christensen2005bologna].
+Following this, an average is recorded as being between 3.5 and 4.5 years, with most institutions having reduced this time over the last 10 years [@hasgall2019doctoral].
+ - Completion rates are shockingly low. As bad as 50% in the USA [@litalien2015improving], 66% (within 6 years) in Europe [@hasgall2019doctoral], but generally better in the sciences [@park2005new].
 
 This has spurred many responsible for postgraduate students to try to ensure that they receive more support for their studies.
-But there is also the need for students to comprehend right from the start that they are undertaken a major commitment that will likely impact many different parts of their lives.
+But there is also the need for students to comprehend right from the start that they are undertaking a major commitment that will likely impact many different parts of their lives.
 
-In this first chapter of the book I intend to make you familiar with what a PhD is, how to approach thinking about it, and what to look for in an institution and advisor.
+In this first chapter of the book, I intend to make you familiar with what a PhD is, how to approach thinking about it, and what to look for in an institution and advisor.
 You may already have some or all of these things fixed, but there is still some use in going over these basic points.
@@
 The time that you spend doing your PhD will likely include participating in conferences, giving talks, participating in journal clubs, seminars, workshops and working groups.
-In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular area, and learn to communicate this to different types of audience.
+In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular subject area, and learn to communicate this to different types of audience.
 You will also learn and develop your skills in presentation, analysis and treatment of data.
@@
 
-## Criteria used to judge PhD studies
-In order to provide you with some general idea of what is required, I list below the (edited) questions asked of examiners for a PhD thesis by my university:
+## Criteria used to judge PhD studies {#criteria}
+In order to provide you with some general idea of what is required, I list below the (edited) questions asked of examiners for a PhD thesis by my university (you should ask for the criteria at your own institution):
 
@@
 - Does the dissertation distinguish clearly between own, new contributions to, and known results in the relevant field of study?
-- Is the candidate capable of evaluating the scientific meaning of his / her results and of placing this in context within existing knowledge in the field of study?
+- Is the candidate capable of evaluating the scientific meaning of their results and of placing this in context within existing knowledge in the field of study?
 - Does the candidate show signs of independent, critical thinking or other signs of originality?
@@
 If you compare these PhD requirements with those of an MSc, the key difference is that for a PhD you are required to produce a "meaningful contribution to the knowledge of and insight into the relevant field of study", and show that you are "...capable of doing independent research".
-Thus, a PhD degree prepares you to be an independent researcher, and so you are expected to have a critical attitude toward other research.
+Thus, a PhD degree prepares you to be an independent researcher, and so you are expected to have a critical attitude towards other research.
 
@@
 - Second is the clear knowledge of how the work that is conducted fits into the bigger picture.
-That is that the work demonstrates that the person is a scientist at a broader level than the narrow focus of their research.
+That is, that the work demonstrates that the person is a scientist at a broader level than the narrow focus of their research.
 - Third is whether the student has been able to bring the study together in a way that shows that it is more than the sum of its parts.
@@
 ## What will you do after your PhD?
-It sounds silly to think about what comes after your PhD just as you are just starting, but the reality is that you always need to be planning ahead in order to make the most from what you are doing now.
+It sounds silly to think about what comes after your PhD just as you are just starting, but the reality is that you always need to be planning ahead in order to make the most of what you are doing now.
 During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas.
@@
 The number of PhDs in a country is considered  (by the [World Bank](https://datacatalog.worldbank.org/dataset/world-development-indicators)) to be one of many measures of the development of that country.
-As the percentage gets larger, so countries can convince the world bank that they are developing (and then get granted its favours, see Figure \@ref(fig:PhDMap)).
+As the percentage gets larger, so countries can convince the World Bank that they are developing (and then get granted its favours, see Figure \@ref(fig:PhDMap)).
 This is all well and good, but strategically, governments should consider why they want these researchers in their populations, and more particularly for what kinds of jobs.
@@
 
-(ref:PhDMap) **Where in the world are there most PhDs?** In this map of the world, countries shaded are those that have data at the declaring the proportion of PhD (or equivalent) in their population (Max percentage between 2010 and 2020 (Data from the World Bank).
+(ref:PhDMap) **Where in the world are there most PhDs?** In this map of the world, countries shaded are those that have data declaring the proportion of PhD (or equivalent) in their population (max percentage between 2010 and 2020 (Data from the World Bank).
 This ranges from 2.97% in Switzerland (dark red) to 0.02% in DR Congo (light yellow).
-Notably there is no data held in these years for many countries in Asia, Africa and South America.
+Notably, there is no data held in these years for many countries in Asia, Africa and South America.
 
@@
 But the reality is that you will never do a PhD for somebody else, you are always going to be doing it for yourself.
-if the only reason you are doing a PhD is because someone else told you that you should then you are likely to have a really hard time.
+If the only reason you are doing a PhD is because someone else told you that you should then you are likely to have a really hard time.
 By the time you graduate, you may have a PhD but without any idea of what you're going to do with it.
@@
 You should be aware by now that there are lots of people in academia who are ‘institutionalised’.
-By this I mean that they have never had a job (or a life) outside of academia.
+By this, I mean that they have never had a job (or a life) outside of academia.
 This is not something that I can recommend for anyone.
@@
 Yet in Belgium, for example, the numbers of PhD graduates entering the market nearly doubled (in the first 15 years of the 21st century) while academic positions remained static [@levecque2017work].
-Universities appear to be exploiting this excess of qualified people to hire them on contracts, reducing the number of full time positions.
+Universities appear to be exploiting this excess of qualified people to hire them on contracts, reducing the number of full-time positions.
 Essentially, this means that in most places your time as a PhD student, postdoc and junior lecturer will keep you outside the employers core staff.
-Joining those on the inside (with well funded posts) becomes increasingly more difficult.
-Some have suggested that the world of academia is becoming more like a gang producing drugs [@afonso2014academia].
+Joining those on the inside (with well-funded posts) becomes increasingly more difficult.
+Some have suggested that the world of academia is becoming more like a gang producing drugs, with those tenured academics earning good, secure salaries at the top of the pyramid, while increasing numbers of insecure poorly paid doctoral students and post-docs do most of the actual work [@afonso2014academia].
 
 ## Doing a PhD is really hard work
-It’s not a MSc with an extra year.
-It really is much harder and tougher on you mentally and physically.
+It’s not an MSc with an extra year.
+It really is much harder and tougher on you [mentally and physically](#healthy).
 The chances that you’ll end up jaded and malcontent with the entire academic system are very high (maybe even obligatory).
-You are probably not going to be very happy while doing it (there are some exceptions of course), and it will likely have negative impacts on all the good things that you currently have in your life (think family, friends and relationships, as well as free time and fun).
+At times, you are probably not going to be very happy while doing it (there are some exceptions of course), and it will likely have negative impacts on all the good things that you currently have in your life (think family, friends and relationships, as well as free time and fun).
+On the other hand, it can simultaneously bring joy and start you off on a lifetime trajectory.
 
@@
 Being aware of this at the outset will allow you to alert close members of your non-working life (think partner, family and friends) that they may well need to act as a support network.
-It is worth tracking your mental health during your PhD to check that you are not getting into difficulties.
+It is worth tracking your mental health during your PhD to check that you are not getting into difficulties, and I show you later [how to do this](#healthy).
 
@@
 
-The country where you study a PhD does not determine the quality of institution (see Figure \@ref(fig:THEreseasrchcountry).
+The country where you study a PhD does not determine the quality of institution (see Figure \@ref(fig:THEreseasrchcountry)).
 Rather look at the merit of the particular institution that you are interested in.
@@
 
-(ref:THEreseasrchcountry) **It’s not the country that determines the quality of your university.** In this plot you see the Times Higher Education (THE) World Ranking (2021) score for Research plotted by country (with ≥ 10 universities).
+(ref:THEreseasrchcountry) **It’s not the country that determines the quality of your university.** In this plot, you see the Times Higher Education (THE) World Ranking (2021) score for Research plotted by country (with ≥ 10 universities).
 Note that most universities in most countries have mean THE Research scores in the lowest quartile.
@@
 ## What to look for in an advisor
-Throughout this book I will call the person who is the academic in charge of your PhD studies your ‘advisor’.
+Throughout this book, I will call the person who is the academic in charge of your PhD studies your ‘advisor’.
 I will not call them your supervisor.
@@
 Many of my students have not followed my advice.
-For some this has been the right thing to do.
-For others it has not.
+For some, this has been the right thing to do.
+For others, it has not.
 No one can know what the best thing to do is in different situations, because they will almost always be unique.
@@
 
-One of the best ways of picking a good advisor is by talking to current or past students.
+One of the best ways of picking a good advisor is by talking to past students, or current students that have been in the lab a while.
 You should ask them how happy they are with their advisor and whether they are receiving the support that they need to do their projects.
@@
 
-At the end of the day there is no such thing as the perfect advisor.
+At the end of the day, there is no such thing as the perfect advisor.
 You will likely have to make a trade-off on the above points with someone who has a project that they are willing to advise you on, and your chosen area of research.
@@
 ## What makes a good project?
-After having made lots of statements about how a project should be your own project the following advice might seem a bit strange.
-Clearly the work that you're going to be doing will need to fit into the group of the advisor that you've chosen.
-
 The most obvious point to make here is that a project should be something that you want to do and that you're prepared to dedicate yourself to for ~3 years.
@@
 ## Funding your PhD studies
-Funding may be one of the biggest obstacles that you come across when wanting to do a PhD. It's important that you consider the funding of your PhD in two separate ways:
+Funding may be one of the biggest obstacles that you come across when wanting to do a PhD. It's important that you consider the funding of your PhD in two separate ways.
 
@@
 It seems to me that many countries offer bursaries that are not actually enough to live on (e.g.
-buy food, rent a house etc.).
+buy food, rent a house, etc.).
 If you're offered a PhD bursary, especially if it's in another country, then do check on the living cost at that location and whether or not the bursary that you are being offered will be sufficient money in order for you to survive.
@@
 There is no one single model, and the more people you talk to the more likely you are to discover one that might suit you.
-Of course conferring with other people can include reading the accounts of many different people who share their experiences on social media.
+Of course, conferring with other people can include reading the accounts of many different people who share their experiences on social media.
 There are blogs and Twitter accounts dedicated to this stuff.
@@
 ### The cost associated with your studies. 
-The money associated with the project that you were doing also needs to come from somewhere.
-Depending on where in the world you can do these studies they will be different expectations of where this money comes from.
+The money associated with the project that you are doing also needs to come from somewhere.
+Depending on where in the world you can do these studies there will be different expectations of where this money comes from.
 This is a very important topic for you and your potential advisor to discuss.
 During the proposal phase of your PhD you'll be expected to put together a budget to determine the total costs of your studies.
-In reality you or your advisor should have a ballpark figure of how much this Is going to cost and have some idea about where the money is going to come from.
+You or your advisor should have a ballpark figure of how much this is going to cost and have some idea about where the money is going to come from.
 
@@
 For example, if your intention is to sequence genomes of every animal or plant that you work with, the finance required will likely be far greater than your bursary.
-If however your intention is to do a PhD that models distributions of populations of animals and plants, then the entire costs may only run to a computer remote access to a computing cluster and data storage.
+If however your intention is to do a PhD that models distributions of populations of animals and plants, then the entire costs for your study may only add up to a computer, remote access to a computing cluster and data storage.
 

### 2021-07-20 1.02-useR.Rmd
Add files via upload
# After copy edits

--- 2021-04-15 1.02-useR.Rmd
@@
 # Reconciling yourself to doing things you've been avoiding {#useR}
-Some 30 odd years ago as an undergraduate I was told that I would have to use computers in one of my 2nd year courses.
+
+Some 30 odd years ago, as an undergraduate, I was told that I would have to use computers in one of my 2nd year courses.
 The thought filled me with dread and I remember putting off going into the computer room for as long as I could.
 It was a steep learning curve, and I struggled.
-Each time the cursor got to the end of the screen I hit the carriage return button.
+As my only previous experience had been with a typewriter, each time the cursor got to the end of the screen I hit the 'carriage-return' button.
 And each machine had an odd white device attached called a mouse, and I had no idea what that did.
 All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC.
-This is practically where we are now with R [@rcoreteam2021r], and indeed a wealth of other alternative programming platforms.
+This is practically where we are now with R (@rcoreteam2021r; Figure \@ref(fig:Rlogo)), and a wealth of other emerging and alternative programming platforms.
 
@@
 If you are not already aware of the prominence that R now has in conducting statistical analyses in biological sciences, then you’ve missed out on a great deal.
-The real point here is not that you can only use R, but that R represents a platform that can be easily reproduced (because it’s free unlike many statistical software packages), and by saving and making the code available, you are making your data analyses reproducible.
-We will see elsewhere the importance of reproducibility in sciences [@baker20161500], and the critical role this has in transparency and therefore best practice.
+The real point here is not that you can use R, but that R represents a platform on which analyses can be easily repeated, and by saving and making the code available together with your data, you are making your data analyses reproducible.
+We will see elsewhere the importance of reproducibility in sciences [@baker20161500], and the critical role this has in transparency and, therefore, best practice.
 
@@
 If you already know all of the above, then you won’t face any major barriers, and you will fit into the new and more transparent world of science with great ease.
-If the thought of using R fills you with dread, in the same way that using a computer did some 30 years ago, then you should reconcile yourself now before you go any further.
+If the thought of using R fills you with dread, in the same way that using a computer did for me some 30 years ago, then you should reconcile yourself now before you go any further.
 Learning R will be a steep learning curve, but it will be made easier by the large amount of excellent video tutorials and online courses that are available.
 
-## And learning lots of other stuff...
+## And learning lots of other stuff
 The same principle that I’ve written about above can be extended to all sorts of other aspects of your PhD project.

### 2021-07-20 1.03-lifescientific.Rmd
Add files via upload
# After copy edits

@@
 Science is a process in which we reach conclusions based on egalitarian reasoning and evidence, something that we can aspire for the larger community around us.
-Already, most societies are making the majority of their collective evidence based decisions through the guidance of science.
-Therefore, if we consider our knowledge based societies to be democratic, then all citizens of that society should understand the processes under which the foundation of that knowledge is acquired [@sagan2011demon].
+Already, most societies are making the majority of their collective evidence-based decisions through the guidance of science.
+Therefore, if we consider our knowledge-based societies to be democratic, then all citizens of that society should understand the processes under which the foundation of that knowledge is acquired [@sagan2011demon].
 Yet globally, public skepticism in science has been rising, and has reached as high as 35%.
@@
 
-Imagine a world in which all of the principle decisions are made as a result of advances in science, but vanishingly small numbers of people actually understand how science works.
+Imagine a world in which all of the principal decisions are made as a result of advances in science, but vanishingly small numbers of people actually understand how science works.
 This is not a world that we would want, but as Carl Sagan [@sagan2011demon] pointed out, it’s one that could all too easily happen.
@@
 Moving forward with the scientific project does not mean that everyone must become a scientist, but instead that the process by which scientists establish knowledge is understood.
-Indeed, as we shall see (below) sceptical thinking, as exemplified in science, is indeed a healthy way for everyone in society to avoid time wasting and potentially being deceived or conned.
+Indeed, as we shall see (below), sceptical thinking, as exemplified in science, is a healthy way for everyone in society to avoid time wasting and potentially being deceived or conned.
 It is also vital that the scientific process be transparent in a way that results can be understood and therefore respected.
@@
 Alien species are those that have been moved by human agents across biogeographic barriers into parts of the world where they did not evolve.
-Once these aliens arrive in their new environment, some breed and become established, and after a lag a fraction of these spread and become invasive, often with substantial impacts on the environment and the socio-economic efforts of people in those invaded areas.
+Once these aliens arrive in their new environment, some reproduce and become established, and after a lag, a fraction of these spread, and become invasive, often with substantial impacts on the environment and the socio-economic efforts of people in those invaded areas.
 Invasive species disproportionately affect poorer people in our society, often completely disassociated with the pathways responsible for the introduction and spread.
-Removing invasive species requires a considerable amount of effort, and unless this is supplied by tax-payers, via governmental policy, falls on those directly impacted.
+Removing invasive species requires a considerable amount of effort, and unless this is supplied by taxpayers, via governmental policy, falls on those directly impacted.
 Invasion science, the expanded transdisciplinary area stemming from invasion biology, is required to engage with all stakeholders in invasions, and most importantly to prevent new invasions.
@@
 Invasions at larger scales will normally require governmental resources to ensure that communities can continue to function, especially to best deliver their services to the society.
-Governments require evidence based decisions that are reached from well researched studies conducted in the field.
+Governments require evidence-based decisions that are reached from well-researched studies conducted in the field.
 Effective measures may necessitate large sums of money, and governments require that this process be transparent so that the taxpayers’ money they are investing is spent efficiently.
@@
 Over time, views from science that appear radical to society become mainstream and accepted.
-Think of the views on the earth being round, or more recently the general acceptance that the universe started with the ‘big bang’.
-The counterculture, also known as post truth, faction in society that espouses views accepted by the mainstream and proved by science beyond reasonable doubt is not a novel phenomenon.
-Indeed, in the scientific project, we need to accept that these views will persist, but empower the general populace to understand why and how they can be shown to be false.
+Think of the views on the earth being round, or more recently, the general acceptance that the universe started with the ‘big bang’.
+The counterculture, also known as post-truth, faction in society that espouses views accepted by the mainstream and proved by science beyond reasonable doubt is not a novel phenomenon.
+Indeed, in the scientific project, we need to accept that these views will persist and empower the general populace to understand why and how they can be shown to be false.
 Hence, your view that the earth is not flat is bolstered if you can provide three ways to demonstrate this as fact (pick the ones you like from 7 listed [here](http://crosstalk.cell.com/blog/seven-ways-to-prove-earth-is-round) or [here](https://www.unlv.edu/news/release/round-earth-clues-how-science-proves-our-home-globe)).
 Otherwise, your assertion that it is not flat will appear as easily denied as a ‘flat-earther’s’ assertion that it is.
-The power of the scientific argument in society is that it is not simply contradiction or gainsay, but that there is evidence.
+Thus the power of the scientific argument in society is that it is not simply contradiction or gainsay but that there is evidence.
 
@@
 Reforms are needed throughout the structures that fund, employ and disseminate scientific research.
-Sweeping this all away will take a concerted effort, but might well be easier in science than in other sectors of society.
+Sweeping this all away will take a concerted effort but might well be easier in science than in other sectors of society.
 Nearly all scientists recognise that science needs to be objective and impartial in order to succeed.
@@
 These are well worth recording here as they provide a guiding light for all scientists whether they be novices, emerging or established.
-They are also easily taught, and the foundation of what we should spread in the Scientific Project.
+They are also easily taught and the foundation of what we should spread in the Scientific Project.
 
 As humans, we are all susceptible to bullshit, hoaxes, charlatans and contrived deceptions as they are often more appealing than the truth.
-However, science is about truth seeking, and by applying a set of simple rules, Sagan argues, we can avoid being misled or distracted by ideas that may waste our time and energy by using ‘skeptical thinking’.
+However, science is about truth-seeking, and by applying a set of simple rules, Sagan argues, we can avoid being misled or distracted by ideas that may waste our time and energy by using ‘skeptical thinking’.
 These rules are useful in your scientific life when considering hypotheses or when building arguments to explain your results.
@@
 1. You should always seek independent confirmation of results. 
-1. Seek out different opinions on the results and consider all possibilities
+1. Seek out different opinions on the results and consider all possibilities.
 1.
 Don’t believe facts simply because they are spoken by an authority on the subject.
-We know that humans are prone to [confirmation bias](type1error.html), and often exhibit motivated reasoning, especially if they have some vested interest.
+We know that humans are prone to [confirmation bias](#TypeI), and often exhibit motivated reasoning, especially if they have some vested interest.
 Even the most experienced have made mistakes in the past and will again in the future.
@@
 1.
-Don’t get too involved with keeping your chosen hypothesis [opening you to potential cofirmation bias](type1error.html).
+Don’t get too involved with keeping your chosen hypothesis [opening you to potential cofirmation bias](#TypeI).
 Remain critical to the possibility that it might be wrong, and provide ideas on how to substantiate or disprove competing hypotheses.
@@
 1.
-If at all possible use a quantitative method to test your hypothesis.
+If at all possible, use a quantitative method to test your hypothesis.
 You should already be well on the way to this one inside your PhD, but it is worth bearing in mind especially when it comes to considering testing other competing hypotheses.
@@
 1.
-Arguments ([as we see later](Argument.html)) consist of a chain of statements that follow one from the other.
+Arguments ([as we see later](#argument)) consist of a chain of statements that follow one from the other.
 This chain is only as strong as its weakest link, and the strength of each link in the chain (your argumentative statements) must be critically considered by you.
 1.
-[Occam’s Razor](Argument.html) dictates that given two hypotheses that equally explain the results, the simpler one is likely more plausible.
+[Occam’s Razor](#argument) dictates that given two hypotheses that equally explain the results, the simpler one is likely more plausible.
 Or put another way, the more links you have in your chain, the more likely that one of them is not as strong as you would like it to be.
 1.
-Your [hypothesis](howhypothesis.html) (or argument) must always be falsifiable.
+Your [hypothesis](#hypothesis) (or argument) must always be falsifiable.
 That is, that your test should be able to find that it is incorrect.
@@
 
-Sagan provides plenty of real world examples of each of these arguments, and I’d suggest that you consult his book [@sagan2011demon] if you are interested in following these up.
+Sagan provides plenty of real-world examples of each of these arguments, and I’d suggest that you consult his book [@sagan2011demon] if you are interested in following these up.
 Similar approaches, but written in different styles, are provided by Richard Dawkins in his [-@dawkins2004devil] book: A Devil’s Chaplain, and Julia Galef in her [-@galef2021scout] book: The Scout Mindset.
@@
 
-I would encourage you not to stop your objective and evidence based thinking when you leave the office.
+I would encourage you not to stop your objective and evidence-based thinking when you leave the office or laboratory.
 By applying the rules of science to other sectors of society, you will find that there are many societal norms that could benefit from similar reform.
-Certainly, a lot of the technology developed in science can be used to help society reform, but scientific thinking will help everyone.
-As well as bringing the world to scientific thinking, we can take scientific thinking to the world.
+Certainly, a lot of the technology developed through scientific inquiry can be used to help society reform, but scientific thinking will help everyone.
+As well as bringing the world to scientific thinking, as scientists we should take scientific thinking to the world.
 
 ## Citizen science
-For a long time I struggled with the term 'Citizen science' as I reasoned that many such ventures did not involve citizens doing science.
+For a long time, I struggled with the term 'citizen science' as I reasoned that many such ventures did not involve citizens doing science.
 To me, most seemed focussed on outsourcing large tasks to willing public participants: i.e.
 having citizens do some grunt work.
-Collecting data, or doing a fraction of the analysis is not science.
+Collecting data or doing a fraction of the analysis is not science.
 However, in light of the scientific project, I have changed my mind.

### 2021-07-20 1.04-healthy.Rmd
# Updating sections to chapters
Add files via upload
# After copy edits

@@
-## A healthy body to go with your PhD mind {#healthy}
+# Keeping track of your mental health {#healthy}
 
-While I’m busy handing out advice, here’s some of the best that I have to offer.
+Stress is a natural part of life and many people are at their most productive when they are under some degree of pressure, such as a deadline. Although deadlines don’t work for everyone.
+
+>Douglas Adams famously claimed: 
+>
+>“I love deadlines. I like the whooshing sound they make as they fly by.”
+
+Problems arise when we become overwhelmed by stress and are unable to fully respond.
+When this occurs, productivity can drop off and survival responses can be triggered as if responding to an actual physical attack.
+These responses include fight, flight or freeze responses.
+Anxiety and panic can be triggered.
+In this state, additional demands on your time may also push your life off balance, so that you start to neglect your personal wellbeing which can negatively impact on relationships, exercise regime, or even nutrition and personal hygiene.
+Some people can find that the additional stress can cause physical symptoms that may even need medical treatment.
+Your sense of competence and mastery can be negatively impacted such that you may even suffer from feelings of inadequacy or imposter syndrome ([see part 2](#fear)).
+
+Although there are not many studies on mental health for PhD students, those that exist (as well as surveys: [Nature 2019](https://www.nature.com/articles/d41586-019-03459-7)) all suggest that there is a significant toll, which is proportionately higher than for others in society [@levecque2017work].
+Whatever your prior experience of stress in a working environment, academia is known to be particularly stressful, and as a PhD student, you are likely to absorb a significant amount of this stress into your own life [@stubb2011balancing].
+
+The General Health Questionnaire (see GHQ-12 in Table \@ref(tab:GHQ)) is an instrument used to measure psychological distress.
+It is quick, reliable and simple to score, so you can use it at any time during your PhD studies as an indicator of whether you need to reach out to personal, occupational or professional support networks.
+
+Right now, I suggest you complete the GHQ-12 (Table \@ref(tab:GHQ)) and record your answers as a baseline.
+Keep the scores somewhere safe.
+During the course of your PhD, if you feel that your scores may have changed, take the test again and compare them with your baseline scores.
+Although there are no hard rules, if three or more of your scores have moved by two or more points it could be worth discussing with your support network to help you decide whether or not to seek professional help.
+
+Table: (\#tab:GHQ) **A General Health Questionnaire with 12 questions (GHQ-12) that you can use to keep track of your mental health.** Answer the questions in this table now, and compare your answers to answers in your future when you are feeling more under strain.
+Write down any explanations you have for your choice of answers.
+
+| General Health Questionnaire: Have you recently... | 0 | 1 | 2 | 3 |
+|-|-|-|-|-|
+| been feeling reasonably happy, all things considered? | Better than usual | Same as usual | Less than usual | Much less than usual |
+| lost much sleep over worry? | Not at all | No more than usual | More than usual | Much more than usual | 
+| been feeling unhappy and depressed? | Not at all | No more than usual | More than usual | Much more than usual |
+| felt you couldn’t overcome your difficulties? | Not at all | No more than usual | More than usual | Much more than usual |
+| felt under constant strain? | Not at all | No more than usual | More than usual | Much more than usual | 
+| felt capable of making decisions about things? | Better than usual | Same as usual | Less than usual | Much less than usual |
+| been able to face up to your problems? | Better than usual | Same as usual | Less than usual | Much less than usual | 
+| been thinking of yourself as a worthless person? | Not at all | No more than usual | More than usual | Much more than usual |
+| been losing confidence in yourself? | Not at all | No more than usual | More than usual | Much more than usual | 
+| been able to enjoy your normal day-to-day activities? | Better than usual | Same as usual | Less than usual | Much less than usual |
+| been able to concentrate on whatever you are doing? | Better than usual | Same as usual | Less than usual | Much less than usual |
+| felt that you are playing a useful part in things? | Better than usual | Same as usual | Less than usual | Much less than usual |
+
+Even if you don't feel you need the support of your institution now, it is worth finding out how they can support your mental health in the future if needed.
+Although there has been some stigma attached to difficulties with mental health in the past, most institutions accept that pressures are mounting on postgraduate students and that they may require support.
+Most institutions have experienced councillors available to support you if needed.
+Importantly, you should realise that none of these symptoms are unusual and that there is a high probability that many of your colleagues may also be struggling.
+Knowing that your problems are shared and reaching out to support networks early is an excellent way to prevent them from escalating beyond your control.
+
+A study into the mental health of PhD students in Belgium exemplifies the kinds of difficulties that they face when compared with other similar groups (@levecque2017work; Figure \@ref(fig:Mental-health)).
+
+(ref:Mental-health) **How does the mental health of people in different sectors compare?** A comparison of the mental health of PhD students [data from @levecque2017work] with highly educated general population, highly educated employees and higher education students using the General Health Questionnaire.
+The Risk Ratio (RR: adjusted for age and gender) in PhD students in Flanders, Belgium, is consistently higher (>1) when compared to any of the other surveyed groups.
+
+```{r Mental-health, echo=FALSE, out.width='80%', fig.cap="(ref:Mental-health)"}
+ knitr::include_graphics("figures/Mental-health.png")
+```
+
+No matter how well you think of your own abilities to cope with mental health issues, doing a PhD will cause you additional stress and can trigger maladaptive coping mechanisms.
+Learning how to cope with additional stress early in your career can be beneficial for future personal development.
+
+Academia is recognised as a particularly stressful environment; you will likely take on some of this environmental stress in addition to any stress associated with your studies.
+Additional stressors come from home and family situations.
+Your best means of coping will be to try and develop a support network and to understand where and with whom you can discuss any difficulties as they arise.
+Knowing who this is and how and when to approach them will put you in a stronger position if you need them in future.
+
+## Being physically active improves mental wellbeing
+
+The positive relationship between the amount of physical activity and higher mental wellbeing is well established [e.g.
+@grasdalsmoen2020physical; @gerber2014increased], but the kind of exercise required to achieve this improved result is varied.
+For example, a study of college students showed that an hour of Thai Chi twice a week for 3 months was enough to have significantly improved physical and mental health scores [@wang2004effects].
+Increased levels of physical activity were also found to be associated with improved sleep quality in another study [@ghrouz2019physical; @gerber2014increased].
+There are plenty of studies out there that suggest there are multiple benefits from physical activity.
+
 If you don’t do it already, then when you start your PhD start regular exercise.
-There’s an old adage that ‘a healthy body leads to a healthy mind’.
-I never gave it much attention until I started exercising regularly some 7 years ago.
 Getting fit was the best thing I ever did to improve my work.
-
-### Being physically active improves concentration
-
 The first thing I found was that I could sit and concentrate on work for longer.
@@
 
-### Time to think
+There are other ways to achieve these benefits, including mindfulness or meditation.
+
+## Time to think
 
@@
 But I find the meditative time I spend exercising to be especially productive.
+Don't take my word for it, see this great new book by Caroline Williams [-@williams2021move].
 
-### Balancing work with life
+## Balancing work with life
 
@@
 By life, I'm referring to everything away from your academic work.
-This might include, family, friends, sport, hobbies (even fishing).
+This might include family, friends, sport, hobbies (even fishing).
 If you don't have friends outside your academic life, then it would be a good idea to find some so that you can keep in contact with the 'real world'.
@@
 If you already exercise, then you likely know how important this is.
-But if you don’t then my best advice (above anything else that you might read in this book) is to start as soon as you can.
+But if you don’t then, my best advice (above anything else that you might read in this book) is to start as soon as you can.
 Do whatever is right for you.

### 2021-07-20 2.01-Hyp.Rmd
Add files via upload
# After copy edits

@@
 I think that there are several reasons for this.
-Writing well comes with practice, and most PhD students have had little experience with academic writing - even if they have already completed a MSc.
-The same could be said of most early career scientists (even after their PhDs), because practice is measured in completed pieces, it usually takes many years rather than weeks or months.
-YOu shouldn't expect that scientific writing will come automatically or immediately.
-In order to get this practice you need to get started, and (again) many students find starting to write challenging.
+Writing well comes with practice, and most PhD students have had little experience with academic writing even if they have already completed an MSc.
+The same could be said of most early-career scientists (even after their PhDs), because practice is measured in completed pieces, it usually takes many years rather than weeks or months.
+You shouldn't expect that scientific writing will come automatically or immediately.
+In order to get this practice, you need to get started, and (again) many students find starting to write challenging.
 This guide was written to help you get started.
@@
 In this part of the book, I start with an explanation of the hypothesis.
-As you will see in [Part III](hypothesis.html), the hypothesis is central to all of the writing in each chapter or paper, and you need to have a very good idea of how to formulate your research question into a hypothesis before you begin.
+As you will see, the hypothesis is central to all of the writing in each chapter or paper, and you need to have a very good idea of how to formulate your research question into a hypothesis before you begin.
 Then I build on helping you think through lots of different aspects of writing in science: style, citations, writing concisely, scientific names, and lots more.
@@
 
-When you read about the formula of writing a chapter (see end of part 2) you will see that everything in your chapter or paper revolves around the hypothesis or questions that you ask, normally at the end of your introduction.
-In this chapter, we will take a look at how to write a hypothesis.
+When you read about the formula of writing a chapter (see [end of part 2](#formula)) you will see that everything in your chapter or paper revolves around the hypothesis or questions that you ask, normally at the end of your introduction (Figure \@ref(fig:Hypothesis)).
+In this section, we will take a look at how to write a hypothesis.
 This is a sticking point for many students.
-We are used to using and writing questions and statements in day to day communications, as well as reading popular media.
+We are used to using and writing questions and statements in day-to-day communications, as well as reading popular media.
 But hypotheses (the plural of hypothesis) only rarely float across our desks.
@@
 
-In addition to this chapter, there is some good information out there on the web, and it's worth looking at this too: (e.g.
+In addition to this section, there is some good information out there on the web, and it's worth looking at this too: (e.g.
 [Wikihow](https://www.wikihow.com/Write-a-Hypothesis), [Wikipedia](https://en.wikipedia.org/wiki/Hypothesis), etc.).
@@
 (ref:Hypothesis) **Generating a good hypothesis for your PhD chapter isn't easy.** A good hypothesis will be invaluable in helping you write the chapter.
-It is important to start working with hypotheses
+It is important to start working with hypotheses.
 
@@
 But there's a little more to it than this.
-The hypothesis becomes a part of the scientific method if it is testable and (importantly) falsifiable, as well as being informed from previous published work on the subject.
+The hypothesis becomes a part of the scientific method if it is testable and (importantly) falsifiable, as well as being informed from previously published work on the subject.
 
-Yes, your hypothesis must be informed by the literature, which is why you spent so much time and effort crafting your introduction to inform your reader of the same.
+Your hypothesis must be informed by the literature, which is why you spent so much time and effort crafting your introduction to inform your reader of the same.
 This is also why your hypothesis usually comes at the end of your introduction, because you spend all of the introduction telling your reader about it.
-There's not much point in writing more after the hypothesis, because once your reader has read that, they are ready to learn about how you went about testing it (in the Materials & Methods).
+There's not much point in writing more after the hypothesis, because once your reader has read that, they are ready to learn about how you went about testing it (in the [Materials & Methods](#methods)).
 The other important point to make is that the literature should dictate how you write your hypothesis, and the variables that you include.
@@
 That's why it is conventional to provide the 'null hypothesis' that is the falsified version of the statement, suggesting that there is no relationship between the variables you have proposed to measure.
-The convention is to label this 'null hypothesis' H0, while the 'alternative hypothesis' (the one that says your variables are related as you suggested) is written as H1.
+The convention is to label this 'null hypothesis' H~0~, while the 'alternative hypothesis' (the one that says your variables are related as you suggested) is written as H~1~.
 When you formulate your hypothesis, it is traditional to write your alternative hypothesis to indicate the directionality of your tested variables.
-This way the reader can simply imply that the null hypothesis is when there is a relationship, but this will need to be stated if the null hypothesis is more complex.
+This way, the reader can simply imply that the null hypothesis is when there is no relationship, but this will need to be stated if the null hypothesis is more complex.
 
@@
 Most importantly, your hypothesis must come first, before you do the experiment or study.
-Hence the reason why this chapter comes at the start of part 2.
+Hence the reason why this section comes at the start of part 2.
 Setting the hypothesis after the work is already done is fraudulent, and goes against the scientific method.
-Obviously, it isn't fair to pose the hypothesis once you already know the answer.
+Obviously, it isn't fair to pose the hypothesis once you already know the answer (also known as [HARKing](#TypeI)).
 This is why there is so much emphasis put on formulating your hypothesis during your research proposal.
@@
 
-Writing a hypothesis isn't easy, but it is essential and once you've understood what to do most of the rest of what you are writing for should make sense.
+Writing a hypothesis isn't easy, but it is essential, and once you've understood what to do most of the rest of what you are writing for should make sense.
 
@@
 Both of these variables should then feature in your hypothesis.
-Next, by paragraph four you will have identified the problem that you are interested in tackling.
+Next, by paragraph four, you will have identified the problem that you are interested in tackling.
 In addition, your introduction will provide all of the pertinent literature that has relevance to this hypothesis, giving the all important context.
@@
 In addition, the causal mechanism is stated.
-You can watch a video about using the "If.
-then.
-because." construction here, or read more here.
 I say that this is a formulaic way to start writing your hypothesis, because it usually ends up as an inelegant statement, which can be better refined for a reader.
@@
 Exogenous causal explanations concern mechanisms that are outside the body of individuals.
-Common exogenous mechanisms are climatic factors (temperature, humidity, precipitation, etc.), or may relate to the availability of food, predators or mates.
+Common exogenous mechanisms are climatic factors (temperature, humidity, precipitation, etc.) or may relate to the availability of food, predators or mates.
 
@@
 
-### Teleological versus Causal Hypotheses
+### Teleological versus causal hypotheses
 
@@
 Although a reduction of predation may be a consequence of vertical migration in water fleas, each water flea does not think about predation and then starts it’s upward movement as a result.
-A common mistake made in biology is to apply teleological arguments to processes that have no purpose or region.
+A common mistake made in biology is to apply teleological arguments to processes that have no purpose or reason.
 Evolution is often mistakenly suggested to have a purpose (e.g.
-to evolve to a more advanced state), but in fact evolution is not a goal-orientated process.
+to evolve to a more advanced state), but in fact, evolution is not a goal-orientated process.
 There is no end-point to evolution, and evolution did not start in order to meet some predetermined form or function.

### 2021-07-20 2.02-TypeI.Rmd
# Updating sections to chapters
Add files via upload
# After copy edits

@@
-## Being aware that you can get it wrong {#TypeI}
+# Being aware that you can get it wrong {#TypeI}
 
@@
 
-With all the will in the world, when you are testing your hypothesis using statistics there is a chance that you will accept your alternative hypothesis when it is not valid.
+With all the will in the world, when you are testing your hypothesis using statistics, there is a chance that you will accept your alternative hypothesis when it is not valid.
 This is known as a ‘false positive’ or a Type I error (see Figure \@ref(fig:TypeITypeII)).
 It is also possible that you will accept your null hypothesis when you should have accepted your alternative hypothesis, also known as a Type II error, or a ‘false negative’.
-While it won't be any fun to get a Type II error, as scientists we should be more worried about Type I errors and the way in which they occur.
+While it won't be any fun to get a Type II error, as scientists, we should be more worried about Type I errors and the way in which they occur.
 This is because following a positive outcome, there is more chance that the work will be published, and that others may then pursue the same line of investigation mistakenly believing that their outcome is likely to be positive.
@@
 Another manifestation of this is publication bias which also tends to be biased towards positive results, with as many as 84% of studies supporting their initial hypothesis [@fanelli2010positive].
-Put together this means that scientists being human are more vulnerable to making Type I errors than Type II errors when evaluating their hypotheses.
+Thus scientists, being human, are more vulnerable to making Type I errors than Type II errors when evaluating their hypotheses.
 Although there are deliberate or inadvertent ways of making of Type I errors, here it is important to understand that simply by chance you can make a Type I error: accepting your alternative hypothesis even when it is not correct.
@@
 The rows are what the researcher finds when testing their hypothesis.
-The blue squares is what we are hoping to achieve when we set and test our hypothesis.
+The blue squares are what we are hoping to achieve when we set and test our hypothesis.
 The grey squares may happen if the hypothesis we set is indeed false.
 The other two possibilities are the false positive Type I error (red), and the false negative Type II error (black).
-In the table it seems that the chances of getting one of the four outcomes is equal, but in fact this is far from reality.
+In the figure it seems that the chances of getting one of the four outcomes is equal, but in fact, this is far from reality.
 There are several factors that can change the size of each potential outcome to testing a hypothesis.
@@
 
-(ref:HighlyUnlikely) **The likelihood that highly unlikely hypotheses you test are accepted.** This figure has been redrawn after figure 1 in Forstmeier et al [-@forstmeier2017detecting].
+(ref:HighlyUnlikely) **The likelihood that highly unlikely hypotheses you test are accepted.** This figure has been redrawn after figure 1 in Forstmeier et al.
+[-@forstmeier2017detecting].
 This is a graphical representation of an argument first made by Ioannidis [-@ioannidis2005why].
@@
 
-### Changing the likelihood of a hypothesis
+## Changing the likelihood of a hypothesis
 
@@
 Because α is set at 0.05, we can expect that 5% of the 990 incorrect hypotheses will give us a false positive (49 or 50 times).
-Although it might seem unlikely that anyone would test such highly unlikely hypotheses, there are increasing numbers of governments around the world that create incentives for researchers to investigate what they term blue skies research, which might be better termed high risk research or investigations into highly unlikely hypotheses.
+Although it might seem unlikely that anyone would test such highly unlikely hypotheses, there are increasing numbers of governments around the world that create incentives for researchers to investigate what they term 'blue skies' research, which might be better termed high-risk research or investigations into highly unlikely hypotheses.
 The real problem with such hypotheses is that you are more likely to get a Type I error than actually find that your hypothesis is truly correct.
 
-(ref:Unlikely) **The likelihood that unlikely hypotheses you test are accepted.** In this figure we see a scenario of unlikely hypotheses that are found to be correct approximately one in 10 times.
+(ref:Unlikely) **The likelihood that unlikely hypotheses you test are accepted.** In this figure, we see a scenario of unlikely hypotheses that are found to be correct approximately one in 10 times.
 Now we see that the possibility of committing a Type I error is roughly equivalent to a Type II error and to the probability of finding that the hypothesis is truly correct.
@@
 
-(ref:Likely) **The likelihood that likely hypotheses you test are accepted.** In this figure we see a scenario of likely hypotheses that are found to be correct approximately every other time.
+(ref:Likely) **The likelihood that likely hypotheses you test are accepted.** In this figure, we see a scenario of likely hypotheses that are found to be correct approximately every other time.
 
@@
 When the negative result is produced, a reasonable assumption is that a Type II error has been found, and that these scientists need to increase the power of their tests, inevitably increasing the cost of the experiment.
-But there are plenty of other ways that researchers could end up finding that their negative result is in fact positive, especially if this is what they are expecting.
-Simmons et al [-@simmons2011false] refer to this as the ‘researcher degrees of freedom’ in which researchers inadvertently or deliberately end up getting positive results because of the way in which they treat either their hypothesis or analyse their data.
+But there are plenty of other ways that researchers could end up finding that their negative result is, in fact, positive, especially if this is what they are expecting.
+Simmons et al.
+[-@simmons2011false] refer to this as the ‘researcher degrees of freedom’ in which researchers inadvertently or deliberately end up getting positive results because of the way in which they treat either their hypothesis or analyse their data.
 It is easy to see how a false scientific culture can be erected in these mistaken beliefs, and it is important to be aware of this possibility.
 
-### Increasing the statistical power
+## Increasing the statistical power
 
-In all of these figures the power of the analysis is set at 40%.
+In all of these figures, the power of the analysis is set at 40%.
 The statistical power of any analysis depends on the sample size (or number of replicates) you're able to use.
 Some research has suggested that in most ecological and evolutionary studies this is actually more like 20% [see references in @forstmeier2017detecting].
-What is important to notice in the next figure \@ref(fig:20PowerHyp) is that when we change the power of the analysis (in this case from 40% to 20%) we influence the proportion of Type II errors over finding that the hypothesis is correct.
-While the overall numbers of Type I errors does not change, if your analysis tells you to accept your alternative hypothesis, there is now a 1 in 5 chance (20%) that it will be a false positive (Type I error).
+What is important to notice in the next figure (Figure \@ref(fig:20PowerHyp)) is that when we change the power of the analysis (in this case from 40% to 20%) we influence the proportion of Type II errors over finding that the hypothesis is correct.
+While the overall numbers of Type I errors do not change, if your analysis tells you to accept your alternative hypothesis, there is now a 1 in 5 chance (20%) that it will be a false positive (Type I error).
 
@@
 Now we can see that the possibility of creating a Type II error is highest.
-Next the blue squares show us the chances that we find the hypothesis is truly correct.
+Next, the blue squares show us the chances that we find the hypothesis is truly correct.
 Lastly, there's only 11% chance of a Type I error.
@@
 
-What you should see when you look at these figures is that there is quite a high chance that we don't in fact correctly assign a true positive hypothesis.
+What you should see when you look at these figures is that there is quite a high chance that we don't correctly assign a true positive hypothesis.
 There's actually much more chance that we will commit a Type II error.
-Worse the more unlikely a hypothesis is, we are increasingly likely to commit the dreaded Type I error.
+Worse, the more unlikely a hypothesis is, we are increasingly likely to commit the dreaded Type I error.
 
-From the outset we should try to make sure that the hypotheses you are testing in your PhD are very likely to find positive results.
+From the outset, you should try to make sure that the hypotheses you are testing in your PhD are very likely to find positive results.
 In reality, this means that they are then iterations of hypotheses that are built on previous work.
@@
 
-In biological sciences we have more options than many scientific disciplines to investigate potentially false culture beliefs by using different model species, and looking at problems through different mechanisms.
-Presenting complimentary results from multiple lines of evidence within your thesis is an extremely powerful way to build evidence for your hypothesis.
-However, this requires creativity of experimental design as well as an open mind.
+In biological sciences, we have more options than many scientific disciplines to investigate potentially false cultural beliefs by using different model species, and looking at problems through different mechanisms.
+Presenting complementary results from multiple lines of evidence within your thesis is an extremely powerful way to build evidence for your hypothesis.
+However, this requires the creativity of experimental design as well as an open mind.
 It may mean that you have to draw upon the most fundamental of scientific abilities: to be comfortable with being wrong.
 It may also require dogged determination in terms of publishing negative results against the currently prevailing culture.
-As we shall see (in parts 4 and 5), publishing has another set of challenges.
 

### 2021-07-20 2.03-NoHyp.Rmd
Add files via upload
# After copy edits

@@
-## What happens if you don’t have a hypothesis? {#nohypothesis}
+# What happens if you don’t have a hypothesis? {#nohypothesis}
 
-So far we have concentrated on thesis chapters or manuscripts that have hypotheses to test.
-I have emphasised that there is a need to focus all of the sections of your writing to helping frame and understand the hypothesis (in the Introduction), explain your approach in collecting and testing the variables (in the Materials and Methods), provide the results of your tests (in the Results), and respond to the hypothesis (in the Discussion) together with placing it in context.
+So far, we have concentrated on thesis chapters or manuscripts that have [hypotheses](#hypothesis) to test.
+I have emphasised that there is a need to focus all of the sections of your writing to helping frame and understand the hypothesis (in the [Introduction](#introduction)), explain your approach in collecting and testing the variables (in the [Materials and Methods](#methods)), provide the results of your tests (in the [Results](#results)), and respond to the hypothesis (in the [Discussion](#discussion)) together with placing it in context.
 But what, I hear you cry, if you don’t have a hypothesis to test?
@@
 
-### Central problematic
+## Central problematic
 If you have a particular problem that your research is trying to solve, this may result in a new methodology or perspective rather than testing a hypothesis.
@@
 Hence, no matter what your central problem is, you still need to identify this and put it at the heart of your manuscript so that all of your sections address this point.
-Of course, there may be more than one, and if so you will need to clearly articulate each one and explain how they are related to each other.
+Of course, there may be more than one, and if so, you will need to clearly articulate each one and explain how they are related to each other.
 If you are unsure about how to move forward, the best thing to do is to look to the literature for other examples of what you’re trying to do.
@@
 
-### Does it matter that you don’t have a hypothesis?
+## Does it matter that you don’t have a hypothesis?
 Having a hypothesis makes your work stronger than if you only have a question or prediction.
@@
 
-### Avoid HARKing 
-Hypothesising After Results Known (HARKing) is the practice of creating your hypothesis once the results are in and you’ve done some preliminary data analyses to see what is significant [@forstmeier2017detecting].
+## Avoid HARKing 
+Hypothesising After Results are Known (HARKing) is the practice of creating your hypothesis once the results are in and you’ve done some preliminary data analyses to see what is significant [@forstmeier2017detecting].
 HARKing has become prevalent in science because of the confirmation bias by journals (only accepting papers that can statistically accept the alternative hypothesis), especially those with higher impact factor.
+Increasing chances of confirmation bias, or a [Type I error](#TypeI), is not desirable as you are more likely to accept the alternative hypothesis when it is not correct.
+Alternatively, you should seriously consider [preregistering your aims](#transparency) to remain transparent.

### 2021-07-20 2.04-BigIdea.Rmd
Add files via upload
# After copy edits

@@
 # What's the big idea? {#bigidea}
-In the previous chapters, I've talked about the importance of having a hypothesis, and building that hypothesis in a logical framework within the introduction (see part 3).
-The introduction serves to inform the reader about why this particular hypothesis was chosen, introducing both the response and determinate variables, as well as the presumed mechanism by which the hypothesis can be falsified (or upheld).
-
-## The funnel introduction 
-But so much of what you read doesn't have any clear hypothesis, and there doesn't seem to be any thought about building on bigger concepts or ideas.
-In this chapter, I suggest that this might not be true only of papers on herpetological invasions (my specialist area), but also of herpetology in general (although I concede that some areas, such as herpetological physiology are actually quite good).
-Further, I contend that using these wider hypotheses or theories would actually be good for the authors concerned, as it would likely garner them a wider audience.
-Moreover, a greater number of biologists might come to realise how valuable reptiles and amphibians are as models in biology.
+In the previous chapters, I've talked about the importance of having a hypothesis, and building that hypothesis in a logical framework within the [introduction](#introduction).
+The introduction serves to inform the reader about why this particular hypothesis was chosen, introducing both the dependent and independent (response and determinate) variables, as well as the presumed mechanism by which the hypothesis can be falsified (or upheld; Figure \@ref(fig:Introfunnel)).
 
@@
 Ask your advisor about relevant reviews of hypotheses in your area of biological sciences.
-The first is by Mark Velland [-@vellend2010conceptual] on theories in community ecology.
-The next is by Jane Catford and colleagues [-@catford2009reducing] on hypotheses in invasion biology, but I encourage you to look for more up to date versions.
-Lastly, consider these wide ranging papers about so-called 'laws' in biology and ecology [@dhar2010laws; @lawton1999are].
+For example, here are some theories and hypotheses in ecology [@vellend2010conceptual; @travassos-britto2021pragmatic; @catford2009reducing], while these wide ranging papers about so-called 'laws' in biology and ecology [@dhar2010laws; @lawton1999are].
 Each of these papers will give you a list of big ideas, together with the citations for seminal papers that have built them.
-You will note that many of these theories are very old with some dating back to Darwin.
+Clearly, you will need to find the papers that set out these big ideas in your own discipline.
+You will note that many of the theories are very old, with some dating back to Darwin.
 You can think of the way that such ideas are structured as a hierarchy of hypotheses (Figure \@ref(fig:JeschkeHagar61)).
@@
 
-(ref:JeschkeHagar61) **A hierarchy of hypothses can be used to help determine your Big Idea.** A useful structure for thinking about how hypotheses are structured was presented by @heger2018hierarchy in what they termed the _Hierarchy of Hypotheses_.
+(ref:JeschkeHagar61) **A hierarchy of hypotheses can be used to help determine your Big Idea.** A useful structure for thinking about how hypotheses are structured was presented by @heger2018hierarchy in what they termed the _Hierarchy of Hypotheses_.
 
@@
 
-**The take home message**
+**The take-home message**
 
 Reading the literature can really expand your mind and broaden your horizons.
-When undertaking a literature review [or when reviewing a paper], take the time to think about not only what has been tested, but what could have been.
+When undertaking a literature review, take the time to think about not only what has been tested, but what could have been.
 Make a list of theories and hypotheses in your own field and try to rank them in a hierarchy.

### 2021-07-20 2.05-paragraph.Rmd
Add files via upload
# After copy edits

@@
-## Writing a paragraph {#paragraph}
-
-In the following chapters, I will explain how to write a scientific paper in a formulaic style, as well as the advantages that this has over starting from scratch each time.
-The suggestion is that once, you sketch out your paper in a 'bullet-point' outline.
-This outline will give you an idea of the subject in each paragraph, as well as how the paragraphs fit in with each other.
-The other suggestion was to flesh out this outline by assigning relevant citations to help you remember those main ideas.
+# Writing a paragraph {#paragraph}
 
 Writing paragraphs has a skill all of its own, and the aim of this chapter is to go over the basics of how to put a paragraph together.
-Other people have also written about this, and you can read an example here.
 
 **This paragraph starts with a topic sentence which contains the principle idea that citations are needed for examples.** The following sentences put this idea into the context of your study.
-This should be based on examples from existing literature with the corresponding in-text citations provided (Smith et al 2017; Jones 2018).
+This should be based on examples from existing literature with the corresponding in-text citations provided (Smith et al.
+2017; Jones 2018).
 You can use examples in your paragraph to make your point.
@@
 
-In the paragraph above, the topic sentence is bold, the supporting sentences are normal font, and the clincher is in italics.
+In the paragraph above, the topic sentence is **bold**, the supporting sentences are normal font, and the clincher is in *italics*.
 
-### Topic sentence
+## Topic sentence
 A topic sentence allows the reader to understand quickly the idea/topic you are putting forward in the paragraph.
@@
 
-### Supporting sentences
+## Supporting sentences
 Supporting sentences convey all the relevant information to the reader.
@@
 If you are writing about differences, then state which way the difference is.
-For example, if bullfrogs are larger than leaf litter frogs, then say this.
-Telling your audience that something is different than something else only ends up leaving them guessing about the way in which the things are different.
+For example, if tulips are larger than lillies, then say this.
+Telling your audience that something is different than something else only ends up leaving them guessing about the directionality.
 There’s no point in drip dripping this information through.
@@
 
-Your supporting sentences are best built in the formula of an [argument](argument.html).
+Your supporting sentences are best built in the formula of an [argument](#argument).
 
-### The magic of threes
+## The magic of threes
 We can learn a lot about writing convincingly in science by borrowing from storytelling, and one of the greatest lessons in storytelling involves the magic of threes [@angler2020telling].
 
-In putting together your paragraph you can use the magic of threes to break your dialogue into three parts: thesis, antithesis and synthesis (a, b and a+b).
+In putting together your paragraph, you can use the magic of threes to break your dialogue into three parts: thesis, antithesis and synthesis (a, b and a+b).
 Think of it this way: there’s a standard explanation of how a phenomenon that you want to investigate interacts with the variable of your choice (thesis).
@@
 What if your list is two instead of three?
-Well then you’ll lose the magic, but maybe you can think it through it again?
+Well, then you’ll lose the magic, but maybe you can think it through it again?
 If you have two, then you normally have three by means of the combination of the two (a, b and a+b).
@@
 
-### Use an example
+## Use an example
 Examples are a very powerful way of conveying ideas in a short amount of space.
@@
 You should be able to do this in a sentence (or two), but if you’re tempted to go on, it’s probably not a good example.
-Because the first paragraph sets out ideas, it's unlikely that an example there will be a good idea.
+Because the first paragraph sets out ideas, it's unlikely that an example there will be a good idea there.
 
-### Avoid lists
+## Avoid lists
 I’m not a big fan of paragraphs which are simply a long list with little or no thought offered.
@@
 
-### The last sentence of the paragraph: the clincher
+## The last sentence of the paragraph: the clincher
 Once you’ve conveyed all of the information that you planned to impart in your outline, it’s time for the last sentence.
@@
 
-### Above all - read it!
+## Above all - read it!
 Your paragraph is not finished until you've read it.
-Reading is an essential part to writing that cannot be emphasised enough.
+Reading is an essential part of writing that cannot be emphasised enough.
 Does your paragraph say what you planned in your outline?
@@
 
-### How does the paragraph fit into the flow?
+## How does the paragraph fit into the flow?
 So now we’ve gone over the formula, it’s time to take a step back and look again at the paragraph in the context of your outline.
-Remember that the paragraph represents a single subject, but that it is still just part of the manuscript as a whole and you need that to flow from beginning to end.
+Remember that the paragraph represents a single subject but that it is still just part of the manuscript as a whole and you need that to flow from beginning to end.
 This means that it’s not enough to write each paragraph in isolation, but to think of the way in which they link together as a whole, especially when you are trying to make a logical argument.
 
-You might get a comment from me about missing logic in the argument or a lack of flow in a paragraph.
-This usually comes from not having a good outline (see below).
-You should be using the outline to provide the logic in the flow of your argument, flesh it out with the relevant references.
+You should be using the outline to provide the logic in the flow of your argument, fleshed out with the relevant references.
 If you already know this well, then writing the paragraph will just be a question of styling the outline.
@@
 
-### Seeding ideas
+## Seeding ideas
 The introduction sets out the established literature in order to put your study in context, but your discussion provides you with an opportunity to present new ideas, or to turn and twist existing ideas in a new way.
-Once you’ve got a good idea what these are, I like to seed the introduction with hints as to what these might be.
+Once you’ve got a good idea of what these are, I like to seed the introduction with hints as to what these might be.
 Sowing seeds early in a manuscript will provide the reader with hints as to where you are going.
@@
 
-### Breaking the rules
-Just as in the other instructions on formulaic paper writing, when writing a paragraph you shouldn’t feel totally constrained so that you can’t break the rules.
+## Breaking the rules
+Just as in the other instructions on formulaic paper writing, when writing a paragraph, you shouldn’t feel totally constrained so that you can’t break the rules.
 Breaking the rules can set you free, and much of what you read that really stands out will do this.
@@
 The idea of this chapter was to help you get started, not to communicate with those who are already writing great stuff.
-So if you're already great, don't break it by doing any of this!
+So if you're already great, don't break it!
 
-### Some common paragraph mistakes
+## Some common paragraph mistakes
 - Don't start your paragraph by linking back to the previous point. Start with your new topic sentence. If you aren't ready to change the topic, you aren't ready for a new paragraph.
@@
 Similarly, don't start with a trivial caveat.
-And don't start with a list or long string of citations.
+And don't start with a list or a long string of citations.
 - Don't let your paragraph get too long.
-It is said that 250 words is the threshold.
+It is said that 250 words are the threshold.
 If you haven't covered what you wanted to in your outline within 250 words, then you need to rethink your outline for that paragraph!
@@
 If you find yourself citing the same paper several times in the same paragraph, then there's something wrong with your outline, and the way that you are reading and citing papers (see below).
-Remember that what you write is about making your own argument from the literature, and not copying others arguments.
+Remember that what you write is about making your own argument from the literature, and not copying others' arguments.
 
@@
 
-- You should not start a sentence with an abbreviated genus. Instead write the genus name in full.
-- You should not start a sentence with a numeral. If you have to use the number as the first word, write it out. Otherwise, rearrange your sentence. 
+- You should not start a sentence with an abbreviated genus. Instead, write the genus name in full.
+- You should not start a sentence with a numeral. If you have to use the number as the first word, write it out. Otherwise, rearrange your sentence. .
 - Make sure that acronyms are spelt out on first use (this may be needed for each section)

### 2021-07-20 2.06-argument.Rmd
Add files via upload
# updated references with DOI and discussion table
Add files via upload
# After copy edits

@@
 Writing is not straightforward.
-Your objective is to communicate with a reader, someone who you’ve likely never met and will never meet.
+Your objective is to communicate with a reader, someone who you’ve likely never met.
 You need to communicate highly complex information.
@@
 The beginning of your introduction starts with the most general concepts in the context in which you are writing.
-The context depends on your audience, and this in turn relates to the particular journal that you are writing for.
+The context depends on your audience, and this, in turn, relates to the particular journal that you are writing for.
 Even if you are writing a thesis, you should pick out a target journal for each of the chapters.
@@
 
-A series of statements that introduce a starting premise, provide evidence for and against that premise (perhaps adding in an example that makes your point), point out what missing information would allow reaching a better understanding of said premise, and logically conclude that what you are doing is going to fill this gap.
+A series of statements that introduce a starting premise, provide evidence for and against that premise (perhaps adding in an example that makes your point), point out what missing information would allow reaching a better understanding of said premise, and logically conclude that what you are doing is going to help fill this gap.
 
@@
 
-The following science argumentation model (Table \@ref(tab:argument)) is modified from Cope et al [-@cope2013science].
+The following science argumentation model (Table \@ref(tab:argument)) is modified from Cope et al.
+[-@cope2013science].
 Try using this model to guide your writing of a paragraph of your introduction (or discussion) where you attempt a logical argument.
+For a more in depth treatment of using arguments in science, see Toulmin [-@toulmin2003uses].
 
-Table: (\#tab:argument) **A framework for a scientific argument.** Modified from Cope et al (2013), this model argument can be followed to provide as a way of getting started.
+Table: (\#tab:argument) **A framework for a scientific argument.** Modified from Cope et al. (2013), this model argument can be followed to provide as a way of getting started.
 
-| *Structure*  | *Explanation*  |
+| **Structure**  | **Explanation**  |
 | ------------ | -------------- |
@@
 | Claim 1 | One potential explanation or interpretation of the original idea.
-Evidence: Literature that agrees with this interpretation (could include an example) Reasoning: Your justification that links the evidence to the claim or interpretation |
+Evidence: Literature that agrees with this interpretation (could include an example) Reasoning: Your justification that links the evidence to the claim or interpretation.
+|
 | Claim 2 | Another potential explanation, of the previous interpretation of the original idea | *Evidence*: Literature that agrees with this interpretation (could include an example) *Reasoning*: Your justification that links the evidence to the claim or interpretation |
-| Counter - claims | Other possible interpretations or counter claims. |
+| Counter-claims | Other possible interpretations or counter-claims. |
 | Conclusions | Evaluation: your judgement on weighing up the evidence for the idea. |
@@
 I find that logical arguments develop best when I’m not sitting at my desk writing.
-They come when I’m reading other papers (especially those with well written arguments), taking a deep, hot bath, or when I’m out running in the mountains.
+They come when I’m reading other papers (especially those with well-written arguments), taking a deep, hot bath, or when I’m out running in the mountains.
 If you are more social, talk through your logical argument with others, perhaps over a cup of coffee.
@@
 Another sentence structure to consider is explaining cause and effect.
-For practical wording examplesof both of these classic arguments, [see part 2](phrasebank.html).
+For practical wording examples of both of these classic arguments, [see part 2](phrasebank.html).
 
 ## Arguments in Latin
-From time to time you may come across comments about arguments that you’ve made in your work made in Latin.
+From time to time you may come across comments in Latin about arguments that you’ve made in your work.
 Why do people use Latin and not simply say it in English?
@@
 
-Here I have provided some of these arguments, and the reasons why they may appear in your thesis in writing biological sciences.
+Here I have provided some of these arguments, and the reasons why they may appear in your thesis in biological sciences.
 
@@
 
-_Non sequitur_ - “It does not follow”: suggests that one statement in your argument is not linked to the next.
-This is also known as a logical fallacy, a formal fallacy or a deductive fallacy (for a full list of fallacy statements and their explanations, see here).
+_Non sequitur_ - “It does not follow”: suggests that one statement in your argument is not linked to the next. This is also known as a logical fallacy, a formal fallacy or a deductive fallacy.
 
@@
 
-**Correlation does not equal causation** - That two events or variables are related when in fact they are co-incidental.
+**Correlation does not equal causation** - That two events or variables are related when in fact, they are co-incidental.
 
@@
 This is also known as a Type II error (false negative), where the lack of finding a significant relationship does not lead to a conclusion that the relationship does not exist.
-Later we discuss how a [lack of finding of a significant result](transparency.html) leads to confirmation bias in science.
+Later we discuss how a [lack of finding of a significant result](transparency.html) leads to [confirmation bias](#TypeI) in science.
 
-**Observational selection** -"The root of all superstition is that men observe when a thing hits, but not when it misses." Francis Bacon [-@bacon1620novum].
+**Observational selection** - "The root of all superstition is that men observe when a thing hits, but not when it misses." Francis Bacon [-@bacon1620novum].
 This refers to the cherry picking of studies that support your hypothesis while omitting those that do not support it.
@@
 * Did you understand it?
-* Was it written in a conventional style (as in the table (\#tab:argument))
+* Was it written in a conventional style (as in Table \@ref(tab:argument))
    + If not, how was the style broken with and did this improve or detract from the understanding?
@@
 However, I encourage you to learn and experiment with writing styles as you become more experienced.
-[Critical reading](criticalreading.html) is one of the best ways to learn about alternative writing styles (see here): reading is probably the best way of improving your writing.
+[Critical reading](criticalreading.html) is one of the best ways to learn about alternative writing styles: reading is probably the best way of improving your writing.
 Alternatively, and especially if the above hasn’t clicked with you, you can read more about writing a scientific argument (see for example [here](https://msu.edu/course/iah/231b/gifford/sciaargs.htm)).

### 2021-07-20 2.07-storytelling.Rmd
Add files via upload
# After some small edits over the week
Add files via upload
# After copy edits

@@
-## Storytelling in science? {#storytelling}
+# Storytelling in science? {#storytelling}
 
@@
 
-“No storytelling” is a comment that I sometimes make when reading drafts of manuscripts, chapters, and even when editing for journals.
+'No storytelling' is a comment that I sometimes make when reading drafts of manuscripts, chapters, and even when editing for journals.
 What do I mean by this?
 Stories are deterministic.
-That is to say that the storyteller has an end in mind when they start telling the story, and the telling is a way to get to their goal.
+That is to say that the storyteller has an end in mind when they start telling the story (i.e.
+teleological; see Figure \@ref(fig:Storytelling)), and the telling is a way to get to their goal.
 A story that’s ‘pointless’ will frustrate the audience and won’t engender them to listen to that storyteller again.
@@
 
-### The danger of a teleological argument
+## The danger of a teleological argument
 
@@
 When I was doing my PhD I was very frustrated as I had the impression that my advisor knew what result he wanted and designed the study to show it.
-What probably happened is that he was displaying “motivated reasoning”, but the result is known as “confirmation bias”.
+What probably happened is that he was displaying 'motivated reasoning', but the result is known as '[confirmation bias](#TypeI)'.
 When an experiment failed to meet the expected result, he declared that it had failed.
@@
 And if it encounters a larger number of more powerful countervailing examples, it either fails to notice them, or disregards them, or makes fine distinctions to dismiss and reject them, and all this with much dangerous prejudice, to preserve the authority of its first conceptions.”__
+>
 >Francis Bacon [-@bacon1620novum]
@@
 
-### SCience is the very opposite of storytelling. Or is it?
-I often tend to think of the answer to the question as the goal of the study.
+## Science is the very opposite of storytelling. Or is it?
+I often tend to think of the answer to the right question as the goal of the study.
 That I don’t know what the answer is, doesn’t spoil the story for me.
@@
 
-A good example of this is the study published by Becker et al [-@becker2018extreme].
+A good example of this is the study published by Becker et al.
+[-@becker2018extreme].
 Francois found a (spookily) strong relationship between probability that Rose’s dwarf toadlets survive and local rainfall in the breeding season.
@@
 It wasn’t until we made the connection between the fact that increased rainfall during the breeding season meant that the toadlets spent more time in puddles, that we understood why their survival was decreasing.
-As the toadlets aren’t feeding during this time they lose weight, and also expose themselves to increased predation pressure.
+As the toadlets aren’t feeding during this time they lose mass, and also expose themselves to increased predation.
 In a dry year, the toadlets will head back to their subterranean refuges much earlier and continue to pursue ‘safer’ feeding and hiding habits.
-While we might intuitively feel that a ‘safe toadlet’ is a better life-strategy, reduced rainfall means reduced reproduction, and so results in a failure for toads that don’t manage to pass on their genes that year.
+While we might intuitively feel that a ‘safe toadlet’ is a better life strategy, reduced rainfall means reduced reproduction, and so results in a 'failure' for toads that don’t manage to pass on their genes that year.
 The result is a variable life-history with the weather, something that was previously unknown.
@@
 
-But the scientific paper that was written about this ‘story’ [@becker2018extreme] doesn’t have a narrative style, and doesn’t fit the description of a narrative that we discussed above.
-Instead, it follows the formula that we set out below: Introduction, Materials and Methods, Results and Discussion.
+But the scientific paper that was written about this ‘story’ [@becker2018extreme] doesn’t have a narrative style, and doesn’t fit the description of a narrative that is discussed above.
+Instead, it follows the formula that is set out below: **Introduction, Materials and Methods, Results and Discussion**.
 This style does not treat the experiment as a complication on the way to the story’s goal.
@@
 There is a role for stories when communicating science as this increases interest, facilitates understanding, and enhances memory.
-This is particularly true when communicating science to the public and making it more accessible, but it also applies to interactions between scientists, for example at conferences.
+This is particularly true when communicating science to the public and making it more accessible, but it also applies to interactions between scientists, for example, at conferences.
 The [presentations](#proposal-presentation.html) that tell a story and entertain are those we tend to remember.
@@
 
-### Can storytelling help memory of science?
+## Can storytelling help memory of science?
 Some argue that it can, and should be an integral part of scientific writing.

### 2021-07-20 2.08-citations.Rmd
Add files via upload
# After copy edits

@@
 > Isaac Newton (in a letter to Robert Hooke in 1675) wrote: 
+>
 > **"If I have seen further it is by standing on the shoulders of Giants."**
 
-Although the phrase is often attributed to Isaac Newton, it turns out that it was well accepted in Newton's time and has been traced back to the 12th century.
+Although the phrase is often attributed to Isaac Newton, it turns out that it was a well accepted phrase in Newton's time and has been traced back to the 12th century.
 
-Essentially, standing on the shoulders of giants (Figures \@ref(fig:Googlescholar) is a recognition that all research is built upon research that has gone before it, and this is the basis for citations in the text of scientific papers.
-Patrick Dunleavy [@dunleavy2017citations] argues that citations are required to meet 7 criteria with respect to academic writing.
+Essentially, standing on the shoulders of giants (Figure \@ref(fig:Googlescholar) is a recognition that all research is built upon research that has gone before it, and this is the basis for citations in the text of scientific papers.
+Patrick Dunleavy [-@dunleavy2017citations] argues that citations are required to meet 7 criteria with respect to academic writing.
 Others have suggested that there are more criteria [see @harwood2009interview], and that these are a fundamental premise of academic work.
 But it was not always thus.
-If we dwell a moment longer, we can look back at the World's longest running scientific journal: *Philosophical Transactions of the Royal Society.*
+If we dwell a moment longer, we can look back at the World's longest-running scientific journal: *Philosophical Transactions of the Royal Society.*
 
-You can download the first of those papers of 1665 and note that there are no citations (other than to books or letters) because there were no previously published articles from which to draw.
-However, even then authors noted that ideas came from previous authors and we can regularly find acknowledgements that refer back to Aristotle (~350 BC).
+You can download the first of those papers of 1665 and note that there are no citations (other than to books or letters) because there were no previously published articles from which to draw (Figure \@ref(fig:Boyle)).
+However, even then, authors noted that ideas came from previous authors and we can regularly find acknowledgements that refer back to Aristotle (~350 BC).
 
@@
 Citations need to be used to back up any statements of fact that you make in your PhD chapter.
-Any examples need citations for where the examples came from, and when you make [arguments](argument.html), you add credibility to your arguments from each side by citing that these are from published works.
+Any examples need citations for where the examples came from, and when you make [arguments](#argument), you add credibility to your arguments from each side by citing that these are from published works.
 
@@
 I've seen many students try to write their text first and then only later fill in citations.
-I've had students hand in thesis chapters and manuscripts that have [ref] written strategically at the end of many sentences.
-I've even had the experience of doing this myself when I was a early career researcher (I probably still have those manuscripts somewhere because it's a sure way of never getting anything finished).
+I've had students hand in thesis chapters and manuscripts that have '[ref]' written strategically at the end of many sentences.
+I've even had the experience of doing this myself when I was an early career researcher (I probably still have those manuscripts somewhere because it's a sure way of never getting anything finished).
 I suggest that you **do not do this.** The reason is that if you write first and try to insert citations later, you are going to spend lots of time looking for the citation that fits your text.
@@
 Indeed, he knows it all so well that he already knows what citations fit what statements without needing to look them up.
-However, later on in the video Pat admits that even he sometimes gets it wrong, and that means that he ends up having to rewrite text when he finds the publication that he was thinking of and notes that what they actually found did not fit what he wrote.
+However, later on in the video, Pat admits that even he sometimes gets it wrong, and that means that he ends up having to rewrite text when he finds the publication that he was thinking of and notes that what they actually found did not fit what he wrote.
 Although I can identify with what Pat is saying, I would find this re-writing far too time consuming to take this approach myself.
 I am familiar with most of the pertinent literature in my specialist area, but I'd rather check things before I start writing than leave it until later.
-I find that my memory is not only wrong about what other people have found, but also about what's in my own work.
+I find that my memory is not only wrong about what other people have found but also about what's in my own work.
 Therefore, I'd suggest that while you may aspire to write like Pat later on in your career, right now it'd be better for you to plan your writing with citations in mind, and cite while you write (not afterwards).
@@
 
-## Citation Styles
+## Citation styles
 
@@
 
-### Vancouver Style
+### Vancouver style
 
-Vancouver style dates back to medical journals in the 1970s and refer to conventions to avoid fraud in medical sciences.
-You are probably familiar with it even if you don't know the name.
+Vancouver style dates back to medical journals in the 1970s and refers to conventions to avoid fraud in medical sciences.
+You are probably familiar with it, even if you don't know the name.
 It is the convention that uses small superscript numbers to denote the place of citations in the main text, and the full reference given at the end of the document is provided in order of the citations (and their numbers), instead of alphabetical order.
@@
 The diminutive numbers take up less space, allowing the reader and writer to concentrate more on their prose.
-On the other hand, if you are familiar with the literature, or you are interested in learning more about the literature, you’ll find that the little numbers take much more time to cross reference with the list at the end, than does Harvard style.
+On the other hand, if you are familiar with the literature, or you are interested in learning more about the literature, you’ll find that the little numbers take much more time to cross-reference with the list at the end, than does Harvard style.
 Reference strings in Vancouver style can either be separate numbers, or numbers listed as a series.
@@
 
-The corresponding references in [Vancouver](vancouver.html) format are in the section on [references](references.html).
+The corresponding references in [Vancouver](#vancouver) format are in the section on [references](#references).
 
-### Harvard Style
+### Harvard Style {#Hstyle}
 
 The standard way is to use the name and year in parentheses at the end of the statement to which the citation is relevant.
-Names of three or more authors are frequently reduced to “et al.” (often written in italics) which is short for the Latin et alia, meaning “and others”.
+Names of three or more authors are frequently reduced to 'et al.' (often written in italics: _et al._) which is short for the Latin _et alia_, meaning 'and others'.
 References strings in Harvard style can go on for an entire line of text or more.
-Some journals have rules that mean on first mention the citation should have all authors (up to a certain number).
+Some journals have rules that mean on the first mention, the citation should have all authors (up to a certain number).
 This can become tedious when citations start taking up more space than text.
@@
 
-You can find the corresponding references in [Harvard](harvard.html) format in the [references](references.html) section.
+You can find the corresponding references in [Harvard](#harvard) format in the [references](#references) section.
 
@@
 That doesn’t mean removing the citations themselves, but it may mean re-writing some sentences.
-Simply changing your reference manager to use the different referencing style will only be the start of your work.
+Simply changing your reference manager to use a different referencing style will only be the start of your work.
 
@@
 
-The impact of all invasive amphibians is similar to that of invasive birds and mammals [@measey2016global].
+>The impact of all invasive amphibians is similar to that of invasive birds and mammals [@measey2016global].
 
 You’ll see that sometimes the names are brought into the sentence and become the central agents of the text.
-The phrase et alii (often abbreviated to et al.), literally means "and others" referring to the other authors (typically used when there are more than two - but check the journal format!).
 
-Sometimes, instead of "et al." you can write "and colleagues" or "and others". This is something to do occasionally when you are looking to diversify some text. Don't over do it though.
+Sometimes, instead of 'et al.' you can write 'and colleagues' or 'and others'. This is something to do occasionally when you are looking to diversify some text. Don't over do it though.
 
@@
 A citation string is a list of two or more citations that all assert the statement given.
-When you provide a string of citations you will need to decide which one comes first.
+When you provide a string of citations, you will need to decide which one comes first.
 Normally, this is given in the order of precedence: the oldest citation comes first, the youngest last.
@@
 
-Taxonomists have special rules for this, and I'll do it in another chapter (see section on [Scientific names and taxonomic authorities](taxonomy.html)).
+Taxonomists have special rules for this, and this will be explained in another chapter (see section on [Scientific names and taxonomic authorities](#taxonomy)).
 These are not the same as regular citations (because they don't appear in the literature cited and you don't have to have read the descriptions), and only some journals ask for them.
@@
 For example, in the introduction of their paper, Measey et al.
-[-@measey2016global] make comments on amphibian decline.
-
->Amphibian populations are currently declining across the globe [@wake2008are; @collins2009extinction; @pimm2014biodiversity] and alien amphibians are at least partially driving these declines through competition [@kupferberg1997bullfrog], hybridization [@dufresnes2015introgressive] and introduction of novel pathogens [@berger1999chytrid; @daszak2003infectious; @lamarca2005catastrophic; @martel2013batrachochytrium].
+[-@measey2016global] make comments on amphibian decline (see [above](#Hstyle)).
 
@@
 Another common mistake is to forget which paper has which information.
-You can try to make sure that you don’t do this by taking better notes or a more accurate plan.
+You can try to make sure that you don’t do this by taking better notes or write a more accurate plan.
 And do the citations before or as you write, not afterwards!
@@
 My suggestion would be to avoid such sources anyway (with particular exceptions – see below), unless it is really important that you include it.
-Other examples of texts to avoid are: text books (use the original paper instead), newspapers or magazines, blog posts, facebook pages (or other social media sites), predatory journals (or any non-peer reviewed article).
+Other examples of texts to avoid are: text books (use the original paper instead), newspapers or magazines, blog posts, Facebook pages (or other social media sites), predatory journals (or any non-peer-reviewed article).
 These are all examples of __grey literature__.
 
-There are of course exceptions. When you are writing about social media sites, or newspaper coverage, you will want to cite those sources.
+There are, of course, exceptions. When you are writing about social media sites, or newspaper coverage, you will want to cite those sources.
 
@@
 Most of what's written in a guidebook has already been documented in the scientific literature, but guide-books generally don't provide sources for the information that they provide.
-Thus, it's easy to look up and find something in a guide-book, but suggests (to me) that you haven't spent enough time or effort reviewing the literature.
+Thus, it's easy to look up and find something in a guide-book but suggests (to me) that you haven't spent enough time or effort reading the literature.
 Some guide-books are excellent and the authors have gone to a lot of trouble to incorporate original data and observations.
@@
 
-One exception to the not citing websites relevant to our work is the IUCN Red List.
+One exception to not citing websites is the IUCN Red List.
 Note that all entries on this site now have DOIs, and this might be a good guide for what is available to cite.
-The Digital Object Identifier ([DOI](doi.html)) is very useful as it means that there is a consistent record of that version.
-Otherwise, you could cite any website (even this one) and then the owner can go and change the site and it no longer says what you thought it did.
+The Digital Object Identifier ([DOI](#DOI)) is very useful as it means that there is a consistent record of that version.
+Otherwise, you could cite any website and then the owner can go and change the site and it no longer says what you thought it did.
 The DOI removes this problem; there will always be an archived version with that particular DOI.
@@
 
-The primary literature are studies or experiments that are done in order to test a hypothesis.
+The primary literature consists of studies or experiments that are done in order to test a hypothesis.
 Secondary literature includes reviews, syntheses or meta-analyses.
@@
 
-## What is primacy and why does it matter?
+## What is rule of priority and why does it matter?
 
-Primacy in the literature is the paper/s where the authors provide the original idea or concept that you are citing.
-It is important in that you should give credit to original ideas over those who copy or simply repeat them, or even those who review them.
-Sometimes primacy is less important, especially if the concept is well known and has changed substantially.
+The rule of priority (or primacy) in the literature is that those paper/s where the authors first provide the original idea or concept, or evidence for this are the ones that are cited [@strevens2003role].
+Papers that are published afterwards, are simply seen as repeating the same story as those who published first, even if the time-lines of the research itself don't reflect this.
+The rule of priority makes sense, in that we have an obligation to name the shoulders on which we stand (Figures \@ref(fig:Googlescholar).
+Thus, it is important in that you should give credit to original ideas over those who copy or simply repeat them, or even those who review them.
+Tracing back the origins of ideas makes for interesting reading and a deeper understanding of your subject.
+
+Sometimes primacy is less important, especially if the concept is well known and/or has changed substantially.
 Then you should cite the most recent work that shows the ideas that you are wanting to show.
-Other times, primacy is important - especially when few have built on ideas or concepts since they were put forward.
-If in doubt place a citation to the original study and the most recent study, then flag this using a comment for your advisor to consider.
+Other times, primacy is more important - especially when few have built on ideas or concepts since they were put forward, or in the [taxonomic literature](taxonomy.html).
+If in doubt, place a citation to the original study and the most recent study, then flag this using a comment for your advisor to consider.
 The important point here is for you to give credit where it's due and not to overlook those who put in the hard work to publish original content.
 
-## How many citations is enough?
+Some have argued that the rule of priority is of benefit to science as it drives competition and this spurs creativity.
+On the other hand, the winner-takes-all approach of being first with something drives secrecy and dishonesty, and does not genuinely reflect the team effort that is the [science project](lifescientific.html) [@casadevall2012reforming].
+
+## How many citations are enough?
 
@@
 
-If you are looking for multiple examples of a statement that you’ve made and there are many possible, I’d suggest that you aim to produce three.
-Make sure that you use a suffix (like “e.g.”) to show that you are aware that these are examples of a widely reported phenomenon.
+If you are looking for multiple examples of citations for a statement that you’ve made and there are many possibilities, I’d suggest that you aim to produce three.
+Make sure that you use a suffix (like 'e.g.') to show that you are aware that these are examples of a widely reported phenomenon.
 You can choose these as you like, but may want to consider using what you consider to be ‘the best’ examples, and/or references that you are planning to use elsewhere in your paper/chapter.
-This can drive the total number down considerably, and helps to keep the citations more relevant to your work.
+This can drive the total number of citations down considerably, and helps to keep the citations more relevant to your work.
 
@@
 In many cases (such as with your thesis work), your own publications are likely to be more relevant to methodology and subject matter than much of the other work that is out there.
-However, if you’ve previously published on termite fungus and now you’re publishing on frog toes, it’s unlikely to be relevant.
+However, if you’ve previously published on termite fungus and now you’re publishing on rat toes, it’s unlikely to be relevant.
 Choosing citations should include your comfort with being transparent.
@@
 They claim that self-citations in the Natural Sciences run at 33% [@vanraan2007scoping], which provides you with a nice idea of what is acceptable.
-Remember that this would include not only your citations to your articles, but all citations to your co-authors articles too.
+Remember that this would include not only your citations to your articles, but all citations to your co-authors' articles too.
 
@@
 Obviously, the inverse is also true, so beware of the politics of citing.
-However, the most important points have already been raised, but will be tackled [in detail later](impactfactor.html).
+However, the most important points have already been raised.
 The study must be relevant and appropriate before it gets included as a citation in your work.
@@
 Papers in journals with high impact factors are more likely to be cited because their contents are already thought to be of interest to a wide range of people.
-Indeed, there is evidence that identical statements published in many journals, those with higher Impact Factors are cited more [@perneger2010citation].
-That is something that the editors of that journal will have considered even before sending the manuscript to review.
+Indeed, there is evidence that for identical statements published in many journals, those with higher Impact Factors are cited more [@perneger2010citation].
 Sometimes (but not always) the impact factor of the journal can be an indication of the quality of the study.

### 2021-07-20 2.09-databases.Rmd
Add files via upload
# After copy edits

@@
 
-Databases of scientific literature have been around since the 1960s when the number of journals and papers started increasing so much that scholars simply couldn’t keep up.
+Databases of scientific literature have been around since the 1960s, when the number of journals and papers started increasing so much that scholars simply couldn’t keep up.
 Quite simply, you should use one or more of these databases to look for relevant literature to read.
-The first database, and what continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of selected journals (titles, abstracts, authors and other metadata), but also the citations that each paper used.
-As we have seen in the last chapter, citations record the way in which people find the literature useful, and are therefore a highly effective way of finding related and pertinent literature in your field.
+The first database, and one that continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of selected journals (titles, abstracts, authors and other metadata), but also the citations that each paper used.
+As we have seen in the last chapter, citations record the way in which people find the literature useful, and are therefore, a highly effective way of finding related and pertinent literature in your field.
 
@@
 If you don’t have this, then check with your advisor.
-Often, this key literature is cited in the advert for the PhD, the grant application that funded it, or if this is a topic that your advisor has already conducted some research on, then their own publications may include part of the key literature.
+Sometimes, this key literature is cited in the advert for the PhD, the grant application that funded it, or if this is a topic that your advisor has already conducted some research on, then their own publications may include part of the key literature.
 If you are not familiar with your advisor’s (including co-advisor’s where appropriate) papers, then a great first step is to search for them on a database and read through titles and abstracts that are relevant to your research topic.
@@
 
-## Keywords
+## Keywords {#keywords}
 
@@
 Any keywords that they share are going to be important for you.
-Make a note of all relevant keywords for your searches, as ultimately this will help you when you need to determine the keywords for [your own chapters](title.html).
+Make a note of all relevant keywords for your searches, as ultimately, this will help you when you need to determine the keywords for [your own chapters](title.html).
 
@@
 Following our example above, the search term *performance AND reptiles NOT snakes* should give you all performance papers on lizards, but you may want to exclude turtles and crocodilians as well if they have a lot of hits.
+*NOT* is a powerful operator and needs to be used with caution, as it could exclude some relevant hits.
+In our example here, an item about interactions between snakes and lizards would be excluded.
 
@@
 You can also use the term NEAR to indicate that two terms need to occur within a specified number of words.
-For example, performance NEAR/5 reptile will capture items where these terms occur within 5 words of each other, but no greater.
-This might be useful to exclude some papers that mention performance and reptiles in the abstract, but are not about reptile performance.
+For example, *performance* NEAR/5 *reptile* will capture items where these terms occur within 5 words of each other, but no greater.
+This might be useful to exclude some papers that mention performance and reptiles in the abstract but are not about reptile performance.
 
 Once you start putting together search strings you will need to use brackets to make sure that the correct terms are grouped.
-For example, performance AND (lizards OR geckos) will get what we described above, whereas (performance AND lizards) OR geckos will give you all papers on lizards and performance plus all papers on geckos (not what we wanted).
+For example, *performance* AND (*lizards* OR *geckos*) will get what we described above, whereas (*performance* AND *lizards*) OR *geckos* will give you all papers on lizards and performance plus all papers on geckos (not what we wanted).
 
@@
 - Use ? when you want to search for any single letter at that place in a word. For example, _sterili?e_ will capture words with sterilize and sterilise.
-- Use $ when there is more than a single letter that might change within a word, or the omission of a letter. For example, _colo$r_ will return both US and UK spelling, color and colour, respectively.
+- Use $ when there is more than a single letter that might change within a word, or the omission of a letter.
+For example, _colo$r_ will return both US and UK spelling, color and colour, respectively.
+Note that Scopus doesn't use the $ symbol as a wildcard: for this example you can use ?
+instead.
 
@@
 Beware that these regularly come with errors, and that the amount of errors is related to your database source (Scopus and WoS are much better than Google Scholar).
-This means that after importing the refrences, you will need to go back to your reference manager and edit them or they will come out incorrectly when you use them.
+This means that after importing the references, you will need to go back to your reference manager and edit them or they will come out incorrectly when you use them.
 Of particular note to biologists is the need to have all species names in [italics](taxonomy.html), and you'll need to do this yourself.
@@
 
-In addition to using keywords, you should search citations of your key references.
-Typically, when you start reviewing the literature in your specific area you will have a few key references that have been given to you by your advisor.
+In addition to using [keywords](#keywords), you should search citations of your key references.
+Typically, when you start reviewing the literature in your specific area, you will have a few key references that have been given to you by your advisor.
 Having read these, you will already know that they cite other works of interest, and you should follow up to look at all that seem relevant.
@@
 Indeed, given the prolific ever increasing nature of scientific publications, it might be the only way to keep on top of what is happening.
-Some literature databases allow you to receive alerts if a paper is cited, and you should set this up for all of your key citations, and perhaps for some of the most prominent authors in your field.
+Some literature databases allow you to receive alerts if a paper is cited (e.g.
+Google Scholar), and you should set this up for all of your key citations, and perhaps for some of the most prominent authors in your field.
 
@@
 
-The above section is a brief guide, and your library is likely to have in depth guides on how to use the specific databases that they house.
+The above section is a brief guide, and your library is likely to have in-depth guides on how to use the specific databases that they house.
 They may also provide training sessions, or online videos to watch.
@@
 This can be used to assess not simply the productivity of a scientist (i.e.
-number of publications), but their usefulness by others (i.e.
+number of publications), but their usefulness to others (i.e.
 number of citations).
@@
 Clearly, the Science Citation Index cost some money to put together, and universities paid annually to receive the books, and then the CDs and finally to access the online versions.
-They were not for profit organisations, but eventually they were bought out by the evil profit making parasites of science.
-Yes, publishers saw the dependency that scientists have on a resource and decided to own it for themselves and then radically increase the price to access it.
+They were a not for profit organisation, but eventually they were bought out.
 
@@
 
-In the life sciences, there are three prominent databases used: Google Scholar, Web of Science (owned by Clarivate Analytics aka Thomson Reuters) and Scopus (owned by Elsevier).
-But new players are entering the market all the time, and two of the newest are Dimensions from Digital Science and Microsoft Academic.
+In the life sciences, there are three prominent databases used: Google Scholar, Web of Science and Scopus.
+But new players are entering the market all the time, and two of the newest are Dimensions and Microsoft Academic.
 
@@
 
-Once you have found a scientist who works within your field and produces influential work that you are keen to read, then you may want to search for all of their other papers to determine whether there is more literature that you have missed.
+Once you have found a scientist who works within your field and produces influential work that you are keen to read, then you may want to search for all of their other papers to determine whether there is more literature that you might have missed.
 The databases make such searches easy, but you will need the (correct) spelling of your scientist's name, and also to be aware that they may have used different iterations of their name throughout their career.
@@
 While Dave Thompson could always add his middle initial to specify his work, as time went by there were more and more scientists publishing as D J Thompson.
-This made it more difficult for administrators and scientits alike.
-This has begun to be a bigger global problem, such that the databases have begun to use profiles of particular scientists that they can 'own' and curate.
-Each database has its own way of doing this, and it can become rather a chore so that not many scientists have their own profiles, which makes it a problem particularly for those with common names.
+This made it more difficult for administrators and scientists alike.
+This has begun to be a bigger global problem, such that the databases have begun to use profiles of particular scientists that individuals can 'own' and curate.
+Each database has its own way of doing this, and it can become rather a chore so that not many scientists have their own profiles, which makes it a problem, particularly for those with common names.
 
-The [ORCID](orcid.html) number has been errected to provide a common platform for [authors to curate](title.html), and now some journals and funding authorities won't allow you to submit or apply without one.
+The [ORCID](orcid.html) number has been erected to provide a common platform for [authors to curate](title.html), and now some journals and funding authorities won't allow you to submit or apply without one.
 
@@
 
-Well it seems that there are some people who have already thought about this, and come up with a good idea of exactly what's different.
-Martín-Martín et al [-@martin-martin2018google] have done a great job of analysing all this stuff from some 2.5 million citations.
-What they found inspired me to write this chapter, in which I've focussed on the results from the Life-Sciences.
+Well, it seems that there are some people who have already thought about this, and come up with a good idea of exactly what's different.
+Martín-Martín et al.
+[-@martin-martin2018google] have done a great job of analysing some 2.5 million citations.
+What they found inspired me to write this section, in which I've focussed on the results from the Life-Sciences.
 But I encourage you to go read the article for yourself.
 
-I have been known to take the odd peak at my Google Scholar profile over the year, and see how the citations are accumulating.
+I have been known to take the odd peek at my Google Scholar profile over the year, and see how the citations are accumulating.
 I rarely check on WoS or Scopus, because it's a bit of a faff getting signed in and doing the search.
 Plus it looks so much smaller when one is habituated to seeing those double digits in GS!
-However, I've always been a bit uneasy about citing my GS citation rate, H-index or i10 (among others that they give) as I've never really known what all that extra represents.
+However, I've always been a bit uneasy about citing my GS citation rate, H-index or i10 (among other mertics that they give) as I've never really known what all that extra represents.
 Something grey and unseemly?
@@
 To directly compare your citation metrics from different platforms, and also to get more unbiased metrics than these platforms provide, consider using the software Publish or Perish [@harzing2007publish].
-This also provides a great platform on which to harvest data from Google Scholar into a database, or put it directly into another database, like EndNote, Zotero or Mendelay.
+This also provides a great platform on which to harvest data from Google Scholar into a database, or put it directly into another database, like EndNote, Zotero or Mendeley.
 
@@
 
-According to Martín-Martín et al [-@martin-martin2018google] The GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus.
+According to Martín-Martín et al.
+[-@martin-martin2018google], the GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus.
 If you deviate from this with a higher score, you can give yourself a pat on the back for having work that's reaching more people in more parts of the world.

### 2021-07-20 2.10-referencesortware.Rmd
Add files via upload
# After copy edits

@@
 Because published papers play such a fundamental role in your [reading and comprehension](critical.html) of a subject, and [citations](citations.html) are pivotal in chapter or paper writing, putting these elements together in a good reference manager makes a lot of sense.
-This is a piece of software in which you can store and view all of your papers (in pdf format or via links), make notes on them, and use them to make citations and build bibliographies.
+A reference manager is a piece of software in which you can store and view all of your papers (in pdf format or via links), make notes on them, and use them to make citations and build bibliographies.
 Unsurprisingly, they become something that you may well use a lot during the course of your PhD, and it is worth thinking carefully about which one you want to use.
@@
 There are quite a lot of different software packages out there to help you do this (e.g.
-[Endnote](https://endnote.com/), [Mendelay](https://www.mendeley.com/guides/desktop), [Zotero](https://www.zotero.org/) - see more about these [here](https://subjectguides.library.american.edu/c.php?g=479020&p=3323781)).
-These are a mixed blessing.
+[Endnote](https://endnote.com/), [Mendeley](https://www.mendeley.com/guides/desktop), [Zotero](https://www.zotero.org/) - see more about these [here](https://subjectguides.library.american.edu/c.php?g=479020&p=3323781)).
+These packages are a mixed blessing.
 Firstly, most of them are not free, so unless your university subscribes to their providers you would have to pay for the privilege of using them.

### 2021-07-20 2.10a-sortware.Rmd
Add files via upload
# After copy edits

@@
 If there's any good time to start with a new word processing package it will be before you start your PhD.
-If you do decide to go with a new writing package then my suggestion would be to choose open source software like [Rmarkdown @xie2018r].
+If you do decide to go with a new writing package then my suggestion would be to choose open source software like [R Markdown @xie2018r].
 More on these later.
 
-Another important consideration is whether or not your chosen word processing package will allow seemless integration with your [reference manager](referencesoftware.html).
+Another important consideration is whether or not your chosen word processing package will allow seamless integration with your [reference manager](referencesoftware.html).
 Your lab may have a communal repository of references, and so this may then constrain what you are comfortable using.
@@
 These certainly have advantages like being able to get a synonym at the click of a mouse (Figure \@ref(fig:synonym)).
-My best suggestion is that you do not save your files in their default ".docx" format.
-Instead make sure that all of your files are saved as .rtf or html.
+My best advice is that you do not save your files in their default ".docx" format.
+Instead, make sure that all of your files are saved as .rtf, .txt or .html.
 This should make little difference to your opening and editing but will mean that all of your texts are forward compatible with other kinds of software.
@@
 
-Many academics are turning to latex in order to write their manuscripts as this gives a lot of possibility to insert links, formulae, references, images, etc.
+Many academics are turning to LaTeX in order to write their manuscripts as this gives a lot of possibilities to insert links, formulae, references, images, etc.
 in a very simple way.
-Once you are ready to print a latex document it can be shared very easily as a PDF.
+Once you are ready to print a LaTeX document it can be shared very easily as a PDF.
 
-Be aware that the problems may come not with what you want but with how your advisor and or co-authors are comfortable with.
+Be aware that the problems may come not with what you want but with how your advisor and or co-authors are comfortable.
 I find that commenting on a PDF takes more time than doing the same on a Word document.
-I don't mind working with latex but I know some advisors are very adverse to having to use this platform.
-Obviously in order to get the feedback from your advisor you are going to need to use a platform that they are also comfortable with.
+I don't mind working with LaTeX but I know some advisors are very adverse to having to use this platform.
+Obviously, in order to get the feedback from your advisor you are going to need to use a platform that they are also comfortable with.
 So please consult them on what they are happy to use.
 You may also need to instruct them about how to make comments.
-For example track changes is a function that a lot of people like to use but it doesn't translate back into other platforms seamlessly.
+For example, track changes is a function that a lot of people like to use but it doesn't translate back into other platforms seamlessly.
 
 As with many things in this book I urge you to keep an open mind with respect to software.
-For example if your lab is progressive and already experimenting with new and flexible software platforms then I suggest that you learn with them.
-The future of computing is certainly likely to be open-source software and so the sooner that you become more familiar with this, the easier you will find it to adapt in future.
+For example, if your lab is progressive and already experimenting with new and flexible software platforms then I suggest that you learn with them.
+The future of computing is certainly likely to be open-source software, and so the sooner that you become more familiar with this, the easier you will find it to adapt in future.
 
@@
 
-Importantly, what you want to try to avoid is conducting a major switch between software right at the end of your PhD. Try making such changes when you have more time, like after you’ve finished!
+Importantly, what you want to try to avoid is conducting a major switch between software right at the end of your PhD. Try making such changes when you have more time.
 
@@
 I like my students to give each chapter of their thesis a nickname or short running title that we both agree on to name the files.
-An example might be “Swimming_performance_V1.rft”.
+An example might be “Swimming_performance_V1.rtf”.
 The reason why I don’t like “Chapter_1.rtf” as a filename is that every student has a chapter 1, so I can’t distinguish between them.
 In addition, I’m not good at remembering which chapter is chapter 1 and which is chapter 2.
-Sometimes, towards the end of a thesis, the chapters get their orders switched around anyway, causing even more confusion.
+Sometimes, towards the end of a thesis, the chapters get their orders switched around, causing even more confusion.
 Notice that I don’t like having spaces in filenames.
@@
 
-Once I’ve commented on a file and sent it back in an email, or uploaded it to their folder, I’ll add my initials Swimming_performance_V1_JM.rtf.
-Each time I comment on this version, I will add a number to my initials and the student adds one back, so eventually we might come to: Swimming_performance_V1_JM3_MM2.rtf.
+Once I’ve commented on a file and sent it back in an email, or uploaded it to their shared folder, I’ll add my initials Swimming_performance_V1_JM.rtf.
+Each time I comment on this version, I will add a number to my initials and the student adds one back, so eventually, we might come to: Swimming_performance_V1_JM3_MM2.rtf.
 After three rounds, we would normally send a clean V2 out to co-authors and once their comments are back in we should be all set.

### 2021-07-20 2.11-paywall.Rmd
Add files via upload
# After copy edits

@@
 
-Once you've understand where the money for science and publishing comes from, you will understand that there is no real shame in trying to access papers that are behind a paywall.
-Essentially, I’m going to recommend that you move from legitimate to illegitimate, and then if all else fails to get an interlibrary loan.
-This is last on the list as someone will have to pay for it.
+Once you understand where the money for science and publishing comes from, you will understand that there is no real shame in trying to access papers that are behind a paywall.
+Essentially, I’m going to recommend that you move from legitimate to illegitimate means, and then if all else fails to get an interlibrary loan.
+This is last on the list as someone will have to pay for it and it will take some time.
 Your library may have an allocation that they can use before the bill comes to your advisor’s desk.
@@
 
-Google scholar has lots of links (on the right) labelled with “[pdf]”, and these are often integrated with your library.
+Google Scholar has lots of links (on the right) labelled with “[pdf]”, and these are often integrated with your library.
 If there’s nothing at first, click on the “All 9 versions” link (below and to the right).
@@
 
+### Write to the author and request a pdf
+Some authors will respond very quickly to reprint requests, while you won't hear from others for sometime, or at all.
+However, this was the traditional way of obtaining a copy that you didn't have access to, and is definitely worth trying.
+You should be able to get the email address of the corresponding author from the journal website.
+Sometimes these are hidden and you need to jump through hoops to find them.
+You could also search for the email addresses of any co-authors, and also write to them.
+
 ### Go to an academic social media site
@@
 You may need to be away from campus and using your own machine, or in an internet cafe to access them (a lot are blocked by universities through pressure from publishers).
-For some, like [Sci-hub](https://sci-hub.se/), you will need the [DOI](DOI.html) (you can get this from the publisher’s website).
+For some, like [Sci-hub](https://sci-hub.se/), you will need the [DOI](#DOI) (you can get this from the publisher’s website).
 At the time of writing, Sci-hub has >85,483,812 papers.
 [Unpaywall.org](https://unpaywall.org/) has >28,112,995 papers.
+[Library Genious](http://libgen.li/) or Libgen claim to have 3 million files.
 [Open Access Button](https://openaccessbutton.org/) don’t say how many they can reach, but “millions” of papers.
@@
 The chances are that you have friends at other universities where the subscription you are looking for is not behind a paywall.
-If not then ask your advisor or labmates, or post it on your lab’s (electronic) notice board.
+If not, then ask your advisor or labmates, or post it on your lab’s (electronic) notice board.
 
-### Go back to you librarian and ask them for help
+### Go back to your librarian and ask them for help
 

### 2021-07-20 2.12-taxonomy.Rmd
Add files via upload
# After copy edits

@@
 This is expressed in a binomial: two names, first the genus (with an initial capital letter) and then the species name (lower case).
-This name is italicised by convention (or in some cases underlined).
+This name is italicised by convention (or, in some cases, underlined).
 Note that no other taxonomic names (e.g.
@@
 The single name is a hard rule as taxonomy can't survive in any other way.
-The most recently published version is the one to use, but happily with many vertebrates [mammals](https://www.mammaldiversity.org/), [birds](https://avibase.bsc-eoc.org/avibase.jsp?lang=EN), [fish](https://www.fishbase.se/search.php), [reptiles](http://www.reptile-database.org/) and [amphibians](http://research.amnh.org/vz/herpetology/amphibia/) there are well curated databases to which you can refer for the latest taxonomic treatments.
-Also see [WoRMS](http://www.marinespecies.org/) for marine species.
+The most recently published version is the one to use, but happily with many vertebrates ([mammals](https://www.mammaldiversity.org/), [birds](https://avibase.bsc-eoc.org/avibase.jsp?lang=EN), [fish](https://www.fishbase.se/search.php), [reptiles](http://www.reptile-database.org/) and [amphibians](http://research.amnh.org/vz/herpetology/amphibia/)) there are well curated databases to which you can refer for the latest taxonomic treatments.
+Also, see [WoRMS](http://www.marinespecies.org/) for marine species.
 My suggestion is that you defer to these (with citation) and you won't go far wrong.
@@
 
-When my former student, Mohlamatsane Mokhatla, started his PhD in 2013 he worked on three species, and over time these changed so that at various points in his thesis.
+When my former student started his PhD in 2013, he worked on three species, and over time these changed so that at various points in his thesis they had a variety of names.
 You can see the combinations he had to contend with in Table \@ref(tab:taxonomy).
@@
 The taxonomy of two species changed such that he had to constantly revise the names in his thesis.
-Luckily, he could simply conduct a "replace all", but it did become quite confusing.
+Luckily, he could simply conduct a 'replace all', but it did become quite confusing.
 There's a chance that they've changed again, so I've provided hyperlinks for you to check out what the current name is.
@@
 
-In taxonomic papers, and certain journals with a taxonomic focus, there is an insistence that you cite the "taxonomic authority" after the species name when used for the first time.
+In taxonomic papers, and certain journals with a taxonomic focus, there is an insistence that you cite the 'taxonomic authority' after the species name when used for the first time.
 This is essentially a citation in which you acknowledge the original description that accompanies the name.
-Let's take Mohlamatsane's examples again:
+Let's take the above examples again:
 
@@
 However, in 2016 new work on the type specimen revealed that it was the same as *Amietophrynus rangeri,* but because Tschudi's description was older, the taxonomy deferred to the rule of precedence.
-If you want to read more about it, you can look at all the changes on the AMNH amphibian database.
+If you want to read more about it, you can look at all the changes on the amphibian taxonomic database.
 
@@
 You might find that you have two species, which belong to different genera that share the same first letter.
-In this case, most authors use a different abbreviation: for example *Ap.
+In this case, most authors use a different abbreviation: for example, *Ap.
 rangeri* and *At.
 angolensis*.
-This is exasperated when both species names are the same.
-Keen readers would have noticed that Mohlamatsane doesn't have this problem any more!
+This is exasperated when both species' names are the same.
 
@@
  - Viruses – [International Code of Virus Classification and Nomenclature](https://talk.ictvonline.org/information/w/ictv-information/383/ictv-code)
- - Bacteria & Archaea [International Code of Nomenclature of Prokaryotes](https://www.the-icsp.org/bacterial-code)
- - Algae, Fungi & Plants [International Code of Nomenclature for algae, fungi, and plants](https://www.iapt-taxon.org/nomen/main.php)
- - Animals [International Code of Zoological Nomenclature](https://www.iczn.org/the-code/the-international-code-of-zoological-nomenclature/)
+ - Bacteria & Archaea – [International Code of Nomenclature of Prokaryotes](https://www.the-icsp.org/bacterial-code)
+ - Algae, Fungi & Plants – [International Code of Nomenclature for algae, fungi, and plants](https://www.iapt-taxon.org/nomen/main.php)
+ - Animals – [International Code of Zoological Nomenclature](https://www.iczn.org/the-code/the-international-code-of-zoological-nomenclature/)
 
@@
 
-Plant taxonomy is considered the taxonomic mother discipline as the work of Linneaus in 1753 was the first botanical work that consistently used Linneus' binomial (two names: Genus species) nomeclatural system [@linnaeus1753species] which pre-dated the more inclusive (and better known) landmark publication in 1758 [@linnaeus1758systema].
+Plant taxonomy is considered the taxonomic mother discipline as the work of Linneaus in 1753 was the first botanical work that consistently used Linneus' binomial (two names: Genus species) nomenclatural system [@linnaeus1753species] which pre-dated the more inclusive (and better known) landmark publication in 1758 [@linnaeus1758systema].
 Until the Botanical Code was revised in 2012, all new botanical diagnoses had to be written in Latin.
-In addition to the Latin binomial name (Genus and species), plants (and some other taxa) often require further descriptors or **ranks** that denote subspecies, varieties, subvareties, etc.
+In addition to the Latin binomial name (Genus and species), plants (and some other taxa) often require further descriptors or **ranks** that denote subspecies, varieties, subvarieties, etc.
 I have included a (non-exhaustive) list of these in Table \@ref(tab:plantrank), together with some examples.
 
-Unlike animal authorities, for plants if the combination of Genus and species has changed, both the authority for the original genus placement, and the author responsible for the new combination are given after the species name (with the former in brackets).
+Unlike animal authorities, for plants, if the combination of Genus and species has changed, both the authority for the original genus placement, and the author responsible for the new combination are given after the species name (with the former in brackets see Table Table \@ref(tab:plantrank)).
 In addition, for plants no year is given, and the name of the authority is usually abbreviated.
-Because Linneaus named so many plant species is his landmark publications, instead of writing out the full taxonomic authority after the species names, botanists abbreviate this to "L.".
+Because Linneaus named so many plant species in his landmark publications, instead of writing out the full taxonomic authority after the species names, botanists abbreviate this to "L.".
 Like Linneaus, other prolific plant taxonomists can have their names abbreviated, but plant taxonomists became so numerous that there needed to be some standardisation of names and their abbreviations.
@@
 
-Table: (\#tab:plantrank) **Plants often need more information than the simple Latin biomial.** This table provides the required abbreviations that you will need to add in order to properly describe the rank of your plant taxa.
+Table: (\#tab:plantrank) **Plants often need more information than the simple Latin binomial.** This table provides the required abbreviations that you will need to add in order to properly describe the rank of your plant taxa.
+Note that the taxonomic authority (last column) can include two names with the first in parentheses if the second changed the placement of this taxon.
 
-| Rank | Abbreviation	| Example |
-|-----|-----|-----|
-| Subfamily | subf. | Poaceae subf. Pooideae |
-| Subgenus | subg. | _Metrosideros_ subg. _Metrosideros_ (Myrtaceae) |
-| Section | sect. | _Senecio_ sect. _Senecio_ (Asteraceae) |
-| Subsection | subsect. | _Stachys_ subsect. _Swainsonianeae_ (Labiatae) |
-| Series | ser. | _Mimosa_ ser. _Paucifoliolatae_ |
-| Subseries | subser. | _Palicourea_ subser. _subcymosae_ |
-| Subspecies | subsp. | _Trichophorum cespitosum_ subsp. _germanicum_ |
-| Variety | var. | _Closterium acerosum_ var. _elongatum_ |
-| Subvariety | subvar. | _Hieracium vulgatum_ subvar. _oblongatum_ |
-| Form | f. | _Prunus cerasifera_ f. _atropurpurea_ |
+| Rank | Abbreviation	| Example | Authority |
+|-----|-----|-----|-----|
+| Subfamily | subf. | Poaceae subf. Pooideae | Benth.  |
+| Subgenus | subg. | _Metrosideros_ subg. _Metrosideros_ (Myrtaceae) |   |
+| Section | sect. | _Senecio_ sect. _Senecio_ (Asteraceae) |   |
+| Subsection | subsect. | _Stachys_ subsect. _Swainsonianeae_ (Labiatae) |   |
+| Series | ser. | _Mimosa_ ser. _Paucifoliolatae_ | Benth.  |
+| Subseries | subser. | _Palicourea_ subser. _subcymosae_ | (Müll.Arg.) C.M.Taylor  |
+| Subspecies | subsp. | _Trichophorum cespitosum_ subsp. _germanicum_ | (Palla) Hegi |
+| Variety | var. | _Closterium acerosum_ var. _elongatum_ | Brebisson |
+| Subvariety | subvar. | _Hieracium vulgatum_ subvar. _oblongatum_ | Sudre |
+| Form | f. | _Prunus cerasifera_ f. _atropurpurea_ | Diffel |
 
 Algae sometimes exist in association with other taxa (such as lichens which are an association between an algae or cyanobacteria and one or more fungi.
-For these associations there is a specific code for their nomenclature: International Code of Phytosociological Nomenclature [@theurillat2021international].
+For these associations, there is a specific code for their nomenclature: International Code of Phytosociological Nomenclature [@theurillat2021international].
 
 There is also a code for cultivated plants, and if you study any kind of agronomical system, you will be familiar with not just the normal binomial name, but also a specific cultivar or variety (*var.*) that follows the binomial: [The International Code of Nomenclature for Cultivated Plants](https://www.ishs.org/scripta-horticulturae/international-code-nomenclature-cultivated-plants-ninth-edition).
+
+## Other times to use italics
+
+In general, italics are used when names are derived from Latin or Greek.
+A number of subdisciplines have international codes for how the names should be written: Anatomy, Histology, Embryology and Neuroanatomy.
+There is often confusion about these correct terms, but luckily four major publications are here to help you.
+These are maintained by the international groups who specialise in standardising the nomenclature, and they are republished from time to time (so look out for updates).
+I provide directions to these citations should your research require them:
+
+- Terminologia Anatomica. [@terminology1998terminologia]
+- Terminologia Histologica. [@terminology2008terminologiaa]
+- Terminologia Embryologica. [@terminology2013terminologia]
+- Terminologia Neuroanatomica. [@terminology2017terminologia]
 
@@
 My suggestion would be to take the most commonly used common name, as that's likely to be the one that most people will recognise.
-However, it's not always that simple and because there's no "correct" common name, you have some leeway to choose the one you prefer.
+However, it's not always that simple and because there's no 'correct' common name, you have some leeway to choose the one you prefer.
 
@@
 In a world of stable taxonomy (such as that enjoyed by ornithologists), I can understand that this is possible.
-But imagine the requirements in common name changes for Mohlamatsane in his thesis over only a few years.
+But imagine the requirements in common name changes for frogs in Table \@ref(tab:taxonomy) over only a few years.
 Surely the useful thing about common names is that they are used commonly by normal people who don't worry about taxonomy?
@@
 
-### View 1.
+### View 1
 
@@
 
-### View 2.
+### View 2
 

### 2021-07-20 2.13-style.Rmd
Add files via upload
# After copy edits

@@
 
-> “The goal of good writing is straightforward: to make your reader's job as easy as possible” 
-> ---  Kevin W. Plaxco [-@plaxco2010art].
+> “The goal of good writing is straightforward: to make your reader's job as easy as possible.” 
+>
+> Kevin W. Plaxco [-@plaxco2010art]
 
 This great quote from a paper by Plaxco [-@plaxco2010art], who cuts to the heart of why it’s essential to write well.
-As an editor, I once experienced an author who thought the opposite, and for me this was the epiphany.
+As an editor, I once experienced an author who thought the opposite, and for me, this was the epiphany.
 The submitted manuscript was an impenetrable mess.
-I could tell that there was good work in there, but as an editor I felt that the information had been made so obscure by the authors that my readers were unlikely to get much out of it.
+I could tell that there was good work in there, but as an editor, I felt that the information had been made so obscure by the authors that my readers were unlikely to get much out of it.
 The author was a colleague, and so I decided to phone him and chat about the need for much greater clarity in his manuscript.
@@
 “Let the reader work to make sense of my data,” he said.
-My epiphany came not because this was a totally alien concept, but because during my PhD days this was exactly how I had thought scientific writing should be done.
+My epiphany came not because this was a totally alien concept, but because during my PhD days, this was exactly how I had thought scientific writing should be done.
 I had spent so many hours slogging through dense and dreary papers by well-respected figures.
@@
 
-The preponderance of inpenetrable guff has led to a general perception that "academics stink at writing" - as Steven Pinker put it [@pinker2014why].
+The preponderance of impenetrable guff has led to a general perception that "academics stink at writing" - as Steven Pinker put it [@pinker2014why].
 But, I think that Pinker [-@pinker2014why] is outdated.
@@
 There are still authors out there who attempt to set puzzles for their readers, but they aren’t in the mainstream any more.
-Instead, biological sciences has some inspiring writers, and many more are taught how to inspire future generations towards communicating great science.
+Instead, biological sciences have some inspiring writers, and many more are taught how to inspire future generations towards communicating great science.
 That’s not to say that there are no puzzles left.
-Much of what we do requires great puzzle solving abilities.
+Much of what we do requires great puzzle-solving abilities.
 However, let’s keep the puzzles away from communicating with our audience, make their job as reader easy or even pleasurable, and they’ll keep coming back for more.
 
-## Don’t bamboozle
+## Develop your writing style
 
-It’s easy to use jargon.
-The whole point of jargon is to convey a (usually complex) idea in a short amount of space.
+During the course of your PhD you should improve your writing style toward proficient communication with your target audience.
+This will involve improving both your technical writing and accessibility writing skills.
+Although the bottom left portion of Figure \@ref(fig:develop-writing-style) is labelled “Poor communication” we might more productively think of it as being the area of the ‘first draft’.
+You have to start somewhere.
+If your current style is too ‘hard’ and technical (top left), then you will need to increase your accessibility.
+If on the other hand you are used to a more popular writing style (bottom right), you will need to increase your technical ability.
+See Dunleavy [-@dunleavy2003authoring] for more insight on achieving your requisite writing style.
+
+(ref:develop-writing-style) **Develop your scientific writing style.** Pitching your writing style will take some practice to reach the desired pitch.
+Wherever your starting point in this technicality and accessibility square, your aim is to move toward the top right corner.
+Adapted from Dunleavy (2003: Fig.
+5.1).
+
+```{r develop-writing-style, echo=FALSE, out.width = '95%', fig.cap="(ref:develop-writing-style)"}
+knitr::include_graphics('figures/develop-writing-style.png')
+```
+
+### Don’t bamboozle {#bamboozle}
+
+It’s easy to use technical jargon (top left portion of Figure \@ref(fig:develop-writing-style)).
+The whole point of jargon is to convey an (usually complex) idea in a short amount of space.
 Using a word (or two) instead of using several sentences clearly has some advantages.
@@
 My old advisor at Liverpool University, the late, great Brian Moss, shared the following example of too much jargon by Bump et al [@bump2009large], when plain English would have been much shorter.
-Do yourself a favor and read the abstract of Bump et al [-@bump2009large].
+Do yourself a favor and read the abstract of Bump et al.
+[-@bump2009large].
 The fact that it gave Moss a chance to write about moose wasn’t lost on anyone (see Figure \@ref(fig:moss))!
 
-(ref:moss) **Bamboozled Moss takes on the moose.** Brian was so unhappy after reading Bump et al (2009) the above that he felt compelled to write a letter to the British Ecological Society Bulletin.
+(ref:moss) **Bamboozled Moss takes on the moose.** Brian was so unhappy after reading Bump et al. (2009) that he felt compelled to write a letter to the British Ecological Society Bulletin.
 
@@
 
-Whatever your writing style, you need to be conciious of your ultimate aim in writing: to communicate with your reader.
+Whatever your writing style, you need to be conscious of your ultimate aim in writing: to communicate with your reader.
 Making your text as easy to read as possible [as @plaxco2010art put it] isn't easy, but it will be rewarding.
-The reward you will get is in willing readers to the message that you want to convey.
-This is something that will stand you in good sted no matter what career you have after your PhD.
+The reward you will get is having willing readers understand the message that you want to convey.
+This is something that will stand you in good stead no matter what career you have after your PhD.

### 2021-07-20 2.14-voice.Rmd
Add files via upload
# After copy edits

@@
-# When is it possible to retain your own voice when writing? {#voice}
+# Retaining your own voice when writing {#voice}
 
-Should one use “I” or “we” when writing a scientific paper?
+Should one use first-person pronouns ('I' or 'we') when writing a scientific paper?
 Although there was a tradition not to use first-person pronouns when writing scientific articles, this has fallen by the wayside in recent years, because the use of I or we makes writing simpler for writers and easier to understand for readers.
-Use “I” if you are the sole author (or in your thesis) and “we” for two or more authors.
+Use 'I' if you are the sole author (or in your thesis) and 'we' for two or more authors.
 
-You will regularly see “I” or “we” being used in the last paragraph of the introduction and the first paragraph of the discussion.
+You will regularly see 'I' or 'we' being used in the last paragraph of the introduction and the first paragraph of the discussion.
 It’s also quite common to use this in the methods section.
-The reason why these sections have “I” or “we” is quite obvious, as it allows you to place your own voice on your actions and decisions.
+The reason why these sections have 'I' or 'we' is quite obvious, as it allows you to place your own voice on your actions and decisions.
 After all, it is your aim and hypothesis for your study, so you should own it.
@@
 
-The first use of “I” or “we” at the end of the introduction is powerful.
+The first use of 'I' or 'we' at the end of the introduction is powerful.
 Up to that point, the reader has read four or five paragraphs of logical arguments outlining the background information that makes up the reasoning for your study.
@@
 So if you were to continuously use it throughout the methodology, the reader may get tired of reading a string of statements stating that you did this, then you did that, and so on.
-The power of using “I” or “we” is to use it sparingly to boost clarity.
+The power of using 'I' or 'we' is to use it sparingly to boost clarity.
 
-## Does writing “I” or “we” help people to understand who you are?
+## Does writing 'I' or 'we' help people to understand who you are?
 
@@
 Having said that you should write about subjects that you feel passionately about, this does not mean that you should write with unsubstantiated passion.
-For example, it makes no sense to write: "The threats to this spectacular species surpass any mechanism known to humankind" ...unless you have substantive citations or results to back this up.
+For example, it makes no sense to write: "'"The threats to this spectacular species surpass any mechanism known to humankind" ...unless you have substantive citations or results to back this up.
 Yes, it's passionate, but that should be balanced by the need to keep what you write factual.

### 2021-07-20 2.15-concise.Rmd
Add files via upload
# After copy edits

@@
 Instead, I simply try to rework a sentence into what I know is correct (and I often don't achieve that).
-The lesson then is that should you come across an opportunity to learn more about writing, do take it.
+The lesson is that should you come across an opportunity to learn more about writing, do take it.
+Otherwise, try writing concisely in all that you do.
 
@@
 A personal plea from me would be to ask that you don't start your email by enquiring after the health of your recipients.
-The fashion to to this has turned into a surge in recent years (the pandemic notwithstanding), and the reality is that this are simply superfluous and gratuitous words.
+The fashion to do this has turned into a surge in recent years (the pandemic notwithstanding), and the reality is that these are simply superfluous and gratuitous words.
 You certainly don't expect them to write back starting with a paragraph about their health.
@@
 Although we talk about writing science here, the grammatical rules of writing (in English) are fairly universal (with some local exceptions), so do practice your writing.
-Maybe write a letter (or long email) instead of making a call sometimes.
+Maybe write a letter (or long email) instead of making a call.
 It'll also help you to [read critically](critical.html).
@@
 You need to stick to your narrative like a highway.
-The highway analogy is useful as although you will see signs to other places that you can point out (to the reader), you shouldn't turn off the highway.
-The best of highways will also be free of traffic jams (think long, complex sentences here).
-and toll free ([no paywall](paywall.html)).
+The highway analogy is useful as although on a highway you will see signs to other places that you can point out to the reader, you shouldn't turn off the highway to get to your destination.
+The best of highways will also be free of traffic jams (think long, complex sentences here), and toll free ([no paywall](paywall.html)).
 
@@
 
-If the highway is 'your message' then getting to the point is the big sign that states the destination.
+If the highway is 'your message' then getting to the point is the big sign that states that you've arrived at your destination.
 
-Some of the best papers I've read manage to encapsulate the whole point in the first sentence, or better still in the title.
+Some of the best papers I've read manage to encapsulate the whole point in the first sentence, or better still, in the title.
 Our formula has you getting to the main question in the last paragraph of the introduction, but you should have already 'got to the point' in the first paragraph - i.e.
@@
 
-You can find a guide to writing your methods in [Part III](part3.html).
+You can find a guide to writing your methods in [Part III](#methods).
 This rule is about not allowing the methods to creep out of this section and into the results, or even the discussion.
 Similarly with the Results.
-There are some specific times when this is permitted (such as a post-hoc test), but generally you shouldn't expect to do this.
+There are some specific times when this is permitted (such as a post-hoc test), but generally, you shouldn't expect to do this.
 
@@
 
-Copy and pasting is very easy, and a good way of suddenly producing vast quantities of text (Figure \@ref(fig:plagiarism)).
-But repeated text is quickly recognised by the reader and appears very boring and cumbersome.
+Copy and pasting are very easy, and a good way of suddenly producing vast quantities of text (Figure \@ref(fig:plagiarism)).
+But the repeated text is quickly recognised by the reader and appears very boring and cumbersome.
 This is likely to happen in the Methods and Results sections.
@@
 The reader will quickly see the repetitions and become bored.
-Similarly, the [conclusion/summary](discussion.html) section is also often copied from lines above.
+Similarly, the [conclusion/summary](discussion.html) section is also often copied from the lines above.
 If it's not worth writing again, then it wasn't worth writing before.
@@
 
-Interesting.
+This rule relates to keeping your text concise, specifically at the start of a sentence.
+You may be tempted to reiterate your point, but this is not always needed.
 I'd say that you need some practice and a critical eye to spot these kinds of errors.
@@
 For some reason, most students avoid this at all cost.
-There appears to be the idea that saying "I" or "we" isn't correct for scientific writing.
+There appears to be the idea that saying 'I' or 'we' isn't correct for scientific writing.
 In fact, it can be the easiest way to avoid the passive voice.
-There are other reasons why [using the first person](voice.html) is helpful.
+There are other reasons why [using the first person](#voice) is helpful.
 
@@
 
-This is really getting down to the nitty gritty.
+This is really getting down to the nitty-gritty.
 It's hard to do this yourself.
@@
 I do get mad when I find that this help is ignored without an indication why.
-My suggestions or comments are simply deleted.
-Certain students become very hard work when they don't even explain why they refuse to change something.
+Occasionally, my suggestions or comments are simply deleted.
+Certain students become very hard work when they don't explain why they refuse to change something.
 Reading the work of these students quickly becomes a chore.
-You should have noted by now that it's all about being flexible, and being stubborn about anything you've written isn't going to work.
+You should have noted by now that it's all about being flexible for your reader, and being stubborn about anything you've written isn't going to work.
 

### 2021-07-20 2.16-foreign.Rmd
Add files via upload
# After copy edits

@@
 
-Reading this book you will learn of some biases in biological sciences and how you need to be aware of them in order to properly interpret your work.
+Reading this book, you will learn of some biases in biological sciences and how you need to be aware of them in order to properly interpret your work.
 Be aware also that biological sciences is biased in where studies take place, and who publishes them [@culumber2019widespread; @nunez2021monolingual].
-What we really need then are more studies from less studied parts of the world, and especially we need these to be researched and published by people from those countries.
+What we really need then are more studies from less studied parts of the world, and we need these to be researched and published by people from those countries.
 Why then do we insist that science be written in English?
 
-English may be one of the most difficult languages to learn how to write as a foreign language because rules are so inconsistent, and the vocabulary is so large.
-At 171476 words, English has an enormous vocabulary, followed closely by Russian (150000), Spanish (93000) and Chinese (85568: although you should note that counting words isn’t very accurate and as for counting dictionary entries, Korean wins).
+English may be one of the most difficult languages to learn to write as a foreign language because rules are so inconsistent, and the vocabulary is so large.
+At 171 476 words, English has an enormous vocabulary, followed closely by Russian (150 000), Spanish (93 000) and Chinese (85 568: although you should note that counting words isn’t very accurate and as for counting dictionary entries, Korean wins).
 I cannot defend the use of English as the language of science, but I can assert that having a universal language for communicating science is massively helpful, and makes a large swathe of the literature available to those scientists who are prepared to become proficient in a second language.
@@
 One of my 'pet hates' is that many non-English speaking editors seem to insist that you have your work checked by a 'native English speaker' (see Figure \@ref(fig:editor)).
-This suggests to me that some of the worst people to comment on whether or not something is written in good English, is another foreign language speaker.
-The lesson here is that you should be careful who you trust.
-Moreover, in my experience, having advice from a 'native English speaker' is no guarantee to getting a well written chapter or manuscript.
+In my experience, having advice from a 'native English speaker' is no guarantee to getting a well-written chapter or manuscript.
 Most of the world’s scientists did not grow up speaking English.
@@
 I have read enough correct English to know when something is incorrect, but that doesn’t mean that I know how to correct it.
-I have spent many long hours trying to decode what others have written, and in some cases this has involved me rewriting entire manuscripts.
-I still do this as a co-author, although I do remember asking one colleague to please send any further drafts in their native Spanish as it would be easier to translate than it would be to re-write.
+I have spent many long hours trying to decode what others have written, and in some cases, this has involved me rewriting entire manuscripts.
+I still do this as a co-author, although I do remember asking one colleague to please send any further drafts in their native Spanish as it was easier to translate than it to re-write.
 
 The time component is at the crux of my reasoning why, as an editor, reviewer and advisor, I will not provide an English language service.
-It is both a time consuming and an unsatisfying experience.
+It is time consuming and an unsatisfying experience.
 We all have our own voice, and correcting while maintaining other people’s voices is a painstaking task.
@@
 And this agrees with the opinions of some non-English speakers who ask for editors and reviewers to concentrate on the science, a sentiment with which I agree.
-
-Given that I’m not going to correct your English, what remarks would you expect me to make when reviewing a poorly written manuscript?
 
@@
 One option is the service offered by **AuthorAID** (http://www.authoraid.info/en/).
-Using AuthorAID, you can find a long term mentor who will help you with your English.
+Using AuthorAID, you can find a long-term mentor who will help you with your English.
 You can read more about this approach here [@freeman2006publishing].
@@
 Most of the world doesn't speak English, and so there is a real need for this to be done in your language.
-See a nice blog from the LSE on this subject here.
+Many journals now allow abstracts to be submitted bilingually.
+Obviously, the second language should have some relevance to the study itself.
+Even if there is nothing in the instructions to authors about submitting the abstract in another language, I'd encourage you to do it.
+The more we scientists are exposed to the reality that other languages exist, the more it will rub off on those that need to understand.
+[Read more on this blog @taskin2020long]
 
@@
 >Kurt Vonnegut [-@vonnegut1982how]:
+>
 >"No matter what your first language, you should treasure it all your life.

### 2021-07-20 2.17-plagiarise.Rmd
Add files via upload
# After copy edits

@@
-## Making sure that you don’t plagiarise {#plagiarise}
+# Making sure that you don’t plagiarise {#plagiarise}
 
-Essentially plagiarism is when you copy somebody else's work.
-We're most commonly familiar with plagiarism of writing and these days this is especially easy with the copy and paste function.
+Plagiarism is when you copy somebody else's work.
+We're most familiar with plagiarism of writing and these days this is especially easy with the copy and paste function.
 Most people are not even aware that they have plagiarised because sometime in the past they copied and pasted the work of somebody else into a document and this later became incorporated into their text without them being aware of it.
 
-Plagiarism is a problem because essentially you are taking somebody else's work without attributing it to them.
+Plagiarism is a problem because, essentially, you are taking somebody else's work without attributing it to them.
 
-### How to know if you have plagiarised
+## How to know if you have plagiarised
 
 Today there are several pieces of software that are used to scan text that's written and available on the internet to discover plagiarism.
-One such example is Turnitin.
+One such example is TurnItIn.
 One of the outputs of TurnItIn is to highlight text that matches other text already on the internet.
-As almost all journals published on the internet TurnItIn can accurately determine if text has been copied from another article or website.
+As almost all journals publish on the internet, TurnItIn can accurately determine if text has been copied from another article or website.
 I usually set TurnItIn to determine plagiarism with five or more consecutive words.
@@
 
-###  What to do if plagiarism is detected in your work
+##  What to do if plagiarism is detected in your work
 
@@
 __Here's what you do:__
+
 - Read the sentence that has been plagiarised several times to yourself. 
 - Now without looking at that sentence, write another sentence that has the same meaning.
-Because it's very hard to replicate somebody else's words without copying them, what you should find is that you've written a fresh plagiarised sentence.
-- This can now be added to your text, changed as appropriately to fit your existing text. And that should be the end of your plagiarism worries.
+Because it's very hard to replicate somebody else's words without copying them, what you should find is that you've written a fresh unplagiarised sentence.
+- This can now be added to your text, changed as appropriate to fit your existing text. And that should be the end of your plagiarism worries.
 
 __Here's what not to do:__
+
 - Take the sentence swap out some of the words for synonyms and pass it off as your own. 
-- The sentence will still have the same structure that you copied and essentially this is still somebody else's work. Moreover TurnItIn will still recognise this as plagiarism.
+- The sentence will still have the same structure that you copied and, essentially, this is still somebody else's work. Moreover, TurnItIn will still recognise this as plagiarism.
 
-### How can you make sure that you never plagiarise?
-Quite simply if you never cut and paste, you won’t plagiarise.
+## How can you make sure that you never plagiarise?
+Quite simply, if you never cut and paste, you won’t plagiarise.
 It’s that easy.
@@
 It still seems amazing to me when I submit a student manuscript to TurnItIn and see that it is completely free of any plagiarism.
-There are so many words in English (>643 000 according to the Oxford English Dictionary) and so many different ways of putting them that you really can have your own writing style.
+There are so many words in English, and so many different ways of putting them that you really can have your own writing style.
 Your writing style will be as unique to you as a fingerprint, and it will be entirely free of plagiarism.
@@
 
-Want to read more about plagiarism to understand what it is: Louw [-@louw2017defining]
+Want to read more about plagiarism to understand what it is: see Louw [-@louw2017defining]
 
-### What is grey plagiarism?
+## What is autoplagiarism?
 
@@
 So what are people talking about when they refer to a grey zone?
-The nominal grey zone within plagiarism comes when you copy your own text, this is sometimes referred to as self-plagiarism, but this also refers to the associated practice of ‘manuscript recycling’ [@geraldi2021self-plagiarism].
-Essentially this is really just the same as plagiarism but obviously if you are the original author then if you use your own text it's not the same as using somebody else's, because the work is still yours.
-However this is frequently complicated by multiple authors on a paper.
-For example, if I wrote a paper with five other people, even though the words were mine, if I use them again they would actually be attributed to all six authors and not just to me.
-Thus, if I use them again with five different authors it would still be plagiarism.
+The nominal grey zone within plagiarism comes when you copy your own text, this is sometimes referred to as 'text-recycling', 'self-plagiarism' or 'autoplagiarism'.
+A new guideline from Hall et al.
+[-@hall2021understanding] sets out the different types of grey plagiarism, or text recycling:
 
-This is a particular problem in biological sciences (and some other sciences) when it comes to the methodology section ([see part 3](methods.html)).
-Many students believe after having written the [Materials and Methods](methods.html) once, they can’t possibly write it again in another way.
-Actually, you can and you should.
-It’s good practice, but remember that you can also cite the methods in a previous chapter.
-For example, chapter 2 has one method written out in great detail.
-Chapters 3 and 4 then reference chapter 2 while giving only a brief section on that methodology.
-You should already have seen some examples of how people do this in the literature, especially when citing their own previous work.
+### Developmental recycling {#develop}
 
-This is further complicated by having different publishers for your papers.
-Let's say I wrote a paper, I was the sole author and it was published by a journal that retained the copyright.
-If I use the same text again in another paper where again I was the soul author, but it was published in a different journal and they retained the copyright, that would be a breach of copyright to the first journal.
-It would be regarded as plagiarism.
+Developmental recycling is when you are reusing text that you have written for example between your proposal and something you intend for publication or in an ethics application that you also want to use in your thesis.
+All of this sort of developmental recycling is permitted and actually encouraged.
+I would further encourage you to use the opportunity of recycling this text to develop it and refine it further, condensing and improving where you can.
 
-Obviously, this situation really shouldn't occur because if you are the author of your own work then you should retain the copyright it should never be given to a publisher or a journal.
-The fact that many journals want to obtain copyright on work that they don't pay for is a complete disgrace but it's something that we're not going to go into right now (you can see rants here and here about science publishing scandals).
+### Generative recycling {#generative}
 
-### Manuscript recycling
+Generative recycling is where you take pieces of already published text for example from the methods when it does not make sense to change the text or actually makes it more obscure to reword it in order to avoid plagiarism.
+In my experience this doesn't amount to more than a few sentences describing technical settings on equipment.
+However this will depend strongly on your own subject area and may amount to larger chunks of text.
+I suggest that it is usually possible to reword most of the methods sections of papers.
+You should really only be generatively recycling material if you cannot avoid it: i.e.
+when the text becomes more obscure by your attempts to reword it.
 
-Manuscript recycling is the practice of re-publishing the same manuscript (or with minimal changes) in different journals.
+### Adaptive recycling
+
+Adaptive recycling is where you are using published text as the basis for a different type of content (e.g.
+a popular article online, a magazine, or op-ed).
+I think that this kind of text recycling is quite unnecessary because you almost certainly need to reword your text for a different audience.
+There maybe times such as figure legends where you need to reuse text that was already published.
+If you do find yourself in such a position then check with the copyright owner of the material that you are able to reuse the text that you want without legal issues.
+
+### Duplicate recycling
+
+Duplicate recycling (also known as 'manuscript recycling') is where large tracts of texts are essentially the same for the same message and audience.
+This is never likely to be sanctioned as it suggests that you are attempting to publish the same work twice.
+It will not be legal or ethical.
 Manuscript recycling is a bigger problem in other subject areas, notably economics [@geraldi2021self-plagiarism], publishing issues in economics appear to need a lot more professionalism.
 
-### Overall - just don’t plagiarise
+## Overall - just don’t plagiarise
 

### 2021-07-20 2.18-phrasebank.Rmd
Add files via upload
# After copy edits

@@
-## Academic phrasebank {#phrasebank}
+# Academic phrasebank {#phrasebank}
 
@@
 They seem so well honed to the situation that I’m envious about how the writer dreamt them up.
-Stealing them isn’t quite [plagiarism](plagiarism.html), but then they don’t regularly fit the exact situation that I’m looking for.
+Using them isn’t quite [plagiarism](plagiarism.html), but then they don’t regularly fit the exact situation that I’m looking for.
 
-Imagine cutting and pasting them into a file, or even better a database that’s ready for any eventuality. Banking them over time so that they are always available.
+Imagine cutting and pasting these phrases into a file, or even better a database that’s ready for any eventuality. Banking them over time so that they are always available.
 
@@
 
-the academic phrasebank
+[**the academic phrasebank**](http://www.phrasebank.manchester.ac.uk/)
 
-### How to find the phrase that you need:
+## How to find the phrase that you need:
 
 - Look across the top for different sections of text that you are currently writing.
-- Then look down the left hand side to see the type of language function that you need. Click on something that looks applicable and watch the database open up.
+- Then look down the left-hand side to see the type of language function that you need. Click on something that looks applicable and watch the database open up.
 
-### Getting started
+## Getting started
 
@@
 Is it plagiarism?
-Given that the chapter before is all about not copying and pasting, am I now telling you that it’s ok to copy and paste?
+The preceding chapter is all about not copying and pasting, am I now telling you that it’s ok to copy and paste?
 Most of these phrases are less than five words or you will find that they are non-consecutive parts of a sentence.

### 2021-07-20 2.19-critical.Rmd
Add files via upload
# After copy edits

@@
-## Why critical reading is crucial for improving your writing {#critical}
+# Why critical reading is crucial for improving your writing {#critical}
 
@@
 As I’ve pointed out elsewhere, your chapter or manuscript is likely to resemble the formula that many others already follow, and standing on the shoulders of those giants will help you again.
-But it’s possible that your reading isn’t helpful if you aren’t being sufficiently ‘critical’ when you read: and by critical I am talking about the methodical approach that you take to reading that improves your comprehension of a paper (Figure \@ref(fig:Criticalreading)).
+But it’s possible that your reading isn’t helpful if you aren’t being sufficiently ‘critical’ when you read: and by critical, I am talking about the methodical approach that you take to reading that improves your comprehension of a paper (Figure \@ref(fig:Criticalreading)).
 
@@
 
-### Making notes on good and bad styles
+## Making notes on good and bad styles
 
@@
 It’s tempting to think that the only aspects of a paper that you should take notes on are the science (questions, approach and findings), but this is only part of your mandate as a critical reader.
-Of course you should be taking notes on the science, but you should also be using it as an opportunity to learn tips and tricks of writing.
+Of course, you should be taking notes on the science, but you should also be using it as an opportunity to learn tips and tricks of writing.
 Once you are actively writing, and I’d suggest that this should have been right from the very beginning of your studies, you should be actively thinking about how the authors are writing.
@@
 
-#### Some examples
+## Some examples
 
@@
 What about the times when authors break the rules?
-Can you see in your reading that the majority of authors follow rules?
+Can you see in your reading that the majority of authors the follow rules?
 Do they follow the formula?
@@
 
-### Make handwritten notes
+## Make handwritten notes
 
@@
 
-### Recognising good or bad in what you read
+## Recognising good or bad in what you read
 
@@
 If you’ve not stopped to check your phone, or think about eating or drinking something, well done.
-It might be worth creating an atmosphere with less other stimulants while you read.
+It might be worth creating an atmosphere with less extraneous stimulants while you read.
 If you only have your reading material and notebook, you’ll be able to concentrate all of your thinking on them instead of getting distracted.
@@
 
-If you are looking for good fodder for your journal club, consider reviewing a preprint and publishing your review with a doi [@teixeiradasilva2017fortifying].
+If you are looking for good fodder for your journal club, consider reviewing a preprint and publishing your review with a [DOI](#DOI) [@teixeiradasilva2017fortifying].
 This will make the whole exercise a lot more worthwhile and produce a net output for your club.

### 2021-07-20 2.20-proposal.Rmd
Add files via upload
# After copy edits

@@
 not including advisors) in your department.
-The department requires that your written proposal is publicly available to everyone in the department to look at (normally deposited in the secretary’s office) at least 7 days prior to the oral presentation of your proposal.
+The department requires that your written proposal is publicly available to everyone in the department to look at prior to the oral presentation of your proposal.
 This is not unique to my university, many institutions require that you formally propose the work that you plan to do in a PhD.
@@
 It is well worth checking out what your institution requires and by when.
-Talk to your adviser about this.
+Talk to your advisor about this.
 
-Your proposal is a collection of documents, where each document is a chapter of the thesis, and is roughly equivalent to a published paper.
+Your proposal is a collection of documents, where each document is a chapter of the thesis.
 Ideally, these should be bounded by an introduction, which gives the bigger picture of how the thesis is placed, and a time-line at the end that demonstrates how you can complete the thesis within the two or three years for your MSc or PhD, respectively.
-Note that your project proposal is placed on your permanent file within the department.
 
@@
 
-We expect that your final thesis will be a set of papers (data chapters sandwiched between an [introduction and discussion](PhDintro.html)), and so we ask you to prepare the proposal in the same way.
+Your final thesis will be a set of papers (data chapters sandwiched between an [introduction and discussion](PhDintro.html)), and so prepare the proposal in the same way.
 That is, you will normally have two (MSc) or five (PhD) data chapters.
-In the proposal, each chapter will consist of three parts:
+In the proposal, each chapter will consist of three parts.
 
@@
 
-## Your Proposal - what is it good for?
+## Your proposal - what is it good for?
 
@@
 This included nearly all of the methods and materials section verbatim.
-Note that this **is not** plagiarism.
-You wrote the original proposal as a single author and it was not published, so you are at liberty to use the text again.
+Note that this **is not** plagiarism, but [developmental recycling](#develop).
 

### 2021-07-20 2.21-proposal-presentation.Rmd
Add files via upload
# After copy edits

@@
-## Making a presentation from your chapter, paper or proposal {#proposal-presentation}
+# Making a presentation from your chapter, paper or proposal {#proposal-presentation}
 
@@
 If you don't know, then you can't make any assumptions about what might be present in the room where you give your presentation.
-I turned up to give my talk with a presentation on a USB drive only to find that there is no computer, no projector and a room of people expecting me to entertain them for an hour.
+I turned up to give my talk with a presentation on a USB drive only to find that there was no computer, no projector and a room of people expecting me to entertain them for an hour.
 Even if you know that a projector is present, it's good to have some extra context:
@@
   + can it advance slides
-  + If in doubt then bring your own accessories.
+  + if in doubt, then bring your own accessories.
 5. **Know exactly where you are giving your talk and plan for plenty of travel time**
@@
 
-Essentially, you have a story to tell, but does not mean you are story telling.
+Essentially, you have a story to tell, but that does not mean you are [story telling](#storytelling).
 It means that your presentation will require you to talk continuously for your allotted period of time, and that the sentences must follow on from each other in a logical narrative; i.e.
@@
 
-### So where do you start?
+## So where do you start?
 
-#### Here are some simple rules to help guide you to build your presentation:
+### Here are some simple rules to help guide you to build your presentation:
 
-- One slide per minute: However many minutes you have to present, that’s your total number of slides. Don’t be tempted to slip in more.
-- Keep the format clear: There are lots of templates available to use, but you’d do best to keep your presentation very clean and simple.
-- Be careful with animations: You can build your slide with animations (by adding images, words or graphics).
+- **One slide per minute**: However many minutes you have to present, that’s your total number of slides. Don’t be tempted to slip in more.
+- **Keep the format clear**: There are lots of templates available to use, but you’d do best to keep your presentation very clean and simple.
+- **Be careful with animations**: You can build your slide with animations (by adding images, words or graphics).
 But do not flash, bounce, rotate or roll.
 No animated little clipart characters.
-- No goofy cartoons – they’ll be too small for the audience to read.
+No goofy cartoons – they’ll be too small for the audience to read.
 No sounds (unless you are talking about sounds).
@@
 They have come to hear about your research proposal.
-- Don’t be a comedian: Everyone appreciates that occasional light-hearted comment, but it is not stand-up.
+- **Don’t be a comedian*: Everyone appreciates that occasional light-hearted comment, but it is not stand-up.
 If you feel that you must make a joke, make only one and be ready to push on when no-one reacts.
@@
 
-#### A formula for presenting a proposal
+### A formula for presenting a proposal
 
 1. You need a title slide (with your name, that of your advisor and institution)
-1. Several slides of introduction
+2. Several slides of introduction
   + that put your study into the big picture
@@
   + give the context of your own study
-1. Your aims and hypotheses
-1. Methods and Materials
+3. Your aims and hypotheses
+4. Methods and Materials
   + Images of apparatus or diagrams of how apparatus are supposed to work. If you can’t find anything, draw it simply yourself.
@@
   + Analyses are important.
-Make sure that you understand how they work, otherwise you won’t be able to present them to others.
+Make sure that you understand how they work, otherwise, you won’t be able to present them to others.
 Importantly, explain where each of the variables that you introduced, and explained how to measure, fit into the analyses.
 There shouldn’t be anything new or unexpected that pops up here.
-1. Expected results
-I like to see what the results might look like, even if you have to draw graphs with your own lines on it. Use arrows to show predictions under different assumptions.
-1. Budget
-1. Timeline
+5. Expected results
+6 like to see what the results might look like, even if you have to draw graphs with your own lines on it. Use arrows to show predictions under different assumptions.
+7. Budget
+8. Timeline
 
-#### A formula for presenting a chapter or paper
-
-1. You need a title slide (with your name, that of your advisor and/or co-authors and institution). Most institutions have logos, and they like you to use them.
-1. Several slides of introduction
-  + that put your study into the big picture
-  + explain variables in the context of existing literature
-  + explain the relevance of your study organisms
-  + give the context of your own study
-1. Your aims and hypotheses
-1. Methods and Materials
-  + You are going to have to skip details of the methods
-  + Be prepared to pare this down to the important stuff
-  ++ variables
-  ++ replicates
-  ++ analyses
-  + Use diagrams and images if you can
-1.Results & Discussion
-  + If you have a lot of results, then it makes sense to discuss them while they are still fresh in the minds of your audience. 
-  + If you had several hypotheses, then place reminders of what they are when you respond to them with the results
-1. Summary 
-  + state initial hypotheses and whether or not you accept them
-  + give some ideas for next steps
-
-### Know your story first, and then make your presentation
+## Know your story first, and then make your presentation
 
@@
 If the audience can follow your story, but wants to know the details at the end, then they can ask questions about the details.
-Indeed, you can tell your audience precisely this as a way of side-stepping chunks of (especially) methodologies.
+Indeed, you can tell your audience precisely this as a way of side-stepping chunks of (especially boring) methodologies.
 
@@
 
-### Slide layout
+## Slide layout
 
@@
 1. Everything on your slide must be mentioned in your presentation, so remove anything that becomes irrelevant to your story when you practice.
-1. Tables: you are unlikely to have large complex tables in a presentation, but presenting raw data or small words in a table is a way to lose your audience. Make your point in another way.
+1. Tables: you are unlikely to have large complex tables in a presentation as presenting raw data or small words in a table is a way to lose your audience. Make your point in another way.
 1. Use citations (these can go in a smaller font: 20 pt). I like to cut out the title and authors of the paper from the pdf and show it on the slide.
@@
 
-### Practice, practice, practice
+## Practice, practice, practice
 
-1. It can’t be said enough that you must practice your presentation. Do it in front of a mirror in your bathroom. In front of your friends. It's the best way of making sure you'll do a good job.
+1. It can’t be said enough that you must practice your presentation. Do it in front of a mirror in your bathroom, or in front of your friends. It's the best way of making sure you'll do a good job.
 1.

### 2021-07-20 2.22-transparency.Rmd
Add files via upload
# After copy edits

@@
 
-In this book, I have tried to consistently encourage you toward transparent and open practices in science.
+In this book, I consistently encourage you toward transparent and open practices in science.
 Here my aim is to introduce this topic and provide an overview of the main areas in which you can currently make a difference by opening up your research.
 As the Centre for Open Science explains (COS: [see here](https://youtu.be/9YuNGB3vNOw)), this will take a shift in the culture of the scientific community, which is why you need to understand and adopt a transparent approach from the outset of your research career.
-There are much better and more comprehensive guides and online resources so I encourage you to look through the literature cited here.
-Although I advocate hard for the COS, please be aware that this is not the only set of guidelines (see also [recent moves by the Public Library of Science](https://plos.org/open-science/preregistration/)) and [Peer Community In](https://peercommunityin.org/).
+There are much better and more comprehensive guides and online resources, so I encourage you to look through the literature cited here.
+Although I advocate hard for the COS, please be aware that this is not the only set of guidelines (see also [recent moves by the Public Library of Science](https://plos.org/open-science/preregistration/) and [Peer Community In](https://peercommunityin.org/)).
 
@@
 The effort put into your proposal is totally disproportionate if it’s never looked at again.
-Yet, this document is of historical importance, because it says what you think before ever doing the experiment and collecting or analysing the data.
-Thus, it can in future prove that you were not conforming to confirmation bias, for example by [HARKing](TypeI.html): rewriting your hypothesis after getting your results in order to get a significant result [@kerr1998harking; @forstmeier2017detecting].
+Yet, this document is of historical importance because it says what you think before ever doing the experiment, and collecting or analysing the data.
+Thus, it can in future prove that you were not conforming to confirmation bias, for example by [HARKing](#TypeI): rewriting your hypothesis after getting your results in order to get a significant result [@kerr1998harking; @forstmeier2017detecting].
 
@@
 Confirmation bias is bad because it violates the assumption that we are answering the hypothesis that we started with, or it could cause scientists to manipulate their data until a significant result is achieved.
-For example another form of confirmation bias is p hacking: repeating tests with different approaches in order to a significant result.
+For example another form of confirmation bias is [p hacking](#TypeI): repeating tests with different approaches in order to obtain a significant result.
 Rubin [-@rubin2020does] gives a good list and set of explanations.
@@
 
-What if new analyses appear during the period, or you are forced to change your experimental protocol.
+What if new analytical methods appear during the period, or you are forced to change your experimental protocol?
 Certainly, most analytical software will be updated over the course of your studies, and some might be superseded.
@@
 Logging the software that you intend to use will demonstrate the approach and type of analyses that you intend to take.
-While it is unlikely that you will ever be held to account for minor changes to equipment or protocol, things do change along the way.
+It is unlikely that you will ever be held to account for minor changes to equipment or protocol as things do change along the way.
 It doesn’t mean that you won’t be able to report your results, or that your research won’t be viable.
@@
 Although at first sight, preregistration appears to confine you to only testing ideas that you had before you started a study, actually the opposite is true.
-When you preregister your proposal (or research plan) it places a time-stamp and therefore a historical trail on your ideas.
+When you preregister your proposal (or research plan) it places a time-stamp and, therefore, a historical trail on your ideas.
 If as you go along with your study you gain new insight and ideas, you can register these with their own time-stamp.
-When it comes to analysing your data, you use a separate section labelled “exploratory analyses”.
-You can then apply post-hoc rules to these findings, which will not stop you from reporting them, but gets around the awkwardness of [HARKing](TypeI.html).
+When it comes to analysing your data, you use a separate section labelled 'exploratory analyses'.
+You can then apply post-hoc rules to these findings, which will not stop you from reporting them, but gets around the awkwardness of [HARKing](#TypeI).
 This serves to reinforce the distinction between prediction and postdiction.
@@
 It is probably too early to tell [@rubin2020does], but it is certainly a good place to start [@nosek2018preregistration].
-The more researchers know and subscribe to [transparency](transparency2.html) in their research, the more it will shift the culture in science for the better [@forstmeier2017detecting].However, there are more biases existing in science, and the system needs buy-in from stakeholders to promote equity by elimination of bias.
+The more researchers know and subscribe to [transparency](transparency2.html) in their research, the more it will shift the culture in science for the better [@forstmeier2017detecting].
+However, there are more biases existing in science, and the system needs buy-in from stakeholders to promote equity by elimination of bias.
 
@@
 If the RR is approved by reviewers then the study is, in principle, given the green light for publication whether or not the hypothesis posed is accepted or rejected.
-I say "in principle" because those same reviewers are shown the ms again once the results are in.
+I say 'in principle' because those same reviewers are shown the manuscript again once the results are in.
 They need to check that the methods proposed were followed, the analyses were conducted in the same way they were proposed, and that the conclusions are justified by the results.
-In terms of figure \@ref(fig:preregistration), Peer Community In are offering to organise the two yellow review arrows.
+In terms of Figure \@ref(fig:preregistration), Peer Community In are offering to organise the two yellow review arrows.
 In addition, there are a bunch of journals that have already signed up to accept RRs that are signed off after completion (notable among these is [PeerJ](www.peerj.com)).
@@
 
-When choosing where to archive your proposal (or any of your data, analyses, etc.) there are lots of platforms to choose from: [Bitbucket](https://bitbucket.org/product), [Figshare](https://figshare.com/), [Github](https://github.com/), [OSF](https://osf.io/), [Zenodo](https://zenodo.org/) and the list will undoubtedly grow.
-Making a decision about what you are going to use now may not require that you stick with this same platform for you entire career, but there are some things that you should consider:
+When choosing where to archive your proposal (or any of your data, analyses, etc.), there are lots of platforms to choose from: [Bitbucket](https://bitbucket.org/product), [Figshare](https://figshare.com/), [Github](https://github.com/), [OSF](https://osf.io/), [Zenodo](https://zenodo.org/) and the list will undoubtedly grow.
+Making a decision about what you are going to use now may not require that you stick with this same platform for your entire career, but there are some things that you should consider:
 
@@
 For example, you could make sure that notes taken during meetings with your advisor on different projects are logged onto the platform.
-This way you both have a record of what decision was made when.
+This way, you both have a record of what decision was made when.
 Remember that you can choose what you make accessible to the outside world.
@@
 Becoming familiar with the entire process now will be to your advantage, so I encourage you to read more about this topic.
-Sullivan et al [-@sullivan2019open] provide a nice overview about how to get started, but be sure to consult documentation at [OSF](https://osf.io/).
+Sullivan et al.
+[-@sullivan2019open] provide a nice overview about how to get started, but be sure to consult documentation at [OSF](https://osf.io/).

### 2021-07-20 2.23-funding.Rmd
Add files via upload
# After copy edits

@@
-## Generating funding for your PhD research {#funding}
+# Generating funding for your PhD research {#funding}
 
-To be honest, writing a funding proposal is an art in itself and should be the subject of an entirely different book written by somebody else.
-What is written here is only really scratching the surface so I suggest that if generating funding for your PhD research is going to be a large part of your work then you should look for another book.
+To be honest, writing a funding proposal is an art in itself and should be the subject of an entirely different book.
+What is written here is only really scratching the surface, so I suggest that if generating funding for your PhD research is going to be a large part of your work then you should look for an additional book.
 
-### Funding databases
+## Funding databases
 
 Funding is one of the most important aspects of doing science, and something that different emphasis is placed on in different parts of the world.
-In North America, in particular, graduate students are expected to be able to demonstrate that they are able to raise their own funds, as this is expected of them in their jobs as academics.
+In North America, in particular, some graduate students are expected to be able to demonstrate that they are able to raise their own funds, as this is expected of them in their jobs as academics.
 No matter where you are, your CV will be greatly improved by showing prospective employees (or labs) that you can generate your own funding.
@@
 
-### Reasons you might want to apply for funding:
+## Reasons you might want to apply for funding:
 
@@
 - You want to attend a conference or workshop. There are great opportunities for travelling around the world, but they cost money and international conferences are often very expensive.
-Publishing in Open Access journals.
+- Publishing in Open Access journals.
 I find the idea of putting cash into the pockets of publishers abhorrent, but you may have little choice.
@@
 - It is clear that without making an application, you won’t get any funding. But where do you start?
-Databases. There are many databases, but here we’ll mention 3 that are easy to use.
+Databases. There are many databases, but here are 3 that are easy to use.
 
-#### Mendeley:https://www.mendeley.com/ 
+### Mendeley:https://www.mendeley.com/ 
 This is probably my favorite. It’s easy to use and apply or remove filters. You can use the same login that you have for the referencing software and SCOPUS.
 
-#### Open 4 Research: https://www.open4research.eu 
+### Open 4 Research: https://www.open4research.eu 
 For this you’ll need an academic email address to register, but you’ll get access to their research database.
@@
 
-#### Research Professional:https://www.researchprofessional.com/ 
+### Research Professional:https://www.researchprofessional.com/ 
 You’ll need to create a login for this site, although you might get an in with your university address.
@@
 
-#### Opportunities in your department or societies:
-Probably the most likely place to get funding is where you already have an “in” (where you are already known).
+### Opportunities in your department or societies:
+Probably the most likely place to get funding is where you already have an 'in' (where you are already known).
 This will include professional societies where you are a member, or your department or university.

### 2021-07-20 2.24-fear.Rmd
Add files via upload
# After copy edits

@@
-## Fear of submitting written work {#fear}
+# Fear of submitting written work {#fear}
 
@@
 It is possible that you are suffering from **impostor syndrome**.
-This chapter is to help you overcome your fear of handing in your written work.
 
-### What is impostor syndrome? {#impostorsyndrome}
+## What is impostor syndrome? {#impostorsyndrome}
 
@@
 
-- Are you fearful that your work won’t be well received? 
+- Are you fearful that your work won’t be well received (Figure \@ref(fig:conflicted))? 
 - Is it really up to the standards that are required?
@@
 
-### Think of it this way
+## Think of it this way
 
@@
 
-### What can you do to help yourself overcome the fear of submitting written work?
+## What can you do to help yourself overcome the fear of submitting written work?
 
-1. Give it to a friend or colleague to read
+1. **Give it to a friend or colleague to read**
 1.
-Remember that most of what you write should be understandable to the majority of people who can read (think of articles in The Conversation as a reasonable reading/understanding level).
+**Remember that most of what you write should be understandable to the majority of people** around you.
 This means that you can give your written work to fellow postgraduates and postdocs and ask for feedback.
-1. Annotate your text with places where you are particularly unsure.
-If there’s a certain part that you are struggling with, annotate it (add comment in word) and point out that you are struggling with this particular section
-1. Ask for a meeting with your advisor
+1. **Annotate your text with places where you are particularly unsure**.
+If there’s a certain part that you are struggling with, annotate it (add comment) and point out that you are struggling with this particular section
+1.
+**Ask for a meeting with your advisor**.
 You can sit down together with your advisor and discuss the points where you are uncertain before handing in the work.
 Or, after you’ve written it you can ask for a session when you are given verbal feedback
-1. Produce a checklist
+1.
+**Produce a checklist**.
 If you specifically worry that there may be errors in what you write (grammatical, spelling, plagiarism), then make a checklist that you can tick off prior to submitting.
 Once your check list is done, don’t mess with the written work again (or you could add more errors), just hand it in!
-1. Set a deadline for yourself
-If you don’t already have one, having a set of deadlines that you give yourself to give your writing to work to colleagues and hand in to your advisor.
+1.
+**Set a deadline for yourself**.
+If you don’t already have one, having a set of deadlines that you give yourself to give your writing to colleagues and hand in to your advisor.
 If you know that you aren’t good with deadlines, share them with as many people as you can!
-1. Provide yourself with a reward for submitting your work. Rewards are a really simple way of helping you to do things. They help simplify otherwise apparently complex fears. 
-1. Tell your advisor about your fears
-There’s nothing like honesty. Your advisor may be able to cut you some slack, or might sit down with you and look at how the two of you can overcome this difficulty.
-If you have more than one advisor, there may be one who you are more confident to read your work, and you can suggest that reading is done in series (instead of in parallel - actually I’d advise this).
+1. **Provide yourself with a reward for submitting your work**. Rewards are a really simple way of helping you to do things. They help simplify otherwise apparently complex fears. 
 1.
-Talk to your advisor about what really gets you the most upset.
+**Tell your advisor about your fears**.
+There’s nothing like honesty.
+Your advisor may be able to cut you some slack, or might sit down with you and look at how the two of you can overcome this difficulty.
+If you have more than one advisor, there may be one who you are more confident to read your work, and you can suggest that reading is done in series (instead of in parallel - actually, I’d advise this).
+1.
+**Talk to your advisor about what really gets you the most upset**.
+In case it's something that your advisor does that makes you anxious.
 If you can’t do this face to face, then you could annotate it in a reply to their comment.
@@
 1.
-Ask your advisor about their fear of handing in and get them (and others in your lab) to share their stories.
+**Ask your advisor about their fear of handing in** and get them (and others in your lab) to share their stories (for example at a lab meeting).
 You might find that you have common ground to start sharing how these problems can be overcome.
@@
 You need to find a way of getting through the fear, and if it’s going to be a persistent problem, you’ll need to work out what works best for you.
+
+## When is it good enough to submit to my advisor? {#goodenough}
+
+Once you've read and revised the draft text three times yourself, and you feel that you can't really improve it any more, hand it over.
+Personally, I need to wait a day or more between drafts when I work on something else, or do some reading.
+I find that it really helps to think about writing while not writing.
+Run through the arguments in your head when you are doing something else.
+That could be time in the gym, out running, cycling or swimming.
+Or talk through the logic of your argument with a friend or colleague while having coffee.
+You’ll find that spending time thinking away from sitting in front of your word-processor will be really valuable in promoting productivity once you sit down again.
+
+If you are [fearful of handing in](#fear) your work, share it with others to give you confidence.
+Hand it to an office mate or other friend first.
+Your reader shouldn't have to be an expert in your field, but only needs to have a loose grasp on scientific writing to follow any paper.
+As a guide, any recent graduate of a biological sciences degree (BSc) should be able to understand.
+This makes it easy to find lots of potential readers for your manuscript before handing it in.

### 2021-07-20 2.25-formula.Rmd
Add files via upload
# After copy edits

@@
 Very few people can sit down at a desk and write an academic paper from beginning to end.
-For most of us it takes a lot of hard work over many weeks or even months to write something that can be submitted to peer review.
-In part two of this book I approach the writing of a chapter or a paper for peer review in the same formulaic way, and will refer to papers or chapters interchangeably.
+For most of us, it takes a lot of hard work over many weeks or even months to write something that can be submitted for peer review.
+In the third part of this book, I approach the writing of a chapter or a paper for peer review in the same formulaic way, and will refer to papers or chapters interchangeably.
 
@@
 Not only is it easier, but because it is so common and widespread, it also gives the reader a familiarity with reading your work.
-This means that editors are able to skim through and find exactly what makes your work worth publishing, and it means that those who are only semi-interested are more likely to cite your work.
+This means that editors are able to skim through and find exactly what makes your work worth publishing, and it means that those who are only semi-interested are more likely to find what they need to cite in your work.
 
@@
 If we were in my office, at this point in the meeting I usually go to my white board and start drawing funnels (as a metaphor) and blocks.
-However, others have much better diagrams and here I’ve adapted an idea from Brian McGill’s blog [@mcgill20165].
-But you can also find other useful material in other papers [-@turbek2016scientific; @plaxco2010art @tomaska2007teaching].
+However, others have much better diagrams and here I’ve adapted an idea from Brian McGill’s blog [-@mcgill20165].
+But you can also find other useful material in other papers [@turbek2016scientific; @plaxco2010art @tomaska2007teaching].
 If you are of the Twitter persuasion, you get lots of useful tips and links from WriteThatPhD and Write4Research.
@@
 
-(ref:formula) **Using a formulaic approach to writing your PhD chapter.** The formula of writing a paper or chapter that we will concentrate on in this book.
+(ref:formula) **Using a formulaic approach to writing your PhD chapter.** This diagram represents the formula of writing a paper or chapter that we will concentrate on in this book.
+Imagine the flow of the reader being like a liquid that passes from one section to the next.
 
@@
 __So now that you've got the idea that writing to a formula is a good idea, what next? __
-Next, I suggest that you begin to plan out your formula as a bullet-point outline.
+Next, I suggest that you begin to plan out your formula as a bullet-point outline (Figure \@ref(fig:outline)).
 Start by writing a general sentence for the subject of each paragraph.
@@
 Personally, I'd prefer to see the bare outline, and then a fleshed outline before you start the writing proper.
-This would allow the advisor and student to get on the same page to be sure that what is planned is thought out.
+This allows advisor and student to get on the same page to be sure that what is planned is thought out.
 
-Example of an outline draft of the [introduction](introduction.html):
-This outline is taken from Measey et al [-@measey2020cost]:
+(ref:outline) **Example of an outline draft of the introduction.** In this outline taken from Measey et al.
+(2020), paragraphs are paragraphs are indicated by the solid dots, and points to make inside the paragraphs by hollow dots.
+Each paragraph has an argument sketched out within it.
 
-__Introduction__
-
-* It is important to assess the impacts of invasive species from published studies in order to compare them
-  + Not all studies are equally rigorous
-  + Inferences from less robust studies should be treated with caution
-* More complex study designs tend to result in greater power to detect effect direction
-  + Increasingly complex designs are likely to cost more money to conduct
-  + Researchers are likely to conduct cheaper (small scale) experiments before large landscape scale ones
-* EICAT and SEICAT rely on studies carried out at scales that are appropriate to the invasive population
-  + Cheaper studies will result in impacts of lower magnitude, and/or confidence and not reflect the species’ true impact
-  + Poorer regions of the world might be expected to produce impacts of lower magnitude
-  + But EICAT is used to inform policy
-* There is evidence of inequality in existing impact assessments of birds and amphibians
-  + geographic disparity for birds
-  + This could come from funding disparities
-  + how can we achieve a representative impact level for all alien species, without enough peer reviewed studies?
-  + will the impact score of species be independent of research costs?
-* Aims
-  + Whether studies that result in higher impact scores cost more to conduct
-  + Whether impact is related to study complexity or time taken to conduct the study
-  + Whether EICAT and SEICAT scores demonstrate changes through time
-  + Whether different EICAT and SEICAT scores occur with equivalent confidence
-
-__Materials and Methods__
-
-Paragraphs are indicated by the solid dots, and points to make inside the paragraphs hollow dots.
+```{r outline, echo=FALSE, out.width = '95%', fig.cap="(ref:outline)"}
+ knitr::include_graphics("figures/outline.png")
+```
 
@@
 
-In the next stage, fleshing out, I ask that citations for statements above be added along with any examples. Below is an example of the citations added to paragraph 4 above.
+In the next stage, 'fleshing out', I ask that citations for statements above be added along with any examples.
+Figure \@ref(fig:outline-fleshed) is an example of the citations added to paragraph 4 above.
 
-- There is evidence of inequality in existing impact assessments of birds, amphibians and plants with lots of DD species (30%; Evans et al.
-2016; 38%; Kumschick et al.
-2017; 79%; Rumlerová et al.
-2016)
-  + Geographic disparity for birds (Evans and Blackburn 2020)
-  + Examples of birds from Evans et al (2020)
-  + This could come from funding disparities
-  + How can we achieve a representative impact level for all alien species, without enough peer reviewed studies?
-  + Will the impact score of species be independent of research costs?
+(ref:outline-fleshed) **Example of a fleshed out outline of a paragraph in the introduction.** In this outline taken from Measey et al.
+(2020), citations and relevant figures have been inserted for each of the statements.
+
+```{r outline-fleshed, echo=FALSE, out.width = '95%', fig.cap="(ref:outline-fleshed)"}
+ knitr::include_graphics("figures/outline-fleshed.png")
+```
 

### 2021-07-20 2.27-data.Rmd
Add files via upload
# After copy edits

@@
-## Data Management {#data}
+# Data Management {#data}
 
@@
 
-At the time of writing, it is very easy to keep copies of documents on the cloud, or even write straight onto the cloud.
+At the time of writing, it is very easy to keep copies of documents in the Cloud, or even write straight into the Cloud.
 Talk to your advisor about what options are available in your institution for backing up your work.
@@
 A really good option is to start working in one of the Open Science platforms.
-For example, OSF allows you to create a project where you can continue to use lots of different cloud based resources, and provides an integrated home for them (for an introductory video see here).
+For example, OSF allows you to create a project where you can continue to use lots of different Cloud based resources, and provides an integrated home for them.
 OSF is also a great platform as it is free to use, and as long as individual files are less than 5 Gb, OSF will host them.
-This means that for most projects, raw data files can be stored on the cloud at no cost to the project.
+This means that for most projects, raw data files can be stored in the Cloud at no cost to the project.
 
-### Should the cloud be the only copy of your data?
+## Should the cloud be the only copy of your data?
 
 No.
-I suggest that you keep a hard drive with a copy of your data that you regularly deposit on a monthly basis, or after significant bouts of data collection such as field trips or experiments.
+I suggest that you keep a hard drive with a copy of your data that you regularly update on a monthly basis, or after significant bouts of data collection such as field trips or experiments.
 A hard drive has the advantage that in the case of an internet outage, you always have access to your data.
@@
 
-### Backing up in the field
+## Backing up in the field
 
-If you are going to be doing lots of field work, then you might want to make a specific plan about backing up data to the cloud during the field time.
+If you are going to be doing lots of field work, then you might want to make a specific plan about backing up data to the cloud during field time.
 It may seem expensive, but I'm sure that it won't be as expensive as losing the data.
@@
 
-I still prefer field notes and data to be written into a book (preferably with waterproof paper such as [rite in the rain](https://www.riteintherain.com/)).
+I still prefer field notes and data to be written into a book (preferably with waterproof paper such as '[rite in the rain](https://www.riteintherain.com/)').
 This allows you to make a photographic backup of the hard copy.
-Similarly, you can design datasheets that you use in the field with a pencil, and then photograph them to make a hard copy.
+Similarly, you can design datasheets that you use in the field with a pencil, and then photograph them to make a soft copy from your hard copy.
 I don't like entering data directly into a laptop or other device.
@@
 
-### Make a data management plan, and stick to it
+## Make a data management plan, and stick to it
 

### 2021-07-20 3.01-Title.Rmd
Add files via upload
# After copy edits

@@
 That’s ok, and it’s not a waste of your time.
-In fact, the first draft is invaluable in getting your ideas and perspective down on paper, allowing you to better organise and sift through your thoughts to bring clarity to your manuscript.
+In fact, the first draft is invaluable in getting your ideas and perspective down on paper, allowing you to better organise and sift through your thoughts and bring clarity to your manuscript.
 Allow yourself a reward for finishing your first draft, something that you won’t do until it’s done.
-Now make life easy for yourself by changing the order of what you write.
 
+You can make life easier for yourself by changing the order of what you write.
 Although I have placed the contents of this chapter as you’d expect to see them in a typical paper, this is not the order I would suggest that you write them in.
@@
 Once you have this, it will help you to know the literature that is already there and so writing the introduction next seems logical.
-Of course, you may already have [your proposal](proposal.html) to work with, in which case you can start by reviewing what you already have and deciding whether or not you need to tweak this, or if you need a fresh start.
+Of course, you may already have [your proposal](#proposal) to work with, in which case you can start by reviewing what you already have and deciding whether or not you need to tweak this, or if you need a fresh start.
 
@@
 Again you should review your proposal and see whether there were any significant changes to your planned methodology and note them down.
-This will not be an issue even if you have [preregistered your proposal](transparency.html), you simply need to provide a rational explanation for why things changed.
+This will not be an issue even if you have [preregistered your proposal](#transparency), you simply need to provide a rational explanation for why things changed.
 
-Next write the results.
+Next, write the results.
 Remember that you are only looking to respond to your hypotheses, preferably using your (preregistered) proposal plans.
@@
 Getting authors' names correct, their correct addresses (many people have more than one affiliation), can be tricky.
-Check that you have all of the information required of a title page before you submit.
+Check that you have all of the information required for a title page before you submit.
 While you need the names of your co-authors to be displayed correctly, you do not need their titles (no Dr.
@@
 Having the correct name and address will be important to each author.
-Along with the author names and addresses you should record the ORCID [see part 3](ORCID.html) of each author.
+Along with the author names and addresses you should record the ORCID of each author.
 The ORCID is simply a unique identification code for individual researchers.
@@
 
-Note that the same information given in the title page will also be needed when you enter the information into the journal submission site.
-
-Work closely with your advisor, and make sure that your co-authors are aware of the correspondence that arrives from the journal.
-Use an address for correspondence that you know you check regularly.
-It;s also worth checking the spam of your email regularly so that you catch any correspondence that might have been misclassified.
-The golden rule for the corresponding author is the timely communication with the other authors when decisions have to be made.
-
-## The Title
+## The title
 
@@
 Your title will be your selling point, and your aim is to use it to draw your readership in.
-Once you’ve managed to invegle your readers to download your paper, your title is also their hook for remembering your paper in their database of thousands of others.
-Having their key word in your title will help here, and as ever with writing your challenge is to think like your reader.
+Once you’ve managed to inveigle your readers to download your paper, your title is also their hook for remembering your paper in their database of thousands of others.
+Having their key-word in your title will help here, and as ever with writing your challenge is to think like your reader.
 The best titles are those that sum up the entire study in five to seven words.
-This is best done in a narrative that tells the story ([see part 2](storytelling.html)) of your manuscript in its entirety.
+This is best done in a narrative that tells the story ([see part 2](#storytelling)) of your manuscript in its entirety.
 This may sound daunting, but you should get into the habit of summing up your story quickly (for friends, relatives and work colleagues).
@@
 
-Some people are excellent at writing titles that contain puns of well known phrases or sayings.
+Some people are excellent at writing titles that contain puns of well-known phrases or sayings.
 These can be brilliant, working both to inform what’s in the paper as well as providing some familiar input that helps retain them in memory.
@@
 It also helps to mix and match from a set of candidate titles.
-Once you have come up with something that looks good to you and your colleagues, test it by entering it into your database of choice (with the default being Google Scholar - see Part 2).
+Once you have come up with something that looks good to you and your colleagues, test it by entering it into your database of choice (with the default being Google Scholar).
 Your first 10 results should include a set of papers that you have likely cited in the upper area of your introduction.
@@
 - Don’t start by looking for the best title
-  + Write a number of candidate titles and ask your co-authors to vote for their favorites.
+  + Write a number of candidate titles and ask your co-authors to vote for their favourites.
 - The shorter and catchier your title can be the better: 6 words (±1) is an ideal.
@@
 - Do include your principal finding if possible
-- Include as many key words as you can
+- Include as many key-words as you can
 
 ### Things to avoid in your title
- - Don’t feel obliged to include [taxonomic terms](taxonomy.html) unless it is relevant or compulsory
+ - Don’t feel obliged to include [taxonomic terms](#taxonomy) unless it is relevant or compulsory
    + Some (taxonomically minded) journals will insist on the species name followed by the taxonomic authority, and/or the family and order, in your title
@@
 
-## Key Words
-The key words are a way for readers to find your content with searches.
+## Key-words
+The key-words are a way for readers to find your content with searches.
 Typically, the advice is to use words that are not in the title or abstract.
-This is because many databases have combined searching facilities for title, abstract and key words.
-I struggle to think of appropriate key words.
-Some of the big idea words and short phrases from the introduction.
-Then I tend to look at articles in the same genre, and see what key words they have used.
-As with the title, I suggest that you enter your chosen key words into your literature database of choice [see part 2](database.html) and see what comes back.
+This is because many databases have combined searching facilities for title, abstract and key-words.
+I struggle to think of appropriate key-words, and so make a list of some of the [big idea](#bigidea) words and short phrases from the introduction.
+Then I tend to look at articles in the same genre, and see what key-words they have used.
+As with the title, I suggest that you enter your chosen key-words into your literature database of choice [see part 2](#database) and see what comes back.
 You should see a group of papers that look wholly familiar and preferably those that are already cited in your manuscript.

### 2021-07-20 3.02-Abstract.Rmd
Add files via upload
# After copy edits

@@
-## The Abstract {#abstract}
+# The Abstract {#abstract}
 
@@
 
-### What is the abstract?
+## What is the abstract?
 The abstract is a concise paragraph that sums up the major points of your manuscript so that the potential reader will be able to assess whether or not they want to read the entire paper.
 It is an abstract of the entire document.
-It should not be abstract!
 
-### So what would a good abstract contain?
+## So what would a good abstract contain?
 A good abstract is a summary of the highlights of the paper.
@@
 It will likely require multiple iterations, and some intense word-smithing to make it as good as it can be.
-Abstracts almost always have a word limit, and that makes it challenging and means that you have to be concise [see part 2](concise.html).
+Abstracts almost always have a word limit, and that makes it challenging and means that you have to be concise [see part 2](#concise).
 
@@
 This means that it contains enough information that someone knows that it can be used to cite for a specific fact.
-Of course, these people should download and read the entire study, but there is no guarentee that they will.
+Of course, these people should download and read the entire study, but there is no guarantee that they will.
 
-### Where do you start?
+## Where do you start?
 Just like planning your writing in general, I’d suggest starting your abstract with an outline.
@@
 
-### When do you write your abstract?
+## When do you write your abstract?
 Although the abstract comes on page 1 of your manuscript, only try to write it once you’ve got to the end of the process of writing your manuscript so that you are aware of all the major points that you need to include.
-Maybe once you've [submitted your second draft](goodenough.html) of the manuscript to your advisor (given that you are both happy with the content), you can formulate the abstract.
+Maybe once you've [submitted your second draft](#goodenough) of the manuscript to your advisor (given that you are both happy with the content), you can formulate the abstract.
 
-### Do abstracts for conferences differ?
+## Do abstracts for conferences differ?
 Yes.
 It’s likely that an abstract for a talk will not be the same, unless you have already published the study: but even then it’s probably worth re-writing it.
-I’d suggest that your conference abstract be more descriptive and thought provoking, such that it is aimed toward the specific audience of the conference.
+I’d suggest that your conference abstract be more descriptive and thought-provoking, such that it is aimed toward the specific audience of the conference.
 Pose questions that you will answer in your talk.
 
-### Where do people go wrong in writing the abstract?
+## Where do people go wrong in writing the abstract?
 - The most common mistake is getting the balance wrong.
@@
 - Text from the main document is copied into the abstract.
-I’ve written about this [elsewhere](plagiarism.html), but you should never copy parts of your writing.
+I’ve written about this [elsewhere](#plagiarism), but you should never duplicate text within your document.
 You could start like this, but wordsmithing must change the words and their order substantially.
 Otherwise, when the first sentence of the abstract is the same as the first sentence of the manuscript, it will really turn off your readers.
-- Too many statistics or no numbers are also common mistakes. Adding numbers and statistics are important, especially when they take the place of more words. 
-- There is only time for one or two comments on the discussion, so pick out something important.
+- Too many statistics or their complete absence are also common mistakes. Adding numbers and statistics are important, especially when they can take the place of more words. 
+- There is only time for one or two comments on the discussion, so pick out the most important.
 Look through your written discussion and rank the points that you make in terms of interest to others.
@@
 - The abstract does not contain citations (or very rarely and only if unavoidable), so don’t put any in!
-  + If you have to have a citation in your abstract, then it must include sufficient information that it can be found without opening up the paper: _first author, abbreviated journal name, volume, first page_
+  + If you have to have a citation in your abstract, then it must include sufficient information that it can be found without opening up the paper: _first author, abbreviated journal name, volume, first page_.

### 2021-07-20 3.03-Introduction.Rmd
Add files via upload
# After copy edits

@@
-## The Introduction {#introduction}
+# The Introduction {#introduction}
 
-### Why do we need an introduction?
+## Why do we need an introduction?
 The introduction and discussion appear to be two common stumbling points for students writing chapters or manuscripts.
@@
 And second how to construct it.
-They are complimentary, and if written well can be read without the Materials and Methods or Results, with the reader still gaining all of the major information of the manuscript.
+They are complementary, and if written well can be read without the Materials and Methods or Results, with the reader still gaining all of the major information of the manuscript.
 
@@
 The meat starts with the introduction.
-And it’s not only important to get all the correct content in there, it’s also important not to give misleading information that might distract the reader.
+And it’s not only important to get all the correct content in there, but it’s also important not to give misleading information that might distract the reader.
 
-When you write the introduction to a paper you must not be tempted to stray away from what we are aiming to introduce.
+When you write the introduction to a paper, you must not be tempted to stray away from what we are aiming to introduce.
 The aim of our introduction is to explain to the reader the hypothesis you are testing, and the approach you have taken to test it.
 
-### So what do we put into the introduction?
+## So what do we put into the introduction?
 
-As I’ve said elsewhere, the [hypothesis](hypothesis.html) itself is made up of different parts, and each of these must be explained in the introduction.
+The [hypothesis](#hypothesis) itself is made up of different parts, and each of these must be explained in the introduction.
 We also need to understand the approach that you’ve decided to take in your study (e.g.
@@
 All of these decisions that you made were informed by the literature, as is your general understanding of the subject that you are studying.
-So making sure that you cite ([read about citing](citations.html)) the relevant literature is a key ingredient of the introduction.
+So making sure that you cite ([read about citing](#citations)) the relevant literature is a key ingredient of the introduction.
 
@@
 
-### And the construction…?
+## And the construction…?
 
-Previously, I’ve described the introduction like a funnel (Figure \@ref(fig:introduction)), where we channel the reader into our hypothesis by starting broad and ending up narrow ([see the formula](formula.html)).
+Previously, I’ve described the introduction like a funnel (Figure \@ref(fig:introduction)), where we channel the reader into our hypothesis by starting broad and ending up narrow ([see the formula](#formula)).
 This can draw criticism that journal articles are expected to fit a certain style, that this style is boring, and that what we really need is for people to write in a more exciting and varied way.
@@
 
-Table: (\#tab:introparagraphs) **Here is a guide to the contents of each paragraph of your introduction.** Although you should not feel constrained by this suggestion, it might help you get started when planning your bullet pointed structure of the introduction before you start writing.
+Table: (\#tab:introparagraphs) **Here is a guide to the contents of each paragraph of your introduction.** Although you should not feel constrained by this suggestion, it might help you get started when planning your bullet-pointed structure of the introduction before you start writing.
 
@@
 
-Remember that while this section on writing the introduction is short, much of [Part II](part2.html) concerns writing style that is applicable to your introduction.
-In particular, sections on composing your [hypothesis](hypothesis.html), writing a [paragraph](paragraph.html), and an [argument](argument.html) should be referred to when you are writing your introduction.
+Remember that while this section on writing the introduction is short, much of [Part II](#part2) concerns writing style that is applicable to your introduction.
+In particular, sections on composing your [hypothesis](#hypothesis), writing a [paragraph](#paragraph), and an [argument](#argument) should be referred to when you are writing your introduction.

### 2021-07-20 3.04-MaterialsMethods.Rmd
Add files via upload
# After copy edits

@@
-## The Materials and Methods {#methods}
+# The Materials and Methods {#methods}
 
@@
 
-### Sections or subheadings
+## Sections or subheadings
 
-As ever, I’m going to recommend inserting section headers into your Materials and Methods to help you break it up into bite sized chunks.
+As ever, I’m going to recommend inserting section headers into your Materials and Methods to help you break it up into bite sized chunks (Figure \@ref(fig:methods)).
 These are principally for the reader, but will also make your job easier.
@@
 Don’t be tempted to scatter variables and how you measured them throughout the Materials and Methods section.
-Keep them clear and maintain a logical flow, using a time-line if that makes sense.
+Keep them clear and maintain a logical flow, using a timeline if that makes sense.
 
-### Study organism or study system
+## Study organism or study system
 
-Having already written the [introduction](introduction.html), you will know that there isn’t too much space to write a lot of information about your study organism there.
-However, often there is important background information that is needed on the species, which would be distracting or lose the flow of the [introduction](introduction.html).
+Having already written the [introduction](#introduction), you will know that there isn’t too much space to write a lot of information about your study organism there.
+However, often there is important background information that is needed on the species, which would be distracting or lose the flow of the [introduction](#introduction).
 By having a ‘Study Organism’ section at the start of your Materials and Methods, you can add in all relevant information in a paragraph or two.
 
-For example, if you are working on African clawed frogs, _Xenopus laevis_, as a model organism you will want to use this space to describe why they are a good system for this particular investigation.
-Alternatively, your study could be detailing their invasion biology, in which case this would be the section where you can explain their distribution, that they are principally aquatic, or that they are invasive on four continents.
+For example, if you are working on mice, _Mus muculus_, as a model organism you will want to use this space to describe why they are a good system for this particular investigation.
+Alternatively, your study could be detailing their invasion biology, in which case this would be the section where you can explain their distribution, or that they are invasive on all continents.
 You probably won’t need all of these points, but just the aspects of this species’ biology that is relevant to your study.
 You should make sure you know whether or not the journal requires the taxonomic authority.
-If you are not sure, include it now ([see here for an explanation](taxonomy.html)).
+If you are not sure, include it now ([see here for an explanation](#taxonomy)).
 
-Similarly, if you are working on invasive zebra fish, _Danio rerio_, you can write in this section the known dates when they were introduced to Japan, and how we know that those animals came from.
+Similarly, if you are working on invasive zebra fish, _Danio rerio_, you can write in this section the known dates when they were introduced to Japan, and how we know where those animals came from.
 You can also give details of their natural range in South Asia.
@@
 
-### Animal husbandry / Growing Conditions / Culturing Techniques
+## Animal husbandry / Growing conditions / Culturing techniques
 If you have kept animals in the lab, grown plants or cultured bacteria in order to do your experiments, then you need to provide details of how you did this.
-For plants this will include all relevant information on their source, propogation, growth medium and climatic regimes.
+For plants, this will include all relevant information on their source, propagation, growth medium and climatic regimes.
 For animals, you need to include their source, welfare, housing, feeding, light:dark cycles, temperature, etc.
@@
 
-### Repetitive methodology
+## Repetitive methodology
 If you’ve used the same methodology for each part of your data collection (like DNA extraction, sequencing, etc.), you should have a section that explains all of this prior to any experimental manipulations done.
-I would suggest that this is detailed in the first relevant chapter of your thesis, and then referred to in later chapters.
+I would suggest that this is detailed in the first relevant chapter of your thesis, and then referred to in later chapters ([see section on generative recycling](#generative)).
 It’s best to go to look to see what other people have reported on in order to know what level of detail to include.
-For example, 20 years ago it would have been important to include detailed information of how you extracted DNA from tissues.
-Today, you can probably say that you used a standard extraction kit or salt extraction without saying more.
+For example, 20 years ago, it would have been important to include detailed information of how you extracted DNA from tissues.
+Today, you can probably say that you used a standard extraction kit without saying more.
 
-### Do you need to include formulae?
+## Do you need to include formulae?
 Some of your methodologies will include calculations that rest on mathematical formula which can be written in mathematical notation.
-Certainly, there are plenty of examples in the biological sciences, especially in modelling, where these formulas are needed because they are developed within the methodology in order to produce the [results](results.html).
+Certainly, there are plenty of examples in the biological sciences, especially in modelling, where these formulas are needed because they are developed within the methodology in order to produce the [results](#results).
 If your thesis involves this type of mathematical progression, then you will already know this is expected because all of the literature that you read will contain equations and formulas.
@@
 
-### Experimental manipulations and data collection
+## Experimental manipulations and data collection
 If you have done more than one kind of experiment, it’s probably a good idea to give each a separate subheading.
-Be consistent with these subheadings between the Materials and Methods and the  [results](results.html).
+Be consistent with these subheadings between the Materials and Methods and the  [results](#results).
 
 One very important point here is to fully explain the collection of all of the variables that are used in your analyses.
-Remember to use exactly the same names for your variables here, as you have in the [introduction](introduction.html) (and will use in your [figures](figures.html) and  [results](results.html) sections).
+Remember to use exactly the same names for your variables here, as you have in the [introduction](#introduction) (and will use in your [figures](#figures) and  [results](#results) sections).
 It’s very important that the reader gets this consistency across sections.
@@
 
-#### Equipment
+### Equipment
 
@@
 This is to help others that might want to buy the same equipment (although it’s a bit outdated in these days of multinational companies).
-You should try to get all of the model numbers of equipment as you do the work, so that you are not scrambling later on to find out what they were.
+You should try to get all of the model numbers of equipment as you do the work, so that you are not scrambling, later on, to find out what they were.
 It’s really quick and easy to take a picture of this at the time.
@@
 
-### Data analysis
-This section is getting increasingly longer as people do ever more fancy statistical manipulations.
+## Data analysis
+This section is getting increasingly longer as people do ever more intricate statistical manipulations.
 As above, it’s very important that you use the same names for the variables as you have introduced and collected them.
@@
 
-You should be familiar with the terms above from when you formulated your hypothesis (see [Part II](hypothesis.html) for a refresher).
+You should be familiar with the terms above from when you formulated your hypothesis (see [Part II](#hypothesis) for a refresher).
 
@@
 Some journals will have particular software conventions that you need to respect and adhere to.
-Also give a citation to R as well as the version that you used in your analysis.
+Also, give a citation to R as well as the version that you used in your analysis.
 If you don’t use the latest version of R, then your reviewers may want to know why, so always remember to update your version of R when you do your analysis.
 
-### Did you leave anything out?
+## Did you leave anything out?
 By following your Materials and Methods, anyone else should be able to repeat your work.

### 2021-07-20 3.05-Results.Rmd
Add files via upload
# After copy edits

@@
 ## The correct approach - responding to your hypotheses
-Yes, the easiest way to write the results is simply to have the questions or hypotheses that you posed at the end of the introduction and have them as subheadings for your results section.
+Yes, the easiest way to write the results is simply to have the questions or hypotheses that you posed at the end of the introduction and have them as subheadings for your results section (Figure \@ref(fig:results)).
 Once you have these subheadings, you simply assemble all of the results with respect to that hypothesis.
@@
 This means stating that one increases with the other (a direct relationship) or one increases when the other decreases (an inverse relationship).
+Note that other types of variable relationships exist.
 Once you have described the result, either make reference to the Table or Figure where this can be seen, or describe the results of the test in brackets.
@@
 Keep results that don't respond to your hypotheses for the Supplementary Information.
+
 ```{r results, echo=FALSE, out.width = '95%', fig.cap="(ref:results)"}
@@
 In genetic studies, for example, this might include the number of samples that were collected and successfully sequenced, and thus entered into the rest of the analyses.
-Your text can then be broken into sections that correspond with your questions, unless you only have one in which case further sections are not warranted.
+Your text can then be broken into sections that correspond with your questions, unless you only have one, in which case further sections are not warranted.
 
 ## Responding to hypotheses
-While I encourage you to present data that corresponds to hypotheses here in the results section, you should not formally accept or reject your hypothesis here.
+While I encourage you to present data that corresponds to hypotheses here in the results section, you should not formally accept or reject your hypothesis in the results.
 This is done in the discussion, and allows you to present caveats and other important issues.
 
-## Common Mistakes
+## Common mistakes
 - Having extra analyses in the results that you didn’t include in your methods.
 If you really didn’t anticipate conducting the extra tests, then these are actually _post hoc_ (after the event) tests (not anticipated before the study) and you should think carefully about how to treat these.
-Otherwise, you’ll need to r-ejig your methods to make sure that all tests are included.
+Otherwise, you’ll need to rejig your methods to make sure that all tests are included.
 - Meaningless precision.
@@
 For example, the output of your analysis might tell you that flight feathers in one population are 2.6484332 mm longer than those of another (± 0.0654562 mm), but in your methods you only use callipers that have a precision of 0.01 mm.
-Reporting results beyond the precision whereby they can be measured quickly becomes meaningless.
+Reporting results beyond the precision whereby they can be measured becomes meaningless.
 Whenever you obtain the output of a statistical package, you need to make sure that you use your practical head to re-interpret these results and think about what they actually mean.
@@
 - Repeating text in different sections.
-As you’ve seen elsewhere in this book, repeating text simply isn’t acceptable.
+As you’ve seen [elsewhere in this book](#plagiarise), repeating text simply isn’t acceptable.
 It doesn’t matter if you did the same thing many times, you can always write something in a different way.

### 2021-07-20 3.06-Discussion.Rmd
Add files via upload
# After copy edits

@@
 
-The discussion should be the final section of your manuscript or chapter that you are going to write (given that the [abstract](abstract.html) will come last).
+The discussion should be the final section of your manuscript or chapter that you are going to write (given that the [abstract](#abstract) will come last).
 This is because for the discussion you must already know all of the rest of the manuscript, starting with the hypothesis, which dictates what will go into all of the other sections.
@@
 I’d make this suggestion for all of the sections of your chapter or paper.
-It doesn’t take that long to do, and it provides an opportunity for you and your advisor to talk about the [results](results.html) of your work and discuss them together.
-Such discussions should be eye opening for both of you, and they provide a great opportunity for you and your advisor to get excited about the work you’ve done, your [results](results.html) and what they mean.
-I get a lot of enjoyment during these discussions, especially when sharing the excitement of the [results](results.html).
-Sharing thoughts before you start writing is important because by talking about it, you and your advisor are more likely to come to a consensus about the best way in which to interpret your [results](results.html).
-Conversely, presenting them with the discussion finished might not be the best way of pitching ideas to them.
-Once you've had time to discuss it, you might have a lot more work to do in rewriting what you've already produced.
+It doesn’t take that long to do, and it provides an opportunity for you and your advisor to talk about the [results](#results) of your work and discuss them together.
+Such discussions should be eye opening for both of you, and they provide a great opportunity for you and your advisor to get excited about the work you’ve done, your [results](#results) and what they mean.
+I get a lot of enjoyment during these discussions, especially when sharing the excitement of the [results](#results).
+Sharing thoughts before you start writing is important because by talking about it, you and your advisor are more likely to come to a consensus about the best way in which to interpret your [results](#results).
+Conversely, presenting them with the discussion finished might not be the best way of pitching your ideas.
 
-There is a convention that you do not refer to the [tables](tables.html) or [figures](figures.html) in the discussion.
-This is because they have already been referred to in the [results](results.html), and a reader should have already consulted them there.
+There is a convention that you do not refer to the [tables](#tables) or [figures](#figures) in the discussion.
+This is because they have already been referred to in the [results](#results), and a reader should have already consulted them there.
 Occasionally, you may want to have a new conceptual diagram in the discussion, but this is rare.
-Most often you do not refer to figures or tables and no new ones appear in the discussion.
+Most often, you do not refer to figures or tables and no new ones appear in the discussion.
 
-In general, the first and last paragraphs of the discussion are key to the reader, but the discussion must also consider caveats and limitations in the experimental design and interpretation of your [results](results.html), as well as providing a concise discussion of the results in the context of existing literature.
+In general, the first and last paragraphs of the discussion are key to the reader, but the discussion must also consider caveats and limitations in the experimental design and interpretation of your [results](#results), as well as providing a concise discussion of the results in the context of existing literature (Figure \@ref(fig:discussion)).
 This is also your opportunity to suggest new hypotheses and how they could be tested.
@@
 ## First paragraph of the discussion
-Your discussion begins by you responding to your hypothesis, clearly stating whether or not it has been accepted, and putting this into the wider context of the study (i.e.
-paragraph one or two of the [introduction](introduction.html) with relevant literature).
-You can then follow these statements by emphasising what you consider to be the most important finding, and explain how it adds to existing literature.
-However, don’t be tempted to over-interpret your results, or claim that they mean more than they do ([see section on speculation](speculation.html) below).
+Your discussion begins by you responding to your hypothesis, clearly stating whether or not you accept it, and putting this into the wider context of the study (i.e.
+paragraph one or two of the [introduction](#introduction) with relevant literature).
+You can then follow these statements by emphasising what you consider to be the most important finding, and explain how it adds to existing knowledge.
+However, don’t be tempted to over-interpret your results, or claim that they mean more than they do ([see section on speculation](#speculation) below).
 
@@
 ## To sub-section or not to sub-section the discussion?
-My preference is to plan the discussion before you write it, just as you did for the [introduction](introduction.html) using a bullet pointed list.
+My preference is to plan the discussion before you write it, just as you did for the [introduction](#introduction) using a bullet-pointed list.
 This will provide you with logical sub-section headers for the discussion under which you can write the first draft.
@@
 It is then sometimes wise to leave sub-sections in your discussion so that your reader can more easily follow the text.
-Where possible, these should be the same sub-sections that you have broken your [methods](methods.html) and [results](results.html) into, especially where these relate to specific hypotheses or aims.
+Where possible, these should be the same sub-sections that you have broken your [methods](#methods) and [results](#results) into, especially where these relate to specific hypotheses or aims.
 Or it may be more appropriate to discuss the different approaches separately, specifically when the literature that you refer to falls into different groups.
@@
 
-The discussion is about explaining the meaning of your [results](results.html) to the reader.
+The discussion is about explaining the meaning of your [results](#results) to the reader.
 I often find that people write a lot of inappropriate information in the discussion.
 Remember that this section is not going to provide background information, and is unlikely to bring up new topics that need introducing.
-It may be that your [results](results.html) prompt you to introduce a new area of research that wasn’t covered in your [introduction](introduction.html), and this is fine.
-But for the main part, you should discuss your [results](results.html) in the context of existing literature.
-You can expect that the literature that you use in your discussion will only partially overlap with your [introduction](introduction.html), with plenty of new citations.
-Similarly, it can be that discussing your [results](results.html) will mean that you end up with paragraphs that have no citations.
+It may be that your [results](#results) prompt you to introduce a new area of research that wasn’t covered in your [introduction](#introduction), and this is fine.
+But for the main part, you should discuss your [results](#results) in the context of existing literature.
+You can expect that the literature that you use in your discussion will only partially overlap with your [introduction](#introduction), with plenty of new citations.
+Similarly, it can be that discussing your [results](#results) will mean that you end up with paragraphs that have no citations.
 
@@
 In some respects, when writing the discussion your walk a tightrope between inferring the meaning of your results for a wider audience, and making claims that are too broad (Table \@ref(tab:discuss)).
-Overreaching in the discussion (and consequently [abstract](abstract.html)) is a common problem that examiners and reviewers regularly ask authors pull back on their claims.
+Overreaching in the discussion (and consequently [abstract](#abstract)) is a common problem that examiners and reviewers regularly ask authors pull back on their claims.
 
-When providing different sides of an [argument](argument.html), try to use your [results](results.html) to conclude that one side is supported more than the other.
+When providing different sides of an [argument](#argument), try to use your [results](#results) to conclude that one side is supported more than the other.
 If your results don't help with this particular point, then it could be that you are trying to discuss something that isn't directly related to the work.
-This is a very common problem in discussions, and a good test is asking yourself how your [results](results.html) add to the point you are trying to discuss.
+This is a very common problem in discussions, and a good test is asking yourself how your [results](#results) add to the point you are trying to discuss.
 If they don't leave it out and move on.
@@
 | Restrain your conclusions to evidence from your data | Make unsubstantiated claims that overreach your evidence |
-| Provide insight into unresolved questions, controversies and biases | [_Reductio ad absurdum_](argument.html)  |
+| Provide insight into unresolved questions, controversies and biases | [_Reductio ad absurdum_](#argument)  |
 | Explain the study design's limitations and uncertainties | Inflate the importance of your findings |
@@
 | Mention literature that contradicts your findings | Use your conclusions as a 'bully pulpit' |
-| Provide arguments that are logical and coherent | Erect a ['straw man' or 'false dichotomy'](argument.html) |
+| Provide arguments that are logical and coherent | Erect a ['straw man' or 'false dichotomy'](#argument) |
 | Make conclusions based on your logic | Make emotional appeals to the reader |
@@
 ## Caveats and limitations
-An important aspect of the discussion is to consider how the interpretation of your [results](results.html) may be incorrect.
-For example, if you have done an experiment, how well controlled was it and how well could it be considered to scale up to real world interactions?
+An important aspect of the discussion is to consider how the interpretation of your [results](#results) may be incorrect.
+For example, if you have done an experiment, how well controlled was it and how well could it be considered to scale up to real-world interactions?
 Could you have measured other variables?
@@
 Reviewers will often be unhappy with speculation in the discussion section.
-Speculation isn’t that hard to spot, as it occurs when you make claims for which your [results](results.html) have no foundation.
-I think that it is healthy to have one or two statements that are speculative, but clearly labelled as such.
+Speculation isn’t that hard to spot, as it occurs when you make claims for which your [results](#results) have no foundation.
+I think that it is healthy to have one or two statements that are speculative, but clearly label them as such.
 After all, after writing this paper, you are going to be one of the world experts in the topic, and thus your deeper understanding is often worth relating to the reader.
 However, I suggest that you speculate in combination with suggesting what work could be done in future.
-If you really feel that the point has to be made, you must clearly label it as speculative.
+Remember, if you really feel that the point has to be made, you must clearly label it as speculative.
 
 Perhaps an easier trap to fall into is over-interpretation.
-This is when you suggest that your [results](results.html) mean more than they do.
-It’s an easy trap to fall into, especially after setting up the study in relation to key topics in the discipline (presented in paragraph one or two of the [introduction](introduction.html)).
+This is when you suggest that your [results](#results) mean more than they do.
+It’s an easy trap to fall into, especially after setting up the study in relation to key topics in the discipline (presented in paragraph one or two of the [introduction](#introduction)).
 You will probably find it hard to see where you have over-interpreted, and this is something that having your work read by your advisor, or another colleague, will really help.
-You may then be asked to ‘tone down’ your claim, or to place it into the direct findings of your [results](results.html).
+You may then be asked to ‘tone down’ your claim, or to place it into the direct findings of your [results](#results).
 
@@
 ## Don’t beat up on others
-Your [results](results.html) may show that other researchers were wrong with their interpretation or findings.
+Your [results](#results) may show that other researchers were wrong with their interpretation or findings.
 Whatever you may think of them, never use your discussion to be disrespectful to other researchers or their work.
-This has been referred to as the “bully pulpit”.
+This has been referred to as the 'bully pulpit' or an [_ad hominem_ attack](#argument).
 As with all aspects of professional interactions, consider how you would like to be treated, and act accordingly.
-This is not to say that you shouldn’t point out mistakes that were made before, but be sure not to become emotive or insulting.
+This is not to say that you shouldn’t point out mistakes that were made before, but be sure not to be emotive or insulting.
 
-Things brings to mind Freud's narcissism of minor differences, in which he pointed out that bitter fights could start between groups when they in fact agree on 90% on issues.
+Things brings to mind Freud's narcissism of minor differences, in which he pointed out that bitter fights could start between groups when they, in fact, agree on 90% on issues.
 So it is likely to be true that someone you might disagree with one one point, is likely to be allied with you on nearly all of their other views and opinions.
@@
 The ‘where next’ aspect of your discussion is important as it may provide the reader with ideas for their own work.
-Of course, these are questions that you may wish to pursue in your own career, or they may be corroborating evidence from other disciplines that you will never undertake.
+Of course, these are questions that you may wish to pursue in your own career, or they may require corroborating evidence from other disciplines that you will never undertake.
 Either way, making pointers for continuing aspects of the research is an important component of the discussion.
-Providing new lines of research may also allow you to speculate about what you consider to be the most important angle of this topic now that you have presented your [results](results.html).
+Providing new lines of research may also allow you to speculate about what you consider to be the most important angle of this topic now that you have presented your [results](#results).
 This should be justifiable and not gratuitous.
@@
 ## Last paragraph
-The last paragraph of the discussion is your take home message.
-It’s a summary paragraph that sets out what you aimed to achieve, and what the new state of understanding of the topic is now that your [results](results.html) are out.
+The last paragraph of the discussion is your take-home message.
+It’s a summary paragraph that sets out what you aimed to achieve, and what the new state of understanding of the topic is now that your [results](#results) are out.
 This should include the key literature that can now be reconsidered.

### 2021-07-20 3.07-Acknowledgements.Rmd
Add files via upload
# After copy edits

@@
-## The Acknowledgements {#acknowledgements}
+# The Acknowledgements {#acknowledgements}
 
@@
 
-### Thank your funders
+## Thank your funders
 
-Any people or organisations that funded you or any component of your work.
+You should thank any people or organisations that funded you or any component of your work.
 Check with them they may have specific wording that they want you to use.
-Remember that your institution may not have given you money but may have supported you in other ways you should probably thank them.
-There are registers of funders, and you should use to see whether your funder is present and exactly how they should be acknowledged.
+Remember that your institution may not have given you money but may have supported you in other ways, and you should probably thank them.
+There are registers of funders, and you should use these to see whether your funder is present and exactly how they should be acknowledged.
 One example is the [Crossref Funder Registry](https://www.crossref.org/services/funder-registry/).
-Funders rely on being able to search [databases](databases.html) for their name and find all of the work that they have funded.
+Funders rely on being able to search [databases](#databases) for their name and find all of the work that they have funded.
 This means that you should use the particular wording that they supplied, especially if they specify this in their funding letter.
@@
 
-### People and organisations to thank
+## People and organisations to thank
 
- - Try to remember all of those people who helped you doing fieldwork, or in the laboratory. Especially laboratory technicians without whom a lot of the work that we do isn’t possible.
+ - Try to remember all of those people who helped you doing fieldwork, or in the laboratory.
+Especially laboratory technicians (if they are not also authors) without whom a lot of the work that we do isn’t possible.
  - People that may have helped administer your funds or other office administration. 
  - Organisations that gave you permission to do your work.
-In some journals the acknowledgements is the correct place to put your permit numbers along with the names of the organisations who provided them.
+In some journals, the acknowledgements are the correct place to put your permit numbers along with the names of the organisations who provided them.
  - Increasingly, permission from ethics committees is placed in the materials and methods but do check with the journal instructions to authors.
@@
 
-### Who not to thank (in your manuscript acknowledgements)
+## Who not to thank (in your manuscript acknowledgements)
  - You do not need to thank anyone who is an author. The exception is in your thesis acknowledgements, where you should mention them and their support.
@@
 
-If in doubt it's probably a good idea to be as inclusive as you can in the acknowledgements.
+If in doubt, it's probably a good idea to be as inclusive as you can in the acknowledgements.
 

### 2021-07-20 3.08-References.Rmd
Add files via upload
# After copy edits

@@
 
-Having the references at the end of your chapter or publication is actually very important because in-text citations are not sufficient to determine exactly which paper you are referring to.
+Having the references at the end of your chapter or publication is actually very important because in-text citations often are not sufficient to determine exactly which paper you are referring to.
 Once you become very familiar with your own specific field you will find that you do glance at the references to make sure that the authors are citing the papers you think they are.
-It's also a great place to learn about literature that you don't know about already, and for this you need the full citation in order to look up the paper.
+It's also a great place to learn about literature that you don't know about already, and for this, you need the full citation in order to look up the paper.
 
@@
 Do not leave this until the end.
-It has happened to me many times that I have written an in-text citation, did not paste the citation of the end of the document, and by the time I had finished writing, had forgotten what I had read and where I had read it.
+It has happened to me many times that I have written an in-text citation, did not paste the citation at the end of the document, and by the time I had finished writing, had forgotten what I had read and where I had read it.
 This then leads to a lot more time wasting trying to find that or another relevant citation.
 
-[Earlier in this book](citations.html), I provided a guide to the differences between Vancouver and Harvard referencing styles.
+[Earlier in this book](#citations), I provided a guide to the differences between Vancouver and Harvard referencing styles.
 Here I provide the output for the references that were given in that section to demonstrate the way in which these different referencing styles take up very different amounts of space.
@@
 ## Vancouver style {#vancouver}
-The following references follow the text written in [Vancouver style](citations.html) in the section on citations.
-Note that the order of citations here corresponds to the order in which they are cited.
+The following references follow the text written in [Vancouver style](#citations) in the section on citations. Note that the order of citations here corresponds to the order in which they are cited.
 
->^1^Wake DB, Vredenburg VT. Are we in the midst of the sixth mass extinction? A view from the world of amphibians. _Proceedings of the National Academy of Sciences_. 2008;105(Supplement 1):11466–73.
+>^1^Wake DB, Vredenburg VT. (2008) Are we in the midst of the sixth mass extinction? A view from the world of amphibians. _Proceedings of the National Academy of Sciences_. 105(Supplement 1):11466–73.
 >
->^2^Collins JP, Crump ML, Lovejoy III TE. _Extinction in our times: global amphibian decline_. Oxford University Press; 2009.
+>^2^Collins JP, Crump ML, Lovejoy III TE. (2009) _Extinction in our times: global amphibian decline_. Oxford University Press.
 >
->^3^Pimm SL, Jenkins CN, Abell R, Brooks TM, Gittleman JL, Joppa LN, et al. The biodiversity of species and their rates of extinction, distribution, and protection. _Science_. 2014;344(6187):1246752.
+>^3^Pimm SL, Jenkins CN, Abell R, Brooks TM, Gittleman JL, Joppa LN, et al.
+(2014) The biodiversity of species and their rates of extinction, distribution, and protection.
+_Science_.
+344(6187):1246752.
 >
->^4^Kupferberg SJ. Bullfrog (*Rana catesbeiana*) invasion of a California river: the role of larval competition. _Ecology_. 1997;78(6):1736–51.
+>^4^Kupferberg SJ. (1997) Bullfrog (*Rana catesbeiana*) invasion of a California river: the role of larval competition. _Ecology_. 78(6):1736–51.
 >
 >^5^Dufresnes C, Dubey S, Ghali K, Canestrelli D, Perrin N.
-Introgressive hybridization of threatened European tree frogs (*Hyla arborea*) by introduced _H.
+(2015) Introgressive hybridization of threatened European tree frogs (*Hyla arborea*) by introduced _H.
 intermedia_ in Western Switzerland.
 _Conservation Genetics_.
-2015;16(6):1507–13.
+16(6):1507–13.
 >
 >^6^Berger L, Speare R, Hyatt A.
-_Chytrid fungi and amphibian declines: overview, implications and future directions.
+(1999) _Chytrid fungi and amphibian declines: overview, implications and future directions.
 Declines and disappearances of Australian frogs_ Environment Australia, Canberra.
-1999;1999:23–33.
+1999:23–33.
 >
->^7^Daszak P, Cunningham AA, Hyatt AD. Infectious disease and amphibian population declines. _Diversity and Distributions_. 2003;9(2):141–50.
+>^7^Daszak P, Cunningham AA, Hyatt AD. (2003) Infectious disease and amphibian population declines. _Diversity and Distributions_. 9(2):141–50.
 >
->^8^La Marca E, Lips KR, Lotters S, Puschendorf R, Ibanez R, Rueda-Almonacid JV, et al.
-Catastrophic population declines and extinctions in Neotropical harlequin frogs (Bufonidae: *Atelopus*) _Biotropica: The Journal of Biology and Conservation_.
-2005;37(2):190–201.
+>^8^La Marca E, Lips KR, Lötters S, Puschendorf R, Ibanez R, Rueda-Almonacid JV, et al.
+(2005) Catastrophic population declines and extinctions in Neotropical harlequin frogs (Bufonidae: *Atelopus*) _Biotropica: The Journal of Biology and Conservation_.
+37(2):190–201.
 >
 >^9^Martel A, Spitzen-van der Sluijs A, Blooi M, Bert W, Ducatelle R, Fisher MC, et al.
-*Batrachochytrium salamandrivorans* sp.
+(2013) *Batrachochytrium salamandrivorans* sp.
 nov.
@@
 _Proceedings of the National Academy of Sciences_.
-2013;110(38):15325–9.
+110(38):15325–9.
 >
@@
 The order of these references is alphabetical.
-If a paper has the same authors the oldest one usually comes before subsequent papers.
+If a paper has the same authors the oldest paper usually comes before subsequent papers.
 
@@
 Full details are always provided by the journal (in their **Instructions to Authors**), and you can refer to these to see how your references need to be formatted.
-I also provide some pointers towards using [reference managers](referencesoftware.html).
+I also provide some pointers towards using [reference managers](#referencesoftware).
 These can save you a lot of time, if you've already invested time in setting them up.
@@
 I really believe that on first submission of your manuscript you should be able to submit your references in any format.
-it really only makes sense for you to do all the work formatting them if you aren't going to pay to have published the paper.
+it really only makes sense for you to do all the work formatting them if you aren't going to pay to publish the paper.
 An increasing number of journals do allow you to use any referencing style (within the generic Vancouver or Harvard format) when you first submit.
@@
 
-There are lots of anal academics out there who will delight in looking for every missing full-stop, comma and capital-letter missing from your references.
+There are lots of anal academics out there who will delight in looking for every missing full-stop, comma and capital letter missing from your references.
 Perhaps it's because they feel secure pointing out these errors, but are incapable of knowing whether or not you've written good or bad science.
-Please don't join them and if you can look for a better world where you spend your valuable time doing more important things.
+Please don't join them and, if you can, look for a better world where you spend your valuable time doing more important things.
 
@@
 
-However, DOIs cannot replace references, otherwise we'd need to be able to click on every link all the time, and couldn't read any paper without a connection to the internet.
+However, DOIs cannot replace references, otherwise, we'd need to be able to click on every link all the time, and couldn't read any paper without a connection to the internet.
 It's still really useful to be able to read a formatted reference at the end of a paper.
 
-### DOI Tools
+### DOI tools
 

### 2021-07-20 3.09-Tables.Rmd
Add files via upload
# After copy edits

@@
-## Tables {#tables}
+# Tables {#tables}
 
@@
 | Make your headings clear and concise. | Omit units of measurement |
-| Summarise extensive data. | Try to include every data point (that's what [Supp Info](SuppInfo.html) is for). |
-| Use tables to provide repeated statistical information on tests or models. | Feel the need to include all of these in the ms  (that's what [Supp Info](SuppInfo.html) is for). |
-| Provide a full table legend that explains exactly what is in it (see [figures](figures.html)).
+| Summarise extensive data. | Try to include every data point (that's what [Supp Info](#suppInfo) is for). |
+| Use tables to provide repeated statistical information on tests or models. | Feel the need to include all of these in the ms  (that's what [Supp Info](#suppInfo) is for). |
+| Provide a full table legend that explains exactly what is in it (see [figures](#figures)).
 | Make tables so large that they have to run over several pages.
-If this can go into [Supp Info](SuppInfo.html) then it should.
+If this can go into [Supp Info](#suppInfo) then it should.
 |

### 2021-07-20 3.10-Figures.Rmd
Add files via upload
# After copy edits

@@
-## Figures
-
-I consider myself to be a visual person, somebody who understands more by looking at pictures than by reading words.
-Not everyone is the same but there are a lot of people who can derive more information from a good figure.
-Therefore for me, a figure is an opportunity to convey maximum amounts of information.
-It can be a relatively simple figure like a graph or bar chart, a map, a conceptual diagram, or my favourite, a composite figure.
+# Figures
 
@@
 Claus points out that while there are ways in which you can go wrong when plotting data (such as drawing lines to infer relationships that are not there), the skills in data visualisation are more in the aesthetic and interpretive environment of what tools to use that make the data easiest to interpret.
-How will the figure enhance understanding of the data?
+How will the figure enhance understanding of the data and respond to your [hypotheses](#hypothesis)?
 
-Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’ books is R ([see part 1](useR.html)), and the many packages that have been developed for use within the R environment [@rcoreteam2021r].
-The distinct advantages that this produces are the almost limitless additions and details that you can make to these figures to produce exactly what you want, in a [vector format](vector.html).
+Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’s books is R ([see part 1](#useR)), and the many packages that have been developed for use within the R environment [@rcoreteam2021r].
+The distinct advantages that this produces are the almost limitless additions and details that you can make to these figures to produce exactly what you want, in a [vector format](#vector).
 The drawback is that you might need to invest some time to learn how the graphics package works, and how to manipulate it to do what you want.
@@
 
-### Graphs
+## Graphs
 
-You should be aware of all the things that make up a good graph already. here I'm going to provide some extra pointers on what I think is important and should be considered when drawing a good graph:
+You should be aware of all the things that make up a good graph already. Here I'm going to provide some extra pointers on what I think is important and should be considered when drawing a good graph:
 
@@
 - __The scale__.
-make sure that the numbers of the scale are large enough to read even when the figure is reduced the size that it will be printed on the page.
+Make sure that the numbers of the scale are large enough to read even when the figure is reduced the size that it will be printed on the page.
 This includes the text that crossed the axes.
@@
 This can be done using the jitter function in R.
-- Don’t draw relationships that aren’t there.
+- __Don’t draw relationships that aren’t there__.
 The ease of plotting graphs makes it possible to display relationships between data points that don’t exist.
@@
 Be especially careful when plotting lines that you can infer a continuous value between points plotted.
-- Compiling graphs.
-if you have a series of graphs that share the same x or y axis then consider compiling them together.
+- __Compiling graphs__.
+If you have a series of graphs that share the same x- or y-axis then consider compiling them together.
 This essentially turns them into a composite figure, but it often means that readers can compare data easily from one graph to another.
-remember that you will have to make sure that the scales on the shared access are all the same.
-- Use consistent symbols and colours throughout your thesis.
+Remember that you will have to make sure that the scales on the shared access are all the same.
+- __Use consistent symbols and colours throughout your thesis__.
 Your thesis will likely plot data on similar variables in many ways throughout.
@@
 - Colour.
-The reason for all the fuss about not having colour in figures is mostly to do with the extra cross cost involved in printing them.
-Typically these days it is possible to use colour in figures and for these to be reproduced without any extra charge online.
+The reason for all the fuss about not having colour in figures is mostly to do with the extra cost involved in printing them.
+Typically these days, it is possible to use colour in figures and for these to be reproduced without any extra charge online.
 Thus if the data that you have is best shown with colour then use colour.
 There are some things to be mindful of when using colour.
-  + Colour blindness: This affects quite a large portion of the population, particularly men. To avoid using colours that can be easily confused simply look them up for example here
+  + Colour blindness: This affects quite a large portion of the population, particularly men. To avoid using colours that can be easily confused simply look them up.
   + Printing colour in monochrome: some people still like to print out papers in order to read them, annotate them or share them at a journal club.
-using a monochrome printer many colours are indistinguishable.
-you can therefore change the hue or intensity of a colour to make sure that it can be distinguished when it is printed.
+Using a monochrome printer many colours are indistinguishable.
+You can therefore change the hue or intensity of a colour to make sure that it can be distinguished when it is printed.
   + Aesthetic quality: some colours don't go well together.
-a lot of people have already determined this and you can find a good guide [here](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3).
+A lot of people have already determined this and you can find a good guide [here](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3).
 
-### Maps
+## Maps
 There are many reasons why you might need to include a map in your paper when writing for biological sciences.
@@
 Especially with QGIS you can watch ‘how to’ videos on the internet to learn most of what you need to know.
-things to remember when you're drawing a map:
-If your map is a detailed view of an area then consider a small context map in the corner showing where your area is
+Things to remember when you're drawing a map:
+
+- If your map is a detailed view of an area then consider a small context map in the corner showing where your area is
 - Include the scale
@@
 
-#### Maximising the content of your figure
+### Maximising the content of your figure
 Sometimes you will have the option of including more information in a figure than the original plan that you started with.
-This is perhaps especially true when it comes to producing maps with GIS, because you have the option of adding extra data layers.
+This is perhaps especially true when it comes to producing maps with GIS because you have the option of adding extra data layers.
 But remember that this information should not be gratuitous.
-In figure \@ref(fig:not-fancy), we can see a map that shows a number of sampling sites.
+In Figure \@ref(fig:not-fancy), we can see a map that shows a number of sampling sites.
 The map has a scale bar and north arrow as needed.
@@
 However, it is relatively simple to add more information to this map in the form of layers.
-In figure \@ref(fig:fancy-map), a digital elevation model (DEM) allows us to see the relative elevation at the sampling points, while transparent colours provide information about the different biomes in which the samples were made.
+In Figure \@ref(fig:fancy-map), a digital elevation model (DEM) allows us to see the relative elevation at the sampling points, while transparent colours provide information about the different biomes in which the samples were made.
 Lastly, I've used a piece of sea to insert an image of the species sampled.
@@
 (ref:fancy-map) **Maps tend to have a lot of space, and GIS provides the opportunity to add layers to give rich information, as well as improving aesthetics.** This map shows lots more information about the biomes in which each of the sample sites (red dots) falls, the elevation and an image of the species sampled.
-Inset of Africa (top left) shows the area sampled with a red square.
+The inset of Africa (top left) shows the area sampled with a red square.
 I've also managed to fit in an image of the species sampled into the sea (bottom right).
@@
 
-### Conceptual diagram
+## Conceptual diagram
 Conceptual diagrams or figures are very useful in the introduction or methods when you have undertaken a complex study design.
@@
 
-I encourage my students to draw conceptual diagrams of how the thesis chapters relate to one another, and then to include this diagram in the introduction to the thesis to help examiners and any other readers to understand how the chapters are interrelated.
+I encourage my students to draw conceptual diagrams of how the thesis chapters relate to one another (Figure \@ref(fig:thesis-construction)), and then to include this diagram in the introduction to the thesis to help examiners and any other readers to understand how the chapters are interrelated (see [Part IV](PhDintro.html)).
 
@@
 - Make use of arrows to show how ideas are related
-- Look at other conceptual diagrams especially those that resonate with you as a reader
+- Look at other conceptual diagrams, especially those that resonate with you as a reader
 
-### The composite figure
+## The composite figure
 For me, composite figures started when I noticed other papers were including small pictures of species in phylogenies.
-As a reader I really liked this way of understanding more about the animals that were being studied.
+As a reader, I really liked this way of understanding more about the animals that were being studied.
 The composite figure also comes about by making use of blank spaces that naturally occur within figures.
@@
 
-#### Ideas for composite figures:
-include a picture of the subject species in a small corner of blank space in one of your figures.
-For example, maps with a large area of sea (If your subject is a terrestrial species) are especially good for this (see Figure \@ref(fig:fancy-map)).
+### Ideas for composite figures:
+Include a picture of the subject species in a small corner of blank space in one of your figures.
+For example, maps with a large area of sea (if your subject is a terrestrial species) are especially good for this (see Figure \@ref(fig:fancy-map)).
 
-### Sending your graphics file to the publisher
+## Sending your graphics file to the publisher
 One consistent problem is the lack of understanding that authors have for different file types.
-I can be rather confusing as these are stated in lots of different ways, and often in units that are not intuitive.
+It can be rather confusing as these are stated in lots of different ways, and often in units that are not intuitive.
 Journals usually have long and detailed explanations of how to do this in their instructions to authors, yet this is one of the most common difficulties that authors have.
@@
 Remember that each figure needs a single file, so if you have produced a composite figure, then you will need to output this as a single graphics file.
+You will need to send them a generic file type (e.g.
+eps, jpg) and not something specific to the software that you used (e.g.
+ai, cdr).
 
-#### Vectors vs bitmaps {#vector}
-The biggest difference among image types are the difference between vectors (a line drawn between two points) and bitmaps (using pixels to create images).
+**Colour** is usually free to produce online, but you may need to pay for printed production. Make sure that your figure makes sense in monochrome if you intend the colour to be online only.
+
+### Vectors vs bitmaps {#vector}
+The biggest difference among image types is the difference between vectors (a line drawn between two points) and bitmaps (using pixels to create images).
 Vector diagrams will produce superior images on screen that will allow you to zoom in without loss of resolution (e.g.
@@
 
-#### Sizing your figure
+### Sizing your figure
 When you export your graphic from the graphics package that you are using to draw it, specify the resolution (most often in dpi: dots per inch) given by the publisher.
@@
 It’s always possible to draw at a higher resolution and convert to a lower resolution.
-Low resolution images need to be redrawn to make them higher resolution.
+Low-resolution images need to be redrawn to make them higher resolution.
 
@@
 
-### The figure legend
-The most important point with the figure Legend is that it should be a stand-alone explanation of what is in the figure.
+## The figure legend
+The most important point with the figure legend is that it should be a stand-alone explanation of what is in the figure.
 The reader should not have to refer to the text in order to understand what is in the figure.
@@
 The explanation of any variables or categories needs to be provided.
-But it should not be taken to absurd levels, is not a reason to repeat everything that's in the materials and methods.
+But it should not be taken to absurd levels, and is not a reason to repeat everything that's in the materials and methods.
 Take a look at some figure legends that are already published in the journal that you are hoping to submit to and you should have a good idea of what to write.
@@
 Remember that the figure legend is placed underneath the embedded figure in your thesis.
-This means that the itself graphic should not have any title.
+This means that the graphic itself should not have any title.
 Traditional submissions to journals require figure legends to be at the end of the manuscript, before the figures themselves.
 Many journals now allow you to embed a lower resolution image of the figure on first submission together with the legend underneath in order to facilitate peer review.
-Check on the requirements for you institution or journal before finalising this.
+Check on the requirements for your institution or journal before finalising this.
 
-A trend with online journals is that each figure (and table) needs a topic or title sentence, followed by a description.
+A trend with online journals is that each figure (and [table](#table)) needs a topic or title sentence, followed by a description.
 This is a good way forward with any figure legend.

### 2021-07-20 3.11-Credit.Rmd
Add files via upload
# After copy edits

@@
-## Who did what? {#credit}
+# Who did what? {#credit}
 
@@
 How do you decide whether or not someone did enough to be included on an author line.
-This is a very important point that we cover elsewhere.
+This is a very important point.
 All that I will say here is that this should have been decided early on in the study.
@@
 
-### CRediT where it’s due
+## CRediT where it’s due
 
@@
 
-If you want to know more about each of these roles and how they Are defined in the CRediT taxonomy then please visit the site.
-You can also ‘claim’ these roles under your [Rescognito profile](https://rescognito.com/), linked to your ORCID account.
+If you want to know more about each of these roles and how they are defined in the CRediT taxonomy then please visit the site.
+You can also ‘claim’ these roles under your [Rescognito profile](https://rescognito.com/), linked to your [ORCID](#ORCID) account.
 
-Even if the journal that you are submitting to does not use the credit taxonomy I would suggest that you use this as the basis for your saying who did what in the publication.
+Even if the journal that you are submitting to does not use the credit taxonomy.
+I would suggest that you use this as the basis for your saying who did what in the publication.
 See a blog article on this [here](http://crosstalk.cell.com/blog/giving-authors-the-credit-they-deserve).
-Otherwise you could end up with what is known as hyperauthorship, a phenomenon evidenced by massive co-authorship levels [@cronin2001hyperauthorship].
+Otherwise, you could end up with what is known as hyperauthorship, a phenomenon evidenced by massive co-authorship levels [@cronin2001hyperauthorship].
 
-### Ordering the author line up 
-In Biological Sciences it is traditional for the author who did most of the work to take the first position.
+## Ordering the author line up 
+In Biological Sciences, it is traditional for the author who did most of the work to take the first position.
 For you and your thesis you are most likely to be the first author, certainly for all of your thesis chapters.
@@
 The last position is normally for the head of the laboratory in which the work was undertaken.
-This would usually be your advisor, but could be another collaborator, if for example, you went to do some work in their laboratory.
+This would usually be your advisor, but could be another collaborator, if, for example, you went to do some work in their laboratory.
 Funding also plays a role here, and if you are unsure about who should be listed last, then ask your advisor.
@@
 One colleague is always first to suggest alphabetical order as he regularly ends up as last author.
-If it isn’t part of your thesis, should you still be an author?
 

### 2021-07-20 3.12-SuppInfo.Rmd
Add files via upload
# After copy edits

@@
-## Supplementary Material {#suppinfo}
+# Supplementary Material {#suppinfo}
 
@@
 
-### What can go into Supplementary Material?
-Practically anything!
+## What can go into Supplementary Material?
+Practically anything can go into supplementary material.
 
@@
 
-### There has been a big movement toward transparency and repeatability of scientific studies
-This is a really positive movement in science and it should receive all of our support as it is part of the greater scientific project [see part 2](transparency.html).
+## Toward transparency and repeatability of scientific studies
+This is a really positive movement in science and it should receive all of our support as it is part of the greater scientific project ([see part 2](#transparency)).
 Being proud of your work is being proud of the entire product, including the data and the analysis.
@@
 Not sure if you haven’t made some mistakes?
-We all make mistakes and you shouldn’t think that you will have an entirely fault free career.
+We all make mistakes and you shouldn’t think that you will have an entirely fault-free career.
 Whenever and wherever possible conduct your professional due diligence and be open and honest with your research.
-Science is a career when we should welcome errors and mistakes that we have made in the past.
+Science is a career when we should openly acknowledge errors and mistakes that we have made in the past.
 There’s no problem with this.
@@
 By doing this you ensure that any people that want to use your work in the future have full access to it.
-Your data deposit will be given a DOI.
-You can use this DOI in your publication so that readers know where to find the data.
+Your data deposit will be given a [DOI](#DOI).
+You can use this [DOI](#DOI) in your publication so that readers know where to find the data.
 
@@
 It’s a great platform, and so much more secure than using a publisher.
-Another example is the [Open Science Framework (OSF)](https://osf.io/) that aims to be a one stop shop for all of your storage needs.
-OSF best practices can be found here.
+Another example is the [Open Science Framework (OSF)](https://osf.io/) that aims to be a one-stop shop for all of your storage needs.
 You will need to make sure that your data is correctly sorted with sufficient metadata for someone else to understand [see @roche2015public].
 
-Once we did statistical analysis with a GUI and a point and click approach.
+Once, we did statistical analysis with a GUI and a point and click approach.
 Forget to click a checkbox and the results were not repeatable.
 As the analyses have become increasingly sophisticated with more and differing pieces of software, it is becoming increasingly important to record the code that you write that gets your results together with your data.
-This is the difference between making your results and therefore your study repeatable.
+This is the difference between making your results and, therefore, your study repeatable.
 
@@
 
-### Why stop with just depositing your data and analyses online?
+## Why stop with just depositing your data and analyses online?
 Let’s face it, you probably put a lot of effort into proposing your thesis hypotheses.
@@
 And it shouldn’t stop you from publishing another study that happened along the way, even though you hadn’t planned it.
-I suspect that in future funders will require us to deposit our successful funding proposals in an online repository before they release funds - hopefully they’ll also be willing to support these facilities financially.
+I suspect that in future funders will require us to deposit our successful funding proposals in an online repository before they release funds - hopefully, they’ll also be willing to support these facilities financially.

### 2021-07-20 4.00a-finishing.Rmd
Add files via upload
# After copy edits

--- 2021-04-17 4.00-part4.Rmd
@@
-# Now that you have finished your data chapters {#part4}
+# Now that you have finished your data chapters
 
 In this last part of the book, I am going to presume that you have done all of the work for the data chapters, and written them up as distinct chapters of your thesis.
-When you look at PhDs that are stored in your institutions libraries, or those that you've referred to in the course of your study, you will notice that they have two additional chapters in addition to data chapters.
-These are often quite short (although lengths vary), and serve to introduce and conclude the PhD thesis (Figure \@ref(fig:thesis-construction)).
-While these are not usually considered to be the 'meat' of the PhD itself, they are important sections to reflect on and carefully consider, as they attempt to pull together data chapters that are written as stand-alone documents into a cohesive thesis.
+When you look at PhDs that are stored in your institution's libraries, or those that you’ve referred to in the course of your study, you will notice that they have two additional chapters in addition to data chapters.
+These are often quite short (although lengths vary), and serve to introduce and conclude the PhD thesis (Figure \@ref(fig:thesis-construction).
+While these are not usually considered to be the ‘meat’ of the PhD itself, they are important sections to reflect on and carefully consider, as they attempt to pull together data chapters that are written as stand-alone documents into a [cohesive thesis](#criteria).
 To this end, they require the same careful planning that any other section in your thesis deserves, with one important condition: that they should not impede the timely submission of your thesis.
+
+A typical construction of the data chapters in your thesis together with a introduction and conclusion.
+The exact construction of your thesis may vary, especially with the number of data chapters that you have.
+The relationship between thesis chapters may also not be linear as presented here.
 
@@
 
-```{r thesis-construction, echo=FALSE, fig.align='left', out.width='90%',  fig.cap="(ref:thesis-construction)"}
+```{r thesis-construction, echo=FALSE, fig.show='hold', fig.align='center', out.width='90%', fig.cap="(ref:thesis-construction)"}
 knitr::include_graphics('figures/thesis-construction.png')
@@
 If funding ceases and your institution allows you to continue your studies, you may have to raise the funding to pay fees as well as to maintain your living costs.
-After submission there will be a period of time while your thesis is in examination, and another period before the final graduation.
+After submission, there will be a period of time while your thesis is in examination, and another period before the final graduation.
 Each of these milestones comes with its own set of deadlines and conditions.
@@
 
-The next chapters deal with writing the [introductory chapter](PhDintro.html), the [concluding chapter](PhDconc.html), and then finally [formatting your thesis for submission](formatting.html).
+The next chapters deal with writing the [introductory chapter](#PhDintro), the [concluding chapter](#PhDconc), and then finally [formatting your thesis](#formatting) for submission.
 These final hurdles require some time and thought.

### 2021-07-20 4.01-PhDintro.Rmd
Add files via upload
# After copy edits

@@
 
-For each of your data chapters, you will have already written an [introduction section](introduction.html), and the same principles that you used to construct this will apply here.
+For each of your data chapters, you will have already written an [introduction section](#introduction), and the same principles that you used to construct this will apply here.
 Instead of aiming to introduce any one of the hypotheses in the chapters, however, your introduction will aim to introduce the ideas and concepts explored inside the data chapters of the thesis.
 
-My suggestion would be for you to start with the conceptual diagram of the thesis that you produced in your [proposal](proposal.html).
+My suggestion would be for you to start with the conceptual diagram of the thesis that you produced in your [proposal](#proposal).
 If you didn't construct one then, you may be reluctant to do it now, but having a good conceptual idea of how the components of your thesis fit together will really aid you in putting together both this introduction and the final conclusion.
-For your [proposal](proposal.html), you may also have written a general introduction, and it would be worth revisiting this now, even if you don't actually use any of it.
+For your [proposal](#proposal), you may also have written a general introduction, and it would be worth revisiting this now, even if you don't actually use any of it.
 
@@
 
-Next produce a hierarchy of hypotheses for all of the chapters and concepts in your thesis (see figure \@ref(fig:JeschkeHagar61)).
-This will help you to make sure that you cover all of the [big ideas](bigidea.html) and concepts in the outline.
-Your outline should follow the standard funnel [formula](formula.html) (see Figure \@ref(fig:introduction)) this time with the highest level ideas in your hierarchy of hypotheses given full prominance in the first paragraph.
+Next, produce a hierarchy of hypotheses for all of the chapters and concepts in your thesis (see Figure \@ref(fig:JeschkeHagar61)).
+This will help you to make sure that you cover all of the [big ideas](#bigidea) and concepts in the outline.
+Your outline should follow the standard funnel [formula](#formula) (see Figure \@ref(fig:introduction)) this time with the highest level ideas in your hierarchy of hypotheses given full prominence in the first paragraph.
 
@@
 
-The way in which to cite your chapters is most straightfoward here by using their chapter number (i.e.
+The way in which to cite your chapters is most straightforward here by using their chapter number (i.e.
 **Chapter n**).
@@
 This doesn't mean that they know what you found in each of the data chapters, but that they know the topics explored within the thesis and what was done to address them.
-Specifically
 
@@
 It is also perfectly legitimate to refer in the introduction to a section of the thesis where a particular topic is dealt with in detail.
-If you are unsure, you'd you best advised by your advisor.
+If you are unsure, you'd best be advised by your advisor.
 

### 2021-07-20 4.02-PhDconc.Rmd
Add files via upload
# After copy edits

@@
 
-By the time you come to write your concluding chapter of your PhD, you should have written all of the [data chapters](part3.html) and the [introductory chapter](PhDintro.html).
+By the time you come to write your concluding chapter of your PhD, you should have written all of the [data chapters](#part3) and the [introductory chapter](#PhDintro).
 Firstly, I should congratulate you for getting this far.
@@
 ## Using the funnel structure
-In principle, you can use the same funnel structure that we described for the [discussion](discussion.html) as a way of structuring your concluding chapter (see figure \@ref(fig:discussion)).
+In principle, you can use the same funnel structure that we described for the [discussion](#discussion) as a way of structuring your concluding chapter (see Figure \@ref(fig:discussion)).
 This would include starting with a synopsis of what your thesis is trying to achieve.
-Clearly, this starting section may be more than a paragraph (as it was in the [data chapters](part3.html)), and the precise structure will be dictated by the complexity of your thesis structure (Figure \@ref(fig:thesis-construction)).
+Clearly, this starting section may be more than a paragraph (as it was in the [data chapters](#part3)), and the precise structure will be dictated by the complexity of your thesis structure (Figure \@ref(fig:thesis-construction)).
 
@@
 
-I suggest that you take your conceptual outline (from the [introductory chapter](PhDintro.html)), and use this to being thinking about how best to construct the outline of your concluding chapter.
-If your data chapters are used to feed into a final output, then consider using the approach of a [logical argument](argument.html) as the outline.
+I suggest that you take your conceptual outline (from the [introductory chapter](#PhDintro)), and use this to being thinking about how best to construct the outline of your concluding chapter.
+If your data chapters are used to feed into a final output, then consider using the approach of a [logical argument](#argument) as the outline.
 
-The main body should review the concluding findings of each of the data chapters, and especially the way in whcih they interact - as these are the aspects that are least likely to have been touched on already in the individual data chapters.
+The main body should review the concluding findings of each of the data chapters, and especially the way in which they interact - as these are the aspects that are least likely to have been touched on already in the individual data chapters.
 It is hoped that the thesis represents a body of work, and not a series of unconnected works.
@@
 Your concluding chapter should provide all of the highlights and insights that have been gained from your PhD thesis.
-It is unlikely to you need to dwell on individuals caveats from the chapters, unless there are general caveats from your approach that span chapters - in which case these probably deserve attention.
+It is unlikely that you need to dwell on individuals caveats from the chapters, unless there are general caveats from your approach that span chapters - in which case, these probably deserve attention.
 
@@
 
-You also have an opportunity to point to where, in your opionion, the direction of your particular field should move next.
+You also have an opportunity to point to where, in your opinion, the direction of your particular field should move next.
 **Do provide your opinion**.

### 2021-07-20 4.03-formatting.Rmd
Add files via upload
# After copy edits

--- 2021-04-15 4.01-formatting.Rmd
@@
 
-If you chose [Bookdown](bookdown.org) or [thesisdown](thesisdown.org) to write your thesis in, then you'll simply be able to download the correct style file (cls) and bibliography formatting file (bib) and press the build book button in [RStudio](www.rstudio.com) and you'll have your formatted thesis in the time that it takes your computer to render the pdf.
+If you chose [Bookdown](bookdown.org) or [Thesisdown](thesisdown.org) to write your thesis in, then you'll simply be able to download the correct style file (cls) and bibliography formatting file (bib) and press the build book button in [RStudio](www.rstudio.com) and you'll have your formatted thesis in the time that it takes your computer to render the pdf.
 You'll still need to look through it and make sure that the output is what you want.
@@
 
-Be patient and don't try to build mega-big files in word-processing documents, they tend to crash.
+Be patient and don't try to build mega-big files in word-processing documents, as they tend to crash.
 Instead, copy the template file and compile one chapter at a time, placing them together only at the end for a final build.
 
-Your university will have their own stipulation for exactly how they want the thesis formatted.
+Your university will have its own stipulation for exactly how they want the thesis formatted.
 Here, I provide a check-list for the last things to do before you submit.
@@
 
-### Must do check-list before submitting your thesis:
+## Must do check-list before submitting your thesis:
 The following points are based on what may irritate or annoy your examiner (and you really don’t want to do that!):
@@
 This will ensure their consistency throughout (within and between chapters).
-- Make sure that your language settings are set to ‘UK English’ (or the English setting for where you are based. 
+- Make sure that your language settings are set to ‘UK English’ (or the English setting for where you are based). 
 - Look out especially for words that have different accepted spellings like those ending in -ise or -ize. Decide which you want and be consistent. Consistency is king!
-- Capitalisation of [common names](taxonomy.html), place names and not adjectives. For example, ‘South Africa’ has two capitals, but ‘southern Africa’ only has one.
+- Capitalisation of [common names](#taxonomy), place names and not adjectives. For example, ‘South Africa’ has two capitals, but ‘southern Africa’ only has one.
 - Grammar check - always good to take a final look, especially for chapters that you wrote some time ago. 
@@
 - Have your computer or another device read the text so that you can hear anything obviously wrong.
-- Pay attention if you have used “we” or “I” and make sure they are consistent in your thesis.
-As a rule I encourage “I” in a thesis, unless the chapter is also a manuscript, in which case “we” is usually correct.
+- Pay attention if you have used 'we' or 'I' and make sure they are consistent in your thesis.
+As a rule, I encourage 'I' in a thesis, unless the chapter is also a manuscript, in which case 'we' is correct.
 - Page layout. Really important to get this right in your template. Make sure that your template has:
@@
 This is very useful for your examiners and may be a requirement for the university.
-Using the [word processor’s](software.html) built in functions will make this task consistent and easy.
+Using the [word processor’s](#software) built in functions will make this task consistent and easy.
 - I dislike writing within a formatted document (as word processors can start getting weird), so my preference is to cut and paste written text into a template at the end. 
-- Remember to give them a check through before handing it in. If you’ve done the sections correctly, then the contents page will come out right.
-- [Title page](title:html) - prescribed very strictly by the university.
-The librarians place the SU watermark after final submission to the library.
+- Remember to give them a check through before handing it in. If you’ve done the sections correctly, then the contents page will come out correctly.
+- [Title page](#title) - prescribed very strictly by the university.
+The librarians place a watermark after final submission to the library.
 It may be tempting to change your title now that you know what's in the thesis, but many universities have strict policies about this.
@@
   + You can check this to make sure that you’ve done all of your sections and subheadings correctly.
-- [Acknowledgements](acknowledgements.html) - this is your time to say thank you to all the special people that have helped during your study.
+- [Acknowledgements](#acknowledgements) - this is your time to say thank you to all the special people that have helped during your study.
 There are probably more than you realise, but in addition to your friends and family (who most people don’t forget), think about the people who administered the work, lab mates (past and present) who were always there to help, and people who gave permission at study sites.
-- [References](references.html) - probably one of the most dreaded sections of any thesis preparation, but they do have to be done.
-If you’re one of these people that has everything in a [database](database.html), then you’ll be laughing or cursing your database throughout.
+- Check that you provide links to any [online repositories](#suppinfo) of your data.
+- [References](#references) - probably one of the most dreaded sections of any thesis preparation, but they do have to be done.
+If you’re one of these people that has everything in a [database](#database), then you’ll be laughing or cursing your database throughout.
 While it might be tempting to only look through the data within the database, spend some time to see how it’s displayed in the thesis.
@@
 
-### Mistakes people make:
+## Mistakes people make:
 Other than the obvious things, all mentioned above, here are some of the mistakes I’ve seen.
 
-- Submitting the wrong version (yes this does happen!). Probably worse if having a mixture of right and wrong versions for different chapters (worse because it takes longer to sort out)
-- Last minute additions to text with incorrect spelling and or grammar
-- Two correctly spelled synonyms sitting next to each other when only one is desired (probably came about when editing)
-- Forgetting to check for plagiarism (see here)
+- Submitting the wrong version (yes, this does happen!). Probably worse if having a mixture of right and wrong versions for different chapters (worse because it takes longer to sort out).
+- Last-minute additions to text with incorrect spelling and or grammar.
+- Two correctly spelled synonyms sitting next to each other when only one is desired (probably came about when editing).
+- Forgetting to check for [plagiarism](#plagiarism).
 - Comments and or edited text (especially when it’s marked as being by someone other than the student).
-- Page numbers that start again and again at different sections
+- Page numbers that start again and again at different sections.
 - Lots of blank pages or spaces (avoid blank pages if you can, and try to limit the amount of blank space (never >half a page).
-- Leaving important people out of the acknowledgements (e.g. advisor, administrators, funders, etc.)
+- Leaving important people out of the acknowledgements (e.g. advisor, administrators, funders, etc.).

### 2021-07-20 4.03-publish.Rmd
Add files via upload
# After copy edits

--- 2021-04-15 4.02-part4.Rmd
@@
-# The obligatoin to publish your work {.unnumbered #publish}
+# The obligation to publish your work {.unnumbered #publish}
 
-Now that you've submitted your PhD, you need still need to do something with it.
+Now that you've submitted your PhD, you still need need to do something with it.
 For most of you this will be publishing it in a journal of your choice.
-It will then take you through what happens to the manuscript and give you insight about where you might want to publish, how to choose the journal.
 Publication is important.
@@
 - By publishing your work you will find that both you, and your work, become known by an international community that may well invite you to participate in the academic process further [@marks2013misuse].
-- Both inside and outside of science as a career, peer reviewed published papers are seen as an important accomplishment in addition to the production of your thesis.
+- Both inside and outside of science as a career, peer-reviewed published papers are seen as an important accomplishment in addition to the production of your thesis.
 - The work that you have already put into your thesis (especially if you have followed the advice in this book!), will mean that you are close to having chapters that can be submitted as publications.
@@
 This is by no means an exhaustive list, but I hope it will give you some insight into the importance of publishing your results.
-To counter this assertion, please be aware that there is also research that suggests that publication driven science is not healthy, and that we need a new way to motivate science [e.g.
+To counter this assertion, please be aware that there is also research that suggests that publication driven science is not healthy, and that we need a new paradigm to motivate science [e.g.
 @stergiou2014impact].
@@
 The challenges of publishing work, like other challenges in academia, are much easier if you possess inside knowledge of the system.
-The demystification of the next steps along the road in academia are the subject of another book:
+The demystification of the next steps along the road in academia is the subject of another book.
+
+```{r cover2, echo=FALSE, fig.align='centre', out.width='60%', fig.link='www.routledge.com/9781032116419'}
+knitr::include_graphics('figures/cover2.png')
+```
+
+**The follow-up to this book, How to Publish in Biological Sciences: A Guide for the Uninitiated**  is available at [www.routledge.com/9781032116419](www.routledge.com/9781032116419).

### 2021-07-20 4.04-lastnote.Rmd
Add files via upload
# After copy edits

@@
 
-If you have found this book useful, then please remember that you can contribute to the project at the Github pages for this book, using bookdown.
+If you have found this book useful, then please remember that you can [contribute to the project](#contribute) at the Github pages for this book, using bookdown.
 If your contributions, no matter how small, improve the book, then you will be improving the experience of doing a PhD in Biological Sciences for future PhD students, and that has to be worth something.

### 2021-07-20 index.Rmd
Add files via upload
# After copy edits

@@
 
-# Under Contract {-}
-This book is currently under contract with CRC Press. 
-```{r fig.align='center', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/Writing_Publishing'}
-knitr::include_graphics('figures/Cover.png', dpi = NA)
-```
-
 # Welcome {-}
@@
 I wrote them because I felt that I was repeating the same advice year after year.
-Instead of rewriting comments on manuscripts time after time, I thought I would consolidate them and put them all in one place.
+Instead of rewriting comments on manuscripts time after time, I decided to consolidate them and put them all in one place.
 Later, I started looking for particular problems that students were having, and then writing blog posts to try to preempt those problems, and to provide solutions before tasks were completed.
@@
 Academics accumulate knowledge throughout their careers, but what can be done for those who are uninitiated?
-What is needed is a guide.
-In the same way that we had guides to foreign countries and cities, postgraduate students need a guide that newcomers can refer to before, during and when making decisions about their time within academia.
+What is needed is a guide that postgraduate students can refer to before, during and while making decisions about their time within academia.
 Note that this is not a rulebook.
 There are times when the guide will be accurate and others when it will be vague, but providing some insight to point you in directions where you can explore more.
+The intention then is to provide you with a starting point from which you can establish your confidence in the academic writing process, and build your own creativity.
 
@@
 [**Part 1**](part1.html)
-Part one is written as something that you should read early on in the process, once you’ve committed to doing a PhD.
-This shortest section of the book deals with what a PhD is, how it’s examined, with whom, where and when to do it.
+-	Part one is written as something that you should read early on in the process, once you’ve committed to doing your PhD.
+This section of the book deals with what a PhD is, how it’s examined, with whom, where and when to do it.
+Perversely, it asks you what you want to do after you’ve finished, pointing out that you will have opportunities to open doors during the process.
 At the start of your PhD is the time to be thinking of the bigger questions both in your study area, and your life, and this section provides some pointers and things to consider.
+Working within academia is particularly stressful environment, and students should be aware of the state of their mental health, how to watch become aware if it is deteriorating, and what you can do to help yourself.
+I also encourage you to participate in the **scientific project**, and become aware of your role as a scientist within society.
 
 [**Part 2**](part2.html)
-Part two of the book sets out a practical guide to getting started with writing.
+-	Part two of the book sets out a practical guide to getting started with writing.
 The approach I've taken here is describing the production of a PhD in biological sciences which is made up of data chapters, and where each chapter is essentially a potential publication in a science journal.
 If you didn't want to do a PhD but just wanted to write a publication, then most of what is here (and in [Part 3](part3.html)) will be of use as you could take the same formulaic approach as given in this book.
-Similarly, if you were writing a MSc project, this section would also be useful.
-There is a lot of technical stuff involved in writing science, and in part 2 I cover as much as I can while still remaining general to the subject area.
+Similarly, if you were undertaking a MSc by research, this section would also be useful.
+The approach taken is ‘hypothesis-centric’, and how the entire chapter will be built around answering the hypothesis posed.
+This section explains how the hypothesis and all logical arguments that you make in the text are constructed from existing literature - **standing on the shoulders of giants** - with a practical guide on how to use citations.
+In addition to some nuts and bolts of how to write paragraphs, construct arguments, remain concise and avoid plagiarising, this section attempts to cover a lot of technical aspects of writing for the biological sciences.
 
 [**Part 3**](part3.html)
-A typical PhD will consist of five data chapters where each data chapter is a publication.
-There is a shortish introductory chapter (sometimes a literature review) that comes before these data chapters, and a concluding synthesis chapter at the end.
-That's not to say that this is the only way of writing a PhD.
-There are many different ways and it is a good idea for you and your advisor to talk about this together before you start.
-
-This part covers how to write the core sections of a paper: abstract, introduction, methods and materials, results, discussion and references.
+-	A typical PhD will consist of five data chapters where each data chapter is a publication.
+The standard formula is to write this is in sections consisting of an introduction, methods and materials, results and a discussion.
+Together with the title, abstract, references and supplementary information, all of these parts also make up most scientific publications.
+In this section of the book, I go through each of these components using a formulaic approach.
+To get you started writing, I suggest how you can build each section from an outline that is fleshed out with literature and works around a hypothesis (from [Part 2](part2.html)).
+In addition to the sections needed for your thesis, I also cover parts needed to submit a paper, including the title page, writing the abstract, supplementary material, authorship, and acknowledgements.
+Therefore, this part of the book will be relevant for anyone setting out to write a scientific paper in the biological sciences, especially if you are having trouble getting started.
 
 [**Part 4**](part4.html)
-The concluding part of this book is written to cover the things that remain after you have finished writing all of your data chapters.
-Most importantly, these include the introductory and concluding chapters that sit as bookends to the data chapters that you have already produced.
+-	The last and shortest section of this book talks about what you still need to do once your data chapters are written.
+For most PhD theses, you will need to write a short introductory chapter (sometimes a literature review) that comes before these data chapters, and a concluding synthesis chapter at the end.
 These chapters make your thesis into a 'body of work', and allow you space to reflect on what you have achieved in your PhD studies.
+I have framed their content at the bare minimum required, as many students will find themselves under a great deal of pressure to submit their thesis to a prescribed deadline.
+I also provide some alternative approaches for those who have the luxury of more time to attack these sections in more detail.
+The very end of the book sets out some reasons for why you need to publish your thesis chapters.
 
@@
 
-## Acknowledgments {-}
+## Acknowledgments {- #acknowledge}
 
@@
 This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help.
-I thank this greater academic community for sharing and inspiring.
+I thank this greater academic community (especially #academicTwitter) for sharing and inspiring.
 Thanks go to the many reviewers and editors who have taken their time to improve my writing.
@@
 Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning.
-James Baxter-Gilbert, Anthony Herrel,  Allan Ellis, James Vonesh all read parts and commented.
+James Baxter-Gilbert, Jack Dougherty, Anthony Herrel, Allan Ellis, Andrea Melotto, Lisa and Mark O'Connell, James Vonesh, Carla Wagener all read or commented on different aspects of the book.
 Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and Stellenbosch University.

### 2021-08-01 00-author.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 ## Creative Commons License {-}
-How to write a PhD in Biological Sciences: a guide for the uninitiated is copyright to John Measey and distributed under a Creative Commons BY-NC-ND 4.0 International License.
+How to write a PhD in Biological Sciences: a guide for the uninitiated is copyright to John Measey and distributed under a [Creative Commons BY-NC-ND 4.0 International License](https://creativecommons.org/licenses/by-nc-nd/4.0/).
 You may freely share this content for non-commercial purposes, with a source credit to [www.HowToWriteAPhD.org](www.HowToWriteAPhD.org).
@@
 The author and publisher have neither liability nor responsibility to any person or entity related to any loss or damages arising from the information contained in this book.
-
-```{r iris}
-knitr::kable(
-  head(iris), caption = 'We must have the iris data and some R code, otherwise it would not seem right',
-  booktabs = TRUE
-)
-```

### 2021-08-01 1.01-beforeyoustart.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make informed decisions from the outset.
-For example, reconciling yourself to using the [statistical package R](useR.html) will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on.
+For example, reconciling yourself to using the [statistical package R](#useR) will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on.
 Preparing yourself into the right mental framework will also be very important for your ability to work under pressure.
@@
 
-I prefer a thesis that fits together within an explicit [conceptual framework](PhDintro.html).
-I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the introduction of their [proposal](proposal.html).
+I prefer a thesis that fits together within an explicit [conceptual framework](#PhDintro).
+I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the introduction of their [proposal](#proposal).
 A useful concept to be aware of is the 'Hierarchy of Hypotheses' [@heger2018hierarchy].
-In this approach, you are encouraged to consider the '[bigger question](bigidea.html)' in your research area, identifying both what studies have produced compelling evidence, and identifying knowledge gaps from a theory driven approach.
+In this approach, you are encouraged to consider the '[bigger question](#bigidea)' in your research area, identifying both what studies have produced compelling evidence, and identifying knowledge gaps from a theory driven approach.
 Generally, it is only possible to consider these types of concepts once you are familiar with the literature (both theoretical and experimental).

### 2021-08-01 2.04-BigIdea.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 You can think of the way that such ideas are structured as a hierarchy of hypotheses (Figure \@ref(fig:JeschkeHagar61)).
-Building such a hierarchy of ideas in your own specialist subject area is good reference material to prompt [critical reading](critical.html) and reflection.
-You may want to take this structure forward when thinking how to [introduce your PhD chapters](PhDreview.html).
+Building such a hierarchy of ideas in your own specialist subject area is good reference material to prompt [critical reading](#critical) and reflection.
+You may want to take this structure forward when thinking how to [introduce your PhD chapters](#PhDreview).
 

### 2021-08-01 2.06-argument.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 Another sentence structure to consider is explaining cause and effect.
-For practical wording examples of both of these classic arguments, [see part 2](phrasebank.html).
+For practical wording examples of both of these classic arguments, [see part 2](#phrasebank).
 
@@
 This is also known as a Type II error (false negative), where the lack of finding a significant relationship does not lead to a conclusion that the relationship does not exist.
-Later we discuss how a [lack of finding of a significant result](transparency.html) leads to [confirmation bias](#TypeI) in science.
+Later we discuss how a [lack of finding of a significant result](#transparency) leads to [confirmation bias](#TypeI) in science.
 
@@
 Better go back and make sure that you understand the original premise and the works that promote or oppose this.
-Having told you to beware, I’m now going to encourage you to read, because [reading is one of the best ways in which you can learn about writing a logical argument](criticalreading.html).
+Having told you to beware, I’m now going to encourage you to read, because [reading is one of the best ways in which you can learn about writing a logical argument](#critical).
 
@@
 However, I encourage you to learn and experiment with writing styles as you become more experienced.
-[Critical reading](criticalreading.html) is one of the best ways to learn about alternative writing styles: reading is probably the best way of improving your writing.
+[Critical reading](#critical) is one of the best ways to learn about alternative writing styles: reading is probably the best way of improving your writing.
 Alternatively, and especially if the above hasn’t clicked with you, you can read more about writing a scientific argument (see for example [here](https://msu.edu/course/iah/231b/gifford/sciaargs.htm)).

### 2021-08-01 2.07-storytelling.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 But this storyis just one interpretation of the correlation that was found.
-[Correlation is not causation](argument.html), and more studies would be needed to test the hypothesis concerning the underlying reasons suggested for the relationship observed.
+[Correlation is not causation](#argument), and more studies would be needed to test the hypothesis concerning the underlying reasons suggested for the relationship observed.
 The key differences in the storytelling style from the scientific formula are the absence of a known goal at the start of the process (see a more detailed discussion of this by Yarden Katz, [-@katz2013storytelling] here).
@@
 This is particularly true when communicating science to the public and making it more accessible, but it also applies to interactions between scientists, for example, at conferences.
-The [presentations](#proposal-presentation.html) that tell a story and entertain are those we tend to remember.
+The [presentations](#proposal-presentation) that tell a story and entertain are those we tend to remember.
 Not easy, but if we do want to communicate well with each other, then we need to learn the art of storytelling, without compromising our scientific objectivity.
 In addition, there are lots of lessons from storytelling that we can use within scientific papers, and to which I subscribe wholeheartedly.
-For example, don’t be boring, don’t [bamboozle](style.html), write an active narrative (not in the passive tense), use the magic of threes, and many more.
+For example, don’t be boring, don’t [bamboozle](#style), write an active narrative (not in the passive tense), use the magic of threes, and many more.
 

### 2021-08-01 2.08-citations.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 
-Within a manuscript for a scientific journal or in your thesis, you can expect that your [introduction](introduction.html) and [discussion](discussion.html) sections are going to be full of citations.
+Within a manuscript for a scientific journal or in your thesis, you can expect that your [introduction](#introduction) and [discussion](#discussion) sections are going to be full of citations.
 The methodology will likely also be citation rich.
@@
 
-Later in the book ([Writing with a Formula](formula.html)), I explain how to start writing with an outline, then flesh out the outline with citations from your reading, and plan which citations you are going to use carefully before you start.
+Later in the book ([Writing with a Formula](#formula)), I explain how to start writing with an outline, then flesh out the outline with citations from your reading, and plan which citations you are going to use carefully before you start.
 This won't stop you from adding more later, but this ensures that your work is built on your scholarly endeavour.
@@
 Then you should cite the most recent work that shows the ideas that you are wanting to show.
-Other times, primacy is more important - especially when few have built on ideas or concepts since they were put forward, or in the [taxonomic literature](taxonomy.html).
+Other times, primacy is more important - especially when few have built on ideas or concepts since they were put forward, or in the [taxonomic literature](#taxonomy).
 If in doubt, place a citation to the original study and the most recent study, then flag this using a comment for your advisor to consider.
@@
 Some have argued that the rule of priority is of benefit to science as it drives competition and this spurs creativity.
-On the other hand, the winner-takes-all approach of being first with something drives secrecy and dishonesty, and does not genuinely reflect the team effort that is the [science project](lifescientific.html) [@casadevall2012reforming].
+On the other hand, the winner-takes-all approach of being first with something drives secrecy and dishonesty, and does not genuinely reflect the team effort that is the [science project](#lifescientific) [@casadevall2012reforming].
 
@@
 
-Obviously, everything you cite needs to be in the References (or [Literature Cited](references.html)) section, and you may well need to spend time deleting extra stuff.
-You can get around this common issue by using a citation manager like [Mendeley, Zotero or EndNote](references.html).
+Obviously, everything you cite needs to be in the References (or [Literature Cited](#references)) section, and you may well need to spend time deleting extra stuff.
+You can get around this common issue by using a citation manager like [Mendeley, Zotero or EndNote](#references).
 I’d always suggest that you use one of these tools as they can really help with your reading too.

### 2021-08-01 2.09-databases.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 This is called a systematic literature review.
-Using this tactic from the outset of your interactions with literature databases will give you a good idea of what proportion of the literature you are reading, and you might then be able to convert your finding into an objective literature review - making a great first [introductory chapter](PhDintro.html) for your thesis.
+Using this tactic from the outset of your interactions with literature databases will give you a good idea of what proportion of the literature you are reading, and you might then be able to convert your finding into an objective literature review - making a great first [introductory chapter](#PhDintro) for your thesis.
 
@@
 Any keywords that they share are going to be important for you.
-Make a note of all relevant keywords for your searches, as ultimately, this will help you when you need to determine the keywords for [your own chapters](title.html).
+Make a note of all relevant keywords for your searches, as ultimately, this will help you when you need to determine the keywords for [your own chapters](#title).
 
@@
 
-An invaluable way in which databases have become extremely useful is the ability to move items that you have found during your search directly into your [reference manager](referencesoftware.html).
+An invaluable way in which databases have become extremely useful is the ability to move items that you have found during your search directly into your [reference manager](#referencesoftware).
 Beware that these regularly come with errors, and that the amount of errors is related to your database source (Scopus and WoS are much better than Google Scholar).
 This means that after importing the references, you will need to go back to your reference manager and edit them or they will come out incorrectly when you use them.
-Of particular note to biologists is the need to have all species names in [italics](taxonomy.html), and you'll need to do this yourself.
+Of particular note to biologists is the need to have all species names in [italics](#taxonomy), and you'll need to do this yourself.
 
@@
 
-The [ORCID](orcid.html) number has been erected to provide a common platform for [authors to curate](title.html), and now some journals and funding authorities won't allow you to submit or apply without one.
+The [ORCID](#orcid) number has been erected to provide a common platform for [authors to curate](#title), and now some journals and funding authorities won't allow you to submit or apply without one.
 

### 2021-08-01 2.10-referencesortware.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 
-Because published papers play such a fundamental role in your [reading and comprehension](critical.html) of a subject, and [citations](citations.html) are pivotal in chapter or paper writing, putting these elements together in a good reference manager makes a lot of sense.
+Because published papers play such a fundamental role in your [reading and comprehension](#critical) of a subject, and [citations](#citations) are pivotal in chapter or paper writing, putting these elements together in a good reference manager makes a lot of sense.
 A reference manager is a piece of software in which you can store and view all of your papers (in pdf format or via links), make notes on them, and use them to make citations and build bibliographies.
@@
 
-My advice here is (like software for [writing your PhD](software.html)) to pick the format that your references are going to be stored in carefully.
+My advice here is (like software for [writing your PhD](#software)) to pick the format that your references are going to be stored in carefully.
 For example, the BibTex format (see below) is a generic format that many of the software packages are capable of importing and exporting.
@@
 
- - Does it integrate with your [word processing package](software.html)?
+ - Does it integrate with your [word processing package](#software)?
  - Does your university have a licence?

### 2021-08-01 2.10a-sortware.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 
-Another important consideration is whether or not your chosen word processing package will allow seamless integration with your [reference manager](referencesoftware.html).
+Another important consideration is whether or not your chosen word processing package will allow seamless integration with your [reference manager](#referencesoftware).
 Your lab may have a communal repository of references, and so this may then constrain what you are comfortable using.
@@
 when part of the file malfunctions and you lose the contents.
-That’s right, in addition to loss of data by back up deficiencies [see Part III](data.html), you also have the risk of having corrupt files that will also mean you lose all of the contents.
+That’s right, in addition to loss of data by back up deficiencies [see Part III](#data), you also have the risk of having corrupt files that will also mean you lose all of the contents.
 Even if you always use Word and never use any other file format, do yourself a favour and save each version as a .rtf or .txt file as well.
@@
 Once you start putting chapters together especially if they include embedded images tables and the like you will find that many word processors start misbehaving.
-This can be especially frustrating in the [closing phases of your PhD](part4.html) when you're trying to submit it.
+This can be especially frustrating in the [closing phases of your PhD](#part4) when you're trying to submit it.
 Especially if there's a tight deadline.
@@
 Nevertheless, text files take up very little space and it’s not a problem to keep more versions if you want to.
-More on this in the chapter on [data management](data.html).
+More on this in the chapter on [data management](#data).

### 2021-08-01 2.15-concise.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 
-In this book, I have put an emphasis on a [writing formula](formula.html) in order to get started writing.
+In this book, I have put an emphasis on a [writing formula](#formula) in order to get started writing.
 Many students have a problem in knowing where and how to get started with what seems to be a daunting prospect: writing a proposal or writing the entire thesis.
@@
 Maybe write a letter (or long email) instead of making a call.
-It'll also help you to [read critically](critical.html).
+It'll also help you to [read critically](#critical).
 If you are a keen reader, then you could do worse than reading some Kurt Vonnegut [-@vonnegut1980how].
@@
 
-It's important to avoid distracting your reader (see [Part II](part2.html)).
+It's important to avoid distracting your reader (see [Part II](#part2)).
 Your aim is to be thorough, which will mean including all relevant information, but don't allow your writing to sideline them into taking the wrong direction.
@@
 The highway analogy is useful as although on a highway you will see signs to other places that you can point out to the reader, you shouldn't turn off the highway to get to your destination.
-The best of highways will also be free of traffic jams (think long, complex sentences here), and toll free ([no paywall](paywall.html)).
+The best of highways will also be free of traffic jams (think long, complex sentences here), and toll free ([no paywall](#paywall)).
 
@@
 
-Also, please don't cut and paste sections of [methods](methods.html) from one part of your thesis to another. Just don't do it.
+Also, please don't cut and paste sections of [methods](#methods) from one part of your thesis to another. Just don't do it.
 
-People also have a tendency to build the [abstract](abstract.html) by copying and pasting text from the main sections.
+People also have a tendency to build the [abstract](#abstract) by copying and pasting text from the main sections.
 Don't do it.
 The reader will quickly see the repetitions and become bored.
-Similarly, the [conclusion/summary](discussion.html) section is also often copied from the lines above.
+Similarly, the [conclusion/summary](#discussion) section is also often copied from the lines above.
 If it's not worth writing again, then it wasn't worth writing before.
@@
 Perhaps it should have been inserted higher up the list.
-It would help you to have your work read through by one of your colleagues before you [give it to your advisor](goodenough.html).
+It would help you to have your work read through by one of your colleagues before you [give it to your advisor](#goodenough).
 But above all, you must be prepared to read __all of your work__ yourself, and be prepared to work on it when you read.
@@
 
-Being concise is a great way to write well, and the [abstract](abstract.html) is a case in point. Please don't forget to read Scott's 10 rules as written by Scott [@hotaling2020simple].
+Being concise is a great way to write well, and the [abstract](#abstract) is a case in point. Please don't forget to read Scott's 10 rules as written by Scott [@hotaling2020simple].

### 2021-08-01 2.20-proposal.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 
-Your final thesis will be a set of papers (data chapters sandwiched between an [introduction and discussion](PhDintro.html)), and so prepare the proposal in the same way.
+Your final thesis will be a set of papers (data chapters sandwiched between an [introduction and discussion](#PhDintro)), and so prepare the proposal in the same way.
 That is, you will normally have two (MSc) or five (PhD) data chapters.
@@
 
-I’m not going to go into what’s inside the introduction here, as it’s already been done in other chapters on formulaic writing (see later in this [part](formula.html)) and specifically about the introduction (see [part 3](introduction.html)).
+I’m not going to go into what’s inside the introduction here, as it’s already been done in other chapters on formulaic writing (see later in this [part](#formula)) and specifically about the introduction (see [part 3](#introduction)).
 Although you may feel that there's no point putting a lot of effort into this introduction it serves as a literature review for your upcoming PhD study, and is well worth the time and effort.
@@
 
-What’s in the [Methods and Materials](methods.html) section is also covered here. Essentially you need to:
+What’s in the [Methods and Materials](#methods) section is also covered here. Essentially you need to:
 
@@
 - Describe exactly how you’ll set up your experiment and/or collect your samples
-- You must explicitly state how each of the variables (introduced in the [introduction](introduction.html)) is collected
+- You must explicitly state how each of the variables (introduced in the [introduction](#introduction)) is collected
 - (If necessary) Describe how you’ll turn your collected data into the variables needed for your analyses
-- Data analyses (how you’ll test your [hypotheses](hypothesis.html)).
+- Data analyses (how you’ll test your [hypotheses](#hypothesis)).
 This should include a description of the statistical or analytical techniques.
@@
 
-Writing grant proposals is essentially the same as your thesis proposal, and so you can use this document (with some modifications to tailor it to the grant objectives) to [raise money](funding.html) for your work.
+Writing grant proposals is essentially the same as your thesis proposal, and so you can use this document (with some modifications to tailor it to the grant objectives) to [raise money](#funding) for your work.
 
@@
 
-This is the subject of another chapter (see [part 3](transparency.html)).
+This is the subject of another chapter (see [part 3](#transparency)).

### 2021-08-01 2.21-proposal-presentation.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 
-Despite previous sections [of this book](storytelling.html) cautioning you against storytelling, you really need to have turn your study into a story in order to present it well.
+Despite previous sections [of this book](#storytelling) cautioning you against storytelling, you really need to have turn your study into a story in order to present it well.
 This means that you know what the message you want to reveal to the audience is.

### 2021-08-01 2.22-transparency.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 It is probably too early to tell [@rubin2020does], but it is certainly a good place to start [@nosek2018preregistration].
-The more researchers know and subscribe to [transparency](transparency2.html) in their research, the more it will shift the culture in science for the better [@forstmeier2017detecting].
+The more researchers know and subscribe to transparency in their research, the more it will shift the culture in science for the better [@forstmeier2017detecting].
 However, there are more biases existing in science, and the system needs buy-in from stakeholders to promote equity by elimination of bias.

### 2021-08-01 3.10-Figures.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 
-I encourage my students to draw conceptual diagrams of how the thesis chapters relate to one another (Figure \@ref(fig:thesis-construction)), and then to include this diagram in the introduction to the thesis to help examiners and any other readers to understand how the chapters are interrelated (see [Part IV](PhDintro.html)).
+I encourage my students to draw conceptual diagrams of how the thesis chapters relate to one another (Figure \@ref(fig:thesis-construction)), and then to include this diagram in the introduction to the thesis to help examiners and any other readers to understand how the chapters are interrelated (see [Part IV](#PhDintro)).
 

### 2021-08-01 index.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

@@
 site: bookdown::bookdown_site
-description: You don’t have to be a genius to write a PhD. It will always involve
-  a lot of hard work and dedication, but the process of writing up is a whole lot
-  easier if you understand the basic ground rules. This book is a guide specifically
-  for writing a PhD in the biological sciences. It will guide you from deciding
-  whether or not you should actually do a PhD, provide practical, up-to-the-minute
-  guides to getting started with planning and writing, and consolidate this with the
-  nuts and bolts of writing for the biological sciences. The book guides you through
-  the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication
-  in peer-reviewed journals. Whether you are a student just starting your studies,
-  or a supervisor struggling to cope, this book will provide the insider information
-  you need to get ahead.
+description: "You don’t have to be a genius to write a PhD.
+It will always involve a lot of hard work and dedication, but the process of writing up is a whole lot easier if you understand the basic ground rules.
+This book is a guide specifically for writing a PhD in the biological sciences.
+It will guide you from deciding  whether or not you should actually do a PhD, provide practical, up-to-the-minute guides to getting started with planning and writing, and consolidate this with the nuts and bolts of writing for the biological sciences.
+The book guides you through the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication in peer-reviewed journals.
+Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
 github-repo: yihui/bookdown-crc
@@
 ```
+# In Press {-}
+This book is _in press_ with CRC Press.
+
+```{r fig.align='centre', out.width = '80%', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/'}
+knitr::include_graphics('figures/cover.jpg', dpi = NA)
+```
+
+You can pre-order your own copy on the [CRC Press website](https://www.routledge.com/How-to-Write-a-PhD-in-Biological-Sciences-A-Guide-for-the-Uninitiated/Measey/p/book/9781032080208), on [Amazon](https://www.amazon.com/How-Write-PhD-Biological-Sciences/dp/1032080205), at [Barns & Noble](https://www.barnesandnoble.com/w/how-to-write-a-phd-in-biological-sciences-john-measey/1139436164), and at other good book sellers.
 
@@
 
-[**Part 1**](part1.html)
+[**Part 1**](#part1)
 -	Part one is written as something that you should read early on in the process, once you’ve committed to doing your PhD.
@@
 
-[**Part 2**](part2.html)
+[**Part 2**](#part2)
 -	Part two of the book sets out a practical guide to getting started with writing.
 The approach I've taken here is describing the production of a PhD in biological sciences which is made up of data chapters, and where each chapter is essentially a potential publication in a science journal.
-If you didn't want to do a PhD but just wanted to write a publication, then most of what is here (and in [Part 3](part3.html)) will be of use as you could take the same formulaic approach as given in this book.
+If you didn't want to do a PhD but just wanted to write a publication, then most of what is here (and in [Part 3](#part3)) will be of use as you could take the same formulaic approach as given in this book.
 Similarly, if you were undertaking a MSc by research, this section would also be useful.
@@
 
-[**Part 3**](part3.html)
+[**Part 3**](#part3)
 -	A typical PhD will consist of five data chapters where each data chapter is a publication.
@@
 In this section of the book, I go through each of these components using a formulaic approach.
-To get you started writing, I suggest how you can build each section from an outline that is fleshed out with literature and works around a hypothesis (from [Part 2](part2.html)).
+To get you started writing, I suggest how you can build each section from an outline that is fleshed out with literature and works around a hypothesis (from [Part 2](#part2)).
 In addition to the sections needed for your thesis, I also cover parts needed to submit a paper, including the title page, writing the abstract, supplementary material, authorship, and acknowledgements.
@@
 
-[**Part 4**](part4.html)
+[**Part 4**](#part4)
 -	The last and shortest section of this book talks about what you still need to do once your data chapters are written.
@@
 It is because you wanted more that I put this together.
-I have also been a student, and have been inspired by colleagues around the world who have been exemplary advisors.
+I have also been a student, and have been inspired by colleagues around the world who are exemplary advisors.
 This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help.
@@
 Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning.
-James Baxter-Gilbert, Jack Dougherty, Anthony Herrel, Allan Ellis, Andrea Melotto, Lisa and Mark O'Connell, James Vonesh, Carla Wagener all read or commented on different aspects of the book.
-Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and Stellenbosch University.
-A special thanks to the librarians who have supported many of my more extreme rantings about publishers.
+James Baxter-Gilbert, Jack Dougherty, Anthony Herrel, Allan Ellis, Andrea Melotto, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, James Vonesh, Carla Wagener all read or commented on different aspects of the book.
+Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and in the library at Stellenbosch University.
 

### 2021-09-05 2.09-rebuttal.Rmd
Add files via upload
# all round update

# Receiving the editor's decision {#decision}

[Peer review](#peerreview) is the basis of guarding and maintaining quality in science. If you’ve just got a decision from a journal, you’ll need to respond to the comments. This to and fro between authors and reviewers usually doesn’t exceed two rounds. As you approach the comments, there are several points that it’s worth bearing in mind:

- The task of the reviewer is to help improve your manuscript for the scientific record and for the reader
- The reviewers and editors have given their time freely to help you and your work
- It's easy to misinterpret what reviewers and editors say, and you need to take the time and space to respond appropriately

If you already know what it’s like to receive reviewer comments, but struggle to understand why reviewers say what they do, then it’s time for you to start [reviewing papers](#howpeerreview). You can ask your mentor to mention you as a potential reviewer (especially if they are going to decline because they are too busy). But probably the best way to get started is to participate with your lab or journal club in the review of a preprint ([see part 2](#preprint)). This has the advantage that you are not the sole reviewer, but will be in a group. Your group will get credit for their review (via a [DOI](#DOI)). You will get to discuss the finer point of reviewing with your team. Especially, you will likely hear remarks that might come across as insulting or unprofessional and it’ll give you a chance to challenge these at source, and so positively influence others.

## What to expect in your decision letter
When you finally receive it, the email will contain some stock text about the overall decision, some information about how to resubmit your revised version (if you have this as an option), and a time-frame. After the editor has signed off, you will find comments first from the Associate Editor (AE) who handled the review, and then from (typically) two reviewers, but sometimes three (or even more). It would be a good idea to add the deadline to your calender, especially if you feel that you won't be able to respond immediately. You can always ask for an extention if you are likely to go over the allotted time.

### The desk rejection {#desk}
A desk rejection is when your manuscript is assessed by a subject editor (it could be the editor-in-chief, an associate editor or any rank in between), and they decide (based on your title, abstract, cover letter, or a quick look at the content), that the manuscript is: not in the right subject area, insufficiently novel (too many similar papers), or unlikely to make it through peer review (not scientifically robust). The editor in question should explain how your manuscript failed to meet the specified editorial requirements for undergoing peer review in their journal [@teixeiradasilva2018establishing]. In my experience, this is rarely the case. Instead, it appears as if these editors are making a quick [confirmation bias](#transparency) decision based on whether or not they consider the submission will receive sufficient interest in order to match their [Impact Factor](#impactfactor) (or even inflate it). Moreover, it has been suggested that these decisions by editors are on the whole poor, with 66% of submissions being accepted in an equivalent journal without changes [@farji-brener2014rejecting]. The editors responded that their desk rejections are far more nuanced, and that equivalent journals don't necessarily have the same scope [@schimel2014editors]. You should have already [checked the scope of the journal](#whichjournal) prior to submission. The reality is that many journals have far too many submissions than they can handle, and so there is a need for an initial triage. In general, desk rejections work well for all concerned when authors know why they have received this decision. For this, there need to be specific guidelines that are followed consistently by the whole editorial team [@teixeiradasilva2018establishing].

On the positive side, a desk rejection is fast (otherwise, it's unfair) and there should be plenty more journals on your list. It is possible to appeal, but you will already have had an editorial decision against your manuscript, hence you will be fighting an uphill battle from the start. In terms of your time and effort, you'd probably be better off accepting this rejection and moving on.

### Take a rejection seriously {#rejection}

Rejections are harsh, but totally normal [@cassey2004publication; @day2011silent]. Everyone gets rejections, including the top scientists in their fields [@cassey2003publication]. As a result, you will feel disheartened, frustrated and even angry [@pannell2002prose], and this is normal too. Don’t take it personally, and you'll find that the more you share the times you fail, the more you'll find that others share the experience [@crew2019here]. You will need to pick yourself up, and take your manuscript back and try again. The results of a survey among top ecologists suggests that scientists need to develop a thick skin when it comes to rejection [@cassey2003publication]. Learn as many lessons as you can from your rejection and quickly move on. Certainly, never dwell on a rejection and feel that this is anything more than a minor setback. Receiving a rejection is not an indication of professional inadequacy, also known as [imposter syndrome](#mentalhealth) [@woolston2016faking]. All of us have been rejected, and continue to deal with rejections throughout our careers.

The most important point to discern as quickly as possible is whether or not your manuscript is not fundamentally flawed. If you receive a rejection because your study is flawed then take the time to learn the lesson. This will be a hard lesson to accept because it speaks about your experimental design, and whether or not you can really answer the questions that you set out to when forming your hypothesis.

Do not start Hypothesising After Results are Known (HARKing). If you have made mistakes in your experimental design, then your data might be useful for a post-hoc manuscript. There is no shame in producing a descriptive note of this kind. Many researchers appear to automatically send a flawed manuscript to another journal to see how it will do there. These manuscripts are very plentiful and take up the time of a great many people (in specialist areas you see them doing the rounds of different journals). I would urge you to learn from mistakes, and not to burden others with them. Many editors try to give good advice when making rejections, and you should take the time needed to absorb this. If you believe their assessment to be wrong, and it may be, then seek out some honest sounding boards for your manuscript, such as a mentor or another colleague or post-doc.

Sometimes, you will receive a Reject decision, and the editors comments will allude to the work not being sufficiently novel, or not sufficiently advancing the field. In other words, whatever it was that they were looking for, your manuscript didn't have it. Tacked onto the end of their email may be the offer to submit to another journal in the same stable, the so-called 'portable peer review'. Typically, these are Open Access [No Impact Factor](#noif) journals where you have to pay an, often hefty, article processing charge to get published. This is rarely a good idea. The standings of these journals is (typically) not great, and you’d probably be better off going back to your [original submission list](#whichjournal) and selecting the next journal there. Having said this, if the journal offers to move your submission with or without the peer reviews, at least you know that your manuscript is not fundamentally flawed. This kind of rejection is simply an example of [publication bias](#badscience), and it is time for you to move on from this journal to another on your [submission list](#whichjournal).

### Time taken to receive your decision {#timetaken}
Usually, you will receive a decision letter (email) after ~60 days. This depends on the journal, editors and reviewers, but it’s worth signing into the editorial management system and looking at the status of your manuscript. If after 60 days the status hasn't changed and you’ve heard nothing, then contact the editorial team (with the email address given in the editorial management system. Time to first decision in the Life Sciences was found to average 11 weeks and 25% of decisions are received within a month [@huisman2017duration]. That might sound oh so slow, but this is relatively quick when compared to the 18 week average first response in economics.

It is worth bearing in mind that authors generally feel more unhappy with the reviewer comments they receive the longer they have to wait [@jiang2021understanding], while fast turn around times produce more content authors even in the face of rejections [@huisman2017duration]. Clearly, the difference is in the resentment of time wasted, especially for a desk rejection, or even for poor or feeble reviewer input with a rejection. If you feel particularly unhappy about long waiting times, then you could consider releasing your manuscript as a [preprint](#preprints2). Otherwise, there's little that you can do, and it can simply be bad luck when you have to wait a long time for a review. Different journals have different mean times to first decision, and some (often those with faster times) advertise this and so this may influence your decision about [where to take your manuscript next](#whichjournal).

In my experience, every manuscript is different and it's very hard as an editor to determine how long it will take to find reviewers, or how long those reviewers will take to produce their reviews. Just because reviewers accept to undertake reviews, does not mean that they will do them timeously, or even at all. After a failure to review on time, a good editor will follow up to seek out other potential reviewers. But there are other reasons why decisions may take time.

In the case that the editor found no reviewers after looking for some weeks, one option is for them to make a '[desk reject](#desk)' and encourage resubmission. In the rejection, they may point out a number of potentially minor faults. For the editor, it takes the heat off of their desk. They may not be allocated the new manuscript after resubmission. For the authors, it's just frustrating. If you suspect this has happened then consider another journal (with a clear editorial stance on peer review), or provide more [potential reviewers](#suggestreviewers).

It is worth remembering that certain times of the year are likely to take longer than others for peer review. Editors find it more difficult to find reviewers in the summer months when many biologists are in the field or on holiday (remember that summer occurs at different times in northern and southern hemispheres). The start of the academic year is also a particularly busy time for most academics (editors and reviewers), and you might expect waiting longer (although academic years start at different times in different countries). In general, academics are busy and finding reviewers is difficult [@perry2012peer].

The delay from the end of data collection to being published can take between 0.5 and 3.5 years, depending on the subdiscipline of biological sciences that you are working in [@christie2021reducing]. This includes the initial write-up and submission time, together with any initial rejections from other journals and multiple rounds of reviews. This adds to many other arguments for using [preprint servers](#preprints) to make your work publicly available as soon as it is completed [@christie2021reducing].

## What to do when you receive your reviewer comments
Your reviewer comments will arrive in an email when you are busy doing something else. If you have time to read them the same day, then my suggestion is that you read without trying anything further. Remember to forward them to your co-authors as soon as possible. Simply read the comments and then close the email and mark it for further attention the next day. Your writing is very personal to you, and you might be surprised at just how hurtful it can feel to have someone critique your writing (and your experiment) without holding back. If you’ve not experienced this before, then prepare yourself. No matter how much effort you put into your text, sending it out for peer review is a really high bar. Make an appointment with your co-authors so that you can discuss the comments together. Whether good, bad, or bizarre, it is best to set aside some time to read through the comments carefully, so that you can respond.

## Responding to reviewers’ comments with a rebuttal {#rebuttal}
The reviewers are usually given two boxes to write comments in. One pertains to the comments that you receive in your letter, and the second are confidential comments to the Associate Editor (AE). Remember that the AE is acting on both of these sets of comments, and so the decision may reflect something that is said in confidence. This is not really in the spirit of [transparency](#transparent) for peer review. The best peer review systems are open and online.

Once you’ve found the time in your week, sooner is better than later, sit and read the comments again. Normally, they will sound much better, and less harsh, on the second read (if not, try third or fourth). They should seem far more approachable than when you first read them. Most reviews will have a set of major (when applicable) and minor comments that you need to address from each of them. Try sketching a few responses down to the major revision comments before your meeting with your co-authors. The easiest way to do this is to copy all of the comments from the email (together with those of the editor), and paste them all into a fresh document. Use a different colour text or a clear set of symbols (e.g. >>>>) to indicate which text is your responses and which is the reviewers' or editors'. Or number each comment and reply. If it isn’t clear enough, then the editor may well get confused about what is the comment and what is the rebuttal. One of the best ways I’ve seen of doing this was to make a table with all the comments in one column (each on a separate row), and the author responses in a new column.

Sketch out your responses to the major comments, and use a tick if you are happy with making the suggested minor comments. If there are comments that you don’t know how to handle, simply leave them with a question mark. By making this start before you meet with your co-authors, you will have an idea of what is likely to be difficult to tackle in the revision. Even if you’ve received a rejection with reviewers’ comments, it’s well worth having this same meeting with your co-authors so that you can decide together what to do next. Skipping on comments from reviewers during a rejection appears to be very common [@crijns2021effect], but is a very uncollegial way of moving forward. I have personally reviewed manuscripts that were rejected, only to see them again as a reviewer in another journal with all of the same errors.

Make a plan of how to handle all of the comments, or where to go, what to read or who to talk to (perhaps another co-author), to sort out those you don’t know. Decide whether you need to send out the journal decision to co-authors now, or wait until you have your rebuttal ready to circulate. For me this decision is largely based on how much time the revision is likely to take: if it’s quick, rather send the revision and rebuttal together with the decision.

Next, when you sit down to write the rebuttal and revise the document, you need to make sure that you have pressed “track changes” on the submitted version of the manuscript. I find it easiest to have both the rebuttal letter and the revised manuscript open side by side on the screen. As you revise the manuscript in response to the comment, make a note to mark that it’s done in the rebuttal letter. Mark any comments that you don’t do. Your revision is written as a rebuttal to the editor. While you don’t write your comments back to the reviewer, it is worth bearing in mind that the reviewer is likely to read them.

Three watchwords should be your guides for your response to the reviewers:

**professional     -----    polite     -----        precise**

In addition to these, make the entire process easier for everyone by:

- Making a note of the line number where the revision is made (note that these can shift around in the revision)
- If you have reworded the text, do copy and paste that rewording into the rebuttal using quotes and corresponding line numbers
- Simply use a word like “done” to indicate changes on Minor comments
- Do be polite with your responses, but you don’t get any extra points for wordy thankfulness or praise. So keep it succinct and to the point
- Signal that you agree with the comment and that you have made a change to the text

Reviewers sometimes use a chatty style, and it may appear to you that they are asking you a question. For example, they might ask you exactly how accurate a piece of equipment you used to measure your organism. Intuitively, it seems like the right thing to do is to simply answer them in the rebuttal. But they expect you to make a change to the ms, and not simply to give them an answer in the rebuttal. Otherwise, it would have been pointless in making the comment.

Do bear in mind that your reviewer is a human, and was likely operating under less than ideal conditions when reading your manuscript. They could have been getting constant interruptions. They could have been reading it after having read another three manuscripts. They could suffer from insomnia and read it in the middle of the night with no sleep for a week. Give the reviewer the benefit of the doubt. Do remember to thank your reviewers and editors in your [acknowledgements](#acknowledgements). They’ve been working and doing the best for your manuscript without any thanks other than what you will give them. So give them a boost and help make their day that much brighter.

### What if you don’t agree with a reviewer?
Most of the time, reviewer comments are sensible, helpful and genuine attempts at improving the quality of your contribution. If you don’t agree with particular points, try skipping them and moving ahead with the easy points or those that you do agree with. Discuss any points that you don’t agree with your co-authors. Try to get another perspective on the comment. Do your best to try to see the comment from the reviewers standpoint.

For example, a reviewer might ask for details on a point in the methods, but they are mentioned in another section of the methods. This is a cue for you to add a flag to that point in the manuscript. For example, write: “see section 2.2.3 for an explanation of how this was done”.

If a reviewer has made a comment that says that they don’t understand something, this means that you need to make a change in your text so that the text is easier to understand. If they don’t understand, then it could be that more people don’t understand and you want your text to be understood by all people that are reading it, so make a change.

If you and your co-authors don’t agree with the reviewer, then make it clear what exactly you don’t agree with. Again, try to see it from the reviewer’s perspective and write a courteous and clear explanation of why they might have misunderstood or misinterpreted what was written. Back up your comments with citations, even if these aren’t cited in the paper. Provide full references for any citations you give. The more thorough your explanation of both sides of the disagreement, the more likely the editor will side with your perspective on the point that you don’t agree with. You may find that you want to include some of this text in the ms, or that you offer to provide it in the [Supplementary Information](#suppinfo) (if there’s a word limit on the ms).

Remember that the reviewer is likely to read exactly what you write in your rebuttal. Your job is to professionally explain why you don’t agree. Forget any of the emotions that you might believe to be there. Revert back to professionalism, because you are a professional.

### When reviewers ask for additional analyses or experiments
It is not unusual for reviewers to suggest additional or different analyses, or even experiments. It will be important for you, as author, to differentiate between requests that are reasonable and stay within the original bounds of your stated hypothesis, and those that do not. Because there are so many ways in which to analyse data, it is not unusual for a reviewer to suggest you use their preferred method over the one that you submitted. Such suggestions are made with good intentions, and unless there are clear reasons for not undertaking these analyses (such as you have already [preregistered your study](http://howtowriteaphd.com/transparency.html#preregistration) or they are inappropriate), you should attempt the analyses and then make a call on whether or not they improve your work. Even if you decide not to include the results, you can present them to the reviewer/editor in your rebuttal, together with your reasoning for not including them.

It is important to be aware of [p-hacking](http://howtowriteaphd.com/transparency.html) even in your rebuttal. Improving your work through peer review should not result in changing the focus of your work, or even including a co-variate that you did not plan to use. Many authors feel pushed into conducting extra analyses for fear of having their manuscript rejected [@hopewell2018influence]. Although it's impossible to determine every potential scenario here, if your work was well prepared and conceived, you should not need to conduct extra experiments and there should be journal policy to prevent this.

Exceptions might include when journals ask for independent experimentation to determine a mechanism detected (or speculated on) in the manuscript. Including multiple lines of evidence is likely to have your article accepted with higher impact. You may or may not have the option of doing this kind of extra work, and may therefore need to settle for another journal. In all cases, discussions with your co-authors should help you decide on the best course of action.

### When reviewers don’t agree {#conflict}
Normally, you will have two reviews (possibly three depending on the journal policy) and comments from the Associate Editor (AE). The AE acts as a judge given the opinions of the reviewers, and so if the reviewers disagree, the AE should suggest the correct direction for you to take. Sometimes this means that the AE will consult a third reviewer (and occasionally even more reviewers). This is one of the many reasons why it is important for [editors to read your work](#editors). If the AE gives you no direction (as is increasingly the case) then make this decision with your co-authors and indicate to the AE the conflict between the reviewers and the reason why you’ve chosen the direction you have.

### More than two rounds of peer review
You should not expect to have more than two rounds of peer review for most articles, but there are instances when this could happen. There are plenty of reasons why your manuscript might go into extra rounds of peer review, chief among these is the acquisition of new reviewers' after others are unavailable. However, the AE should be doing everything that they can to avoid this. In cases where it happens, you should be receiving helpful and specific comments from your AE. In the instance that you are making all possible changes to reviewer comments, but not receiving a clear and directed decision from your AE after three rounds of review, you can reasonably [appeal to the Editor-in-Chief](#unfair).

Note that on occasion it is the authors that are refusing to implement changes in the manuscript demanded by reviewers and editors. In this case, you should expect that your manuscript will be rejected. The most common problem that I see, is that editors fail to state exactly what they want. Then the authors and reviewers end up in unnecesscary rounds of reviews.

### What if you feel that your reviewer is being unprofessional?
Reviewers’ comments can come across as harsh, upsetting, rude and even arrogant. While it isn’t ok for reviewers to be rude, it does sometimes happen (see [Part 4](#problempeerreview)). If you really feel that a reviewer is being unprofessional, it is worth flagging this with the editor. I’ve never had to do this myself, but I am aware that there is some unprofessional behaviour out there (I’ve seen it on [ShitMyReviewersSay](https://shitmyreviewerssay.tumblr.com/)). Discuss it with your co-authors, but here are two potential options:

- If it’s just one or two comments, then simply indicate to the editor that you don’t feel that you don’t know how to respond. Ask the reviewer to try again, or ask the editor to interpret the comment for you.
- If it is every comment from one reviewer, write an email to the handling editor and ask for their guidance. You should find their email address in the journal submission site. They will flag it with the editor and come back with a solution for you.

### Appealing against a decision that you think is unfair {#unfair}
From time to time, a decision comes from an editor that is clearly unfair. I’ve had a few. As I’ve mentioned before, scientists are humans and humans do have [biases](#problempeerreview) that manifest into their professional lives. This is the reason for [double-blind review](#double). Scientists in STEM are predominantly white and male, and express the views of this minority but powerful group. Their prejudices do manifest in their decisions, and it is important to push back against this when you feel that this is the reason for a decision.

Most (good) journals will have an appeals process and you should look this up and see what’s involved. While doing this, it is worth reviewing the journal’s policy on how they handle manuscripts; again, good journals should have a clear policy. Of all the rejections and poor decisions I’ve had on my manuscripts over the years, I’ve only felt that decisions were unfair and worth appealing two or three times.

Normally, an appeal is made to the editor in chief. Be very clear about why you are appealing and what in the decision does not tally with the journal’s own policy. Remain professional and detached from the decision itself, and instead appeal on how the journal’s own policy was not followed. For example, a journal may have a policy that the editor will sum up all of the reviewers’ comments and use this as the basis for their decision. If the editor seems to have sided with one reviewer while not considering others, this can be the basis of an appeal.

Any appeal should be agreed with your co-authors before sending it.

### 2021-09-05 00-author.Rmd
Add files via upload
# all round update

@@
 
+## If you haven't already, read the other book {- #otherbook}
+
+**[How to write a PhD in Biological Sciences: a guide for the uninitiated](http://www.howtowriteaphd.org/)** by John Measey
+
+```{r fig.align='left', out.width = '45%', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://www.howtowriteaphd.org'}
+knitr::include_graphics('figures/Cover1.jpg', dpi = NA)
+```
+
+Embarking on a PhD is intimidating as, for most students, it will be their first experience working within the academic system.
+The voyage of discovery is often made very frustrating as much of what goes on in academia is assumed knowledge.
+Academics accumulate knowledge throughout their careers, but what can be done for those who are uninitiated?
+What is needed is a guide that postgraduate students can refer to before, during and while making decisions about their time within academia.
+Note that this is not a rulebook.
+There are times when the guide will be accurate and others when it will be vague, but providing some insight to point you in directions where you can explore more.
+The intention then is to provide you with a starting point from which you can establish your confidence in the academic writing process, and build your own creativity.
+
+## How to cite {-}
+Measey, J. (in press) _How to publish in Biological Sciences: a guide for the uninitiated_. CRC Press, Boca Raton. ISBN: 9781032116419
+
 ## Creative Commons License {-}
-How to write a PhD in Biological Sciences: a guide for the uninitiated is copyright to John Measey and distributed under a [Creative Commons BY-NC-ND 4.0 International License](https://creativecommons.org/licenses/by-nc-nd/4.0/).
-You may freely share this content for non-commercial purposes, with a source credit to [www.HowToWriteAPhD.org](www.HowToWriteAPhD.org).
+"**How to publish in Biological Sciences: A guide for the uninitiated**" is copyright to John Measey and distributed under a Creative Commons BY-NC-ND 4.0 International License.
+You may freely share this content for non-commercial purposes, with a source credit to www.john.measey.com
 

### 2021-09-05 1.00-Part1.Rmd
Add files via upload
# all round update

\mainmatter

# (PART) Part I: Getting ready for submission {-}

# Submitting your work to a journal {- #part1}

Part 1 of this book is about the publication arena. I provide simple explanations of what scientific journals are for, and who the gatekeepers are. The publishing world is at a turning point, and before you start publishing you should be aware of the current trends and what is possible.

### 2021-09-05 1.00a-closedtoopen.Rmd
Add files via upload
# all round update

# The transition from closed to open {#closedtoopen}

There are a lot of problems in publishing in the biological sciences, but it doesn't have to be this way. The aim for this book is firstly to help you navigate the current 'closed' system (while acknowledging that parts are already open), and act as a catalyst toward a more open, transparent and equal system for the future not only of publishing science, but permeating throughout the culture of the scientific project. We have all the tools to make this transition now, and I think that this change will likely come within the time frame of the careers of you as an Early Career Researcher. But as you will see, this change needs to be driven.

While I'm going to pitch the transition from closed to open publishing as a simple process, and as a move from darkness into light, I acknowledge that it might best be defined as a [wicked problem](#wicked). I hope that a lot of these complexities will come out in the book, but acknowledge that a lot will be left unsaid. A large unanswered question is what happens to all the downstream impacts of losing a lucrative academic publishing business that employs tens of thousands of people. I like to think that many of these skilled people, who themselves are not recipients of the large fees acquired by the publishers, would be absorbed into the repurposed institutional libraries. No doubt, there will be casualties. But my belief is that the importance of the scientific project, and the [wicked problem](#wicked) we currently face in academic publishing, far outweighs the problems that we will see during the transition.

Throughout this book, I will make reference to the '[scientific project](http://howtowriteaphd.com/lifescientific.html)' as a broader philosophical stance in which our studies in biological sciences is simply a part [see @measey2021how]. This is a stronger institution without the culture of assumed knowledge and elitism that we see today. I consider movement away from the current model of publishing to be at the heart of this essential institutional transformation.

## Three fundamentals of publishing in Biological Sciences

There are three fundamental concepts that lie at the heart of understanding of current publishing models in the Biological Sciences (Figure \@ref(fig:publishing-triangle)). In themselves, none of them should be particularly influential as they don't relate to your study, how well the study was done, or what your results were. Nevertheless, these three aspects of publishing are key in your understanding of the nuances of publishing, and your understanding will likely make the difference between publishing your work being an obstacle that is occasionally insurmountable, and finding your way with a lot more ease through the process.

### Gatekeepers {#gatekeepers}

The journal editor is the kingpin, and sits at the centre of this triangle, and has the power, backed by their gatekeeping advisory board and associate editors, to continue the current model, or oversee the change. [Editors make decisions](#editors), not simply whether to accept or reject your manuscript, but also to implement policies that take their journals in one direction or another. In some models, they are given this power (usually democratically) by the [scholarly society](#society) that they represent, and oversight is granted by an editorial board who mediate in any dispute, but also in theory have influence over the editorial policy. Editors appoint associates that handle many of the manuscripts that are submitted, shuffling them between reviewers and authors until they feel that they are worthy of publication (or not). These associate editors are also responsible for implementing the policies of the editor, the editorial board, and the scholarly society. In models where there is only a (for-profit) publisher, the publisher appoints the editor and together they appoint the editorial board. In both models, the editorial board, editors and associate editors are thought of as being the gatekeepers to the scholarly integrity that permeates scientific publishing. There is more information on the advisory and editorial boards in [Part I](#boards).

(ref:publishing-triangle) **A simple schematic for the three fundamentals of publishing in the current Biological Sciences model**. Citations, peer review and impact factors each have direct impact on each other and your understanding of each one and how they relate to the other will be pivotal in clarifying your understanding of how to publish your work. At the heart of the process are the gatekeepers: Editorial board, associate editors and the editor.

```{r publishing-triangle, echo=FALSE, out.width = '85%', fig.cap="(ref:publishing-triangle)"}
 knitr::include_graphics("figures/publishing-triangle.png")
```

Gatekeeping takes a lot of time and effort, and there are plenty of places where the current system lacks the transparency that is needed. In theory, there's nothing wrong with this model, were it not for the dark forces of for-profit publishers that have used the system for their own gain. Instead of the gatekeepers focussing on science, they have become distracted by metrics and hype promoted by publishers. Each of the three facets that surround the gatekeepers in the current publishing model (Figure \@ref(fig:publishing-triangle)) need to be changed to open up the system for a more equitable future, and eliminate the current biases that favour the select few.

### Is it possible to purge the publishers?

Editors don't have a complete free reign over what to do with their journal. They may be constrained by the publishers, if their journal is run on a for-profit model, or by a contract with a publisher, for most society journals. Journals which are independent of societies and publishers are very rare, but do exist. However they work, editors sit as king-pins of the system. There are examples of editors who have taken all of their associate editors and authors and moved their entire platform to a not-for-profit system. This has also meant changing the name of the journal, as the publisher often owns this. The first example, that I'm aware of, happened in 2015 in the social sciences where the editor of _Lingua_ walked away from publishers Elsevier [see @bakovic2017language]. As Elsevier owned the name, the editor, Johan Rooryck, started a new Open Access journal [_Glossa_](https://www.glossa-journal.org/). There was a fight (see [Rooryck's website](https://www.rooryck.org/interaction-with-elsevier)), but Rooryck showed that it could be done, and has therefore paved the way for others. Not only did Rooryck show proof of concept, but he formed the [Fair Open Access Alliance](https://www.fairopenaccess.org/), who have managed to pull 6 titles away from Elsevier since _Lingua_ flipped. It is important to add here that when the gatekeepers leave, the publisher simply approaches new people to take over, and the old journals remain alongside the new. It appears that there is no shortage of academics who are prepared to step into the shoes of editors who part company with the publishers. See what draws editors in [here](#boards).

All editors are answerable to their advisory board, and to the scholarly society from which they were (usually) voted into office. The society (often through the editor) signs a deal with the publisher, and this usually runs for a period of 5 years (see [Part IV](#money)). Society journals usually own their own content and title, and so can decide to leave the publisher whenever their contract expires. It just takes will power, and being prepared to say goodbye to that [income stream](#money).

**So why do the gatekeepers stay with the publishers?**
It's mostly smoke and mirrors. Editors and gatekeepers in general are busy people. Their gatekeeping roles are not their primary jobs (for the most part), and if they are then the publishers are paying their wages. What they are most interested in is a smooth system that works with minimum effort on their part. This is what the publishers have established so well, and the principle way in which they will try to persuade gatekeepers to stay with them. Next is the [money](#money), which flows from the publishers into the accounts of the societies, with occasional small amounts to editors (and in rare cases associate editors) as expenses. Such perks used to be more substantial, like trips to conferences, hotel stays and wining and dining. But I understand that this is largely gone now. The perceived prestige and professional advancement are [discussed in detail later for all gatekeepers](#boards). Last is the inertia on the part of gatekeepers to change, as they don't experience the pain of the authors or the libraries that pour money into the publishers. This is a reason why bringing societies into closer contact with institutions is an important step in the process.

## Peer review
Few would argue that peer review is at the heart of publishing in science today. Consequently, there are several chapters in this book concerned with the subject: [What is peer review](#peerreview); [what to expect from your peer reviewers](#peerreview2); [how to respond to peer review with a rebuttal](#rebuttal); [and how to conduct peer review](#howpeerreview). The last chapter covers [problems with peer review](#problempeerreview) and this digs into some of the real biases that occur during peer review, and with the reviewers themselves. Peer review isn't a perfect system, but in order to get the most from it we need to understand the weaknesses, both as authors and as reviewers. Only through this understanding can we reinvent the publishing system. We should not expect to do away with peer review (although this has been suggested many times in the past, and no doubt will be suggested again in the future), but by understanding the the biases that exist we will be able to make sensible choices despite the limitation.

Changes in the peer review system can make the difference between a highly biased system where editors manipulate the content for their own purposes ([networks](#networks)), or a system that is fair and equitable to all.

## Impact factor
[Impact factor](#impactfactor) is a simple metric, and as such there is no need for it to be anything more. But the publishers have managed to use impact factor to their advantage such that it has become of overriding importance for many publishing in the biological sciences today. But it doesn't have to be this way. To make the most of the system, you will need a thorough understanding of the way in which impact factor can [control other aspects of publishing](#money), and how the behaviour of the [editor can have a profound effect on the impact factor](#increaseIF). The higher Impact Factor results in more submissions, and this in turn will mean that the editor will have more power over the content of their journal.

## Citations {#cite}
Sitting above peer review and Impact Factor in Figure \@ref(fig:publishing-triangle) are [citations](#citations). Compared to the other two parts of this [wicked problem](#wicked), citations seem to be blameless and without the potential biases of the others. However, citations are the units of control for Impact Factor and can be manipulated by both editors and peer reviewers. [Metrics](#citations) driven by citations are also at the heart of many of the problems in today's publishing world.

-----------------------------------------

The relationships between Citations, Peer review and Impact Factor are indicated with double headed arrows in Figure \@ref(fig:publishing-triangle). There follows an attempt to describe these subtle differences, and the way in which they impinge on the gatekeepers at the centre.

#### Citations and impact factor  
The more citations a journal gets (in the first 2 years following publication), the higher its Impact Factor. Higher Impact Factor journals will want any submissions that they choose to have a good chance of garnering as many or more citations than their current impact factor following publication. This can mean that if you have co-authors that are in active groups, publishing a lot, they are more likely to have their work accepted in higher IF journals (they are also likely to have more experience at science and publishing).

#### Impact Factor and Citations

The relationship flows in the other direction as the higher the impact factor of the journal that you publish in, the more likely your work will be seen, read and cited. Citations garner more citations, and your work bearing your name will become better known within your discipline impacting your chances of getting a job, being invited to give talks at conferences or at visits to departments.

#### Impact Factor and Peer Review

Your peer reviewers are likely to be directed differently depending on the Impact Factor of the journal. As the Impact Factor increases, so the Peer Reviewers will be asked about the novelty and impact (i.e. citations) of your work.

#### Peer Review and Impact Factors

The impact factor of the journal may well influence which peer reviewers are prepared to review your work. In theory, better read (usually more senior) reviewers will provide more insight and may well refuse to review for journals that they consider do not garner sufficiently interesting (novel and impactful) work.

#### Peer Review and Citations

Your reviewers are likely to look at the papers that you cite and this will inform them of the scholarly quality of your work. When you miss important citations, or omit contrasting viewpoints, you demonstrate a failure in your scholarly undertaking to read the literature.

#### Citations and Peer Review

Peer reviewers may well suggest citations for you to consider in your work. These suggestions can be legitimate attempts to increase the value of your work, and through these citations its visibility. There are also reviewers who will suggest that you cite their own publications, and even editors that suggest that you cite papers in their journal.

## Open Science - a vision of the future {#openscience}

The way I advocate in this book is toward a vision for a future of Open Science (Figure \@ref(fig:open-science)). This future is both open and transparent. The transparency means that there is no need for this book, as there will be no hidden agendas or assumed knowledge needed for Early Career Researchers. Instead, it will be a 'what you see is what you get' system.

(ref:open-science) **A simple schematic for an Open Science publishing model**. Open science relies on the open nature of publication, communication and data. The gatekeepers still lie at the heart of this model, but are principally involved in ensuring the open and free flow of scientific information. This ideal world is free of the metrics that have dogged research in the past leading it into a profit making game for private publishers, and promoting the careers of a few elite scientists.

```{r open-science, echo=FALSE, out.width = '85%', fig.cap="(ref:open-science)"}
 knitr::include_graphics("figures/open-science.png")
```

The simple schema for Open Science shown in Figure \@ref(fig:open-science) is modified from O'Carrroll et al. [-@ocarroll2017providing]. The three areas of Open Science start with Open Data, the need to share both data, the code to analyse data, and the details for open source software with which to do the analysis all within open data repositories. The sector on Open Communication replaces the current closed peer review system with an Open framework where all actors are named and any interests declared. These include any journal gatekeepers (if involved). Lastly the publishing of the work is Open Access for other scientists and the public. This includes both proposals, preprints and published articles.

At the heart of Open Science publishing is ownership of the content. Like many aspects of changing the model from closed to open, implementation of the Rights Retention Strategy [see @janickehinchliffe2021explaining], through mechanisms such as [CC-BY](https://creativecommons.org/licenses/by/4.0/), is available today. This is something that you, as an author, should insist on [anywhere you publish](#copyright). If the journal refuses, then **think about what are they saying to you.**

In my opinion, the [Open Science framework](https://osf.io/) is incompatible with the for-profit scientific publishing model that drives the current model (Figure \@ref(fig:publishing-triangle)). It can be hoped that the transition from Figure \@ref(fig:publishing-triangle) to Figure \@ref(fig:open-science) will be swift and happen within your career. However, the actors at play in this system are not neutral, and need both bottom up challenges (from yourselves as ECRs) as well as top down pressure (especially from large funding agencies). We should acknowledge that changing our research culture within science is not an immediate process, but requires a suite of cultural changes (Figure \@ref(fig:culture-change)), that will start with early adopters and ultimately end with regulated policy [@nosek2019strategy]. These changes will be as exciting as they are challenging, and I hope that the contents of this book will equip you to participate fully.

(ref:culture-change) **Changing the prevailing culture in science will take time and effort, but is currently possible**. This book attempts to provide information about how a change towards Open Science is currently both possible and easy (green), but requires widespread adoption among the biological sciences community in order to make it normative and rewarding (orange), until we reach ubiquity through policy (red) [Redrawn from @nosek2019strategy].

```{r culture-change, echo=FALSE, out.width = '85%', fig.cap="(ref:culture-change)"}
 knitr::include_graphics("figures/culture-change.png")
```

### 2021-09-05 1.01-whatjournalfor.Rmd
Add files via upload
# all round update

# What is a scientific journal for? {#whatjournalfor}

It’s worth reflecting on why we have scientific journals, and what they are for. Primarily journals are a means of communication for the academic community. The academic community here should be regarded in the broadest of interpretations. For example, some journals also serve to disseminate research to other those who draw-up and implement policy. Others are aimed at managers who want to base their actions on scientific findings. The direction and inclusivity of the audience is pivotal in both [what](#whattowrite) and how we write.

- Journals record and disseminate the findings of individuals and teams of academics from all over the world. 
- By having a date when they are published, together with the names of the authors, they record primacy; that is who came up with the finding or idea first. Also see important arguments against the need for primacy [@casadevall2012reforming].
- They attempt to register legitimacy by collating and integrating comments and concerns through [peer review](#peerreview).
- Lastly, they archive these findings so that in future people can build on the work.

There are so many scientists in the world publishing so many papers that it's not possible for all scientists to read everything. Today annual growth in scientific papers is 5.1%, equivalent to a doubling time of 13.8 years [@bornmann2020growth]. Contrast this with the early days when there were only two journals and they published all of the studies that were being undertaken at the time. Over time there has developed a natural hierarchical system of what scientists will read. This is reflected in citations, and the simplest measure of journals is something called the [Impact Factor](#impactfactor).

It has been said that ‘authorship’ is a relatively modern concept, emerging from the empiricism of England’s middle-ages [see @cronin2001hyperauthorship]. In our recent history, it is considered to be important for individuals to record who thought of what and when. From these authors, we give societal ‘author-ity.’ This gives credit where it's due. In the big scheme of things of course it's not important who did it. We know from historical examples like Darwin's theory of natural selection and Wallace’s very similar thoughts will merely be a product of many people who were thinking about these ideas at the time. Although certain authors may be ‘ahead of their time’, the majority of thoughts and ideas that come around today are a product of their time. However, for individuals and their institutions it can be important to claim credit as this may translate into some monetary value (e.g. with patents) or a prestige value. The regulated system of a taxonomy puts a lot of importance on the priority of who described what and when.

The system of editors and peer reviewers determining whether or not a manuscript possesses sufficient merit to be published is still regarded as the gold standard in science [@mayden2012peer]. As you will discover, it is often a very high bar to achieve. Of course both editors and peer reviewers are human and so the system is not perfect. We'll talk more about peer review in [another chapter](#peerreview).

Archiving the findings of scientists is perhaps one of the most important roles of publishers that we should be most concerned about. In my career, I have seen the changes from strictly paper dissemination of scientific findings, as it was for the past 350 years, to primarily electronic findings many of which are never printed by the majority of readers. We should be concerned about how long these records will last. If you have never thought about the longevity of data storage, then this is something that you should give some thought to. We all need to change our perspectives on long-term thinking as this impacts almost every societal function (see the [Long Now Foundation](https://longnow.org/)).

### 2021-09-05 1.02-peerreview.Rmd
Add files via upload
# all round update

# What is peer review? {#peerreview}

Peer review is often considered to be the 'gold standard' of science [@mayden2012peer]. Manuscripts that have passed peer review are often considered to have been scrutinised to the highest level. If one's peers in the scientific community consider that a manuscript is worthy of publication, then it meets the high standards of peer review. In theory, the review of peers acts as the gatekeeper to all that is good in science, and excludes all that is bad. A lot has been written about peer review (>23000 articles!), and there's plenty more to read out there [@eve2021reading].

While the views in the above paragraph are generally held, there is also a general acknowledgement that there are a lot of problems with peer review. That this has been widely acknowledged is probably an understatement, as most people who have experienced would likely already know. These problems will be address in a [another chapter](#problempeerreview) in the last section of this book. The myriad of failures means that peer review shouldn't ever be exulted as the 'gold standard' touted by many publishers. Peer review does provide a filter of sorts, with the result being better considered as a '[silver standard](#fixing)'.

But peer review is here to stay and will remain as a fundamental aspect of publishing, and so there are several chapters in this book that are dedicated to different aspects. In this chapter, I attempt to explain what peer review is. Elsewhere there are descriptions:

- [what to expect from peer reviewers](#peerreview2)
- [how to respond to peer review](#rebuttal)
- [how to conduct peer review](#howpeerreview)
- [problems with peer review](#problempeerreview)

This chapter provides an overview of the topic, but you may need to refer to the other chapters first depending on what your current need is.

## History of peer review
The history of peer review is surprisingly modern. We have already seen that journals themselves only date back to the 17th century (see chapter in [Part IV](#whatjournalfor)). These journals included a form of peer review in that letters concerning studies could be published, along with comments made at presentations. However, the type of systematic enforced peer review described in this book is very recent [@eve2021reading]. The journal Nature for example only started systematic peer review for its articles in 1973, and mainstream editor led peer review only really started in the late 1940s [see @tennant2017open]. Typical society journals have followed a similar form of evolution from newsletters to scholarly journals [@measey2011past].

## How high is the peer review bar?
It is difficult to emphasise how high the peer review bar is. When your manuscript is scrutinised by your peers, it is very rare (practically unheard of in the careers of most researchers) that it will get accepted without modifications. This is because the experience of academics tends to be so wide, and vary so much from individual to individual, that it is almost impossible to predict what a peer reviewer will see when they read your manuscript.

You should expect that your manuscript will not receive an easy ride through peer review. But you should also expect that it will be improved. As [we will see later](#rebuttal), this improvement might not be immediately obvious to you when you first read the comments.

It is also important to note that as the author, you are the net beneficiary of the peer review process, and that once you press the submit button (free for the vast majority of us, but see [part IV](#paywall)), a cascade of events happen, all of which are done in the name of you and your submission. It stands to reason then that you should be sure that your manuscript is as ready as it can be for submission.

## Who are your peers?
Essentially the peers in peer review are people that editors find and persuade to conduct the peer review. It can be difficult to find people to conduct a peer review. Although only two or three reviews are needed sometimes as many as 20 or 30 individuals can be approached. Perry et al. [-@perry2012peer] lamented on the increasing difficulty in persuading colleagues to conduct peer review of manuscripts.

While your 'peer' might sound like someone who is in an equivalent position to you, this may well not be the case. If you are junior, your peer reviewers may be very senior. Equally, senior authors may have peer reviewers that may be very junior. Does this make a difference? For some people it might, especially when they know the other party and assign some level of competence associated with their seniority. Of course, both junior and senior researchers are capable of getting points in peer review wrong, just as both are also capable of providing insightful feedback. The editors are those in the hotseat about what it all means.

### Professionals
Peer reviewers are normally professionals. Academics postdocs or postgraduate students. Occasionally there are specialist amateurs who have very high academic standards and who can be contacted to conduct peer review. For journals with a special remit, industry professionals may be approached to provide feedback on applied aspects of manuscripts.

### Scholars
Peer reviews should be familiar with the subject area to a good level of scholarly achievement. Undergraduates and many postgraduate students would not be considered eligible by many editors as selection for peer review. Personally I found that many PhD students, especially those in their final stages of studying are very good peer reviewers.

### Specialists
Peer reviewers should be specialists to some degree of the area on which the manuscript is based. Often it's not possible to be a specialist in every area of a manuscript. But in the case where you are not proficient it is important to inform the editor.

## The role of the editor
The editor has an important role to play (see Figure \@ref(fig:One-round-of-peer-review)):

 - To assess your submission
   + Does it align with the journal?
   + Is it sound enough to send to peer review?
   + Whether to use a specialist associate editor
 - To choose the peer reviewers
   + Without potential conflicts of interest
   + Who can cover the content of the manuscript
   + Who agree to doing the review within the prescribed time-frame
 - To assess the reviews of the reviewers
   + Mitigate for potential bias in the reviews
   + Judge what is in the manuscript against what reviewers have found
   + Determine whether sufficient merit remains in order to undergo a decision (including more peer review)
 - Write the decision: When you bear in mind that the decision is likely to involve some arbitration between different reviewer opinions, and to direct the authors about what changes need to be made to a ms in order to make it acceptable, the decision is not a simple exercise. Making an editorial decision will require [careful reading of the manuscript](#editors), as well as looking past potential [biases of reviewers](#problempeerreview).

## Who are the Gatekeepers (Advisory & Editorial Boards) {#boards}

**The Advisory Board** (sometimes referred to confusingly as the Editorial Board) together with the editor and associate editors, make up the [gatekeepers](#gatekeepers) of scientific knowledge. Through their combined influence, they determine what knowledge is published by screening submissions and allowing only a proportion of them to be published. The Advisory Board are invited academics who are often considered to be leaders in the field relating to the journal subject area, and are invited to join by the editor and/or the society. In theory, they are ambassadors for the journal, encouraging authors to submit their manuscripts (for example following talks at conferences), identifying new topics for editorials and special issues, and generally supporting the editor and associate editors. The Advisory Board are also there in the case of dispute (especially between authors and editors), or for complaints coming from third parties (against editors or published articles). In general, most issues involving a journal are dealt with by the editor. Only in exceptional circumstances are the Advisory Board consulted. Practically, the role of the Editorial Board is largely passive, and hence many think of the gatekeepers as being only the editor and associate editors.

Here, I refer to the editor and associate editors collectively as the 'Editorial Board'.

**The editorial board** are said to support orthodox views in their fields and could be thought of as representing the 'establishment' [@crane1967gatekeepers]. The argument continues that like supports like, and that editorial boards in science tend to be composed of white men at US universities. One real problem with gatekeepers is their lack of diversity [@potvin2018diversity]. In a study of 250 science journals, only one country on the African continent had gatekeepers represented at 0.16%, while the USA had 53.87% of the gatekeepers [@braun2005counting]. Increasing the geographic diversity of the editorial board leads to an increase in the diversity of the authors [@demeter2018changing; @goyanes2020how; @potvin2018diversity], something as biologists we can all appreciate a real need for.

**Editors**, sometimes referred to as editors-in-chief (presumably to distinguish them from associate editors) are usually alone in their position at journals. Increasingly, journals with large numbers of submissions have joint editors-in-chief, or even another tier of editorial oversight under the editor-in-chief. The major part of their job is screening the submissions to the journal and assigning the most appropriate associate editor. In most journal models, they review the information provided by the associate editor and make a final decision on whether or not a manuscript is accepted to the journal (see Figure \@ref(fig:Editorial-work-flow)). Their gatekeeping role comes with the many decisions that they have influence over, for example what [kinds of articles](#scope) they will accept, the decision to amend the description of the journal on the website, which will impact how you [choose your journal](#whichjournal). This is the reason why you are always advised to look at the current content of your journal choice. There is also the important role that the editor takes [when things go wrong](#pruittdata), which can be very time consuming.

**Associate Editors** are most likely to be tenure-track faculty in a US research intensive university, according to an illuminating study by Kelsey Poulson-Ellestad and colleagues [-@poulson-ellestad2020illuminating].Most are within 10 years of earning their PhD, and are therefore still Early Career Researchers, and have published ~20 papers and conducted >50 reviews. Rewards include a better understanding of the publication system, improved communication skills, keeping current in the journal area and giving back to the scientific community. Most of the costs involved time around finding reviewers, reading manuscripts and making difficult decisions, especially around [conflicting peer review comments](#conflict). But [editing takes time](#editors), and can be burdensome in this respect especially for Early Career Researchers who have so much else on their plate. The advice of many Associate Editors in the study was for others not to take on editing unless they are sure that they can commit enough time [@poulson-ellestad2020illuminating]. Gender equality in Associate Editors of ecological journals has been improving over time, as has the gender equality of reviewers [@fox2019gendera], although other surveys have found huge disparity of only 16% of subject editors being women in 10 environmental biology and natural resource management journals [@cho2014women]. However, women are more likely to refuse an invitation to become an Associate Editor [@fox2019gendera]. The gender imbalance in gatekeepers is indicative of the general imbalance in gender across STEM subjects (for more studies see [here](https://collections.plos.org/collection/gender-imbalance/)).

(ref:One-round-of-peer-review) **A simple schematic for one round of peer review**. In this figure, you (yellow circle) start by submitting your manuscript (ms) to an editor of a journal (pink circle), who assesses it and send it out to 3 reviewers (green circles). Each reviewer independently generates an opinion in the form of a review (rev). They each pass this back to the editor, who then makes a decision on your ms. Each light grey arrow going out may be quite quick (a few days or weeks), but the dark grey return arrows might take a long time (often counted in months). Reviewers come in different flavours ([see below](#flavours)), generally at the discretion of the editor.

```{r One-round-of-peer-review, echo=FALSE, out.width = '85%', fig.cap="(ref:One-round-of-peer-review)"}
 knitr::include_graphics("figures/One-round-of-peer-review.png")
```

All of the processes in Figure \@ref(fig:One-round-of-peer-review) need to be worked around the editor's existing job, professional and research commitments (i.e. the day job), and their home life. The motivations for shouldering this additional work-load will be as individual as there are people in these roles. However, I have summarised some of the acknowledged motivations in Table \@ref(tab:motivations).

Table: (\#tab:motivations) **What are the main motivations for the various gatekeepers?** In this table, I summarise what I consider to be the main motivations of the different types of gatekeepers and reviewers of a typical biological sciences journal.

| Motivation | Editors | Associate editors | Board member | Reviewers |
| --------- | -- | -- | -- | -- |
| The prestige associated with the journal | X | X | X | - |
| Increasing their professional network | X | X | - | X |
| Increase their soft power | X | X | - | X |
| Participate in the production of knowledge | X | X | - | X |
| Give back to a system in which you've benefited | X | X | - | X |
| Part of obtaining tenure | - | X | - | X |
| Editors-in-Chief are usually selected from among the ranks of the Associate Editors | - | X | - | - |
| A better understanding of the publication system | - | X | - | - |
| Keeping current in the journal area | X | X | - | X |
| Prestige of having your name listed on the journal website | X | X | X | - |
| An opportunity to use soft power at conferences and other meetings | X | X | X | - |
| Recognition of being influential in your field | X | X | X | - |
| Another lever to use in arguing for promotion | X | X | X | - |
| Complimentary access to the journal  | - | - | X | - |

## Reviewer models {#flavours}
Reviewers themselves come in different flavours that are (mostly) predetermined by the journal regulations.

### Blind reviewers
Blind reviewers know who the authors are, but are anonymous to the author, but known to the editor. This can be considered the 'standard model' in peer review. There are plenty of problems with this model as reviewers may use their anonymity to [hide their biases](#problempeerreview) and are even known to [become abusive](#problempeerreview). Although reviewers may be anonymous, sometimes communities are so small that authors might guess who these people are, simply by their comments and suggestions. Although this is the most common type of review format it is the least recommended. If you feel that there may be potential reviewers who bear a grudge to your laboratory, your institution or your work then it may be better to avoid this kind of review system.

### Double blind reviewers {#double}
Double blind reviewers do not know who the authors are, and are anonymous to the author, but known to the editor. The double blind model was conceived to remove some of the potential biases (particularly around gender, nationality and race) that might come about through the identification of the authors and their addresses. Again, it has been mooted, and it is my also experience, that in a small community one tends to know who authors and reviewers are simply by the subject of the manuscript and the comments [see also @eve2021reading]. However, even when groups can be identified, it is not always possible to determine the author or author combination, and so biases around gender and race may still be avoided with this model.

### Triple blind reviewers {#triple}
In theory, it is possible for the editor, after having chosen the reviewers, to be blinded from their identity once they submit their review. This may prevent the editor putting more importance to a more senior reviewer, and ignoring more junior viewpoints.

### Open reviewers
Open reviewers know who the authors are, and are known by the authors and editor. Note that this is a simplification of a complex set of potential openness. For a thorough discussion see Ross-Hellauer [-@ross-hellauer2017what]. The open reviewer model does encourage good behaviour (or the avoidance of some of the worst problems) on the part of reviewers. However, reviewers remain brutally direct even when they are named, such that even open comments may be construed as bruising by the authors [@eve2021reading].

### Public reviewers {#publicreviewers}
Public reviewers know who the authors are, and are known by the authors and editor, and their names (and often their reviews) are made available to the public. This model is relatively recent, and comes along with the possibility of making the reviews with their own [DOIs](#DOI) available along with the accepted manuscript. It is worth noting that, to date, reviews for manuscripts that are rejected do not get posted using this or any other current publishing model. It does exist in the world of [preprints](#preprints2).

This could be considered the most transparent system for any journal. It has also been called Open Evaluation (OE) by Kriegeskorte et al. [-@kriegeskorte2012emerging]. [_PeerJ_](www.peerj.com) and [_eLife_](www.elifesciences.org) are among a very small handful of journals that have tried to instigate this model. Nevertheless it can be very difficult to find reviewers who are prepared to reveal their names to the authors. A study that compared _PeerJ_ publications in which reviews were made public, compared to those that authors chose to keep closed, suggested that the subsequent number of citations increased by a third for open reviews [@zong2020does]. It is only possible to speculate about why this might be. The decision at _PeerJ_ to open reviews is made first both by reviewers (who opt not to be anonymous), and then by authors (who opt to open reviews). In studies where both groups co-operate, we might hope that this results in a higher quality product. Indeed, public reviews tend to be longer, although positive comments are more frequent in closed reviews [@bornmann2012closed]. It is equally possible that authors who choose to open their peer review are more progressive and active within research (leading to more citations).

## Learn more about peer review by doing it {#reviewit}

As an early career researcher, you may well be asked to conduct peer review of an article in your specialist field. If you have never been asked, then tell your [mentor](#mentor) to recommend you. Usually when they turn down an opportunity to conduct peer review, they have an opportunity to name someone else. If you have told them that you want some manuscripts to review, it should be straightforward for them to add your name when appropriate.

Another option is to volunteer to conduct peer review for an independent peer review site, like [Review Commons](www.reviewcommons.org) or [Peerage of Science](www.peerageofscience.org). Here you can register your interest and then take your pick of articles that get submitted. A nice aspect is that Review Commons have Referee Cross-commenting, so that you get to see the other reviewer comments and make additional comments on these as you see fit. Review Commons and Peerage of Science are both excellent platforms on which to get some experience with reviewing.

Similarly, you could post your reviews of [preprints](#preprints2) online. You might be shy to do this at first, so consider sitting with a lab mate and doing a review together. You can always ask your [mentor](#mentor) to take a look at your review if you are unsure whether or not it should be posted publicly.

When you register in the editorial manager software for journals that you submit to, there is often an option to state what areas of your field you are particularly specialised in, and whether or not you are interested in conducting peer review in these areas. This is also worth doing if you want to generate requests for conducting peer review.

Another way of getting noticed to to sign up to society training programmes for peer review. These might happen at conferences, or could be through [web based courses](#onlinecourse). Although it should be noted that such courses may have little impact to improve your peer review [@schroter2004effects], you will certainly gain more insight. You will need to register to conduct such training, with the result (sometimes) that your name will be entered into the editorial management software, together with your trained status. Some courses actually have 'live' mentors who read through and critique reviews that you conduct. All of these are a good idea, but be sure to check out the time commitment required before you start.

Good reviews get noticed by editors, and it is a good way of increasing your network through soft power.

We will look in more detail about how to conduct peer review [later in this book](#howpeerreview).

### What do peer reviewers get out of participation?
Table \@ref(tab:motivations) suggests that reviewers receive the least out of the system for their efforts. Having simplified this in the table, there are lots of exceptions, and certain journals do provide incentives for reviewers, including free access to their content. Doing a good job of peer review will generate soft power if you renounce your anonymity. Doing this for a journal where your name is displayed alongside the article will help increase your profile.

Note that what reviewers get out of participating in peer review is almost identical to editors (see Table \@ref(tab:motivations)). Editors benefit at a higher level, mostly as their names are seen more often by more people, but with the drawback that they do a lot more work than the peer reviewer.

I have heard some members of the community claim that they will continue to accept peer review requests until these cover what they are demanding of the community (i.e. reviews offered per year = submissions per year x ~2.5). Although this sounds very fair, I would suggest that the reality is more subtle. You shouldn't be accepting to conduct peer reviews for articles where you feel that you lack specialist knowledge. Neither should you be conducting peer review when you feel that you have a conflict of interest.

If you do have to turn down the invitation of peer review, then do it as soon as possible, and try to suggest someone else that you think could do it.

### 2021-09-05 1.03-transparency.Rmd
Add files via upload
# all round update

# Transparency in publishing {#transparency2}

The need for transparency in science stems from the fact that most societies are making the majority of their decisions based on evidence coming through the guidance of science. If scientific evidence is at the heart of decision making, then the collection of this evidence must be transparent to those who make resulting policies, together with those who challenge them in a democracy. Moreover, as public funded scientists we should be setting a community standard of transparency for the rest of society to follow. Publishing of science is currently in transition to address the [wicked problem](#wicked), and it is vital that the scientific community leads the way forwards, and that we are not led by for-profit publishers. One of the ways to achieve transparency is through [preregistration](#commitment) of your research project to avoid [confirmation bias](#PruittData), which is, in part, a product of commercial publishers and the [metrics they promote](#impactfactor). In order to make this effective, we need the [gatekeepers](#gatekeepers) of our journals to support the [preregistration](#commitment) of research hypotheses and methods. Right now, journals should be openly advocating and encouraging preregistration with a plan to transition their journal in future to a system embracing rigour, reproducibility, and transparency [RRT @valdez2020improving]. However, many editors are resisting this move as they feel that there is no support from the community. This may well be the case, but inequalities in science, and particularly in publishing, mean that editors can either be instruments of change or at the heart of inequality in publishing (see [Part IV](#editors) and Figure \@ref(fig:publishing-triangle)). Either our editors will lead us toward transparency, or we as a community simply need to demand that they change their practices. Currently, editors are responding to calls for transparency by making small steps [for example asking for open coding: @powers2019open], rather than adopting transparency wholesale through the [badge system](#transparent) set up by Kidwell et al. [-@kidwell2016badges].

## Removing the prejudice around confirmatory bias
Confirmatory bias is the phenomenon increasingly seen in science that most studies published accept the alternative hypothesis, even though this is the least likely outcome of any experiment. Confirmation bias happens in publishing as editors prefer to accept papers that have a positive outcome. It has been suggested that this leads to a culture of '[bad science](#badscience)', and even [fraud](#pruittdata). One convincing set of evidence of confirmation bias is the decline of null results over time [@fanelli2012negative].

### Accepting the alternative hypothesis
At the outset of our scientific research we pose a hypothesis with the expectation that we will be able to accept or reject our null hypothesis. We often think of rejecting the null hypothesis as the only result that we are interested in, but if we only ever reported these results we would not be responsible in moving our field forwards. That is, in a world where we only report significant results (i.e. reject the null hypothesis) we would necessarily keep repeating experiments where the null hypothesis is accepted, because there would never be the evidence that the hypothesis had been previously tested in the literature. This is actually practised by the majority of scientific journals who won't consider a null result, and results in '[Publication Bias](#badscience)'. It's easy to see why this is a bad policy, but it is the prevailing culture in science publishing.

If journals only publish manuscripts that reject the null hypothesis [*cf* @franco2014publication], researchers are more likely to mine their data for positive results (P hacking), or re-write their hypothesis in order to reject the null (HARKing) [@measey2021how]. Deceptive practices such as p hacking, HARKing and [salami-slicing](#salami) are not in the interests of any journals, or the scientific project in general [@ioannidis2005why; @forstmeier2017detecting; @measey2021how].

### Inadvertent bias
But positive results don’t only come from deliberate manipulation of results. As humans we are predisposed towards positive results [@trivers2011folly], and there are plenty of reasons why researchers might inadvertently reach a false positive outcome. Forstmeier et al. [-@forstmeier2017detecting] draw attention to cryptic multiple tests during stepwise model simplification, and the two types of researcher degrees of freedom [_sensu_ @simmons2011false]: stopping rules and flexibility in analysis.

__Cryptic multiple tests during stepwise model simplification__ relates to the way in which adding predictors to models inflates the total number of models to test, making it necessary to adjust alpha accordingly (for repeated tests). However, Forstmeier and Schielzeth [-@forstmeier2017detecting] report that even with Bonferroni adjusted alpha levels, they found using random data that models with one significant effect happen around 70% of the time. The only way to keep this under control is to use sufficient sample sizes to maintain the power to distinguish between true positives and false positives. A handy rule of thumb from Field [-@field2013discovering] is that sample size needs to be eight times the number of model predictors plus 50. Better would be to run a power analysis on your study design, and to critically reassess your predictors to eliminate as many as you can before you begin your study.

__Researcher degrees of freedom__ is the way that Simmons et al. [-@simmons2011false] described ways in which researchers may inadvertently increase their chances of getting false positive results during analysis. The first is simply the way in which researchers decide to stop collecting data. Clearly, if preliminary collections showed a trend, but not a significant result, then collecting more data sounds like a good idea. However, as the collection of data is not independent (when the first set is kept) then the first test is not independent of the second, and so the chance of getting a Type I error is cumulative. Even if multiple datasets are collected, those that are insignificant should also be considered and reported in order to get an unbiased estimate. The second major way in which analyses can turn out with false positives is through potentially infinite flexibility in analyses. There are lots of ways to analyse your data and given enough trials, it is quite likely that you’ll find one that gives you significant results. Moreover, on the road to conducting the test, there are many options that can change the outcome of the analysis:

- Inclusion or exclusion of an outlier
- Inclusion or exclusion of a covariate
- Transforming dependent variables. 
- Inclusion or exclusion of baseline measures
- Controlling for sex (or another variable) as a fixed effect
- Excluding individuals with incomplete datasets

The potential list of ways in which the outcome of your analysis could change quickly grows as the number of ways in which you could analyse the data also grows. But don’t despair. Transparent help is at hand.

### Novel research
One criterion for many journals is that the research should be novel. This is increasingly practiced by journal editors as you move up the [Impact Factor](#impactfactor) levels. Novelty sells (just think of the meaning of “new” in newspaper), and that’s the basis for selling novel stories from higher Impact Factor journals. The perils of testing increasingly unlikely hypotheses and how this inflates Type II errors as well as increasing the proportion of Type I errors are widely acknowledged [@forstmeier2017detecting; @measey2021how]. Novelty also stifles repeatability. If we can never repeat studies in science, then a fundamental tenet of the methodology is repressed. Reproducibility in science has received a lot of attention recently, as attempts to reproduce the results of highly cited research have failed (cites). This has been followed by general outrage among scientists that things should change [@anderson2007normative; @munafo2017manifesto], including a majority of those in biological sciences [@baker20161500]. The irony that these reports and requests are published in exactly the journals that will refuse to publish research that seeks to repeat work (is not novel) is clearly lost on the editors. However, more nuanced views are also coming forwards to actively introduce variable conditions and sampling of biological variation into the study design to more fully represent the nature of biological variation making studies more likely to be replicated [@voelkl2020reproducibility].

## Introducing transparency in peer review
The way in which [editors choose and interpret reviewers](#problempeerreview) can either reinforce their own prejudices, or help to make publication more open and transparent for everyone. The first step along this road is to move from [double-blind review](#double) to [triple-blind](#tripple) where editors cannot make decisions with prejudice towards certain reviewers. Next is the need for [public reviews](#publicreviewers) with [DOIs](#DOI) that allow open assessment of what reviews contained. For more details about problems in peer review, see [Part IV](#part4).

In order to change this culture to a more transparent selection of scientific studies for publishing, we need journals to sign up to be transparent. Sadly, when most journals are approached, the editors either ignore the email or make an excuse about why it is not possible [@mellor2019advocating]. Of course, some journals have adopted the road to transparency, and we should be encouraged by the fact that they still exist, and that we could build on these initial front runners. In addition, there are a growing number of excellent frameworks that are pointing the way forwards [e.g. @macleod2021mdar]. This is a cultural change that we can expect will take time (Figure \@ref(fig:culture-change)).

## Removing profit from academic publishing
Taking out the profiteering from publishers will take a more concerted approach. But the reality is that we have only ourselves to blame. The publishers are able to prey on us, as biological scientists, because we are used to getting all of the "frills" associated with their publishing model. This includes the designer layout, custom websites and editorial management systems.

But the reality is that we really don't need any of these frills, and if we cared more about our science and less about the prestige that publishers have worked so hard to con us into thinking that their products have higher value, we'd all be better off. Mathematicians and physicists are way ahead of us. Given that they've shown the way, it's simply up to us to [follow their models](#overlay), and embrace the change in culture needed.

### 2021-09-05 1.04-whattowrite.Rmd
Add files via upload
# all round update

# What can you publish? {#whattowrite}

As you become more and more familiar with the academic literature you will quickly realise that actually you can publish just about anything. In this part of the book I'm going to talk about some of the most common articles that you can get published. But don't feel constrained by what you see here. When it comes to publishing you're only constrained by your imagination.

As an early career researcher, you will have a body of work from your thesis that you may have already published, or be in the process of publishing. These likely contain a number of data chapters that will be published as a series of papers in scientific journals. However, it's worth reflecting here about what it is possible to publish and how this might complement your existing and future publications, as well as increasing your publication portfolio with which to further your career. Although publishing is not the only way to do this, having more publications is likely to increase your visibility in your community, as well as giving you more practice in academic writing.

## Standard articles
You should already be familiar with the concept of publishing standard articles and you may already have a number of these published both as a first author and as a co-author. As an Early Career Researchers, you should consider what and how you publish. For example, you should consider whether or not publishing more articles is always the right strategy for you. You will find a chapter that discusses this concept in detail in [Part IV](#publishmore). In particular, you should be aware of the concept of [salami-slicing](#salami) a standard article into two or more different papers.

## Reviews
When considering the way in which citations work one of the things that you should notice is that reviews and in particular meta-analyses are cited many more times than most individual papers. For this reason one of the best things you can do is an early career researcher is to author a review on the topic of (or around) your thesis, or even better a meta-analysis.

I provided an in-depth chapter on how to approach writing a review including different aspects of meta-analyses in [another chapter of this book](#review). The importance of a timely and much used review can be seen in a lot of citation maps such as in Figure \@ref(fig:review-article). In this example, Ellender & Weyl [-@ellender2014review] was the first comprehensive review that sums the knowledge to that date on invasive fishes in South Africa, and so was cited most times that anyone published anything on invasive fish species in the country thereafter. Because this subject was the focus of a lot of research that happened in the area, you can see that it would logically sit at the centre of this subdiscipline. If you are unsure about whether or not a review is needed in your subdiscipline, then constructing a citation network based on the key-words in your subdicipline, such as that in Figure \@ref(fig:review-article), might well be useful.

(ref:review-article) **A timely review can be at the heart of a citation network, such as this one on "Invasive fish" AND "South Africa"**. In this citation network, you can see that the best cited paper (largest circle - green and centre) is a review by Ellender & Weyl [-@ellender2014review]. It has good connections with all of the three subject areas of this citation network, and although it was published in 2014, by 2021 it had been cited 111 times. Drawn with VOSviewer [@vaneck2010software].

```{r review-article, echo=FALSE, out.width = '85%', fig.cap="(ref:review-article)"}
 knitr::include_graphics("figures/review-article.png")
```

## Commentaries or Opinion pieces {#commentary}
Your opinion is important, or at least as important as anyone else's. Critical reading is a very important part of science and something that you should maintain throughout your career. From time to time you will come across articles and papers that you know are wrong, or fail to represent sufficient balance. Many journals will accept commentaries or opinion pieces based on articles that they have already published. This is an opportunity for you to make a correction to something that's already published in the literature. Please know that here we are not talking about anything you think might be fraudulent for that there is [another process](#fraud).

There are several things worth considering before putting pen to paper on your commentary and sending it to the editor.

1. If the people that wrote the article are in your network when a network close to yours then consider approaching them first about what you see as their error. You may end up getting along with them much better when you seek a solution together than writing something that antagonises them. Even if they aren't in your network you may find a way to increase the influence of your network through soft power instead of with a commentary.

1. A commentary should never be an _ad hominem_ attack. Never comment on the authors, only their content.

1. Check with your [mentor](#mentor) or another colleague that your interpretation of their error is correct and that pointing this out will have some value. Always try to do more than just say: "no it isn't". Many journals won't be interested in a commentary that does nothing more than show an error. If possible try and include some original data or some original analyses in your response.

1. Remember that your commentary will likely be sent immediately to the authors that you're commenting on before it is accepted by the editor. This means that they will also get a chance to comment on your commentary. However you will not get a chance to look at their comment.

1. Have a look through at some instances of where this has happened in the literature in your field. If you can talk to the people involved and try to find out whether things worked out positively for them. Although I do not want to say that you shouldn't do this, you should know that what you're doing is not going to backfire on you especially as an Early Career Researcher.

1. If you do decide to go ahead with this then consider asking other members of your network to join you. Although it's not a sheer game of numbers it may help you to gauge a better and more equitable stance on your commentary.

The other option you have is publishing a commentary that is very positive about the findings of a particular paper. Some journals published such commentaries about the contents of their journal as well as the contents of journals outside. Again this may be a better way of influencing soft power.

There are also lots of possibilities about publishing pieces on what it is like to work within your area of the biological sciences. This could be about towards your experience as an early career researcher, but may take on just about any stance that you feel is important in your area of biological sciences (e.g. language, covid, racism, colonialism, etc.).

## Letters
These are generally very short pieces that you can write, often to high profile journals with letters pages. They can be used to raise the profile of all sorts of issues within your subject area or profession.

Letters are going to have to be super polished and concise in order to get your point across in very few words.

## Editorials
You are unlikely to be able to publish an editorial without first being an editor, but there may be a potential for you to become an editor, associate editor or junior editor in a number of society or publishers' journals. Once in position, the editorial is a powerful place to launch your opinion to subscribers.

There are such things as guest editorials for special issues and special issues are particularly useful if you are an early career researcher. If you want to edit a special issue of a journal in your area then approach the editor well and advance of when you want to do it (possibly more than a year in advance). It's often good to have these things linked to an event like a [symposium](#symposium) that you are organising.

Once all the contents of the special issue are in you can put an editorial together to explain what the idea was of the symposium and co-authored along with your other symposium organisers. Special issues of symposium often get cited more than other articles just because it is a collection of similar things altogether in one place.

### 2021-09-05 1.05-impactfactor.Rmd
Add files via upload
# all round update

# What is Impact Factor, and why is it so important? {#impactfactor}

The impact factor of a journal relates to the number of times each publication from the journal gets cited in the two years preceding the date of the Impact Factor (IF) (Equation \@ref(eq:Impact-Factor)). Thus, if you are thinking of publishing in a journal that has an IF of 1 you might expect that in the two years following the publication of your article you may get one citation. But as this is an average for all publications appearing in this journal, it is not necessarily true for your paper. [As discussed elsewhere](#citations) you might be very good at publicising your work and have it extensively cited. One or two extensively cited papers might even change the impact factor of the journal if it doesn't have so many publications per year. If on the other hand you are thinking of publishing your article in a journal that has an IF of 5, you might expect that your article will be cited five times more than if you published in the first journal (IF = 1).

It's a relatively simple calculation as seen in equation \@ref(eq:Impact-Factor):

```{=tex}
\begin{equation*} 
\frac{\mbox{The sum of all  citations  in  journal X for  year Y }}{(\mbox{No. pubs in journal X for year Y-1}) + (\mbox{No. pubs in journal X for year Y-2})}
(\#eq:Impact-Factor)
\end{equation*}
```
Because all citations for year Y are needed before the IF can be calculated for each journal, IF for the preceding two years is typically not released until June of Y+1.

Impact Factors are published by a number of different literature databases. For example, for the impact factor calculated by the [Web of Science](https://www.webofknowledge.com/), if your journal is not even listed in the Web of Science then it will not have any Impact Factor. The Web of Science is continually policing the quality of the journals, and this means from time to time journals are excluded. This tends to happen at the lower end of the Impact Factor scale. But recently it happened to some very well known journals and there was a big stink. You can read more about this [here](https://retractionwatch.com/2020/07/13/stunned-very-confused-two-more-journals-push-back-against-impact-factor-suppression/).

Note that there are potential conflicts of interest with publishing impact factors. For example, a publisher, Elsevier, owns [Scopus](www.scopus.com) and can decide whether or not a publication can get an Impact Factor. Given two equivalent journals, do you think is it more likely that journals published by Elsevier will be included or excluded from Scopus? Organisations that are used by many of our employers in their means of evaluating our effectiveness are owned by private companies. This is certainly cause for concern. There is a group of people who are trying to replace Impact Factor with a group of other metrics, so perhaps by the time you read this Impact Factor will no longer be relevant. If you're interested read more [here](https://www.nature.com/articles/d41586-019-01643-3).

Do remember that there are a number of other citation indices including Dimensions and Google Scholar. Google Scholar doesn't currently calculate the Impact Factor for all journals, only the top 100.

## From a simple score to a way of life

When IF was originally devised by Eugene Garfield in 1955, it wasn't supposed to govern the lives of academics, it was simply intended to be a way of deciding which journals to include in the Science Citation Index [@garfield1999journal]. It then became useful for librarians to help them decide which journals to keep and which to ditch under ever constrained budgets (caused by publishers' ever increasing prices). But along the way, this very simple index is now considered by many people to be a measure of quality, prestige and even academic success [@garfield1999journal]. Many people have highlighted [how wrong these beliefs are](#money), but the growing trouble is that not only have many academics been misled, but so have administrators responsible for [hiring and promotions](#DORA).

> "Like nuclear energy, the impact factor has become a mixed blessing."
>
> ---Eugene Garfield [-@garfield1999journal]

In a paper by McKiernan et al. [-@mckiernan2019use], they found that IF features in the guidelines of many university panels responsible for the fate of academics jobs and therefore lives. Worryingly, many of these institutions don't actually talk about what IF measures. Instead they equate it with values and qualities that it certainly does not represent. Thus, you may find that your career is influenced by a simple metric that almost all who use it don't actually understand. The undue influence on lives of scientists that IF has led directly to the San Francisco Declaration on Research Assessment (known as [DORA](#DORA)) to which many institutions and publishers have signed up. You should read this very simple declaration and find out whether your institution is a signatory. If they are, then remember to hold them to the DORA principles during any assessment that you undergo.

The Impact Factor now dominates many aspects of life for Early Career Researchers, where the pay-off for a high impact publication might make the difference between having a job or leaving science altogether. The pressure is so high that it leads to [misconduct and fraud](#pruittdata).

## Five year Impact Factor

Many journals report their 'Five Year Impact Factor' in addition the the standard two year timeframe (as seen in equation \@ref(eq:Impact-Factor)). This is because many disciplines, such as biological sciences, don't have maximum impact of articles within the first two years of publication as do subjects like medicine and physics. Papers with immediate impact can be equated with sensational breaking news stories that instantly grab headlines, making money for those news outlets. These 'hot papers' or 'fast breaking papers' (i.e. left skewed distribution) will inflate the IF of the journal, which, if sustained, will earn the publisher more [money](#money). Most papers in biological sciences have a 'slow burn' (i.e. low frequency and long tails), consistently gathering small numbers of citations over long periods of time. Papers that suddenly become popular after a long time with no or very few citations are termed 'sleeping beauties' [i.e. right skewed; @bornmann2012anna].

Although the five year Impact Factor might be much more appropriate, most people ignore this metric. While this metric might be more appropriate, because of the bad ways in which people have used impact factor, it is probably better to [push back](#pushback) against this metric as you should against others.

## What can you do if you publish a journal with high IF?
Very high ranking journals for Impact Factor are publications like Cell, Nature, Science and PNAS. This is because these publications are read by a very great number of people, and so are widely cited. Articles that get published in them receive a lot of attention from the press and media. This results in the prestige that a hiring institution might be looking for. If your academics publish in this journal, your institution may well receive lots of positive publicity. In some countries, notably China, there may be a cash incentive towards publishing in a journal with a high IF [@quan2017publish].

One frightening trend that we are seeing in biological sciences is that the higher the Impact Factor the more the journal will charge you to publish in it. At the time of writing in November 2020 _Nature_ has just announced that they will charge $9500 to publish in their highest-ranked journal (see [Part II](#openaccess)). This is more money than it cost to publish in any other journal at the moment, and will be greater than the cost of many research projects, or even salaries of Early Career Researchers in some parts of the world.

## Why is IF so important?
Academics are measured by their productivity but also about the quality of their output. Because there are so many different academic disciplines the bean counters who administer us need some way of ranking academics against each other. This is why they use the Impact Factor of the journals in which their academics publish in order to determine the quality of their output. Even though there are other metrics of the actual quality of an academic, most administrators continue to cling to IF and their beliefs of what it stands for.

Some countries reward their academics if they publish in high ranking journals. This can result in a salary bonus. It may also help with promotion, getting tenure or even just getting an interview for a job. If you're going to publish and you want a career in academia then you need to be aware of Impact Factors and what they mean to different stakeholders.

Many people will complain that their particular sub-discipline has a range of very low ranking journals with low impact factors. Others complain that journals with high impact factors tend to be edited by an old boys club that facilitates the members. In some cases like PNAS this is certainly true.

## Editors try to increase IF {#increaseIF}
It's important to remember that editors care about Impact Factor [see @ioannidis2019user]. There are several reasons for this. Firstly, the impact factor of the journal can be used (by the publisher or society) as a simple measure of how well the editor is doing. Secondly, the higher the impact factor of the journal, the better quality of submissions of manuscripts to publish. Being the editor of a journal with a low (or no) impact factor can result in a lot of mediocre manuscripts being submitted. Editors can only choose their content from what is submitted. Poor manuscripts take up much more time than good ones: more rounds of review, more disagreements among reviewers and more time spent making editorial decisions. Thus, by increasing the IF of the journal that you edit, you are likely to increase both the number of submissions (allowing you to reject poorer ones) and retain better ones.

All this means that if editors believe that your paper will not garner the same or more citations in two years as the current Impact Factor of their journal, they may [desk reject](#deskreject). This is just one of the ways in which editors are known to manipulate impact factors for their journals. Established ways [@metze2010bureaucrats; @martin2016editors] of editors increasing impact factor for their journals include:

1. Ask authors to cite publications from your journal published within the last 2 years. 
1. Ask reviewers to suggest publications from your journal published within the last two years to authors on which their review is conducted.
1. Encourage the submission of papers from laboratories with high output and citation rates.
1. Reject papers that are likely to have no citations. This effectively reduces the size of the denominator in the above equation.
1. Publishing issues in January means they have a maximum period of the year to get cited. This is now being inflated to having issues published online well ahead of the January date all the time gathering citations. 
1. Encourage review articles which themselves garner more citations.
1. Editorials that cite every paper in the journal. This tactic is frequently used in special issues.

### Negotiating your IF
As the number of citations from your published content is divided by the number of papers, one way of improving the IF of a journal is to reduce the number of papers that are counted towards the denominator in the calculation made by the citations database. It has been known for some time that those journals with the highest IF negotiate the removal of all of their editorial and news content from their denominators, making the number of publications much smaller and hence the IF larger [@adam2002counting; @garfield1999journal]. On the other hand, if any of these news articles or commentaries get citations, these are included in the addition to the numerator. Thus our favourite high impact journals, _Nature_ and _Science_, can publish very citable news at the front of their magazine, but negotiate with the commercially minded database owners about exactly which of their content counts towards their IF [@brembs2013deep]. Later, we will see evidence of the financial leverage that these negotiations can reward these top tier journals (see [Part IV](#paywall2)).

## Push back against IF {#pushback}
Just like any metric, Impact Factor is liable for abuse. You need to be aware of how IF is used and abused by many people in the academic community. You also need to be aware of what the rewards are for these individuals. Our problem with Impact Factor is not really the way in which it is manipulated by individuals to achieve their own ends. Instead, we should be worried about the way in which it leads the scientific community towards [bad science](#badscience) and [dishonesty](#pruittdata). People who have benefited from using IF to measure their careers are likely to object if their institutions abandon it [e.g. @chawla2021scientists], even though they are already signatories of [DORA](#DORA). Retaining IF benefits these senior academics, their closed practices and the publishing industry who use this metric (that they own and police themselves) to extort money from taxpayers earmarked for research into their own pockets (more on this in [Part IV](#part4)).

If you must calculate IF, one very simple way that you can push back against industry calculated IF is to calculate IF scores for your own papers, and show how they relate to the IF of the journal that you publish in. In this way, you are simply comparing your actual citations in the two years after your paper is published with the mean for the journal. There is an even chance that you generally get more citations than the mean for the journal, and you can convincingly show that your citations are consistently higher than the journal IF. For this to be true, you might need to help your work get cited, and [that's the subject of another section](#citations).

### 2021-09-05 1.06-authors.Rmd
Add files via upload
# all round update

# When should you be an author? {#authors}

Who should be an author is being increasingly regulated because of widespread abuse, including ‘honorary authors’ or excluding ‘ghost authors’. Back in the day there were people that used to add fictitious characters to their author line (including their dogs!) as a joke. These days we have such a thing as [ORCID](#orcid) that attempts to register all authors, turning them into numbers ([see Part III](#orcid)). One of the benefits with the ORCID system is that you can have your name the way you want to have it, even if it is not a standard western style surname [@goyesvallejos2021what]. The need for all this regulation is because publications have turned into a kind of currency for academics. And when there is currency involved, abuse quickly results in human systems followed by the need for regulation. Hence, one of the results of going transparent is that we all become registered numbers (and our dogs won't get authorship any more). Perhaps this is just the loss of an age of innocence, at least on the part of our dogs!

A lot of the literature on the subject of ghost and honorary authorships has come from the medical profession, perhaps as this profession is prone to ghost authorship via the pharmaceutical industry [@matheson2016ghostwriting], and honorary authorship from heads of large research groups [see @rennie1994authorship].

## Ghost authorship {#ghost}
Ghost authorship is when people, not included in the author line, have contributed substantially to the study [@matheson2016ghostwriting]. Note that ghostwriting outside science is usually when someone who has the talent to write, writes the ideas of someone who doesn’t for a fee, and in return the latter takes the author line. These are called "paper mills" and they are discussed in [Part I](#papermills)).

There are many potential sources for ghost authors, including past students whose thesis work is taken by unscrupulous advisors, and published without their inclusion. More commonly, I believe, is that those who contribute substantially are not included as authors for political reasons (they have fallen out of favour with those who are the authors), or they are simply forgotten because they have moved away from the institution. Ghost authorship is a land of the disenfranchised. This is becoming increasingly prevalent in the world of contributions of data, which is also freely accessible. Some authors will take and use the data, only referencing the DOI for where they obtained it. Others will include the original people who created the data as authors because they value their continued insight and input. Where the situation becomes very messy is when some people are included and others are excluded. This is my experience where a paper simultaneously contains [honorary authors](#honorary) and excludes ghost authors. A study by Wisler et al. [-@wislar2011honorary] found ghost authorship in medical publications at 7.9%, although I’d argue that their methods (contacting corresponding authors) mean that the real levels are likely much higher.

While the world of inclusion (see below) has a warm and friendly glow about it (everyone appears to benefit), exclusion is characterised by lack of information, contact and reasoning. If you are excluded from a publication even when you have contributed, you will not be getting an email from the authors detailing their decision. You’ll be lucky if they even send you a copy once it’s published. Meanwhile, those who are included will remain in the loop.

### Paper mills {#papermills}
The concept of a paper mill is rather different from [ghost authors](#ghost) or even salami-slicing ([Part IV](#salami)). Paper mills involve third parties, often not included on the author line, producing material for publication from scratch: i.e. companies that specialise in producing content that will pass peer review for those who want to buy authorship. The world of paper mills is particularly shady and it is not clear what sums of money are exchanged for these types of goods. What we do know is that there are benefits to those who are put on the authorline, while those that produce the manuscripts generally rewrite text and pull protocols from manuscripts that are already published. Results are often images that have already been published and/or are manipulated to suit the content (see [Part IV](#fabrication)). Hence, paper mills are a systematic and deliberate manipulation of the publication process [see @teixeiradasilva2021abuse].

## Honorary authorship {#honorary}
Honorary authorship happens when people who have not contributed meaningfully to a study are included in the author line. If publications are the currency of science then you can see how being added to other people's publications increases your apparent productivity. While this might sound surprising to you, you should know that does happen and might be more common than you think. The Wisler et al. [-@wislar2011honorary] study found that honorary authorship was as high as 17.6% in medical publications. It’s worth noting that this may vary between disciplines as there are various traditions in some disciplines whereby the head of a large team may always be included as an author of a paper that emerges from the team, whether or not they were involved. In biological sciences, teams tend to be quite small with a single or rarely multiple Principal Investigators (PIs). This means that your PI will likely be directly involved with your research and therefore also an author. Imagine a very large team with multiple PIs working under a head who insists that they are an author on every publication. This could add up to hundreds of publications in a year (e.g. [Yuri T. Struchkov is currently credited with >1600 publications on Scopus](https://www-scopus-com.ez.sun.ac.za/authid/detail.uri?authorId=35427689100)), and such prolific authorship has been questioned [e.g. @rennie1994authorship]. However, there appear to be very different levels of what could be considered credible and what incredible [e.g. 25 papers a year: @wager2015too], and I respond to this, and the general question of how prolific authors are becoming, [later](#publishmore).

In the biological sciences, the area of molecular phylogenetics has traditionally honoured those who collect tissue samples with authorship on the resulting phylogenies when published. As I mentioned above, this is not always equal and has also been used to politically [honour](#honorary) or [ghost](#ghost). The reason given for honouring contributors in this way is that the studies could not have been done without the tissues. On the other side, some people have long lists of publications based on tissue donations and very little else.

## A need for transparency - DORA {#DORA}
Given the problems with both ghost and honorary authorship, there is clearly a need for transparency about who the authors are and what they have actually contributed. This was recognised by the [Declaration of Research Assessment](DORA: https://sfdora.org/), which has a growing number of signatories as well as some solid ideas on the way forward in assessment of research and researchers. In particular, DORA is against the use of Impact Factors ([see below](#impactfactors)) and other journal based metrics, and instead assesses the research on its own merits. DORA also encourages everyone to embrace the opportunities offered by online publication, including colour figures and unencumbered word lengths. DORA also encourages specific information to be published about individual author contributions. In short, DORA stands for transparency and it would be worth you looking at their statement and finding out whether your institution is a signatory.

There are no universal rules about what or how much you should contribute in order to become an author of a scientific paper. However, some journals are independently initiating their own standards, and this might become mainstream. Thus, it is worth discussing the criteria for being an author with your advisor preferably in a lab meeting so that everyone knows where they stand. there's a chance that your thesis chapters (when they are published) will be authored only by you and your advisor. But this can become much more complicated if you collaborate with more people in order to do the work. In turn you may (or may not) be invited onto their papers as an author. Confused? Then let’s look at this more closely.

## Who should be an author?
There is an increasing number of journals that now give clear instructions on who qualifies as author on a paper, and these have been formalised by the International Committee of Medical Journal Editors (ICMJE), and this area has been rapidly evolving [@baskin2011honorary]. Although these approaches are clearly made to prevent [honorary authorships](#honorary), it is hard to see how they help include [ghosted authors](#ghost). Scientists are notorious in their abilities to make _post hoc_ rationalisations. 
As an early career researcher, I'd suggest that you use questions around authorship as a way to formulate how participating researchers fit into the study. By openly attributing roles to a checklist, everyone buys into what is expected of them. Any additions to this author line-up should be openly discussed amongst all authors, and attibuted to the checklist, as the project proceeds.

### Rescognito, ORCID and the CRediT Contributor Checklist {#rescognito}
A new initiative under the name [Rescognito](https://rescognito.com/) has teamed up with [ORCID](#ORCID) to formally list the ways in which researchers are recognised per publication. Rescognito maintains the Data Availability Checklist, Contributor CRediT Checklist and Funder Information Checklist.

The [CRediT Contributor Checklist](http://credit.niso.org/) contains 14 fields:

Conceptualisation       Data curation       Formal Analysis       Funding acquisition       Investigation       Methodology       Project administration        Resources       Software          Supervision       Validation        Visualisation       Writing – original draft          Writing – review and editing

Visit their website to learn more about these fields. This area has new initiatives and registries opening up. We will see how many journals adopt these and what becomes of existing initiatives in the future.

### If you suspect irregularities in authorship

The [Committee on Publication Ethics](https://publicationethics.org/) (COPE) has published a useful flowchart to guide researchers on [how to recognise potential authorship problems](https://publicationethics.org/resources/flowcharts/how-recognise-potential-authorship-problems) [@cope2018how]. Or more specifically, how to recognise [ghost, guest, or gift authorship in a submitted manuscript](https://publicationethics.org/resources/flowcharts-new/changes-authorship) [@wager2006suspectedb].

### 2021-09-05 1.07-citations.Rmd
Add files via upload
# all round update

# Citations and metrics {#citations}

Citations matter as they reflect the number of times people who found your research useful. Although you can look at the citations of each paper, your publications will most likely be summarised into a citation metric to get an overview of how your work is used. Here I provide a guide to some of these [metrics](#metrics), how they are calculated and used. Growing your citations by getting your research cited is not always straightforward. Publishing your research in journals with higher [Impact Factors](#impactfactor) will likely help you get your research read, but there are lots of other tactics and I give some suggestions for [increasing your citations](#increasingcitations).

## What are your citation metrics? {#metrics}
A number of different databases compile metrics which you can access. In addition, they compile a bunch of extra performance metrics that you would have to pay to access, and in general these are available to a select number of recruitment staff at your institution. Here I will talk about Google Scholar, Scopus and the Web of Science [see @martin-martin2018google; @martin-martin2021google on their differences]. In summary, for Biological Sciences, the Web of Science (WoS) and scopus are likely to give you very similar statistics. Scopus favours social-sciences a little more, so if your subject area does cross-over then you might see some extra citations there. Google Scholar is going to give you the highest metrics, but it doesn't mean that they are particularly false (although this is certainly the not the best curated database). The joy of Google Scholar is that it is much more cosmopolitan on the journals and theses that it includes. So it actually gives you a better idea of who is using your research. It will also pick up grey literature and predatory publishers, so you should interpret the outputs with care.

Each of these databases offers you the opportunity of curating your own author profile that (in the case of Scopus and WoS) will link with your ORCID, and (in the case of WoS) link with your Publons account. Your institution may want you to actively curate your profile to help them with assessing your institution's metrics. It is surprising how bad these are when left to the bots, especially if you have your name recorded in more than one way. I also think that it's worthwhile creating and curating a Google Scholar profile, and link it directly to your [website](#website). As an editor, this is one of the easiest ways of looking up an author or potential reviewer. For anyone that uses Google Scholar, your name will then be underlined, and they can quickly find your other published works. Google Scholar also indexes preprints, which may be a boon to you as an Early Career Researcher.

Because all of these indices are cumulative, you can expect that anyone who started their career before you will have higher numbers. It is also worth being aware that some fields get higher citations because there are more people working in them, and therefore more articles are being published.

### Total Citations
Your total number of citations is simply the total number of times that anyone has cited one of your works within the respective database. Note that, the cited article itself doesn't have to appear in the database in order to be counted in total cites, but (clearly) the article that cites it must. For this reason, you are likely to have a different number of total citations in different databases.

### H-index
Very simply, the H-index is the ranked count of the number of your publications that have received at least the same number of citations [@hirsch2005index]. The easiest way to calculate your H-index is to have a list of all your papers and their citations in a list, ranked from the highest to lowest cited (see  Table \@ref(tab:cite)). Counting down from the highest cited paper, you stop when the number of citations reaches the same as the numbered citation in the list. For example, your H-index will be 5 if you have authored 5 or more publications with 5 of them having received 5 citations or more. If 6 have received 5 citations, you'll still only have an H-index of 5, because for the H-index to grow to 6, the top 6 articles will need to be cited 6 times or more.

Table: (\#tab:cite) **A table of fictional papers ranked by their citations to show how H-index is calculated.** In this Table, the H-index is 5 as the 5th paper has 5 or more citations, and no other lower ranked paper has been cited >5 times. Note that the H-index is not 6 even though the 5th paper has 6 citations. Once the paper currently ranked 6 receives one more citation, the H-index of this author will rise to 6.

| Rank | Title | Cited By | Year |
| --- | --- | --- | --- |
| 1 | A review of most important findings | 48 | 2017 |
| 2 | The first paper I presented at a conference | 19 | 2017 |
| 3 | Another significant finding | 8 | 2018 |
| 4 | First Open Access paper with snappy title | 8 | 2020 |
| **5** | **This paper spent a long time online first** | **6** | **2019** |
| 6 | Another review receiving some attention | 5 | 2020 |
| 7 | A paper that shows important findings | 2 | 2020 |
| 8 | Obscure note that I managed to self-cite | 1 | 2018 |
| 9 | The only citation is my own | 1 | 2019 |
| 10 | Hot off the press | 0 | 2021 |
| --- | **Total Citations** | **98** | --- |

Your H-index will grow very quickly as you publish articles from your thesis, but the growth slows as the H-index grows as it requires more citations of particular papers (of your top cited papers). Growing an H-index is therefore very difficult, and it's probably one of the hardest of these metrics to manipulate. Because the H-index is cumulative, people with very large H-indices (>100) are usually very senior academics with very good [networks](#networks), and low R numbers.

Note that while the H-index is usually calculated for individuals, but groups of people can also have an H-index, such as a department, a school, faculty or even university. Indeed, you can calculate an H-index for any set of articles using the citations that they receive.

#### Normalisaing the H-index and other citation metrics
The H-index (and other indices) are sometimes normalised for field of study (fixed - dynamic; broad - narrow), types of citing documents and year of publication [see @ioannidis2016citation]. You can expect to see some of these field normalisations, for example by [Dimensions](https://badge.dimensions.ai/) (see Figure \@ref(fig:Dimensions)) who provide you with, total citations (as the name implies), Recent citations (citations within the last 2 years), field citation ratio (comparing the article to ones of a similar age in in a similar field), and the Relative citation ratio (relative to 1.0, this gives you an idea of how well cited the article is compared to others in its area of research). As an Early Career Researcher, you may feel that your H-index is not particularly impressive (Table \@ref(tab:cite)), in which case, you may want to consider some of these normalised metrics if they show your articles in a more favourable light.

(ref:Dimensions) **Normalising citation metrics.** Some databases will provide you with normalised values for citations, like the one shown here by Dimensions for a paper cited  45 times, and 8% of these citations are from the past two years (normalised for time since published). The Field Citation Ratio (FCR) of this paper is 4.07 times more citations, when compared to other papers in the field. The Relative Citation Ratio (RCR) compares this paper to other NIH funded publications in the same area of research and year, where they will be the same as the mean (at 1.0), more or (in this case) less at 0.57.

```{r Dimensions, echo=FALSE, out.width = '50%',  fig.cap="(ref:Dimensions)", fig.link='https://badge.dimensions.ai/details/id/pub.1001281366'}
knitr::include_graphics('figures/Dimensions.png')
```

### G-index
The G-index is similar to the H-index, but places more emphasis on the actual number of citations in your top cited articles [@egghe2006theory]. To calculate this, you will need to organise your publications by the number of times that they have been cited. For ease of explanation, start with your H-index, and add up citations to all of the preceding articles (1st, 2nd, 3rd, 4th & 5th). For example, in order for your G-index to be 5, all citations up to 5 will need to be more than or equal to 25 (5^2^). For your G-index to increase to 6, the sum of all citations to your first 6 papers will need to be equal or greater than 36 (6^2^). Our fictional researcher in Table \@ref(tab:cite) has a G-index of 9 as citations to the first 9 articles add up to 98 which is greater than 81 (i.e. >9^2^).

Because the G-index uses all citations in your top cited articles, not simply the minimum used to obtain the H-index, your G-index should always be equal to your H-index and will often be higher. This is because of the phenomenon that articles that get cited more, continue by getting increasing numbers of citations (known as the [Matthew Effect](#winnertakesall), see [below](#winnertakesall)). Therefore, you may favour using the G-index if your top papers are particularly well cited.

### R-index
Your R-index is a measure of your active network [@ioannidis2008measuring]. It is rather like the H-index it that it records the count of the number of people that you've collaborated with on a number of papers. But these need to be the same people. This is then divided by your total number of papers. For example, if your thesis produced 5 papers with the same 5 authors and this is your total number of papers, your R-index will be 1. However, if two of the papers lacked one of the people, your R-index will drop to 1.66. If you don't work with those 5 other people again, then your R-index will grow continuously until you have another productive network. See the section on [networks](#networks) for a fuller explanation of the R-index and the importance of networks.

### a-index
The a-index is calculated from the total number of citations divided by H^2^ [@sidiropoulos2007generalized]. In the example given in Table \@ref(tab:cite), Total citations is: 98. H^2^ is 25 (5^2^), and so the a-index is 98/25 = 3.92. This index gives a measure of how top-heavy someone's citations are. The a-index gets larger as the highest ranked papers receive a disproportionate amount of citations. This is known as the [Matthew Effect](#winnertakesall) (see [below](#winnertakesall)), and this is a general feature of a lot of researchers' citation ranks.

### i10-index
Your i10 index represents the number of articles that you have written which have been cited 10 times or more in that database. The i10 is often expressed over a time frame (for example the last 5 years) which may well exclude papers that you write that have an immediate but not a lasting impact. Moreover, looking at an index calculated over the last 5 years will let you know how active older colleagues are. Given that most indices are cumulative, older researchers almost always look better. But a time constrained index may provide additional insight into their relative productivity.

### H-frac
H-frac is the same as the H-index, except that citations for articles are divided by the number of authors [@koltun2021h-index]. Koltun and Hafner [-@koltun2021h-index] propose that H-frac is more robust as it represents a better correlation with awards to individual scientists than the H-index (see their [excellent website](https://h-frac.org/) for an explanation). Their analysis shows that it accounts for changes in author behaviour, such as [hyperauthorship](#authors), based on their retrospective analysis of authors in physics. I am wary of advocating a new index that may drive future behaviour in ascribing [authorship](#author). If citations are divided by number of authors, this will likely make lead authors more circumspect about adding potential authors to their byline. Given that there are already difficulties in some [authors](#authors) being represented, I prefer not to have a disincentive for adding authors. Like other analyses that are made on vast amounts of existing data in [citation databases](http://www.howtowriteaphd.org/databases.html), this does not predict how future behaviour will change when the incentive changes.

### Other metrics
There are more metrics for authors behind paywalls, but there are also metrics on social-media sites such as ResearchGate's 'RG score'. The way that ResearchGate calculates it's RG score is rather opaque and is not simply done on publication metrics, but on the way that you interact with your ResearchGate community. Similarly, many publishers will grant you metrics that will be boosted by the number of times that you publish with them.

There are also Altmetrics handed out as a means of showing the alternative impact for articles that you've written. These are explained in [Part III](#altmetrics).

## How to increase your citations {#increasingcitations}
Increasing your total number of citations is likely to impact positively on all of your metrics, but this is not necessarily true, and hence why you may be assessed on more than one metric. In the following subsections, I go through some of the well known ways in which you can increase the readership of your paper, and with this the likelihood that someone will cite it.

### Media release
Some journals will want you to write a media release that they may use to promote your paper. You can also proactively [write a press release](#pressrelease) and send it to the media office at your institution. Writing a release for the media is very different from scientific writing, and you can find a guide in [Part III](#pressrelease).

### Social media {#social}
Social media is another way of getting your article to the attention of more people. Twitter has been adopted by many scientists and tweeting out an article that gets retweeted by the right people (i.e. social media hubs - with lots of followers) can get you tens or even hundreds of thousands of views. This is likely to be far in excess of anything your article will get by passively sitting on the journal website.

At this point it is worth telling the tale of Neil Hall, who in 2014 wrote a slightly mischievous short article proposing a new citation metric which he called the **K-index** to reflect the number of followers on Twitter divided by the total number of citations of the same researcher [@hall2014kardashian]. In the article, he found (with a non-random selection of Twitter researchers) that there was indeed a correlation between well-cited individuals and the number of their Twitter followers. He then postulated that some people with large numbers of Twitter followers (high celebrity) were not authoritative as they did not have lots of citations (a measure of scientific value):

> "... a high K-index is a warning to the community that researcher X may have built their public profile on shaky foundations,..."
>
> --- Neil Hall [-@hall2014kardashian]

A social media backlash followed swiftly. Within days of the online publication, the Tweets started rolling in, with a great many people very upset with what they saw as an insult to younger researchers who had a good social media following on Twitter [@woolston2014clash]. The social media eruption over this article made it score an [Altmetric](#altmetrics) of 2489, ranked 1st of articles published at that time in that journal, and 9th in all journals. Although Hall had written the article in a light hearted way (poking fun at metrics and social media), the backlash was massive (although some did appreciate the joke).

There are several lessons here: First, is that what seems like a joke to you (including those immediately around you) may be offensive to others, especially those from other cultures. Second, is that items in social media can go badly wrong, and be taken in ways that you had never imagined. Third, is that social media goes well beyond academia - as is the intention with the dissemination, and that this may also have repercussions.

The bottom line is that social media can be an excellent way to reach a very wide audience, but that you must use it responsibly, or risk that it might do you and your reputation considerable damage.

#### Academic social media sites
Sites such as ResearchGate and Academia.edu will play a role in increasing citations to your work, especially if you have a large following. Rather like sharing on traditional social-media sites, the more people who are exposed to your work, the more likely they are to read and cite it. In the case of academic social media sites, the audience is more niche and so more likely to find your work of interest.

### Popular articles
[Writing popular articles](#populararticle) is a good way to ensure that your paper gets some media attention. There are outlets like [The Conversation](https://theconversation.com/) that specialise in publishing popular articles written by academics. But you might also find that academic societies that you belong to have newsletters that you can contribute a popular article to about your paper. As long as the circulation of these newsletters is big it should enhance the numbers of people that eventually read about your work.

### Self-citations {#self}
Self citation is when you cite a paper that you or one of your co-authors have written or co-authored. This is probably one of the oldest and most widely practised ways of increasing citations to your articles. It is generally frowned upon to have gratuitous self citations, but the situation may be complex and biased [see @flatt2017improving].

However, it is also true that you are more than likely to cite your own work because you have probably published in the same area previously. By the end of your PhD you will already be citing publications that you produced at the beginning. Self-citations are also known to facilitate the '[Matthew effect](#matthew)', whereby because you increase your level of citations of a particular paper, others will also cite it [@fowler2007does; @flatt2017improving]. There is evidence to suggest that the proportion of self citations will change in different disciplines, and that they will increase with increasing numbers of collaborative co-authors [@davarpanah2009author]. Self-citations are also known to increase with increasing career length [@mishra2018self-citation], presumably with the growth of ones own publication base in relevant areas. Because of this effect, self-citation can be regarded as a measure of highly productive authors [@mishra2018self-citation].

There has even been a call for another metric, the **s-index** which is the equivalent of the h-index but for self citations [@flatt2017improving]: "the total number of s papers that an author has published that have at least the same amount of s self-citations".

How do you make sure that their citations are not gratuitous? Simply by only citing a work when you need to. Note that many citation indices will provide your citation scores both with and without self citations. As a rule of thumb, metrics suggest reasonable levels self-citations in the biological sciences are ~30% [@aksnes2003macro]. In short, there is nothing wrong with citing yourself, but you should only do so when you need to.

### Citing papers submitted and in press
If you, or others, have papers that are 'in press', then you can cite these in a manuscript submitted to a journal for peer review. However, 'in press' does mean that they have been formally accepted by an editor at a journal. These days the concept of 'in press' is not quite the same as it was. It used to take months before an accepted paper would appear in a journal issue. These days it is likely that you'll have an online version very quickly after acceptance. Hence, by the time you submit a manuscript with citations to papers that are 'in press', by the next round of revisions or acceptance, they should be fully available. If papers are not available by the time you send back your proofs, then you may well be asked to remove these citations. My impression is that many authors claim that a paper is 'in press' when the reality is that it's simply been submitted and that they are hoping it will be 'in press'.

You should not add citations to manuscripts that have simply been submitted to other journals. If you really need to, then submit and cite as a [preprint](#preprint) is really your only option. Some editors will not accept citations of preprints. An exception to this might be when papers are submitted together to the same journal at the same time. If this is the case, then you can point this out to the editor in your [cover letter](#letter).

If the contents of this manuscript are pivotal for understanding the submitted version, then you should make sure that there is a copy of the 'in press' or 'submitted' manuscript available for the reviewers. The easiest way of doing this will be as a [preprint](#preprint). If this is not possible, then you may need to upload the ms as an additional file - but you should seek guidence from the journal editor.

### Presenting the results of your paper at a confrence {#conference}
Probably the most traditional way of generating a bigger audience for your work is to talk about it (or present it as a poster) at a conference with an audience that is likely to appreciate your work. Everyone else at the conference will likely also be wanting you to know their work, so you'll need to target your audience carefully (possibly in a [symposium](#symposium)), and make sure that the audience can access your work (for example through a QR code in your slide or on the poster). Making your talk memorable will likely have people remember your work, and cite it if they are undertaking a similar or relevant study.

### Organising symposia and having special issues {#symposium}
Having your research published in a special issue is a great way of getting citations. This is because special issues have research on a theme and so many people will be drawn to the issue and are then more likely to see your work. This includes other people who participate in the special issue and are _de facto_ already interested in this line of research. This increased visibility of your research is, in my opinion, a good reason for joining a special issue if you have the chance.

Special issues are often edited by people who have an interest in pulling together this kind of research. They may encourage other authors in the special issue to cross cite articles, and may write an editorial that includes citations to yours and other papers. Of course, the best position to be in is when you are one of the organising members. Your name will be associated with the special issue as well as your contribution. This means that when the special issue is published, each of the papers in it will already have citations leading to a positive [Matthew Effect](#matthew) from day one of publication.

Probably the easiest way of pulling together a special issue is to organise a symposium at a conference and ask those participating to contribute to the special issue. If this is done long enough in advance and you are good at organising then it can work very well. Talk to your [mentor](#mentor) about the possibility of their help in organising a symposium.

### Traditional Media 
To get your work cited you need to get people reading it. As scientists read the [traditional media](#pressrelease) it can be a way of bringing your work to their attention, and result in additional citations. There are other benefits of having your work highlighted by the traditional media. You should always make sure that you get a plug in for your institution and any prominent funders. There are now additional metrics (e.g. [Altmetrics](#altmetrics)) that track traditional and social media and in the future you might be able to use these altmetrics to your advantage in getting a job, a promotion or tenure.

### Not so legitimate ways of increasing citations {#notlegitimate}
As well as the legitimate ways of increasing citations and visibility of your work that I've listed above there are other illegitimate ways that I have heard about. I have heard of certain laboratories that have _quid pro quo_ arrangements where they agree to cite each other's publications [see @ritchie2020science].

Another way of increasing citations might be to have a co-author that is very well known. Certainly having a co-author that is very well known is likely to increase your citations but only if they actually do something to contribute to the work. See the section on [honorary authorship](#honorary).

Gratuitous self citations. [As mentioned above](#self), you should really only cite your own work when necessary, and not more than this.

Journals have also been known to manipulate citations in order to increase their [Impact Factors](#impactfactor) and therefore their perceived level of quality. A group of physics journals from Romania were found to have clearly manipulated self-citations to increase their own impact factors [see @heneberg2016excessive]. There can also be a level of coercive pressure from journal editors to cite papers from their journals published within the two year citation window [@chorus2016large-scale]. Such practices are now systematically analysed by the larger literature databases, and clear levels of [Impact Factor](#impactfactor) manipulation result in delisting of the journal from the database.

## Well cited articles are likely to be cited more {#matthew}
Articles that already have a lot of citations are more likely to pick up extra citations than the ones that have very few. This phenomenon, sometimes referred to as the Matthew Effect [see @teixeiradasilva2021matthew; @merton1968matthew], is exaggerated by search engines like Google Scholar that order the results of searches by the number of times an article is cited. Like people looking for a website, academics looking for a paper are more likely to choose one on the first page of Google Scholar than to keep searching through Google Scholar until they find your study. In effect, this means that those that already have large research groups and that can generate initial citations, will get more citations and increase their own standing. Or, in other words, those that are already in a commanding position will automatically generate more [@casadevall2012reforming] - the winner takes it all.

As you read through papers in your specialist area you will notice that there are some papers that seem to get cited over and over again as standard examples of a particular phenomenon. These are often some of the first examples published and have also been published in higher ranking journals. If you are the 20th person to have shown a particular phenomenon then it is unlikely that your paper is going to end up being cited the most.

There have been investigations into what makes certain papers more likely to be cited further. One of the factors that appears to be significant is the "small-study effect" - low precision studies that produce large effect sizes, either spuriously [i.e. a Type I error: @forstmeier2017detecting, or genuinely @fanelli2017meta-assessment].

### 2021-09-05 1.08-advisorless.Rmd
Add files via upload
# all round update

# Growing your network {#advisorless}

During your PhD you will have had an advisor who would have been central to many of the decisions that you made. Don't give up on that relationship, it is something that you have at the heart of your network. Therefore it is worth maintaining in the best professional relationship possible, as this is someone that you will need to go back to (probably lots of times) for references and other professional help. In short, always keep your advisor sweet. The same could be said about your thesis committee members, if you had one.

Having said this, as an early career researcher, you are now advisor-less, yet it is likely that you will still need advice from someone more experienced. Hopefully, there will always be your advisor, but the chance is that you may well have changed institutions, and so they may not be so close or easy to contact. This is a good reason for expanding your network by looking for a mentor to your role as an early career researcher. Increasing numbers of institutions are now recognising that early career researchers need mentors to advise them about how to move forwards within their academic field. Moreover, mentors can help promote equality and inclusively [@davies2021promoting].

## What to look for in a mentor? {#mentor}

The first thing to look for in a mentor is someone that you feel you are able to talk to easily. As you will see, there are a wide range of topics to discuss in terms of building your career, and you will need to find someone that you feel you can talk to about nearly everything safely and without fear that it might be used against you. The mentor should feel happy to talk and spend time with you. They should be someone that is genuine in their desire to help you and your career. You won't have a fruitful relationship with someone that doesn't really like you, or doesn't want to spend time talking to you.

Next, your mentor should be someone who is already well established within an academic network that you want to (or already are) a member of. You will need to have common ground to talk about, and getting their advice on building your network will necessitate talking about how to manage others in the network.

The personal touch is nice to have, so if you can find a mentor who has an office in the same building, or on the same campus, then it's great to be able to meet up and chat. However, your mentor doesn't have to be in the same institution, or even in the same country. In these days of post-pandemic communication we are all a lot happier hooking up to meetings online.

If you still have choices, my suggestion would be to look for someone that you share non-academic interests with. For example, you may both enjoy playing squash together, hiking or scuba-diving. Meeting outside academia will make your mentor-mentee chats easier for both of you. Otherwise, try not to make your meetings with your mentor a burden on their work time. All academics are busy and core-working hours are hard to come by. Instead you could suggest taking your mentor out for a coffee or a sandwich during which you have your chat.

Note that a mentor doesn't have to be someone that is particularly senior, although they should be well connected. Having said this, they are likely to be at a more advanced career stage than you.

## Questions to put to your mentor?

Your mentor's time is valuable, so always have an agenda when you go to talk to them. Have ideas that you want to pitch and get feedback from. Have some way of making notes about what they say, especially if you meet over a beer. You don't want to have to go back and ask them again for the same advice. Be prepared to get the answer that they give. This seems an odd thing to say, but they may not always like your idea, and you should be prepared to listen to their advice whatever it is, not only when it agrees with your own ideas.

Typical things to discuss with your mentor are:

- Where to publish
- What meeting to attend
- How to extend your network
- Whether (or not) to write a [reply or commentary](#commentary) on another groups' work
- Whether to respond to a call for grant applications
- Whether to apply for a job

You may also want to share some of the more exciting aspects of your work, your findings and how you interpret them. Try to keep away from moaning about other academics, the amount of administration you have, teaching burdens, etc. Make your discussions as positive as possible so that your mentor is more likely to want to maintain and grow the relationship.

## The importance of networks in your academic career {#networks}
There is clear evidence that researchers with good networks collaborate more, publish more and are better cited [@parish2018dynamics]. Your mentor can be one way to increase your academic network. Perhaps more importantly, they can be a guide to help you through the network that you are involved with.

There is a good chance that you are already in a network where your advisor, and perhaps some members of your thesis committee, sit at the hub of an extended network that you have had partial exposure to during the course of your studies. If you haven't realised already, these networks are of profound importance in every step of your academic career.

**Benefits of a good network include:**

 - increased citations for your work
 - better chances that your work will be edited or reviewed by a member or an associate of your network, resulting in a higher chance of publication or a grant application accepted
 - increased potential to be nominated for an award
 - increased potential to be invited to give talks at conferences
 - invitation to apply for positions and jobs in the labs of partners at institutions in the network

If you are already in a network, you should be thinking of ways in which you can get more from your network now that you are an early career researcher. If, on the other hand, you are not in a good (sizeable and supportive) network, you should be thinking about how to join an existing network or how to form a network around yourself and your work.

In many cases, managing your network is about providing opportunities for yourself and others in your network. There are many ways in which you can do this, but you should be prepared to put in a lot of work from your own side.

**Potential initiatives include:**

- writing a grant application that involves others in your network
- leading a piece of written work that includes others in the network
  + a review
  + a commentary on another article (not from your network!)
  + a viewpoint or position article on an emerging topic
- organising a symposium aimed a central or emerging question within your network
- organising a special issue of a journal (potentially as a result of the symposium)

Your chosen mentor is the best person to discuss how to make waves in your network. Remember that relationships have already been forged in networks and you should never try to destabilise these. Instead, you should aim to become another node of the network by growing your own usefulness as a positive force for new and interesting initiatives that others will want to join. Because you are new, don't expect to be included in every initiative. If you feel that you are being left out of other initiatives that you've heard about unnecessarily, ask your mentor whether or not it is appropriate for you to bring this up.

Always bridge and build relationships. As you should have become aware by now, there are many potential ways of becoming involved in combative elements of academic life. Your own sub-discipline will likely have groups of people that write commentaries about the work of other groups. Networks are powerful places to be, but they can also become destructive when another network opposes a position. I have seen this type of behaviour happen within and around my own limited subdisciplines. One group will publish a major finding in _Nature_ or _Science_ and a week or two later, the other group will publish a commentary pointing out what are (to them) flaws in the first group's publication. I always have a sneaking admiration for those individuals that appear to swim between these groups, always maintaining good relationships with both sides. Theirs is really an interesting, but perhaps precarious, position.

If you feel that you don't have any network, then I think that you should reassess your position. You likely do have people that you talk to and interact with that are already part of a network, but your question is how to increase the size, and potential scope, of your network. In the same way that everyone will have their own path through their career, the best way to grow your network will be unique to you.

### Calculating the size of your network
A simple metric for your network is the total number of co-authors that you have from all of your publications. Clearly, as you increase the number of people that you work (and hopefully) publish with, you will have a bigger, better functioning network. However, scientometricians have their own ways of interpreting and mapping networks (see Figure \@ref(fig:Olaf-SA-invasive-fish)).

(ref:Olaf-SA-invasive-fish) **An author network for the Web of Science search for "Invasive fish" AND "South Africa"**. In this author network you will see my friend and colleague, the late Prof. Olaf Weyl, who is at the centre of a large group of people who worked on invasive fish in South Africa. Olaf's wasn't the only research group, but certainly the most influential and joined up. Olaf's collaborations were extensive, spanning continents and generations of researchers. Drawn with VOSviewer [@vaneck2010software].

```{r Olaf-SA-invasive-fish, echo=FALSE, out.width = '85%', fig.cap="(ref:Olaf-SA-invasive-fish)"}
 knitr::include_graphics("figures/Olaf-SA-invasive-fish.png")
```

If you aren't sure about who's in what networks within your own researcher area, you can use appropriate keywords to download literature within the area that you work, and plot (using [VOSviewer](https://www.vosviewer.com/)) a similar author and co-author network. I'd suggest using enough keywords to call ~1000 articles (including their citations) that include your research area (all or the majority of your papers should be included). Using a _map based on bibliographic data_, read the bibliometric files and analyse using co-authorship to describe the linkages between authors. This can be very revealing (see Figure \@ref(fig:Olaf-SA-invasive-fish)), and while you should already be well aware of the big names in your area, some of the people that span areas and sub-subdisciplines might well be worth seeking out. When looking into their publications, ask yourself how they managed to span between networks (conducting a postdoc in each lab is one such way). More than this, you can use collaborations to combine networks of other Early Career Researchers in your subject area.

Another method for determining the size of your network, or those of potential collaborators around you, is to calculate your R value [@ioannidis2008measuring]. **R** is made up of two figures, _I~1~_ which is the number of authors that appear in at least _I~1~_ papers, divided by _N~p~_ which is the number of papers that the author has published. _I~1~_ is a little like the H-index in that it increases each time you add a person in your network. For example, to have a _I~1~_ of 4, you need at least 4 publications in which you and the same 4 other authors occur. Parish et al [-@parish2018dynamics] show that as _I~1~_ grows, so the value of **R** decreases and researchers become most productive as **R** approaches 1.

Once you start to calculate your own value of **R**, you will appreciate that not only are large networks important, but that it is useful to start these networks early and maintain working with an every increasing group throughout your career.

### Using social-networks to grow your nework
The ways in which you can grow your network have never been so great. Social networking is a massive way to reach other researchers and you don't have to wait for a meeting to do this. The #academictwitter groups are already very large and established. Investing time in getting connected is worthwhile, especially when you can meet up in person through conferences and symposia. I don't want to make out that Twitter, and (no doubt) other social networking is easy, but it is a good way of hooking up with a whole lot of other people without having to leave your desk. You can pitch ideas, and get collaborations moving in your area. Once you have it working for you, it is very empowering; remaining aware that it can [go wrong](#social). If social networking is your thing, then I'd suggest trying to find yourself a 'social networking mentor' - someone that already has a good following and can guide you through the various nuances of (often unspoken) rules that happen on these platforms.

## Creating your own website {#website}
Whether or not you have your own lab as an Early Career Researcher you would be doing yourself a favour by creating your own website. Having a website is a way of creating your own brand and making you distinct from whatever institution or group you are a member of.

Websites don't have to be fancy. There are plenty of ways in which to make good-looking websites using templates (Weebly, Wix, etc.). You can make these unique to you simply by using your own photographs. If you're not a photographer then consider using some of the images from your work, or ask a colleague whether you can use some of theirs. Remember to place the most important keywords for your area in the website metadata so that it will pop up in any general web searches. There's a lot of information out there on how to build a successful website that people can find, so I won't go into this here.

**What to include on your website**

- Make sure that you have your professional contact details, including your institutional email address. This will help any editors for other colleagues that are trying to contact you. Avoid having a form is the only way of making contact with you
- Write some general blurb about yourself and your work including all relevant keywords for your subject area
- Have a list of your publications, conference presentations, popular articles, patents, and any other outputs that are relevant to your professional profile
- Provide a summary of your professional research experience listing where you did your degrees and any post doctoral work
- Make sure that your website works on lots of different platforms. Preferably it will work on mobile devices as well as desktops and laptops
- Link to your various academic profiles: Publons, ORCID, Scopus, ResearchGate, Academia.edu, GitHub, etc.
- Provide a link back to your institutional website and your current lab, as this will help legitimise your independent website standing. People want to see that you are who you claim to be

**Optional items**

- You could write a blog
- You could have short articles about some of your publications
- Downloadable version of your CV
- Future plans for projects
- Collaborator page - with links to collaborators in your network
- Data from projects (e.g. images, sounds, videos, R code, etc.)

**Things to avoid**

- Be careful about mixing personal and professional profiles
- Don't provide links to any personal social media accounts, only professional ones
- Don't allow your site to be too static (static sites rank lower in search results)

### 2021-09-05 1.09-preprints.Rmd
Add files via upload
# all round update

# Preprints {#preprints2}

## What are Preprints?

So far in this book, I have explained about the process of [peer review](#peerreview) how important this is and later I explain why it's the [silver](#fixing) (and not gold) standard in science. Preprints came to us from the world of physics. An academic world that is moving so quickly that many inside it don't want to wait for peer review before making their work public. They date back to 1991, and were the brainchild of Paul Ginsparg with his preprint server, *arXiv*. Today, *arXiv* hosts nearly 2 million articles in 8 subject areas: Physics, Mathematics, Computer Science, Quantitative Biology, Quantitative Finance, Statistics, Electrical Engineering and Systems Science and Economics.

In addition to making your work open access, it allows anyone to read and review it. This is unlike the traditional publishing model where editors invite selected reviewers. Many consider preprints as a kind of open peer-review system.

> "The life science community needs to return to a culture of evaluating scientific merit from reading manuscripts, rather than basing judgment on where papers are published."
>
> ---@vale2015accelerating

In biological sciences the most prominent preprint server is called [bioRxiv](https://www.biorxiv.org/).

## Who posts preprints on bioRxiv? {#bioRxiv}

In an analysis by Abdill and Blekhman [-@abdill2019tracking] back in 2018, there were 37648 preprints uploaded to *bioRxiv*. But in the world of preprints, this information is quite dated as back then there were only \~2000 uploads each month. Luckily, these same authors regularly scrape data regarding submissions to *bioRxiv* and I'm able to reproduce these live data below. It is interesting to look back at the concern back in 2015 that preprint servers like *bioRxiv* might not catch on in the biological sciences [@vale2015accelerating], and it is still true to say that the number of preprints in the life sciences is still dwarfed by the annual number of publications, whereas physics has seen the opposite trend.

(ref:bioRxiv-line-annotated) **The growth of bioRxiv submissions in time.** Note how the rate of submission rocketed at the beginning of the pandemic lockdown, and has remained high ever since.

```{r bioRxiv-line-annotated, echo=FALSE, fig.cap="(ref:bioRxiv-line-annotated)"}
if(knitr::is_html_output(excludes="markdown")) knitr::include_url("https://johnmeasey.github.io/bioRxiv-line-annotated/index.html", height="350px") else knitr::include_graphics("figures/bioRxiv-line-annotated/annotated-line-example.png")
```

What you see is a doubling of bioRxiv submissions dating back to May 2020 (Figure \@ref(fig:bioRxiv-line-annotated)), a few months after the start of the global COVID pandemic. (Go to the [RXivist website](https://rxivist.org/stats) to see the incredible spike in [medRxiv](https://www.medrxiv.org/) data at the same point in time.) Many scientists had spent a few months working from home. Some had been productive, and many decided to move this productivity for the first time onto bioRxiv.

In answer to the question above, **just about everyone now posts preprints on bioRxiv**. Those that don't likely use other preprint servers, or are not moving with the times.

The preprint revolution has not gone unnoticed by the tech giants. Back in 2017 the [Chan Zuckerberg Initiative](https://chanzuckerberg.com/) made an undisclosed donation to bioRxiv. The advisory board also hoata the architect for Google Scholar, Anurag Acharya.

## Why might you want to post a preprint?

One of the advantages of posting a preprint is that it gets a [DOI](#DOI) [(digital object identifier)](https://www.doi.org/). You can then use this DOI to refer to your work even though it is not published in a peer review journal.

For example imagine that you've just finished your thesis and only one of your chapters is published. How can you show to prospective employers how good your work is? Or if you are applying for money how can you refer to your work even though it's not published? A preprint is a simple solution to this problem.

Other benefits include having a wider scope of peer reviewers. If you know that in your subject area there are many people who may want to comment on your work constructively, then this would be an opportunity to give them access. Importantly, because a preprint has a DOI, your work is not vulnerable to theft. It also allows you to stake your claim on the work that you've already done even though there may be a lag time between this and it coming out in a journal with full peer review.

If you do want feedback on a manuscript that you have posted as a preprint then you will need to tell people about it. A good example of this would be after providing a talk, or a poster, at a conference you might show a QR code where people can read your manuscript as a preprint. you can also publicise it to your community on social networks like Twitter.

If you get lots of feedback on your manuscript then you should expect to incorporate it. So be careful what you wish for, because you could be opening yourself up for a lot of comments.

Posting a preprint on bioRxiv is also a shortcut to submission to a growing number of traditional journals. Manuscripts and Supplementary Information can be [transmitted directly](https://www.biorxiv.org/about-biorxiv) from the preprint server to many journals without the need to upload the files and metadata a second time.

## Upload newer versions

If you, or others, spot errors in your preprint, or you find new literature to cite, you can update your manuscript with a new version. Indeed you should do this for as long as your preprint remains active. Once published, make sure that there is a pointer from your preprint to the published version.

## Will you have to post a preprint?

This field is moving quickly. In December 2020, at least one journal (eLife ‘publish then review’) announced that they won’t accept a submission until a preprint has been registered. Thus all reviews are made on preprints. Other journals, like PLoS, are announcing in house preprint servers. You should expect this area to rapidly change in the coming years, so no matter when you are reading this, you are more likely than ever to need to submit a preprint.

At the time of writing, there are still some journals that make it a condition of submission that there is no preprint. Make sure you check within your target journal list.

If you choose to publish in an [overlay journal](#overlay), then you'll have to deposit your submission onto a preprint server.

## Could these comments pages really replace peer review?

Peer review is often regarded as a gold standard in scientific publishing (although a [silver standard](#fixing) would be a more realistic), and there’s certainly a lot to that. It ensures that published material has been read and its contents assessed independently. But peer review is fallible, because scientists are all humans. These problems are discussed at length in [Part IV](#problempeerreview).

In 2003, Stefano Mizzaro proposed changing peer review to the format that we now see in preprint journals. Let every reader become a reviewer. Another take on this same theme is provided by Heesen and Bright [-@heesen2020peer] who argue for a more subtle change in the date of publication (prior to peer review as seen in preprints) instead of after peer review. Here their emphasis is on removing the wasted time spent reviewing and then rejecting manuscripts that will never be published. Their discourse is very persuasive, yet given that both models currently exist, we need more ideas on how we could drive a preprint model forwards. More ideas do exist, and I encourage you to explore those proposed in a special issue edited by Kriegeskorte et al. [-@kriegeskorte2012emerging].

In the preprint model, [several peer review problems](#problempeerreview) might be overcome as no-one chooses the reviewers. Instead they choose themselves, and are motivated to do the work. Their competence to cover all aspects of the manuscript is not assured, but one assumes that independently motivated reviewers will only comment on parts that they are able to assess.

All of this is very good, but will people actually read and comment?
A quick look at the sites will tell you a lot about the level of reviewing that is currently going on in biosciences preprints. Looking at the top 10 articles on _BioRxiv_ (zoology section) confirmed my suspicions. Plenty of tweets about the articles, but none of them had any comments, let alone reviews. Indeed, a further trawl through PeerJ Preprints, also found no comments.

When reviewers aren't chosen, there's potential for manipulation. This could promote a culture for comments to preprints for well known labs, and (conceivably) a certain amount of trolling for labs with ongoing disputes or rivalries. This would make preprint peer review a sort of trial by popularity. But I don’t see a situation where potential reviewers will take time-out once a week (for example) and hunt for manuscripts that have received no comments. It seems far more likely that the authors will have reciprocal agreements with other groups to review each other’s manuscripts. This nepotistic tendency then puts us back into the area of [problems in peer review](#problempeerreview) that we’ve been working hard to overcome now for sometime.

## Preprints are here to stay

It is clear that preprints are with us to stay. The year of the COVID-19 pandemic (2020) saw an explosion of preprint papers on the topic, but also saw the misunderstanding of what these articles mean by the press and general public alike. Rapid sharing of results via preprint servers has already been put in place following the outbreak of Zika virus in South America (back in 2015-16), but the global nature of the Covid public health crisis saw much larger numbers of preprints being placed online.

But the value of preprints will always be limited for as long as there is no peer review. Moreover, comments won't suffice for peer review as there is no editorial oversight. As you'll read elsewhere, the [role of the editor](#editors) is pivotal in publishing.

## When should evaluation end?

One point raised many times in the special issue edited by Kriegeskorte et al. [-@kriegeskorte2012emerging] is that evaluation should be open ended: __ongoing evaluation__. There was a consensus to see reviewers continue to question the contents of papers long after publication. But these authors don’t appear to have a realistic perspective on the time taken by authors to rebutt their work. Imagine the effort that you currently put into a [rebuttal letter](#rebuttal). Now consider that your first rebuttal might come after a few months, and then you need to compose another after a few years. Perhaps you are the only author still working (especially if it is the work of your students). Perhaps all of your co-authors are dead! Suddenly you are called upon to defend your work, potentially decades after completion. Can you do it? Would you want to do it? What would be the consequences of not doing it? Would people start dismissing your contribution?

While I am regularly the first in the queue to criticise the current peer review system. I am also very grateful that publication represents a line in the sand under which I won’t have to continue working on a project. In a world in which I had continually documented every step of every experiment, I can imagine that it is potentially possible to find a _post hoc_ defence for every step in a protocol. But the painstaking nature and time involved in going through old work would be an added burden that I cannot welcome with any enthusiasm. Personally, in a world when I have the option of working on a new project or endlessly and repeatedly defending old ones, I’d pick the new project every time.

## Are preprints published?

As they each have a DOI ([Digital Object Identifier](#DOI)), they are in their own way already published.

Another point is that these articles are picking up citations. And there is a new concern that these articles are being cited, even when they are subsequently available through a published journal. This is one of my personal concerns with using a preprint service. I’m happy to put the paper out there for public comment, but the idea that it’ll remain there and that readers won’t necessarily be redirected to the peer-reviewed version does concern me.

Another question is what happens to manuscripts that are placed on preprint servers, are then sent out for review but not published because they are fundamentally flawed? It’s not as if the reviews are not made, but there is no automatic link to the reviews by the journal that conducted them.

There are certainly a lot of manuscripts out there with fundamental flaws. These are often sent for peer review, but those reviews pointing out the errors won’t necessarily make it back to the comments page on the preprint server. I think that this is a serious problem. The reviewers have spent time and effort and the very reason they do this is so that manuscripts with fundamental flaws don’t find their way into the literature. However, preprint servers have, perhaps unwittingly, found a loophole that allows manuscripts that are not scientifically robust a backdoor to citations.

**If preprints are fundamentally flawed, can’t everyone spot it?**

No. Reviewers are chosen by an editor with great care because of their speciality area is in their particular domain. They have insights that not everyone will be aware of and these are an important aspect of the purpose of peer review.

I edit for the journal [PeerJ](www.peerj.com). Although there can be various reasons to be rejected from PeerJ, normally it means that your paper is not scientifically sound. As PeerJ has no selection for impact, rejection does not normally mean that it can be simply submitted to another journal. I have noticed that manuscripts that I have rejected from PeerJ are still available as preprints without any comment on their failure during peer review. In my opinion, this is not good as it essentially ignores the input given by both reviewers and editors. The article appears as if it has had no comments or attention, when this is not the case. In a system where we move to relying more on preprints, why would we want to ignore chosen peer reviewers for whom this article was within their specialist area? According to Google Scholar, the rejected article is gaining citations, again raising concerns that rejection by peer review is not a hurdle to entering the scientific literature. All of this calls for reviews to be linked more directly to preprints, no matter where they are published. A model that deals directly with this is [overlay journals](#overlay).

## The exciting new world of Overlay Journals {#overlay}

Having said that preprints won't replace the role of peer review, what if we did have good, editorially coordinated peer review of preprints? What if, instead of these manuscripts effectively leaving the preprint system, they were updated together with the reviews that prompted the updates, each with their own linked [DOI](#DOI)? What if the journals themselves were simply pointing to collections of papers that had been curated in this way? Simply a website that throws a veneer of a journal as waypoints to peer reviewed journals. This world has already been imagined and is functioning in mathematics, where Overlay Journals have begun to prosper.

According to Brown [-@brown2010introduction], the idea of overlaying has been with us for some time, and exists as websites that offer a series of links to other papers. In this way, a review article could be considered an 'overlay paper', the contents of Web of Science as an 'overlay database'. But, for me at least, this is not where the real potential lies. Instead, imagine the overlay journal as a way in which academics entirely remove the need for publishers. The need for this is increasingly evident as we become more familiar with the ways in which we rely on traditional publishing models to pervade our scientific project with confirmation bias. Overlay journals no longer require a publisher to store the publication. This is done at the preprint server. The reviews are housed at the same arXiv site (or would be in an ideal and transparent version)[@rittman2020fast], as is the manuscript in its final form after being accepted by the overlay journal editor. The authors themselves are responsible for the final layout. The Overlay Journal co-ordinates the reviews and conducts the editorial work, and then simply acts as a pointer to the finished product: no papers, no publishers, no editorial management software, no costs and all papers are [Diamond OA](#diamond)!

The journal [_Discrete Analysis_](https://discreteanalysisjournal.com/) (indexed in both Web of Science and Scopus) was the first of these new 'arXiv overlay journals' (since 2017), and following this link will allow you to quickly appreciate what an Overlay Journal is. Each 'published' paper still sits on its original preprint server. The overlay journal itself offers a brief editorial summary of what you'll find if you click through to the paper. This is a fantastic idea in that it pitches editors back into being responsible content curators. As an editor I'd want to be motivated to publish a paper that I liked in order to write an editorial summary about it.

Because the only the accepted version is provided with an 'article number' and the style file of the journal layout, the author then produces the final version of record (VoR) of the accepted manuscript by running the style file with LaTeX. All of this is possible with free software, for example by using R Markdown [@xie2018r].

Using preprint servers also allows the entire process to be [transparent](#transparency2), very quickly becoming associated with other great initiatives like the Centre for Open Science - [OSF](https://www.cos.io/).

### What do traditional publishers think of 'Overlay Journals'?

Surely, the onset of 'Overlay Journals' should have publishers quaking in their boots? Strangley not. But their response should really be enough to wake us up.

> They're probably only going to succeed in disciplines where a no-frills approach to publishing is acceptable... 
> I think that the real threat to our traditional model... is if Overlay Journals have **Impact Factors** and can provide the same services, and they are free... then I think that that does pose a threat.
>
> ---[Anonymous from Highwire Press Inc. 2017](https://youtu.be/ADecAXLrKoA)

As this has already happened (for the journal _Discrete Analysis_), it would be interesting to know how traditional publishers are going to prevent an Overlay Journal take-over.

### What is happening in biological sciences? {#PCI-EvolBiol}

One of the original electronic journals, [Journal of Medical Internet Research (JMIR)](https://www.jmir.org/), announced in 2019 that it will launch an overlay journal covering biology, a so called 'superjournal', JMIRx [@eysenbach2019celebrating]. This overlay journal operates by editors choosing preprints that they want to publish ('editorial prospecting'), and then approaching authors and reviewers, and also by authors pitching their preprints to the editors. Today [JMIRx|Bio](https://bio.jmirx.org/) accepts any preprint published on [bioRxiv](#bioRxiv). Although JMRIx|Bio, and sister journal [JMRIx|Psy](https://psy.jmirx.org/), were launched in 2020, I cannot find any articles submitted (by mid-2021). The sister journal [JMIRx|Med](https://med.jmirx.org/), launched at the same time and in the same area as other JMIR journals, and has a rapidly expanding publication base.

Another preprint led service is _F1000Research_ (f1000research.com), which grew (via a buy out by Taylor & Francis) from the peer recommendation site Faculty Opinions (facultyopinions.com). _F1000Research_ requires preprint submission on their server, and then co-ordinates postprint peer review which is all accessible online in the same location. Like other [Gold OA](#gold) journals, _F1000Research_ charges an impressive Article Processing Charge, and does not aspire to an [Impact Factor](#noIF).

There's an excellent tie-in here with [transparency](#transparency). Because preprints are [Diamond OA](#diamond), and reviews are OA, the process is all transparent.

A nearly 'overlay' model is [Peer Community in Evolutionary Biology](https://evolbiol.peercommunityin.org/), launched in January 2017. This comes very close to the 'arXiv Overlay Journal' model described above. These preprints are submitted to PCI-Evol Biol, and are reviewed and (if they aren't rejected), a recommendation is given. The site then publishes the recommendation from peers as well as pointing to the preprint. However, unlike _Discrete Analysis_, the preprint remains 'unpublished' despite the peer review and can then be taken onto a traditional journal. There is a  growing [list of journals](https://evolbiol.peercommunityin.org/about/about) whose editors will accept recommendations from PCI-EvolBiol, and may use the reviews when appropriate. However, it's also worth noting that there are a small number of journals that will not accept preprints recommended by PCI-EvolBiol. While Peer Community in Evolutionary Biology does not publish their peer reviewed articles, another initiative from Peer Community In takes a step backwards to get a step closer ([see below](#commitment)).

The journal [eLife](https://elifesciences.org/) is also taking steps toward becoming an Overlay Journal, with their implementation of a **preprint only** submission route [@eisen2020implementing]. Although eLife appears to embrace all the advantages of transparency in their use of preprints, there is still a significant barrier that has recently jumped from $2500 to $3000 for the privilege to publish (eLife has an APC waiver system which is not seen by editors). Again the question of what exactly scholars are paying such high fees for comes to the forefront.

### Preregistration and a commitment to publish {#commitment}
Registering your proposal (or any research plan), means that you can present a historical document to a journal (probably four to five years later) to show that you have tested the hypotheses that you originally intended to. This is simply a way of being transparent in your science, and enables you to demonstrate that you have not been [p-hacking or HARKing](#transparency). Similarly, you can demonstrate that you are not '[salami-slicing](#salami)' your results.

Another initiative from Peer Community In is the possibility of submitting to their [Registered Reports](https://rr.peercommunityin.org/), which goes much further towards removing the confirmation bias. The Registered Report (RR) is in effect the registration of a proposal (i.e. preregistration) with review. If the RR is approved by reviewers then the study is, in principle, given the green light for publication whether or not the hypothesis posed is accepted or rejected. I say “in principle” because those same reviewers are shown the manuscript again once the results are in. They need to check that the methods proposed were followed, the analyses were conducted in the same way they were proposed, and that the conclusions are justified by the results. Peer Community In are offering to organise the two sets of peer review. In addition, there are a bunch of journals that have already signed up to accept RRs that are signed off after completion (notable among these is PeerJ). To me, this represents an important step in the right direction toward [transparency](#transparency2), and the elimination of confirmation bias. What would be great to see is the number of conventional journals sign up with RRs based on the quality of the study design and execution, and the concomitant abandonment of [Impact Factor](#impactfactor) as a driving force in publishing.

### Beware of publishers' preprint servers
Publishers have a lot to be worried about preprint servers as they have the potential to take away their business. For this reason, you will see that some publishers have launched their own preprint services. Don't be lulled into submitting to the preprint server of a publisher. There are plenty of preprint servers that have nothing to do with publishers, including the most ubiquitous. These are non-profit transparent organisations, and we all have an interest in them staying that way. Simply put, there's no need to use a publishers' preprint server (this is the same principle as your data, [see here](#suppmat)).

## Peeriodicals - another twist on the idea of an Overlay Journal

The launch in 2018 of [Peeriodicals](https://peeriodicals.com/) puts another twist on the idea of an Overlay Journal. This time, the Peeriodical is a site where anyone puts together a collection of any published papers or pre-prints that they curate themselves on a topic. They don't pass them out for any further review, but they are in effect another kind of curated Overlay.

## Should you submit your manuscript as a preprint? {#preprint}

Here are some reasons for and against submitting your manuscript as a preprint:

### For:

- Your manuscript needs to be cited by another that you are submitting and you are worried that the peer review process will take too long.
  + This is especially appropriate in the need to be [transparent](#transparency2) to demonstrate that you are not '[salami slicing](#salami)' your work.
- You are applying for a scholarship, a grant or job and want to be able to show that you have a body of work that is ready to be published, even if it is not formally published yet. Using preprints, you can allow the hiring committee or potential employer access to your work. This is really much more impressive than claiming manuscripts are 'in preparation' on your CV.
- You are presenting some unpublished work at a [conference](#conference) and you want people to be able to access it (e.g. through scanning a QR code)
- You are submitting a grant application and want to demonstrate that you have sufficient data, although it isn't yet published
- You are aware that another lab is working on a similar project and are worried that submitting to peer review will scoop your findings. 
- Your work has immediacy that it might not have after (potentially) 3 months of peer review. It may be that by releasing your preprint you can contribute to an ongoing debate that otherwise you’ll potentially miss.
- It's free. No APC or other fees are involved with depositing your preprint
- You have the potential to increase your network when people you have never met read your work.
- You are concerned that you’ve missed something important or perhaps analysed something in a novel way that others might be able to help with. You want this chance at feedback before submitting to peer review.
- Your manuscript crashed out of peer review with comments that you felt were unfair or unsubstantiated. You are looking for more balanced comments.
- In the above case, you might be able to use your preprint as leverage to persuade an editor that your contribution should be fast tracked into their journal.
- If you can generate enough buzz and positive feedback, you might be able to get leverage on an editor for submission to a journal with a higher impact factor.
- You have a working group that you actively want to share your publication with, even before it is published
- This is the only submission route for 'Overlay Journals'
  + [PCI-Evol Biol](##PCI-EvolBiol) in Biological Sciences
  + [JMIRx|Bio](https://bio.jmirx.org/) accepts any preprint published on [bioRxiv](#bioRxiv) since 2020
- This is the only submission route for some other journals [e.g. _eLife_ @eisen2020implementing]

### Against:

- Any of your co-authors don’t want the manuscript submitted as a preprint before peer review
- You feel that public access might mean that your results are misinterpreted
 + this should be on you to get it right before you submit it
- There is a real chance that others can use the access to your work and publish it before you
 + It's worth adding here that while you might believe that there are lots of people out there who might want to steal your work, this is a general paranoia that is very common in early career researchers. Few fields in biological sciences really have valid examples of data theft or idea theft. 
 - Preprints are another example of how everything is too rushed these days.
  + I've heard this opinion, but wonder why these authors wouldn't simply hold onto their manuscript until they feel that it's ready.

I can’t really come up with a lot of reasons against submitting a preprint (I've had to add some I have heard other people saying). This is possibly because I’m broadly in favour of preprints and see that there is value there. However, I’ve done it with only a fraction of papers submitted in the last 5 years. Why?

My experience of preprints, in terms of feedback and reviews, is disappointing. Although these get widely shared on social media, and garner a large number of downloads, they don’t generate comments from colleagues. Even when we have sent links of preprints to colleagues asking directly for feedback, we’ve received little to nothing. This does not mean that preprints are worthless. I think that they have great potential, and they may work better for you in your field than for me in mine. Moreover, preprint servers now hold the potential to free academics from the tyranny of [profit hungry publishing houses](#overlay).

At this point, I should say that I have not (yet) made any public comments on a preprint. When I have looked at preprints, I (generally) have downloaded them in order to look at some of the details (often the methods or analyses), when there’s a dearth of peer reviewed (published) material. There are a few references to preprints in this book. I’ll replace them if I find that they have been published. But what should I do if the published version doesn’t contain the point that I’m citing on? In this case, I’ll delete the citation and no longer make the claim because there is the chance that the result did not stand up to the rigours of peer review.

### 2021-09-05 1.10-review.Rmd
Add files via upload
# all round update

# Writing a literature review {#review}

The first chapter of many PhDs is often a place for a literature review. Writing a review of the literature is actually a very useful task to undertake during your PhD. Firstly, it will get you familiar with all of the relevant literature in your field. It will give you some idea of the historical context of your subject area. And it's especially good to help you spot gaps in the literature that may not have been realised before, perhaps giving you ideas about pitching grant proposals. As an Early Career Researcher, if you haven't undertaken a review in your subject area, then you should seriously consider it.

## Having a purpose
Writing a literature review is easiest when you have a clear purpose. The objective can be tight, or can aim to use the literature in order to answer a broader question. Either way, having a clear purpose for your literature review will help you to define exactly what literature you're looking for.

Without an aim, a literature review can be extremely daunting. There will be literally no end to the literature that you could include and it may well overwhelm you. If your question is more broad, it will help to define a clear time period in which you are searching for literature. For example, you may want to just consider the last 10 years. Remember that the amount of literature is increasing almost exponentially and so if you are covering the same ground as a previous review a lot can happen in 10 years.

If your aim is quite narrow then you can legitimately cover all of the literature. This may require delving back more than 100 years. But as we know the first journal only began around ~150 years ago.

If you want your review to have an objective or statistical angle on the literature then consider writing a systematic review or a meta-analysis.

## Systematic reviews
A systematic review is when you define your search criteria and the databases that you want to use and then consider all of the literature that comes out. You will need to provide clear criteria for the inclusion of literature. As well as explaining exactly how you decide to include or exclude papers that result from your search terms. Systematic reviews require a flow diagram to show how this literature was included or excluded [@moher2009preferred see Figure \@ref(fig:Prisma-flow)]. Many journals (especially those linked to PubMed) now insist that a Prisma flow-diagram has to be 'Figure 1' of any systematic review submitted. Simply replace the text in Figure \@ref(fig:Prisma-flow) with relevant numbers as appropriate in your systematic literature review. You can also add boxes as necessary. Note that the need to produce the Prisma flow-diagram means that you should keep good records of the results (and date) each time you conduct a literature search. You should also retain the actual search term used together with a file of the results.

(ref:Prisma-flow) **The Prisma flow diagram after Moher et al (2009)**. Text inside the boxes is replaced by the author with the corresponding numbers of records, articles and studies as they were filtered during their systematic literature review, and (if appropriate) the final meta-analysis.

```{r Prisma-flow, echo=FALSE, fig.cap="(ref:Prisma-flow)"}
 knitr::include_graphics("figures/Prisma-flow.png")
```

In my opinion, a systematic review is less subjective than a review where the author is essentially cherry picking the literature in order to tell the story as they see it. It also allows you to conduct statistical tests in order to answer some of your questions: i.e. a meta-analysis. Lastly, it avoids one of the oldest problems in science: confirmation bias. Non-systematic reviews tend to simply look for articles that back up the author's own view. Too many old reviews are really just written by academics seeking to confirm their standpoints, and so moving forwards we must place the emphasis on being objective through systematic reviews. This view is now growing so that increasing numbers of journals will only accept a review if it is systematic.

## Using Google Scholar
Google scholar has distinct advantages over other literature databases as it searches words inside the contents of an article, not just in the title, abstract and keywords (like Scopus and WoS). However, the output style of Google Scholar is infuriating as it resists being able to capture search results into a simple table (database) format. In addition, it does not support Boolean search operators, and includes a far broader range of literature, and linking citations [@martin-martin2021google].

Help is at hand for those of you who would like to use Google Scholar and get a database output with the publication of [Publish or Perish software](https://harzing.com/resources/publish-or-perish) [@harzing2007publish]. The original purpose of this software was to calculate personal research metrics for promotion purposes, and it still does all this. But it also provides a single platform for searching many different literature databases: Scopus, Web of Science, Microsoft Academic (now discontinued), Google Scholar.

Google Scholar has an important attribute over many other databases in that it returns results in languages other than English. Using only English to provide the basis for your review might lead you to draw an unreasonable conclusion with a literature bias [@nunez2021monolingual]. Language is only one of several potential biases in a literature review [@amano2013four].

## The meta-analysis
A meta-analysis is a type of systematic review that uses the results of all of the studies included in order to provide a synthesis. While more powerful than a systematic review alone they take an awful lot of work. If you are expecting to get a lot of literature then consider putting a team together in order to conduct a meta-analysis.

However, conducting a meta-analysis on a small number or articles (<20) could lead to spurious results. In other words, they might not be a good solution for a very small subject area. Unless your questions have had sufficient numbers of investigations, you would be better to assess it qualitatively.

### Combining topics by combining people
Reviews become especially insightful when you combine topics together. To do this you may want to find somebody else who is an expert on the literature in that field. This might be somebody in your lab group, or it might be someone in your wider [network](#networks). Bringing together a set of like-minded people for a review is a very powerful way in which to create an important and potentially powerful [network of collaborators](#networks).

This brings me to another very important point. Share your enthusiasm for your topic. Remember to talk to people around you, both in your laboratory and in your department about the work that you're doing, and tell them about the interesting things that you're finding. If they also tell you about their work, then you may find that you have somebody to combine your literature review with in a novel angle that's never been thought of before. The paper that results could be a smash hit.

## Don't try to do too much
Whatever your aims were before you started compiling the literature, remember to remain flexible as you proceed. It's hard to know exactly how many papers you will encounter on a particular subject until you've conducted the search. If you're prepared to remain flexible while conducting your review it may save you from overreaching. Always be open to reducing the scope of your review or spotting new questions or issues while you're reading that may be better than your initial idea.

## In summary
If you have got to this stage of your career and have not yet done a literature review then now is the time. You should already know a large chunk of the literature in your particular area and you will be looking for future questions to tackle in your early career. Conducting a literature review at this stage  will not only help you to highlight unanswered questions and interesting topics, but it will get your name out there. Once you've done it you'll be glad you did.

### 2021-09-05 2.01-formatting.Rmd
Add files via upload
# all round update

# Formatting your submission  {#formatting}

Let’s face it, after spending so much time writing your manuscript, it's worth making sure that you don't make silly mistakes in finalising the manuscript.

## Must do check-list before submitting your manuscript:
I am very fond of check lists, in part because their utility has already been proven many times [@haynes2009surgical]. I have developed this check list of things to go through before submitting a manuscript. The journal that you submit to is likely to have their own, so you should add that onto the end of this one. Plus, in your subdiscipline there are likely to be some extra points that you can add, so I suggest that you add to this list when making your own.

- **Spell check** - yes, it sounds obvious but doing a final spell check is a good idea. Not only this, but take the time to have your word processor ignore or add all of the special words (e.g. species or site names) that it doesn’t otherwise recognise. This will ensure their consistency throughout (within and between chapters)
  - **Language settings** - Make sure that your language settings are set to ‘UK English’ (or the English setting for where you are based). Remember that the journal that you submit to may well be in a different region to your own (so you might need to change your spelling settings) 
  - Look out especially for words that have different accepted spellings like those ending in -ise or -ize. Decide which you want and be consistent. Consistency is king!
  - **Capitalisation** - of place names and common names, place names and not adjectives. For example, ‘South Africa’ has two capitals, but ‘southern Africa’ only has one
  - **Species names** - will useually come up as spelling errors in a word processor,  but a good idea to add them to your dictionary so that you can check they are uniform throughout the manuscript
  - **Grammar check** - always good to take a final look, especially for chapters that you wrote some time ago
    + Use the word processor automated options to help you
  - **Listen to the text** - Have your computer or another device read the text so that you can hear anything obviously wrong
  - **Pronouns** - Pay attention if you have used “we” or “I” and make sure they are consistent in your manuscript. You are likely to have used “I” in your thesis, in which case you will need to change to “we” if there are multiple authors
- **Page layout** - Really very important to get this right in your template. Make sure that your template has:
  + Correct paper size (A4 and not US letter - or visa versa!)
  + Margins
  + Line spacing
  + Page numbers
  + Line numbers (really helps your reviewers - add these even if the journal adds their own)
- **Headers and Footers** - If you can manage a chapter specific header, it’s useful to show your name and a short chapter title
- **Sections and subheadings** - Check with the journal instructions to authors about what sections they allow and where
- **Figures** - many journals charge for printing colour figures. This does not mean that you have to reformat your figures to be in monochrome. You will have an option of submitting colour figures for online use only (in pdf and online), while the limited print run will have them in monochrome. You should make sure that your figure is still understandable when printed in monochrome
- **Acknowledgements** - this is your time to say thank you to all the special people that have helped during your study. There are probably more than you realise, but in addition to your friends and family (who most people don’t forget), think about the people who administered the work, lab mates (past and present) who were always there to help, and people who gave permission at study sites
- **References** - probably one of the most dreaded sections of any manuscript preparation, but they do have to be done. If you’re one of these people that has everything in a database, then you’ll be laughing or cursing your database throughout. While it might be tempting to only look through the data within the database, spend some time to see how it’s displayed in the manuscript. A mistake in the reference database will be multiplied many times in the manuscript. Remember that examiners love to take a random look through the reference section to make sure that it’s all good. After years of painfully entering references themselves, they know just what to look for
- **Journal checklist** - Now go to the website of the journal where you want to submit your manuscript and check to see their submission guidelines and whether or not they have a checklist

## Mistakes people make:
Other than the obvious things, all mentioned above, here are some of the mistakes I’ve seen.

- Submitting the wrong version (yes this does happen!). Probably worse if having a mixture of right and wrong versions
- Last minute additions to text with incorrect spelling and or grammar
- Two correctly spelled synonyms sitting next to each other when only one is desired (probably came about when editing)
- Forgetting to check for plagiarism
- Comments and or edited text (especially when it’s marked as being by someone other than you)
- Page numbers that start again and again at different sections
- Lots of blank pages or spaces (avoid blank pages if you can, and try to limit the amount of blank space (never >half a page)
- Leaving important people out of the acknowledgements, especially your funders!

### 2021-09-05 2.02-letter.Rmd
Add files via upload
# all round update

# Writing your cover letter {#letter}

Many journals require that you write a cover letter to the editor on submission. Each journal should tell you what this letter should contain. This will vary from journal to journal as some may have a set of check boxes that cover some points in their editorial management software, while others may want a written declaration. It is very important that your letter contains all of the information required by the journal. If not, it could lead to an automated [desk rejection](#desk).

Some journals have a text box in their submission software for you to paste the text of this letter. If there is a text box, then I suggest that you do not compose it on the fly, but plan it out and write it in a word processor so that you can catch any silly spelling or grammatical errors. Others journals might expect you to upload a pdf letter on headed paper from your institution. Make sure you know what is required and that you are prepared. The reason for a formal signed letter on headed paper is that in some submissions, this letter is used as a legal declaration.

Your cover letter is a professional, formal letter, and should be written as such. Use the conventional letter writing format, including addresses, date, signature, etc. If you are unsure what this looks like, there is some great advice on the web, including templates to use (see [here](https://wordvice.com/journal-submission-cover-letter/) and [here](https://www.springer.com/gp/authors-editors/authorandreviewertutorials/submitting-to-a-journal-and-peer-review/cover-letters/10285574)).

Address your letter to the journal editor. One of the nice things about academic titles is that they are gender neutral, so use them: **Dear Dr. Jones**, or **Dear Prof. Smith**. Even if this person has an office down the hall and you see them every day, keep the cover letter professional. It will go on file with the publisher.

The following points should be covered (where relevant) in the order given as a default. Any content and style requests from the journal, about how the cover letter should be written, clearly takes precedent.

1. The title of your manuscript, the type of submission (review, research article, letter, etc.) and the name of the journal that you are submitting to. Note that it is all too easy to forget to change the journal name when you are submitting to another journal. In my time as editor, I've seen some very nice letters explaining how the manuscript is appropriate to a completely different journal. Not a good start!

1. Statement that your manuscript has not been published or submitted elsewhere, and that all authors have approved the submission. There are other statements that are required by certain journals, and sometimes the journal requests that you copy and paste their text as the cover letter is kept and used as a legal declaration. Also it may include information about ethics clearance and any permits required having been obtained and available should the paper be accepted. Note that you don't need to include all of this information unless the journal that you are submitting to requires it.

1. Include information about why your research is suitable to the [scope](#scope) the journal that you have submitted it to. This requires you to have checked the scope of the journal itself and have thought about exactly how your paper aligns to this. Bear in mind that this is a real problem. Editors get far too many papers that do not fit the scope of their journal, and it takes time to process and reject these (so they are not appreciated).

1. Novel and innovative research. For journals where the [impact factor](#impactfactor) is important, you may want to emphasise what is novel about your study, and therefore deserving of the impact.

1. Important information that the editor must know. For example, if the manuscript was previously rejected with an invitation to resubmit (then also include the manuscript number that it was given previously). Or (rarely) if you (or others) have previously retracted a similar study, with reasons why this manuscript is not affected.

1. Connections to other ongoing research. If relevant, state what other manuscripts are already submitted to this or other journals. For example, is your work part of an ongoing consortium or research programme?  The idea here is to demonstrate to the editor that your manuscript [will be cited](#citations) in a timely manor.

1. A final brief statement to the effect that: **"We declare that we have no conflicts of interest."**

## Do editors read cover letters?

This is a moot point in that you will never know. Some editors definitely read every cover letter [@kenar2016dear]. Even new journals have retained the need for a cover letter when they have stripped down the submission process to a minimum (e.g. _eLife_), and so my suggestion is that it is needed and, hopefully, it will be read. However, every editor at every journal is likely to have their own method of what they regard as important to read [@moustafa2015does], with many actively trying to [minimise this](#editors). Where there are legal requirements, it's not likely that editors will read them, but editorial managers may well make sure that these are present in order not to get a pre-review ([desk](#desk)) rejection.

Chief editors may read parts of the cover letter along with the abstract before deciding which Associate Editor should be assigned. Some chief editors will even leave this to the Associate Editor. The Associate Editor should read your cover letter in full, together with the metadata and other salient information therein. They should also [read the entire manuscript](#editors).

## What not to do in your cover letter

- Try not to make your cover letter too long. A single page should be enough to cover all of the points above.
- Don't copy and paste your abstract. There will be a place for this in the metadata on the submissions site. 
- Avoid specialist terms that the editor may not know.
- Don't try and oversell your study, or make excessive claims. 
- Avoid "first ever" claims, as they won't impress the editor.
- Don't suggest potential reviewers unless the journal rubric specifically requests this in the cover letter.
- Similarly, don't suggest people that you don't want to review your article, or start any history of why your submission is complicated by third parties. 
- Don't deviate from a formal letter style.
- Don't have spelling and grammatical errors.
- Avoid formatting the letter in a way that might make it difficult to read: 
  + Don't be tempted to reduce font size to get more in (keep to Ariel 11 point)
  + If you need more space to keep your letter to 1 page, change the margin sizes
  + Don't fully justify text (left justify only)
  + If you are recycling your letter, check that you have changed **everything**.

### 2021-09-05 2.03-suggestreviewers.Rmd
Add files via upload
# all round update

# Suggesting reviewers {#suggestreviewers}

You will often be asked to suggest reviewers as you submit your manuscript. This can be part of your [cover letter](#letter), or data that you need to fill in during the submission process. Keep in mind that supplying a name alone will not be considered sufficient, you will need their name, the name of their institution and their email address. Some journals will not accept addresses that do not come from institutions unless they can be verified. The reason for this is that there has been an outbreak of fraudulent reviewers, where authors suggested fake reviewers or gave email addresses that they manufactured [see @brainard2018what].

In keeping with the [spirit of peer review](#spirit), you should suggest people that you think will provide insightful reviews. These could be people that work in the same area (but that aren't involved in your study). Personally, I would be likely to name the same set of people that I would invite myself if I were editing. People that I believe would provide a constructive and unbiased review.

## Who should you not suggest?

 - Anyone who is an author or in the acknowledgements
 - Anyone else you feel may have a conflict of interest 
   + connected with the work but not an author
   + someone who was on your review committee
 - Someone in your department or even at your institution (there may be exceptions here, but often journals will not consider people with matching institutional emails)
 - friends, labmates or even relatives (even if you genuinely think that they would do a good job)
 - people who are regular co-authors (grant panels specifically ban you from naming these people as potential reviewers. If you have a large network, this can be problematic)

## Who should you oppose?

In addition to allowing you to suggest reviewers, many journals allow you to oppose reviewers. I usually leave this blank. If you have a particular lab or persons who you know will not follow the [spirit of peer review](#spirit), you could enter them here.

## Who will the editor use?
It is worth bearing in mind that whoever you suggest, the editor may not use them. I have talked to editors who will never consider anyone recommended by authors as a matter of course, because they assume that these people will be positively biased towards the authors. This same editor said that they'd always use at least one of the opposed reviewers. This rarely fits with journal policy.

As an editor, I will often use one of the reviewers suggested by the authors if they fit my (following) criteria. I will make sure that I balance this with another reviewer or two not suggested by the authors. To select reviewers, I will [read the submission](#editors) and look for citations of similar studies, or techniques (depending on the type of paper). From the citation, I will find the paper (preferably published in the last 3 years) and use the corresponding author's address. I will also try to visit the website of the senior (last) author to see whether they have a lab that is still active in this area, and especially I am looking for post-docs or Early Career Scientists who cover the same topic. If they are there, I will invite them. If not, I will write to the lab head in the hope that they may well recommend one of their post-docs. This process can take a long time, especially [hunting down email addresses](#website) that constantly bounce back from the journal's editorial manager software.

I will also use people that I know in my own network, especially an extended network. People whose conference talks I've seen or other papers in my area that I have read or cited myself. I try not to bombard my own close network too much with demands for reviews. I try not to use reviewers auto-suggested by editorial management software. In my experience, these are not suitable people. Whether this is the same for other editors, I cannot say.

If the manuscript is a resubmission, I will try to use the same reviewers that made previous reviews. This isn't always possible, and I know that it is a source of upset for authors when they receive new reviews on a second or even third round of review. I don't think that any editors will do this deliberately, but reviewers do have the option of indicating that they are not willing to reread a resubmission. If this is the case for most or all of the reviewers, then you are likely to face an entirely fresh round of peer review.

### 2021-09-05 2.04-whichjournal.Rmd
Add files via upload
# all round update

# Choosing the right journal {#whichjournal}

In this section I provide steps to go from a potentially very long list of journals where you might publish, to choosing the one where you will submit your manuscript first. As you work through the steps, keep notes on why you exclude journals at each step. You may want to revisit these criteria later, or explain to your coauthors why a particular journal was not considered.

Although your article can be found from anyone outside of the normal journal readership, it is important to publish your paper in the correct journal for your study for the following reasons:

- The editor will [desk reject](#desk) your manuscript if it is not appropriate (and you'll waste their time)
- Many reviewers give feedback tailored to the journal, and this will not be appropriate in the wrong journal
- Perhaps most importantly, you are likely to get additional readers in the correct journal that might not find your paper otherwise
- Note that there are no-Impact Factor journals for which this above is not relevant, and these are [discussed separately](#noif)

This section is geared toward the 'current' model in the life sciences of journal title as a measure of quality. The reality is that this model is shifting, and we may well end up with another (better) [open model](#closedtoopen) in the future based around [preprint servers](#preprints2) and [overlay journals](#overlay). The current reliance on journal titles is driving a [toxic culture](#PruittData) in biological sciences that is becoming increasingly recognised, even by those who are [gatekeepers](#gatekeepers) for this model (see [Part IV](#retract)). To reflect this viewpoint, and help drive a culture of change, I have placed [Impact Factor](#impactfactor) outside of this list of steps. In doing so, I acknowledge that you and your co-authors may well be operating inside the current publisher driven model advocating for Impact Factor as a step with high priority. Even if this is your reality, I hope that you will acknowledge and appreciate that this is not the way that journals should be chosen.

I'm going to suggest that you keep a spreadsheet with the answers to the steps below as columns, and different journals as rows. If your submission is rejected from the first journal choice, you can then use this spreadsheet so that you know where to submit next (i.e. just do the work once). This spreadsheet may well come in handy for future similar submissions.

### Indexing

You want other people to be able to find the work that you publish, and so selecting a journal that is already indexed in one of the major literature databases (Web of Science or Scopus) is important. Be aware that when new journals are indexed, they are usually done with their entire back catalogue. Thus, if a journal advertises that it will be listed by one of major literature databases, your submission will likely be listed eventually, even if not immediately.

### The Subject Area

No matter what your paper is about it will fall within an existing subject area. A good way of determining your subject area is to look at those listed by literature database like Web of Science or Scopus. These databases have subject areas which contain groups of journals. You can look through the journals that you cite in your references, then check with Web of Science or Scopus to see which subject area the majority of them are grouped into.

Journals have different hierarchies of scope. Some journals attempt to take on the full gambit of science (e.g. Science and Nature) while others are only interested in a particular taxonomic group. In general, the impact factor of the journal is likely to be linked to the diversity of the scope. This is not always the case. There are some taxonomically specific journals with high impact factors, and there are some general journals with low impact factors.

Once you have determined your subject area make a list of the journals within this area. Based on your reading of literature pertinent to your manuscript, try to decide whether your manuscript is likely to be accepted by a journal with a high ranking impact factor, a medium ranking journal, or a low-ranking journal. Note that in your discipline it may be more appropriate to use a five or ten year impact factor to rank your journals, instead of the normal two. Sort the list of journals within the subject area by impact factor and then select the ones that are either high (>10), medium (>4), or low (<4).

### The Journal Scope {#scope}

Every journal has a scope that is stated on their website. Your manuscript must fit into the scope of the journal that you select. In order to get your candidate list of journals you now need to visit their websites to look at their scope in detail. Ill fitting scope is the most common cause of a [desk rejection](#desk) [@schimel2014editors; @teixeiradasilva2018establishing].

If your potential journal list is long, I suggest that you lead with the journals whose articles you have already cited in your manuscript. In the journals that you enter into your spreadsheet, quickly summarise the scopes of your choice journals and make some notes on how your manuscript fits them. This will help when writing your [cover letter](#letter).

### Ownership of the content {#copyright}

In this book, I have made a plea for the movement to an [Open publishing model](#openscience). Note that your choice of journal will dictate who ends up owning the content: i.e. copyright. Most for-profit publishers insist on owning the content of the journals that they publish. This means that if you ever want to use the content that you created (such as a figure in a book), then you will need to approach those publishers for permission. Traditionally, for-profit publishers have retained copyright on their content as this makes for another potential [income stream](#money) for the future (think text books that use graphs from journal papers). In the case of society owned journals, the copyright can be retained by the society. Most Open Access publishers have been forced into a Rights Retention Strategy by [cOAlition S](https://www.coalition-s.org/the-rrs-and-publisher-equivocation-an-open-letter-to-researchers/), and consequently, this should be a general trend in all journals: copyright is retained by the authors with a Creative Commons licence. If you want to know more about what the different Creative Commons licences mean, they have a very good explanation on their website: https://creativecommons.org/

Although I've listed ownership of content as Step 4 in this scheme, I would suggest to you that it is not incompatible with any journal. As owner of your content, you have the right to archiving your [Green OA](#green) wherever you please. Wherever you publish, make sure that you retain ownership of your content, and if this is not the case, then insist that you want to retain the right to your own content once your article is accepted. Think of it from the other side of the desk: if the author wrote to you as editor and said that they want to retain ownership of their content - on what grounds do you (as editor) have to refuse them? Owning your own content seems like a trivial step in the massive movement from [closed to open science](#openscience), but it is a very significant step forwards that all scientists should own their content.

### Current Contents

Now that you have a shorter list, it's time for you to look at the current contents of the journals that are on their website. The contents for the last two years should reflect the policy of the current editor for accepting manuscripts. You should be looking for papers that look similar to your manuscript in their scope.

If you see papers that look directly comparable to your own then make a note of what they are. Your list of journals should now be less than ten.

### Society journals {#society}

Academic publishing started with society journals, and I think that they are still worth supporting if you can. Your mentor or co-authors may be members of particular academic societies, and may have a preference therefore to publish in their society's journal. See below for other potential advantages in publishing in a society journal.

### Transparent credibility

[Transparency](#transparency2) in science is very important and should be part and parcel of your own work. When you have made a real effort to be transparent, you should look for journals that do the same. There is a badge system used for transparency in science. See or @kidwell2016badges, or see @marshall2021make for an example of how this can be applied.

### Knowing the journal from the inside
Your mentor or co-authors could be an Editor or Associate Editor (past or present) of one of your target journals. Or there may be someone in your department or institute that you could consult. I am not suggesting that you use their influence to help you get published, this conflict of interest strictly prohibited by most journals. Instead, these people can help you decide whether or not your submission will be welcomed or [desk rejected](#desk).

Reject without review is typical for manuscripts whose authors have not followed steps 2 to 4. It can still happen, even if you have. Reject without review is such a waste of everybody’s time that you should avoid it if at all possible.

### Is there a specail issue?
Special issues are a great way for your work to get better exposure, both to others that are participating in the special issue, and to those who come across the entire issue later (see [Part I](#citations)).

### Avoiding the predators
You must make sure that the journal you plan to submit to is not a predatory journal. Follow the check list in [Part IV](#predatory).

### Financial considerations to publishing

In an ideal world, there wouldn’t be any more barriers to you publishing your contribution to the collective of scientific knowledge. Not only is it not an ideal world, but I would argue that there has never been a less ideal time for publishing science. Greedy publishers have taken publicly funded science and made the public pay for it time and again. Read more on this [later in the book](#paywall).

### Type of peer review
Another factor that may influence your choice of where to publish is the type of reviewing done by each journal. These [reviewer flavours](#flavours) are discussed in detail elsewhere. See a full description of these types of peer review in [part 1](#flavours).

### Time to first decision
Sometimes you will be in a hurry to publish, and journals have different mean times to first decision. In general, this is driven by the responsiveness of the editorial team. Some journals, with fast turnaround times, advertise their time to first decision and so this may influence your decision about whether or not to publish there. Remember that if you are desperate to get a manuscript out into the public domain, you can always submit a [preprint](#preprint).

### Open access

Some journals are exclusively [Open Access (OA)](#openaccess1) meaning that you will need to pay (unless they are [Diamond OA](#diamond)) in order to publish your accepted manuscript (author pays). Different types of OA are covered in a subsequent chapter (see [Part II](#openaccess1)). More information on OA is provided in [Part IV](#openaccess), and I encourage you to read this before you decide to pay any research money toward OA publishing Article Processing Charges (APCs).

Your university may have a deal with some OA publishers (especially if you are based at a European institution) so it's well worth making a note of this. If in doubt, talk to your librarian. However, watch out for [manipulative publisher deals](#deals) that make you more likely to publish OA when your institution has a “read-and-publish” deal. Remember to ask yourself your motivation for why you are choosing your journal. Morally, choice by publishing company should never be high up on anyone’s list.

### Page charges

Note that some journals that remain behind paywalls still demand page charges. These can be quite substantial if you come from a lab with no money for publishing charges. In my experience American society journals regularly have page charges. These may well be reduced for members, or you may be eligible for a page charge fee waiver.

### Fee Waiver

Note that with all of the above you may be eligible for a waiver to page charges or Open Access fees. Who gets a waiver will be discussed [later in the book](#waiver).

### Impact factor

The impact factor of the journal may be an important motivation in your choice, or that of your advisor. I’ve left it off my list of stepwise criteria as I hope that it’s not going to influence you according to the San Francisco Declaration on Research Assessment (known as [DORA](#DORA)). If [Impact Factor](#impactfactor) is important to you, include it in the list that you produce and make a note of the most recent impact factor for each journal. There's more about the impact factor in another [chapter](#impactfactor).

### No impact factor journal #NOif

In this book I refer to journals that have no interest in their [Impact Factor](#impactfactor) as 'no impact factor'. Of course, if these journals are listed by impact factor providing databases (and most are), then they will have an impact factor and this can be high - even higher than other journals that attempt to inflate their impact factors (see [Part IV](#PLOS)). Instead, what I mean is that these journals are operating at minimum standards of technical soundness (so called '[negative selection](#AKP)'), instead of the AKP selection of only papers that are thought by their gatekeepers to uphold or increase the journal IF.

There's no problem to publish in a journal that has no interest in impact factor. Indeed, it can be seen as 'the right thing to do', with respect to confirmation bias. This is the way that all science was published prior to the dominance of [citation metrics](#citations). If you are submitting to a 'no impact factor journal' then you should be aware that this is what their speciality is: an emphasis on the technical soundness, rather than whether or not a significant result was found. You need to be aware as this is how inclusion of such publications in your CV will be regarded by peers. The list of 'no impact factor journals' is growing, and they tend to be online only, [Gold OA](#gold) and carry substantial [article processing charges](#gold).

## Shortlist
Once you have your shortlist of journals to consider, take it to your mentor. Together with your coauthors rework your list into something that you both all agree with. Rank your list by journals that you want to try first and those that are your last options at the end.

Keep the list so that if you are rejected by the first journal on your list you know where you're going to next.

### 2021-09-05 2.05-openaccess.Rmd
Add files via upload
# all round update

# Open Access or a paywall for your manuscript? {#openaccess1}

There are many well established benefits to publishing an Open Access (OA) paper where there is no pay wall to any readers. These include, increased citations, increased exposure and coverage by the media, not to mention increased interactions with the public and the moral and ethical duty to be able to share research as widely as possible. Many studies are now suggesting that the advantages go even deeper [see a collection of studies by @tennant2017open]. However the current reality is that most publishers will then require you to pay in order to produce your publication open access. There are chapters later on in this book that discuss the importance of [open access](#openaccess). In this chapter, I review the different OA models available. At this period in time, the names are somewhat fluid, and you may not find the specific term mentioned here on the publishers’ website.

(ref:Paywall-graphic) **The world of Open Access (OA) is built on [shaky foundations](#openaccess).** Much more of journal content is still within the shadow of the paywall, even for journals that practice Hybrid and [Bronze](#bronze) OA. When you are inside your institution, you may not notice the paywall due to seamless integration with many literature databases. You will be far better acquainted with the paywall if you are from a smaller institution or trying to access literature from outside your institution. Once you are the wrong side of the paywall, your options for most content will be Grey and Black OA. There are regular assaults on these two initiatives by publishers.

```{r Paywall-graphic, echo=FALSE, out.width = '95%', fig.cap="(ref:Paywall-graphic)"}
 knitr::include_graphics("figures/Paywall-graphic.png")
```

## Closed Access: i.e. The Paywall {#closed}
This refers to the need for your institution or you personally to be subscribed to the journal in order to access the content. This is the traditional model in academic publishing. If you are a member of an academic society then you may get access to their society journals through your membership. This is still effectively a paywall that is maintained by the publisher and the society together (Figure \@ref(fig:Paywall-graphic)).

It is sometimes hard to know if there is a paywall if your university subscribes to the publisher or the journal that you are interested in. There has been a lot of headway made in having seamless integration and access to articles behind paywalls from within university IP addresses.

If you try working from home then you will quickly find out which journals exist behind paywalls. There is another chapter with ideas on how to [get around the paywall](#paywall) if you need to.

Although publishing an article behind a paywall is often frowned upon these days it usually means that there'll be no cost for you as the author, and so for many academics this is still the only real option in terms of publishing their scholarly work in an academic journal.

There follows a brief description of each of the OA models. See Piwowar et al. [-@piwowar2018state] for a historical review of these different models:

##Open Access 
There is a bewildering array of different Open Access types. Entries in the list below are not mutually exclusive. Moreover, you may not be aware how your manuscript may be treated, and it is not always possible to predict the level of OA your article will be published with in many journals. Each journal will likely have their own policy, tied to the publisher in many cases.

### Gratis OA {#gratis}
Gratis OA simply refers to the ability for anyone to access the text free, and without any [paywall](#paywall).

### Libre OA {#libre}
Some definitions of OA extend beyond simply being allowed to access the text for free ([Gratis OA](#gratis)) to it bearing a license so that you can share, redistribute and reuse (Libre OA). With Libre OA everyone is able to share, redistribute or copy the content. The Creative Commons (CC-BY 4.0) licence allows others to “remix, transform, and build upon the material for any purpose, even commercially”.

A good example of this is my ability to publish the following Figure \@ref(fig:Piwowar) from Piwowar et al. [-@piwowar2018state] as the content of _PeerJ_ is published under CC-BY licence.

(ref:Piwowar) **Reproduction of material is possible if it is published under CC-BY**. Figure 2 from Piwowar et al. [-@piwowar2018state] shows the increasing number (A) and proportion (B) of a random sample of 100 000 papers published with a CrossRef DOI with an increasing trend in Gold and Hybrid OA publications since 2005.

```{r Piwowar, echo=FALSE,  out.width = '95%', fig.cap="(ref:Piwowar)"}
 knitr::include_graphics("figures/Piwowar.png")
```

### Green OA {#green}
This is an option no matter what journal you publish in, including [closed journals](#closed) behind the paywall. The accepted manuscript, before it has been typeset, is deposited in your institutional repository; usually your library hosts this. This is sometimes referred to as a **postprint** (manuscript with changes made following peer review) - the same version that has been accepted by the journal. Once your paper is published by the journal anyone can access the manuscript that was accepted because it is available free from your institutional repository. However the typeset paper will remain behind the publishers’ paywall (i.e. [closed](#closed)). Note that this version will not be the [Version of Record](#VoR), it may be that some of the text will change with [copy editing](#proofs). However, for people who want Open Access your work, this Green OA version should suffice.

When you deposit your author version to your institutional repository, make sure that you add a [Creative Commons Attribution](https://creativecommons.org/) license (CC BY). In theory, this turns your work into something that is [Libre](#libre) Open Access. Although the wording will be the same in the [Closed OA](#closed) version eventually available from the publisher, the prior attribution of CC BY to your Green OA work should mean that all subsequent versions are [Libre](#libre) OA. There are a number of reasons why this may not be the case, and you should be aware of two of these [@khoo2021why]. First are your obligations to your funders. For example, your funders may have specific requirements about how you publish, and prior rights attribution may conflict. Funders may have other restrictions on your publishing including embargo periods. With many publishers you may be required to sign an acceptance letter that specifically states that it's supersedes any previous licensing commitments including self archiving copies that you make CC BY. Your previous CC BY commitments to your pre-print may then place you in a legal conflict. Publishers are unlikely to assert their license unless you try to assert yours.

If you are unsure about whether or not to deposit a copy in your institutional repository and what license should be attributed to this, then ask your librarian. You may need to furnish them with details of your funding and potential target journals where you'd like to publish.

### Bronze OA {#bronze}
Bronze Open Access is an option at the discretion of the publishers. Only some publishers do this for some of their content. For example, they may decide that a certain thematic issue should be open access, an editorial, or review article. There is some debate about whether this bronze model is truly OA as it often still carries copyright restrictions.

### Delayed OA {#delayed}
Some journals choose to have their archived content available for all readers without a paywall. This could be after two or five years. Like [Bronze OA](#bronze), there is some debate about whether delayed OA is truly open as it often still carries the publishers’ copyright restrictions.

### Gold OA {#gold}
In a journal with gold open access it is compulsory for you to pay the open access fee (or Article Processing Charge, often referred to as APC) in order to publish your paper. These APCs can be very large, often more than \$1000 (see [Part IV](#money)). If your choice of journal is gold open access then make sure you know where the fees are coming from. The history of Gold OA is an interesting one that you can read in [Part IV](#PLOS).

Examples of gold open access journals are _PLOS ONE_, _F1000Research_ and _PeerJ_, and an increasing hoard of [no impact factor journals](#noIF). The advantage to these journals is that as soon as you publish your work in them, everyone will have access to it without any paywall. The disadvantage is that you may not be able to afford to publish there. The disadvantages of these journals are now becoming very prominent in the Biological Sciences, such that you may be excluded from increasing numbers of journals in your field. Alarmingly, journals with higher impact factors are also charging ever increasing sums to publish Gold OA [@gray2020sorry]. Gray [-@gray2020sorry] makes the important point that prestige (often confused with [Impact Factor](#impactfactor)) is being allocated a higher price in Gold OA, that is likely to disadvantage and disenfranchise scientists from less wealthy institutions and countries. This is likely to reinforce an increasing dichotomy between rich and poor researchers. We will take another look at who pays for OA [later in the book](#openaccess).

Watch out for predatory journals among those journals with Gold OA. Predatory publication will be [discussed in full later](#predatory)).

### Hybrid OA
Hybrid OA journals are increasingly the norm. You can decide upon acceptance of your manuscript whether or not you want to pay the fee to make your article open access. Again, note that your institution may have a deal with the publisher that means that anything you publish there is open access. It is well worth knowing these things in advance before you submit. If you can't afford to pay the open access fee then your manuscript will remain behind a paywall and be only available to subscribers.

A slight variation on hybrid OA is when you are a member of an academic society that allows its members to publish open access without extra payment. As a student you are likely to get very discounted membership to an academic society Which might make it very cheap for you to publish open access with them.

### Platinum or Diamond OA {#diamond}
This is without doubt the best OA model, and the one that we should all strive for. In a platinum or diamond open access journal you do not have to pay any money but everything that is published is open access. In order to do this these journals are often subsidised by governmental or philanthropic agencies. Some university presses are also in the habit of publishing platinum or diamond OA when it meets with their stated mission. These journals are very rare but they do exist, and it is well worth looking for them.

Another great new model that is diamond OA is the concept of [Overlay Journals](#overlay). Although there aren't any overlay journals in the biological sciences at the time of writing, I think that it's only a matter of time.

### Black OA
This refers to the placement of published material onto a pirate website such as [Sci-Hub](https://sci-hub.se/). Sci-Hub is considered by most governments to be illegal and may be blocked by your institution or country. However they do a great job in making scientific publications open access and need to be supported for that. Many scientists all over the world depend on Sci-Hub in order to access literature and therefore conduct research. In addition, there are a number of other Black OA sites: [libgen](http://libgen.li/),  [Unpaywall.org](https://unpaywall.org/), [Open Access Button](https://openaccessbutton.org/), [freefullpdf](http://www.freefullpdf.com/).

### Grey OA
You can find grey OA repositories of published material on Academic Social Network sites in which you need membership to access such as [ResearchGate](https://www.researchgate.net/) and [Academia.edu](https://www.academia.edu/). The legality of such sites is regularly questioned [see @piwowar2018state for more details]. There has been legal action with thousands of members being issues with take-down notices.

## Supplementary Information and Data
Whichever route you decide to go for your manuscript, please do not place your data with the publisher. There are some examples where publishers choose to place both data and supplementary information deposited with them behind a paywall, even if the article is available Open Access. We also need to ask whether the publisher has the long term vision to curate data, especially when the expense associated with this will rise over time as datasets accumulate behind their paywall. Elsewhere in this book you will find some suggestions about [what to do with your data](#transparency) to make it available for all.

## Unsure what you can legally do with a published paper?
If you don’t know what level of copyright exists on something that you have published, then you can find an aggregated set of publisher policies at [Sherpa Romeo](https://v2.sherpa.ac.uk/romeo/). This is a really nice database which provides a very simple summary by journal. You can also use this to check out a journal that you are thinking of publishing with. If you are still in doubt, then consult your librarian.

### 2021-09-05 2.07-submit.Rmd
Add files via upload
# all round update

# Submitting a paper to a journal for peer review {#submit}

This chapter deals with the process of submitting a manuscript to a journal (Figure \@ref(fig:Editorial-work-flow)), and what to expect once this is done. It takes time to submit a paper using most existing editorial management software. Detailed information is given on many of the parts, mentioned here, in later sections of the book. Be aware that although it does take a lot of effort to get a manuscript ready for submission, once it is submitted, there is a lot more work taken on by a larger group of people who are (usually) not paid, and who are undertaking the work associated with your manuscript in addition to their 'day jobs'.

Most of the information required for submission is easy enough to provide, but they insist on having it entered in such an unfriendly way that it makes it all very painful. Believe it or not, they have improved over time. As this work is so tedious, it's worth reflecting why they need all of this information upfront before anyone even decides whether or not they want your paper. The reality is that all of this metadata (data associated with your manuscript) is really only of use if the manuscript is accepted. Otherwise, you are really just stuffing the database of the publisher full of information that they may (and likely will) use to spam you in the future. The only data that they must have on submission is your name and contact details, and the verification that you've adhered to the journal's ethical requirements (which you could do in a [letter to the editor](#letter)). Some of the rest will be of use to the editor when deciding who to allocate your submission to, but the vast majority of the metadata are only used if your article is accepted - and then they become vital.

Once accepted, metadata about your manuscript lies at the heart of the ability for [CrossRef](#crossref) to link you, your co-authors and their [ORCID](#ORCID) accounts together with your manuscript using a [DOI](#DOI). All of this information is held in a header file of the published webpage so that Google Scholar can scrape it into their database. It's also used by all of your automated referencing software plugins. Having this data entered accurately means that the following processes will flow nicely. Doing a sloppy job will mean that those who rely on such services might well mis-cite your paper, get your name wrong, or one of lots of other potential issues. Most of the major publishers now use this metadata to make up the author information (and addresses) on the front page of the typeset paper. Be aware that when they appear wrong on your proofs, it's likely because you (or the corresponding author) didn't enter the metadata correctly.

(ref:Editorial-work-flow) **This schematic demonstrates the editorial work-flow of a 'typical' journal.** Ovals show actions by authors which have dotted lines, while rectangles show work done by the editorial team of the journal with solid lines. Referees are shown as diamonds. Letters in circles refer to the sections of the text (below). All arrows are potential places for delays.

```{r Editorial-work-flow, echo=FALSE, out.width='100%', fig.cap="(ref:Editorial-work-flow)"}
 knitr::include_graphics("figures/editorial-work-flow.png")
```

## A typical submmission work-flow

The steps A to G (below) all refer back to letters in circles in Figure \@ref(fig:Editorial-work-flow). Remember that most of the actions described below are also the subject of later sections in [Parts II](#part2) [and III](#part3) of this book.

**A - formatting and submitting**

 - Targeting journals for submission: there are a lot of journals out there, and you need to make sure that you are submitting your paper to a journal in the right subject area ([there is a detailed chapter on this subject](#whichjournal)). Remember to keep your ordered list of journals that you prepare so that you can refer back to this in the case of [rejection](#rejection).
 - Prepare your manuscript according to the journal guidelines: this may require a lot of work especially if the journal requires full formatting on first submission. Some journals require additional items such as graphical abstracts, so make sure that you know what is needed before you start to submit. A checklist to run through before submission is [available here](#formatting), to which you should add any journal checklist from their instructions to authors. Circulate this final version among your co-authors. This is a good time to gather the needed meta-data for submission. 
 - Get all the files and metadata ready for submission. In addition to the manuscript, figures and tables, you’ll (usually) need a [cover letter](#letter), key-words, recommended (or opposed) reviewers, and addresses (with [ORCID](#ORCID) numbers) for all authors. All these items should already have met with the approval of your co-authors.
 - For the purposes of this section, I am assuming that you are the corresponding author. This is something that you should learn to do. Being the corresponding author carries some extra duties as they are responsible for making sure that all the other authors are in agreement about the contents of the paper before submission. They are responsible for gathering all of the necessary information about each of the authors on the title page. 
 - Uploading your manuscript to the editorial management software requires time and preparation. Give yourself a good couple of hours for this process, and be aware that it can be frustrating. Friday afternoon might not be the best time. You may well need to refer back to your co-authors if you don’t have their relevant information. As corresponding author, it is courteous at this point to send a copy of the submitted version (usually the journal provides a pdf of the submission) to all of the other authors for their records.

**B - Editorial assistant and/or editor check the manuscript**

 - Once submitted to the system, your manuscript is usually checked. In bigger journals, this will often be done by an editorial assistant, while in smaller journals, the editor-in-chief may be responsible for moving all manuscripts.  before being passed to the editor. You may get it sent back if the meta-data is wrong. 
 - Once the editor (often Editor-in-Chief) has your manuscript, they will decide which Associate Editor (AE) will handle it.

**C - Associate Editor takes over**

 - The AE [should read the manuscript](#editors) and may [desk reject](#desk) it if they feel that it won’t make it through review. An AE [rejection](#rejection) isn’t great news as AEs don’t always have the best experience in knowing what will and what won’t make it through the review process. The editor usually has more experience. Hopefully, this won’t have taken long ([1-2 weeks](#timetaken)) and so won’t be very painful. This rejection may be fair or unfair, but it’s done and there’s nothing much you can do except return to  **A** (above), but see [later sections](#rejection).
 - Normally, your manuscript will be sent out to review and you can expect to wait 4-6 weeks (good), but sometimes [up to 3 months](#timetaken), for a decision. If it’s away for over 3 months, you should definitely make a query on the editorial management system. Unsurprisingly, authors feel more unhappy with the review results the longer the process takes [@jiang2021understanding].

**D - Reading the reviews**

 - Once back from review, you’ll get an email from the editorial management system with the decision.
 - __Reject and resubmit:__ This is a category that means you need major revision, but the journal doesn’t want the time that it takes to do this on their journal processing statistics. In many journals, this result has replaced ‘major revision’. Back to step 3 with a track changes manuscript and response letter to reviewer comments.
 - __Major revision:__ essentially the same as 6b. Both 6b and 6c result in your manuscript being reviewed again. You’ll need to carefully prepare both the manuscript and the response to reviewers as the reviewers will see both. Back to step 3 with a response letter.
 - __Minor revision:__ is unusual after the first round of review, so if you get this decision first time it's something to celebrate. You may have already done one (or more) rounds of Major revision before you get here. After addressing the Minor revisions, your manuscript should now only be assessed by the AE, so you should address your responses to them.
 - __Accept without further revision:__ is practically unheard of on first submission. Your manuscript may have already undergone some peer review (maybe as a preprint or in another journal), or in previous rounds of review. This decision is likely a recommendation from the AE to the Editor-in-Chief.

**E - Revising and resubmitting your manuscript**

 - If you are resubmitting, aim to prioritise this to get it done in 2 to 3 weeks if possible. 
  + The reason is that the same reviewers are likely to be willing to look at your manuscript again within a month, and will remember all the points that they made. Similarly, the AE will remember all of the issues that they had. It’s hard to stress how valuable this is, as keeping it all fresh will result in a swift response.
  + If you don’t or can’t manage to get your responses back quickly, you might expect a rocky ride through the review process when you go back for the second round. The reviewers you had before might not be available, but the AE will be obliged to have at least 2 reviews again. This means that you may get new reviews. New reviewers are likely to throw up new issues, and could result in your manuscript getting rejected at this stage, or that you’ll have another major review decision, sending you back to step 3 with a track changes manuscript and response letter to reviewer comments. This drags the whole process on for much longer and reviewers and AEs are likely to look less favourably at your manuscript. 
  - A better result is when there are only minor revisions. In this case the manuscript is simply bouncing between you and the AE and even if this happens more than once, it’s fine as long as you can keep the response time reasonable (within a couple of weeks).
  - In either case, your [rebuttal](#rebuttal) letter will be a very important part of your resubmission, and this will be covered in a [later section](#rebuttal).

**F - Associate Editor recommends to Editor-in-Chief**

- When the AE is happy with all of your Minor Revisions, they will make a recommendation to the Editor-in-Chief. The Editor-in-Chief may have some extra revisions that they would like to see, but these will likely be minor. 
- It is unlikely that the Editor-in-Chief would disagree with the acceptance of any paper that an AE has signed off on. 
- It is worth bearing in mind that the Editor-in-Chief does have the final say on whether or not your manusctipt can be accepted to the journal.

**G - acceptance**

- Hopefully your manuscript will now be accepted, and you are entering the last stages of the process. Your accepted manuscript should be sent to the publishers for typesetting, and you can get the proofs back very quickly (for some publishers). Most demand that the proofs are returned very quickly (often within 48 hours), and you should try to prioritise this if you can. If you can, please also send it to the co-authors. The more eyes the better at this stage for spotting errors. Don’t expect to be able to change a lot in the proof process, it’s really just for catching errors. Carefully check all figures, tables and legends. It’s not unknown that typesetters cause problems when they make proofs (tables can be disasters). A detailed section is provided in [Part III](#accepted) on what to do once your manuscript is accepted.
- If you (or a co-author) spot a fundamental error with your data or analyses at this point (or any of the other steps above), you should discuss it with all co-authors and decide what to do. It’s better to withdraw the manuscript now than to have to retract it later ([see part IV](#retract)).

**Rejected**

- If your manuscript is [Rejected](#rejection), take the comments of the reviewers on board. Think about it for a couple of days, and then set about revising the manuscript. However unfair you think the reviewers have been, there should be some important messages for you to consider carefully and discuss among the co-authors before going back to step 2 with the next journal on your list. More information to reflect on regarding a rejection is provided in a [later section](#rejection).

### Remember that peer review is conducted by humans

The peer review process is not ideal, but it is worth remembering that it’s there to help improve your manuscript. The most prominent problems involve the [time](#timetaken) that the editorial team take to find reviewers and have them agree to complete reviews in a [timely manor](#timetaken). In Figure \@ref(fig:Editorial-work-flow), only four of the arrows are in the control of the authors, while at least 16 are within the editorial management process. Each one could be the source of your manuscript getting stuck. Throughout the process, you should be able to track the progress of your manuscript using the editorial management software online. More information on delays in the handling of manuscripts is provided in a [later section](#timetaken).

### 2021-09-05 2.08-peerreview.Rmd
Add files via upload
# all round update

# Expectations of peer review? {#peerreview2}

Once your manuscript is submitted and has passed the initial rounds of checks (A to C in Figure \@ref(fig:Editorial-work-flow)), you can expect that you will receive a written review of your manuscript. Sometimes, you may also receive an annotated manuscript back, but you can interpret this in your own way - most likely as minor comments.

It is widely acknowledged that peer reviews are [likely to be biased in some way](#problempeerreview), and so you should expect this of every review that you get. Try to look over these biases and aim to receive the wisdom that they likely also contain.

## What are peer reviewers asked to do?
The peer review report consists of three major parts: the review, confidential comments to the editor, and the reviewer's opinion of what decision should be made. You will only have access to the first of these three. Understanding peer review is greatly aided by conducting peer review, which is covered by a section in [Part IV](#howpeerreview). Reading about [how to conduct peer review](#howpeerreview) may help with your interpretation of the reviews that you receive.

### The review
Peer reviewers should sum up the manuscripts in their own words to demonstrate that they have understood the contents. This is important because being able to summarise what you have read demonstrates the reviewer’s comprehension. If the reviewer gets this summary wrong, then it is either a flag to the editor that they lacked the necessary comprehension to make their review meaningful. Or, because there are two sides to comprehension, a flag to the authors that they failed to write their manuscript in a way that made it easy for the reader to comprehend.

The reviewer should then provide a general critique including positive and any negative aspects of the manuscript. They should provide detailed information on exactly how the manuscript should be improved, including any significant literature that might be missing from the manuscript. Lastly, they can provide a list of minor comments along the length of the manuscript that require further attention from the authors. When I undertake this last part, I tend to do it with page and line numbers (which is one reason why submitting a manuscript with line numbers is so important. If the minor comments get too numerous, then I tend to stick with major comments.

### Reviewer's confidential comments to the editor
The reviewers are provided with a box where they can write to the editor without text being seen by the authors. It is worth bearing in mind that the editor can act on these (unseen) comments. I'm not a big fan of confidential comments but sometimes they are warranted. Many reviewers avoid making comments in these boxes, in accordance with trends for [transparency](#transparent).

### Reviewer's opinion and the editorial decision {#AKP}
In many journals, the reviewers are directly asked whether the manuscript should be rejected or resubmitted (major or minor revisions). This opinion is given in a set of 'radio buttons' in the editorial management software. Depending on the journal, a number of additional qualitative questions are asked of reviewers that may be pivotal to the progress of a manuscript. These questions are set by the gatekeepers for the journal in question, and may ask whether the manuscript is worthy of publication in that journal, in the top 5% of important findings, or similar. Bornmann and Marx [-@bornmann2012anna] found that some journals require that all reviewers respond with a positive criterion on these questions, or the manuscript will not be considered further. They term this the 'Anna Karenina Principle' (AKP) of peer review because it requires 'positive selection' on all criteria, whereas previously papers were selected by meeting minimum standards; 'negative selection', which is now the overriding principle for '[no impact factor](#noIF)' journals.

Unless you know someone who can tell you the editorial practice of the journal you submitted to, you will not know whether or not your manuscript has been judged on the answers to these questions. However, if you receive all positive reviews and a rejection, it may well be that one or more of the positive reviewers did not select a criterion high enough on one of these questions. Personally, I don't think that reviewers should be asked these subjective questions, as this potentially trivialises decision making for gatekeepers. [Editors must read](#editors) the manuscript and all reviewer comments in order to reach decisions. Certainly, all journals need to provide [transparent criteria](#transparent) for selection of manuscripts, without which the black box nature via AKP workings of peer review will only perpetuate.

## What do good reviews look like?
Good reviews are those where authors improve their manuscript. It may be that the good review doesn't immediately show the way on the first reading (although clearly the best ones should), but it may be that the authors require some work rethinking their manuscript before they understand the comments of a reviewer. I would say that usually on first reading, even a good review might not sound that good.

### The myth of the 'shit sandwich' {#shit}
The shit sandwich is where a review where the beginning and end are generally positive, while the very critical appraisal happens in the middle. Although the shit sandwich might be seen as a way for a reviewer to sugar coat their negative message, an analysis of PLOS ONE reviews by Eve et al. [-@eve2021reading] suggests that some of the best reviews (think Minor Revisions) actually have this format. Here the reviewer will be positive in the outset and the summing up, but then in the middle have a set of issues that need correcting. Hence, they found that the shit in the shit sandwich wasn't that bad. In the same analysis, Eve et al. [-@eve2021reading] found that truly bad (i.e. Reject) reviews could be bad at any point in their length.

## What are reviewers not asked to do?
In their analysis of PLOS ONE reviews, Eve et al. [-@eve2021reading] found that reviewers are good at ignoring the directions that are provided to them by the journal. It is therefore necessary to be aware of what reviewers are not asked to do, because sometimes they do it anyway!

### Peer review is not a trial by committee
Those of you who have experienced manuscripts being critiqued at a book journal club will know that there are very few published papers that leave a journal club without having many negative, critical comments. Instead peer review is conducted by an individual, on their own and with their own personal limitations. Peer reviewers are forbidden from sharing the contents of a manuscript with others, without express permission from the editor.

### Correct English grammar
It's not the job of a reviewer to correct any faulty grammar on a manuscript. similarly it is not up to the reviewer to correct stylistic aspects of the manuscript. However, as English is such a subjective language, it is important that ambiguity is removed, and grammatical aspects can be important for this.

I have noted that some reviewers have become quite obsessive about things like the Oxford comma - insisting that the Oxford comma should be inserted at every possible juncture. Eve et al. [-@eve2021reading] call these reviewers ‘peer copyeditors’. They are likely to comment on your split infinitives and may be pedants for all grammatical concerns.

I would say that it is up to you as an author to decide whether the suggestions from these 'peer copyeditors' are warranted. At the same time, the comments from these same people might drastically improve the readability of your manuscript.

### Tell the authors to cite their papers
Sadly, this is something that a lot of peer reviewers do, and are part of a suite of not so legitimate ways in which people [manipulate citations](#notlegitimate). The reviewers’ own work can be cited, but only when relevant, and the authors should be able to choose whether or not they want to make any additional citations.

### Tell them to cite other papers from the journal
This also happens [see section on Impact Factor](#increaseIF) often at the request of the editor. It is part of IF manipulation, and you should not act on this.

### Justify comments with their own beliefs and opinions 
Peer review should always be an objective critique of a manuscript. It's not really the place of the reviewer to express their opinion or their beliefs about a particular study that they are reviewing. Watch out for statements that begin with "I believe..." or "In my opinion...".

Reviewers should stick to the evidence that they're provided with. If they are not provided with sufficient evidence then they should draw attention to the lack of information rather than extrapolate to what they believe might be the case.

Authors should be provided the benefit of the doubt and opportunity to respond to such criticisms especially when information is missing. It's unfair for reviewers to act as judge and jury. This is the job of the editor.

### Assess aspects of the manuscript that are beyond their competence
Some manuscripts are cross-cutting across several subjects or may contain analyses outside it with the experience of a reviewer. In these cases reviewers should not attempt to review areas that are beyond their competence. Instead they should bring these aspects to the attention of the editor when they are submitting their review.

### Ignore what is good
Is often thought that peer review provides only negative criticism [see @eve2021reading]. This should not be the case as peer reviewers should also be able to accentuate the positive aspects of manuscripts that they read. Even if a manuscript is not considered publishable the positive attributes should be bought the attention of the editors as well as those that are negative.

It is important for authors to understand what aspects of their manuscript are good. This kind of feedback from peer review will likely influence future versions of this manuscript as well as future studies from these research groups.

### 2021-09-05 2.10-editors.Rmd
Add files via upload
# all round update

# Why should an editor read your submission? {#editors}

There is a worrying increase in poor editorial decision making, because editors are not reading submissions. In their survey of Associate Editors, Poulson-Ellestad et al. [-@poulson-ellestad2020illuminating] found that they advised Early Career Researchers not to take on the position of Associate Editor unless they know that they have enough time in  their jobs. This is an assessment with which I agree. You must be prepared to carve out dedicated time in which you can concentrated several times a week, or nearly every day for the Editor-in-Chief.

When a manuscript is submitted to a journal, the submission goes to either the editor-in-chief or a handling editor based on the key words or journal section implied during submission (see [Part II](#submit)). In some journals (like [PeerJ](www.peerj.com)) the submissions are offered up to a whole group of editors who can take their pick. It seems that the next thing that happens is that the manuscript is sent out for peer review. But stop. That’s not correct and it’s really not a good way to proceed. Before sending it out, the designated handling editor needs to read the submission.

## Why is reading so important?
The title and abstract really don’t allow a handling editor to decide whether or not a manuscript should go out to review. There are a lot of manuscripts out there that should not have been submitted, because their authors do not have sufficient judgement of their own or because they believe that there is a reason to just ‘chance it’ (especially if the manuscript has been [rejected](#reject) by somewhere else already). It is very important that handling editors read the submission, because without that they are moving editorial responsibility from themselves to the peer reviewers.

Some years ago, I co-authored a series of articles [@perry2012peer] that were published across many journals about how peer review was becoming very difficult for editors because so few colleagues accept to do reviews. This was a problem then, and it’s a problem that has grown in time. I’ve recently sent out manuscripts to more than 15 people before getting two reviewers accept the invitation. That peers are not prepared to review, or in many cases even to respond to the request, is very poor. However, more recently I’m experiencing a sharp increase in manuscripts to review that should never have been sent out.

My time is precious, and it’s becoming quite expensive for my employers. I am happy to conduct peer review because it is an important part of the scientific publishing process, and I expect others to review my own work. However, I expect that any manuscript that I receive is worthy of my attention and time. If the handling editor has not read it, they cannot decide this and I really wonder what makes them think that they can send it to me (and presumably others) to read while they don’t feel that they have time to do it themselves. Moreover, this appears to be a trend among younger, less experienced, editors (often Associate Editors) that have either not received any guidance in what their job as editor is, or how to do it.

## If you are going to be an editor, then you must be prepared to read
I must admit that I’ve done it. I’ve sent out manuscripts to be reviewed even when I didn’t have the time to properly read the article myself. A a superficial skim suggested that it seemed fine. Not good. It’s embarrassing to have sent out manuscripts that should be rejected without peer review. In the case I’m thinking of, once I’d found time to read through the manuscript later on that day, I realised how bad it was and immediately wrote to those I’d asked to do the reviews and asked them not to. The article was rejected. It is important that this burden is taken on by the editor than burden two or three times as many other reviewers to make the same call.

Sometimes, it’s not clear whether or not a manuscript will pass muster. Articles can stand or fall on good or bad single judgements of the authors. But misjudgements aren’t always obvious to editors. That’s why peer review is important, and that’s why it’s hugely important for editors to send manuscripts to appropriate reviewers that have some expertise in a subject.

Science is built on the work that others have done before, but basing your work on what someone else has written will mean that you have a good understanding of what they have done and how they have done it. Assumptions have to be made to get anything done, and it’s a good exercise to sit down with a published paper (or even a manuscript of a colleague or your own) and read through listing all the assumptions that are made. Physicists might have a very long list if they read a biologist's manuscript, but with some practice you learn to see the assumptions that the authors have made when designing their experiment, or going out to the field to conduct their study. An incorrect assumption could lead to the entire manuscript losing its value.

In my example above, the authors might assume that they had correctly identified the species when recording its call. Such assumptions should be backed up with museum and/or tissue bank accessions. But when they are not, the assumption that the authors are recording what they think they are, is vital. If this is placed in doubt, then the entire premise (description of a call to distinguish this species in the field) simply falls apart. In a case without vouchers, the assumption needs backing up by someone who knows the identification from another study, or without any foundation it becomes worthless.

## I’ve been on the other end too
I’ve submitted manuscripts to journals where the editor clearly never read the manuscript. Editors who have made a decision without any guidance of their own gives this away. If your decision comes as a single sentence that asks you to revise according to the reviewers’ comments, then you can be reasonably sure that your editor hasn’t read the manuscript (and possibly not even the reviews).

It’s not surprising that the editors have little to nothing to say; without reading the manuscript, the reviewer comments aren’t really very helpful. Without reading, the editor has no idea whether the reviewer is biased. As an editor, you simply have to read. And if you don’t have time to read, you shouldn’t be an editor.

## There is worse that goes on in economics {#economics}
If the above makes some editors in Biological Sciences look bad, then I apologise. Being an editor for a journal is a pretty thankless task and there is no financial gain when doing an editorial stint. However, if you're going to do it, then you must do it well. The half measures that I describe above are simply not good enough. But biological journals are a huge cut above those in economics. I've always had my doubts about economics as a subject. Rather like theology, it's based on a fanciful construct that puts its own practitioners in positions of power when we'd do just as well to flip a coin.

In May 2018, I was pursued for some weeks by the _International Journal of Finance and Economics_ to conduct a review of an article submitted there. Even though I raised the flag that I was not an appropriate reviewer, the editorial assistant (not the editor) still wanted me to conduct the review. Apparently, 'the system recommended me' and this was enough for me to be selected. It appears that the problem of non-expert reviewers is on the increase. Consider this blog post by Ivan Oransky [-@oransky2021elsevier], one of the authors of [Retraction Watch](https://retractionwatch.com/) who was invited to review papers on COVID-19! Essentially, this is the result of editorial management systems ‘auto-suggesting’ reviewers, and editors not doing their due diligence to determine whether any of these reviewers is worthy of conducting peer review on that submission.

Clearly, selection of reviewers must be done by the handling editor, and those people must be chosen based on their expertise (not lack of it). While editorial management systems might help editors, they can't replace diligence on behalf of those who are responsible for the upholding integrity of the peer review system.

## Summing up on editorial blunders
The way to get round making the kind of editorial blunders I describe above is simply for editors to read their manuscripts. The guidance of how to read a manuscript should be explained to editors when they take up the position. There is plenty of information out there on the internet, but the journal’s editorial policy should be understood by all of the editors (and preferably open to authors and reviewers too), and that should include reading manuscripts before sending them out for peer review.

### 2021-09-05 3.00-part3.Rmd
Add files via upload
# all round update

# (PART) Part 3: After your paper is accepted {-}

# Now that your manuscript has been accepted {- #part3}

This part of the book gives you information on some actions that you might want to do after your paper is accepted for publication. Much of this section is concerned with publicising the results and content beyond academia and for non-academic audiences. Even if you feel that there would be no interest beyond your academic niche, it would still be worth making some effort to publicise and popularise your study.

## The Version of Record {#VoR}

An important concept to understand in the publication of your article is the Version of Record (VoR). This is the final typeset version of your article that is published.

In this millennium, the VoR had changed from a hard copy that was printed and bound into an issue of a journal into a pdf that appears at a journal website online. In addition, the VoR no longer has to belong to a volume or an issue, and is usually the first version available online [@haustein2015when]. The VoR can appear online long before it appears in an issue or volume (i.e. without page numbers), but still be the VoR.

### Registration to obtain a DOI

If the publisher is a [Crossref](https://www.crossref.org/) member (most are), they will register the content and assign it a [Digital Object Identifier (DOI)](#DOI).

### What does having a VoR mean?

 - This means that if you wanted to make any changes to this first printed version (the VoR), you'd need to publish a separate [corrigendum](#corrigendum)
 - The date of the VoR has an impact on [primacy](#whatjournalfor)
  + in taxonomy, for example, if there are two descriptions of a species, the earliest one counts as the valid one

### 2021-09-05 3.01-accepted.Rmd
Add files via upload
# all round update

# Once your paper is accepted {#accepted}

The day your paper is accepted, tell your co-authors. If they are in the same physical location as you then buy a cake, and celebrate with them at tea time. If cake isn’t your thing, then find another appropriate treat for you and your co-authors. It’s a great achievement and something you should share together. If you aren’t in the same location then make a plan for next time you meet together, even if that’s only online. There is real value in celebrating the positive times in your time as an Early Career Researcher. We all know that sharing the times when you are [rejected](#reject) will improve that experience. Similarly, celebrating the good times will give you more impetus to keep pushing through the down times. Extending this positive feeling to your co-authors, or those in your laboratory will always be appreciated.

Remember to make sure that all of your co-authors have a copy of the accepted version of the manuscript. Sometimes referred to as a **postprint**, you and your co-authors should submit this to your institutional repository so that the article can be reached as [Green Open Access](#green) by anyone who is interested in reading it.

At the same time as your manuscript is accepted, or shortly thereafter, the publisher (if you are using a traditional style publisher model) will tell you when you can expect to receive the proofs.

## Take your time with proofs {#proofs}

The next step in the publishing process, once your paper has been accepted, is that it will go for type setting. Depending on the journal and the publisher, this process can proceed in several ways. Typically you will receive a notification that your paper proofs are ready and that you need to check and return them within 48 hours.

Some journals may have an additional __copy editing__ (also known as sub-editing) step. This is essentially an extra step checking journal style, syntax, punctuation, spelling and grammatical consistency. The copy editor may be tasked with checking additional issues, like species names (taxonomic authorities), names of chemical compounds used, etc. All will check citations against literature cited, and check that literature cited is sufficient to link with [CrossRef](#CrossRef). Most journals that I have worked with roll copy editing and proofing into a single step.

If you haven't checked proofs before then it is important that you read the instructions from the publishers, that come with the proofs, carefully. They should tell you exactly what to do and if you are unsure about anything then talk to your co-authors.

Typically the publishers will send you a set of queries (copy edits) that relate to your proofs. They always ask you to check every author's name and affiliation. Other typical errors are that there are citations in the text that are not in the references. Or that there's literature in the references that are not cited.

The process of checking the proofs is very important. Errors can creep in during the type setting stage. Equally, copy editors can get it wrong, and you should read the text very carefully to make sure that errors have not been introduced. Pay special attention to the tables, table legends and figure legends.

You may also have the opportunity to change the size or orientation of figures, especially if it looks like they are not well presented in the proof. If the journal prints into columns, they may choose to put your figure in one column instead of across two. Another option is to have your figure in landscape across the whole page. Journals are generally pushed for space and so may refuse some requests for more room for larger figures. But you might get lucky if you make a good case.

Although the proofs are the responsibility of the corresponding author, it's good to get as many eyes on them as possible in order to spot any possible errors. Some journals let you know when proofs are likely to arrive, in which case it's a good idea to alert your co-authors and ask them whether they are prepared to look at them within that temporal window. I usually suggest that you make all your own corrections first before sending them around. Worst is for everyone to make their own corrections independently, as you'll have to put them all onto one set of proofs, and everyone is likely to spot some of the same errors.

Many publishers want to proofs back in a hurry (typically 48 hours). If you don't have the confidence to correct proofs yourself, or cannot pass it around your co-authors within this deadline, then you can simply write back and ask for an extended deadline for your proofs. It is important to get it right, and much better than having to correct the paper later.

It is important that you make any corrections needed on the proofs. They are important to get right because once the proofs are submitted and the [Version of Record](#VoR) is produced, any changes that you may want to make will require an official correction in the form of a [separate publication](#corrigendum).

Probably the easiest way of doing proofs is to print them out and go through them with a pencil first. This allows you to take your time and you're more likely to spot errors this way than on the screen. However, some publishers will require you to submit proofs in an online system (effectively working with their LaTeX document). You should still have an opportunity to print and take your time with the proofs though, and these systems do allow for this.

Sometimes, there are a lot of problems with proofs, and you may not have the confidence that the publishers will make all the corrections as indicated. In this case, you should ask to see another round of proofs before committing. Most publishers will happily send you a second round of proofs.

## The DOI for your paper {#DOI}
The Digital Object Identifier (DOI) is an international standard (ISO) unique character string to identify physical, digital or abstract objects. The DOI is a "persistent identifier" or "permalink" which means that it remains unchanged even if the document itself changes location. For example, if the society that owns the journal changes their publisher, the DOIs of all content remain with the same documents even when they leave the old publisher's website. This means that although the publisher will provide a link to their website, and you may have your own repository with a link to your paper, you should rather use the DOI as the link to give out to everyone.

When you cite the DOI, don't take it apart, but provide the full link. For example, https://doi.org/10.1242/jeb.233031 should never be abbreviated to 10.1242/jeb.233031 or doi.org/10.1242/jeb.233031.

You will note that old DOIs look different (they used http instead of https protocol, and typically have dx.doi.org addresses), but you should not edit the start of their address, but leave them. For example: http://dx.doi.org/10.1098/rspb.2018.2528

Sometimes your article may acquire several DOIs, for example from the publisher and from the preprint server. Sites like figshare and ResearchGate, assign different DOIs to content uploaded to their platforms. These are DOIs but they’re assigned by DataCite, another DOI registration agency. The different registration agencies provide different services that relate to DOIs and their associated metadata (from construction to the movie industry), and have different requirements for their member organisations. To prevent ambiguity, you should always use the DOI associated with the [Version of Record](#VoR). The biggest player registering content in scholarly output is [CrossRef](https://www.crossref.org/).

There are still some journals that do not issue DOIs. But the cost associated with adding DOIs by small publishers is changing, and so it is likely that in future a DOI will be associated with all scholarly output.

DOIs are extremely useful as you can usually click on an active link DOI and go straight to the article in question. Therefore it is well worth adding the DOI to your references if you can. Most reference databases will do this automatically, but if they don't pick up the DOI for some reason, then it's worth adding it yourself. Some journals will demand it, while others have yet to come around to how useful they are.

However, DOIs cannot replace references, otherwise we’d need to be able to click on every link all the time, and couldn’t read any paper without a connection to the internet. It’s still really useful to be able to read a formatted reference at the end of a paper.

Publishers normally deposit DOIs and other metadata (authors' names and addresses, publication dates, title, licence, funders, etc.) around the time the article is published online. This is called content registration. This is not just for articles, but also preprints, conference proceedings, books, book chapters, peer reviews and more can all have their content registered with CrossRef. The standardisation of this metadata means that not only is it possible to immediately find your article (the DOI is a link), but all of the metadata associated with it can be cross-referenced. This is of great help to funders, for example, who can look up all the products that their funding has produced without having to contact all of the people that they funded.

### DOI Tools
If you can't find the DOI for an article that you want to cite, then there is this very useful online software that will provide the DOI if there is one for every reference you enter: https://doi.crossref.org/simpleTextQuery or use the Crossref Metadata Search: https://search.crossref.org/references

There’s an equally useful database that provides BibTex for DOIs that you enter: https://doi2bib.org/

## Add your paper to your ORCID account {#ORCID}
The ORCID (https://orcid.org/) number was devised to provide a common platform for authors to curate, and now some journals and funding authorities won’t allow you to submit or apply without one. This helps when there are authors with common (or even identical) names. As an NGO dedicated to helping scientists ascribe [credit for their work](#rescognito), should be supported. ORCID will also help prevent author fraud (see [Part I](#authors)).

If you added you ORCID number to the metadata when you uploaded your manuscript, your ORCID record should be updated automatically (through Crossref) when the DOI of your article appears online. Inside your ORCID account, you can grant permission for Crossref to update the records automatically, otherwise you will need to log onto ORCID from time to time and approve changes suggested there. Alternatively, you can add works to your ORCID record via [Crossref Metadata Search](https://search.crossref.org/references).

Be careful not to create duplicate accounts, and make your ORCID available to your collaborators so that they aren't left guessing at your identity when it comes to submitting a paper. ORCID allows you to create some useful links for [your website](#website) or a QR code for adding to posters and presentations. If you find you have a duplicated account, it is simple to remove it (see [here](https://support.orcid.org/hc/en-us/articles/360006896634)).

## Once you have a publication date
Once your paper is published you have an opportunity to publicise it yourself. There are lots of different ways to do this, [see the next chapters](#populararticle). Some funders and institutions will want to know about press worthy publications before they are published, so that they can prepare a [press release](#pressrelease).

## On the day you publish
This is a great opportunity to contact all the people who helped you in your study and send them a PDF of your paper. The easiest way to do this is to go to the acknowledgements section and write an email that includes everyone that is mentioned in the acknowledgements. Write them a nice email in which you thank them for their help and explain briefly the significance of the paper.

It is a very good idea to keep all of these people informed about your publication as soon as it is published. You really want people in [your network](#network) to hear about it from you first and not from somebody else. This includes contacting any authorities that have issued permits. You may also want to contact funders.

### 2021-09-05 3.02-pressrelease.Rmd
Add files via upload
# all round update

# Writing a press release {#pressrelease}

Many of the aspects to do with writing a press release are similar to those needed when writing a [popular article](#populararticle). Try to make your text newsworthy. Remember that journalists are looking for new things that's why it's called 'The News'. Your press release must be about something that has happened recently. There's no point in writing a press release about a paper that was published six or nine months ago. It’s very unlikely that you’ll find anyone interested in writing something after that amount of time.

Here are 10 simple steps to consider when writing your press release:

1. __Choose your hook__. The paper that you wrote may have several important findings. You are going to need to choose one easy to understand finding for your media release. It's usually quite simple to decide; take the thing that would most impress your Auntie Fanny. 
1. __Write your headline__. Like choosing a good journal title or a popular story title the headline should try to encapsulate the study perhaps with a witty angle. Don't make it too long, eight to ten words at most. Most importantly your headline should connect with a wide and general readership. There's no need to get too fond of your headline because, if they take your story, news outlets are likely to want to write their own.
1. __Crafting the first paragraph__ is important. You need to sum up the study together with the finding. Even if your reader only reads the first paragraph they should have an understanding of what you've done and found. This first paragraph should not be longer than 30 words.
1. In the second paragraph, you should __state who you are and where you are from__, both geographically and the name of your institute. Here you need to concentrate on getting across the information on why you're finding is interesting. A typical second paragraph might read:

>Dr Frankella Smith from FitsSimon’s University found a new species of woodlouse when bending down to tie her shoelaces last month. She published her findings today in the journal _Cobblers Oniscids_.

5. In the next two paragraphs you should simply explain more about the background to your story and why the finding is interesting. Don't be tempted to deviate from the hook that you've chosen. After reading these two paragraphs your reader should be able to answer the question: So what?
1. Finally sum up your finding with a quote from you, the author. Either use the quote to emphasise the study results, or you can try and humanize your findings. This means a way of connecting with the reader, especially if you feel that the rest of your texts won't:

> "I never expected to find such a pretty woodlouse in my shoe", said Frankella. "I was flabbergasted when it turned out to be new to science."

7. Include your name and contact details of the person that the press should contact in order to find out more about the story. 
1. Give the full citation to the paper with all the author names and the journal name plus a link so that any journalist can find the full text online.
1. Include one or two photographs or relevant graphics that the press can use. If they are not taken by you then make sure that you have permission to use them. If you can, include a picture of the study organism, or even better of you with the study organism. 
1. Seek feedback. Send your press release to your university’s press office and ask for feedback. Those are the professionals and they should be able to help you.

Of course, the better your press release is, the more likely it will be that people will write about it. Remember that it also matters a lot about the subject of your paper. The media are likely to be far more interested if your work is on dolphin communication than if you are writing about isopod communication (just like your Auntie Fanny). Having said this, never be put off just because your organism or system isn't cute and cuddly. Try asking your non-academic friends about the newsworthiness of your press release and see what they say.

### 2021-09-05 3.03-populararticle.Rmd
Add files via upload
# all round update

# Why write a popular article? {#populararticle}

There are many reasons why it is important to communicate science beyond your own discipline and into the wider public forum. Primary among these is that in a society based on decisions made on the basis of science, it is our responsibility as scientists to make sure that we make the findings of our work, upon which basis political decisions are made, understood to the widest of audiences. I do not mean that we are sharing science pejoratively to an ignorant public, but instead as equals in our collective scientific society. We share with a wider public in the same way that we share with those who are used to reading a well reasoned newspaper article, or listening to an informed political debate. By sharing our work, we help affirm that decisions should be made objectively, and we make the most important connection by reaching out to the rest of our society and to join them in the scientific project.

**Here are some extra reasons why communicating by writing a popular article might be right for you:**

- Inform tax-payers who funded research what you found
- Increase profile of your work and you as a researcher
- Reach other researchers (who also read popular articles)
- Reach other stakeholders like practitioners or policy makers
- Open more doors to other potentially cross-disciplinary work
- Gain new insights into how your work appears to the general public
- Public communication is a key part of social responsibility, quickly becoming a key aspect of an academic career
- Maintaining and furthering the Scientific Project

The sooner that you come to terms with the need to communicate your work more widely, the more comfortable you will be when you are contacted by a reporter, a vlogger or someone from TV or radio.

## Here’s a quick guide on how to get started writing a popular article.

Just like any writing project, there is no one way to write a popular article. I provide the following advice in order to get you started. You are, of course, free to write however you think your work will be best understood and appreciated.

### What’s the hook?
Your popular article will not be the same as your paper. You should plan to have a single fact or message that you want the public to walk away with after reading your article. This is likely to be the same as the main result in your paper.

When composing your article, you need to be single minded about achieving the understanding of your hook. The article cannot take any side roads or distractions, but must stick to the main point. Once that’s done, provide the “so what” that allows the reader to see the bigger picture, and maybe where you would go next.

### Don’t get complex or technical
If your whole article hinges on something technical, you might have to start by explaining it simply. If you can’t easily explain it, then this is probably the wrong subject for a popular article. Don’t worry about leaving out (what might be to you) key details, you can always refer the reader to your article if they want to know more.

### Always refer to your published work
Make sure that you always have some reference to your work that’s published. Provide a hyperlink, but preferably give the full citation. Be aware that news items count towards the [altmetrics](#altmetrics) of your article, so be sure to link it correctly.

### Pictures, videos and even sound files
These are great to help readers engage with your work. Try to choose images that tell the story with the same information as you have in your article. If you don't have any, then try asking your co-authors, and then try to remember for the next project that you need to collect these when doing your research as it really helps when publicising your work.

### 2021-09-05 3.04-altmetrics.Rmd
Add files via upload
# all round update

# Altmetrics from traditional and social media  {#altmetrics}

In recent years, more emphasis has been placed on the way that scientists communicate their work. Many institutions now consider the degree to which scientists communicate their work as one of several key performance areas on which they are judged. Because administrators are always looking for simple solutions to evaluate the work of many different types of academics, commercial solutions to measuring the degree of communication for each publication have sprung up. The most ubiquitous of these in biological sciences are **'altmetrics'** [@priem2012altmetrics]: alternative metrics that aim to measure activity on the internet through social media (e.g. [Twitter](www.twitter.com), [Facebook](www.facebook.com)), online reference managers (e.g. [Mendeley](https://www.mendeley.com/guides/desktop), [Zotero](https://www.zotero.org/)), blogs and news outlets. Because of the immediacy of these activities, altmetrics tend to accumulate much faster than traditional citations, giving a near immediate impression of the interest generated in an article.

A prominent company producing altmetrics for many biological journals is Altmetric (www.altmetric.com). The whirls they produce, known as ‘Altmetric badges’ are coloured to show the proportions of different media that have been scrapped from the web (Figure \@ref(fig:atlmetric)). Although Altmetric are widely used, the calculation of their impact score is not transparent. For example, in March 2021 they changed the weighting of their impact score so that Tweets (which were weighted at 86.9% of the score), lost 75% of their effect [@anderson2021altmetric]. They did this without an announcement or the knowledge of their staff. Verifying the calculation of scores doesn't always add up, and I've noticed that even their web crawling news coverage can be hit and miss.

A better option would be an open source tool to track altmetrics data transparently. Such tools have been developed (e.g Paperbuzz: paperbuzz.org and ImpactStory: profiles.impactstory.org), but at the time of writing these initiatives are unfunded and adrift.

(ref:atlmetric) **The whirl output for a single paper from an Altmetrics analysis**. In this example, a paper by Baxter-Gilbert et al. [-@baxter-gilbert2020shrinking] was covered by many Tweets, news outlets, some blogs and a Facebook mention. Altmetric provides an overall score, but different types of mentions are not, so a news outlet is awarded a higher score than a tweet. Although this paper did not garner interest due to a charismatic species, the story was of general interest to the public as it centred on island dwarfism. Reproduced with permission.

```{r atlmetric, echo=FALSE, out.width='40%', fig.cap="(ref:atlmetric)"}
 knitr::include_graphics("figures/altmetric.png")
```

You can track your own altmetric profile on ImpactStory (profiles.impactstory.org) - use your [ORCID](#ORCID) number to see the altmetric impact of your articles. You'll need to have your account visible and your profile information will get pulled through automatically to ImpactStory, although you can login and customise your profile (Figure \@ref(fig:impactstory)).

(ref:impactstory) **An altmetrics profile from ImpactStory**. This is [my researcher profile on ImpactStory](https://profiles.impactstory.org/u/0000-0001-9939-7615). This gives you a summary of mentions on different social media outlets for all of the publications that you have logged on your [ORCID](#ORCID) account.

```{r impactstory, echo=FALSE, out.width='85%', fig.cap="(ref:impactstory)"}
 knitr::include_graphics("figures/impactstory.png")
```

In biological sciences, there is a traditional bias in media coverage towards species with higher charisma [@bonnet2002taxonomic; @ducarme2013what]. This means that if you work on whales or roses, your work is likely to generate much higher altmetrics than if you conduct equivalent work on phasmids or grasses. Traditional media is starting to make an effort away from only reporting on science with charismatic species, but they are driven by a public with insatiable demand for kittens and flowers.

There is a lot that you can do to improve the level of your altmetrics, like writing a [popular article](#populararticle) or a [press release](#pressrelease). As communication is becoming so important in the careers of scientists, then I’d suggest that you remain aware of altmetrics and how they are used by your institution. Be aware of how to influence and increase your score. For example, if you and your friends tweet about your article, make sure that there is a live link to the article on the publisher’s website. Similarly, if you are contacted by a news outlet about some of your research, you can insist that they place a link to your paper in their article. If the Alemetric scraper cannot find coverage on your paper, you can inform them [here](https://www.altmetric.com/about-our-data/our-sources/).

### 2021-09-05 4.01-openaccess.Rmd
Add files via upload
# all round update

# Is Open Access good? {#openaccess}
Open access appears to be a great initiative that acknowledges that everything should be free to view. Neither scientists nor the public that fund them should be barred from accessing the knowledge that they produce. What could be wrong with Open Access? Many people have written a great deal on what has gone wrong, and what I provide here (and elsewhere in this book) is not the last word as this is a dynamic topic that is changing all the time.

## So what is open access?
I have covered the many different kinds of Open Access (OA) elsewhere in this book ([see part II](#openaccess1)). Here I concentrate on the fiscal implication of OA. Someone needs to pay for the work done. Who should pay and how?

The principle of Open Access is something that is easy to appreciate, and lies very much at the heart at the movement from [closed to open science](#openscience). Some history is appropriate when we talk about Open Access, because the movement started as a genuine attempt to change the publishing model for the better [some key texts including those by @poynder2019open; @poynder2020information]. Hopefully in the future this experiment will be seen as a success. In the meantime, Open Access has been hijacked by publishers as a way in which they can earn large amounts of additional income. The story of _PLOS ONE_ is useful to see how this change came about.

### _PLOS ONE_ {#PLOS}
The Public Library of Science (PLOS) started soon after the turn of the new millennium, primarily as a concern to the [problems with peer review](#problempeerreview) that are highlighted elsewhere in this book. This really was a new experiment in publishing, that was to radically alter the publishing model in the 21st century, and in only ten years, _PLOS ONE_ became the world's biggest journal. The idea was to do away with all the normal waiting times in publishing, and to have articles published immediately online, and without any charges associated with accessing them: Open Access. At the time, this was all fairly radical, and it became very successful with a large amount of interest from scientists who submitted their articles to PLOS. But the founders were frustrated that they were rejecting a lot of papers that were technically good, but not selected as they were not novel, or did not significantly advance their field. In 2006, PLOS dreamt up a new journal to take all of these technically sound manuscripts, and this was the birth of _PLOS ONE_.

From the outset, _PLOS ONE_ was not interested in acquiring any [Impact Factor](#impactfactor). Instead, it was mostly interested in taking technically sound manuscripts irrespective of their results. This meant that reviewers for _PLOS ONE_ were (and still are) asked to assess the technical soundness of a manuscript, and not to judge the 'value' of the results. The original idea was that reviews should be post publication, through comments made on the website after the publications were posted. The initial review then, was simply meant as a preliminary check for technical soundness. This was again a radical departure from the publishing norm by the PLOS group, when many other journals were still [rejecting](#rejection) manuscripts that had no faults other than they were not attractive to editors. _PLOS ONE_ was opened to accept all such manuscripts, effectively doing away with [publication bias](#badscience). As you might imagine, at a time when academics were under pressure to ['publish or perish'](#publishmore) rejections were plentiful and so manuscripts quickly found their way to _PLOS ONE_. Because _PLOS ONE_ was also Open Access, it's papers received an extra boost of citations through visibility, and ironically after five years it found that it had an Impact Factor to rival some of the journals whose rejections it was picking up. This meant that more and more scientists started submitting to _PLOS ONE_ as their journal of choice, mostly because it was publishing Open Access.

_PLOS ONE_ became the worlds biggest scientific journal, and in 2013 it published 32058 papers [@davies2019plos]. It's worth taking a moment to do the arithmetic on the income that PLOS received for these at > USD 1000 per article (so it seems inconceivable that they could make a [loss](#loss)). By this time, other academic publishers had noticed the rocketing ascent of _PLOS ONE_ and had responded by starting their own 'no impact factor' journals that could be fed by the rejection piles of journals already in their stables. Notable among these was _Scientific Reports_ from the Nature Publishing Group which started in 2011 with 208 papers, and after 5 years ramping up to nearly 20000 papers in 2018. For whatever reason, the Impact Factor of _Scientific Reports_ is roughly double that of _PLOS ONE_ for around the same article processing charge (APC), drawing far more researchers in. During this same period, since 2013, a great many such examples of 'no impact factor' journals that accept any article that is technically sound began to appear. They are not only the general behemoths like _PLOS ONE_ and _Scientific Reports_, but you will find that even niche journals that make significant numbers of rejections of technically sound papers have their own version that they will offer to you in the same email that informs you of a [rejection](#rejection) from their flagship journal.

The bottom line, from this story of _PLOS ONE_, the highly innovative journal with an attempt to fundamentally change the publishing landscape, is that this is exactly what it did. However, somewhere along the way, the well meaning ideas were hijacked by mainstream publishers' greed. They saw a massive and untapped market. Unsurprisingly, they decided not to call these new entities 'no impact factor' journals, or even the 'rejection pile', but they were binned under the [Open Access](#openaccess) banner.

The bandwagon movement to Open Access that happened as a result of _PLOS ONE_ success, is probably one of the best scams that the publishers have come up with to date. Now the scientists pay for making their own content open for anyone to read. They pay a once off fee to the publishers to typeset the manuscript and host it on their site without a paywall. And how much do the publishers want for this service. Prices start from USD 1000 and go up to around USD 12 000. More frightening, and clear evidence that publishers are simply taking advantage of scientists, is that prices increase with the [Impact Factor](#impactfactor) of the journal [@gray2020sorry], although the costs involved to the publisher remain static. What further evidence do you need?

## So does that mean that OA journals are now free?
Mostly no. The majority OA model ([hybrid OA](#openaccess1)) means that a minority of articles in these journals are free, but the universities are expected to subscribe to those same journals at ever increasing prices because much of the rest of the content is still behind the [paywall](#paywall). This is because most authors cannot afford to pay the exorbitant fees charged by the journals (although some countries now have this payment as mandatory - [cOAlition S](https://www.coalition-s.org/), they and their scientists are still in a minority). There are some journals that are entirely open access ([gold OA](#gold) and [diamond OA](#diamond)). These are (almost) exclusively online and have never been part of traditional packages that university libraries spend so much of their budget on. Hence the fact that they are entirely free does not impact library budgets. The new way to fleece our libraries is through '[transformative publishing agreements](#paywall2)' [@janickehinchliffe2019transformative], estimated to cost up two to three times as much [@poynder2020information].

But paying for open access has not reduced the cost of access to scientific journals for libraries. This cost constantly goes up. [Hybrid OA](#openaccess1) was a brilliant scam dreamt up by the publishers, because for much of this content we pay not twice but thrice [@buranyi2017staggeringly]!

There are people that are trying to stop this constant escalation of publishers wanting money. You, as an Early Career Researcher, are in one of the best positions to do something about this change from [closed to open science](#openscience). Currently, [diamond OA](#diamond) journals are a very small and quite unusual component of the publishing scene. But it is totally possible for these to become the mainstream using initiatives like the [overlay journal](#overlay) system. The current dynamic will only change when academics submitting their manuscripts change from for-profit publishers to other models. The power is literally in your choice of [where to send your manuscript](#whichjournal).

## Nature Publishing - pay for review
The Nature Publishing Group has now proposed to go one step further: pay to submit your paper for review! In this new twist on the OA scam, the exclusive Nature publishing group is offering a discount to their new USD 11390 OA charge if you pay USD 2665 upfront before review. In this case, you will have the benefit of being told that your manuscript (most likely) is not publishable in Nature, but you can pay some more money to them in order to publish in one of their other journals. The most lowly (Nature Communications) would be another USD 3160, leaving you a few hundred dollars short than if you had submitted there originally. Or you could be rejected outright, and your USD 2665 is, of course, non-refundable.

- Are Nature Publishing Group taking advantage of authors who aspire to publish in their exclusive titles? __Yes__. 
- Will they use the money to plough back to the benefit of science, scientists and the taxpayer? __No__. 
- Will the cover price of Nature go down when some of the articles inside are free to read? __No__.
- Will it be cheaper for libraries to subscribe to Nature? __No__.
- Will they line their pockets? __Yes__.
- Will they pay the scientists who conduct the peer review that they want to charge for? __No__.
- Can you see the common thread in the greed of publishers and the willingness to prey on scientists who need to publish in high IF journals for their career? __If not, read [Part IV](#part4) again.__

For a lot of academics there is a trade-off: vanity for cash. We know that for a lot of academics, their vanity will mean that they are prepared to find the money (even if it comes from their own pockets). However, for as long as enough are willing to pay, we will continue to be exploited. Instead, we really need a new model for publishing. For some academics they will be paying directly for a chance at [getting a job](#jobinacademia) or [getting tenure](#habilitation). The chances of these individuals reaching their goal from a privileged institution then simply becomes reinforced.

This is not a system that we should ever support, and one that we should do all we can to remove for the sake of transparency and equality in the greater [scientific project](3beforeyoustart).

**We need a new model for publishing without for-profit publishers**

## The wicked problem {#wicked}

There are growing problems with publishers. Publishing has become very expensive for scientists, and the public who fund science. Many funders are now unwittingly funding publishers instead of science. The history of scientific publishing includes a role for commercial publishers right from the very beginning ([Part IV](#paywall2)). This relationship continues to the present day, but while the origins saw learned societies commissioning publishers and then distributing their content, today the publishers have climbed into the driving seat, conceiving and owning many of the current journal titles. Driving the hiring and promotion of scientists that feed their own products ([Part IV](#badscience)), ultimately leading to [fraud and misconduct](#pruittdata). We, as scientists, have allowed this to happen and we are certainly the people that continue to perpetuate the dominance of for-profit publishers in our scientific domain.

A wicked problem is one that is not only complex, but lacks clarity (with respect to solutions) or a way to scientifically test and study it. According to Rittel and Webber [-@rittel1973dilemmas], there are ten important characteristics, all of which are met by the ongoing situation in academic publishing. There are multiple stakeholders in our wicked problem, and they include our employers, our funders (political and societal), our peers as well as the gatekeepers and societies that police the system. I maintain that there is no solution that will benefit all parties. A practical solution that does not include the for-profit publishers will result in levels of disinformation that will rival those from the oil and tobacco industries. It is up to you to be aware of the options and become part of a solution that benefits science.

As scientists, we need to go back to our core concerns around **the scientific method: rigour, independence, transparency and reproducibility**. We need to persuade our institutions that our core concerns are what we should be measured by, and that the new range of Open Science tools be used to determine the extent to which we live up to our scientific values. Getting jobs, tenure and grants should be based on our Open Science credentials instead of the publishing industry serving [Impact Factor](#impactfactor) and other [publishing metrics](#citations).

### What do publishers do? {#deals}
The publishers would claim that they do an awful lot. All they really do is pay for the layout, printing and distribution of journals. These days ‘printing’ really means hosting electronic pdfs only, as there’s very little paper that’s printed, and you can be sure that paper subscribers now pay the extra cost of any additional fees. The layout from the manuscript (most often a MS Word document) into a pdf does take some skill and talent, although nothing like what you might expect given how much the publishers charge. You can be sure that they don’t pay much for this service as it is almost all outsourced ([cOAlition S](https://www.coalition-s.org/plan_s_principles/) have insisted that OA journals become transparent with their costs). Quality can be good, but more often quality control is completely lacking. Most authors have stories of how manuscripts have come back mangled, although my own impression is that the worst days appear to be over.

This is not to say that there are no skills in the publishing world. But the reality is that these days it is possible to use free software like R Markdown [@xie2018r] to write papers that can quickly and easily be made into any sophisticated layout using LaTeX, the same language used by the publishers. Many journals allow submission of articles already formatted this way.

I also want to make it clear that I am not condemning all those employed by all publishing houses. I have interacted with many excellent staff at publishing houses that have some of the worst practices. It’s important to note that these staff do not get the fiscal benefit of the aggressive and unethical behaviour of their publishing companies. These profits are retained for the directors and shareholders. One might also question whether the sales reps of these companies who make deals with academic libraries don’t also receive unethical bonuses for their undisclosed financial arrangements. But as these deals are not disclosed, who would know or admit this?

### What is publishing then?
Once publishers have ‘printed’ the manuscript, they ‘publish’ it by placing it behind a paywall on their website. I would be the first to admit that there are massive costs in doing this properly, and big journal companies have invested a lot to do this very well. The electronic hosting of journals is (in my opinion) truly excellent, except for that paywall. However, once they’ve set this system up, adding another 10 or 20 journals comes at practically no cost compared to the revenue that each one can be expected to gain.

To get behind that paywall, university libraries need to subscribe. Publishers bundle journals together and sell subscriptions at very high prices. If you are inside the university IP address, this access should be seamless. If you are outside, you might need to log in through your university’s library. There are other tactics for getting around a paywall ([see part IV](#paywall)).

So far, the publisher hasn’t produced any content. The scientific content has been produced by the scientist at the cost of the public purse. The editing and peer review ([see part IV](#preprint)) has all been done by the scientists, which has also cost the public purse but has been completely free to the publisher. OK, so there are some small costs associated with manuscript handling software subscriptions that the publishers normally pay. The publisher has also paid for the typesetting (although they’ve done this as cheaply as possible - see above), and they’ve paid for the servers that distribute the pdfs maintaining that all important [paywall](#paywall). What else? Nothing else. Now they simply maintain the system, and charge everyone to look at the content (and because it’s by subscription), actually charge everyone whether or not they are looking at the content. Bergstrom and Bergstrom [-@bergstrom2006economics] showed the difference between the costs of subscribing to society journals in 2005 was USD 0.29 while for-profit journals cost US$ 1.42, irrespective of any degree of quality. Fifteen years on, most society journals have now been captured by for-profit publishers, so that now we all pay the higher price.

But don't take my word for it. Read the excellent article by Stephen Buranyi [-@buranyi2017staggeringly]: Is the staggeringly profitable business of scientific publishing bad for science? In it, Buranyi makes the point that the profit margins of academic publishers are in excess of 40%. Something that even drug dealers, pimps and the mafia struggle to achieve. This situation seriously needs to change.

## Making the change {#change}

My message, throughout this book, has been that the tools to make the shift from [closed to open science](#openscience) are available to us now. There is nothing to stop us changing the system other than the need to act collectively.

[Scholarly societies](#societies) offer an excellent opportunity to organise. Some societies still dictate terms to publishers (for example, [The British Ecological Society](https://www.britishecologicalsociety.org/), and they have the power to finish their contracts and move their journals to new [diamond Access](#diamond) formats, using tools such as [overlay journals](#overlay). In turn, they have to give up the income that their for-profit publishers supply. This does not mean that they would lose all income, they would still have subscription funds and conferences. But there would be a loss of some of the comfort that they have become used to, making giving up hard to do. But societies need to carry out the wish of their members, and I would suggest that there are more members that would benefit from [diamond access](#diamond) society journals, than currently see any benefit in the kick-back from for-profit publishers.

### 2021-09-05 4.02-Howpeerreview.Rmd
Add files via upload
# all round update

# How to conduct peer review {#howpeerreview}

Here I will assume that you have been given a peer review assignment. If you aren't getting asked to do any reviews, there are some suggestions on how to get started in [Part I](#reviewit).

Most modern journals provide reviewers with a guideline for the expected to do. I encourage you to read the specific instructions that are given by journals on how to conduct peer review for them. There are also a number of excellent blogs to read about peer review [including this on @raff2013how]. A systematic assessment of these requirements in biomedical journals has been undertaken by Glonti et al. [-@glonti2019scoping] and Eve et al. [-@eve2021reading]. These accounts are worth dipping into for an overview on the different sorts of statements that peer reviewers come up with. You can see, in this quantitative analysis of Eve et al. [-@eve2021reading], that the overwhelming number of comments are those of skilled critics. This paper also makes it clear that the role of peer reviewer is often ambiguous and that reviews are not consistent in what they deliver.

Essentially peer reviewers are tasked with determining whether or not the manuscript is credible.

- Could the study be repeated?
- Are the methods legitimate in order to produce the results provided
- Are the results sufficient to respond to the hypothesis posed?
- Can it be improved?
- Is the content of the manuscript appropriate to the journal?
- Does the experimental design contain sufficient controls?
- Did the authors try and stretch the implications of their results beyond the credibility of the manuscript?

Once you have conducted your peer review, you can log it [publons](https://publons.com) in order to get credit later on. Publons also carries your publication output and citations (tied to Web of Science), so can be a useful way of keep track of your own productivity for reporting purposes ([but see this section on Impact Factor](#impactfactor)).

## Novelty or repeatability?
At the heart of scientific enquiry is that studies done should be repeatable, with the presumption that if they have been done in the same way they should achieve the same results [given the bounds of significance testing - see @forstmeier2017detecting]. Hence, you must examine and report on whether or not the study's materials and methods are sufficient for someone else to repeat the study. This appears to be surprisingly rare in peer review [at ~4% according to @eve2021reading] For some [no impact journals](#noif), this technical soundness will be enough to allow them to pass peer review.

Many journals that aspire to increase their [Impact Factor](#impactfactor), ask for a comment on how novel the study is. This is a somewhat subjective question, as individuals have biased opinions of what constitutes something novel, noteworthy, of significance or of relevance to the audience of a particular journal. Hence, this is really going to be a point that you can decide based on your own understanding of the literature (remembering that you have been chosen because your opinion counts).

You can gain important insight into what is relevant, and what not, in peer review from the analysis of PLOS ONE reviews by Eve et al. [-@eve2021reading].

## Parts of your review
Just as we discussed the different parts of a review when thinking about [writing a rebuttal](#rebuttal), here we discuss the same parts from the view of writing the review. I think that, in addition to reading this section, it is worth refreshing your memory about receiving a [peer review](#rebuttal), when thinking about writing one.

### A positive appraisal of the study
Summarising the study in your own words, to the tune of a single paragraph, is a useful way to start a review.

- It is usually positive being skewed towards what you understood, and what was well communicated
- Concentrates the minds of reviewers, making them think about the whole manuscript (and not simply focussing on minutiae)
- Lets authors know exactly what came across (and by omission what didn't)
- Allows editors to contrast your understanding with what other reviewers contributed, as well as their own interpretation. Usually different people find different issues, but overall the impression should be broadly congruent

If you have not understood everything, then you should concentrate on what you have understood to be reported in this section. Although this might set you up to produce a ['shit sandwich'](#shit), this is not necessarily a bad thing.

### Major comments
The next section concerns major positive and negative comments. Try to be even handed here. This is a place to point out any major short-falls of the manuscript, but it should also be used to point out where the authors have done a good job. You may need to resort to a list, where each major item gets it's own paragraph, but these may turn into sections if your reasoning takes longer.

For each major comment, give an example of what you mean with line numbers. These should be tangible points that you can tie down to things that are present in (or even missing from) the text. In this section, I would urge you to keep away from providing subjective statements (like "I think..." or "I feel..." or "It seems to me..."). If you need to voice these feelings, then keep them for a final paragraph where you make it clear that these are impressions given by the manuscript to the reader.

### Minor concerns
List these out under a new heading **Minor Comments** starting each one with a line number where it occurs. Figures and tables should receive their own comments (no line numbers required, but give the Figure or Table number).

## The spirit of peer review {#spirit}
In their heart, a peer reviewer should be trying their best to improve the manuscript they read as much as they possibly can. This may simply represent an improvement in the way the text is worded. But it may also mean adding extra analyses or even experiments (within the bounds of reason).

As McPeek et al. [-@mcpeek2009golden] put it, the golden rule of reviewing is to do unto others as you would have them do unto you. You could also read Baglini and Parsons [-@baglini2020if], who provide some useful insight into how to remain neutral when making reviewer comments. Again, the emphasis is on being professional.

## There are ethical considerations for reviewers
- Reviews may not share manuscripts with other scientists unless specific permission is given by the editor
- Similarly reviews should not discuss the content of manuscripts that they are reviewing
- Reviewers should not try to take the work presented in the manuscript and copy it for publication (i.e. do not steal the ideas)
- Reviews should be conducted within a reasonable time frame. No reviewer should hold on to a manuscript especially if they have a vested interest (like a rival study) in not seeing it published. This should have been declared as conflict of interest
- Any other potential conflicts of interest, including those that might make you positively predisposed to the authors, should be declared
- Reviewers should be aware of their own prejudices and biases and not bring them through to the review process 
- You must decide whether or not to sign your review. Given the opportunity ~43% of reviewers will provide published open reviews [@wang2016open]. Would you want your reviewer to sign?

In essence these ethical issues are overcome when reviewers conform to [transparency](#transparent). In order to facilitate transparency in peer review, Parker et al. [-@parker2018empowering] have produced a checklist that I encourage you to use if and when you are asked to conduct a review.

## Remain objective and rational
Your job as a reviewer will be to remain objective about the manuscript that you are reading, point out its merits and problems without succumbing to bias. Forming your own world view of your topic within the biological sciences does mean that you likely need a form of directionally motivated reasoning. For example, this is why you decide to investigate one hypothesis before another, or feel that one line of investigation is more salient to your area than another. These could be made through observations or experiences that you have had during your research, or they may come from schools of thought within your discipline. But it is important that you remain **intellectually honest**, to allow others to hold alternative, valid arguments. Just as it is important in your own work that you are always prepared to accept the null hypothesis as readily as you do the alternative hypothesis. One lesson revealed from reading lots of peer reviews is that reviewers find it hard to remain centred using accuracy motivated reasoning, all too often resorting to attacking the authors or their experiment [@eve2021reading]. Your principle task it to remain intellectually honest in your review, such that you can point out faulty arguments without perverting the direction that the authors planned to take. Equally, it is important that the authors acknowledge alternative viewpoints, but not to the extent that they should be made to abandon their own interpretation.

Psychologists have argued that as humans we cannot be expected to be rational, and that we are not particularly good at being objective. But this doesn't mean that we shouldn't try, and it also means that by being aware of the [potential problems in peer review](#problempeerreview), we are in a better position to learn how to avoid them.

**Write every review as if it will be public.** Ask yourself whether every statement that you make can be backed up either by other references in the literature, or with line numbers corresponding to erroneous logic on the part of the authors. Although this strategy is not guaranteed to produce an unbiased review, it will be an intellectually honest way to approach the manuscript. If you have the choice, then do make your review public to hold with the Open Communication ethos of [Open Science](#openscience).

## Remember to accentuate the positive
Peer review is often thought of as being brutal, where anonymous reviewers have the opportunity to vent their darkest thoughts. Certainly, there are plenty of reviewers who are unprofessional in what they say [@eve2021reading; @hyland2020this]. When conducting peer review, you have the opportunity to be one of the goodies. You can point out where the authors have done due diligence, in their experimental design, reporting, analyses, etc. This is likely to benefit the authors far more than pointing out only the problems - especially those that cannot be fixed.

## How long should your review be?
Quite simply, you need to write enough until you have reviewed the manuscript. The length of peer review varies wildly, from 200 characters to 43000 (likely more than the article itself), according to Eve et al. [-@eve2021reading]. The distribution peaks between 2000 to 4000 characters, and this should be a good guide.

If you submit your review to a service like [publons](https://publons.com), you can compare the lengths of your reviews get charted against those of the 'average reviewer', and I would suggest that you should aim to keep your reviews above average length when the extra words are used to help the authors. There's no point in just writing extra words to increase your character count!

## What to do if you suspect fraud?
The [Committee on Publication Ethics](https://publicationethics.org/) (COPE) has published some useful flowcharts to guide reviewers who suspect fraud in manuscripts they are reading. A list of these is provided [in Part IV](#fraud).

## Further help with conducting peer review {#onlinecourse}
A number of publishers and academic institutes have provided online resources to help train those undertaking peer review. I provide links to some of these here:

 - [ACS Reviewer Lab](https://www.acsreviewerlab.org/)
 - [Publons Academy](https://webofscienceacademy.clarivate.com/learn/signin)
 - [Nature Masterclass](https://masterclasses.nature.com/focus-on-peer-review-online-course/16605550).

Remember that these are suggestions, and should provide sufficient instruction to get you started. Not all journals ask the same of their reviewers, and so instructions may vary. Your review should follow the recommendations provided by the journal that you are providing the review for.

### 2021-09-05 4.03-problempeerreview.Rmd
Add files via upload
# all round update

# The problems with peer review {#problempeerreview}

There is already a lot covering peer review in this book, and I have placed this chapter last not because it is the least significant, potentially it is the most significant, but because I think that it is important that you appreciate exactly what peer review is, and to some extent experience it, before you begin to consider the problems with the peer review system.

At the heart of the problems with peer review is that individual humans are themselves biased. Because peer review relies on a small number of individuals providing their assessment of a manuscript, it is quite likely that these biases might align, and that the manuscript is rejected along those lines, rather than being considered along purely objective lines. This likelihood of aligned prejudices comes about because the pool of people that conduct peer review in biological sciences, and in many other disciplines, is mostly white, western (i.e. Europe and North America) and male. These people hold a very similar cultural set of biases.

Some people have argued that peer review is untested and that the effects are uncertain [@jefferson2002effects]. Perhaps more worryingly, studies designed to test peer review (by deliberately sending out manuscripts with errors) have shown that most reviewers are unable to find all errors and some find none [@rothwell2000reproducibility].

If peer review was effective, then reviews of grant applications should closely align with the productivity of grants given. Fang et al. [-@fang2016nih] found that percentile scores awarded by peer review of NIH grant applications were poor at predicting the the productivity of >100 000 grants awarded.

Essentially, the major problem with peer review is that it is conducted by humans, and that like humans in societies everywhere, reviewers tend to have their own set of biases. The above sections should have given you some idea about the frailties of the peer review system.

## Upsetting comments

I think that the reason why we find reivew comments so harsh is usually because we put so much effort into the writing process that it feels very personal whenever we receive criticism. Indeed, I think that there might be a correlation between how much effort you put in and how harsh the reviewers’ comments seem. Another study suggests that authors consider the competence of their reviewers to be closely aligned to the editorial decision [@drvenica2019peer]. Just be aware that this is normal. Remember that the reviewers are humans, and they have sat down and given freely of their own time to read your work. The most important thing to be aware of is that all they had was what you had written. No background information, and possibly no information about the species or the system involved. They will be experts at some level, but perhaps not the type you might expect. Importantly, the editor asked them because they thought that their opinion would be of importance in helping them make their decision on your paper. This means that you also need to respect their opinion and comments, even if you don’t agree with them or find them to be offensive, arrogant or even rude. Remember also that some apparent rudeness may just be a reviewer who has a sense of humour that you don’t understand. There are lots of examples of this at [ShitMyReviewersSay](https://shitmyreviewerssay.tumblr.com/). So no matter what you think of each comment, you should respond to it in a professional and courteous manner that shows that you are a professional scientist.

Why do scientists make disparaging or unprofessional remarks to their colleagues in peer review? Whenever two or three scientists get together, you hear tales of recent woes associated with peer review. The retelling of such stories is all part of the collective, cathartic unburdening of what can be a traumatic experience especially when we put so much effort into each piece of work [see @hyland2020this]. Reading through a lot of these reviewers’ comments, I can see that there is an attempt at humour. This humour is not appreciated by those who receive the reviews. Perhaps I understand the humour, because I also come from that same culture that dominates STEM, but that is not understood or even recognised as humour by others. Writing humorous reviews is unprofessional, especially if it is used to accentuate negative aspects. Needless to say, we could all do without unprofessional reviews.

## _Ad hominem_ attacks

One of the shocking results of a very large study of peer review of _PLOS ONE_ articles is the large number of comments that are written directly attacking the authors as a group or personally [i.e. _ad hominem_ attacks, see @eve2021reading]. This should not happen. Reviewers should be confining their objective comments to the work and its presentation. However, this is an aspect of peer review where authors (especially the corresponding and leading authors) will need to acquire a thick skin, because unprofessional comments are made to people across gender and racial groupings, but especially toward traditionally underrepresented groups [@silbiger2019unprofessional]. Sadly, these same groups feel that such comments disproportionately impact their productivity and career advancement [@silbiger2019unprofessional]. Reading comments that are sent to other authors can be cathartic as these allow you to see that everyone receives such negative comments. [ShitMyReviewers](https://shitmyreviewerssay.tumblr.com/) is a good source of these, or see Eve et al. [-@eve2021reading], or Silbiger and Stubler [-@silbiger2019unprofessional]. When _ad hominem_ attacks are made, it would be good if editors openly and explicitly identified these as **bad behaviour**. It would certainly improve the understanding of authors if editors intervened when such _ad hominem_ attacks are made. This would not necessarily involve deleting these comments, but directing authors to ignore the same.

**Why do academics make all of these terrible comments?** I can't pretend to know the answer for all of the cases, but I can speak from personal experience. Time is at a premium, and time spent reading and reviewing manuscripts tends to be quality time - best when it is quiet and uninterrupted. If these manuscripts are not of a quality that will pass peer review (i.e. will be rejected), then this feels like an abuse of professional time - especially when editors should have spotted the same mistake in their [first reading](#editors). Editors that fail to see manuscripts that [should be rejected](#desk) do the reviewers a dis-service by increasing the amount of work for everyone (more people and more time is involved). Resentment and frustration may follow on the part of reviewers that manifests itself in the form of _ad hominem_ attacks.

## Demonstrated biases in peer review

Although Table \@ref(tab:Bias) shows that many kinds of bias have been explicitly demonstrated, that’s certainly not their limit. Given that over 280 biases have already been catalogued (I encourage you to look through the [online catalogue](www.catalogofbias.org)), many more different types of bias are likely to exist in peer review. Let’s not forget that our biases have evolved because they are very useful. They exist as a way of shortcutting exhaustive decision making based on random variables. But maybe peer review needs some more of this. And perhaps that means that I should be tolerant when I’m asked to review an economics journal, as these folk clearly weren’t exhibiting any biases associated with economists when they picked me (see [section on editors](#editors)).

Table: (\#tab:Bias) **There are as many biases in peer review as there are humans that conduct them.** This table demonstrates some of the biases that have been proven in studies.

| Bias for which there is evidence | Study demonstrating bias | 
| - | - | 
| Against female authors | @tregenza2002gender; @manlove2018authors; @fox2019gender; @budden2008double-blind; @morgan2018foundation; @hagan2020women |
| Against female reviewers | @helmer2017gender; @fox2019gendera |
| Towards author reputation, favouring acceptance of manuscripts despite poor reviews | @bravo2018hidden; @okike2016single-blind |
| Towards authors from more prestigious institutions, also called prestige bias | @ceci1982peer; @travis1991new; @garfunkel1994effect; @tomkins2017single; @manlove2018authors ; @lee2013bias |
| Nationality and language bias | @song2000publication; @lee2013bias; @manlove2018authors; @nunez2021monolingual; @link1998us |
| Confirmation bias (the tendency for journals and reviewers to favour significant results) | @mahoney1977publication; @fanelli2010positive; @fanelli2012negative; [see Part I](transparency.html) |
| Publication bias (the literature contains a bias in published results) | @jennions2002publication; @munafo2007association; @dongen2011associations; @franco2014publication; @fanelli2017meta-assessment; @sanchez-tojar2018meta-analysis; [see Part IV](#badscience) |

Perhaps the biggest problem facing those who wish to reform the peer review system is that it all starts with editors who are choosing reviewers. Those editors themselves have their own inherent biases. When they look for reviewers, they are likely to sample from within their own group of peers who have the same biases. Interestingly, bias (in general) is more easily perceived by early career scientists [@zvereva2021biases]. My experience is that soliciting reviews from people that I don’t know and have no connection with (are outside of my field) are more likely to fail - they will say no, or they won’t reply to the request [see @perry2012peer]. This is even for academics that are publishing within the same area.

Editors are the people who select reviewers, and inspection of most editorial boards will reveal that they reflect the same biases found in peer review. That is, editorial boards are mostly made up of white men from Europe and North America. Rectifying this bias will take time and the acknowledgement that there is a problem together with the willingness to do something about it. In 2020, I have seen that there has been a big movement to redress the imbalance in science at all levels. I hope that this will continue into the future so that at least some of the biases in peer review will fall away.

## All reviews are not equal
If you are an editor and you receive three reviews from three researchers each suggesting something different, I have argued ([in another section](#editors)) that the editor should make their own decision on what action to take. But what if one of the reviewers is very negative and is a leader in their field? Should their review count equally with the others? Should their opinion be given more weight than the others? Of course, they could be using their position to influence their field, to make sure that opinions they hold are reinforced. Lee et al. [-@lee2013bias] provide a good overview of the potential way in which influential reviewers could bias the peer review system. But the power sits with the editor to make this decision. Interestingly, Thurner and Hanel [-@thurner2011peer] make the point using an agent based model (much as you might use in biological sciences) to show that only a small number of biased (for whatever reason) reviewers are needed to seriously degrade the quality of peer review, and thus the science system as a whole.

The truth is that all reviews are not equal because some reviewers will put in more effort than others. Some will know the literature better. Some will be experts in the field that should be better placed to comment. These people are actually more likely to be less senior, PhD students, post docs or early career researchers. However, the importance for the editor is not to take account of the names of these people, their rank, their institution, or other demographics such as their gender, race or nationality. There are great editors out there who can do this, but my impression is that the majority fail. In this case, the only way to do this is for the [triple blind method](#triple). Here the editors will invite the reviewers (by name) but the reviews that result will not be marked with the reviewers’ names. This will make forgetting who they are easier, especially for busy editors.

## Decisions rest with editors
A good editor will look at the reasoning in the reviews and make a decision in an unbiased way. A poor editor may be swayed by the perceived influence of an important reviewer irrespective of their argument. An increasing trend that I’ve noticed is that editors will simply take a decision that follows the consensus of all reviewers: that is, they rate all reviewers equally [see also @rothwell2000reproducibility]. However, I would argue that this is also bad editing. Irrespective of the bias from reviewers, guarding the integrity of the process of peer review lies with editors.

Today, editors are so busy with the other duties in their jobs as academics that their decisions are hurried and expecting them to take the time and space to overcome their personal biases might be a lot to ask. Instead, I think that it is time for the [triple blind review](#triple) concept to move into the mainstream so that editors can more easily not be led by potential biases of their reviewers, and instead be swayed by the quality of the review and their own reading of the manuscript.

Another important problem with peer review comes when editors are not independent of authors. This can happen when an editor is known well by the authors. They could be in the same department or even in the same research group. Similarly, there could be a group of editors for different journals that have some _quid pro quo_ arrangement, that might even be unstated, whereby their manuscripts do not undergo equal scrutiny to other manuscripts that are submitted. One could argue that whenever editors know the names of the authors, there is a conflict of interest that should be declared or the possibility for the system to be corrupted.

Despite all of the problems with peer review that are acknowledged above, we stick with it as the majority system in science. It could be that peer review favours exactly the same people who uphold the system and prevent it from moving into something more transparent, equal, just and fair. These are the editors and reviewers who have, for the most part, managed to make their careers inside the system, and have therefore mastered it to some degree.

To you, dear reader, I can only suggest that you be aware of all the potential pitfalls with peer review, and never stop striving for something better.

## The social side of peer review
There is so much more to peer review than peer review. Being selected by an editor to review a manuscript represents an important standing amongst your peers. Literally it means that your opinion is valued. But there's much more to it. Doing a good job at peer review means that you improve other people's work. This help can be valued to the point where those colleagues get in touch and want to work with you. That this can happen has now been shown in a study, and has been termed the 'invisible hand' of peer review.

### The 'invisible hand' of peer review
Dondio et al. [-@dondio2019invisible] found that reviewers were more likely to provide positive review comments to authors who were close (less than or equal to three steps) to their collaborative networks [see @adams2012rise]. In this case, a close reviewer to the author was calculated by a social network where a distance of one meant that they had co-authored together (one step), co-authors of the reviewer may have collaborated with these authors (two steps), or co-authors of reviewers and authors had collaborated (three steps). Surprisingly, they obtained this result even though the journal practised a strict double-blind review system (reviewers didn’t know who the authors were, and vice versa). Referees that were not close (i.e. greater than or equal to four steps] were more likely to provide more negative review comments. Those who helped authors more during peer review (i.e. asked for major revisions), were more likely to cite the manuscript, once published, and eventually more likely, than random, to publish with those same authors, even if manuscripts were eventually rejected. The authors concluded therefore that peer review may accelerate the potential for collaboration in science [@dondio2019invisible].

This finding appears to be based on the fact that peer review can/should be constructive. Authors and peer reviewers are in fact collaborating to improve the quality of a manuscript. The process is orchestrated by an editor who can and should join in to improve the manuscript. Dondio et al. [-@dondio2019invisible] make the point that this interaction is inherently social, and the peer review therefore has a function that develops relationships within and between [networks of researchers](#network).

This evidence that peer review is a collaborative system towards the betterment of science is, to me, a sign that all is alive and well in science, and that peer review is acting as it should. However with any social network comes the fragilities of human bias. This means that while peer review may function well for some, for others it may more often than not fail. The bigger problem is that it might depend on your sex, the colour of your skin, the name of your institution, or your country as to whether you are selected as a potential reviewer (i.e. to join the club; see Table \@ref(tab:Bias)), or having submitted your manuscript, find that peer review is going to work for you [see @davies2021promoting]. In addition, if you are never asked to review then you will never benefit from this network.

Casnici et al. [-@casnici2017assessing] tracked the fate of rejected manuscripts and showed that if the reviewers had several rounds of peer review before rejection, these manuscripts benefited later by being accepted to journals with a higher Impact Factor, and/or obtaining greater numbers of citations, even if the reviewers were instrumental in rejecting the manuscript. This suggests that in working collaboratively on a manuscript, reviewers are more likely to promote, cite and help authors. The alternative is that reviewers agree to re-review an article again because they see merit in it, even if they also see flaws. And having spent considerable effort reading the manuscript, they are more likely to remember and cite it. But this doesn’t take away from the idea that reviewers and authors are collaborating in a social way.

## Fixing peer review {#fixing}
Fixing peer review will rest with the community of biological scientists, at the level of the gatekeepers: editors and the scholarly societies that they represent. To me, it is clear that we won't fix peer review by asking our peers to be less biased, or by asking them to be more rational. We should know by now that we can't fix people in this way. For example, Khoo [-@khoo2018there] found that there was little improvement in reviews after [reviewer training courses](#onlinecourse), even when these included feedback on previous reviews submitted. Instead, we have to plot a course for peer review whereby we accept that reviews will contain bias and irrational content, and train those in editorial positions to try to spot these, instead of falling victim to them.

There are lots of ways in which to do improve editorial oversight. My intuition is that the crux is to find a way that makes it more efficient and objective for the editors. For example, to try to pin down reviewers on where they find fault and exactly what that fault is. There is a difference between:

 - insufficient information to decide whether the experimental design was faulty and needing this clarified before a decision can be made
 - finding a fundamental error in the experimental design such that the manuscript can be rejected
 - insufficient power (in replicates or sampling) to reach the conclusion generated in the manuscript

A manuscript having each of these outcomes should have different fates: The first is Reject & Resubmit, the second is Reject, while the third may warrant either major/minor revisions (depending what else is problematic), or movement to another journal.

However, because peer reviewers represent a minority, this means that at times prejudices and biases will align, a more inclusive world might mean that they diverge, prompting more differences in opinions about what should happen to manuscripts. Given that it is already quite difficult to find enough reviewers, simply asking more reviewers won't fix this. Instead, we need ways in which editors can more easily come to decisions on  manuscripts taking into consideration the potential faults. This really entails journals being more transparent about what flaws in manuscripts will be considered fatal. For journals where methodological competency is all that's required (see here for [commiment to publish](#commitment)), this is simple, but for many more journals (particularly those that are important for Early Career Researchers because advancement in their careers will depend on publishing there), these will be more ill defined, editor-centric choices that are more about fashion in Biological Sciences, than good science _per se_. Removing the systematic biases is important, and something that is well worth fixing.

Hence, fixing peer review comes back to fixing problems associated with the publishing culture (and all that that entails), rather than any a simple fix-all for the myriad of existing publishing options. [Preprints](#preprints) combined with [Overlay journals](#overlay) offer one solution that keeps reviews with the original submission. This stops that practice of authors resubmitting their manuscripts over and over to countless journals until a pair of reviewers fail to spot a fundamental flaw. Keeping these manuscripts with their reviews on BiorXiv represents one solution, together with another problem that good articles with biased reviews still need to overcome. I feel that this is more likely to come right when authors resubmit to another Overlay journal with a valid rebuttal.

Lastly, but perhaps most importantly, we have to be more realistic about the limitations of peer review. We would be better to think of peer review as a '**silver standard**' - something that some scientists agree has merit. Our problems come because our expectations of peer review are too great - a 'gold standard' it is not. Instead, we can keep the 'gold standard' for those papers that have withstood the test of time, the repeatability of the community, and acceptance into the mainstream.

### 2021-09-05 4.04-predatory.Rmd
Add files via upload
# all round update

# What are predatory journals? {#predatory}

Predatory journals are publications that purport to be from scholarly publishing houses, but have little or no editorial oversight or peer review. They exist in order to extract the Article Processing Charge (APC) that is so ubiquitous in [Open Access](#openaccess) journals. They continue to exist because publishing is moving towards APCs, and there is little difference between what they do and some supposedly ‘legitimate’ journals. For example, some definitions of predatory journals include that their APCs far exceed their publishing costs, but this can now certainly be said of some legitimate journals.

## If the line is so grey, how is it possible to tell whether or not a journal is predatory?
This is a surprisingly difficult question to answer. Predatory journals have become so sophisticated at what they do, that it can be very difficult to determine whether or not they are legitimate. Moreover, their electronically published journals can create new titles faster than the time that it takes to check that they are legitimate. An example of a publisher in the grey zone is Hindawi, that were once considered predatory but were later removed from Beall’s list. At the time of writing, I considered that Hindawi still had some questionable practices and so suggested continuing to avoid them. However, in a new (January 2021) twist, Wiley (a well respected publisher) [acquired Hindawi for USD 298 million @michael2021wiley] (yes folks, your science is big business because of the big profits that Wiley expect to make out of you and your tax paying funders). This news goes to show how quickly the publishing world is reacting to the new world of Open Access (limitless numbers of titles), but also how some publishers intend to dominate the market and maintain their stranglehold on scientific publishing.

Most academics have an interaction with predatory journals through email. Indeed, the numbers of emails that academics are spammed with is astounding [@vandenberg2021solution], and while the best advice is to use a spam filter, I still find that too many legitimate emails go there too. These spamming emails from predatory publishers are not straight forward and seek to manipulate the reader into submitting a manuscript [@bett2020predatory]. Moreover, predatory journals target the most vulnerable in the community, poorer researchers who do not have English as their first language are deliberately targeted [@bett2020predatory, @lund2020analysis].

At the same time, legitimate journals have become increasingly predatory in their habits, and it’s difficult to tell them apart from predatory journals. There was a time when it was possible to categorically state that no journal will ever approach you with a general email that invites you to contribute. However, there are now several legitimate journals that send out unsolicited emails. Realistically then, in the current publishing world, there is a continuum from predatory to legitimate. It was not always this way, and that means that as an early career researcher you are facing difficulties not faced by your mentor or other more senior academics. Not only do you need to avoid publishing in predatory journals, but you should also avoid citing their articles.

However, help is at hand. There are some definite ways that you can determine whether you are choosing a legitimate journal. Here are my **five steps to take to safeguard your submission against a predator**.

### To spot a predatory journal, use the following list in a stepwise fashion

1. **Use an index**. Web of Science and Scopus both curate contents of legitimate journals. If your journal of choice appears in one of these, then it is very likely to be legitimate. Note that Google Scholar includes many predatory journals, so please never use this to determine whether or not a journal is legitimate. Note also that it takes a journal several years to gain enough kudos to get accepted onto Web of Science and/or Scopus. Therefore, it can still be legitimate and not be there. How to choose the right journal for your publication is covered in an earlier section ([Part I](#whichjournal)). If the journal you want to publish in is not in Web of Science or Scopus, then proceed to the next step.
1. **Ask your librarian**. Librarians are fantastic sources as well as custodians of information, and journals are one of their key knowledge areas. Don’t hesitate to get in touch with your librarian and ask their advice. They are likely to be very well placed to respond to your request. They may also be guardians of granting APCs at your institution, so it is in their interest to make sure that these valuable monies don’t fall into the wrong hands.
1. **Ask your mentor or an experienced colleague**. It’s worth doing this with them so that you can see the steps that they follow. Given that steps 1 and 2 have already come back with uncertain answers, spreading your net more widely will help with step 4. However, be warned that there are increasing numbers of senior scientists that have been caught out by predatory journals, so checking their contributors is now no cut and dry way to differentiate between them and legitimate journals.
1. **Who is on the editorial board?** Journals publish names of their editors, associate editors and the editorial board. Look through these lists and see whether there are names that you recognise. If you know any of the people, you (or your advisor) can contact and ask them about the journal (they should be happy to respond). Be warned that it is easy to place someone’s name on a website, so unless they have personally told you, keep away.
1. **Check against a known list**. In the past, this might have been the first thing to do, but the number of predatory journals is proliferating so quickly that it’s hard for any list to keep up. Beall’s list retired in 2017. The next best list now has more than three times the number of journal names. A new list, by Simon Linacre, is sadly behind a paywall, so you can’t expect to access it. One of the reasons why Beall gave up is that the new tactic for these publishers is to produce lots of new journals. Curating a list is real work and has implications for the publishers on it, hence you now need to pay to access an up to date list. There are more lists: Cabell's Predatory Journal Blacklist, Beall’s list, DOAJ delisted journals, Scopus discontinued sources, etc.

### If predatory journals are becoming more like legitimate journals, where’s the harm in publishing with them? 
Your reputation is important. As an early career researcher, your publishing record is what many people will look at first. It is all that is shown in your ORCID, Google Scholar or ResearchGate profile. It’s your shop window or showroom. What prospective employers will want to see is that there are plenty of publications (appropriate for your career stage), and that they are in appropriate journals with good reputations. You might confuse having a good reputation with a high impact factor. The two need not be the same. High impact factor journals don’t accept all types of submission, and you may have data that simply doesn’t fit into one of their mandates. I would say that it’s still important to publish, and there are many journals with good reputations where you can publish (discussion about [impact factor is in another section](#impactfactor)).

The other reason why you would be best to avoid a predatory journal is that they attract very little in the way of scientific impact: few people will read or cite them. One thing that you definitely want for your work is for people to use it. To do this, they must read and cite it. If you publish in a predatory journal, many scientists won’t even consider reading the content as it has not been, nor will it be, peer reviewed. Thus, unlike a preprint, it is not being openly offered to the community for review.

Due to the ambiguity of whether or not these papers have been peer reviewed, I would also suggest that you do not cite publications that you think may be from predatory journals. You can use the same steps (above) to determine whether or not what you want to cite is from a legitimate journal.

## What do you do if you have already published in a journal that others consider predatory?

The first step would be to write to the publisher and withdraw the article. Whether or not you paid an APC, having it on their website is not good for your reputation. Beware, these journals don’t adhere to an ethical code, and so they might refuse to withdraw your paper. Or they may want to charge an additional fee to remove it (remember that they are in it for the money).

Do not cite the paper, or put it on your CV. You can easily remove such articles from your ORCID, Google Scholar or ResearchGate  profile. Don’t put it in your showroom.

Prepare a statement that explains how it happened. You may not have been responsible for the submission, or aware that the publication was from a predatory publisher. However, in time you are likely to forget the exact reasons. It would be a good idea to prepare a statement, so that if you are asked (for example in a job interview), you can explain how it happened. People can be very understanding when provided with an explanation, but if you say that you can’t remember or can’t give any details, then you may sound evasive.

It’s not just publishing where the predators are lurking, they are also waiting to invite you to a conference. Your work is valuable to you and to your lab, so please try to make sure that it doesn’t end up in the hands of a predator!

### 2021-09-05 4.05-paywall.Rmd
Add files via upload
# all round update

# Why did some journals go behind paywalls? {#paywall2}

Academics are not (usually) superstars, nor looking for enormous numbers of readers, but there would be little point to recording our work if we had no readers, or if our work were inaccessible, and so publishing is a necessity. However, we have got into a state in which much of our work is behind a paywall, and thus inaccessible to most would be readers. Whether or not we need our work to look pretty and appealing speaks more to our readers as humans than academics. Perhaps it goes without saying that an audience is likely to be larger the more appealing the presentation, and that’s not just the writing style but the layout and presentation of the text itself. And this is not new.

The first scientific journals appeared in 1665: _Journal des Sçavans_ (Figure \@ref(fig:scavans)), and three months later _Philosophical Transactions of the Royal Society of London_ (Figure \@ref(fig:PhilTransRoySoc)), which is still in print today. It is clear that the papers were type set and presented in the manner of a book, perhaps analogous to a collection of short stories. At the start, these were reports of studies that were presented at meetings. Producing proceedings of learned societies became the way in which most scientific journals began. Only later did it become possible to submit a manuscript that had not been presented. And later still when publishers began to manufacture their own scholarly journals in the absence of any academic society.

(ref:scavans) **The first issues of the World's first scientific journal**. The _Journal des Sçavans_ ceased publication in 1792 due to the French Revolution.

```{r scavans, echo=FALSE, out.width = '50%',  fig.cap="(ref:scavans)"}
knitr::include_graphics('figures/scavans.png')
```

(ref:PhilTransRoySoc) **The world's longest running journal**. The _Philosophical Transactions of the Royal Society of London_ continues to this day as the world’s longest running scientific journal.

```{r PhilTransRoySoc, echo=FALSE, out.width = '50%',  fig.cap="(ref:PhilTransRoySoc)"}
knitr::include_graphics('figures/PhilTransRoySoc.png')
```

Being the editor of a society journal means being elected by members of that society, and being responsible to an editorial board, normally made up of the society’s members. Until very recently, and I’m thinking back to my first interactions with editors for my first few publications, submitting to the journal meant producing three (or sometimes more) double spaced hard copies of a manuscript and mailing a large and heavy envelope to the editor. Editors of bigger journals had secretaries dedicated to handling the administration of the paper. Following a telephonic enquiry, the copies were sent out to referees by post and sent back to the editorial office with a typed report often together with the marked up manuscript. Once the editor had received all reports, they communicated their decision back to the corresponding author (i.e. the one to whom correspondence was addressed) and, once accepted, the article went into production. Prior to personal computers being commonplace (only 25 years ago), each journal would have had to have had a publisher to set the type and print the pages. Clearly, this was beyond the scope of individual societies and the publisher was a necessity. Libraries had to pay for copies of journals, as the cost of publishing had to be offset by the society.

The advent of desktop publishing changed the need for publishers and brought many small society journals onto a larger platform where they could produce attractive content (over the typewritten documents that had been stencil duplicated or similar) and sent around to members. However, for small societies there was no professionalism involved as it devolved to the editor to publish society material. This is where I entered the stage in 2009 when I took over as editor of the _African Journal of Herpetology_. Thankfully, email had taken the place of the postal service, but once a manuscript was accepted I was the one who needed to type-set the documents (in Quark) and send out proofs to authors (just as previous editors had done before me). Once proofs were corrected and the issue was ready, I had to find quotes from 3 printers, deliver discs and ultimately collect boxes of printed journals. Back at home, I also packed all of the copies into labelled envelopes (with some help from friends) and carried the boxes once again to the post office.

(ref:journal-video) Stuffing envelopes with journal copies seemed like an endless task. View [on YouTube](https://youtu.be/sJReMVz7Yxk).

```{r journal-video, echo=FALSE, fig.cap="(ref:journal-video)"}
if(knitr::is_html_output(excludes="markdown")) knitr::include_url("https://www.youtube.com/embed/sJReMVz7Yxk")
```

After the first issue, I realised that I could not do all of this work indefinitely. I knew that there were publishers who were interested in acquiring the journal into their stable, and I contacted them and started negotiations. In 2011, the first copy of _African Journal of Herpetology_ from a professional publisher emerged, and allowed me to go back to editing the content through peer review.

At the time, I was aware of Open Access and considered this as an option for the journal. Open Access would have required that someone pay for the type setting, and the society would still have to pay for an online dissemination platform, given that they did not have their own platform. This would have meant that authors paid for getting their manuscripts into print. And then, like today, the decision was that our authors would not be willing to pay. Other societies did opt to follow this Open Access route, with the costs being covered by the authors. For some this became an incredibly successful model, with submissions increasing as well as the [Impact Factor](#impactfactor). They demonstrate that Open Access is possible on an independent platform.

## Why don’t all society journals do without publishers, and go Open Access independently?
The first problem is that societies generate income from journals. Subscribers to print or virtual copies pay the society, and this can defray a large part of the cost of publishing otherwise carried by the members. Some make a profit, and this profit can enable the society to do more for their community of members. This could include providing small grants, subsidising conferences for students and other much appreciated initiatives. Going Open Access means losing this revenue, as well as taking on the extra costs.

The second problem for many societies is that their members are paying, and the councils or committees elected to represent the members do not feel that it is fair for their members to pay for open access for everyone else. Part of the privilege of being a member of the society is having a free (or more accurately paid through membership) copy of the society’s journals. The costs are not high, and the exclusivity of members having subscription access while it is denied to others is perhaps just a hangover from the days when the only other copies were in the library. Certainly, the costs are nothing like those which authors are now charged by publishers to turn their accepted manuscripts into Open Access.

But the current situation is unsustainable. Tax-payers (in the main) pay for science to be conducted at universities and other institutes. In places where Open Access is mandatory, the tax-payer pays again to have the research published at a cost that is far inflated from the actual production costs. Publishers are getting rich, and the losers are the scientists (whose funding is reduced to pay publication costs) and the tax-payers who end up footing the bill.

## How do publishers make their money? {#money}

In the original publishing model, scholarly societies, such as The Royal Society, paid publishers to type set and print their publications (see Figure \@ref(fig:PhilTransRoySoc)). These in turn were bought by the society for their members and any libraries who subscribed to the society. As institutions and their libraries grew internationally, so too did the subscription base such that there was a profit to be made from the cost of the subscriptions for the paper content. Even back in these early Victorian days of scholarly publishing, publishers to started their own journals (aside from any society) as they recognised their superiority in distribution [@brock1980development]. This model continued over many decades with some publishing houses growing as they acquired more journal titles from societies because there was a lot of profit to be made. The peculiarity of scholarly publishing includes the unpaid nature of authors, editors and reviewers (all of the content), and the non-equivalence of the output: that is to say that just because you have access to articles in _Nature_, they are not equivalent to those being published in _Science_ - as a scientist you need to read articles from both of these publications.

Interestingly, the dawn of the internet was predicted as the downfall of the academic publisher Elsevier, because it promised the ability for academics to share their content for free [@lariviere2015big]. Despite this Elsevier's profit in 2012-13 was USD 2 billion, with a profit margin of ~40% [@lariviere2015big]. This profit margin is so large that some have compared it favourably with street gangs selling drugs [@buranyi2017staggeringly]. In the Biological Sciences, Elsevier, Springer-Nature, Wiley and Taylor-Francis control around 50% of all papers published [@lariviere2015big]. Selling to university libraries is no longer made on print copies, but on packages of electronic journals from each of these big publishers. The deals made by individual libraries were (and are) made in strict confidence, the exact sums are unknown, but the deals made fall into the hundred and thousands of USD for a set of publications that cover (for example) the science content of one of these publishing houses. The penalty for not subscribing is that your researchers reach a [paywall](#paywall) when they try to access a particular paper. You will find, as I have, the luxury of visiting a very wealthy institution and having access to just about every journal title you can think of. If you are not at one of these elite institutions, then you will face a demand for money when you try to access a title that your institution does not subscribe to.

The cost of accessing a single article comes at between 25 and 50 USD [@hagve2020money]. For me, and a great deal of other researchers, there are other ways to [get around the paywall](http://www.howtowriteaphd.org/paywall.html). But for some people, especially members of the public who simply want to access the information, they may simply pay the price. Imagine then the potential profit to be made from a society that has a large back-catalogue of content. There has been a scramble to sign up editors from society journals with tens of thousands of old journal articles at their command. The publisher offers to scan the entire back-catalogue and then offer it on sale. The society gets a paltry 10%, while the publisher fishes with tens of millions of articles. Even if they only sell 10% of these in 1 year, that makes their profit in the millions of USD.

In 2014, universities in Germany abandoned all subscriptions of journals by the Dutch publishing house Elsevier, as they couldn't keep up with the 30% price increase over the previous five years, making the average cost ~$4700 per journal per year [@vogel2014german]. Despite this, Elsevier continued to allow these universities access to their content for free, before negotiating another (secret) more amenable deal to them. Somehow the publishers had managed to turn the ease of distribution of articles over the internet to their advantage, via effective paywalls. But this was not the last of their master strokes.

[Open Access](#openaccess) titles came about in the early 2000s as the need for hard printed copies fell away and academics became comfortable handling all of their publications as pdfs. In the Biological Sciences, _PLOS ONE_ was the title that revolutionised the market, quickly becoming the single biggest journal. What caught the eyes of the publishers was the article processing charge (APC) that academics seemed to be prepared to pay. When they first started, _PLOS ONE_ cost ~1300 USD for its APC [@khoo2019article]. Publishers quickly started titles with similar [no impact factor](#noif) goals. Competitive titles began at the same cost as _PLOS ONE_, but once these titles became established, with higher Impact Factors, they quickly increased their APC [@khoo2019article]. Commercial publications (as opposed to non-profit journals) have become adept at pricing their titles toward what libraries and authors are prepared to pay, such that commercial publications are regularly more expensive than non-profit journals [@bjork2015article]. In their analyses of drivers of APC for OA journals, Budzinski et al. [-@budzinski2020drivers] found several variables that explained the vast difference in APCs of different journals. Impact Factor played an important role, but so too did market power of the publishers (presumably the ability to distribute the title and content), the hybrid model ([Gold OA](#gold) or [Hybrid OA](#hybrid)), and the concentration of disciplines that they cover.

Librarians analysed content that was [Open Access](#openaccess) finding that it was more highly cited, and so academics began to look for the potential to publish [Open Access](#openaccess). The publishers were happy to oblige, and in the early days conducted 'double dipping' charging both the APC for the authors, and collecting full revenue for subscription to hybrid journals by universities [@pinfield2016total]. It was only in May 2018 that [Knowledge Unlatched](https://knowledgeunlatched.org/2018/05/knowledge-unlatched-announces-its-fifth-pledging-round-ku-select-2018-2/) formed an alliance to prevent double-dipping in Open Access publications.

The end to double-dipping, as mandated by governments [@pinfield2016total], has led to even more complicated publishing deals known as **transformative publisher agreements** (Table \@ref(tab:profit). Unsurprisingly, these deals are said to lead to even greater profits for the publishers [@poynder2020information]. With these agreements, researchers from universities that have full subscriptions to certain publishers will receive free Gold OA for any journals in the package and their authors will be able to publish OA inside the same publisher's journals. In effect, the price of the OA is offset against the cost of the subscription. As you may have noticed, this will quickly start a bias in the publishing trends of academics from some (middle income) institutions publishing with certain titles, because they receive the benefit of OA. Those from privileged institutions will continue to publish anywhere they can, and the real losers will be those from poorer institutions and especially middle-income countries, who cannot afford transformative agreements, and cannot afford to provide their researchers with costs to cover APCs [@measey2018europe]. For these researchers, there will be a reduced set of journals that they can subscribe to. This is the current reality that increasing numbers of researchers find themselves in.

On the face of it, Plan S makes total sense [@schiltz2018science]. If all papers published by European researchers are Open Access, then everyone can read them and there is no more need for paywalls. Unfortunately, the way in which Plan S has been implemented has only been for wealthy European institutions. Others are left to continue to pay a hefty APC price for OA publishing, and their libraries pay a subscription to access journal content [@hagve2020money]. There is a real possibility that the USA and China will take other routes to OA, causing a schizm in publishing, more complex agreements and even more profits for publishers [@poynder2020information].

Another way that publishers make money is through [copyright](#copyright) on the content of their very large back catalogues. There are a lot of text books written for undergraduate students, and this is a large profit making publishing model in and of itself. When you look at the examples given in text books, they are often taken directly from peer reviewed papers. The publishers of the text book will pay for the content that they get from the pages of journals where they are under copyright. This can be very lucrative for the publishers who retain copyright from the authors. Of course, the authors get nothing even though it was they that created the content. The governments that funded the authors also get nothing. The advent of Creative Commons licensing has meant that it is now very easy to make the content of journals available for anyone to use. Happily, most Open Access journals have taken this route.

Table: (\#tab:profit) **How publishers make a profit**. Publishers are keen to move away from old models of publication fees, because their new transformative publisher agreements allow them to charge across the board.

 | Funding source | From society journals | From their own journals | From [Open Access](openaccess.html) | From [copyright](#copyright) |
| --------- | -- | -- | -- | -- | 
| Old subscription model (mostly obsolete) | X | X | - | - |
| Selling bundles to university libraries (undisclosed sums - millions of USD) | X | X | - | - |
| Selling articles of large back catalogues (~10% goes to society) | X | X | - | - |
| Article Process Charging | - | - | X | - |
| Double dipping  | X | X | - | - |
| Transformative publisher agreements | X | X | X | - |
| Copyright content  | X | X | - | X |

### How could such a profitable business lead to making a loss? {#loss}
To be fair, we should also consider that some publishers have not (always) been able to turn a profit, even when their turnover is in millions of USD. One such beleaguered publisher is PLOS. Despite growing exponentially in the early noughties, to become the world's biggest academic publisher in 2013, PLOS has not managed to turn a profit since 2015 [@davies2019plos]. At the time of writing (May 2021), the APC on a _PLOS ONE_ article was USD 1749, while other PLOS titles ranged from USD 2100 to 4000 (see [PLOS fees site](https://plos.org/publish/fees/)). This begs the question, how much does it cost to process an article? Data from the PLOS Financial Overview suggests that processing your article costs USD ~315, with another USD ~262 to keep it online.

So where does the rest of the money go? PLOS spend the bulk of your APCs on the editorial aspects of publishing. Remember that PLOS don't pay academic editors, but they do have huge financial burdens of the salaries of their own editorial staff. PLOS is based in San Francisco, California, which just happens to be one of the most expensive cities in the world to live in. Chief among their many options are to raise the costs of your APCs in order to pay their staff, or move their operation to somewhere cheaper. Interestingly, the other two large OA outfits, MDPI and Frontiers, are both based in Switzerland, possibly one of the most expensive countries in the world in which to pay salaried staff.

Most recently, PLOS announced a new funding model to keep themselves afloat that spreads the cost of their APC among the authors, and aims to cap profits at 10% [@else2021open-access]. The new scheme, Community Action Publishing, sees a return to the institutional subscription model where institutions will be charged proportionately at the rate their academics publish in PLOS. Great if your institutions can afford to opt in. If not, you can expect to face your proportion of the APC now as a co-author (instead of only the corresponding author paying).

Time will tell whether PLOS can make their new financial model work in order to keep their staff in the comfort to which they have become accustomed. No doubt we will see more of these new funding models as the increasingly bloated OA market attempts to consolidate itself, while still feeding on the funds of the global tax payer.

In case you hadn't realised, there is no advertising (at least in Biological Sciences - some medical journals do have advertising). The only time you see adverts is when you are browsing the content of Nature and Science. Oddly, this doesn't make them any cheaper to publish in, but presumably does increase the profits for these particular titles.

## What about academic books?
The story about the move from print to eBooks for academic books shows some even more remarkable disparities in the prices of eBooks compared to their physically printed paper counterparts (see Figure \@ref(fig:eBook-prices)). University libraries are required to buy versions that are specifically licensed for university use. The copyright law means that universities are not allowed to scan their paper copies and make these available. But in order to buy the university library copy, the price paid might be as much as 10 times that of the paper edition (Figure \@ref(fig:eBook-prices)).

(ref:eBook-prices) **The cost of eBooks compared to printed copies of the same books.** In crowdsourced data compiled by people in the Campaign to investigate the academic eBook market, you can see that most of the titles have a much more expensive eBook version. Note that in all cases the eBook licence is the equivalent of a single printed book: i.e. a 1 user copy. Data as entered on 17 May 2021.

```{r eBook-prices, echo=FALSE, fig.cap="(ref:eBook-prices)"}
if(knitr::is_html_output(excludes="markdown")) knitr::include_url("https://johnmeasey.github.io/eBooks-prices/index.html", height="450px") else knitr::include_graphics("figures/eBook-prices/eBook-prices.png")
```

More than this, publishers appear to inflate the prices of their eBooks without warning, by as much as 200% [@anderson2021campaign]. In addition, there are moves to make licensing annual for universities, with examples where once-off purchases have become unavailable as the publisher has moved them to an annual pricing model. While copyright law would protect you as an individual from any publisher that was undertaking such terrible behaviour, UK universities appear to be without recourse, such that they are now petitioning the Minister for Higher Education [@anderson2021campaign].

## What will it take to break the vicious cycle?
We need new models for publishing. Society journals are still kings in this game and ultimately hold the cards for moving away from filling the pockets of publishing companies. What we have seen in recent years is that journals can come from nowhere to become dominant players in the system. Think [_PLOS ONE_](#PLOS), and the even more recent meteoric rise of _Scientific Reports_. These mega Open Access journals didn’t exist ten years ago. And they don’t need to exist ten years from now. What is needed is for the actual costs of publishing (not that currently inflated by publishers) to be covered by the institutions that employ the academics. This could cover type-setting (fulfilling our irrational desire for '[fancy layouts](#layout)' and the additional IT infrastructure (on-line submission system and online dissemination platform). Most (if not all) societies are not-for-profit organisations, and only need to cover the costs of publishing.

### A role for scolarly societies and libraries {#societies}
Scholarly societies have an important role to play in leading change in the scientific publishing model. Not only do societies organise their own membership, and elect members to positions, but they have the potential to make sure that there is equity in their executive committees. Indeed, previous studies have shown that women leaders are promote more equitable societies, but globally most societies in the biological sciences are still dominated by men [@potvin2018diversity]. Promoting equity and representation of a society's membership is an important step on the road to [transparency](#transparent), and to make the shift from [closed to open science](#openscience). Societies be led by the desire of their members for [diamond Access](#diamond) society journals, now that it has been established that this is possible through [overlay journals](#overlay).

without publishers we still need to organise academics at a level beyond the individual and spanning institutions. This means elevating the importance of scholarly societies [@harington2020importance], with all of the additional benefits that these bring (e.g. networking, conference organisation, newsletters, socialising, mentoring, stewardship, community of practice, ethics, outreach, etc.). However, far from relying on the publishers be the sources of income for societies [@harington2020importance], we need a professional society model (such as is seen in medicine and other vocationally orientated disciplines) whereby the relationship between academics and their societies is strengthened through their institutions.

## Can we afford not to change?
If you are from a rich country or institution, then you can probably afford the current system. Those who cannot are researchers in developing countries and disadvantaged institutions. In some cases, the cost of publishing Open Access is greater than the cost of the research. These are insurmountable costs for many researchers. We have a massive hole in scientific contributions from the poorest of nations, and the current Open Access models will see their work being the most hidden from view, while the countries paying for their work do so disproportionately. But even developing countries could be winners in a new Open Access model. By sourcing the relevant IT skills in the country, governments of middle-income countries could facilitate the content of their own society’s with relative comfort.

## Societies need money. Editors can’t be publishers.
We desperately need good, free editorial management software. There are some free versions out there, but what we need are free versions that are at least as good, if not better, than existing platforms (e.g. ScholarOne; Editorial Manager). Galipeau et al. [-@galipeau2016scoping] make the point about the ever expanding role of editors in the modern publishing era. There is no scope for editors to take on extra work.

We need a free LaTeX interface with robust templates for all societal journals. Ideally, this would be packaged with the above editorial management software. This must have the ability to cope with figures and equations, and the unusual demands that some society journals have.

We need a solution for hosting and disseminating the Open Access society journals (and their supplementary information if not hosted elsewhere) in perpetuity. This last point is perhaps the most expensive, and almost certainly requires government assistance. Maybe this is an interesting use of the block-chain with libraries keeping the data. It would be an interesting way to build a doi with editor and referee unique IDs, and the document's information hanging off.

We need to take back our content stuck behind paywalls. Yes, it’s time for you all to dig up those old submitted manuscripts and submit them to an institutional repository where it can be accessed for free.

## A paywall is never acceptable wherever you put it
Any paywall, whether it be high (i.e. EUR 9500 for Nature in 2020), or ‘considerably lower’ (i.e. EUR 900 for NeoBiota in 2020) is a wall that excludes many researchers, and certainly those from many developing countries. While previously I lobbied for the source of these publishing fees to become public [@measey2018europe], which I suggest would show that for a great majority of researchers, publishing fees are coming from research funds. Funds that would otherwise further knowledge are going directly into the pockets of publishers. Publishers instantly refused to do this. Now I think that our energies would be better spent demolishing the paywall.

Until we have [Diamond Open Access](#diamond) for all, having the paywall after publication is the preferable reality for most of us, as most of us cannot afford to pay anything as we simply don’t have this type of discretionary funds. We do have to publish our work, and would rather that it was out there behind a paywall, than not out there at all. We have learned how to live with paywalls, and have inventive means of getting around them. Paying before publication simply stops us from publishing.

I am not advocating a paywall, but I disagree that by placing the paywall before publication (i.e. on acceptance) solves anything for anyone other than the most privileged researchers. In the words of Peterson et al. [-@peterson2019open], "do not replace one problem with another". Instead, what we need is to tear down the paywall with a completely new publishing model for academia. We are all quick and ready to agree that Open Access is the best way forward for all scientific results, so this means [Diamond Open Access](#diamond) for everyone.

If current trends continue, scientists from low-income countries will be granted full fee waivers. Many journals use the [Hinari Eligibility list of countries](http://www.who.int/hinari/eligibility/en/) to separate Group A (free access) and Group B (low cost access – normally billed at a 50% reduction in fees). The lists are made up from five global economic and development criteria. Middle-income countries are missing from these lists, and receive no support for fee waivers. Their governments provide scientists with no means of paying fees. Scientists who pay fees often do so from their own research budgets. The increasing number of journals that charge unjustifiably high publishing costs are forcing middle-income scientists away from Open Access journals.

Moreover, it is not just scientists from middle-income countries that are missing out. Many scientists do their research on a shoestring budget because they were unlucky with grant applications, or are not at a university that can provide any research grants. These scientists do great research, and are in some of the most wealthy countries. For them, the new paywall before publishing also represents a barrier to communicating their science.

## The answer lies inside our University Libraries 
University libraries have undergone a massive transformation over the last 20 years. During my PhD, I made a weekly visit to the library to physically pick up the latest issues of all the journals that came through the postal service from all over the world.Librarians arranged these issues on the shelves and eventually sent them off for binding into volumes and then worried about the physical space that was available inside the library as every year publication inflation [@lariviere2016how] meant more pages to be supported within their walls.

For papers that I found out about but were not in the library, I had a stack of postcards that were specifically for reprint requests, and I enthusiastically filled them out and posted them off to researchers the world over. The post held a weekly haul of fat envelopes filled with offprints that were frequently more than I'd asked for. Many researchers would sign the top with personal messages - halcyon days perhaps. But there are better days ahead.

Like others, I think that the logical solution to our problem with publishers is to turn to our university libraries to curate our academic outputs [@fang2012reforming]. There are clear reasons why it makes sense for libraries to take on the roll as publishers. Most of us are employed by universities or research institutions that also fund libraries. Linking the work we do (writing, reviewing and editing) more closely with our institutions would result in a greater appreciation for this part of our workload. Libraries have fantastic networks, and are our professional long-term storage partners. They developed efficient and impressive information technology (IT) long before it hit most academic departments. Their inter-library networks are what we now need to disseminate the knowledge that we generate without any walls. The idea of libraries as the new publishers isn't new. Raju and Pietersen [@raju2017library] proposed this as a solution in Africa. Here I extend the same idea as an exclusive way of publishing academic journals for the world.

## Do we need a fancy layout? {#layout}
Once the storage and dissemination of our contributions are taken care of, the only service left from the publishers is a fancy layout. This is mostly a historical legacy. I have to admit that I really like seeing my work being nicely produced and printed. But I’m happy to give this up if it means demolishing paywalls. In reality, LaTeX can solve most of these problems so that we simply use the journal (library) produced template, that will need minimal manipulation afterwards. I feel sure that those who are hung up on the importance of their layout can find undergraduates at their own institutions who will be happy to provide layout services for a reasonable fee. No doubt, there will be some institutions that will invest extra to have nicer layouts for their journals. But I feel confident that this will not change the impact factor, or any other journal metrics, as academics will value the papers for what they contain rather than what they look like. Nothing about the contents of the highest ranking journals suggests that impact factor is consistently related to research quality.

If our futures lie with the [overlay](#overlay) model of publishing, then I suggest that we need to have less reliance on international storage of (what are currently) 'preprints', and instead a closer relationship between university libraries and scholarly societies is needed for preprint storage. If you have read this far, then I hope that you will join the call for [Diamond Open Access](#diamond) - no paywalls for anyone.

## Giving up the obsession with metrics
Another key move will be having all academic institutions and funders adopting the [San Francisco Declaration on Research Assessment](https://sfdora.org/read/) (known as [DORA](#DORA)) in order to stop hiring staff based on their publishing metrics, and the negative impacts that this is known to have [@casadevall2012reforming]. This will have an impact on many aspects of science including the current way in which science is funded on a '[winner takes all](#winnertakesall)' basis [@casadevall2012reforming]. Many suggestions for reforming science are out there, but when the participants are still ruled by the cult of metrics, it is hard to imagine the necessary reforms taking place [@fang2012reforming].

### 2021-09-05 4.06-publishmore.Rmd
Add files via upload
# all round update

# Are researchers writing more, and is more better? {#publishmore}

The idiom ‘publish or perish’ suggests that researchers will increase their output in order to obtain positions and promotions. And if a researcher’s productivity is measured by their publication output, shouldn't we all be writing more papers? Certainly, it appears that more papers are being published (see Figure \@ref(fig:authors-writing-more)). An estimate for the total number of scholarly articles in existence was 50 million in 2009, with more than half of these published in the years from 1984 [@jinha2010article].

Similarly, if we are all be writing more, then wouldn't some people start publishing two (or more) papers, when one would be adequate? This idea of ‘[salami slicing](#salami)’ to inflate outputs would be an understandable strategy if researchers were all trying to increase their output. Alternatively, the names of authors might be added to papers in which they did not make significant input via ghost [authorship or hyperauthorship](#authors) [see @cronin2001hyperauthorship for an interesting historical perspective, and [Part I](#authors)].

(ref:authors-writing-more) **The growth in the number of papers published in Life Sciences over time.** The number of papers published (blue line) compared to a standard growth rate (black line). The data come from www.scopus.com.

```{r authors-writing-more, echo=FALSE, fig.cap="(ref:authors-writing-more)"}
if(knitr::is_html_output(excludes="markdown")) knitr::include_url("https://johnmeasey.github.io/authors-writing-more/index.html", height="450px") else knitr::include_graphics("figures/authors-writing-more/authors-writing-more.png")
```

A study by Fanelli and Larivière [-@fanelli2016researchers] has a new take on the above questions, by asking whether researchers are actually writing more papers now than they did 100 years ago. They used the Web of Science to look for unique authors (more than half a million of them) and determine whether the first year of publication and the total number of publications resulted in an increasing trend. Fanelli and Larivière’s [-@fanelli2016researchers] trend line for biology is very stable at around 5.5 publications whether you started publishing in 1900 or 2000 (note that earth science and chemistry do both increase dramatically).

But it is possible that these figures could be explained by the fact that the culture in publishing in biological sciences has changed a great deal since then. One hundred years ago, it was very unlikely that any postgraduate students would publish articles in peer reviewed journals. Moreover, it was also acceptable for advisers to take the thesis work of their students and write it up in monographs. This has certainly changed with the ranks of authors now being swelled considerably so that many more authors are likely to be included on only a single publication in which they participated [see @measey2011past]. I interpret this as the biological sciences becoming more democratic, with more of the people that contribute to the work receiving the [credit](#authors).

### At what rate is the literature increasing?
Using several databases (Web of Science, Scopus, Dimensions and Microsoft Academic) back to the beginning of their collections at the start of scientific journals in the mid 1600s, Bornmann et al. [-@bornmann2020growth] caluclated the inflation rate of scientific literature to run at 4.02%, such that the literature will double in 16.8 years [@bornmann2020growth]. This means that there is literally twice as much published in 2020 as there was in 2003.

Although the early period of scientific publishing was notably slower than today, it is since the mid-1940s (following the end of 'World War II') that science has seen an exponential growth in productivity, with annual growth of 5.1%, and a doubling time of 13.8 years [@bornmann2020growth].

### If more is being published, will Impact Factors increase?
Yes, there is inflation of impact factors. If the numbers of citations per paper remains constant, then the [Impact Factor](#impactfactor) of journals should increase annually at 5%. My impression [see @measey2011past] is that citations are increasing in papers as the literature increases, which suggests that Impact Factor will grow at a faster rate .

## Are some authors unfeasibly prolific?

How many articles could you publish in a year before you could be described as 'unfeasibly prolific'? This question was posed in a study that examined prolific authors in four fields of medicine [@wager2015too]. This publication piqued my interest as it turns out that the authors decided that researchers with more than 25 publications in a year were unfeasibly prolific, as this would be the equivalent of “>1 publication per 10 working days”. Their angle was to suggest that publication fraud was likely, and that funders should be more circumspect when accepting researchers productivity as a metric. Looking back through the peer review of this article (which is a great aspect of many PeerJ articles), I’m astounded that only one reviewer questioned the premise that it’s infeasible to author that number of papers in a year.

I have published >25 papers and book chapters in a year, and I know other people who do this regularly. To me, there is no question that (a) it is possible and (b) that they really are the authors - with no question of fraud. Firstly, the idea that prolific authors constrain their activity to “working days” is naïve. Most will be working throughout a normal weekend, and working in the early morning and late evening, especially in China [see @barnett2019working]. A hallmark of a prolific author would be emails early in the morning and/or late at night. This gives you an indication of their working hours, and how they are struggling to keep up with correspondence on top of writing papers.

Authorship of a publication is often the result of several years of work, and it can be at many different levels of investment (see [Part 1](#authors)). Thus, from my perspective, when I look at authoring a lot of publications it reflects the activity of the initial concept for the work, raising of money, conducting the field work or experiment, analysing the data and then writing it up (with the subsequent submission and peer review time - see [Part I](#authors)). Often, the work conducted by students who lead the publications, many years later, are the culmination of many years of investment of both time and money. And in the Biological Sciences, good study systems keep giving.

## Salami-slicing {#salami}
'Salami-slicing' is different things to different people. While many people refer to salami-slicing as the multiplication of papers into as many papers as possible, or slicing research into 'least publishable units', others are referring to publishing the same paper more than once in different journals. In the latter case, this should be considered self-plagiarism at best and fraud at worst. If you find examples of dual publications, these should result in [retractions](#retract). Instances don't need to be exact copies. I edited a submission where one of the referees alerted me to the fact that the same data with a very similar question had already been published two years previously. In this instance, I passed the submission to the ethical panel of the journal who rejected it, and flagged the authors for scrutiny in the case of future submissions. Note that instances where conference abstracts are printed in a journal does not prevent you from publishing a full paper of this work. I would maintain, however, that you should rewrite the abstract to prevent [self-plagiarism](http://howtowriteaphd.com/plagiarise.html) [@measey2021how].

During the production of any dataset, you are likely to find that you are able to answer more questions that you originally set out to ask when you first proposed the research (i.e. preregistration; see [Part I](#preprints2)). The question you will be left with is: whether you should be adding these _post hoc_ questions (questions that arise after the study) to the manuscripts that you planned to write when you proposed the research, or whether these should be published as separate publications - clearly identified as _post hoc_ questions? The realities of publishing in scientific journals means that in many instances you will be restricted by the number of words a journal will accept. This will mean that for certain outputs it will not be possible for you to ask additional _post hoc_ questions, or potentially all the questions you wanted to report in the [preregistered plan](#commitment).

There is nothing wrong with writing papers based purely on findings that you came across during the study: _post hoc_ questions. There is a clear role in scientific publishing of natural history observations. But that any publication (or part of a publication) that results must indicate that it results from a _post hoc_ study. To my way of thinking, it would be more useful if journals had separate sections for such studies, with other publications only stemming from from those that can show a [preregistered plan](#commitment). This would clearly improve [transparency](#transparency2) in publishing, and avoid accusations of [p-hacking or HARKing](#transparency).

At what point does the separation of research questions into different papers become 'salami slicing'? There is no simple answer to this question, and editors are likely to disagree [@tolsgaard2019salami-slicing]. However, there are ways in which you as an author can make sure that your work is [transparent](#transparency2), and therefore that you are not accused of 'salami slicing'. First is the [preregistration](#commitment) of your research plan. Second is to [preprint](#preprints2) any unpublished papers that are referenced in your submission. There are also guidelines from COPE on the: '[Systematic manipulation of the publication process](https://publicationethics.org/resources/flowcharts/systematic-manipulation-publication-process)' [@cope2018systematic]. And the last is to be transparent when you publish _post hoc_ research.

In manuscripts where another very similar study is cited by the authors, but not available to reviewers or editors, there should be a 'salami slicing' red flag. Obviously, when you produce a number of outputs from a research project, they are likely to be linked and therefore [cited by each other](#self). However, when these are not available to reviewers and editors (as [preprints](#preprints2) or as [preregistration](#commitment) of the questions), authors should expect to be asked for these manuscripts to demonstrate that they are not salami-sclicing. Perhaps worse, however, is when authors deliberately hide any citation to another very similar work. In the end, we have to rely on the integrity of the researchers not to be unethical or dishonest.

## Is writing a lot of papers a good strategy? 
This is a question of long standing, and one that you may find yourself asking at some point early on in your career. I'd suggest that the answer will be more about the sort of person that you are, or the lab culture you experience, over any strategy that you might consciously decide. If you tend toward perfectionism, this will likely result in fewer papers that (I hope) you'd consider to be of high quality. If on the other hand your desire were to finish projects and move on, you'd be more likely to tend toward more papers. It is clear that the current climate leads towards the latter strategy, with increasing numbers of early career researchers bewildered at the idea of increasing their publication metrics [@helmer2020what]. But what should you do?

Given that the 'best' personality type lies somewhere in the middle, you can decide for yourself whether you identify with one side more than the other. But which is the better strategy? Vincent Larivière and Rodrigo Costas [-@lariviere2016how] tried to answer this question by considering how many papers unique authors wrote and seeing how this relates to their share of authoring a paper in the top 1% of cited papers. Their result showed clearly that for researchers in the life sciences, writing a lot of papers was a good strategy if you started back in the 1980s. However, for those starting after 2009, the trend was reversed with those authors writing more papers less likely to have a 'smash hit' paper (in the top 1% of cited papers). Maybe the time scale was too short to know. After all, if you started publishing in 2009 and had >20 papers by 2013 then you have been incredibly (but not unfeasibly) prolific. Other studies continue to show that in the life sciences, writing more papers still provides returns towards having papers highly cited: the more papers you author, the higher the chance of having a highly cited paper [@sandstrom2016quantity].

One aspect not considered Larivière and Costas [-@lariviere2016how] is that becoming known as a researcher who finishes work (resulting in a publication) is likely to make you more attractive to collaborators. Thus, publishing work is likely to get you invited to participate in more work. Obviously, quality plays a part in invitations to collaborative work too. Thus pulling the argument back to the centre ground.

There are other scenarios in which you might be encouraged to write more. In Denmark, for example, research funding is apportioned to universities based on the number of outputs their researchers generated in a point system, where higher ranked journals get more points. This resulted in researchers in the life sciences changing their publication strategy with a notable increase in publications in the highest points bracket following this change [@deutz2021quantitative].

You may find yourself becoming preoccupied about which is the best strategy for you, not because you want to, but because your institution is relying on you to pull your weight in their assessment exercise. University rankings are now very important, and big universities like to be ranked highly for research, which depends (in part) on the quantity and quality of their output.

### Natural selection of bad science {#badscience}
In [-@smaldino2016natural], Smaldino and McElreath proposed that ever increasing numbers of publications not only leads to bad science, but is currently selected for in an academic environment where publishing is considered as a currency. They argued that the most productive laboratories will be rewarded with more grant funding, larger numbers of students, and that these students will learn about the methods and benefits of prolific publication. When these ‘offspring’ of the prolific lab look for jobs, they are more likely to be successful as they have more publications themselves. An academic environment that rewards increasing numbers of publications eventually selects towards methodologies that produce the greatest number of publishable results. To show that this leads to a culture of ‘bad science’, Smaldino and McElreath [-@smaldino2016natural] conducted an analysis in trends over time of statistical power in behavioural science publications. Over time, better science should be shown by researchers increasing their statistical power as this will provide studies with lower error rates. However, increasing the statistical power of experiments takes more time and resources, resulting in fewer publications. Their results, from review papers in social and behavioural sciences, suggested that between 1960 and 2011 there had been no trend toward increasing statistical power. Biological systems, whether they be academics in a department or grass growing in experimental pots, will respond to the rewards generated in that system. When grant funding bodies and academic institutions reward publishing as a behaviour, it is inevitable that the behaviour of researchers inside that system will respond by increasing their publication output. Moreover, if those institutions maintain increasing numbers of researchers in temporary positions, those individuals are further incentivised to become more productive to justify their continued contracts, or the possibility of obtaining a (more permanent) position elsewhere. Eventually, this negative feedback, or gameification of publishing metrics, produces a dystopian and dysfunctional academic reality [@helmer2020what].

An example of this kind of confirmation bias driven publishing effect toward bad science can be found in the literature of fluctuating asymmetry, and in particular those studies on human faces [@dongen2011associations]. Back in the 1990s, there was a flurry of high profile articles purporting preference for symmetry (and against asymmetry) in human faces. The studies were (relatively) cheap and fast to conduct as the researchers had access to hundreds of students right on their doorsteps. The studies not only hit the top journals, but were very popular in the mainstream media as scientists were apparently able to predict which faces were the most attractive. Stefan van Dongen [-@dongen2011associations] hypothesised that if publication bias was leading to bad science in studies of fluctuating asymmetry in human faces, there would be a negative association between effect size and sample size when fluctuating asymmetry in human faces. Effect sizes are expected to be smaller in larger in studies with high sample sizes as these come with less accurate measurements [see @jennions2002publication], but we should not expect to see this relationship change depending on the stated focus of the study. However, van Dongen found that this relationship was diminished in studies when fluctuating asymmetry in human faces was not the main aim of the study, suggesting that there was important publication bias.

Where others have looked, publication bias has been found and is particularly associated with a decreasing effect size that correlates with journal [Impact Factor](#impactfactor): i.e. once the large effect is published in a big journal, the natural selection of bad science results in publication bias, and diminishing effect sizes that ripple through lower impact factor journals [@munafo2007association; @brembs2013deep; @smaldino2016natural], while negative results disappear almost entirely [@fanelli2012negative]. One can presume that negative results exist, but their authors either do not bother to write up the results, or that journals won't publish them.

The direct result of a system driven by [Impact Factor](#impactfactor) and author publication metrics is that we will have a generation of scientists at the top institutions that are trained not to conduct the best science, but to generate publications that can be sold to the best journals. We should be deeply suspicious of any claim of linkage between top journals and quality [@brembs2013deep]. Indeed, what we see increasingly is that the potential rewards of publishing in top [Impact Factor](#impactfactor) journals leads not only to bad science, but increasingly to [deliberate fraud](#pruittdata). Continuing along this path threatens to undermine the entire [scientific project](http://howtowriteaphd.com/lifescientific.html), and places science and scientists as just another stakeholder in a system ruled by economic markets, and their promotion of the fashion of the day [@brembs2013deep; @casadevall2012reforming].

### 2021-09-05 4.07-retract.Rmd
Add files via upload
# all round update

# When should you correct or retract your paper? {#retract}

Once your [Version of Record](#VoR) is published, any changes that you may wish to make will result in a separate correction publication. In extreme cases, you may even need to retract the paper. The different options available in many journals to correct your paper are shown in  Table \@ref(tab:retract). If you are in doubt about exactly what kind of correction you need, read the guidelines from the [Council of Science Editors](https://www.councilscienceeditors.org/resource-library/editorial-policies/retraction-resources/) and the [Committee on Publication Ethics](https://publicationethics.org/) (COPE) [@barbour2009guidelines].

Table: (\#tab:retract) **Journals have a number of different ways in which to correct the published Version of Record.** You many not need a full retraction to set your study straight. Source: Council of Science Editors.

| Action | Example | Conclusions impacted | Issued by |
| --- | ------- | --- | --- |
| Corrigendum / erratum / correction | Important typos / incorrect figure legends or tables / author name or address issues | Not | Author |
| Expression of concern | Data appear unreliable / misconduct suspected | Undetermined | Editor (perhaps through information received) |
| Partial retraction | One aspect of the study is corrupted / Inappropriate analysis | Overall findings remain | Author or Editor |
| Full retraction | Majority of the study is corrupted / Evidence of misconduct / Work is invalidated | Yes | Author or Editor |

## Making a correction to a published paper {#corrigendum}

It is very unlikely that you will be in a position where you will need to think about retracting your paper. If you notice a mistake, especially one that results in a difference to how the results are presented, then you should approach the editor about publishing a correction (also termed a _corrigendum_ or erratum - plural errata).

Because a _corrigendum_ is a change to the [VoR](#VoR), it will result in what is in effect an additional separate publication (with its own [DOI](#DOI)). This will succinctly point out the error and how this should be rectified. In the journal, this will only be a few lines. In addition, on the site of the original publication, the journal will place a notice that there has been a correction, and provide a permalink to the correction. However, it results in a lot of extra administrative work for everyone, so it's best avoided if at all possible. This is another reason why it's worth taking your time when checking your [proofs](#proofs).

Another way to avoid having to make a _corrigendum_ is to ensure that all co-authors are happy with the [original submission](#submit), the [resubmission](#rebuttal) and the [proofs](#proofs) (i.e. get it right before you submit it).

Mistakes do occur, and it is likely to be some time after the publication that you might find that there was an error. Errors such as typos, or mistakes in the introduction or discussion are unlikely to warrant a _corrigendum_. However, if the error is in the way that the results were calculated, or causes a change in the significance, then you should consider making a _corrigendum_. If you feel that it is necessary, do consult your co-authors before taking it to the editor. It is well worth having someone else check your new calculation, as the last thing you want is a compounded error.

If the mistake is systemic, and changes all of the results, their significance and/or the validity of the conclusion, then you need to consider a full retraction.

## Expression of concern

An expression of concern lies between a correction and a retraction. An expression of concern can precede a retraction, but suggests that the editors are seriously worried about something being wrong with the publication. For example, a paper published in _Proceedings of the National Academy of Sciences_ that used an unusual mutant strain of _Chlamydomonas_ (a genus of green algae) was placed under an official 'Editorial Expression of Concern' when the editors learned that the authors would not share their strain with any other researchers [@berenbaum2021editorial]. Clearly, this notice could be removed if, for example, the authors agreed to share their strain and their findings were replicated. However, if they continue to refuse, the editors could also fully retract the paper.

This is an unusual case, but shows how the editors are prepared to take their journal's policy seriously, as a requirement of submission is that authors must be prepared to share any unique reagents described in their papers, or declare restrictions up front. Failure of these authors to follow through with the [transparency](#transparency2) declared on submission could mean that these authors have their paper retracted. Finding out this sort of information takes time, and so there is often a lag in the retraction window (Figure \@ref(fig:life-sciences-retractions)).

## A retraction is unusual

> “A person who has committed a mistake, and doesn't correct it, is committing another mistake.”
>
>  --- Confucius

A retraction of a paper is when your paper is effectively 'unpublished'. This happens at the discretion of the editor (and often the entire editorial board), and is a very serious issue. Retractions are rare. Reasons for retractions vary. It could be that a piece of equipment was later found to have malfunctioned or was calibrated incorrectly [@anon2018retraction]. A cell line was misidentified. Or they can be through no fault of the authors. For example, Toro et al. [-@toro2019retraction] had their manuscript rejected by _Journal of Biosciences_, but due to an administrative error, the article was printed in an issue, and later retracted. However, the top reason for retraction is now misconduct [@fang2012misconduct; @brainard2018what], and this is hardly surprising given the crazy incentives that many scientists received to publish in journals with top impact factors. Another important factor with retractions is that they appear to be more common in journals with higher [impact factors](#impactfactor) [@brembs2013deep], and this should not surprise us as these journals are prone to publishing studies with [confirmation bias](http://www.howtowriteaphd.org/TypeI.html) [@forstmeier2017detecting; @measey2021how].

Although retractions are so rare in life sciences, 0.06% of all papers published between 1990 and 2020, they appear to be on the increase in the last 30 years: from around 0.5 to over 15 papers per 10 000 published (see Figure \@ref(fig:life-sciences-retractions)). It takes a mean time of nearly 2 years between notification of problems with a paper, and issuing a correction or retraction [@grey2021timeliness], but this belies a bimodal distribution in retraction times with the first hump coming from self-correcting authors, or clerical errors from journals coming within months of the original date of publication. The second hump is usually associated with [fraud](#preittdata), and comes after several years of investigations by institutions often with added legal frustrations.

(ref:life-sciences-retractions) **The growth of retractions in life sciences journals over time.** The three lines show when the original paper is published (blue line) and when a correction, expression of concern or retraction are made (black line). Articles that are finally retracted (green line) are only a part of those with other issues after 2008. While the number of papers published that are later retracted appears to take turn downward from 2014, this may simply represent the lag before they are later retracted. This data came from the Retraction Watch database, selecting only data from  Basic Life Sciences and Environment. The data is normalised by dividing the total number of publications (as taken from SCOPUS) in this area for the year by 10 000, and multiplying this by numbers of Original or Retracted papers.

```{r life-sciences-retractions, echo=FALSE, fig.cap="(ref:life-sciences-retractions)"}
if(knitr::is_html_output(excludes="markdown")) knitr::include_url("https://johnmeasey.github.io/Life-sciences-retractions/index.html", height="450px") else knitr::include_graphics("figures/life-sciences-retractions/life-sciences-retractions.png")
```

Retractions showed a steep uptake in 2011, when a number of laboratories that made multiple retractions in 2011 [see @fang2011retracted; @brainard2018what]. A single publisher was responsible for pulling a great many of the retracted papers in 2011, and this spike in retractions isn't seen in the life sciences [@oransky2011year; @brainard2018what]. What is clear from Figure \@ref(fig:life-sciences-retractions) is that the rising trend in retractions (black line) appears to have been unaffected by the 2011 spike in other subject areas. Claims that retractions are levelling off [@brainard2018what], are not matched by the data.

### How do you know if a paper you cited is later retracted?
Citations to retracted papers are not uncommon, and often positively cite the paper even when the retraction has been made for misconduct [@bar-ilan2017post]. This suggests that most authors are simply not aware of the retracted status of many publications. Of course, if you visit the publisher's website, you should see a clear notification at the [Version of Record](#VoR) (e.g. Figure \@ref(fig:retraction1)), that points to the retraction notice (see Figure \@ref(fig:retraction2)), but this is not always the case. In general, publishers seem very shy about their retractions, and it can be difficult to track down retractions that should be clear for everyone to see. Indeed, if publishers did their due diligence on notifications on retractions, and this were entered into CrossRef, we wouldn't need an independent database like the [Retraction Watch database](https://retractionwatch.com).

(ref:retraction1) **You should not cite a retracted paper.** Once papers are retracted they don’t disappear. They continue to be available at the publishers' website, but with a clear notice that they have been retracted (see below). In addition, a separate publication is made announcing the retraction of the work, as shown in Figure \@ref(fig:retraction2).

```{r retraction1, echo=FALSE, out.width = '50%',  fig.cap="(ref:retraction1)"}
knitr::include_graphics('figures/retracted1.png')
```

If you downloaded the article before it was retracted, then you will not be aware of what has happened unless you are following that particular publication. Similarly, if you get your search results from Google Scholar, there is no indication that a paper has been retracted. Scopus and Web of Science clearly indicate some articles that have been retracted, but the vast majority go unrecorded even on these databases. Perhaps this is why even highly publicised retractions continue to be cited by articles that follow [@piller2021disgraced]. Clearly, the community is still responsible for watching what happens to the literature, even once a paper is cited. Of course, the publishers could be using items such as [DOIs](#DOI) to track retracted papers and query their citations. So why don't they?

Publishers are bad at sending through the correct metadata with their content. For example, for the same 30 year period as Figure \@ref(fig:life-sciences-retractions), Web of Science lists only 133 retractions. Really what this means that if you aren't using Zotero, and you want to be sure that there hasn't been a retraction in any of your source material, you need to run a search on the [Retraction Watch database](https://retractionwatch.com) - better to use Zotero says Ivan Oransky. Right now, the chance that any paper published in the last 30 years that you have cited will get retracted is still low (one in 1750), but if it was published in 2020 this rises to one in 650.

Some literature databases will notify you if a paper that you have in your database is retracted - but don't count on this unless you use [Zotero](https://www.zotero.org/). Zotero takes [DOIs](#DOI) from the [retraction watch database](#retractionwatch), and uses them to notify uses of any retractions of articles that have occurred. A large red bar (that's hard to ignore) is shown across the top of the citation window. This is a strong and very positive reason for using Zotero, leaving you to get on with your research.

### Notification of retraction
The notification of retraction is supposed to explain exactly why a paper has been retracted. For example, if you have cited or used this work you should know whether it is because data has been fabricated, or more innocently there was a mistake with the equipment or another aspect of the investigation. However, it seems that some journals are issuing retraction notices that fall short of the guidelines required by COPE [@barbour2009guidelines], and that these delays are not in the interest of anyone involved [see @grey2021timeliness; @teixeiradasilva2021abuse]. Clearly, there is need for improvement here on the part of the journals.

But we must be cautious about playing a blame game when it comes to journal retractions [@smith2021reimagining]. We already know that peer review has shortcomings (see [Part IV](#problempeerreview)), and even the best of peer reviewers and/or editors cannot be expected to spot potentially fatal errors, especially when these come about from deliberate deceit on the part of the authors. Retraction will remain a necessary part of the scientific publishing process, and as retractions become more commonplace among journals, we can hope that guidelines will be followed in a timely manor [@grey2021timeliness].

(ref:retraction2) **You can however, cite the retraction which is published under a separate citation string.** You might want to do this as an example of the type of work that is retracted, like the one shown [@costa-pereira2020retraction]. This retraction notice refers the the Version of Record seen in Figure \@ref(fig:retraction1).

```{r retraction2, echo=FALSE, out.width = '50%', fig.cap="(ref:retraction2)"}
knitr::include_graphics('figures/retracted2.png')
```

## Retraction Watch {#retractionwatch}
To learn more about retractions in science, I encourage you to read the blog at Retraction Watch (https://retractionwatch.com/). This will give you an idea of the reasons why retractions are made, and give you some perspectives about the practices (and malpractices) that go on in the scientific environment.

## Fabrication of data {#fabrication}
The fabrication of data does happen. A growing body of retractions and alleged evidence on the tampering of data in spreadsheets has led to the suspension of a top Canadian researcher, [Jonathan Pruitt](#Pruittdata). The detection of fraudulent (usually duplicated) data in spreadsheets is not too difficult to spot (e.g. by application of Benford's law - the last digits of naturally occurring numbers should approach uniformity), and has become the subject of some contract data scientists who specialise in finding such instances of fraud. To some extent, the automated assessment of fraudulent practices has been or could be implemented for many infringements [@bordewijk2021methods].

The existence of [paper mills](#papermills) should also be mentioned at this point. Papers, produced commercially and to order, are entirely fabricated by third parties to improve the CVs of real paying scientists [@teixeiradasilva2021abuse]. Hundreds of these papers have been discovered that appear to come from the same source [@bik2020tadpole], but the true size of such additions to the scientific literature is unknown.

Some have suggested that the pressure to obtain a permanent academic position is enough to drive some scientists to commit fraud [@husemann2017publicationism; @kun2018publish; @fanelli2015misconduct]. The idiom ‘[publish or perish](#publishmore)’, and the importance of publishing is made elsewhere. However, I hope that by shedding some light on unethical practices, this book equips you to avoid these together with those that may espouse them, and instead show you that there is a better path to success.

[Pruitt’s case](#PruittData) highlights a good reason for increased transparency in the publication process. A blog post from someone caught up in the Pruitt retractions makes the point that journals that insisted on full data deposits for publication were well ahead of those that hadn't [@bolnick202117]. Of growing concern in many areas of Biological Sciences is the potential to manipulate results that are essentially images of results, for example blots on a gel. However, it turns out that manipulated images are also not too hard to discern.

Images are increasingly being used in journals to demonstrate results, and the manipulation of images in published papers appears to be rife. In a study of more than 20000 papers from 40 journals (1995 to 2014), Bik et al. [-@bik2016prevalence] found that nearly 2% had features suggesting deliberate manipulation. These could include simple duplication of an image from supposedly different experiments, duplication with manipulation (e.g. rotation, reversal, etc.) and duplication with alteration (including adding and subtracting parts of the copied image). The authors suggested that as they only considered these types of manipulations from certain image types, the actual level of image fraud in scientific papers is likely much higher [@bik2016prevalence]. An R package (FraudDetTools) is now available for checking manipulation of images [@koppers2017systematic], but there are other steps that reviewers and editors can take themselves [@byrne2020digital].

### Who is responsible?
In the case of fraud, retraction statements should indicate who the perpetrator is in order to exonerate the other researchers. Some research into the likely source of the fraud has been conducted. There are clearly [serial fraudsters](#PruittData), and the presence of their names in the author list is a red flag for those investigating fraud. Data from papers that are known to be fraudulent suggest that the first author is the most likely to be responsible for the fraud committed, and middle authors the least [@hussinger2019scientific]. This suggests that in a collaboration, you should be very careful who you collaborate with. While you might not be responsible, the discovery of (particularly large scale) fraud might well harm your career.

In a study looking for patterns about types of authors involved in retractions (all reasons), suggested that Early Career Researchers were particularly likely to be involved in retractions [@fanelli2015misconduct], although the exact reason why remains obscure.

It is worth noting that while the journals (and ultimately the [journal editor](#gatekeepers)) are responsible for retractions from journals, this is not the same as punishing individuals who have committed fraud. As we have already seen, there can be many reasons for a retraction, and it is not up to editors or journals to punish researchers as there will be innocent researchers who may also need to make retractions. Moreover, it should never be the role of the journal to have any punitive action over a researcher. There are other mechanisms for this with the employer and (where applicable) the academic society involved. Different institutions and governments will have different rules when it comes to fraud being committed by their employees. These processes may be legal and take some time to finish. As we saw with the case of [Pruitt](#Pruittdata), once lawyers get involved, the process may become mired in a lot more beaurocracy.

Whilst you may consider that your employers are not a group that you are particularly concerned about, you should consider the possibility that any scientific fraud could penetrate deeper than you are aware. For example, if you used fraudulently obtained data in order to make a grant application look better, then this would be deemed very serious by the (probably) government that you were applying to. Essentially, this becomes financial fraud as you are using false data in order to obtain money. While your employers may simply remove you from your position (and their employment), the government might prosecute and you could find yourself in prison. This does happen in some countries, and so (obviously) it is a really bad idea to commit fraud.

## What to do if you suspect others {#fraud}
If you suspect that someone in a lab in your department, faculty or university is fabricating data, find out whether your university has a research integrity officer (RIO); most universities in the US have one. Document your evidence if you can and approach the RIO or person in the equivalent position. If you can't find such a person, then ask at you university library for the most relevant person. Libraries are usually neutral places where you can find out information without arousing suspicion. You do need to be careful that you do not place yourself in harm's way when reporting, so be prudent about sharing until you are assured protection from any potential retaliation. It isn't easy to be a whistleblower - but it is the right thing to do.

If the research is published, and you think it is fraudulent, approach the editor directly. If there is some conflict of interest (like the person is at your institution), then you can try to sort it out internally (as described above). Otherwise, you can approach the editor directly yourself, anonymously or by using a third party.

The [Committee on Publication Ethics](https://publicationethics.org/) (COPE) has published some useful flowcharts to guide researchers who suspect fraud in manuscripts or published articles:

 -  ['Image manipulation in a published article'](https://publicationethics.org/resources/flowcharts/image-manipulation-published-article) [@cope2018what]
 -  ['How to recognise potential authorship problems'](https://publicationethics.org/resources/flowcharts/how-recognise-potential-authorship-problems) [@cope2018how]
 -  ['Systematic manipulation of the publication process'](https://publicationethics.org/resources/flowcharts/systematic-manipulation-publication-process) [@cope2018systematic]
 - ['How to recognise potential manipulation of the peer review process'](https://publicationethics.org/resources/flowcharts/how-recognise-potential-manipulation-peer-review-process) [@cope2017how]
 - ['If you suspect fabricated data in a submitted manuscript'](https://publicationethics.org/resources/flowcharts-new/what-do-if-you-suspect-fabricated-data-0) [@wager2006suspecteda]
 - ['Ghost, guest, or gift authorship in a submitted manuscript'](https://publicationethics.org/resources/flowcharts-new/changes-authorship) [@wager2006suspectedb]
 - Undisclosed conflict of interest 
   + [in a submitted manuscript](https://publicationethics.org/resources/flowcharts-new/conflict-interest)
   + [in a published article](https://publicationethics.org/resources/flowcharts/undisclosed-conflict-interest-published-article)
 - ['Plagiarism in a submitted manuscript'](https://publicationethics.org/node/19701) [@wager2006suspected]

## Confirmation Bias and the paradox of high-flying academic careers {#PruittData}
Jonathan Pruitt had it all going for him. His studies of spider sociality were producing novel and significant results that opened the door to publications in high impact journals. In turn, this opened the door to getting prizes and funding. The funding allowed him to conduct more studies and soon a prestigious chair in Canada with more funding to pursue his rocketing career. Things started to unravel for Pruitt when colleagues raised concern about the data in some of his publications. Things gathered pace very quickly, and doubt gathered around more and more of his publications. Although there is much written about the Pruitt debacle on the internet, the blog by *American Naturalist* editor and former Pruitt fan and friend, Dan Bolnick, is particularly enlightening [@bolnick202117]. Pruitt's case is becoming increasingly untenable as more editors backed by co-authors are retracting papers where he contributed data [see @marcus2020spider]. For Pruitt, this has become a threat to his career and livelihood [@pennisi2020embattled]. Similarly, his university is facing the possibility that they hired a fraud. Consequently, this whole debacle has slipped into the legal world. Bolnick has clearly suffered personally from the affair, but has set out to provide as transparent an account as possible.

The harrowing part of Bolnick's account is when he, co-authors and other editors started to receive letters from Pruitt's Lawyer (see one example on [Bolnick's blog](https://ecoevoevoeco.blogspot.com/2021/05/17-months.html)). At the point that the lawyer steps in, the functioning academic community that had raised itself to meet the demands of the concerns began to get muted. Bolnick then makes an important point that the legal threats from Pruitt's Lawyer were stifling the freedom for academics (in this case the co-authors and editors) to publish, and therefore their [academic freedom](#tenure).

Another example of a rising star with high profile papers, allegedly making a habit of fabricating data, comes in the world of marine biology [@clark2020ocean]. The researchers in question, Danielle Dixson under the supervision of Philip Munday, made counter-claims that the detractors were unimaginative or that they are attempting to make a career from criticism. Meanwhile, students from their own labs continue to raise concern about the culture of fraud [see @enserink2021does]. In this case, the tide of evidence against the marine biologists appears to have turned, with forensic data specialists finding multiple examples of suspect data.

Neither case is fully resolved as the cases against these scientists still rest with their institutions. The lives of co-authors, former students and colleagues are put on hold, until some unforeseen point in the future.

It is clear from these reports that there are systemic problems when high profile scientists are accused of fraud. Journals say that it's the responsibility of the institutions, and the institutions have no impetus to find fraud as that might lose them a very productive (think research income) and high profile scientist. What university would want to have its name dragged through the mud, and on top of this lose a large amount of grant income? Top researchers become untouchables in many institutions because they are essentially cash cows that no-one wants to disturb. Allegations against such individuals also include [bullying](#bullying) and sexual misconduct. For those interested in reading more high profile misdemeaners in science Stuart Ritchie [-@ritchie2020science] has put together a popular book on the subject.

Another important issue that arises from (alleged) scientific fraud is that it creates a [culture of research](#badscience) that pushes towards an extremely unlikely hypothesis, in the misbelief that the hypothesis is likely given the nature of the publications [also see @fanelli2017meta-assessment]. Indeed, this natural selection of '[bad science](#badscience)' has permeated the hiring system so that researchers like this are more likely to be hired [@smaldino2016natural]. Forsmeier et al [-@forstmeier2017detecting] have an excellent review that outlines the problems with a culture that pushes towards increasingly unlikely hypotheses [see also @measey2021how on [Type I errors](http://www.howtowriteaphd.org/TypeI.html)].

At the heart of all of this is the cult of the [Impact Factor](#impactfactor) and the research mentality that it generates.

### 2021-09-05 4.08-bullying.Rmd
Add files via upload
# all round update

# Are you bullying or being bullied? {#bullying}

I am going to talk about academic bullying because it is currently prevalent in academia, and because most of the bullies are unaware of what they are doing. In a 2019 survey of graduate students (see [here](https://www.nature.com/articles/d41586-019-03459-7) Figure \@ref(fig:bully)), 22% felt that they had experienced bullying during their PhD program. The only way of improving the situation around academic bullying is for everyone to become more aware. It may not be happening to you, but it may be happening to people around you either in your lab or in another lab in the same department or faculty. If you think that this is very rare behaviour in academia, think again [@devlin2018hundreds].

(ref:bully) **Do you feel that you have experienced bullying in your PhD program?** Responses to a survey of graduate students (in 2019) demonstrates the changes in bullying propensity in different research cultures.

```{r bully, echo=FALSE, out.width = '95%', fig.cap="(ref:bully)"}
 knitr::include_graphics("figures/bully.png")
```

## How to spot a bully
A bully is anyone who abuses or misuses their position of power in order to humiliate, denigrate or injure another. This does not need to be your advisor, or someone in your lab. It can be anyone in your working environment (watch the video [here](https://youtu.be/bZmmp7i9Tsc) or [here](https://youtu.be/lKUONMm-pWo)). These people are usually in positions of power with influence over you and your future. As in my example (below) you may be worried that the power they have could be used by them to negatively impact your future. If you are worried about this, then their behaviour most likely conforms to bullying. Bullying often involves harassment that is designed to undermine your dignity, often through sexism, racism, or another prejudice [@krishna2020playing]. Even if you think that your bully didn’t mean to cause offence, the fact that they did upset you and that this behaviour was unwanted is enough to fulfil the criteria for bullying. Thus, it is not what they intended, but what you felt that is important in bullying. A direct consequence of this is that bullies often don’t recognise this as a description of their behaviour. It is worth bearing in mind that bullies are often damaged individuals who are repeating behaviour that they have themselves experienced from others. This doesn’t excuse their behaviour, but they may think that such behaviour is normal.

What you should ask yourself is if someone were to observe this behaviour from the outside, would they recognise the interaction as normal or see that something was not correct? Of course, if you observe this going on with someone else in your group, or outside your group. Take the initiative to approach the person after and determine whether they feel like a victim: remember that not all interactions are as they appear from the outside. This is where a good set of institutional rules about bullying is important.

## What to do about bullying
First, you need to find the rules that your institution has regarding bullying. If your institution has no rules, then they will need them and so helping them achieve this would be a good place to start [@mahmoudi2021filling]. No one wants to be the first case study, but there may need to be a first in order to set up a protocol.

Avoid the bully when asking for your institutional rules, but you should be able to find them via your departmental secretary, administrative support staff in the department or faculty. Read the documentation carefully and learn about how and by whom such reports are dealt with. Become aware of resources that are available; [paritymovement.org](https://paritymovement.org/) is a great place to start. Read more about other people’s experiences and be aware that you are not alone [@mahmoudi2020survivor; @malaga2004meyer].

Next, document your case. Make some notes about the incident(s), when they happened, what was said, how you were made to feel, and what power you feel the person has over you.

Share your burden with a trusted friend or colleague. It is worth sharing the incident with others to see what they think about your predicament. In the survey mentioned above, more than half of the respondents who said that they experienced bullying felt that they could not discuss their experience for fear of reprisals. However, you do not have to discuss it inside the workplace, and often it’s better to talk to people outside as the context is not so important in bullying. In your description, attempt to strip down the interaction into the component parts.

Follow your university’s rules about who to go to with your complaint. Don’t leave it to the next person in your lab to experience, they may not be as strong or as resourceful as you. It may not be your career that is destroyed, and the next student might not be so lucky.

## What to do if the procedure against bullying at your institution doesn’t work?
My worst experience of bullying happened when I was a PhD student. It happened to me, and I witnessed it happening to other members of my lab. It wasn’t hard to spot. Students would come out of my advisor’s office in tears, and recount horrific stories of how he had debased and humiliated them. At the time, our department had no specific code on bullying, but there was a complaints procedure which started with the head of department. Unfortunately, as my advisor was also the head of the department, I could not follow the procedure as it was supposed to be done. Instead, I went to the academic who was responsible for postgraduates. The first two times, that staff member simply went to the head of department (yes, my advisor) and I was called in both times and bullied some more: how dare I complain about him?

The last time I tried to complain, once I had finished my PhD and felt much safer from the bully, I went to the dean of the faculty. He was the line manager of my advisor (still head of department) and was a lot more sympathetic. After listening to my story, and how my other lab mates were still suffering, he called them in one by one. And one by one they each denied all of the bullying that had happened. They were afraid. Unlike me, who had finished, they were still relying on their advisor to get their postgraduate degrees. The result was that without any corroborating evidence, there was no case for the dean to take forwards.

What should you do if the procedure at your institution doesn’t work? You become a survivor. You also become more vigilant against bullying in the future. Whatever happens, don’t be tempted to become the bully yourself. Support other survivors and make progress to improve the system for future postgraduate students. Bullying is in human nature, and it won’t stop. But we can make people more aware of it, and we can have procedures that work both for the bullied and the bullies.

### 2021-09-05 4.09-mentalhealth.Rmd
Add files via upload
# all round update

# Keeping track of your mental health {#mentalhealth}

Stress is a natural part of life, and many people are at their most productive when they are under some degree of pressure, such as a deadline. Although deadlines don’t work for everyone. Problems arise when we become overwhelmed by stress and are unable to fully respond. When this occurs productivity can drop off and survival responses can be triggered as if responding to an actual physical attack. These responses include fight, flight or freeze responses. Anxiety and panic can be triggered. In this state additional demands on your time may also push your life off balance, so that you start to neglect your personal wellbeing which can negatively impact on relationships, exercise regime, or even nutrition and personal hygiene. Some people can find that the additional stress can cause physical symptoms that may even need medical treatment. Your sense of competence and mastery can be negatively impacted such that you may even suffer from feelings of inadequacy or imposter syndrome. Impostor syndrome is an experience you have when you believe that you are not as competent as you think others perceive you to be. It is not uncommon in many professions, and especially prevalent in academia [@clance1978imposter]. This is now widely recognised and there are lots of useful shared experience out there to read [e.g. @dickerson2019how].

The General Health Questionnaire is an instrument used to measure psychological distress. For example, you could use the GHQ-12 in [Table 4.1](http://www.howtowriteaphd.org/healthy.html#tab:GHQ) [see @measey2021how]. It is quick, reliable and simple to score, so you can use it at any time during your career as an indicator of whether you need to reach out to personal, occupational or professional support networks. My suggestion is that you complete a General Health Questionnaire now and record your answers as a baseline. Keep the scores somewhere safe. If you feel that you are getting overly stressed, take the test again and compare them with your baseline scores. Although there are no hard rules, if three or more of your scores have moved by two or more points it could be worth discussing with your support network to help you decide whether or not to seek professional help.

Even if you don't feel you need the support of your institution now, it is worth finding out how they can support your mental health in the future if needed. Although there has been some stigma attached to difficulties with mental health in the past, most institutions accept that pressures are mounting on their staff and that they may require support. Most institutions will make experienced councillors available to support you if needed. Importantly, you should realise that none of these symptoms are unusual and that there is a high probability that many of your colleagues may also be struggling. Knowing that your problems are shared and reaching out to support networks early is an excellent way to prevent them from escalating beyond your control. Your best means of coping will be to try and develop a support network and to understand where and with whom you can discuss any difficulties as they arise. Knowing who this is and how and when to approach them will put you in a stronger position if you need them in future.

## Being physically active improves mental wellbeing

The positive relationship between the amount of physical activity and higher mental wellbeing is well established [e.g. @grasdalsmoen2020physical; @gerber2014increased], but the kind of exercise required to achieve this improved result is varied. There are plenty of studies out there that suggest there are multiple benefits from physical activity [@williams2021move]. This does not have to be the most hectic exercise possible. You'll get huge benefits from simply taking a walk.

This does not exclude you from participating in other less physical activities, including mindfulness or meditation. Try different types of activities and then do what's right for you.

## Time to think

Use the time that you spend exercising as thinking time: turn ideas over in your head, think through the logic in arguments, and potential flaws in experimental design. If you can, spend this moving time with others in your group discussing projects and talking about ideas. There are reasons why we do some of our best thinking when physically active, and you should seek to exploit these when you can. For example, if you have a meeting scheduled then why not do it as a walk - even if it's on the university campus this walk will get you and your colleague in a much better space to plan ideas.

This isn’t to say that you won’t need time at the computer writing it all down, or talking around a table with colleagues. Be inventive in how you spend your time, and don't resign yourself to spending all day sitting behind a desk. Don’t take my word for it, see this great new book by Caroline Williams [-@williams2021move].

### 2021-09-05 4.10-jobinacademia.Rmd
Add files via upload
# all round update

# Getting a job inside academia {#jobinacademia}

I have already written at the start of this book (see part 1) that competition is high to get jobs in academia and your chances are therefore diminishing. However, academia needs new academics, and so I wanted to try to provide some key insights into what might be important in getting a job as an academic. Of course, there is no single way in, and your path is likely to be different to mine and others that you meet.

During lockdown in 2020 my lab held some meetings with invited guests from all over the world. What follows is a list of key insights that came from a meeting with four academics on four continents. if you are interested in getting a job in academia then take a look and consider these.

## Key Insights if you want a career in academia
Here is a list of key insights that our team shared. They are in no particular order, but each one probably deserves a lot more information.

- Academics often suffer from [imposter syndrome](https://en.wikipedia.org/wiki/Impostor_syndrome)
- A suite of skills that are all are required:
  + Need a logical mind: even OCD
  + Good writing skills
  + Computational skills
  + Attention to detail
  + The importance of finishing the job
  + Being creative (not just for arts students)
  + Need to read (a lot)
- The coolest job in the world as you’re paid to learn
- Research has to be fun 
  + Once you have a tenured position metrics are unimportant to you (but remain important to your students!)
  + Have to produce a publication that you are proud of
- You have to tolerate rejections
  + Papers
  + Conferences
  + Positions
  + Promotions 
- Don’t be harsh on yourself - it happens to everyone
- Don’t rely on how people used to get hired, or that positions you see now will become available in the future 
- Don’t ignore the importance of natural history observations
  + Write the notes as they demonstrate productivity
  + They make your CV look stronger in the early stages
  + Use them to get things published strategically
  + Don’t fill your CV exclusively with these notes (there are more important things)
- There is still a glass ceiling in employment institutions
  + Things are improving
  + Don’t allow comments phase you - many people don’t understand their own prejudices or discriminations.
  + Share the down sides, you’ll find out that lots of other people experience them, and not just you
- Role models are very useful in science
  + Especially someone with whom you can identify, preferably from your own background
  + Use your network to explore which role model might fit best
- Someone needs to fit into the context of the job
  + This means that not every job will be right for you as other people might fit better [not the fault of you or your CV]
  + Could be why lots of people get jobs from the inside - they are already known to fit into the team
  + Many institutions pride themselves on their position in the community, and will look fondly on people who are clearly committed to the place. This could include:
  ++ Speaking the local language
  ++ Writing popular articles for a local audience
  ++ Engaging with the local press and media
  ++ Giving talks to local groups
  + Give back into the faculty or institute - can you demonstrate that you do more than just conduct research and write papers. Are you active in your community (both big - academic community and small - departmental community)
- Learn about opportunities and take advantage of them [you make your own luck]
  + Many people don’t know about opportunities that are out there. 
  + Meet people one on one at smaller meetings [big meetings aren’t good for this]
- Moving around [in postdocs or between jobs] can make it more difficult to get accepted into some places that have a culture of staying put
  + If you know where you want to be, it’s worth investing time to that institution and community [but don’t count on it!]
  + Try to keep multiple irons in the fire
- Even though career paths look linear, this is really only in hindsight. In reality they are wondering paths that sometimes wonder right out and back in again
- Writing small grants isn’t a waste of time as it develops this as a skill 
  + Grant writing is different to thesis writing or paper writing
  + Having a CV with evidence of lots of grants gained (even if they are small) demonstrates to people that you know this stage of the process
  + Showing that you can finish the same projects and produce outputs is even better
- Try saying yes to opportunities (especially early on in your career)
  + You never know where it will take you
  + Establish collaborations outside your direct circle
- Creativity includes reinventing yourself and your science as you move through your career
  + Each grant proposal is to do different work and take you into new paths and directions
  + Some will work out and open up whole new areas or specialities. Others won’t
- It’s easier if you can describe exactly what you do early on in your career
  + A very mixed up CV leaves some people unclear about who they are hiring and for what
  + This doesn’t mean that you have to be overly focussed, but early on it’s useful to have a tag (or a few specific categories like “systematics”, “comparative anatomy”, “natural history” to organize your publications if they’re really different from each other)
- What type of job do you want?
  + Teaching - then get teaching experience 
  + Research - then make sure your CV is strong
  + There are more types of jobs out there, but if you know what you want (or don’t want) then make sure that your CV reflects this
- Your first job is not necessarily your last job 
  + But it could be if you love it
  + You can use it as a springboard to go elsewhere
  + You might need to take the first job to get somewhere else
- Learn about what you are good at and embrace it
  + This might require some honest reflection
- It’s totally possible to have a job outside academia and then move back in
  + Some jobs might even give you an advantage in getting an academic position
  + Many academic subjects are applied, and so experience in the relevant jobs really help [industry relevant experience]
  ++ You may then have inside knowledge to subjects that are taught
  + Maintain your understanding of the field (to get back in)
  + Make sure that there is a continuing narrative, a reason why you left and why this helps you come back in 
- Keep irrelevant jobs off your CV
- Knowing what you don’t want to do can be as helpful as knowing what you do want to do
  + Internships are great opportunities for this
- Get feedback on the letters that you write when applying for jobs
  + Ask people whether they will give you a letter of reference, and if that will be positive 
  + Take note of what the job is asking for and ask specifically for those aspects to be mentioned
  + Different regions of the world have different styles for letters of reference
  ++ US letters are thorough [and often over the top, even flamboyant] and very long
  ++ Europeans tend to be understated and more direct
  ++ Some parts of the world may provide just a few sentences
  + I ask students to draft their own letters that accentuate what they themselves want to underline about their experiences. I won’t use the same words, but it will help remind and inform me of what colleagues have done.

### 2021-09-05 4.11-habilitation.Rmd
Add files via upload
# all round update

# Habilitation, DSc and Tenure {#habilitation}

These are qualifications post-PhD that exist in some countries and might be prerequisites to getting some positions within academia. I do not go into these in detail here because they are country specific, and you are likely to learn much more about them at your own institution. The brief inclusion here serves as a guide to their existence, and for you to be aware that different rules apply in different countries, and that if you are mobile in your career there may be additional steps that are required of you before you can apply for a job or certain promotions.

## Habilitation 
Habilitation from the Latin _habilitare_, "to make fit", started in Germany first as part of the PhD process, and later as a separate post-doctoral qualification in the 1800s. This process has been adopted by a number of (mostly European) countries as a requisite step in teaching or directing research. In countries where this qualification exists, it is usually a prerequisite before being able to apply as a candidate for a professorship. In some countries, most notably Germany, the habilitation comes after having already held a job as a researcher and lecturer, and comes with a serious expectation that this will lead to a promotion to become a professor. In this way, it could be seen as a similar process to going for tenure in the US.

In France, the related qualification is the Habilitation à Diriger des Recherches (Accreditation to Direct Research or HDR). Like the German system, the HDR is applied for by someone who already holds a position as a lecturer (Maître de Conférences) for several years, when they hold sufficient research to put together a portfolio. You need to be accredited with HDR before you can advise PhD students. Ironically, this portfolio should include the supervision of at least one PhD student. Thus, you’ll need to arrange to be a co-advisor when you are actually the main advisor, before moving forwards with your HDR portfolio. Given that your first PhD student may take some years to finish, it would be worth finding a sympathetic person with an HDR sooner rather than later, once you are in your Maître de Conférences post.

In the biological sciences, most requirements for habilitation are cumulative, meaning that you can assemble a set of published research papers that you have written or led. The number and quality of such publications will depend on where you are submitting this thesis, meaning that in some places it may take as long as 10 years. Importantly, the habilitation is not advised.

## Doctor of Science (DSc)
In the absence of any requirement for habilitation, there is the possibility (at many universities) of compiling published papers, that you have written or directed, into a thesis that can be examined for a Doctor of Science (DSc). Like the PhD, the DSc allows you to call yourself Doctor (although you likely already can) and put the letters DSc after your name. The DSc is touted as an advanced doctoral degree. You will need to register as you would for a PhD, but in most cases your thesis will not be advised.

One interesting point to note is that registration for a DSc need not have possession of a PhD as a prerequisite. If you are in a position where you have never done a PhD, but have worked within or alongside academia, including publishing papers, for a considerable period, you might be in a position to register for a DSc.

## Tenure {#tenure}
Obtaining tenure (in the USA and Canada and some other countries) gives you a special kind of academic freedom such that it is very hard for you to be removed from your post. In some states this means that you are not required to retire (a job for life - although there are [increasingly attractive offers for professors](https://www.theatlantic.com/business/archive/2016/06/colleges-offer-retirement-buyouts-to-professors/487400/#:~:text=Since%20about%2090%20percent%20of,traditional%20retirement%20age%20of%2065.&text=They%20were%20able%20to%20impose,past%20the%20age%20of%2070.) to [retire](https://www.thecrimson.com/article/2018/5/23/yir-aging-faculty/)). Tenure exists around the need for independent academic freedom: that as an academic scholar you are free to hold your own, scholarly views, and as such cannot be censured by the state. Getting tenure, therefore, at the university where you are employed is an important step, vital if you want to move from contract to permanent employment. In practice, if you don’t get tenure it will most likely mean that you won’t get to continue at that university: tenure or bust.

In order to obtain tenure in most US universities you will need to provide:

- A portfolio of peer-reviewed published research
- The proven ability to attract grant funding
  + A significant amount of which goes to the university
- Teaching excellence
  + As assessed by undergraduate and postgraduate students
- Academic visibility
  + The recognition of your research by peers through inclusion in conferences, invitations to give seminars, etc.
- Administrative and/or community service
  + This includes roles such as being an editor for a journal
  + Peer review for journals and grant awarding bodies
  + Serving on your university’s committees and panels

The relative importance to each of the above aspects will depend on the type of college where you are trying to get tenure. Unsurprisingly, a teaching college will require you to have excellence in teaching, while a research university will place more emphasis on your research portfolio and your standing as an academic in the international community. If you aren’t from North America, it is important that you know what the priorities of your institution are before you apply for a position there, or even before you try to do a postgraduate degree.

Once you are in an untenured post at a US university, you will have a limited amount of time to achieve the above portfolio in order to achieve tenure. Getting tenure often comes together with promotion (to professor) and a reduction in (undergraduate) teaching load. The time limited nature of getting tenure is such that even after you have received your PhD, this is a much higher hurdle to attain.

### 2021-09-05 4.12-leavingacademia.Rmd
Add files via upload
# all round update

# Leaving academia for a job outside {#leavingacademia}

In addition to thinking about jobs inside academia we also invited some past members of my lab who have got jobs outside of academia. They also provided us with their key insights into how the world outside academia is different to that inside. Again, if you are thinking about it getting a job outside academia take a look at some of these key insights.

## Key Insights if you are looking for a career outside academia
- The importance of the networks that they had made during their times as academics. In addition, the importance of how to manage and grow a network. 
- Many jobs these days involve project work, and include generating the funding from donors as well as completing the project and writing the report. Post graduate degrees really help with learning how to start, manage, and complete projects.
- Although papers and citations gained during academic life won’t help with some jobs, they allow flexibility in the job market (potentially to re-join academia). They also demonstrate your ability to write. More papers are likely to improve your chances and some jobs include writing research papers as part of the job.
- Employers are interested in the experience and skills that you’ve acquired during your academic work. Instead of just listing papers you need to sell what you’ve done in cover letters and interviews:
  + What kind of experience do you have with your specialist area?
  + Do you understand about management?
  + Do you have good organisational skills?
  + Have you done fieldwork?
  + Have you managed students?
- Regardless of your academic background, you should expect to enter into your job at quite a junior level or even as an intern, and then work your way up
- Employers are looking for ‘emotional intelligence’ (the ability to understand and manage your own emotions, and those of the people around you). They will expect you to be a good team member, work with different stakeholders and clients. Conflict management skills are important. 
- You might not need to wait for a job advert. Use your contacts and write to people who are employers
- You might need to become ‘comfortable with feeling uncomfortable’: your work might be so different from what you’ve done before that you should and you should be able to adapt
- There are expectations from employers that you will meet challenges that your employers place before you (and not shy away)
  + Pay special attention to the Key Performance Areas (KPAs) of the position you apply for. These are what you will need to report on in order to have your performance in your job assessed.
- You might need to get used to different working cultures that are meeting focussed (even when you have meetings about having meetings)
- The working culture might not be static, and could change with the replacement of a manager or director.
- If you are employed by a governmental agency, it will be expected that you are accountable to your employers as well as the public that pays through their taxes
- The position you are employed in will likely involve you constantly acquiring new skills, such that you feel like more of a student then when you were studying. This really adds to the interest in the working life, and allows you to meet new and unexpected challenges.
- New subjects and other areas might be well outside your expertise, but can be just as rewarding once you rise to meet the challenges. 
- Jobs outside academia are especially challenging in working out how to apply the results of scientific studies. 
- Your employers might expect you to conceive your projects, as well as carrying them out.
- You are expected to be an authority in your work, and interpret your results in context and with reasonable confidence

At the end of the day, you need to discuss what you want and what you expect from your job outside academia with as many people as possible in that profession.

### 2021-09-05 5.00-lastnote.Rmd
Add files via upload
# all round update

# Last note {- #lastnote}

I really hope that this book has been helpful and that it has achieved what I set out to do: provide you with the guide on how to publish in biological sciences. If you feel that this book has important items missing, is out-of-date or simply wrong, then please [contribute](#contribute). Any good guide relies upon the people that use it to keep it viable. My special plea to those of you who are Early Career Scientists is to become part of the solution for shifting from the [closed to open models](#openscience) of scientific publishing. In doing so, you will make science more transparent, open and equitable. While there is much that is currently wrong in the system, the vast majority of the actors in it are working for the good of science. While I conceded that publishing science is a [wicked problem](#wicked) which is not easily solved, there are some simple steps that we can all make toward a resolution that benefits science.

When seeking a solution to publishing science, and publishing in other academic disciplines, we cannot compromise on the cornerstones of the scientific method: **rigour, independence, transparency and reproducibility**. This book should have highlighted aspects of scientific publishing that currently violate these cornerstones and so need to be changed. When you can, select those journals and publishers that maintain the [highest standards](#transparent). Join your disciplinary [academic society](#societies) and stand for election to a position. Issue your clarion call for change. Join the [gatekeepers](#gatekeepers) of the society's journals, and use your soft power to help bring about change.

1. Make more use of archiving services: [preprints](#preprints2), data and proposals
1. Make use of independent peer review services: peer community in, and  peer commons
1. Persuade your society to move to a [Diamond Open Access](#diamond) [overlay](#overlay) journal model
1. Become part of the solution

### 2021-09-05 _404.Rmd
Add files via upload
# all round update

# Page not found

**How to Publish in Biological Sciences**

The page you requested cannot be found (perhaps it was moved or renamed).

You may want to try searching to find the page's new location, or use the table of contents to find the page you are looking for.

**To continue, you can simply click on the contents (left),  or click on the image (below) to take you back to the beginning of the book.**

```{r fig.align='centre', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='https://johnmeasey.github.io/How-to-publish-in-Biological-Sciences/'}
knitr::include_graphics('figures/cover.jpg', dpi = NA)
```

### 2021-09-05 index.Rmd
Add files via upload
# all round update

--- 2021-04-24 index.Rmd
@@
 --- 
-title: 'How to write a PhD in Biological Sciences: a guide for the uninitiated'
+title: "How to publish in Biological Sciences: a guide for the uninitiated"
 author: "John Measey"
 date: "`r Sys.Date()`"
-documentclass: krantz
-bibliography:
-- book.bib
-- packages.bib
+documentclass: book
+bibliography: [book.bib, packages.bib]
 biblio-style: apalike
+csl: peerj.csl
 link-citations: yes
@@
 site: bookdown::bookdown_site
-description: You don’t have to be a genius to write a PhD. It will always involve
-  a lot of hard work and dedication, but the process of writing up is a whole lot
-  easier if you understand the basic ground rules. This book is a guide specifically
-  for writing a PhD in the biological sciences. It will guide you from deciding
-  whether or not you should actually do a PhD, provide practical, up-to-the-minute
-  guides to getting started with planning and writing, and consolidate this with the
-  nuts and bolts of writing for the biological sciences. The book guides you through
-  the do’s and don’ts of writing a thesis, and taking your thesis chapters to publication
-  in peer-reviewed journals. Whether you are a student just starting your studies,
-  or a supervisor struggling to cope, this book will provide the insider information
-  you need to get ahead.
+output: bookdown::gitbook
+description: "Publishing a paper in an academic journal should simply consist of submitting a publication worthy manuscript.
+But having a working knowledge of publishing will enable you to make better decisions about what, where and how to submit manuscripts.
+This all comes with experience, and in this book, I unravel areas of assumed knowledge and furnish them with explanations pitched at the Early Career Researcher, along with links and citations where you can read more.
+This book explores the wicked problem of academic publishing, and how to move from a closed to an open, transparent system.
+I explain the many choices that exist for those wishing to submit a manuscript for publication in the Biological Sciences.
+I explore the world of publication bias, and how this is evidenced by reviewers and editors.
+Impact Factors, citations and how the desire to track the performance of academics has led to unethical practices and exploitation of science and scientists.
+This guide provides an 'everything you wanted to know about publishing but were afraid to ask' approach for anyone who still feels that submitting a manuscript is like posting it into a black box.
+This book is written to get you onto an even footing."
 github-repo: yihui/bookdown-crc
 graphics: yes
+cover-image: images/Cover.png
+favicon: images/favicon.png
 ---
@@
 # Under Contract  {-}
-This book is currently under contract with CRC Press. 
-```{r fig.align='center', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/Writing_Publishing'}
-knitr::include_graphics('figures/Cover.png', dpi = NA)
+
+This book is currently under contract with CRC Press. This version was **submitted** for publication on: __27 August 2021__
+
+```{r fig.align='right', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/'}
+knitr::include_graphics('figures/cover.jpg', dpi = NA)
 ```
 
+**Cover illustration**: The Tower of Babel was adapted by the author from an original work by Pieter Bruegel the Elder (1563) painted to represent the demolition and scattering of peoples so as to impair their communication, much in the same way that modern day publishers have turned the scientific project into a business that profits only them.
+The collapse is coming, but can we save the scientific project?
+In my version, a group of Early Career Researchers can be seen (bottom left) deciding how to rebuild the crumbling edifice that represents scientific publishing (centre), with a ruined entrance guarded by gatekeepers (centre right).
+The skies above reflect the choice open to them; to stay in a closed system or to choose a future of open science.
+
 # Welcome {-}
-Welcome to this guide on how to write a PhD in biological sciences.
-I will not tell you what to write.
-My intention is to tell you how to approach writing with respect to style, format and formula (planning).
 
-Writing a PhD is a daunting prospect.
-One does not put pen to paper, or, rather, finger to keyboard, only after finishing all of your experiments and/or labwork.
-Writing is actually something you should start from day one of your PhD, or even before day one, when you make the first initial pitch of big ideas to your advisor on what you want to do.
-If you can find the discipline to start writing from day one, then you are much more likely to make a much better go of it than waiting until the final, pressured, stretch.
-But to do this, you need to understand how to write a PhD thesis.
-You need to understand what you are aiming to achieve, and, crucially, you need to understand why you are doing your PhD.
+Welcome to my guide on how to publish in the biological sciences.
+This guide is pitched at the early career researcher.
+It is not going to tell you what to write, but to open and examine the black box of scientific publishing, and more broadly explore how this impacts the academic context.
+My intention is to demystify publication in the biological sciences, so that readers become aware of what is happening once they have submitted a manuscript, and how to better interpret the decisions made by colleagues who are reviewing and editing your work.
 
-This guide intends to demystify the writing process so that the eventual goal is clear, and provide the necessary information on how to get there.
-It takes you from those initial ideas, all the way through proposal writing, posing hypotheses, testing them, and then how to write it all up.
-
-This book started as a series of blog posts written by me for my students ([john.measey.com/blog](john.measey.com/blog)).
-I wrote them because I felt that I was repeating the same advice year after year.
-Instead of rewriting comments on manuscripts time after time, I thought I would consolidate them and put them all in one place.
-Later, I started looking for particular problems that students were having, and then writing blog posts to try to preempt those problems, and to provide solutions before tasks were completed.
-Finally, I have tried to fill in the missing gaps in the blog posts.
-But there are still plenty of gaps, and I regard this book as a work-in-progress.
-It is my intention that this book will always be added to, and improved by a community of practice, and I hope that this will include you.
+Publishing has become vital for all academics, such that it is widely recognised that we inhabit a 'publish or perish' academic landscape.
+For some the process appears effortless, while for others publishing represents a black box leaving them outside in a highly stressful environment.
+This book is meant to be a guide to those uninitiated members of the academic community, postgraduate students and early career researchers, to bring them up to speed with all the necessary background information on publishing, providing links and references for reading and learning more.
 
@@
 
-Embarking on a PhD is intimidating as, for most students, it will be their first experience working within the academic system.
-The voyage of discovery is often made very frustrating as much of what goes on in academia is assumed knowledge.
-Academics accumulate knowledge throughout their careers, but what can be done for those who are uninitiated?
-What is needed is a guide.
-In the same way that we had guides to foreign countries and cities, postgraduate students need a guide that newcomers can refer to before, during and when making decisions about their time within academia.
-Note that this is not a rulebook.
-There are times when the guide will be accurate and others when it will be vague, but providing some insight to point you in directions where you can explore more.
+Publishing a paper in an academic journal should simply consist of submitting a publication worthy manuscript.
+But having a working knowledge of publishing will enable you to make better decisions about what, where and how to submit manuscripts.
+This all comes with experience, and in this book, I try to explore the areas of assumed knowledge, and furnish them with explanations pitched at the Early Career Researcher, along with links and citations where you can read more.
+I explain the many choices that exist for those wishing to submit a manuscript for publication in the Biological Sciences.
+Where possible, information in this book are tailored for the biological sciences, and when this information was not available for the Life Sciences, STEM or science (in that order).
+I explore the world of publication bias, and how this is evidenced by reviewers and editors.
+Impact Factors, citations and how the desire to track the performance of academics has led to unethical practices and exploitation of science and scientists.
+This guide provides an "everything you wanted to know about publishing but were afraid to ask" approach for anyone who still feels that submitting a manuscript is like posting it into a black box.
+This book is written to get you onto an even footing.
+
+## What's not in this book {-}
+
+Depending on just how early you are in your career, there is a lot missing from this book that has been provided in another book [**How to write a PhD in Biological Sciences**](www.howtowriteaphd.org).
+That book concentrates on getting PhD students started writing data chapters, while this book concentrates on publishing manuscripts.
+Hence, if you want extra information about writing in the biological sciences, I would point you to the other book.
+If you are happy with what you have written, but want help to demystify the publishing process, then this is the right book for you.
+
+There is common ground in both books, and I will point to important chapters relevant to publishing in the other book that are not reproduced here:
+[Plagiarism](http://www.howtowriteaphd.org/plagiarise.html), [Being aware that you can get it wrong](http://www.howtowriteaphd.org/TypeI.html), [Transparency](http://www.howtowriteaphd.org/transparency.html), and many more.
 
@@
 
-This book is intended to be a guide.
-Don't sit down and read it from cover to cover.
-If you do, then I think you have likely misunderstood what this guide is for.
-Instead, I want you to think about this book as a guide much in the same way that you might pick up a guide to a foreign city, country or even like Douglas Adams' fabled electronic guide _The Hitchhiker's Guide to the Galaxy_ @adams2017ultimate (if you are not familiar [see here](https://en.wikipedia.org/wiki/The_Hitchhiker%27s_Guide_to_the_Galaxy)).
-However, instead of having hostels or restaurants recommended by name with a description of what makes them particularly praiseworthy, this guide provides a generic idea of what such places are for, what they tend to be like, the best ways to get to them, and pitfalls to avoid.
+This book is written in four parts:
 
-None of the topics is explored in any great depth.
-This is deliberate as this guide is here to provide you with a starting point for you to explore if you think it is necessary.
-I have provided literature (where possible), and you should use this in the standard way to research the topic yourself should you need or want to know any more.
-In the same way, none of the advice provided in this book should override what you decide for yourself or with your advisor.
-Take your decisions from an informed position.
+[**Part I - Getting your manuscript ready for submission**](#part1)
+Although you may have already done your research, and written your manuscript, getting it ready for publication will require a new set of hurdles for you to jump over.
+In this section, I discuss what you need to know before entering into the publication arena.
+What are scientific journals for, and who are the gatekeepers?
+How does peer review work?
+The publishing world is at a turning point, and before you start publishing you should be aware of the current reality in Biological Sciences around the currency of citations and how these relate to other metrics such as the Impact Factor and career advancement.
+You also need to know potential directions for publishing, including the need for transparency in your work, whether or not you should deposit your manuscript as a preprint, and who you should invite to be an author.
+Tactical chapters for Early Career Researchers provide information on how to actively build and maintain a network to facilitate and support your work.
 
-**The book is divided into four main parts:**
+[**Part II - Submission, reviews and reviewing, revising and resubmitting**](#part2)
+Sending a paper to a journal is like posting it into a black box where, after some time, you might simply get a rejection and have no idea what has happened.
+In this section, I take you through the mechanisms of submitting a manuscript from choosing the right journal for your submission, writing a letter to the editor, suggesting reviewers, entering metadata about your manuscript into the editorial management software, all the way to pressing the submit button.
+I explain how the editorial submission system works, and what you can expect from editors and peer reviewers.
+I take a practical approach to writing a rebuttal and explain how and why you should expect to revise your manuscript for the editor.
+The eventual goal of this section is to demystify the entire process between submission and acceptance, and to understand the process from the viewpoint of an author, editor and reviewer.
 
-[**Part 1**](part1.html)
-Part one is written as something that you should read early on in the process, once you’ve committed to doing a PhD.
-This shortest section of the book deals with what a PhD is, how it’s examined, with whom, where and when to do it.
-At the start of your PhD is the time to be thinking of the bigger questions both in your study area, and your life, and this section provides some pointers and things to consider.
+[**Part III - Once your paper is published**](#part3)
+Once your article is accepted, you can celebrate together with your co-authors!
+You will need to submit the final version of your manuscript, have this type-set and then approve the proofs before a Version of Record appears.
+At this point, you can start to share your paper, but there are still some key steps that you can take to improve the dissemination of the research both to the academic community, to your funders, and the public at large.
+Who is it best to share your research with, and what would be the best form to share it in?
+In this section, there are chapters that explain how to write a press release and a popular article on your paper, and how you can improve and monitor its circulation both in academia and in the general media and social-media, focussing on those stakeholders who might use your findings.
 
-[**Part 2**](part2.html)
-Part two of the book sets out a practical guide to getting started with writing.
-The approach I've taken here is describing the production of a PhD in biological sciences which is made up of data chapters, and where each chapter is essentially a potential publication in a science journal.
-If you didn't want to do a PhD but just wanted to write a publication, then most of what is here (and in [Part 3](part3.html)) will be of use as you could take the same formulaic approach as given in this book.
-Similarly, if you were writing a MSc project, this section would also be useful.
-There is a lot of technical stuff involved in writing science, and in part 2 I cover as much as I can while still remaining general to the subject area.
-
-[**Part 3**](part3.html)
-A typical PhD will consist of five data chapters where each data chapter is a publication.
-There is a shortish introductory chapter (sometimes a literature review) that comes before these data chapters, and a concluding synthesis chapter at the end.
-That's not to say that this is the only way of writing a PhD.
-There are many different ways and it is a good idea for you and your advisor to talk about this together before you start.
-
-This part covers how to write the core sections of a paper: abstract, introduction, methods and materials, results, discussion and references.
-
-[**Part 4**](part4.html)
-The concluding part of this book is written to cover the things that remain after you have finished writing all of your data chapters.
-Most importantly, these include the introductory and concluding chapters that sit as bookends to the data chapters that you have already produced.
-These chapters make your thesis into a 'body of work', and allow you space to reflect on what you have achieved in your PhD studies.
-
-Throughout this guide you will find links and references of things to read elsewhere.
-The reality is that there is a lot out there for you to read.
-I am not the only person to write a guide, and I can't pretend that all the answers are here in this book.
-You will find that in your particular situation you need to look around.
-What I aim to do in this book is provide you with enough background information to help in your onward search, and provide you with enough inside knowledge to do a PhD.
+[**Part IV - Further challenges in academia**](#part4)
+ The last part of this book discusses the wicked problem posed by current publication models in academia.
+This section deals with the growing number of issues driven by a 'publish or perish' culture, and what this means for Early Career Researchers.
+Special focus is given to the paywall erected by many publishers, Open Access publishing and predatory publishers.
+I also explain the problems in the current system of biases in peer review, and the confirmation bias in scientific publishing.
+Instead of just presenting you with problems, this section provides insight into ideas that the academic community has produced in order to get over the current problems.
+Other important hurdles that you might meet, such as retractions, fraud and bullying, receive in depth focus.
 
 ## Why 'A guide for the uninitiated'? {-}
-
-I think that most people with doctorates would agree that a PhD is not awarded to people because they are particularly bright or smart.
-If you had to be a genius then I wouldn't have a PhD.
-Indeed, I don't consider myself to be particularly clever, but I worked very hard to get my PhD.
-I was hampered by the fact that I didn't know anything about the goals and aims of the academic process of working towards a PhD, so it took a lot more work, wasted time, and (let's not mince our words) real pain.
-The end product was a fraction of the potential that I could have achieved, if I had understood more about the process.
-If I had only had a guide to tell me what it was all about, I could have saved myself so much time and energy.
+Early Career Researchers are within eight years of getting their PhD or within six years of their first academic job.
+At this time, you will have already experienced the academic life, including publishing, but there will be far more to it than you are aware of.
+This book considers Early Career Researchers as colleagues who simply lack the experience of a system that has changed in many ways over the last 20 years.
+To those of you who know the current publishing scene in biological sciences, it offers the perspective of where things have come from.
+I have written this book as I feel that I would have been able to achieve more had I understood more about the publishing process early on in my own career.
+If I had only had a guide to tell me what it was all about, I could have saved myself so much stress, time and energy.
 In short, I feel that I was uninitiated, and this is the guide I wish that I had had.
-
-So, this guide is my practical attempt to help you to get the most that you can out of your PhD time.
-After all, it should be an amazing journey that culminates with you achieving academic maturity: the ability to conceive and carry out scientific investigations.
-When you have a PhD, you become a scientist.
-Too often, however, it's a nightmare journey of cul-de-sacs, and groping in the dark.
-
-You should be aware from the outset that doing a PhD is a major undertaking.
-It will take several years of your life and will impact heavily on your day to day life, to the extent that your mental health may suffer.
-A PhD represents the culmination of an incredible amount of work, not to mention that you've already had to complete one to three previous degrees beforehand.
-It may not be something that you ever finish to your own satisfaction.
-But finish it you must.
+So, this guide is my practical attempt to help you; to get you up to speed in the world of academic publishing, specifically in the biological sciences.
+After reading this book, I hope that you will avoid the nightmare world of publishing of constant effort ending in dead end rejections that so many academics describe.
 
@@
 
-Before I go on, there are a great many people that I need to thank.
-First and foremost are my students, past and present, who have inspired me to put together first the blog posts and then the book.
+There are a great many people that I need to thank.
+First and foremost are the members of my lab, past and present, who have inspired me to put together first the blog posts and then the book.
 It is because you wanted more that I put this together.
-I have also been a student, and have been inspired by colleagues around the world who have been exemplary advisors.
 This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help.
-I thank this greater academic community for sharing and inspiring.
+I thank this greater academic community (especially \#academicTwitter) for sharing and inspiring.
 Thanks go to the many reviewers and editors who have taken their time to improve my writing.
@@
 Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning.
-James Baxter-Gilbert, Anthony Herrel,  Allan Ellis, James Vonesh all read parts and commented.
+James Baxter-Gilbert, Dan Bolnick, Jack Dougherty, Anthony Herrel, Michael Hochberg, Allan Ellis, Rachael Lammey (CrossRef), Andrea Melotto, Lisa and Mark O'Connell, Ivan Oransky (Retraction Watch), Heather Piwowar (ImpactStory), Claire Riss (Center for Open Science), Johan Rooryck (cOAltition S), James Vonesh, Carla Wagener all read or commented on different aspects of the book.
 Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and Stellenbosch University.

### 2021-09-05 1.00-Part1.Rmd
Add files via upload
#general update of all files

@@
+# (PART) Part 1: Right from the very start {-}
+
 \mainmatter
 
-# (PART) Part I: Getting ready for submission {-}
+# Introduction  {#part1}
 
-# Submitting your work to a journal {- #part1}
-
-Part 1 of this book is about the publication arena.
-I provide simple explanations of what scientific journals are for, and who the gatekeepers are.
-The publishing world is at a turning point, and before you start publishing you should be aware of the current trends and what is possible.
+Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you plan to do afterwards.
+There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make informed decisions from the outset.
+For example, reconciling yourself to using the [statistical package R](#useR) will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on.
+Preparing yourself into the right mental framework will also be very important for your ability to work under pressure.
+Getting into good habits at the beginning will put you in a commanding position later.
+The aim of this first part of the book is to help you assess your current position, and realign yourself toward an optimum mindset.

### 2021-09-05 1.01-beforeyoustart.Rmd
Add files via upload
#general update of all files

@@
----
-output:
-  html_document: default
-  pdf_document: default
----
-
-\mainmatter
-
-# Introduction  {#intro}
-
-Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you plan to do afterwards.
-There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make informed decisions from the outset.
-For example, reconciling yourself to using the [statistical package R](#useR) will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on.
-Preparing yourself into the right mental framework will also be very important for your ability to work under pressure.
-Getting into good habits at the beginning will put you in a commanding position later.
-The aim of this first part of the book is to help you assess your current position, and realign yourself toward an optimum mindset.
-
-## So you are doing a PhD? {#beforeyoustart}
+# So you are doing a PhD? {#beforeyoustart}
 

### 2021-09-05 2.00-Part2.Rmd
Add files via upload
#general update of all files

--- 2021-07-20 2.01-Hyp.Rmd
@@
+# (PART) Part 2: Get writing {-}
+
 # How to get started with writing  {#part2}
@@
 Later, you may need to refer back to sections once you’ve started writing.
-
-## How to write a hypothesis  {#hypothesis}
-
-When you read about the formula of writing a chapter (see [end of part 2](#formula)) you will see that everything in your chapter or paper revolves around the hypothesis or questions that you ask, normally at the end of your introduction (Figure \@ref(fig:Hypothesis)).
-In this section, we will take a look at how to write a hypothesis.
-This is a sticking point for many students.
-We are used to using and writing questions and statements in day-to-day communications, as well as reading popular media.
-But hypotheses (the plural of hypothesis) only rarely float across our desks.
-So how do we write one, and how do we know if our hypothesis is good?
-
-In addition to this section, there is some good information out there on the web, and it's worth looking at this too: (e.g.
-[Wikihow](https://www.wikihow.com/Write-a-Hypothesis), [Wikipedia](https://en.wikipedia.org/wiki/Hypothesis), etc.).
-There's also some less good stuff out there, so read critically.
-
-(ref:Hypothesis) **Generating a good hypothesis for your PhD chapter isn't easy.** A good hypothesis will be invaluable in helping you write the chapter.
-It is important to start working with hypotheses.
-
-```{r Hypothesis, echo=FALSE, fig.align='center', out.width='90%', fig.cap="(ref:Hypothesis)"}
-knitr::include_graphics('figures/hypothesis.jpg')
-```
-
-### What is a hypothesis?  {#whathypothesis}
-
-A hypothesis is a statement of your research intent.
-It tells the reader (because just like all of your other written work, it is intended for an audience who will read it), what you plan to do in your research.
-But there's a little more to it than this.
-The hypothesis becomes a part of the scientific method if it is testable and (importantly) falsifiable, as well as being informed from previously published work on the subject.
-
-Your hypothesis must be informed by the literature, which is why you spent so much time and effort crafting your introduction to inform your reader of the same.
-This is also why your hypothesis usually comes at the end of your introduction, because you spend all of the introduction telling your reader about it.
-There's not much point in writing more after the hypothesis, because once your reader has read that, they are ready to learn about how you went about testing it (in the [Materials & Methods](#methods)).
-The other important point to make is that the literature should dictate how you write your hypothesis, and the variables that you include.
-If, for example, you know that temperature is the most important variable, but all of the literature suggests that it is oxygen, you can't ignore oxygen and you should also frame your hypothesis using this variable (you can have more than one hypothesis after all!).
-In this case, you will also need to provide a sufficient introduction to temperature as a variable to justify its inclusion in your hypothesis.
-Perversely, your aim is not to prove that your idea is right, but to show that the hypothesis can be refuted.
-
-We try to write a hypothesis that is falsifiable: i.e.
-you can prove (usually using statistical tests) that it is not correct (or at least show that the likelihood that it is correct is very low).
-That's why it is conventional to provide the 'null hypothesis' that is the falsified version of the statement, suggesting that there is no relationship between the variables you have proposed to measure.
-The convention is to label this 'null hypothesis' H~0~, while the 'alternative hypothesis' (the one that says your variables are related as you suggested) is written as H~1~.
-When you formulate your hypothesis, it is traditional to write your alternative hypothesis to indicate the directionality of your tested variables.
-This way, the reader can simply imply that the null hypothesis is when there is no relationship, but this will need to be stated if the null hypothesis is more complex.
-
-Karl Popper [-@popper2005logic] was the philosopher who proposed that without being able to refute or falsify a scientific problem, it ceases to be scientific.
-This is the reason for our null hypothesis.
-If the null is not available as a possible outcome, then logically, there is no science.
-
-> Karl Popper [-@popper2005logic]:  
-> **"...it must be possible for an empirical scientific system to be refuted by experience". **
-
-Most importantly, your hypothesis must come first, before you do the experiment or study.
-Hence the reason why this section comes at the start of part 2.
-Setting the hypothesis after the work is already done is fraudulent, and goes against the scientific method.
-Obviously, it isn't fair to pose the hypothesis once you already know the answer (also known as [HARKing](#TypeI)).
-This is why there is so much emphasis put on formulating your hypothesis during your research proposal.
-Getting it right will determine what you do and how you test it.
-If you think of an extra hypothesis that would be really useful to test once you've already done your study, you can conduct a post hoc test, but this should have more stringent levels of statistical assessment.
-
-Writing a hypothesis isn't easy, but it is essential, and once you've understood what to do most of the rest of what you are writing for should make sense.
-
-### What a hypothesis isn't
-
-It is not a question and so should never have a question mark after it.
-
-It isn't really a simple prediction: if this then that.
-You will see on the internet that hypotheses are explained in this simple predictive framework.
-I say that a hypothesis is not a simple prediction because it lacks the mechanistic and scholarly aspect of a good hypothesis, which is what we want to achieve.
-
-### A formulaic way to start writing your hypothesis:
-
-__"If. then. because."__
-
-Later in the book, I will emphasise that you must have introduced all the variables that you plan to use to test your hypothesis in your introduction.
-This usually comes in the second paragraph of your introduction, where you emphasise the utility of the dependent variable/s (what you are planning to measure) and your independent variable (what you will manipulate).
-Both of these variables should then feature in your hypothesis.
-Next, by paragraph four, you will have identified the problem that you are interested in tackling.
-In addition, your introduction will provide all of the pertinent literature that has relevance to this hypothesis, giving the all important context.
-
-A simple way to consider making your hypothesis is to adopt an **"If.
-then.
-because."** construction where you add in your problem statement using your independent variable after 'if', and your prediction using your dependent variable after 'then', and finally the expected mechanism after 'because'.
-Using our example above with the "If.
-then.
-because." construction, we would say: "If environmental temperatures in which tadpoles develop are increased then tadpole development rate is faster because they follow the classic metabolism of ectotherms".
-Both independent variable (temperature) and dependent variable (tadpole development rate) are present in this hypothesis, and the predicted relationship between them is clear.
-In addition, the causal mechanism is stated.
-I say that this is a formulaic way to start writing your hypothesis, because it usually ends up as an inelegant statement, which can be better refined for a reader.
-A citation for your stated mechanism might also help clarify exactly where the justification for this comes from.
-
-Mechanisms (or causal explanations) fall into three main areas: endogenous, exogenous and evolutionary [@allen2017scientific].
-
-#### Endogenous causal explanations
-
-Endogenous causal explanations focus on the mechanisms happening inside an organism, such as physiological processes, hormones, reproductive state, etc.
-
-#### Exogenous causal explanations
-
-Exogenous causal explanations concern mechanisms that are outside the body of individuals.
-Common exogenous mechanisms are climatic factors (temperature, humidity, precipitation, etc.) or may relate to the availability of food, predators or mates.
-
-#### Evolutionary causal explanations
-
-These mechanisms have evolved through time, and often relate to exogenous mechanisms triggering endogenous processes over multiple generations.
-
-Note then that the above mechanisms are not mutually exclusive in their nature, and it may be useful to combine different approaches within biology to ask hypotheses across all of these levels.
-Mechanisms in biological sciences are rarely simple or act on multiple organismal levels, so designing a controlled experiment in order to test a specific mechanism thoroughly can be very demanding.
-In other words, can you be sure that the cause is really responsible for the effect that you are measuring?
-
-A good hypothesis will often take an existing hypothesis further, to try to better refine the knowledge on a subject.
-Hence, it is perfectly acceptable to state that you are building on existing hypotheses (and giving the appropriate statement) when making your own.
-
-### Teleological versus causal hypotheses
-
-A teleological argument refers to the reason or a purpose of a particular process.
-For example, you may measure vertical migration of water fleas and suggest that diurnal migrations are made because the water fleas want to avoid predation.
-This is a teleological hypothesis because you are suggesting that the reason behind a process is the desire by water fleas to avoid predators.
-Although a reduction of predation may be a consequence of vertical migration in water fleas, each water flea does not think about predation and then starts it’s upward movement as a result.
-A common mistake made in biology is to apply teleological arguments to processes that have no purpose or reason.
-Evolution is often mistakenly suggested to have a purpose (e.g.
-to evolve to a more advanced state), but in fact, evolution is not a goal-orientated process.
-There is no end-point to evolution, and evolution did not start in order to meet some predetermined form or function.
-On the other hand, a causal hypothesis focuses on the factors about A that cause B.
-
-You should have realised that biologists are principally interested in causal hypotheses, because most mechanisms that are studied in biological sciences have no predetermined goal.
-If you are a behavioural ecologist, then you will need to be particularly aware of these two types of hypotheses, and when teleological explanations may be appropriate: many types of behaviour are goal orientated.
-
-### How to evaluate your hypothesis
-
-Once you've written your hypothesis, how do you decide whether or not it is good?
-To do this, you might think that you need plenty of experience (and yes, that does help).
-But really, you just need to look for the elements that are discussed above.
-So once you've written your hypothesis, try to objectively answer the questions below (for more see [here](https://youtu.be/9KoSGQlIFkM)):
-
-- Is there a clear prediction (if. then. statement)?
-- Does the prediction use independent and dependent variables correctly?
-- Is the mechanism supported by the literature?
-- Is the hypothesis testable/falsifiable?
-- Does the hypothesis use concise wording and precise terminology?
-
-If your hypothesis meets all of the criteria above, then you've done a good job!
-
-Probably one of the hardest issues that you will face in biological sciences is to determine whether your dependent variable is reliant only on your independent variable of choice.
-For example, a lot of variance in biological sciences will relate to the climate (especially with global change studies), but if your independent variable is temperature, this means that you will need to keep all other climatic variables the same.
-That is, if temperature is your independent variable, it is the only variable that can change in your experiment.
-This type of experiment is challenging as temperature often affects other variables (especially that they may vary in an unpredictable way).
-As soon as you have more than one independent variables, you can no longer test your dependent variable because you don't know which independent variable it is reacting to.
-Isolating variables is notoriously difficult, especially when we move from the laboratory to the field.
-You will need to think very carefully about what variables other than those of interest are potentially impacted by your experimental design.
-If you cannot control for them, this will likely mean that you need to change your hypothesis, or change your experimental design.
-If you are unsure, then I would encourage you to look carefully at the experiments that others have conducted in the literature.

### 2021-09-05 2.01-Hyp.Rmd
Add files via upload
#general update of all files

@@
-# How to get started with writing  {#part2}
-
-__Before you start writing__
-
-The writing part of the PhD is often the most difficult for many students.
-I think that there are several reasons for this.
-Writing well comes with practice, and most PhD students have had little experience with academic writing even if they have already completed an MSc.
-The same could be said of most early-career scientists (even after their PhDs), because practice is measured in completed pieces, it usually takes many years rather than weeks or months.
-You shouldn't expect that scientific writing will come automatically or immediately.
-In order to get this practice, you need to get started, and (again) many students find starting to write challenging.
-This guide was written to help you get started.
-
-However, before you start writing, it's a good idea to understand some of the ways in which academic writing differs from other writing styles.
-Good planning is the key here, along with knowing exactly what it is that you'll be writing about.
-That sounds facile, but at the foundation of each scientific study lies the hypothesis, which is central to being able to build the rest of the chapter.
-
-In this part of the book, I start with an explanation of the hypothesis.
-As you will see, the hypothesis is central to all of the writing in each chapter or paper, and you need to have a very good idea of how to formulate your research question into a hypothesis before you begin.
-Then I build on helping you think through lots of different aspects of writing in science: style, citations, writing concisely, scientific names, and lots more.
-Finally, I introduce you to the idea of using a formula to structure your writing.
-
-Before you start writing in earnest, it’d be good to read this part of the book and make sure that you are familiar with the ideas herein.
-Later, you may need to refer back to sections once you’ve started writing.
-
-## How to write a hypothesis  {#hypothesis}
+# How to write a hypothesis  {#hypothesis}
 

### 2021-09-05 2.12-taxonomy.Rmd
Add files via upload
# corrections after July 2021 Copy Edits
Add files via upload
#general update of all files

@@
 This is actually deliberate.
-Tschudi, 1838 isn't in brackets because the genus name hasn't changed since the description by Tschudi [1838](tschudi1838classification) placed *S.
+Tschudi, 1838 isn't in brackets because the genus name hasn't changed since the description by Tschudi [-@tschudi1838classification] placed *S.
 capensis* in *Sclerophrys* in 1838.
@@
 
-### View 1
+### View 1 - Yes, use capitals for common names
 
@@
 The Western Leopard Toad is not any toad, and the African Clawed Frog isn't any frog.
-The Cape platanna is any platanna living in the Cape, but the Cape Platanna is *Xenopus gilli*!
+The Cape platanna is any platanna living in the Cape, but the Cape Platanna is *Xenopus gilli*.
 In this last example, you can see how capitalisation reduces ambiguity and differentiates between adjectives and nouns.
 
-### View 2
+Personally, I find that capitalising common names helps them stand out in the text, and as a reader it helps me immediately recognise that a word is a name, and not a regular adjective.
+As many common names contain strings of words including adjectives, I find that I might need to re-read a section when I realise that a common name is being used, and that the author is not describing an attribute of the species.
 
-Common names are not formal names (like scientific names) so why does it matter?
+Note that many common names are hyphenated, and if this is the case then you should only capitalise the first part of two in a hyphenation: i.e.
+**White-throated Sparrow** (_Zonotrichia albicollis_), not White-Throated Sparrow.
+
+### View 2 - No, don't use capitals for common names
+
+Common names are not proper nouns (like scientific names) and so should not be capitalised.
 Many journals do not capitalise common names and insist on lower case unless they are proper nouns, like African in African clawed frogs.
@@
 
+Other reasons why capitalisation might not be implemented for common names include:
+
+- They are not proper nouns, and so (according to traditional English grammar rules) should not be capitalised
+  - Countries publish style manuals for what is acceptable (e.g. see [here](https://www.stylemanual.gov.au/)), and these often state that common names should not be capitalised
+  - I suspect that many publishers follow these official national guidelines when making decisions about scientific journals
+  - Many publishers and their editors hate superfluous capitalisation and attempt to eliminate it. This simply comes from sloppy writers who tend to overcapitalise. 
+- The capitalisation of common names implies some sort of official recognition: that it is **the** common name, and not **any** common name.
+Some groups, such as ornithologists, have only one common name for each species that they officially recognise.
+They take this to absurd lengths when they insist that all common bird names **must** be capitalised while common names for other taxonomic groups must not be capitalised (see [here](https://www.worldbirdnames.org/new/english-names/spelling-rules/capitalization/)).
+  - Even when ornithologists get their own way, and bird names get capital letters, these are changed to lower case when they are in a list with other animals and plants that carry lower-case (see [here](https://www.stylemanual.gov.au/style-rules-and-conventions/names-and-terms/plants-and-animals#write_common_names_in_lower_case_and_roman_type))
+- You may notice that Wikipedia does not use capitalisation for common names (see [here](https://en.wikipedia.org/wiki/Wikipedia:Manual_of_Style/Capital_letters#Common_names)).
+They state that this is because they avoid superfluous capitalisation in English (much as editors do - see above).
+
 On balance, I'd say that it's likely that your usage will be dictated by the policies of your faculty, advisor or journal.
+
+### Changing scientific names to English names
+
+There are many examples of names of scientific taxa that are changed to English derivatives of the original Latin names (see Table \@ref(tab:derivative)).
+While the official Latin version of the scientific name almost always carries a capital letter (see [above](#taxonomy)), these anglicised versions do not.
+
+Table: (\#tab:derivative) **Latin names can be changed into English versions.** You may not realise it, but many English names have been derived from scientific names.
+While the Latin names usually carry capital letters (see [above](#taxonomy)), these anglicised versions do not.
+
+| Taxonomic rank | Latin name | English derivative |
+| ---- | -- | --- |
+| Kingdom | Animalia | animal |
+| Phylum | Arthropoda | arthropod |
+| Order | Carnivora | carnivore |
+| Superfamily | Vespoidea | vespoids |
+| Family | Bufonidae | bufonid |
+| Subfamily | Felinae | feline |
+| Genus | _Eucalyptus_ | eucalypt |
+
+Turning a Latin taxonomic rank into an English derivative is a useful mechanism in which to bring variation into your writing.
+While there are distinct rules about the use of taxonomic ranks (see [above](#taxonomy)), there are not with these anglicised derivatives, and you can use them as you want.
+Even if they haven't been used before for your study group, there is no reason why you should not erect an English derivative for your own purposes - as long as this enhances readability and does not cause more confusion.

### 2021-09-05 2.17-plagiarise.Rmd
Add files via upload
#general update of all files

@@
 This is not what we're looking for in plagiarism.
+
+## Free online plagiarism tools
+
+If your institution doesn't subscribe to a plagiarism tool, then you can check your work online free using one of a number of tools available.
+There are advantages and disadvantages to each, so I suggest that you try them out, and see which suits your needs best.
+
+- **Grammerly** ([www.grammarly.com](www.grammarly.com)) Checks your work for plagiarism for free, but you need to create an account (even for the free version) and you have to pay if you want it to check your grammar.
+- **Plagiarisma** ([plagiarisma.net](plagiarisma.net)) You'll need to create an account (to see where your copied text came from) and paste your text into a box, but you'll get a good sense of whether or not there is plagiarism in your text.
+- **PaperRater** ([www.paperrater.com](www.paperrater.com)) No need to create an account to paste text into a box (but it didn't recognise my content which is also on my blog and so doesn't get top marks from me).
+Some nice different output with suggestions on readability and grammar.
 

### 2021-09-05 3.00-Part3.Rmd
Add files via upload
#general update of all files

--- 2021-07-20 3.01-Title.Rmd
@@
+# (PART) Part 3: The formula of a chapter {-}
+
 # Writing the sections that make up your data chapter  {#part3}
@@
 While you can leave the formatting of references until the end, it’s dangerous to not put citations in as you write as you are likely to forget which paper is which by the end.
-
-## Title page {#title}
-
-Most of the following information in this section on the title pertain to getting a manuscript ready for submission to a journal.
-For your thesis, I would recommend that you have a title page for each chapter that lists any collaborators and their contributions so that this is transparent for the examiners.
-It also helps break up the document and can be a good spot for a non-academic illustration (if your institution permits).
-
-Nearly all journals will require you to have a title page for your manuscript.
-This may or may not include authors.
-Check to see whether the journal that you are submitting to conducts double-blind review.
-If they do, they will require a title page without any indication of authors or institutions (they will also likely ask you to remove the acknowledgements section).
-This should be clearly specified in the instructions to authors of your chosen journal.
-
-### Names and addresses are important {#ORCID}
-
-Getting authors' names correct, their correct addresses (many people have more than one affiliation), can be tricky.
-Check that you have all of the information required for a title page before you submit.
-While you need the names of your co-authors to be displayed correctly, you do not need their titles (no Dr.
-or Prof., etc.).
-
-Indicate clearly who is the corresponding author.
-Normally, if you have done all of the rest of the work, the corresponding author should be you.
-You need to learn how to start this role at some point, so it might as well be now.
-Having the correct name and address will be important to each author.
-Along with the author names and addresses you should record the ORCID of each author.
-The ORCID is simply a unique identification code for individual researchers.
-ORCID is a non-profit organisation and there’s nothing sinister in signing up.
-If they (or you) don’t have one, then you should ask them to create one before the submission.
-It takes less than 5 minutes.
-Go to www.orcid.org
-
-## The title
-
-The title of your paper, chapter or book is the first thing that any reader will read, and so should be well considered.
-If your reader cannot quickly understand your title, the chances are that they won’t bother reading any further.
-Your title will be your selling point, and your aim is to use it to draw your readership in.
-Once you’ve managed to inveigle your readers to download your paper, your title is also their hook for remembering your paper in their database of thousands of others.
-Having their key-word in your title will help here, and as ever with writing your challenge is to think like your reader.
-The best titles are those that sum up the entire study in five to seven words.
-This is best done in a narrative that tells the story ([see part 2](#storytelling)) of your manuscript in its entirety.
-This may sound daunting, but you should get into the habit of summing up your story quickly (for friends, relatives and work colleagues).
-Then it’s a question of refining this story into the short single sentence that makes up the title.
-While the narrative approach may not work for you, you do want the title to provide enough information so that the potential reader knows what they will find before they open it.
-Your title doesn’t just have to work for you, it needs to work for a wide audience.
-
-Some people are excellent at writing titles that contain puns of well-known phrases or sayings.
-These can be brilliant, working both to inform what’s in the paper as well as providing some familiar input that helps retain them in memory.
-However, many fail to do either and are simply a waste of space.
-If you are tempted to use a pun as your title, make sure that it is widely appreciated, and not just among your co-authors and lab.
-
-You don’t have to come up with a killer title from day one.
-Most of my manuscripts have a working title that gets revised as I write, and is always open for change before submission.
-If you have great ideas for a title, do note them down.
-I find that the more options I have, the more likely I am to come up with something that works for everyone.
-It also helps to mix and match from a set of candidate titles.
-Once you have come up with something that looks good to you and your colleagues, test it by entering it into your database of choice (with the default being Google Scholar).
-Your first 10 results should include a set of papers that you have likely cited in the upper area of your introduction.
-If you don’t recognise any out of the top 10, it’s time to look at another of your candidate titles.
-
-### Some title ideas to start you off
-- Don’t start by looking for the best title
-  + Write a number of candidate titles and ask your co-authors to vote for their favourites.
-- The shorter and catchier your title can be the better: 6 words (±1) is an ideal.
-  + Allow yourself a longer subtitle if needed, but don’t go over 20 words total (some journals may limit your total to less). 
-  + Consider the (former) 120 character limit of a Tweet as an upper limit
-- Do include your principal finding if possible
-- Include as many key-words as you can
-
-### Things to avoid in your title
- - Don’t feel obliged to include [taxonomic terms](#taxonomy) unless it is relevant or compulsory
-   + Some (taxonomically minded) journals will insist on the species name followed by the taxonomic authority, and/or the family and order, in your title
- - Avoid obscure or specialist words that won’t be understood by your readership
-   + There are times when key words are necessarily specialised and your readership will expect this, but simpler words in your title will open up your readership which will otherwise remain narrow
- - Don’t simply define the scope of your work without including your content.
- - Don’t have your title as a question. Rather provide the answer!
-
-## Key-words
-The key-words are a way for readers to find your content with searches.
-Typically, the advice is to use words that are not in the title or abstract.
-This is because many databases have combined searching facilities for title, abstract and key-words.
-I struggle to think of appropriate key-words, and so make a list of some of the [big idea](#bigidea) words and short phrases from the introduction.
-Then I tend to look at articles in the same genre, and see what key-words they have used.
-As with the title, I suggest that you enter your chosen key-words into your literature database of choice [see part 2](#database) and see what comes back.
-You should see a group of papers that look wholly familiar and preferably those that are already cited in your manuscript.
-If not, it’s time to look again.

### 2021-09-05 3.01-Title.Rmd
Add files via upload
#general update of all files

@@
-# Writing the sections that make up your data chapter  {#part3}
-
-Getting over the cognitive inertia of starting to write is top of the list for most of us.
-The approach provided here is to present you with a formula that breaks down each section into smaller more reasonably sized chunks.
-However, you still need to make a start.
-Procrastinating starting to write is normal.
-It seems a massive task, even if you’ve done it lots of times before.
-You want to write the best manuscript that you can, but you aren’t feeling on top form today, so doesn’t it seem a better idea to wait until tomorrow?
-It’s not.
-You must start today and start now.
-Don’t set the bar so high for your first draft.
-In fact, don’t set any bar other than a personal goal of getting the first draft done.
-
-For the first draft, type away confident in the knowledge that 90% of the first draft will get junked or changed so much that you won’t recognise it.
-That’s ok, and it’s not a waste of your time.
-In fact, the first draft is invaluable in getting your ideas and perspective down on paper, allowing you to better organise and sift through your thoughts and bring clarity to your manuscript.
-Allow yourself a reward for finishing your first draft, something that you won’t do until it’s done.
-
-You can make life easier for yourself by changing the order of what you write.
-Although I have placed the contents of this chapter as you’d expect to see them in a typical paper, this is not the order I would suggest that you write them in.
-The most important part of writing up your chapter or paper is to have your hypothesis or question clearly stated.
-Once you have this, it will help you to know the literature that is already there and so writing the introduction next seems logical.
-Of course, you may already have [your proposal](#proposal) to work with, in which case you can start by reviewing what you already have and deciding whether or not you need to tweak this, or if you need a fresh start.
-
-The methods and materials are the easiest section to write first, especially if you've already done the practical work.
-This section will also help you focus on the analysis that you are doing for the results.
-Again you should review your proposal and see whether there were any significant changes to your planned methodology and note them down.
-This will not be an issue even if you have [preregistered your proposal](#transparency), you simply need to provide a rational explanation for why things changed.
-
-Next, write the results.
-Remember that you are only looking to respond to your hypotheses, preferably using your (preregistered) proposal plans.
-Tables and figures are compiled along with the results.
-Now you have the materials and methods together with the results in full.
-I suggest that you make a list of all the relevant discussion points you want to make and then turn your attention to the introduction.
-Begin with an outline, before you flesh it out.
-
-Lastly, write the abstract.
-Populate your article with references as you write.
-While you can leave the formatting of references until the end, it’s dangerous to not put citations in as you write as you are likely to forget which paper is which by the end.
-
-## Title page {#title}
+# Title page {#title}
 
@@
 
-Nearly all journals will require you to have a title page for your manuscript.
-This may or may not include authors.
-Check to see whether the journal that you are submitting to conducts double-blind review.
+Nearly all journals will require you to have a title page for your manuscript. This may or may not include authors. Check to see whether
+the journal that you are submitting to conducts double-blind review.
 If they do, they will require a title page without any indication of authors or institutions (they will also likely ask you to remove the acknowledgements section).
@@
 The ORCID is simply a unique identification code for individual researchers.
-ORCID is a non-profit organisation and there’s nothing sinister in signing up.
-If they (or you) don’t have one, then you should ask them to create one before the submission.
+ORCID is a non-profit organisation and there's nothing sinister in signing up.
+If they (or you) don't have one, then you should ask them to create one before the submission.
 It takes less than 5 minutes.
@@
 
-The title of your paper, chapter or book is the first thing that any reader will read, and so should be well considered.
-If your reader cannot quickly understand your title, the chances are that they won’t bother reading any further.
+The title of your paper, chapter or book is the first thing that any reader will read, and so should be well considered. If your reader
+cannot quickly understand your title, the chances are that they won't bother reading any further.
 Your title will be your selling point, and your aim is to use it to draw your readership in.
-Once you’ve managed to inveigle your readers to download your paper, your title is also their hook for remembering your paper in their database of thousands of others.
+Once you've managed to inveigle your readers to download your paper, your title is also their hook for remembering your paper in their database of thousands of others.
 Having their key-word in your title will help here, and as ever with writing your challenge is to think like your reader.
@@
 This may sound daunting, but you should get into the habit of summing up your story quickly (for friends, relatives and work colleagues).
-Then it’s a question of refining this story into the short single sentence that makes up the title.
+Then it's a question of refining this story into the short single sentence that makes up the title.
 While the narrative approach may not work for you, you do want the title to provide enough information so that the potential reader knows what they will find before they open it.
-Your title doesn’t just have to work for you, it needs to work for a wide audience.
+Your title doesn't just have to work for you, it needs to work for a wide audience.
 
-Some people are excellent at writing titles that contain puns of well-known phrases or sayings.
-These can be brilliant, working both to inform what’s in the paper as well as providing some familiar input that helps retain them in memory.
+Some people are excellent at writing titles that contain puns of well-known phrases or sayings. These can be brilliant, working both to
+inform what's in the paper as well as providing some familiar input that helps retain them in memory.
 However, many fail to do either and are simply a waste of space.
@@
 
-You don’t have to come up with a killer title from day one.
+You don't have to come up with a killer title from day one.
 Most of my manuscripts have a working title that gets revised as I write, and is always open for change before submission.
@@
 Your first 10 results should include a set of papers that you have likely cited in the upper area of your introduction.
-If you don’t recognise any out of the top 10, it’s time to look at another of your candidate titles.
+If you don't recognise any out of the top 10, it's time to look at another of your candidate titles.
 
 ### Some title ideas to start you off
-- Don’t start by looking for the best title
-  + Write a number of candidate titles and ask your co-authors to vote for their favourites.
+
+-   Don't start by looking for the best title
+
+    -   Write a number of candidate titles and ask your co-authors to vote for their favourites.
+
 -   The shorter and catchier your title can be the better: 6 words (±1) is an ideal.
-  + Allow yourself a longer subtitle if needed, but don’t go over 20 words total (some journals may limit your total to less). 
-  + Consider the (former) 120 character limit of a Tweet as an upper limit
+
+    -   Allow yourself a longer subtitle if needed, but don't go over 20 words total (some journals may limit your total to less).
+    -   Consider the (former) 120 character limit of a Tweet as an upper limit
+
 -   Do include your principal finding if possible
+
 -   Include as many key-words as you can
@@
 ### Things to avoid in your title
- - Don’t feel obliged to include [taxonomic terms](#taxonomy) unless it is relevant or compulsory
-   + Some (taxonomically minded) journals will insist on the species name followed by the taxonomic authority, and/or the family and order, in your title
- - Avoid obscure or specialist words that won’t be understood by your readership
-   + There are times when key words are necessarily specialised and your readership will expect this, but simpler words in your title will open up your readership which will otherwise remain narrow
- - Don’t simply define the scope of your work without including your content.
- - Don’t have your title as a question. Rather provide the answer!
+
+-   Don't feel obliged to include [taxonomic terms](#taxonomy) unless it
+    is relevant or compulsory
+
+    -   Some (taxonomically minded) journals will insist on the species
+        name followed by the taxonomic authority, and/or the family and
+        order, in your title
+
+-   Avoid obscure or specialist words that won't be understood by your
+    readership
+
+    -   There are times when key words are necessarily specialised and
+        your readership will expect this, but simpler words in your
+        title will open up your readership which will otherwise remain
+        narrow
+
+-   Don't simply define the scope of your work without including your
+    content.
+
+-   Don't have your title as a question. Rather provide the answer!
 
 ## Key-words
+
 The key-words are a way for readers to find your content with searches.
@@
 This is because many databases have combined searching facilities for title, abstract and key-words.
-I struggle to think of appropriate key-words, and so make a list of some of the [big idea](#bigidea) words and short phrases from the introduction.
-Then I tend to look at articles in the same genre, and see what key-words they have used.
+I struggle to think of
+appropriate key-words, and so make a list of some of the [big idea](#bigidea) words and short phrases from the introduction. Then I
+tend to look at articles in the same genre, and see what key-words they have used.
 As with the title, I suggest that you enter your chosen key-words into your literature database of choice [see part 2](#database) and see what comes back.
 You should see a group of papers that look wholly familiar and preferably those that are already cited in your manuscript.
-If not, it’s time to look again.
+If not, it's time to look again.
+
+### Selecting appropriate key-words
+
+Key-words are very useful in your studies, because if you have a good
+selection, they can help nail down a good proportion of the literature
+that you will need to read during your studies. Moreover, if you have
+the best selection of key-words, then you can set up some automated
+notifications for when new items are published. The only problem then is
+pulling together the correct key-words.
+
+When you submit a conference abstract, or a paper for publication, you
+will also be faced by a demand for key-words. This is when I often draw
+a blank. What will be the appropriate key-words for my study? I can
+often think of one or two, but regularly journals want at least five.
+
+A method for finding key-words Here is my method for finding appropriate
+key-words. It's quick, but it does require a piece of free-ware:
+[VOSviewer](https://www.vosviewer.com/), which for me has become an invaluable research tool.
+
+-   Go to your literature database of choice (i.e. Web of Science,
+    Scopus, PubMed, Dimensions.)
+
+-   Search for documents using the key-words that you are sure are
+    appropriate
+
+-   Constrain the results to \~500 documents
+
+    -   You can either do this with the search date (e.g. the most
+        recent 5 years)
+
+    -   Or you can simply take the first 500 documents that are found
+
+-   Download these \~500 documents in a tab delimited text file. In
+    VOSviewer Press the 'create' button
+
+-   Create a map based on bibliographic data
+
+-   Read data from bibliographic database files
+
+-   Select the file that you downloaded in the appropriate tab
+
+-   Type of analysis
+
+    -   Co-occurrence
+
+    -   All keywords (you can choose here)
+
+    -   Full counting
+
+-   You should see the total number of key-words in your downloaded file
+    now
+
+    -   Change the minimum number of occurrences of a key-word so that
+        you have \~100 results (again you can choose what suits you)
+
+-   Press finish
+
+You should end up with a network like the one below. Here I have used the key-words "invasive" and "fish" (Figure \@ref(fig:key-word-selection)).
+
+(ref:key-word-selection) **A network of key-words for "invasive" and "fish".** This network represents key-words that are repeated 10 times or more in the Web of Science together with "invasive" and "fish".
+
+```{r key-word-selection, echo=FALSE, fig.align='center', out.width='90%', fig.cap="(ref:key-word-selection)"}
+knitr::include_graphics('figures/key-word-selection.png')
+```
+
+In VOSviewer, you can highlight any one of these key-words and see exactly what combination they have been used in. I have selected
+examples that occur 10 or more times in my downloaded file. This means that I can be fairly sure that these are relatively common key-words to use in combination with the ones I know are good.
+
+The larger the panel in this network, the more frequently the word is used (see "invasive species" in Figure \@ref(fig:key-word-selection)).
+This should help you when you select your own key-words.
+For example, even though I had used the keywords "invasive" and "fish" to generate this network, one of the first things I noticed is that the term "invasive species" is far more common than "invasive".
+Hence, the first thing I should do is to change the first of my key-words.
+
+You should note in Figure \@ref(fig:key-word-selection) that some of the key-words relate to specific taxonomic groups.
+Others include the habitat in which the fish were sampled.
+Now I have a shortlist from which to pick the remaining 5 key-words that I need in order to submit my abstract or manuscript.
+Once you've made your selection, you can go back to your literature database, add this combination of key-words into a search and see what comes out.
+If you've done it right, you should see some familiar papers on similar topics to your own.

### 2021-09-05 4.03-publish.Rmd
Add files via upload
# corrections after July 2021 Copy Edits
Add files via upload
#general update of all files

@@
 
-```{r cover2, echo=FALSE, fig.align='centre', out.width='60%', fig.link='www.routledge.com/9781032116419'}
+```{r cover2, echo=FALSE, fig.align='centre', out.width='60%', fig.link='http://www.howtopublishscience.org/'}
 knitr::include_graphics('figures/cover2.png')
@@
 **The follow-up to this book, How to Publish in Biological Sciences: A Guide for the Uninitiated** is available at [www.routledge.com/9781032116419](www.routledge.com/9781032116419).
+
+You can read this book now and contribute on GitHub at [www.howtopublishscience.org](http://www.howtopublishscience.org/).

### 2021-09-05 _404.Rmd
Add files via upload
#general update of all files

@@
 
-The page you requested cannot be found (perhaps it was moved or renamed).
+The page you requested cannot be found (perhaps it was moved or renamed - most likely it's my fault and not yours).
 
@@
 
-```{r fig.align='centre', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='https://johnmeasey.github.io/How-to-publish-in-Biological-Sciences/'}
+```{r fig.align='centre', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://www.howtowriteaphd.org/'}
 knitr::include_graphics('figures/cover.jpg', dpi = NA)

### 2021-09-05 index.Rmd
Add files via upload
#general update of all files

--- 2021-08-01 index.Rmd
@@
 biblio-style: apalike
+csl: peerj.csl
 link-citations: yes
@@
 Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning.
-James Baxter-Gilbert, Jack Dougherty, Anthony Herrel, Allan Ellis, Andrea Melotto, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, James Vonesh, Carla Wagener all read or commented on different aspects of the book.
+James Baxter-Gilbert, Ellie Brown, Jack Dougherty, Anthony Herrel, Allan Ellis, Andrea Melotto, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, James Vonesh, Carla Wagener all read or commented on different aspects of the book.
 Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and in the library at Stellenbosch University.

### 2022-08-22 index.Rmd
Update index.Rmd

@@
 graphics: yes
+cover-image: figures/Cover.png
+favicon: figures/favicon.png
 ---

### 2022-09-25 2.07-storytelling.Rmd
Add files via upload
# updating chapters

@@
 From this study, I can spin the story (above), because I know what the endpoint is.
-But this storyis just one interpretation of the correlation that was found.
+But this story is just one interpretation of the correlation that was found.
 [Correlation is not causation](#argument), and more studies would be needed to test the hypothesis concerning the underlying reasons suggested for the relationship observed.
@@
 In addition, there are lots of lessons from storytelling that we can use within scientific papers, and to which I subscribe wholeheartedly.
-For example, don’t be boring, don’t [bamboozle](#style), write an active narrative (not in the passive tense), use the magic of threes, and many more.
+For example, don’t be boring, don’t [bamboozle](#style), write an active narrative (not in the [passive voice](#passive)), use the magic of threes, and many more.
 
@@
 If you like this way of thinking, you can get deeper into the depths of story thinking with Angler [-@angler2020telling].
+
+## Of babies, werewolves and silver bullets
+Another take on the need for a narrative to engage the reader when writing science is provided in a [blog post](http://ecoevoevoeco.blogspot.com/2014/10/how-to-writepresent-science-baby.html) by Andrew Hendry.
+These unlikely characters in a scientific document act as a metaphor to help writers remember that when communicating their results they require a subject area to pull the reader in (everyone is interested in a cute baby), then to introduce the central problematic that needs to be solved, or the villain to the peace of the cute baby (the werewolf).
+Lastly, the silver bullet is a metaphor for the approach taken that will resolve the threats of the werewolf to the cute baby.
+Once you've laid out your study with this narrative, you should be able to go on to show exactly how your silver bullet results in a dead werewolf, or at least a horribly wounded werewolf that is a lesser threat to the cute baby.

### 2022-09-25 2.08-citations.Rmd
Add files via upload
# updating chapters

@@
 
+But there may be a lot of papers that have found the same result, and you might want to only cite one as an example.
+
+>Invasive amphibians have some serious impacts on their recipient environments [e.g. @measey2016global].
+
 You’ll see that sometimes the names are brought into the sentence and become the central agents of the text.
@@
 
->Measey et al. [-@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals.
+>Measey and others [-@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals.
 
@@
 
->Measey et al.
-[-@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals.
+>Measey and colleagues [-@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals.
 They did this by constructing GISS scores for all individuals in all groups.
@@
 
-What about page numbers?
+### What about page numbers?
 Sometimes you'll see a citation with a colon and page number after. This really only needs to be used if you are quoting specific text on a particular page:
@@
 >Measey et al. [-@measey2016global: 976] proposed that using GISS scores could show that "some amphibians can have devastating impacts to the environment".
+
+## The problems with brackets {#brackets}
+Citations use brackets, and brackets are also used for other parts in writing.
+The problems with brackets often come when trying to combine citations that have brackets with the other parts of writing that also need the brackets.
+This is compounded when trying to do all this in [citation software](#referencesoftware) when you don't have complete control over it.
+
+Most [citation software](#referencesoftware) allows for flexibility of how you will use your citation.
+In the above examples we have seen the full citation, a version where just the year is included in the brackets, some text occurs before (prefix) or after (suffix) the citation within the brackets, or even a combination of these:
+
+ - The classic: [@measey2016global]
+ - Omitting the authors: Measey et al. [-@measey2016global]
+ - Using a prefix: [e.g. @measey2016global]
+ - Using a suffix: [@measey2016global: 976]
+ - A combination of the above: Measey et al. [-@measey2016global: 976]
+
+[Citation software](#referencesoftware) handles producing these different ways by omitting the author, adding a prefix, or adding a suffix, respectively.
+However, the default will be to use the classic version, and you will need to alter this to one of the others in the software as you are citing it (i.e.
+as you add it to the document).
+
+It is very common that people simply don't know how to do these different ways of citing within their [citation software](#referencesoftware).
+The easiest is to look up "how to add suffix" or whatever it is that you want to do in the respective manual or in your search engine.
+The result of not being sufficiently skilled in your [citation software](#referencesoftware), is that you end up with confused text - and usually a surfeit of brackets.
+The examples below are all mistakes that I regularly see in theses and manuscripts:
+
+>Measey et al. [@measey2016global] found that the impact of invasive amphibians is comparable to that of birds and mammals.
+
+>Invasive amphibians have some serious impacts on their recipient environments (e.g. Measey et al. [-@measey2016global]).
+
+>Invasive amphibians have some serious impacts on their recipient environments (e.g. [@measey2016global]).
+
+The above examples are obviously wrong (i.e.
+obvious to an examiner or reviewer), and immediately suggest that the writer hasn't invested enough time in mastering the [citation software](#referencesoftware), or isn't aware that what they have produced is incorrect.
+I'd encourage you to take the time to correct these kinds of errors in a timely manor, or at least before submitting your thesis or manuscript.
+
+### Avoid back to back bracketing
+
+It is better to combine information into one set of continuous brackets instead of having back to back sets of brackets. You can do this by using a prefix or a suffix as below:
+
+>Some invasive amphibians have been investigated insufficiently compared to their impact [e.g. Asian common toads:   @measey2016global].
+
+Instead of back to back bracketing as follows:
+
+>Some invasive amphibians have been investigated insufficiently compared to their impact (e.g. Asian common toads)[@measey2016global].
+
+If you must have brackets nested within each other (and I would argue that this is very unusual), then rather use different types of brackets, such as square brackets inside round brackets.
+But make sure that you do this consistently throughout your thesis or manuscript.
 
@@
 @wake2008are; @collins2009extinction; @pimm2014biodiversity], and read around those to find studies on amphibian decline that are appropriate for your context.
-This underlines one important aspect of choosing citations where the statement that you make relates directly to the study carried out in the citation.
+This kind of mis-citation has been called an 'empty citation' [@todd2007citing], and underlines one important aspect of choosing citations where the statement that you make relates directly to the study carried out in the citation.
+
+Todd et al.
+[-@todd2007citing] looked at mis-citation rates in ecology, concluding that only around three quarters of papers were correctly cited.
+They found that 11% of citations were ambiguous, and 7% had citations that did not support the statement.
+Around 5% of citations were classed as empty - citing works that cite works with relevance - also known as 'lazy author syndrome'.
 
@@
 You can even cite both when relevant.
+Todd et al.
+[-@todd2007citing] suggested that citing review articles might occur when postgraduate students do not have access to sufficient literature because of [pay-walls](#paywall).
 

### 2022-09-25 2.10-referencesortware.Rmd
Add files via upload
# updating chapters

@@
 
-Because published papers play such a fundamental role in your [reading and comprehension](#critical) of a subject, and [citations](#citations) are pivotal in chapter or paper writing, putting these elements together in a good reference manager makes a lot of sense.
+Because published papers play such a fundamental role in your [reading and comprehension](#critical) of a subject, and [citations](#citations) are pivotal in chapter or paper writing, putting these elements together in a good reference manager (aka citation software) makes a lot of sense.
 A reference manager is a piece of software in which you can store and view all of your papers (in pdf format or via links), make notes on them, and use them to make citations and build bibliographies.
@@
  - Is there good support available?
+
+## It's worth putting in the time to learn relevant parts of the package you choose
+
+These days, reference managers are powerful and you might not need all of the functions that they have to offer.
+However, it is worth learning immediately what you will use, and this might be more than you think at first.
+For example, [using brackets](#brackets) is not as straightforward as it seems as you might often have to use a prefix of suffix inside the brackets.
+If you don't sort this out from the start, you might get into a mess that you forget to fix later.

### 2022-09-25 3.01-Title.Rmd
Add files via upload
# updating chapters

@@
 
-The title of your paper, chapter or book is the first thing that any reader will read, and so should be well considered. If your reader
-cannot quickly understand your title, the chances are that they won't bother reading any further.
+The title of your paper, chapter or book is the first thing that any reader will read, and so should be well considered.
+If your reader cannot quickly understand your title, the chances are that they won't bother reading any further.
 Your title will be your selling point, and your aim is to use it to draw your readership in.
@@
 
-Some people are excellent at writing titles that contain puns of well-known phrases or sayings. These can be brilliant, working both to
-inform what's in the paper as well as providing some familiar input that helps retain them in memory.
+Some people are excellent at writing titles that contain puns of well-known phrases or sayings.
+These can be brilliant, working both to inform what's in the paper as well as providing some familiar input that helps retain them in memory.
 However, many fail to do either and are simply a waste of space.
 If you are tempted to use a pun as your title, make sure that it is widely appreciated, and not just among your co-authors and lab.
+
+In general, there are three types of title:
+
+-   Declarative - indicating the results of the work.
+
+-   Descriptive - stating the research topic.
+
+-   Interrogative - Asking a question.
+
+Each title type has its pros and cons, but it might be worth starting by knowing which title type you want to achieve.
 
@@
 
+If you are interested in reading more about titles and how people use them, there is some interesting analysis out there to get you started [e.g.
+@soler2007writing, @jamali2011article, @cook2016scholars, @krausman2020writing].
+Remember that title type and use does vary by discipline, so it would be worth being aware of what is generally used in the journals that you intend to target.
+
 ### Some title ideas to start you off
@@
     -   Allow yourself a longer subtitle if needed, but don't go over 20 words total (some journals may limit your total to less). 
+        -   Titles is a colon are downloaded and cited less [@jamali2011article].
+    -   Artiles with longer titles are downloaded less [@jamali2011article].
     -   Consider the (former) 120 character limit of a Tweet as an upper limit
@@
 -   Don't feel obliged to include [taxonomic terms](#taxonomy) unless it
-    is relevant or compulsory
+    is relevant or compulsory [see @krausman2020writing]
 
@@
 -   Avoid obscure or specialist words that won't be understood by your
-    readership
+    readership [@krausman2020writing]
 
@@
 
--   Don't have your title as a question. Rather provide the answer!
+-   Don't use an implicit words in the title: a study of..., the effect of..., factors affecting..., etc. [@krausman2020writing].
+
+-   Don't imply a series of works (e.g. Part I) [@krausman2020writing].
+
+-   Don't be too funny. Sagi and Yechiam [-@sagi2008amusing] found that highly amusing titles got less citations.
+
+-   Don't capitalise every word, the title is a sentence.
+
+    -   Some journals will have a house style that may change your title to capitals, all caps or some other format. However, the title is a sentence, and sentence case makes it easier to read.
+
+### Should your title have a question?
+
+[Betteridge's Law](https://web.archive.org/web/20090226202006/http://www.technovia.co.uk/2009/02/techcrunch-irresponsible-journalism.html) relates to newspaper headlines and the idea that any headline that is a question is most simply answered by a single word "No" even though the writer has written the question with the expectation that readers will be interested in their explanation of why the answer is "Yes".
+In academic circles, this same idea is known as Hinchcliffe’s Rule: “If the title of a scholarly article is a yes–no question, the answer is ‘no’” [@shieber2015this].
+The concern in both studies is that simple questions are being used as 'click-bait' to encourage potential readers to discover the simple answer.
+Indeed, there is some evidence that using a question in the title results in more downloads but less citations [@jamali2011article].
+
+Ball [-@ball2009scholarly] found that the use of question marks in titles has been steadily rising in the life-sciences over 40 years.
+His assumption is that, like a newspaper headline, using a question in the title is a way of the authors trying to market the contents of their article.
+Cook and Plourde [-@cook2016scholars] analysed nearly 8000 titles (although not in the Biological Sciences) for Betteridge’s Law and Hinchcliffe’s Rule, but found that academics in the disciplines they analysed did not conform to either.
+
+In summary, avoid having a simple question as your title.
 

### 2022-09-25 index.Rmd
Add files via upload
# updating chapters

--- 2021-09-05 index.Rmd
@@
 Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
-github-repo: yihui/bookdown-crc
+url: howtowriteaphd.org
+github-repo: How-to-write-a-PhD-in-Biological-Sciences/docs
 graphics: yes
+cover-image: figures/Cover-400wide.png
+favicon: figures/favicon.png
 ---
@@
 ```
-# In Press {-}
-This book is _in press_ with CRC Press.
+
+# Preface {-}
+
+*Published by CRC Press, Bota Raton, USA.*
 
@@
 
-You can pre-order your own copy on the [CRC Press website](https://www.routledge.com/How-to-Write-a-PhD-in-Biological-Sciences-A-Guide-for-the-Uninitiated/Measey/p/book/9781032080208), on [Amazon](https://www.amazon.com/How-Write-PhD-Biological-Sciences/dp/1032080205), at [Barns & Noble](https://www.barnesandnoble.com/w/how-to-write-a-phd-in-biological-sciences-john-measey/1139436164), and at other good book sellers.
+**Cover illustration**: Saint George and the Dragon was adapted by the author from an original work by Paolo Uccello (~1470).
+While the student (St George), supported by their laboratory (White Horse), is attempting to bring about
+the end of their PhD (Dragon) using this Fool’s Guide (lance), the advisor (Princess) is leading the PhD by a leash from the darkness into the community towards the Scientific Project (town).
+
+You can order your own copy on the [CRC Press website](https://www.routledge.com/How-to-Write-a-PhD-in-Biological-Sciences-A-Guide-for-the-Uninitiated/Measey/p/book/9781032080208), on [Amazon](https://www.amazon.com/How-Write-PhD-Biological-Sciences/dp/1032080205), at [Barns & Noble](https://www.barnesandnoble.com/w/how-to-write-a-phd-in-biological-sciences-john-measey/1139436164), and at other good book sellers.
 
@@
 Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning.
-James Baxter-Gilbert, Ellie Brown, Jack Dougherty, Anthony Herrel, Allan Ellis, Andrea Melotto, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, James Vonesh, Carla Wagener all read or commented on different aspects of the book.
+James Baxter-Gilbert, Ellie Brown, Jack Dougherty, Anthony Herrel, Andrea Hurst, Allan Ellis, Joe McArthur, Andrea Melotto, Max Mühlenhaupt, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, Ben Stevenson, James Vonesh, Carla Wagener all read or commented on different aspects of the book.
 Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and in the library at Stellenbosch University.

### 2023-03-30 1.03-email.Rmd
Add files via upload
# 2023 updates to chapters and book

# Communicating by email {#email}

Throughout your PhD, you will be expected to communicate with your advisor and other professionals inside and outside of academia using email. The chances are that you have already written a great deal of emails in the past. However, this does not mean that the style and content of those emails will be appropriate when communicating during your postgraduate studies.

In addition to meeting in person, email has become a standard way of communicating between students and their advisors. If your project has any collaborators, then you will likely also interact with them over email, perhaps without ever meeting them in person. Although other platforms for communication are available (e.g. Slack), email has the advantage that it provides a paper trail (time stamped, and institutionally traceable), is rapid, can include attachments and can include additional advisors or collaborators.

Email is a very low-cost method of communication, but it isn’t free. Every email sent takes energy to send it, and the bigger the size (especially sizes of attachments), the more electricity is used to send them through your institutional server, or beyond. This is actually a reason why you should not be sending superfluous emails as the bigger they are, the more energy they use.

An important point about emails is that they can be kept and/or forwarded to third parties. This means that you are best advised not to include any content that might be considered (by anyone) as risky or contentious. In particular, avoid anything written about any third party, unless you are happy for them to see what you have written. There is also a legal implication in that they can be used in court, for example, if you have written something libellous about a third party. This is true even if you didn't write the email but simply forwarded the content that someone else wrote. Remember that what is funny to you might be offensive to someone else.

## The subject
Do include a succinct subject for your email. Do not leave this blank. Remember that your advisor, and other professionals that you communicate with, likely receives many emails in a day. They may not be able to respond to you immediately, but may want to know what the email is about before they open it.

The subject is especially important if you need something quickly (e.g. something signed, or for permission). If this is urgent, you might consider saying when the deadline is in the subject line. Obviously, you should avoid making urgent demands from anyone in your professional world, but sometimes this is necessary and everyone will understand. However, if you fail to communicate this in the subject of your email, don’t be surprised if your request is sidelined until after it’s too late.

## Writing style
In general, it’s best to keep your style brief. If you feel that you need to write a long email, consider ways in which you could better communicate the same information in a meeting, or in responses to comments in a chapter draft. I would discourage you from writing long emails as reading them will tend to take up more of your advisor’s time than other forms of communication.

I used to get annoyed with people who insist on starting every email with “I hope this email finds you well” or some such banality. I now understand that this is culturally important for some people who are used to using longer greeting formats and who are genuinely sending a message of well wishing. Some time back, there was an administrator in our small unit who used this in every email that they sent, irrespective of whether we'd just been talking over a coffee or had spent the entire morning in a meeting together. Hence, this overuse of something that could otherwise have been nice to read became irritating. My advice would be that if you do have a personal connection with someone and want to make this clear, then by all means do so. Try to make it personal and not generic. Certainly, refrain from excessive use where any phrase could become meaningless or irritating

There are whole books on how to write emails [e.g. @taylor2009email], so I'm not going to go into too much detail here. Suffice it to say that just like any other form of communication:

 - try to make the reason why you are communicating clear
 - read through what you've written before you press send
 - let brevity be your watchword

### Don't use slang or SMS language
I remember an email that a colleague received (many years ago) asking for a bursary. I don't remember the exact wording, but it went something like this:

> "Hey Nanni [her family name] - U gonna give me 1 bursary? I is the best student."

Needless to say, this student didn't receive a bursary, although they could well have been the best student in their class and fully deserving. Their skills at communication were not sufficient to warrant another look. And while this example may seem ludicrous, it appears that many undergraduate students do correspond with their lecturers in this way [see @stephens2009r].

As with all communication, you need to be focussed more on the recipient and their ability to comprehend the message that you intend. Although I will reiterate that you must keep your emails short, this should not be at the expense of comprehension.

### Formalities {#formal}
The level of formality that you use will largely depend on your existing relationship with your correspondent. The formal tone is set immediately with the greeting that you use.

I am happy for people to start any email with “Hi”, but other advisors might prefer something else. If you are unsure about your advisor, then ask them. As a default, you should use the same style that they are using to address you.

When emailing someone that you don't know, use the most formal style: "Dear Dr. Blogs" or "Dear Prof. Bliggs". If you don't know what their rank is, use an appropriate level that you consider likely. Remember that there are conventions in different places (e.g. every academic in the US is a Prof.; in Europe Prof. is reserved for the highest rank only, and everyone else is Dr.). Similarly, when you [sign off](#signoff), use a formal [sign off](#signoff). As a general rule of thumb, you can use how they [sign off](#signoff) their email to address them. For example, if I [sign off](#signoff) an email "With best wishes, John", I expect that you will reply with Dear John or Hi John. However, if I [sign off](#signoff) "With best wishes, John Measey", then I am looking for a more formal response starting with "Dear Dr. Measey".

Remember that if you cc in others, you should move your formality accordingly. Bear in mind that emails can be kept or forwarded without your knowledge. You do not want to regret a particularly lax use of formality at a later date. I would suggest that the more people that you address your email to, the more formal you'll need to be.

In general, emails that you write as a postgraduate student are professional communications, so keep them as such. This does not mean that you need to start using citations in your emails, or that you can't use contractions (like can't, won't or didn't). You emails don't need to be formal as in the style of a published paper. As with the initial greeting, you should look at emails you receive to see how to set the formality of the tone that you will write.

## Why exactly are you communicating?
The first two or three sentences of your email should state why you are communicating and whether or not you are expecting a reply. Emails can be just for information, keeping someone up to date on your progress or letting them know that something they asked for has been done. If this is the case, then start your email with a statement letting the recipient know the reason why you are keeping them informed, and that you don’t expect a reply.

If you require a response from your recipient concerning a particular question, then you should flag this in the first line of your email. If their response is time sensitive (i.e. you need a reply before a certain date) then state exactly when this is. In the subsequent information that you write in your email, you need to make the question or questions clear.

If you need to write a long explanation in your email, then you might want to draw attention to the question that you want answering by making it __bold__. Another useful means of concentrating your recipients attention is to use structure.

### Using Structure
I like it when people use some structure in their email. If it’s going to be a long email (which I’d advise against) then do include a short summary explanation at the start so that the reader knows what the email is about (especially if the subject line doesn’t fully explain). Using bullet points or numbered points in the email is also a great way of using structure, especially when you expect a reply on each point.

#### Replying inline
This simply means that the previous email is reproduced when you hit reply, often with some marker for what was written originally. I like inline replies as it’s simple to see that each point has been addressed. It’s also a useful way to be reminded of what was originally written.

Some email chains start to become a bit tedious when many inline responses make them so cluttered that it’s hard to see what’s going on any more. Use your common sense to decide when it’s enough.

#### HTML emails or plain text
Most email software now defaults to emails written in HTML, which means that you can use **bold**, *italics*, bullet points and most other features that you would in a word processing package. Some people prefer plain text only, but HTML is becoming more standard.

#### Email chain
Most email software has an option to include the original message in the reply. This is quite useful in that it can remind everyone about what was asked. As these kinds of conversations get longer, they can become very large and unwieldy.

#### Bullet pointed or numbered lists

I think that both of these are appropriate when you have a number of points that you’d like addressed.

### Use of all Capital Letters - DON’T DO THIS!!!!

When reading any email when parts are written in All Caps, this comes across as someone who is shouting (see use in subheading). Not something that you should be doing in any professional communication. If you need to, then use bold or italics to emphasise points.

Similarly, I’d appeal for you not to use excessive exclamation points, slang (especially swearing) or other such frivolities in your professional email. It’s simply not a way of communication that you should be doing at work. Write emails to your friends like this if you want to, but I’d suggest that you do this from a private email account (and not your institutional address).

## Signing off {#signoff}

How is it best to sign off? Most professional emails sign off with one of the following:

 - With best wishes
 - Kind regards
 - All the best

Any of these is fine to use, as are the more formal classic letter writing “Yours faithfully” or “Yours sincerely”.

I quite like it when people sign off using their own language, as it (somehow) makes it more personal and sincere. I have worked in a number of countries around the world, and when there tried to use local language words to start and end emails.

## Who else should your email be sent to?

You may need to send emails to more than one recipient. If you are planning to send an email to a lot of recipients, then it might be worth checking this with your advisor first. You will also need to decide whether the recipients should be listed in the **To** box, CC or BCC.

If the email is sent to someone, then their name should be listed in the initial greeting: "Dear Drs. Bliggs and Blogs"

## When to CC someone

Only Carbon Copy (CC) someone if there is a real need for them to be included on the email. In reality, multiple addresses can be added in the To field, and anyone that the email is written to should be listed there.

In the business world, a lot of email correspondence includes the entire team in CC. This is usually not needed for academia. If you are in any doubt about whether or not to include someone in CC, then ask your advisor.

If you do put someone additional in CC, then it is polite to note this in the email. For example, in the initial salutation you could write “Dear Sue and Bob (and Warren in CC)...” I like to indicate the reason for including someone in CC at the appropriate point in the body of the email. This lets everyone know exactly why they have been CC’d.

> “Dear Sue and Bob, I am writing regarding the delivery of the 33 g steel ball that Warren (in CC) and I ordered on 12th November this year.”

Or

> "Dear Drs. Bliggs and Blogs (Prof. Blab in cc)"

Remember that someone in CC will be in CC again if the person you are writing to hits [Reply All](#replyall).

### Use of BCC

Blind Carbon Copy (BCC) is a very useful way of copying in another person without the original person or anyone in CC knowing. But you should use this feature of email with extreme care. Many people who receive an email in BCC are not immediately aware, and it has happened to me that people I have BCC’d have then hit “[Reply All](#replyall)” and immediately disclosed their presence on the email. This is very embarrassing. Thus I would urge you to only use this when the person you intend to BCC is aware of the fact.

A common use of BCC is when emails are sent to a great number of people and to avoid having anyone use [Reply All](#replyall). If you have a great number of collaborators (for example on a literature review or a large author list), then you can address the email to yourself (or your advisor) and BCC everyone else.

Note that if someone is on BCC, then they won’t get included when any of the rest of the group (addressee and CC) hit [Reply All](#replyall).

## Reply or Reply All? {#replyall}

Hitting Reply All instead of Reply is perhaps one of the most common mistakes that I see with email. Some email software has Reply All as the default option whenever you hit reply. If yours does this, I’d urge you to change it immediately. The use of emails to large numbers of recipients is common (especially in companies), but it gets extremely annoying when half of this list hits Reply All just to confirm receipt or even worse give some inane chatty response.

If you are going to use Reply All, before you send the email check:

 - that it is appropriate everyone on the list receives your reply
 - that there is not inappropriate information in the email chain
 - that you have added the correct attachment

If you have an email inbox that serves multiple email addresses, then it is worth noting that when you hit Reply or Reply All, the default is for it to send from the address that the email was sent to. You can change this default, or you may want to change from (for example) gmail to your institutional address.

### High Importance
Some email software has the ability to flag an email as important. Avoid using this unless you have something with real importance, like some decision or response needed that same day. Otherwise, try not to use this feature.

### Sending emails outside office hours
Personally, I have no problem with people sending emails whenever they want, as long as they don't expect replies to come out of office hours. I have often noted that chapter drafts appear in my inbox late on a Friday night. My suggestion to students is that this is not a good use of their time. Rather keep the draft and read it later on Sunday evening themselves after having left it alone for the weekend. You are more likely to spot errors and improve your own writing this way. It will certainly be better than it sitting untouched in my inbox all weekend.

If you don't know whether or not you will get back to your draft on Sunday evening, this would be a great time to use the [send later](#later) feature.

### Send Later {#later}

This is a new feature on many pieces of email software that I really like. Essentially, it allows you to compose an email and stipulate the time and date that you want it to be sent. I find this very handy if I know that someone is on leave. I'd prefer that my email not sit in their backlog, but arrive once they are back from being away and have already cleared their inbox.

Another scenario that I use for send later is when I have a deadline, but am awaiting feedback. I can make sure that I meet the deadline by sending the email with report or manuscript attachment or whatever it is that I have to send before the deadline. However, if at the 11th hour I get the feedback from the third party that I'm hoping to include, I can still find the email with report (it's usually in the Draft folder), make any amendments and then send it off.

### Forwarding emails

One of the most common ways of emails being inappropriately shared is when they are forwarded, or additional people are added to a reply much later in an email chain. I have received many emails where the sender was clearly unaware of what the chain contained.

In general, it’s not often appropriate to forward emails to others.

## Out of Office autoreply
I find the autoreply saying that someone is away very useful. It lets you know whether or not to expect a reply, and might mean that you then write to someone else instead. If setting your own Out of Office autoreply:
 - Do include the date when you will be back (and check that it’s correct when you send it)
 - Do state whether you are on holiday or in the field 
 - If there is another means of reaching you (when you aren’t on holiday) then give this

 Becase you don't know who will receive your autoreply, make it [formal](#formal) and polite.

## Using email signatures and banners

An email signature is a useful way of explaining who you are at the end of your email. I find them very useful sources of information like websites, telephone numbers, physical locations (i.e. address). They can also include more information. For example, my signature currently contains a list of journals that I edit for and links to this (and other) books. For this reason, signatures are useful, especially on first contact. Thereafter, signatures don’t really need to be sent with replies and in email chains. Most email software contains options for when signatures are sent.

Some institutions will automatically add a pictorial banner to the bottom of every email sent, advertising how great they are. Others will add long legal statements with disclaimers. Obviously, you will have little power over whether or not these are included by your institution. However, some students create and add their own banners to their emails. I would discourage this simply because it takes more energy to send, and uses up considerably more space than text only signatures.

## Using emails as a record of due diligence {#deliverables}

For students that are having problems with their advisor, especially with lack of communication, or non-communication, email provides a very important record to document your due diligence, as well as being able to document your advisor’s lack of timely response (hopefully you already have a student-advisor agreement to refer to). Make sure that you use your institutional email address and the institutional address of your advisor. If they prefer another address, use this in cc in addition to their institutional address. The reason for this is that if you are going to use this paper trail to document your communication, you will need to produce these emails for your institution and they may only consider emails that have been sent using their services.

If you suspect that you may need to use your emails in this way in the future, it is worth re-reading them, prior to sending them, and think about how they could be interpreted by a third party.

**If you are communicating with your advisor on a web based platform (like Slack), then do you need to send them emails as well?**

In most institutions, postgraduate students and advisors sign agreements before they start working together. These agreements are likely to come along with deliverables that may include progress reports and chapter drafts. It is well worth looking at this agreement and see what you've signed up to. I would maintain that at any point that your agreement states that you need to 'communicate' with your advisor, this should be done via email. Similarly, if the agreement talks about you needing to send thesis drafts, do these via email. As stated above, emails are useful because they can be traced, and institutional emails especially so. If either party claims that they 'did no receive' an email, this can be independently verified. If you sent something via Slack and at some point your advisor claimed that you didn't, how do you verify this later?

Of course, everyone hopes that there won't be a breakdown in communication during your studies. However, it's worth starting your professional career as you mean to go on.

## Keeping emails

There is a vast array of options when it comes to keeping emails. I know people who consistently keep an empty inbox and do not keep anything. I know other people who keep everything. Some people put every email they received (or send) into a folder. Others keep them all in the inbox (ok - I'm the only person I know who does this).

Many institutional email servers will have a limit of what you can keep. These days it can be quite large (e.g. 1 TB). So should you keep all correspondence? I would say that it is worth keeping anything that relates to [deliverables](#deliverables) that you have.

### Folders

I like to use tags to group particular subject areas together. I see no point in using folders for people as any search for a person can quickly recover all emails from someone.

I find that after 2 years, most correspondence is no longer useful. There is the odd occasion when I delve further back in time, but older institutional emails get archived after 2 years.

### 2023-03-30 2.05a-AI.Rmd
Add files via upload
# 2023 updates to chapters and book

# Can an AI write my PhD? {#AI}

The emergence of artificial intelligence (AI) platforms brings with it new issues around writing generally, including writing your PhD thesis. These bots are already widely available in early 2023, with both Microsoft (through their company OpenAI) and Google launching [ChatGPT](chat.openai.com) and [Bard](bard.google.com), respectively. There are already publications that list ChatGPT as an author [@oconnor2023open; @mijwil2023artificial] in acknowledgement that a large amount of the text therein had been generated by the bot. More AI writing tools are going to be coming online in the near future. Microsoft's plan to integrate the ChatGPT into Bing and the Bing platform directly into their office suite, means that as an author you may unwittingly use AI generated automated suggestions inside your word processor without realising it. Thus, you should be interested in knowing whether such tools can be used to help you in your own writing. If you haven't already tried [@owens2023how], once you do you will understand the tantalising possibility that they could be used to replace your needing to write.

In this chapter I set out some ideas about how you might use AI to help you write, as well as some things that you should consider if you do go that route.

## What is a Large Language Model (LLM)?
I decided to ask ChatGPT and Bard to contribute to this chapter by describing themselves. My input text: "I am writing a chapter for a book on academic writing and I want you to describe ChatGPT in your own words". Here's what they said:

**ChatGPT**

>As a language model based on the GPT-3.5 architecture, ChatGPT is a sophisticated artificial intelligence program that has been trained on a massive amount of text data from various sources. It is designed to understand natural language and generate responses that are contextually relevant and coherent.

>ChatGPT uses deep learning algorithms to analyze text inputs and predict the most probable next word or phrase, based on patterns and associations it has learned from its training data. This allows it to generate human-like responses to a wide range of questions and prompts, from simple queries to complex conversations.

>One of the unique features of ChatGPT is its ability to understand and generate text in multiple languages, making it a valuable tool for communication and collaboration across different cultures and regions. Additionally, it has been used for a wide variety of applications, including language translation, content creation, and customer service.

>Overall, ChatGPT represents a significant breakthrough in artificial intelligence research and has the potential to transform the way we communicate and interact with technology.

**Bard**

>Bard is a large language model, also known as a conversational AI or chatbot trained to be informative and comprehensive. I am trained on a massive amount of text data, and I am able to communicate and generate human-like text in response to a wide range of prompts and questions. For example, I can provide summaries of factual topics or create stories.

>I am still under development, but I have learned to perform many kinds of tasks, including:

>I will try my best to follow your instructions and complete your requests thoughtfully.

>I will use my knowledge to answer your questions in a comprehensive and informative way, even if they are open ended, challenging, or strange.

>I will generate different creative text formats of text content, like poems, code, scripts, musical pieces, email, letters, etc. I will try my best to fulfill all your requirements.

>I am excited to be able to help you with your academic writing, and I hope that I can be a valuable tool for you.

To summarise, LLMs are trained on an exceptionally large amount of data and use AI to first understand what you are asking it, and second pull out of their training set what they need to reply to your user input. The key then is that whereas previous chat bots had (relatively) small training sets and limited ability to understand user input, these new LLMs have vast amount of data.

## Large language models have been around for a while
Although at the beginning of 2023 we are seeing a large number of popular articles about ChatGPT, the use of language platforms in writing in general, including writing in the Biological Sciences, has been around for a while. If English is not your first language, then you already know how incredibly useful it is to enter text into a translation app in your own language and get English back.

Some years ago, these translation platforms were rather literal, simply translating each word and providing text back that was barely usable. Today, with the help of AI, the translation that returns is more often than not a reasonable sounding (near) grammatically correct translation of what was entered. Hence, if you have used such translations in your own writing, then you should have already considered the a lot of contents of this chapter, and if you haven't then you need to read on.

Interestingly, when ChatGPT was asked to provide text in French and Arabic, it failed to produce coherent text [@seghier2023chatgpt], although it is possible that this will change over time as it becomes more sophisticated. It is interesting to note the difference between the claim that ChatGPT made (above) and this assertion from Seghier [-@seghier2023chatgpt] a few months ago. Whether or not ChatGPT has improved has yet to be tested.

## Output from ChatGPT cannot be trusted
The first, and perhaps most important issue, about using ChatGPT outputs in your own writing is that you cannot trust it to generate text that is factually correct or unbiased. Indeed, OpenAI who own and administer ChatGPT readily admit that their bot can produce factually incorrect statements and is often biased in its output [@owens2023how]. Similarly, when you start to use Google's Bard, it forces you to click on an button to acknowledge that text generated may not be correct.

Because these bots use data on the internet to generate the basis of the text that they generates, they are prone to providing the internet consensus about popular topics, and studies have already shown that there is a bias in what is produced [@mcgee2023capitalism; @mcgee2023chat].

This means that you should be carefully checking any AI text that is generated before you add it to your chapter or manuscript. In particular, you should never trust a AI generated text with [citations](#citations) (see [below](#AIfail)).

### Plagiarism by AI
There is the chance that your AI tool is going to plagiarise text that it finds on the internet, and so you may unwittingly be pasting the words of another author into your own text (see more on plagiarism [here](#plagiarise)). Some authors consider that even if the AI is programmed to avoid plagiarism, using this text in any way should be considered as plagiarism [@salvagno2023can]. However, on balance, my own view is that any attempt to be overly restrictive on the use of AI generated text is bound to fail as AI itself becomes increasingly more sophisticated and widespread. Instead, I share the view [@lim2023generative] that it is better to embrace and learn how to use AI as a tool. Learn about what it can do, as well as its limitations.

Perhaps a more fundamental consideration is whether AI will require us to redefine what we mean by plagiarism. Certainly, this is a philosophical consideration that right now (at the start of 2023) you will need to decide for yourself, although as more AI tools come online and are more commonly available, I think that many journal guidelines will be prescriptive on their use (for example see [here](https://www.emeraldgrouppublishing.com/journal/sgpe)).

Plagiarism can be thought of as taking someone else's writing without attribution. Another definition of plagiarism is pretending that someone else's work is your own. While these two definitions may appear interchangeable at first sight, when considering use of text from an AI, you might only infringe the attribution rule while the AI is programmed to infringe the pretense. In other words, if the AI is not a person, then following the first definition of AI you have not taken someone else's writing. However, you might consider that as the AI itself was the product of someone else's work, then they are the person (or group of people) that generated that text, and so the work should be attributed to them. This then brings you to the second definition of plagiarism. But what if that group of people don't ask for any attribution or acknowledgement? Should you still give them as an author or put them into the acknowledgements? Moreover, could this team really be said to have generated the text when they have created software that searches text written by others on the internet and then collate and rewrite it? The huge number of content creators on the internet can never be acknowledged individually.

Another way of thinking about this might be to consider human interactions in writing text. If, for example, I help some colleagues with the English text of a manuscript and they offer to place my name in the acknowledgements, but I tell them that there is really no need. Should they still acknowledge me even though I have told them not to?

A further example might be the use of different layers in GIS that have been generated by different people. Many journals now insist that there is attribution to these layers in the legend of the figure or in the acknowledgements. We might legitimately ask whether this is necessary when those layers are freely available and have a non-attribution Creative Commons licence (see [here](https://howtopublishscience.org/whichjournal.html?q=creative%20commons#copyright))?

Clearly, the definition of plagiarism will need some work in the light of new abilities for AIs to write text.

To me, the possibility that by using AI generated text verbatim you could be using someone else's words means that you should avoid this. The same goes whether this is a translation from a large language platform or a new chat bot style AI. Using such platforms as tools is certainly acceptable, while using their text verbatim is probably an example of false attribution.

Plagiarism is the subject of another chapter, and you can read more about it [there](#plagiarise).

### Ethics
In addition to the attribution of authorship, and potential plagiarism issues, you should also be aware of the ethical component of using text from an AI tool.

The kind of ethics points that come up are similar to those that are raised by paper mills [@salvagno2023can]. This issue is covered comprehensively elsewhere: see [Chapter on when you should be an author](https://howtopublishscience.org/authors.html#papermills) in @measey2022how.

## Positive aspects of using AI
It will not come as any surprise that many authors often find it difficult to get started with their writing. The classic image is starting at a blank piece of paper, or these days a flashing cursor on an empty screen. In this book, I provide a lot of suggestions about how to get started with your writing task:

 - using a [phrasebank to generate text](#getting-started)
 - formula for [writing a logical argument](#argument)
 - [using a formula](#formula)
 - [making an outline](#start-with-an-outline)

but can ChatGPT provide another opportunity when you don't know how to start? I think that it can.

### Use AI as a tool 
Essentially, if you want to try using AI to help you write, use it as a tool. Don't try to use it to replace your own writing. That will likely end in a poor product, and could possibly end with factually incorrect or biased text that you won't want to be associated with your name - and clearly not something that you should want your advisor to read.

Getting started is often a struggle and I regularly suggest that students start by writing something/anything to fill out their outline, even if all of that text gets replaced at a later date. This is a way that you could use a tool like ChatGPT. Produce your own outline, because you do still need to plan your chapter, but then try using an AI like ChatGPT to flesh this out. Hopefully, this will end by helping you to write your own text, replacing everything that the AI has produced.

It is worth noting that the input for ChatGPT is dated to a portion of the internet from September 2021 [@azamfirei2023large]. It cannot access the internet and so you will not get anything more recent. Whereas Bard is trained from Google search results, Wikipedia, [The Common Crawl](commoncrawl.org) and other publicly available datasets.

#### Using ChatGPT to get started: an example

I have tried this approach below and you can judge the outcome for yourself. First, I write the text outline, and second I provide each point to ChatGPT. For the following outline, I provided the outline statement (each bullet point) in quotes and prefixed it with "Write a sentence with citations on: "..."

In this example, the outline below is similar to what I would produce when writing [an outline](#start-with-an-outline) for the first two paragraphs of an introduction to a manuscript on invasive fish:

 - Invasive species are a growing global problem
    + invasive species cost a huge amount of money to governments
    + invasive species are a threat to biodiversity
    + invasive species impact people's quality of life

 - Freshwater habitats have been particularly impacted by global change
    + important changes have occurred to freshwater systems after the introduction of invasive fish
    + invasive fish have been added to both rivers and lakes
    + large numbers of introductions of invasive fish have already occurred globally, and are ongoing

The following text was **generated by ChatGPT** (on 19 March 2023) given the primer text above, and the stipulation that it should produce a sentence with citations. I have simply combined the output text into the paragraphs as might be expected in the introduction to a manuscript:

> According to the International Union for Conservation of Nature (IUCN), invasive species are a growing global problem that pose a significant threat to biodiversity and ecosystem health. The National Wildlife Federation (NWF) reports that invasive species cost the United States government an estimated $120 billion annually in damages and control measures (Masterson, 2018). The Convention on Biological Diversity (CBD) states that invasive species are one of the main threats to global biodiversity, causing species extinctions, altering ecosystems, and reducing ecological resilience (CBD, 2021). A study conducted by the University of California, Davis, found that invasive species can have negative impacts on human health, recreation, and property values, affecting people's quality of life in areas where they are present (Koch et al., 2011).

> A global analysis of freshwater ecosystems found that they are among the most vulnerable to the impacts of climate change, pollution, and invasive species, with declines in biodiversity and ecosystem services being reported worldwide (Vörösmarty et al., 2010). Studies have shown that the introduction of invasive fish species can cause significant ecological and economic changes in freshwater systems, including alterations in food webs, declines in native fish populations, and reduced recreational and commercial fishing opportunities (Leprieur et al., 2008; Ricciardi & MacIsaac, 2011). Invasive fish species have been intentionally or unintentionally introduced into both rivers and lakes, causing significant ecological impacts and economic costs globally (Savini et al., 2010; Phelps et al., 2016). A study by Jeschke et al. (2014) found that over 12,000 introductions of invasive fish species have already occurred globally, with new introductions continuing to happen at an increasing rate, highlighting the urgent need for effective prevention and management strategies.

The following text was **generated by Bard** (on 27 March 2023) using identical input format:

>Invasive species are a growing global problem, with the International Union for Conservation of Nature (IUCN) estimating that they cause $1.4 trillion in economic damages annually (IUCN, 2016; WWF, 2017). Invasive species cost governments a huge amount of money to control and eradicate, with the US National Invasive Species Council estimating that the cost in the US is $120 billion annually. (NISC, 2017). Invasive species are a major threat to biodiversity, as they can outcompete native species for resources, introduce new diseases, and alter habitats (IUCN, 2016). Invasive species impact people’s quality of life by causing economic losses, disrupting ecosystems, and harming human health. Economic losses can occur when invasive species damage crops, forests, and other natural resources. Disrupted ecosystems can lead to the decline of native species and the loss of ecosystem services, such as water purification and flood control. Harm to human health can occur when invasive species spread diseases or cause allergic reactions.

>Freshwater habitats have been particularly impacted by global change, with climate change, invasive species, and habitat loss all playing a role. Climate change is causing changes in temperature, precipitation, and water flow that are altering freshwater habitats. Invasive species are non-native species that can outcompete native species for resources, introduce new diseases, and alter habitats. Habitat loss is occurring due to human activities such as development, agriculture, and mining. The impacts of these changes are already being felt, with many freshwater species in decline. If we do not take action to address these challenges, the consequences for freshwater ecosystems and the people who rely on them will be severe. "Invasive species can cause major changes to freshwater ecosystems, including the decline of native species, the alteration of food webs, and the spread of diseases. For example, the introduction of the invasive Asian carp to the Great Lakes has led to the decline of native fish populations and the alteration of the food web" (IUCN, 2016). "Invasive fish, such as Asian carp and largemouth bass, have been added to both rivers and lakes, often through illegal releases or escapes from aquaculture facilities." (IUCN, 2016). "Large numbers of introductions of invasive fish have already occurred globally, and are ongoing. Between 1970 and 2000, an estimated 7,000 species of fish were introduced to new areas (IUCN, 2016)."

For full details of the input and output of each sentence, please see my blog [here for ChatGPT](http://john.measey.com/Blog/2023/03/19/Will-Chat-GPT-write-my-PhD) and [here for Bard](http://john.measey.com/Blog/2023/03/27/Can-Googles-Bard-do-any-better).

#### The results of my experiment are in - both bots failed {#AIfail}

As an exercise, I have tried to find each of these citations in an attempt to determine whether ChatGPT was accurate. You can find the results in full [here](http://john.measey.com/Blog/2023/03/24/The-results-are-in---ChatGPT-is-a-fraud). So, it’s bad news for anyone who might want to try to rely on it for writing referenced statements for their PhD. Although my initial assessment was that the text looked good most of what was attributed to the citations given was fraudulent, when the paper could be found. The text sounds good, but the [citations](#citations) are bogus, and that's no use for you, or anybody else. Potentially it's even worse than bad sounding text as it sounds credible. Similar attempts to fact check ChatGPT generated text have also found it to be completely fabricated [@azamfirei2023large].

I have not attempted to verify the output for the Bard output as there were very few citations given and all are reports, not original works, and so would not be readily accepted by anyone. Interestingly, Bard repeated my input text as part of its output, even encasing it in quotes. Bard's excessive use of quotes when I asked it to write a sentence is also noteworthy. It is unclear whether it had problems generating text, or whether using quotes was considered a viable alternative to generating text. Bard also had problems restricting itself to a single sentence output.

Getting the [citations](#citations) right, and making your text fit what they find is where the skill comes in. Most statements sounded good, and were certainly on message, but the citations were sometimes vague and when they could be pinned down turned out to be about other subjects or did not contain the information claimed.

For information on how to cite, see the chapter in [Part II](#citations).

### Were the bots a complete fail? 
Yes, I would say that having checked the references for ChatGPT, most of what was said sounded good, but could not be backed up by the [citations](#citations) given, in the cases where they could be found. For this reason, I am confident in calling this experiment a complete fail, not because there was no truth, but because it was mostly misleading (incorrect [citations](#citations) and unhelpful misdirection to correct [citations](#citations)).

For most of you who are trying to write their PhD thesis, I would suggest that you are completely wasting your time to try this approach with ChatGPT. This doesn’t mean that you can’t use it as an to writing, but that you can’t rely on it to produce anything that is factually correct. Perhaps ChatGPT is getting all these citation uses wrong because it is looking to see how other people use citations? This is not unreasonable, as mis-citing is a real problem. But having this platform reinforcing false citation habits is the last thing you need.

Given that it’s taken me several hours to check two paragraphs, I feel no hesitation in suggesting that you avoid using ChatGPT in this way – and if you do, beware – it probably got it all wrong.

### Can I use AI to create a first draft?
If you are all out of ideas and staring at a blank page, then you could use the AI bot to help you generate some first ideas for text. However, I would add a heavy caution not to rely on anything it generates, and to be very cautious with respect to plagiarism. Overall, I'd suggest that you only use it as a last resort, and then only do so with a plan to substantially change and rewrite the text afterwards.

### Did the AI pass the plagiarism checker
Using a plagiarism checker (TurnItIn), the output from ChatGPT suggested that: **No plagiarism was found.** However, text generated by Bard was found to have **plagiarism at 21%**. Some of this included text that was already in quotes, but there was additional text that had been lifted from non-cited sources on the internet.

**NB** This was done before TurnItIn launched the trial of their "AI writing detection capability". It will be interesting to repeat this exercise and test whether or not plagiarism detection is possible.

I suspect that if plagiarism checkers, already in use in many institutions and publishing platforms, release additional tools to detect AI writing generation, then this may set the policy at these same institutions. However, this also raises some concerns.

 - What if I write my own original text and it is flagged as AI generated?  
 - Will plagiarism tools keep up worth AI improvement?
 - Can I still use AI as a tool to get started?
 - What is the aim of policing AI text generation? 
    + are we trying to improve science or improve an ability to write English?

## Will others detect ChatGPT if I use it?
Some people claim that they can detect the output from ChatGPT as it lacks the depth and insight that original authors usually have. In other words, the aim of ChatGPT is to produce the words (in a grammatically correct manor), while the aim of an author is to transmit an idea to the reader. As you might expect and hope, AI is not at the point where it can generate the same intent to communicate.

There will likely be better AI chat bots in the future, and ChatGPT itself is constantly getting feedback from users that should improve its own output.

### ChatGPT, or other AI platforms, cannot be an author
You may be aware of several articles that are already published with ChatGPT listed in the author line [@oconnor2023open; @mijwil2023artificial]. Decisions on whether or not this is allowed are likely to be journal specific. However, some publishing houses are already making journal wide decisions excluding the use of AIs as authors (for example, see [here](https://www.emeraldgrouppublishing.com/journal/sgpe)).

This is clearly a dynamic process and you should be aware of what is happening in your own field.

## Transparency {#AItrans}
However you use an AI platform, you should be transparent to your advisor and any colleagues that you publish with about the exact level of use that was involved in generating your text. As a general rule with writing and academia, [transparency](https://howtopublishscience.org/transparency2.html) is the best policy. Some authors have already called for regulations [@salvagno2023can]. Be aware that journals do update their instructions to authors, so you may need to look for statements on AI text generation. Similarly, familiarise yourself with your institutions requirements for thesis submission, and be careful you do not transgress any recently added rules.

### COPE have already provided guidelines
The Committee on Publication Ethics (COPE) have already provided some guidelines to their members (mostly publishers) on the use of AI [@cope2023artificial] which you can find [here](https://publicationethics.org/node/50766). In general, these rules emphasise the importance of oversight and transparency in the use of AI tools in decision making.

## Publishers using AI bots
Publishers are very interested in using the text from AI bots to generate lay summaries of articles that are published on their platforms. Certainly, there is evidence that scientists are already using the bots to create their own summaries [@owens2023how]. This is certainly a possible creative use for AIs, but I would be concerned that without careful curation they may be prone to producing factually incorrect or misleading content. If you plan to use AI software to popularise your own science, then I suggest that you carefully read anything that is created and ensure that the text is correctly attributed when you use it.

## Last note

This text was written in RMarkdown, without the use of suggestive prompts. AI generated text is written as quotes in the above text. All other text is my own.

It will be interesting to look back on this chapter in a decade and see the changes that have emerged in that time.

### 2023-03-30 2.17-plagiarise.Rmd
Add files via upload
# 2023 updates to chapters and book

@@
 
+## Is it plagiarism if an AI bot writes it?
+Currently, we are moving into the new territory of AI generated text.
+Whether or not you can use AI generated text (e.g.
+ChatGPT or Bard) in your PhD or in manuscripts submitted to journals is an open question.
+You can read more on this topic in another chapter [here](#AI).
+
 ## How can you make sure that you never plagiarise?

### 2023-03-30 4.03-publish.Rmd
Add files via upload
# 2023 updates to chapters and book

@@
 - It is hoped that by publishing your work it will become more accessible to the scientific community.
-- By publishing your work you will find that both you, and your work, become known by an international community that may well invite you to participate in the academic process further [@marks2013misuse].
+- By publishing your work you will find that both you, and your work, become known by an international community that may well invite you to participate in the academic process further [@marks2013misusea].
 - Both inside and outside of science as a career, peer-reviewed published papers are seen as an important accomplishment in addition to the production of your thesis.

### 2023-03-30 index.Rmd
Add files via upload
# 2023 updates to chapters and book

@@
 Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
-url: howtowriteaphd.org
-github-repo: How-to-write-a-PhD-in-Biological-Sciences/docs
+github-repo: yihui/bookdown-crc
 graphics: yes
-cover-image: figures/Cover-400wide.png
-favicon: figures/favicon.png
 ---

### 2023-04-29 1.01-beforeyoustart.Rmd
Add files via upload
# with addition of conceptual framework chapter

@@
 
-I prefer a thesis that fits together within an explicit [conceptual framework](#PhDintro).
-I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the introduction of their [proposal](#proposal).
-A useful concept to be aware of is the 'Hierarchy of Hypotheses' [@heger2018hierarchy].
+I prefer a thesis that fits together within an explicit [conceptual framework](#concept).
+I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the [introduction](#PhDintro) of their [proposal](#proposal).
+A useful idea to be aware of is the 'Hierarchy of Hypotheses' [@heger2018hierarchy].
 In this approach, you are encouraged to consider the '[bigger question](#bigidea)' in your research area, identifying both what studies have produced compelling evidence, and identifying knowledge gaps from a theory driven approach.

### 2023-04-29 2.19a-concept.Rmd
Add files via upload
# with addition of conceptual framework chapter

# Creating a conceptual framework of your PhD thesis {#concept}

Explaining your PhD is going to be a constant throughout the duration, and beyond. Getting that explanation correct, and being able to do it for audiences at different levels will be really critical from the outset. The best way that you can do this is to get the conceptual framework for your thesis straight in your own mind as soon as you can.

Once completed, your PhD will have an audience of [examiners](#who-are-the-examiners), but eventually you'll get to [publish](https://howtopublishscience.org/) all the chapters of your PhD to an audience of scientists and potentially a much greater audience of people around the world. In each case, if you can contextualise your work and how it fits into your thesis, it will help increase your confidence.

## Planning your thesis as a concept and a practical timeline

An important phase at the outset of your PhD project is to plan the chapters that are going to be in it. Think about the contents of your PhD and commit as many thoughts and ideas as you can to paper. If you aren't sure how to get started, then think about how to answer the following questions:

- What questions are you going to ask?

- What is your study system?

- How will you collect your data?

- What are the most important variables that you will measure?

- What techniques will you use?

- Do you have hypotheses?

All these ideas (and more) are going to be swirling around your mind at the beginning of your PhD and you're going to need to commit them to paper, and doing this right now at the start is the best time. If you haven’t done so already, write them all down. If you can do a [mind map](https://www.mindmaps.com/how-to-make-a-mind-map/) or some kind of graphical representation, this will be good for you if it suits the way that your mind works. Otherwise, you can use a series of lists and bullet points, if that is more your style. The important point here is be able to move from a jumble of ideas and thoughts into a formal plan for your thesis.

A thesis typically has five data chapters that are presented in a linear fashion (book format), bookended by an [introduction](#PhDintro) and [conclusion](#PhDconc). At the heart of each chapter is going to be a [hypothesis](#hypothesis), a [question or a technique](#nohypothesis) that the following chapters implement in order to get their results.

It may help you to use distinct colours and short titles (just a couple of words) for each chapter so that you can efficiently communicate them to your advisor, and use them as file names for the sections (don’t use “Chapter 1” as your chapter name!). You should then use these same colours and short titles in your thesis timeline (Gandt diagram) used in [your proposal](#proposal).

Once you have some rough ideas for your thesis plan, discuss it with your advisor and get their opinion. Then map it out on a piece of paper. But, be warned, it probably won't be anything simple or linear - it’s likely to be more complex that the one illustrated in Figure \@ref(fig:concept). It should have a lot of links and arrows that join all of the chapters to each other in different ways. It may help to code these links so that it’s clear what they represent. For example, you may want to use one colour for data and another for techniques and results. Although the detail is important, remember that the conceptual thesis plan is supposed to give an overview of the way in which the thesis works. This means that you might need to remove some of the minutiae in order to provide a clearer overall picture that others can easily follow.

(ref:concept) **A conceptual thesis plan will be placed in the introduction to your thesis**. In this (fictional) example, I show how how the different chapters are linked by a single technique (developed as a chapter of the thesis: grey arrows), and data that results from each of the chapters (blue arrows).

```{r concept, echo=FALSE, out.width = '95%',  fig.cap="(ref:concept)"}
knitr::include_graphics('figures/concept.png')
```

### Dependencies

Make sure that your conceptual thesis diagram makes clear exactly where and how chapters depend on each other to provide data, the outcome of results and/or the successful development and implementation of techniques. These dependencies must be very clear to you and your advisor at the outset so that you are able to make contingency plans in case any of these don’t work out in the way that you think that they will.

### Your thesis must add up to more than the sum of its parts

You can't just have a thesis that consists of five manuscripts bound together. Your PhD must be something greater, and you will need to be very clear about how it does this from the start. Indeed, this is a key requirement in the [rubric of most universities](#criteria) for you to pass your PhD examination. To do this, the thesis is definitely going to have the kind of interconnectedness shown in Figure \@ref(fig:concept) from the outset.

### Using your conceptual thesis plan

Once you've got your conceptual thesis plan, and both you and your adviser agree on it, keep it. It's going to be a figure in the [introduction](#PhDintro) of your thesis (like Figure \@ref(fig:concept)). You are also going to be able to use this conceptual plan as a reference guide for your own explanation to yourself, and anyone else that you propose your thesis to. If you have a thesis committee, this may well be a diagram that you use in every meeting to remind them of what your thesis is about and how it fits together. As you progress along the course of your thesis timeline, you will be able to demonstrate what has been done. It may also be useful if you need to [raise funds](#funding) for your work, and to anyone who needs to understand your entire thesis such as an ethics committee.

Perhaps the most important need for your conceptual thesis plan will be to explain to you PhD [examiners](#who-are-the-examiners), the layout of your PhD and how the chapters relate to each other.

### Evolving your thesis concept plan

Don’t be afraid of making changes to your plan during the course of your thesis. If you need to rely on one of your contingency plans, it is likely that you will have to change the layout of your timeline as well as your conceptual plan. This is normal in the course of any PhD thesis.

## Your thesis timeline {#Gendt}

Whether or not you are used to managing your own time, you are going to need to learn how to plan your time during your PhD. This is not only because you will need to know what you must do on a month to month basis, but more that you will need to plan what you need (materials, chemicals, equipment, etc.) and when you will need it. If you plan to do a lot of fieldwork, then you may also need other people to help you with this. Thus you will need to know when to train and organise this group.

### Producing a Gendt diagram
The timeline of your thesis will be an important guide throughout the 3 years that you are conducting your PhD. This means that it is worth getting this onto the record as soon as you are able. One of the standard ways of doing this is to produce a Gendt diagram. This takes each of the chapters along the entire timeline of your PhD and shows when you will work on each one. It may be that you want to break down each of these chapters into sub areas such as literature review, data collection, data analysis, write-up.

Below you will find a very simple version of a Gendt diagram (Figure \@ref(fig:Gendt)) that I put together for a fictional PhD. Note how the colours used for each chapter are the same as the thesis concept diagram (Figure \@ref(fig:concept)). It's always useful to follow such colour conventions once you've started them. Maintaining consistency throughout your thesis will help you when you explain it to others, such as during your proposal or whenever you present to your thesis committee.

(ref:Gendt) **A Gendt diagram used to help thesis plan how you will use your time during your PhD**. In this (fictional) example, I show how how the different chapters (Q1, T1, Q3, Q4, Q5) will take up different time during the three years of a PhD. Each year (1 to 3) is divided into four quarters (Q1 to Q4). This example is set out to show the need for seasonal sampling (in the second quarter of each year). Stars represent goals that you have set.

```{r Gendt, echo=FALSE, out.width = '95%',  fig.cap="(ref:Gendt)"}
knitr::include_graphics('figures/Gendt.png')
```

As you're doing your thesis in the biological sciences, you're much more likely to have some kind of seasonal quotient that constrains the work that you're doing. So in Figure \@ref(fig:Gendt) I've used the second quarter of each year as the key field period that is needed for all of the chapters.

You will note that as soon as you build in a seasonal aspect to your timeline, your life will get much more complicated, especially if you have more than one chapter that depends on the results from a field season. This should help you know when you need to plan each field campaign well in advance of actually going to do it.

### Set yourself goals

It's really useful to set down goals that you want to achieve on the Gendt plan, and to share these with your advisor. Milestones in your PhD are really useful when planning, and psychologically enhance your ability to finish different components.

Note that none of the goals occur in the same quarter as any other goal. This is important as it enables you to know where to concentrate your time and when. Note also that the goals don't occur in the same order as the chapters. Part of this is the need to be realistic about what you need to achieve and the time that it will take to do this.

### Map on significant events

If you plan to attend workshops or conferences during your PhD (or even plan to organise these yourself), you should plan such significant events onto your timeline. For example, If you are presenting the results of your first chapter at a conference, you need to make sure that you have the necessary results well in advance of needing to write the abstract. In Figure \@ref(fig:Gendt), the first conference is in the third quarter of the second year and you should have already achieved your main goal for this chapter well in advance to write the abstract. You might also be able to add in the results of the second year's seasonal field work to present at the conference. Similarly, for the second conference, you should have all the results from the third chapter well in advance to write the abstract.

### Contingency planning

If you can, it's a good idea to build in some contingency planning during this planning stage so that you know where you have options, and especially if your thesis relies on weather or having a particular set of results to continue.

In Figure \@ref(fig:Gendt) I have shown how a contingency might be mapped onto Chapters Q1, Q3 and Q4 with the potential addition of another field season in the second quarter of the third year in case one of the previous field seasons fails.

### Dynamic timeline

As your thesis progresses, so too might you need to change your timeline to reflect changes in its implementation. Make sure that your advisor (and possibly your thesis committee) is aware of any significant changes to your timeline.

### 2023-04-29 2.20-proposal.Rmd
Add files via upload
# with addition of conceptual framework chapter

@@
 
+A key part of the introduction will be your [conceptual framework](#concept) that will explain how your thesis chapters fit together, and crucially, add up to more than a sum of their parts.
+
 ## Methods and Materials
@@
 This could be a bar chart, a map, a scatter-graph, etc.
+
+## Thesis timeline
+
+You should be able to state from the outset when you are going to find the time to fit in all of the components of your thesis.
+Being able to plan when and how you will do the work is a key aspect of your proposal.
+I discuss how to construct a [timeline](#Gendt) in the form of a Gendt diagram in another chapter (see [Part II](#Gendt)).
 

### 2023-04-29 2.21-proposal-presentation.Rmd
Add files via upload
# with addition of conceptual framework chapter

@@
   + give the context of your own study
+  + provide your [conceptual framework](#concept) for your thesis
 3. Your aims and hypotheses
@@
 7. Budget
-8. Timeline
+8. Show your [timeline](#Gendt)
 

### 2023-12-28 1.02-useR.Rmd
Add files via upload
# updates to several chapters

@@
 All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC.
-This is practically where we are now with R (@rcoreteam2021r; Figure \@ref(fig:Rlogo)), and a wealth of other emerging and alternative programming platforms.
+This is practically where we are now with R (@rcoreteam2021language; Figure \@ref(fig:Rlogo)), and a wealth of other emerging and alternative programming platforms.
 

### 2023-12-28 2.09-databases.Rmd
Add files via upload
# updates to several chapters

@@
 
-Databases of scientific literature have been around since the 1960s, when the number of journals and papers started increasing so much that scholars simply couldn’t keep up.
-Quite simply, you should use one or more of these databases to look for relevant literature to read.
-The first database, and one that continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of selected journals (titles, abstracts, authors and other metadata), but also the citations that each paper used.
+Databases of scientific literature have been around since the 1950s, when the number of journals and papers started increasing so much that scholars simply couldn’t keep up [@garfield1955citation].
+The first database (from 1962), and one that continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of selected journals (titles, abstracts, authors and other metadata), but also the citations that each paper used.
+
+Many more literature databases have emerged over the years.
+Perhaps most notably [PubMed](https://pubmed.ncbi.nlm.nih.gov/) contains a database of biomedical literature and has been free for public use since 1996 and is funded by the US National Institutes of Health ([NIH](https://www.nih.gov/)).
+The Science Citation Index was acquired by the Thompson Corporation in 1992, and this scholarly information part of the business was sold to private for-profit owners, Clarivate Analytics, in 2016.
+A rival literature database, [Scopus](https://www.scopus.com/) was founded by publishing company Elsevier in 2004, claiming to be the largest citation database [@schotten2017brief].
+Similarly, [Dimensions](https://www.dimensions.ai/), was launched in 2018 as a project owned by Holtzbrinck Publishing Group, owners of Springer-Nature [@vannoorden2018science; @hook2018dimensions].
+These initiatives of major science publishing houses are thought to be linked to their attempts to capture the scientific work-flow [@brembs2021replacing]: and see [Part I](#capture).Google launched [Google Scholar](http://scholar.google.co.za/) in 2004, and while this database is free to use, the owners are a for-profit company and the data behind the database remains frustratingly closed.
+
+In 2022, a new Open and free database, [OpenAlex](https://openalex.org/about), successfully aggregated data from [ORCID](https://orcid.org/), [ROR](https://ror.org/), [DOAJ](https://doaj.org/), [Unpaywall](https://unpaywall.org/), [Pubmed](https://pubmed.ncbi.nlm.nih.gov/), [Pubmed Central](https://www.ncbi.nlm.nih.gov/pmc/) and [the ISSN International Centre](https://www.issn.org/) among [others](https://docs.openalex.org/api-entities/sources/get-lists-of-sources) [@singhchawla2022massive].
+[OpenAlex](https://openalex.org/) is remarkable for two reasons: (1) by aggregating works from so many different Open Source databases, it has amassed [243 million works](https://openalex.org/about) - nearly 3 times that of Scopus or Web of Science collections.
+(2) OpenAlex is a free index (specifically it is [CC0](https://creativecommons.org/public-domain/cc0)), it is [Open Source](https://github.com/orgs/ourresearch/repositories?language=&q=openalex&sort=&type=public) and the data is [free to use](https://docs.openalex.org/additional-help/faq#how-is-openalex-licensed).
+Contrast this with the other for-profit databases, and you will understand that we should support this new initiative.
+
+As a postgraduate student, you should use one or more of these databases to look for relevant literature to read.
 As we have seen in the last chapter, citations record the way in which people find the literature useful, and are therefore, a highly effective way of finding related and pertinent literature in your field.
@@
 Luckily, literature databases have some tricks for you to add to combine your keywords in ways that you can include or exclude certain topics.
+
+**Struggling to define your own keywords?** Help is at hand with a new workshop designed specifically to enable you to determine your own keywords specific to your study, and set up a citation alert from your favourite database (click [here](https://johnmeasey.github.io/Thesis-Writing-Guide/keywords.html)).
 
@@
 
+## For-profit databases vs Open Source
+
+After the purchase of the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) in
+
 ## Administrators want metrics
@@
 
-## Google Scholar, Web of Science or Scopus?
+## Google Scholar, Web of Science, Dimensions, Scopus or OpenAlex?
 
@@
 
-Well, it seems that there are some people who have already thought about this, and come up with a good idea of exactly what's different.
-Martín-Martín et al.
-[-@martin-martin2018google] have done a great job of analysing some 2.5 million citations.
-What they found inspired me to write this section, in which I've focussed on the results from the Life-Sciences.
-But I encourage you to go read the article for yourself.
-
-I have been known to take the odd peek at my Google Scholar profile over the year, and see how the citations are accumulating.
-I rarely check on WoS or Scopus, because it's a bit of a faff getting signed in and doing the search.
-Plus it looks so much smaller when one is habituated to seeing those double digits in GS!
-However, I've always been a bit uneasy about citing my GS citation rate, H-index or i10 (among other mertics that they give) as I've never really known what all that extra represents.
-Something grey and unseemly?
-Well, it turns out that it's all good stuff, and perhaps GS is the better one to cite as it's a more inclusive index: more inclusive of different document types and different languages.
+According to Martín-Martín et al.
+[-@martin-martin2018google], who analysed some 2.5 million citations, the GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus.
+If you deviate from this with a higher score, you can give yourself a pat on the back for having work that's reaching more people in more parts of the world.
+Google Scholar results reflect a more inclusive citation index [@martin-martin2018google].
+While WoS and Scopus aren't exclusively English or journal publications, they are mostly.
+But that extra third that GS gives you allows you to show the extra scope that your work is getting outside that English journal mainstream.
+Is your GS score more than a third higher than your WoS or Scopus score?
+If yes, then your work is having a greater impact elsewhere in the world, and there's nothing wrong with that.
+Perhaps GS is the better one to cite as it's a more inclusive index: more inclusive of different document types and different languages.
 
@@
 To directly compare your citation metrics from different platforms, and also to get more unbiased metrics than these platforms provide, consider using the software Publish or Perish [@harzing2007publish].
-This also provides a great platform on which to harvest data from Google Scholar into a database, or put it directly into another database, like EndNote, Zotero or Mendeley.
+This also provides a great platform on which to harvest data from Google Scholar into a database, or put it directly into [your own database](#referencesoftware), like EndNote, Zotero or Mendeley.
 
-## What’s the difference between indices?
-
-According to Martín-Martín et al.
-[-@martin-martin2018google], the GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus.
-If you deviate from this with a higher score, you can give yourself a pat on the back for having work that's reaching more people in more parts of the world.
-Google Scholar results reflect a more inclusive citation index [@martin-martin2018google].
-While WoS and Scopus aren't exclusively English or journal publications, they are mostly.
-But that extra third that GS gives you allows you to show the extra scope that your work is getting outside that English journal mainstream.
-Is your GS score more than a third higher than your WoS or Scopus score?
-If yes, then your work is having a greater impact elsewhere in the world, and there's nothing wrong with that.
+We anticipate future studies that will elucidate how Google Scholar, Web of Science, and Scopus differ from Dimensions and OpenAlex.
+We already know that larger databases should have potential for larger citation numbers.
+Because OpenAlex is Open Source, this means that you have the potential to build your own API to generate any metric that you are interested in completely free of charge.
+The prestigious French Sorbonne has been the first university to abandon their subscription to for-profit literature databases [@jack2023sorbonne], and it will be interesting to see how OpenAlex will transform the scene of literature databases in the future.

### 2023-12-28 3.02-Abstract.Rmd
Add files via upload
# updates to several chapters

@@
 It is an abstract of the entire document.
+The abstract is a standalone text.
+The reader should not have to consult more widely in order to make sense of what they are reading  - hence there are no citations in an abstract.
 
@@
 None of this is easy, and you should not expect to write your abstract in a single sitting.
-It will likely require multiple iterations, and some intense word-smithing to make it as good as it can be.
+It will likely require multiple iterations, and some intense wordsmithing to make it as good as it can be.
 Abstracts almost always have a word limit, and that makes it challenging and means that you have to be concise [see part 2](#concise).
@@
 
+## Can I use citations in the abstract?
+No.
+This is not a good idea as it defeats the object of the abstract as a standalone text.
+There are some legitimate times when you might need to use a citation in an abstract (although they are very rare).
+For example, if all of your manuscript refers to a single previously published paper, and the abstract does not make sense without knowledge of this publication.
+If you really need to, you can read more about how to cite in the abstract in [Part 1](#citations-in-the-abstract).
+
 ## Where do people go wrong in writing the abstract?
@@
   + If you have to have a citation in your abstract, then it must include sufficient information that it can be found without opening up the paper: _first author, abbreviated journal name, volume, first page_.
+
+**Still struggling with your abstract?** Take advantage of a new workshop that specifically aims to help you write the abstract of your own thesis chapter (click [here](https://johnmeasey.github.io/Thesis-Writing-Guide/abstractwriting.html)).

### 2023-12-28 3.03-Introduction.Rmd
Add files via upload
# updates to several chapters

@@
 |--------|---------|
-| Paragraph 1 | Overview of the major theme. |
+| Paragraph 1 | Overview of the relevance of the major theme  |
 | Paragraph 2 | Identify knowledge gaps in the field. |
@@
 In particular, sections on composing your [hypothesis](#hypothesis), writing a [paragraph](#paragraph), and an [argument](#argument) should be referred to when you are writing your introduction.
+
+**Still struggling with your introduction?** Take advantage of a new workshop that specifically aims to help you write the introduction to your own thesis chapter (click [here](https://johnmeasey.github.io/Thesis-Writing-Guide/formula1.html)).

### 2023-12-28 3.10-Figures.Rmd
Add files via upload
# updates to several chapters

@@
-# Figures
+# Figures {#figures}
 
@@
 
-Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’s books is R ([see part 1](#useR)), and the many packages that have been developed for use within the R environment [@rcoreteam2021r].
+Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’s books is R ([see part 1](#useR)), and the many packages that have been developed for use within the R environment [@rcoreteam2021language].
 The distinct advantages that this produces are the almost limitless additions and details that you can make to these figures to produce exactly what you want, in a [vector format](#vector).

### 2023-12-29 00-author.Rmd
Add files via upload
#general update of all files
Add files via upload
# with new chapter on planning

--- 2021-08-01 00-author.Rmd
@@
 
-John Measey is Associate Professor of Biological Sciences at Stellenbosch University.
-He has authored or co-authored more than 200 peer reviewed scientific papers and book chapters, and five books.
-He has been the Editor-in-Chief of an ISI journal for 9 years, and currently serves as Associate Editor for 4 other journals.
-He has graduated more than 20 postgraduate students, and his blog on writing and publishing in biological sciences is read by thousands globally.
-British born and educated, he lives and works in the beautiful Western Cape, South Africa.
+John Measey is Professor at the Institute of Biodiversity, Yunnan University, and Professor Extraordinary at Stellenbosch University.
+He has authored or co-authored [more than 250](https://orcid.org/0000-0001-9939-7615) peer reviewed scientific papers and book chapters, and six books.
+He has been the Editor-in-Chief of an ISI journal for 9 years, and currently serves as Associate Editor for three other journals.
+He has graduated [more than 25 postgraduate students](http://john.measey.com/About-Me/Measey-Mini-CV), and [his blog](john.measey.com/blog) on writing and publishing in biological sciences is read by thousands globally.
+British born and educated, he lives and works in the Spring City, Kunming in Yunnan Province, China and in Cape Town, South Africa.
 
@@
 
+## How to cite {-}
+Measey, J. (2022) *How to write a PhD in Biological Sciences: a guide for the uninitiated*. CRC Press, Boca Raton. Online Version: **`r Sys.Date()`**  ISBN: 9781032080208
+
 ## Creative Commons License {-}

### 2023-12-29 1.01-beforeyoustart.Rmd
Add files via upload
# with new chapter on planning

@@
 
-## Think about your career before you start your PhD
+## Think about your career before you start your PhD {#career}
 The last thing a PhD should be is a stop gap for you to fill your time because you can’t think of anything else to do.
@@
 
-## Don't rely on becoming an academic
+## Don't rely on becoming an academic {#relyacademic}
 Academia is getting more competitive.
@@
 On the other hand, it can simultaneously bring joy and start you off on a lifetime trajectory.
+
+>I don't think students are made to understand how hard it is to do research.
+And how very, very hard it is to do important research.
+It's a lot harder than taking even very demanding courses.
+What makes it difficult is that research is immersion in the unknown.
+We just don't know what we're doing.
+We can't be sure whether we're asking the right question or doing the right experiment until we get the answer or the result.
+>
+>Martin A. Schwartz [-@schwartz2008importance]
+
+## The importance of being productively stupid
+
+There will be many points during your PhD when you feel stupid because you don't know what is the right thing to do.
+This is normal.
+Indeed, when you feel that you do know exactly what you are doing, this might be a more worrying moment that gives you more pause for thought.
+You will need to get your head around consistently feeling stupid and feel comfortable with saying 'I don't know' without it causing you to feel like giving up.
+Schwartz [-@schwartz2008importance] talks about the importance of embracing that feeling of absolute stupidity when faced with many aspects of science.
+Facing that stupidity on a daily basis is necessary in order to have the correct attitude to your own studies.
+You can only really expect that your thesis will make you a little less stupid in the subject area of your choice.
+However, it may end up with you embracing more stupidity in ever increasing areas.
+When you feel comfortable with this stupidity, you'll know that you have achieved one of the most important qualities of a scientist.
 

### 2023-12-29 1.015-plan.Rmd
Add files via upload
# with new chapter on planning

# Make sure you have a plan {#Plan}

> “The Cylons were created by man. They rebelled. They evolved. There are many copies. And they have a plan.”
>
> [Battlestar Galactica](https://www.imdb.com/title/tt0407362/)

Having a plan is an essential part of your PhD strategy. You might think of it as a shining star guiding your way through the dark times. Or, perhaps like me, you are prompted to remember the Cylons in the sci-fi series Battlestar Galactica. However it takes you, having a plan will hold you in good stead throughout your PhD as it will ensure that you have something beyond your PhD to focus on.

There is a temptation, once you have started and invested many years in academia, to simply fall into step and see where it leads you. This lack of a plan will see your life collapse around you once you finish your PhD and wonder what it was all for. Or, more likely, you will get part way through your PhD and start questioning exactly why it is taking up so much of your life.

This chapter aims to help you build a plan for [your future career](#career) and aid you to see how you can improve your chances on reaching your goal.

## What is a long-term plan?

A long-term plan sets a goal for you in 10-15 years from now. This is likely to be a career choice, like joining a particular business or having a position in a certain company. Beware of having a long-term plan to become an academic, as there may not be opportunities (see [Part I](#relyacademic)). Equally, your long-term plan could be about the lifestyle you want to achieve in the long-term, or about where you want to be in the world. All of these are valid plans and they are totally up to you to envisage.

This may be the first time you are thinking about it, or you may have always known what your long-term plan was. In either case, once you have decided on your plan you next need to decide on medium- and short-term goals that will guide your decisions during your PhD years, and the moves that follow.

## Long-term plans need medium- and short-term goals

You need to consider your three time periods:

 - **Long-term** (10 - 15 years)
 - **Medium-term** (5 years)
 - **Short-term** (2 years)

The first thing to realise when you are planning for your future is that long-term plans need action in the short- and medium-term. It is only by achieving these actions that will help you reach your long-term goal. Personally, I like the three time period idea as short-term plans might only include getting something ready for the medium-term, and not necessarily doing that action.

If you don't like to think of specific time periods, you can think of the short-term as the time during your PhD. Medium-term is the years after your PhD and long-term is your career path. Short-term planning will consist not of having to find your primary occupation, but adding tools to your tool-box in order that you have a complete tool box for the long-term. Similarly, you can think of medium-term not as being employed in the actual job you want to do, but in a step along that particular career path.

(ref:plan) **Having a plan during your PhD is invaluable** This diagram shows how your short-term plan is related to your medium-term planning, and how that will influence your long-term goals. In turn, your long-term plans will impact on what you do in the medium- and short-term.

```{r plan, fig.show='hold', out.width = '65%', echo=FALSE, fig.align='centre', fig.cap="(ref:plan)"}
knitr::include_graphics('figures/plan.png')
```

***

### Teaching example

If your goal after completing your PhD includes teaching in secondary or tertiary education, you are going to want to make sure that you have teaching experience. Planning to have a teaching experience sufficient to put on your Curriculum Vitae (CV) in the next 2 years is probably unrealistic as you may well be in the midst of your PhD studies (obviously, in some institutions you may get immediate opportunities). But you can plan that within the coming two years you will know exactly when in the medium-term your teaching experience will happen. In turn, this will require that in the short-term you need to actively look for those opportunities.

### Research example

If your goal after completing your PhD is for a career in research, then there are definitely steps that you can take in the short- and medium-term that will increase your chances of success. Most research positions require fundraising as a central component of, what is essentially, project work. This means that you will want to have experience of fund raising, probably through [grant writing](#funding). You might not need to write a grant in the short-term, but you should be looking for opportunities of granting bodies to apply to, and projects that you have the capacity to undertake in the medium-term. Such projects might not be you on your own, but with other people in your group or institution, or connections in [your network](https://howtopublishscience.org/advisorless.html) from another institution.

You will also want to look for opportunities to publish. Publishing your thesis chapters is likely to follow in the medium-term (i.e. once you've finished your thesis), but there are likely to be opportunities before this to start getting publications on your CV. Remember that these might not be research papers, but could be publishing a [literature review](https://howtopublishscience.org/whattowrite.html#reviews) or [another type of paper](https://howtopublishscience.org/whattowrite.html) [@measey2022how].

### Other additions to your CV

Teaching, grant writing and research papers are all solid items to add to your CV during your PhD studies, but you may want to consider having some additional items if these will be skills required in your long-term plans. For example, you may be interested in having experience in administration and management. It may well be that you can volunteer to fulfil these roles in your research group or department. For example, you may want to get elected into a student representative role and sit on the department's management committee. Or it may be that someone has some administration exercise that you can help them with. You won't know unless you ask.

***

## Consider the letter of reference that your advisor will write

An important consideration for your medium-term plan is that you will likely be asking your advisor to write a letter of reference for you. In the short-term you can find out from people who already work in your long-term career choice what the ideal reference letter consists of. Once you know what is needed, you can in the short-term begin to make sure that you have all of these elements in place. For example, your advisor can't write a letter of reference that says you have great organisational skills unless you have helped them with organising something. If, like me, you have a bad memory for things that you have done, you will also need to write down these activities and examples of when you did them so that you can remind your advisor when the time comes for them to write a letter of reference.

Advisors are busy people, and they don't always notice all of the different things that you are doing. So make sure that you remind them what you have achieved during your time with them, especially when it comes to writing that letter of reference. Don't rely on them to remember!

## Changing your long-term plans

Don't worry about changing your long-term plans. Life often happens, and you may want or have to change your plans. In the short-term, changing long-term plans is easy. You should realise that the further you go down a particular career path, the more difficult it becomes to make long-term changes. However, if you do, then it is not impossible, and will need you to make some short-term plans in order to make those changes.

The other book has advice for [getting a job inside academia](https://howtopublishscience.org/jobinacademia.html), as it also does for [leaving academia for a job outside](https://howtopublishscience.org/leavingacademia.html) [@measey2022how].

## Share your plans

Don't forget to tell the people around you of your plans for the future. Remember that they may know about opportunities that they can only pass onto you if they know what your plans are. They may also be able to provide you with specific advice about exactly what experience you need.

Having a plan is probably one of the most valuable assets that you can hold throughout your PhD days. If you have read all of this and still don't know what yours is, then I'd suggest that you start to think about this now, and don't push it into the future when not having a plan is going to be increasingly problematic.

### 2024-03-15 2.03-NoHyp.Rmd
Add files via upload
# 2024 updates

@@
 So far, we have concentrated on thesis chapters or manuscripts that have [hypotheses](#hypothesis) to test.
-I have emphasised that there is a need to focus all of the sections of your writing to helping frame and understand the hypothesis (in the [Introduction](#introduction)), explain your approach in collecting and testing the variables (in the [Materials and Methods](#methods)), provide the results of your tests (in the [Results](#results)), and respond to the hypothesis (in the [Discussion](#discussion)) together with placing it in context.
-But what, I hear you cry, if you don’t have a hypothesis to test?
+I have emphasised that there is a need to focus all of the sections of your writing to help frame and understand the hypothesis (in the [Introduction](#introduction)), explain your approach in collecting and testing the variables (in the [Materials and Methods](#methods)), provide the results of your tests (in the [Results](#results)), and respond to the hypothesis (in the [Discussion](#discussion)) together with placing it in the context of other work.
+But what if you don’t have a hypothesis to test?
 
 There are lots of reasons why manuscripts don’t have hypotheses, and the chances are that in a PhD thesis (for example) you will have at least one chapter where there isn’t a hypothesis tested.
+It's also possible that you won't have any hypotheses tested, and this is the approach in some disciplines.
 So, given all the emphasis above, what can be done?
+
+## Is having a hypothesis the gold standard?
+
+The [hypo-deductive method](#whathypothesis) is certainly a powerful one to use in science but, no matter what you have read or been taught elsewhere, it is not the only scientific approach.
+Much of what we do in science comes from observational work, and these observations are usually the basis for many of the hypotheses that we pose.
+But some disciplines are entirely based on observations, and this does not make them obsolete or unscientific.
+Such disciplines simply don't lend themselves to experimentation, and without an experiment it is hard to control and hypothesise.
+But we should not confine ourselves to an entirely experimental approach.
+Here are some reasons why using non-experimental approaches will benefit our science:
+
+1.
+Observational science (something that I refer to elsewhere as [Natural History Observations](https://howtopublishscience.org/whattowrite.html#natural-history-observations)) can teach us a lot more about systems and inspire greater creativity in our work than plodding from experiment to experiment.
+1.
+We can infer patterns from process.
+As humans, we are evolutionarily programmed to see patterns (even when they aren't there) and recognise patterns in the world around us [e.g., @linhares2013what].
+It would be crazy to put this ability aside and ignore it when conducting science.
+Clearly, we are also able to make mistakes when we infer patterns from processes, but this does not mean that it should not be done.
+But we need to engage in a much more cautious manor when using this approach, always being aware of our bias towards certain explanations so that we encompass others.
+I would also argue here that this is instrumental in forming hypotheses to test experimentally or in models.
+1.
+Models of systems are powerful tools to test questions.
+Models are always simplified versions of complex biological systems, but their power as tools in our scientific approach should no longer be questioned.
+
+Indeed, there are many shortcomings to using a hypo-deductive approach, as we have seen in the [previous chapter](#TypeI).
+By allowing ourselves a liberal approach to science methodology we are far more likely to make new connections and infer novel patterns about biological processes.
+We are more able to be inclusive of different disciplines.
+It is likely to be advantageous to be open to these diversity of approaches in others' research, just as you should in your own approach.
+
+## Why am I always banging on about hypotheses?
+Throughout this book (and in [other books](https://howtopublishscience.org/)) I constantly refer to hypotheses and how important they are when writing your thesis chapters.
+Why do I do this when I am clearly so positive about using other methods?
+In many respects, this comes from my attempt to simplify.
+In most cases, you could replace the word "hypothesis" with "central problematic" and get the same effect.
+You could also replace some instances with "aim" or "objective" (but not all).
+There are some points that are unique to hypotheses, and that's why I made my choice about referring to these as they are, generally speaking, the most complex of these alternative approaches.
+I reason that if you can understand how to approach writing to respond to hypotheses, then you can turn this same approach to your own needs.
+Add to this that most people, scientists included, misunderstand hypotheses, so there is a need to discuss them and their uses more.
+
+If you are a biological scientist that does not use hypotheses, then this does not make you inferior, and I hope that you will still find this book useful.
+If this is you, then let me apologise here and suggest that you mentally shift whenever you read "hypothesis" to the alternative approach that you prefer.
 
@@
 ## Does it matter that you don’t have a hypothesis?
-Having a hypothesis makes your work stronger than if you only have a question or prediction.
+Proponents of the hypothetico-deductive method will argue that having a hypothesis makes your work stronger than if you only have a question or prediction.
 In each case, while your results might help answer your question or confirm your prediction, they won’t test the mechanism (so that you can’t explain why you have the results that you do), and they may not be falsifiable/refutable.
@@
 
-To me, it is important for you to know exactly what your study is before you start writing about it, or even before you do the experiments.
+## Avoid HARKing 
+There is one very important aspect of your approach whether or not you have a hypothesis, and that is that you know exactly what your study aims are before you start.
 It is essential for you to frame the context of your study.
-The central reason why you are writing the study will lie at the heart of the manuscript.
-In the formulaic approach, it will be verbalised at the end of the introduction.
-All parts of the methodology will explain how you approached this central aim.
-So if you can’t pin down what this is, then make sure you discuss it with your advisor.
+The central reason why you are conducting the study will lie at the heart of the manuscript, once you've started writing.
+In the formulaic approach (see [Part III](#formula)), it will be verbalised at the end of the [introduction](#introduction).
+All parts of the [methodology](#methods) will explain how you approached this central aim.
+So if you feel that you still don't know what the objective of a certain chapter is, then make sure you discuss it with your advisor so that you know before you start.
 
-## Avoid HARKing 
 Hypothesising After Results are Known (HARKing) is the practice of creating your hypothesis once the results are in and you’ve done some preliminary data analyses to see what is significant [@forstmeier2017detecting].

### 2024-03-15 2.05a-AI.Rmd
Add files via upload
# 2024 updates

@@
 In this chapter I set out some ideas about how you might use AI to help you write, as well as some things that you should consider if you do go that route.
+Bear in mind that this field is moving extremely quickly and it will not be possible to keep up with every change happening in AI all of the time.
+Instead, I am to discuss more generically about how to use these tools.
 
 ## What is a Large Language Model (LLM)?
-I decided to ask ChatGPT and Bard to contribute to this chapter by describing themselves.
-My input text: "I am writing a chapter for a book on academic writing and I want you to describe ChatGPT in your own words".
+Back in February 2023, I decided to ask ChatGPT and Bard (as it was then) to contribute to this chapter by describing themselves.
+My input text: "I am writing a chapter for a book on academic writing and I want you to describe ChatGPT/Bard in your own words".
 Here's what they said:
@@
 
+### Failure to learn
+Lastly, a great reason not to use AI to help you write you thesis chapters is that you will learn next to nothing this way.
+Even if you aren't going to continue in science, writing a PhD is a significant learning experience, and you will benefit in being able to write reports and other types of documents since you will have had the experience of going from having nothing, planning what to write and then executing this.
+Don't underestimate the importance of this ability.
+The reality is that once you have accomplished writing your thesis you will have an important and marketable skill.
+
+Using AI to generate text, even if it gets to the point in the future when it generates something useful, will impact your ability to generate text yourself.
+I can imagine a future where far fewer people are learning how to write because there is so much AI generated text.
+In this future, having the ability to write yourself will give you an advantage that should be valued.
+
 ## Positive aspects of using AI
@@
 
-#### The results of my experiment are in - both bots failed {#AIfail}
+#### The results of my experiment are that both bots failed {#AIfail}
 
@@
 
+In late 2023, an article in Nature [@conroy2023scientific] highlighted research by Guillaume Cabanac that had found the phrase "regnerate response" in more than 30 published papers.
+It is suggested that the authors have been so sloppy with their use of AI that they have also included the text on the button that occurs with the output.
+The examples spotted include the most crude errors and likely represent a very small proportion of AI generated text now published.
+The real question remains why so few people are critically reading manuscripts before they are published.
+Let us not forget that this is not the first time computer generated text has been published by "reputable publishers".
+An article from 2014 talks about how use of the crude text generation software *SCIgen* had been outed in dozens of published papers [@sample2014how].
+
+Now (in March 2024) that I've had more than a year to evaluate AI tools that write text, I am still of the opinion that AI is not a solution, but can be used as a crude tool.
+In essays that students have submitted for assessment, **it has been easy to spot those who have used AI** to generate their text.
+How?
+Because AI gets the facts wrong, but still makes it sound good.
+These mistakes are really unlikely for a human to make.
+For example, AI will use real species names to summarise the results from an experiment, but uses the wrong name.
+The AI interpretation of what the experiment is about and what the findings are is often faulty, and can sometimes be wildly inaccurate.
+I suspect that students who were more diligent in what they handed in, were unlikely to have gained much advantage to using AI.
+Maybe it helped them get started, or maybe it helped them find some useful phrases, but there is still little help from generating text and using it without considerable reworking.
+
 There will likely be better AI chat bots in the future, and ChatGPT itself is constantly getting feedback from users that should improve its own output.
+Hence, we can expect that current problems will change.
 
@@
 
+### Universities are asking you to declare that text in your thesis is not AI generated
+It would be worth checking with your institution to find out what their latest guidelines are.
+I have already received PhD theses to examine where there are declarations made that no text is AI generated.
+It is very likely that most institutions will make policies on this soon, if they have not done so already.
+
 ### COPE have already provided guidelines
@@
 In general, these rules emphasise the importance of oversight and transparency in the use of AI tools in decision making.
+
+### Publishers now make it a requirement that you declare no text was AI generated
+Many publishers are now making it a requirement that you declare on submission of a manuscript that it was not AI generated.
 

### 2024-03-15 2.17a-courses.Rmd
Add files via upload
# 2024 updates

# Choosing your courses

As a postgraduate student, your university should offer you a number of courses designed for postgraduate students. In many countries, these postgraduate courses will be compulsory and admission to the graduate school (Admission to Candidacy) will only be made after you have completed sufficient courses with grades to qualify you. For example, in the USA there are normally two years of these postgraduate courses followed by three years of working on your thesis project.

Given that this is actually a really long time, it's worth thinking about how you can get the most out of the courses that you choose, and how best to use this time in general. In this chapter, I will not tell you what courses to take, but try to give you some insights into how to pick the courses that you are going to get most use from, both during your PhD studies and in your life thereafter.

## How to choose your courses

1. There will be some courses that align directly to the contents of your thesis and you should do them
1. There should be a [writing course](https://johnmeasey.github.io/Thesis-Writing-Guide/) (and if not you can follow a link to my own) that you should attend as everyone needs help with writing
1. A time management course would also be very helpful, if your university has one
1. Look back at your short and medium term goals (see [here](#Plan)) and consider which courses could be relevant in your future
1. Do some courses just because you are interested. Given that you have already done a lot of courses at university, you are going to need some that hold genuine interest for you. Remember that diversity is an important aspect of science, so think as widely as possible.

### Attending a postgraduate course at another institution

Some universities allow their students to attend postgraduate courses at other (usually nearby and affilitated) universities. If you know of a course that it being conducted at another university, but that you would like to attend, it might be worth approaching your postgraduate teaching co-ordinator to ask them if you can go.

There is also the opportunity to conduct a postgraduate course online. Again if you know of one, or put in the time to find an appropriate one, it may be worth approaching your postgraduate teaching co-ordinator to ask if you can do it.

The most important consideration is whether or not your institution is able to use marks generated at another institution in their own system to assess your postgraduate courses. Even if they don't allow you, it's worth talking to them about it as it may help those that come after you to find the course that they want. Or it might help stimulate to offer something similar at your institution in future.

## Due diligence during courses

You need to make sure that you get the most out of these courses.

Looking back at being an undergraduate, you may have been focussed on grades or shortcutting work on courses, but for postgraduates this must have changed.

## Getting the most our of your courses

Remember that it is not direct applicability that is of the only importance.

## Building your network

Courses are designed for postgraduates and usually taught by the most experienced professors. Use your time with them to [increase your network](https://howtopublishscience.org/advisorless.html#networks).

In addition to the professors that are giving your courses, you will also learn about other academics at different institutions. While these people may be far from you, their work can be very relevant to your own. Using your time wisely on your course also involves reading work by these people. If your advisor agrees, then it may be possible for you to approach some of these top researchers to participate on your committee.

## Time off from courses

As you are in the biological sciences, there may be good reason why you need to go on a trip and not attend part of a course. For example, your advisor may ask you to participate in lab fieldwork that is relevant to your project. You may want to attend a conference to present the results of your MSc work. You may have any number of private or professional reasons why you need to miss lectures or parts of your course.

 - Inform the course co-ordinator. Your institution may have a system where you require to be signed out of the class. Make sure that the course co-ordinator knows, and the lecturer of the missed classes (if this is a different person).
 - Catch up. Be diligent and catch up with all of the missed course work. 
 - Try to miss as little as possible. Taking your courses seriously means missing as little as possible.

## Make the most out of learning

I you haven't already worked this out for yourself, learning is a real gift and the postgraduate courses you are offered at universities are probably the ultimate in their level of sophistication and knowledge transfer. Take this opportunity seriously and you should find it extremely rewarding.

### 2024-03-15 2.19a-concept.Rmd
Add files via upload
# 2024 updates

@@
 
-- Do you have hypotheses?
+- Do you have hypotheses, or do you have another approach?
 
@@
 
-## Your thesis timeline {#Gendt}
+## Your thesis timeline {#Gantt}
 
@@
 
-### Producing a Gendt diagram
-The timeline of your thesis will be an important guide throughout the 3 years that you are conducting your PhD.
+### Producing a Gantt chart
+The timeline of your thesis will be an important guide throughout the 3+ years that you are conducting your PhD.
 This means that it is worth getting this onto the record as soon as you are able.
-One of the standard ways of doing this is to produce a Gendt diagram.
+One of the standard ways of doing this is to produce a Gantt chart.
 This takes each of the chapters along the entire timeline of your PhD and shows when you will work on each one.
@@
 
-Below you will find a very simple version of a Gendt diagram (Figure \@ref(fig:Gendt)) that I put together for a fictional PhD.
+Remember that [planning  your time](#Plan) is empowering and something that you should be considering not only during your PhD (short-term), but also mid-term and long-term (see [Part I](#Plan)).
+To achieve your goals, you need to (a) know what they are, and (b) actively pursue them.
+
+Below you will find a very simple version of a Gantt chart (Figure \@ref(fig:Gantt)) that I put together for a fictional PhD.
 Note how the colours used for each chapter are the same as the thesis concept diagram (Figure \@ref(fig:concept)).
@@
 
-(ref:Gendt) **A Gendt diagram used to help thesis plan how you will use your time during your PhD**.
+(ref:Gantt) **A Gantt chart used to help thesis plan how you will use your time during your PhD**.
 In this (fictional) example, I show how how the different chapters (Q1, T1, Q3, Q4, Q5) will take up different time during the three years of a PhD.
@@
 
-```{r Gendt, echo=FALSE, out.width = '95%',  fig.cap="(ref:Gendt)"}
-knitr::include_graphics('figures/Gendt.png')
+```{r Gantt, echo=FALSE, out.width = '95%',  fig.cap="(ref:Gantt)"}
+knitr::include_graphics('figures/Gantt.png')
 ```
@@
 As you're doing your thesis in the biological sciences, you're much more likely to have some kind of seasonal quotient that constrains the work that you're doing.
-So in Figure \@ref(fig:Gendt) I've used the second quarter of each year as the key field period that is needed for all of the chapters.
+So in Figure \@ref(fig:Gantt) I've used the second quarter of each year as the key field period that is needed for all of the chapters.
 
@@
 
-It's really useful to set down goals that you want to achieve on the Gendt plan, and to share these with your advisor.
+It's really useful to set down goals that you want to achieve on the Gantt chart plan, and to share these with your advisor.
 Milestones in your PhD are really useful when planning, and psychologically enhance your ability to finish different components.
@@
 For example, If you are presenting the results of your first chapter at a conference, you need to make sure that you have the necessary results well in advance of needing to write the abstract.
-In Figure \@ref(fig:Gendt), the first conference is in the third quarter of the second year and you should have already achieved your main goal for this chapter well in advance to write the abstract.
+In Figure \@ref(fig:Gantt), the first conference is in the third quarter of the second year and you should have already achieved your main goal for this chapter well in advance to write the abstract.
 You might also be able to add in the results of the second year's seasonal field work to present at the conference.
@@
 
-In Figure \@ref(fig:Gendt) I have shown how a contingency might be mapped onto Chapters Q1, Q3 and Q4 with the potential addition of another field season in the second quarter of the third year in case one of the previous field seasons fails.
+In Figure \@ref(fig:Gantt) I have shown how a contingency might be mapped onto Chapters Q1, Q3 and Q4 with the potential addition of another field season in the second quarter of the third year in case one of the previous field seasons fails.
 

### 2024-03-15 3.03-Introduction.Rmd
Add files via upload
# 2024 updates

@@
 Point out the relevant sites on the way, but don't stop until you get where you're going.
+
+### The hypothesis at the heart of the formula
+
+Having made the point that the formula is funnelling the readers' interest, the question arises: to what?
+What are we working towards?
+Why are we so interested in having the reader focus on a particular thing, and more importantly: what is that thing?
+
+> The Hypothesis
+
+Your hypothesis is the reason that you are conducting your study.
+You are trying to respond to it, trying to accept or reject it.
+Your hypothesis is why you are doing the study and you need your reader to get the context for this, and to understand exactly why it was that you posed this specific hypothesis.
+
+The hypothesis is then at the heart of everything that you write in your chapter (Figure \@ref(fig:formula-concept)):
+
+ - Your introduction explains why you posed the hypothesis
+ - Your methods & materials describe how you went about testing your hypothesis
+ - Your results provide the outcome of the tests of your hypothesis
+ - Your discussion responds to your hypothesis and provides the context
+
+(ref:formula-concept) **Your hypothesis lies at the heart of your PhD chapter.** Your hypothesis (red star) is at the heart of everything that you are writing about.
+The entire reason for providing the chapter is to test your idea.
+
+```{r formula-concept, echo=FALSE, out.width = '95%', fig.cap="(ref:formula-concept)"}
+ knitr::include_graphics("figures/formula-concept.gif")
+```
 
@@
 
+I suggest that you keep your outline with you as you write and develop the introduction.
+It is important that you remember the objective of each paragraph.
+Sometimes, you can put it away, then when you bring it out again and start working on it you find that it changes into something else.
+At this point, you need to make a judgement on whether it has changed into something better.
+You can only really make this judgement if you remember what your original objective was.
+When you think that it has become a better paragraph, you need to check that it still meets your original objectives.
+If it doesn't, then you'll need to decide whether or not that objective was met elsewhere.
+
 Now that you’ve fleshed out your outline with relevant citations, it’ll be time to pass it by your advisor to check that you are moving in the right direction before you start writing.
@@
 
-Remember that while this section on writing the introduction is short, much of [Part II](#part2) concerns writing style that is applicable to your introduction.
-In particular, sections on composing your [hypothesis](#hypothesis), writing a [paragraph](#paragraph), and an [argument](#argument) should be referred to when you are writing your introduction.
+### Should you include the results when introducing your study?
+
+> How do you keep an idiot in suspense? I'll tell you later...
+
+Increasingly, more scientists are including a concise statement during the last paragraph that responds to the hypothesis and states what the major outcome of the study is.
+The arguments for and against this have been made elsewhere [@heard2024scientific], and I will attempt to summarise them here:
+
+**Against:**
+
+ - This is not the convention in science 
+ - Results do not appear in the introduction
+ - You are giving away the result
+
+**For:**
+
+ - Science writing is not like writing a murder mystery novel
+ - If it helps readers then we should do it
+ - Convention is not a good reason for adherence
+
+In this guide, I have emphasised that the [formulaic approach](#formula) to writing advocated here is to help you get started as a writer.
+One of your goals should be to experiment with your writing and if you think that moving away from the formula is going to improve clarity and readability, then I think you should go for it.
+Next, I think that you should elicit feedback from peers and your advisor.
+The formula I have presented is conventional, and that's deliberate.
+I am not going to tell you that it is the best way of writing your particular thesis chapter, and I would go so far as to say that shaking up convention is a very positive thing to do.
+
+But if you are struggling with getting started, then I am going to continue to advise you to stick to the conventional way.
+As it's tried and tested you can be confident that this way is going to advance you towards your goal of writing your PhD.
+
+## Summing up
+
+The aim of your introduction is to provide the reader with all of the background information necessary to understand your study.
+Although the contents of this section on writing the introduction are somewhat brief, much of [Part II](#part2) concerns writing style that is applicable to your introduction.
+In particular, sections on composing your [hypothesis](#hypothesis) ([or not](#nohypothesis)), writing a [paragraph](#paragraph), and an [argument](#argument) should be referred to when you are writing your introduction.
 

### 2024-03-15 3.05-Results.Rmd
Add files via upload
# corrections after July 2021 Copy Edits
Add files via upload
# 2024 updates

@@
 However, there are still many mistakes that are made.
-Let’s start with the approach that you should take, and then go through some of the common mistakes.
+Here, I will start with the approach that you should take, and then go through some of the common mistakes.
 
-## The correct approach - responding to your hypotheses
-Yes, the easiest way to write the results is simply to have the questions or hypotheses that you posed at the end of the introduction and have them as subheadings for your results section (Figure \@ref(fig:results)).
-Once you have these subheadings, you simply assemble all of the results with respect to that hypothesis.
-You should already have all the tests that you used in the methods section.
+## Description of your data
+Some studies have a set of descriptive results that, by convention, are placed without a section header at the start of the results.
+These normally convey the size of the dataset that entered into the rest of the analyses.
+If the size of the dataset varies by design in the methods, then you don’t need it here.
+But you can’t always control the sample size, so it needs to be established before moving on to the rest of the results.
+It may be that your methods do not always produce predictable results.
+In genetic studies, for example, this might include the number of samples that were collected and successfully sequenced, and thus entered into the rest of the analyses.
 
-The results section often includes a lot of statistical testing, resulting in lots of numbers inside brackets within the text.
-This is all expected and fine, but consider consolidating multiple iterations into a table (if appropriate).
-And if your tests are already in a table, then they should not be repeated in your text.
-Instead, simply make your statement asserting the results and add “(see Table 1)”.
+Whenever you provide a summary of a variable (e.g.
+the mean), remember to include the variance; the range,  standard deviation, or standard error in brackets following the mean.
+If you collected a lot of variables, it may be more relevant to have them summarised in a table.
 
-When describing the relationship between two variables, make sure that you state the direction of the effect.
-This means not simply stating that they have a significant relationship, but what that relationship is.
-This means stating that one increases with the other (a direct relationship) or one increases when the other decreases (an inverse relationship).
-Note that other types of variable relationships exist.
-Once you have described the result, either make reference to the Table or Figure where this can be seen, or describe the results of the test in brackets.
+Once you have described the result, either make reference to the [Table](#tables), [Figure](#figures) or [Supplementary Information](#suppinfo) where this can be seen.
+If your tests are already in a [table](#tables), then they should not be repeated in your text.
+Instead, simply make a summary statement asserting the results and add “(see Table 1)”.
 
-If you do have to insert the results of a statistical test into the text, include the test statistic, degrees of freedom (which incorporates N - the sample size) and the value of P.
-Only if P is less than α (alpha is usually 0.05) can you use the word ‘significant’ when describing the effect.
+It is important to include mention of how much data was missing compared to your aim in the [Material and Methods](#methods) section.
+For example, an entire replicate may have been compromised, or a collection date might have been skipped for logistical reasons.
+
+If you are [not testing an explicit hypothesis](#nohypothesis) your results will concentrate on the description of your data, and your analysis of the pertinent information that it contains.
+This may be done through tables, figures or models.
+If you modelled your data, you will need to show all models tested either in a table in the results, or in the [Supplementary Information](#suppinfo).
+You will also likely need to describe how you selected the best model and describe it.
+Although it is correct to describe your model here, do not try to interpret your findings.
+This is done in the [discussion](#discussion).
+
+## Responding to your hypotheses
+The easiest way to write the results is simply to have the questions or hypotheses that you posed at the end of the introduction and use them as subheadings for your results section (Figure \@ref(fig:results)).
+Once you have these subheadings, you simply assemble all of the results that address that hypothesis.
+You should already have all the tests that you used in the [Material and Methods](#methods) section.
+No additional tests should be listed in the results if they were not already described in the [Material and Methods](#methods) section.
 
@@
 ```
+
+Remember that the most important aspect to your results is the size of any effect that you find and the confidence intervals, and **not** whether or not it is significant [@dushoff2019can].
+This is not to say that significance should not be reported, this is important, and care should be taken to report the significance in a non-binary fashion [@dushoff2019can; @muff2021rewriting] - see [below](#binary).
+
+The results section often includes a lot of statistical testing, resulting in lots of numbers inside brackets within the text.
+This is all expected and fine, but consider consolidating multiple iterations into a [table](#tables) (if appropriate).
+
+When describing the relationship between two variables, make sure that you state the direction of the effect.
+This means not simply stating that they have a statistical relationship, but what that relationship is.
+For example, explaining that one increases with the other (a direct relationship), or one increases when the other decreases (an inverse relationship).
+Note that other types of variable relationships exist.
+When describing this relationship, you should also include the effect size (how much the relationship differs from no relationship), and the [power of your test](#increasing-the-statistical-power) (usually the number of replicates that allows you to avoid a [Type II error](#TypeI)).
+
+### Moving away from binary significance {#binary}
+
+Traditionally, we would say that only if P is less than α (alpha is usually 0.05) should you use the word ‘significant’ when describing the effect.
+However, the statistical community (including the biological community) is waking up to the need to replace the idea of the binary reporting of significance ("significant" or "not significant"), with wording that better reflects the reality of statistical testing " [see @muff2021rewriting].
+The reason for this is that it represents a misunderstanding of statistical testing, and can be very misleading to those that read the results.
+Instead, we must appreciate that the position of our estimate of the P statistic lies somewhere along a continuum of potential results, and should be reported as such.
+The suggested change in language can be simply implemented as shown in Figure \@ref(fig:results).
+
+(ref:replace-significance) **Help your reader to better interpret the meaning of your statistical testing by changing your language.** The figure shows some suggested language to use when writing about your results depending on the value of P that you find.
+These words should replace the binary outcome "significant" or "not significant" [see @muff2021rewriting].
+
+```{r replace-significance, echo=FALSE, out.width = '45%', fig.cap="(ref:replace-significance)"}
+knitr::include_graphics('figures/replace-significance.png')
+```
+
+While this approach succeeds in getting away from the significance dichotomy, there are still dissenting voices with the concept of P values representing continuous measures of evidence [@goodman1988evidence].
+Another approach is to talk about the results of your significance tests in terms of the clarity that they provide about the relationship between the variables measured [@dushoff2019can]: whether they show a "statistically clear relationship" (aka significant), or "statistically unclear" (aka insigificant) [see @dushoff2019can for examples].
+Here, I stress that if you have a lower value for P (e.g.
+equating to strong evidence) you haven't provided any additional evidence, you simply have more confidence in the non-compatibility of your data with your null hypothesis (H~0~).
+Your result is confined to the dataset that you are presenting.
+If you want additional evidence, then you need to run replicate experiments.
+Or, better still, find another mechanism in the biological sciences through which you can measure or test the same relationship.
+
+There are better alternatives when it comes to statistical testing, but getting the inferential wording correct may still represent a challenge for non-statisticians.
+
+If you have a statistical test with a P value, you do have to insert the results into the text, include the test statistic, degrees of freedom (which incorporates N - the sample size) and the value of P.
+My suggestion is that you also consider exactly what you can say about your result, the effect size, confidence intervals and inferred relationship between the variables, as well as any statistical significance.
 
@@
 
-See sections on [tables](tables.html) and [figures](figures.html) below.
-
-## Descriptive results
-Some studies have a set of descriptive results that, by convention, are placed without a section header at the start of the results.
-These normally convey the size of the dataset that entered into the rest of the analyses.
-If the size of the dataset varies by design in the methods, then you don’t need it here.
-But you can’t always control the sample size, so it needs to be established before moving on to the rest of the results.
-In genetic studies, for example, this might include the number of samples that were collected and successfully sequenced, and thus entered into the rest of the analyses.
-Your text can then be broken into sections that correspond with your questions, unless you only have one, in which case further sections are not warranted.
+See sections on [tables](#tables) and [figures](#figures) below.
 

### 2024-09-11 00-author.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

@@
 This book is written in bookdown [@xie2016bookdown] specifically to make it a 'live project' that will be open to anyone who wants to contribute, improve, or use as the basis for your own book.
-The easiest way for readers to contribute content directly is through a [GitHub pull requests](https://help.github.com/articles/about-pull-requests/).
-At the repository for this book, you will find Rmd files for each chapter, and as a GitHub user, you can simply edit the Rmd file and submit the changes.
+The easiest way for readers to contribute content directly is through a [GitHub pull request](https://help.github.com/articles/about-pull-requests/).
+At the repository for this book, you will find Rmd files for each chapter, and as a [GitHub user](https://github.com/signup?ref_cta=Sign+up&ref_loc=header+logged+out&ref_page=%2F&source=header-home), you can simply edit the Rmd file and submit the changes.
 If I am happy with the changes proposed, I will merge your content with that of the book and add your name to the [Acknowledgements](#acknowledge).
@@
 One of the amazing potentials for [bookdown books](https://bookdown.org/) is that all the files for this book are hosted in a repository on [Github](https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences/tree/main).
-You have the opportunity to fork this repository and write your own version for a different discipline, a different language or for a different region of the world.
+You have the opportunity to fork this repository and write your own version for a different discipline, a different language, or for a different region of the world.
 It is also my hope that this guide can grow to become a community of practice for those conducting PhDs in Biological Sciences.
-It will not be possible to cover every aspect of writing a PhD in Biological Sciences, but it may be that I have missed out ones that are very important to you.
+It will not be possible to cover every aspect of writing a PhD in Biological Sciences, and it may be that I have missed out ones that are very important to you.
 Equally, parts of what is currently written will become obsolete as new initiatives begin, and old problems are resolved.
 For this reason, this guide needs to be a 'living document', and anyone who wants to provide feedback or contribute new sections is more than welcome.
-Please feel free to open an issue, or make a Pull Request if you spot a typo.
+Please feel free to open an issue, or make a Pull Request if you spot an error.
 
 ## How to cite {-}
-Measey, J. (2022) *How to write a PhD in Biological Sciences: a guide for the uninitiated*. CRC Press, Boca Raton. Online Version: **`r Sys.Date()`**  ISBN: 9781032080208
+Measey, J. (2022) *How to write a PhD in Biological Sciences: a guide for the uninitiated*. CRC Press, Boca Raton.  262pp ISBN: 9781032080208  DOI:10.1201/9781003212560
+
+Measey, J.
+(`r format(Sys.time(), '%Y')`) *How to write a PhD in Biological Sciences: a guide for the uninitiated*.
+Online Version from https://howtowriteaphd.org/: **`r format(Sys.time(), '%d %B, %Y')`**
+
+[Download Bibtex file here](https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences/blob/main/citation.bib)
 

### 2024-09-11 1.00-Part1.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

@@
 Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you plan to do afterwards.
-There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make informed decisions from the outset.
-For example, reconciling yourself to using the [statistical package R](#useR) will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on.
-Preparing yourself into the right mental framework will also be very important for your ability to work under pressure.
+Both short-term and long-term planning will be one of the [core skills](#coreskills) that are the subject of Part I.
+You are likely to have had experience of each of these core skills before, but this part of the book tries to show you the importance of growing these skills and having the discipline to apply them regularly to achieve your goal of finishing  your PhD.
+Some of the most important core skills involve critical reading and regular writing, skills that you will need to begin acquiring and practising from the outset of your studies.
+Managing your time day-to-day will help you find time to practice these skills alongside your other commitments.
+There are a lot of things to think about as you start your PhD, and for many of these, it is far easier to make informed decisions from the outset.
+Preparing yourself for the right mental state will also be very important for your ability to work under pressure.
 Getting into good habits at the beginning will put you in a commanding position later.
-The aim of this first part of the book is to help you assess your current position, and realign yourself toward an optimum mindset.
+This first part of the book aims to help you assess your current position, and realign yourself toward an optimum mindset for doing your PhD in Biological Sciences.

### 2024-09-11 1.01-beforeyoustart.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

@@
-# So you are doing a PhD? {#beforeyoustart}
+# From the start of your PhD {#beforeyoustart}
 
-Universities are worried about two major aspects of those who start PhD studies:
-
- - The time to completion in many countries worldwide is much longer than it should be.
-For example, 7.5 years is the median time to completion in the USA [@geven2018how], Europe universities have largely adopted the Salzburg Principles that doctoral programmes should be three to four years for full-time candidates  [@christensen2005bologna].
-Following this, an average is recorded as being between 3.5 and 4.5 years, with most institutions having reduced this time over the last 10 years [@hasgall2019doctoral].
- - Completion rates are shockingly low. As bad as 50% in the USA [@litalien2015improving], 66% (within 6 years) in Europe [@hasgall2019doctoral], but generally better in the sciences [@park2005new].
-
-This has spurred many responsible for postgraduate students to try to ensure that they receive more support for their studies.
-But there is also the need for students to comprehend right from the start that they are undertaking a major commitment that will likely impact many different parts of their lives.
+The start of your PhD studies are a very turbulent time.
+You may be in a new institution with new rules to follow.
+You may be far away from the people at home and friends who make up your [support network](#support).
+You will be meeting many new people and finding lots of other people who are interested in the same subjects as you.
+You should be talking about your shared ideas and aspirations for your PhD studies.
+Although it is a turbulent time, it is also a very exciting time.
 
@@
 Therefore, to say that this is a big decision is an understatement.
-It is not a substitute for getting a job or a way of remaining a student for longer.
+It should not be a substitute for getting a job or a way of remaining a student for longer.
 If after reading this section you are in any doubt about whether or not a PhD is right for you, then I would suggest that you do need to think about it further.
@@
 Especially if their experience is in your focal research area, they are likely to know a lot more about it.
+
+Your advisor and your institution are also happy to have you start your PhD.
+Your research and findings will be important to them.
+But they will also want to make sure that you finish your studies in a timely manner, as too many students start but do not finish their work, and many take too long to complete their studies.
+Universities are worried about two major aspects of those who start PhD studies:
+
+ - **The time to completion** in many countries worldwide is much longer than it should be.
+For example, 7.5 years is the median time to completion in the USA [@geven2018how], Europe universities have largely adopted the Salzburg Principles that doctoral programmes should be three to four years for full-time candidates  [@christensen2005bologna].
+Following this, an average is recorded as being between 3.5 and 4.5 years, with most institutions having reduced this time over the last 10 years [@hasgall2019doctoral].
+Similarly, in China doctoral studies take three to four years to complete, although there are no official available statistics on time to completion [@powell2007doctorate; @zhu2024graduate].
+ - **Completion rates** are shockingly low.
+As bad as 50% in the USA [@litalien2015improving], 66% (within 6 years) in Europe [@hasgall2019doctoral], but generally better in the sciences than other subjects [@park2005new].
+Once again, there appear to be no official statistics for completion rates for those enrolled in Chinese PhD programmes [@zhu2024graduate].
+
+This has spurred many responsible for postgraduate students to try to ensure that they receive more support for their studies.
+But there is also the need for you as the student to comprehend right from the start that you are undertaking a major commitment that will likely impact many different parts of their lives.
 
@@
 Suffice it to say that this would be a very poor place to start, and if this were your attitude there may be lots of other reasons why doing a PhD really is not appropriate.
+
+It may be worth briefly considering the history of this qualification.
+Fifty years ago, a PhD was only taken by a small number of elite academics for whom it was treated as an apprenticeship to themselves become an academic [@houston2024final].
+This has now changed, and in the 21st century, a PhD has become a qualification that can also be sought after outside academia.
+In some countries, where industry has an important voice in the content of a PhD, the synergy between industry and education appears to lead to national prosperity [@frick2017death].
+This, in turn, leads to important questions about what a PhD is and who it is for, especially when one considers the use of PhD statistics by the World Bank (see [below](#lowpay)).
 
@@
 That is, that the work demonstrates that the person is a scientist at a broader level than the narrow focus of their research.
+
 - Third is whether the student has been able to bring the study together in a way that shows that it is more than the sum of its parts.
@@
 ## What will you do after your PhD?
-It sounds silly to think about what comes after your PhD just as you are just starting, but the reality is that you always need to be planning ahead in order to make the most of what you are doing now.
+It sounds silly to think about what comes after your PhD just as you are just starting, but the reality is that you always need to be planning ahead in order to make the most of what you are doing now (see more on planning [here](#Plan)).
 During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas.
@@
 
-### Doing a PhD actually reduces your chance of employment and lowers your pay
+### Doing a PhD actually reduces your chance of employment and lowers your pay {#lowpay}
 The reality is that most employers don’t want someone with a PhD, because it is a very academic qualification.
 It will take you so far down an academic rabbit hole, that for an employer there’s very little to use.
+PhDs were originally designed by academics as an apprenticiship for future academics [@houston2024final].
 Most employers want people who understand the nuts and bolts of doing research, but who can apply this knowledge outside of the academic system.
 Employers who want people with PhDs (and there are a few) generally want professional researchers who are essentially academics.
-Good work if you can get it, but it tends not to pay well.
-You might end up only being fit to be an academic, and that life is getting increasingly harder without any end to the expectations of the employers… like the equivalent of doing a PhD every year.
+Good work if you can get it.
+You might end up only being fit to be an academic (a traditionally low paid sector which is likely what skews the figures of salaries of PhD holders), and that life is getting increasingly harder without any end to the expectations of the employers.
 Something only very few people will enjoy.
@@
 The number of PhDs in a country is considered  (by the [World Bank](https://datacatalog.worldbank.org/dataset/world-development-indicators)) to be one of many measures of the development of that country.
-As the percentage gets larger, so countries can convince the World Bank that they are developing (and then get granted its favours, see Figure \@ref(fig:PhDMap)).
+This is also true in East Asia, where the dramatic increase of postgraduate degrees in the last two decades has primarily been driven by efforts to accelerate national economic development [@jung2022expansion].
+As the percentage of PhD holders per 100 000 of population gets larger, so countries can convince the World Bank that they are developing (and then get granted its favours, see Figure \@ref(fig:PhDMap)).
 This is all well and good, but strategically, governments should consider why they want these researchers in their populations, and more particularly for what kinds of jobs.
@@
 
-(ref:PhDMap) **Where in the world are there most PhDs?** In this map of the world, countries shaded are those that have data declaring the proportion of PhD (or equivalent) in their population (max percentage between 2010 and 2020 (Data from the World Bank).
-This ranges from 2.97% in Switzerland (dark red) to 0.02% in DR Congo (light yellow).
-Notably, there is no data held in these years for many countries in Asia, Africa and South America.
+(ref:PhDMap) **Where in the world are there most PhDs?** In this map of the world, countries shaded are those that have data declaring the proportion of PhD (or equivalent) in their population (max percentage between 2010 and 2024 (Data from the [World Bank](https://databank.worldbank.org/reports.aspx?source=Education%20Statistics)).
+This ranges from 3.138% in Switzerland (red) to 0.008% in Guatemala (light yellow).
+Notably, there is no data (white) held in these years for some countries.
 
@@
 Men earn around 10% less.
+Studies with data from other countries are broadly in agreement [@polonyi2022it].
 The point here is that you should not be doing a PhD in Biological Sciences because you think you will earn more money in your following career.
-Having made this point, I also feel that it is important to point out that how much you earn may not be directly related to job satisfaction or lifetime happiness.
-To my knowledge, this is not (yet) reported, but could be a lot higher for those with PhDs, even though they earn less.
+Similarly, another qualitative study of the perception of PhD holders in the UK suggests that while doing a PhD provides a lot of personal value for the student, it lacks any monetary value from the career that follows: i.e.
+their PhD Worth doing but not worth having [@guccione2022worth].
+Another way of thinking about this is that it might be more important to you to have job satisfaction or lifetime happiness over a well paid job.
+To my knowledge, comparative levels of happiness is not (yet) reported, but could be a lot higher for those with PhDs, even though they may earn less.
 
@@
 But the reality is that you will never do a PhD for somebody else, you are always going to be doing it for yourself.
-If the only reason you are doing a PhD is because someone else told you that you should then you are likely to have a really hard time.
+If the only reason you are doing a PhD is because someone else told you that you should, then you are likely to have a really hard time.
 By the time you graduate, you may have a PhD but without any idea of what you're going to do with it.
@@
 You’ll also have some experience and connections to go back to once you’ve finished.
+
+If you are planning a career outside academia, it is worth contacting potential employers and finding out how interested they are in employing people with a PhD in the subject area you are considering.
+Similarly, if you are considering a career inside academia, it is worth talking to people who are already in that job market (probably post-docs in your own group) about how easy or difficult they are finding it.
+Whatever you are planning to do, the right time for planning is before you begin your PhD.
+There is more written about planning your future [here](#Plan).
 
@@
 Doing a PhD will be the cause of stress not only in your working life, but will also impact other areas of your life.
-Being aware of this at the outset will allow you to alert close members of your non-working life (think partner, family and friends) that they may well need to act as a support network.
+Being aware of this at the outset will allow you to alert close members of your non-working life (think partner, family and friends) that they may well need to act as a [support network](#support).
 It is worth tracking your mental health during your PhD to check that you are not getting into difficulties, and I show you later [how to do this](#healthy).
@@
 Again no matter how high profile these people are, if they can't advise their own students they are best avoided.
-Chamberlain [-@chamberlain2016ten] suggests that there are 10 types of academic advisors, implying that 9 of these are rather disfunctional.
+Chamberlain [-@chamberlain2016ten] suggests that there are 10 types of academic advisors, implying that 9 of these are rather dysfunctional.
 
@@
 Especially if you have a limited amount of time once you have registered in order to complete your PhD studies.
+
 Some projects are much cheaper than others.
@@
 If you haven't already got the message (see above) then these are important decisions and you may be better off postponing the start of your PhD in order to raise money, and do the project you want, in the place you want, and with the person you want to advise you.
+
+## Core skills {#coreskills}
+
+Although this book is about writing your PhD in Biological Sciences, there are a number of core skills that you, or anyone doing a PhD will need.
+This section of the book aims to give you a foundation on these core skills and encourage you to use them from the very start of your PhD studies.
+You may well find that you already have one or more of these core skills to a certain degree, but I would encourage you to read the chapters in this section and reflect on your proficiency of these core skills and build them up where you need to.
+
+ - [Time management](#time)
+ - [Planning](#plan)
+ - [Communication](#email)
+ - [Upskilling](#useR)
+ - [Mental health](#healthy)
+ - [Advisor relations](#advisor)
+ - [Critical Reading](#critical)
+ - [Regular writing](#regularwriting)
+ - [A life scientific](#lifescientific)
+
+Improving Each of these core skills will make doing your PhD easier.
+These core skills will extend beyond your PhD into your working life beyond such that you can consider them to be life-skills.
+None of them are independent, instead they are all integrated with each other and require careful thought and reflection from you about how you can improve each of them.

### 2024-09-11 1.02-time-management.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

# Time management {#time}

How much time during the week do you spend doing nothing? Do your friends and colleagues ever find you sitting staring into space? My guess is that because you are starting a PhD, you are used to some level of time management and you can fill all of your waking hours every day of the week without the need for sitting staring into space because you have nothing to do.

We all have exactly the same amount of time in a day or a week. People who accomplish more in that time are doing so because they manage that time efficiently. You may well know such people and be amazed that they not only use their time at work wisely, but they also have time for sports, socialising and other hobbies. This is because once you have learned the skills of time management, including focussing on tasks that you have been assigned to do when you choose to do them, you'll find that you can accomplish much more. Research has shown that time management is empowering having a positive relationship with job satisfaction and work-time balance and negatively with stress [@claessens2007review]. Moreover, postgraduate students who use time management techniques were found to have improved [mental health](#healthy) [@liu2023postgraduates].

This chapter is meant to encourage you to manage your own time and help you start planning so that you can conduct all of the work needed for your PhD, and also find time for your life's needs during your PhD studies.

## Start planning your time today

At the start of your PhD, you are likely to have to attend classes, and so you will already have a fairly full time-table. This makes the task of planning your time slightly easier as you already know when you will be busy attending class. This period will give you a limited amount of free time (i.e., time that you have to manage), and you can start using this wisely to schedule both critical reading and academic writing tasks there.

### Planning the total duration of your PhD studies {#whole-period}

Once you know the subject of your PhD and what questions you need to answer with what experiments and observations, you will have to plan your remaining years of study to make sure that you can get all of these experiments done in the time allotted. Figure \@ref(fig:long-term-planning)) gives you a visual idea of what this plan might look like. In addition to showing the time when you will be busy and on what tasks, it also includes goals and conferences. Of course, your own planning chart might look much more complicated and have many more rows of tasks to achieve. For example, you might want to split the time spent for each chapter by time spent acquiring data from that spent analysing and writing about it as separate rows. My suggestion would be to keep this document dynamic so that you can add to it and adjust it as time goes by. Doing so will ensure that your planning is flexible and can take into account the unexpected.

(ref:long-term-planning) **Knowing when you are going to do the different work that makes up your PhD is critical**. In this Gandt figure, I have broken the different tasks that are needed to complete a PhD into different chapters (rows) which include four data Chapters (C1-4) and a technique chapter (T1). For planning purposes over three years, I have split each year into quarters (Q1-4) shown as columns. In this example, collecting the data for the four data chapters is seasonally sensitive (data can only be collected during Q2), meaning that these are immediately prioritised for this work. The technique chapter involves intensive lab time, and so can be completed during Q3 and Q4. Also included are the goals for completing chapters (shown as yellow stars) and important conferences where the results of chapters T1 and C1 are presented. Also indicated is time to write the introduction (Intro) and concluding chapters (Conc).

```{r long-term-planning, fig.show='hold', out.width = '85%', echo=FALSE, fig.align='centre', fig.cap="(ref:long-term-planning)"}
knitr::include_graphics('figures/long-term-planning.png')
```

However you choose to plan the period of your PhD, you will need a plan for the entire duration of studies that ensures that you have time to collect all of the data you need, analyse and write it up in the time you have available.

### Day-to-day planning {#day-plan}

Calendars are extremely useful for planning, and you should find that your phone or email app already has a built-in calendar that you can easily use. Web or cloud-based calendars are particularly good as you can access them from different devices and choose how they send you alerts and notifications for events. You probably already have one that you use, so I'd encourage you to keep using that same software and build on the framework you already have for your day-to-day planning.

When it comes to day-to-day planning it is really useful to know your own rhythms and abilities. Although everyone is different, everybody will have times in the day when they are more or less productive. I find that I can concentrate most in the early mornings while drinking a cup of coffee. Other friends do their best work late at night. Many of my colleagues sleep after lunch as their bodies concentrate on the serious task of digesting their large lunches. Becoming aware of your own best (and worst) times for concentrating is an important pre-requisite to your day-to-day planning as you will be able to set your best times for the work that requires the most concentration, like [critical reading](#critical) and [scientific writing](#part3).

If you are new to time management, then what follows is a suggestion to get you started. However, there are many many different approaches and you should find what works for you.

1. Add all current commitments (lab meetings, classes, etc.) to your calendar.
1. Divide your day into 40 minute blocks. These are the blocks of time that you have to allocate to doing different tasks or assignments. The chances are that you already have commitments (such as classes), so you may have an existing framework of start and end times that fit logically for you. Of course, the time blocks don't have to be 40 minutes - they can be longer or shorter. The aim is to find a period through which you can concentrate from beginning to end on a single task. If you can only manage 20 minutes at the beginning of your PhD, you will need to be extra strict with your time management so that throughout your studies you can slowly increase your concentration abilities up to 40 minutes and even beyond.
1. Make a list of things that you find rewarding. Rank these from top to bottom with your favourite at the top. Examples might be playing a game on your phone, catching up with your instant messaging, drinking tea, going for a walk, having a snack or playing table-tennis. These rewarding activities will need to be achieved in 5 to 10 minutes, so don't include things like baking a cake which would take much longer, which would require you to return home via the supermarket etc. You also need to consider that these activities should not rely on other people too much as their schedules may not match your own. 
1. Dedicate the 40 minute blocks to different tasks in your day. Focus on having regular tasks for skills such as [critical reading](#critical) and [scientific writing](#part3). In the 5-10 minute gaps between the blocks allocate the best rewards (highest ranked) for the tasks that you like least. 
1. Plan for at least two weeks in advance, but also keep an eye on the [planning of your total study duration](#whole-period) so that you know when you need to go to the field and accomplish your goals.

The idea of the above planning with tasks and rewards is to use motivation (Figure \@ref(fig:motivation)) to drive you to do the tasks that you like the least, and eventually to make them your favourite tasks. This is a really very simple but powerful way in which you can train yourself to manage your time, become more productive and start enjoying your least liked tasks the most.

(ref:motivation) **Using rewards after doing tasks is a very easy way to manage your time enjoyably**. Humans respond very well to rewards, and you can use this system to help yourself do your least liked tasks by rewarding yourself with your most liked activities.

```{r motivation, fig.show='hold', out.width = '45%', echo=FALSE, fig.align='centre', fig.cap="(ref:motivation)"}
knitr::include_graphics('figures/motivation.png')
```

My suggestion is to use colour coding in your calendar so that you can quickly see what you are doing next. You could use colours to code for the places where you will conduct each task block. For example, you could code your desk blue, the lab yellow and meetings red. This will allow a quick glance to let you know where you need to go between tasks. It will also determine the kind of environment that you need. Try to keep your desk environment for tasks that require concentration, and keep your reward environment separate. Associating certain spaces with different tasks and rewards will really help you quickly transition between roles when the time comes.

#### Don't reward failure to do activities

The idea of using chunks of time to do certain tasks and then rewarding them will only work if you actually do the tasks that you allocate to yourself. If you sit down but fill all of your time with [displacement activities](#displacement) then **do not** reward yourself afterwards! Make sure that you keep the rewards for positive reinforcement only. If you start rewarding yourself for not working, then you'll actually be training yourself not to work. Obviously, this is not what you want but the discipline of working and then rewarding yourself is entirely in your own hands.

At the same time, don't be too hard on yourself. If you filled the time with the task given, you can reward yourself with your intended reward - even if you did not finish the task. The important point is to fill the time with the task at hand. There will be more time in your week for coming back to unfinished tasks.

### Remember to include time to be healthy

When planning your day, you do not have to include details like sleeping, eating, sports and time with your family and [support network](#support). However, as you get busier it is a good idea to make sure that you do have time to get adequate sleep, food and exercise time. There is research that shows that if you fail to get enough sleep or nutrition, post-graduate students cannot maintain their concentration on normal tasks, eventually leading to burnout [@renu2024time]. Hence, when you are especially busy, time management becomes a critical skill in order to ensure that you remain [healthy](#healthy).

### Thinking time {#thinking}

As outlined elsewhere in the core-skills, much of what you do during the day will need time to be processed in your brain. This could be overnight, during REM sleep, or it could be during discussions with friends and colleagues. This [thinking](#time-to-think) or [reflective](#reflection) time is important. You may want to include topics to think about or talk about in your day-to-day activities. For example, a workout at the gym might become a session to think about the most recent paper you read in your [critical reading](#critical) session, or a walk with a friend might be an opportunity to chat with them about some difficult problem you encountered during a [writing session](#regularwriting). Remember to balance such work times with times when you give yourself a break from thinking about work (actually avoid thinking about it). These times are also important to give your brain a rest from the intense thinking that is involved in reflection.

Over time, you will learn which activities are best for [thinking](#time-to-think) and [reflection](#reflection). Once you know what they are, be sure to use them to the best possible effect. During your week, you should plan to have at least three of these periods when you have time set aside to think and a topic on which you are reflecting.

### Beware of displacement activities {#displacement}

These are things that you do that take up your time when you should be focusing on the work that you planned to do.

Common Examples:

 - arranging your working space and equipment
 - social media
 - emails and correspondence 
 - direct messaging apps
 - chatting with friends
 - watching TV or listening to the radio

The danger with displacement activities is that they can become habit forming. Remember that social media sites are designed to take your time (for the purpose of you watching their advertisements). They are well-known to be addictive, and if you are already hooked the best way to manage them is to place your time on them strictly within bounds.

Learn to recognize displacement activities - once you know what they are either avoid them (if they are not productive) or carve out special time for them (if they are necessary).

### To-do lists

Humans find it very satisfying to cross out tasks on a list once they are completed. Having a to-do list alongside your calendar might really help motivate you to a task based system of time management. You can write a list out for your week, adding items as they come up and crossing them out as you complete them. Then carry the uncrossed out items on the list over to the next week.

I encourage my students to have these lists in their meeting books so that we can check through them when we have student-advisor meetings.

To-do lists don't work for everyone, so this is something that you might want to try and see whether it helps motivate you.

## Multitasking (is a myth)

First, I am going to emphasise that the old idea of multitasking (trying to proficiently conduct several tasks at the same time) is a myth. Secondly, I am going to encourage you to multitask when the tasks that you are trying to accomplish are non-critical and do not require all of your attention.

Investigations have consistently shown that those who are trying to split their attention between two or more tasks do so at a cost to the tasks themselves [@gilbert1993you; @wang2012myth]. For example, a study of students who were instant messaging while reading showed that this multitasking increased their reading time by more than 20% [@bowman2010can]. Similarly, trying to watch television while conducting academic work resulted in poor comprehension and memory recall [@armstrong1991background]. The number of studies that show that humans are not capable of conducting several tasks proficiently at the same time is now overwhelming.

The above information should be of concern to us when we are conducting critical procedures that require all of our attention, such as driving a car, piloting a plane or conducting an operation. However, there will be tasks that you need to do that do not require all of your attention. Some aspects of your PhD might be incredibly boring and repetitive so you can do them with your eyes shut!

Your PhD will require a lot of activities that require 100% of your concentration. They will also have activities that require less of your attention. Know what these relevant tasks are and adjust your time accordingly throughout the day. High-concentration activities should be done when you are most alert. See below to see how to recognise this.

Low-concentration activities provide you with opportunities to conduct additional tasks. Note that although you can do different things at the same time, you won't actually be multitasking, because your concentration will be flipping from one task to another constantly during this period. That's why I'd advise only doing low concentration tasks together when your brain can cope with the constant switching.

Some examples might help. If you work in a lab where you need to fill pipette boxes or conduct a lab clean every week, schedule these tasks alongside a colleague and use the time to catch-up on their research and tell them how yours is going. This would be a more productive use of your time than listening to music. If you don't have an appropriate colleague there, then consider downloading a podcast from your subject area, or an audiobook on something relevant. Be sure that you have note taking equipment close-by so that you can quickly jot down ideas as you are listening.

### Priority activities

I cannot list all of the priority activities that you will need to perform during your PhD, instead I will list some general ones that everybody is likely to have. When you schedule time for these priority activities, you will not only need to put away things that might distract you from your task (such as a phone), but you should also try to clear your mind of worries and other aspects of your life that might impinge on this time. If there's something happening in your life that you really can't pull your mind away from, then it would be worth rescheduling any priority activities that you might have for that day. If such worries stretch for weeks or even months, then this is something that you should alert your advisor to.

#### Academic writing

Perhaps it is no surprise to find out that I consider writing a priority activity from which you should not be distracted (or attempt multitasking). For this reason, it is important to prioritise time for writing in your weekly schedule. Throughout this book I emphasise that academic writing is a skill that you need to practice, meaning that you can't wait until the end of your PhD to start writing your thesis. Similarly, at the beginning of your PhD you should not expect to sit down and write perfectly. It takes time and practice, and this is why writing regularly is the best way to improve your writing skills.

Note here that I am talking about academic writing. I am not talking about writing an email or sending IMs to your friends. Academic writing, the subject of this book, is something quite different and no amount of writing IMs or emails will help you.

#### Critical reading

Like writing, [critical reading](#critical) requires all of your attention and is a skill that you will need to learn (see [Part II](#critical)). Critical reading is going to help the most with your writing skills, without actually writing. Of course, critical reading does involve writing, and I encourage you to read the chapter on this topic later in the book to find out more.

#### Meetings with your advisor

Your regular [meetings with your advisor](#advisor-meet) should be times when 100% of your attention is focussed on getting all of the information and advice that you need in the shortest time possible (see [Part I](#advisor-meet)).

#### Running equipment and machinery

Much of the equipment that you will use during your PhD will be very costly very likely accompanied with expensive consumables. Mistakes will happen, but if you break equipment or ruin consumables because you are distracted, then I don't consider this a mistake, but your own folly. You now know that some tasks require all of your attention, and this will include many tasks involving equipment and expensive consumables. Give these the full attention that they deserve.

### Non-priority activities

Having put together your timetable (above), you will already know what your non-priority activities are, because you had to find time to fit them in. My suggestion is to use a different colour for these tasks. They are the tasks that will drop away when you get especially busy, or have another additional task that you didn't foresee.

Non-priority tasks are often important, but don't need to be done immediately. Even if you post-pone tasks like these, it is a good idea to keep an eye on them to make sure that they don't get dropped entirely.

These non-priority tasks will allow you to keep your calendar flexible so that you can accommodate additional or unforeseen tasks. Remember to make sure that your advisor agrees with you about what is a non-priority task. For example, if you consider lab meetings as non-priority and stop turning up, your advisor might not agree.

## Time management hacks

You will find lots of advice out there on the internet about how to hack your time management and improve your efficiency. These ideas are numerous, and some are based on research that shows that breaking up intensive concentration time into short sessions followed by rewards is very efficient.

My suggestion is to give such hacks a go when you are not too busy. If they work for you, then that's great. However, don't rely on a hack that you have never used before during an especially busy time. During busy times you should be using tried and tested methods of time management that you already know work for you.

### The Eisenhower Matrix

Using this method, you divide every task into one of four different quadrants as shown in Figure \@ref(fig:Eisenhower-Matrix)). Once you have decided where a task sits, you can decide how to handle it.

(ref:Eisenhower-Matrix) **The Eisenhower Matrix helps you decide how important the different tasks are, and then how to deal with them**. Prioritising tasks will be difficult at first, but as you get busier it will become essential for you to unsure that you accomplish all of the work to complete your PhD.

```{r Eisenhower-Matrix, fig.show='hold', out.width = '65%', echo=FALSE, fig.align='centre', fig.cap="(ref:Eisenhower-Matrix)"}
knitr::include_graphics('figures/Eisenhower-Matrix.png')
```

- **Urgent and Important**: Do this right away as it is likely to impact on other goals.
- **Important but Not Urgent**: Schedule such tasks into your calendar.
- **Urgent but Not Important**: Depending on how busy your calendar is, you might want to consider asking someone else to help with these tasks.
- **Neither Urgent nor Important:** You won't have time to do these tasks when you are busy, so keep them in a list for when times are less intense.

### The Pomodoro Technique

This is a time management method that breaks work into focused 25 minute intervals, called "Pomodoros". These are then followed by short five minute breaks. Once you have completed four Pomodoros, you can take a longer break of 15-30 minutes. The technique is said to help improve focus, reduce distractions, and prevent burnout by encouraging regular breaks while maintaining a structured approach to work. It is particularly effective for managing tasks that require sustained attention like thesis writing.

This might be especially useful if there are a group of your who all time your Pomodoros together. You get the reward of sharing the breaks, plus the calm during those concentrated efforts.

### SMART goals

You are likely to find SMART goals everywhere, as they are a very popular way ensuring that goals are met, including time management. Originating in 1981, SMART goals were first conceived as a framework for setting clear and achievable goals in management contexts [@doran1981theres]. They are now used practically everywhere and are well worth knowing about:

 - **Specific**: Clearly define the goal with a detailed and focused objective.
 - **Measurable**: Ensure the goal can be quantified or tracked with specific metrics.
 - **Achievable**: Set a realistic and attainable goal given available resources and constraints.
 - **Relevant**: Align the goal with broader objectives and ensure it is meaningful and valuable.
 - **Time-bound**: Set a clear deadline or time-frame to accomplish the goal.

I will not go into SMART goals in detail here, but there is plenty to read including books by Stevens [-@stevens2013turn] and Hyatt [-@hyatt2023your].

### Batch tasking

You may find that your to-do list contains a number of similar tasks that would be better if performed together. For example, you may need to make five different telephone calls in relation to different parts of your studies. By batching these calls together into one of your 40 minute task slots it will ensure that you get them all done without eating into other productive time.

### Saying no

Be aware that while your time management skills might be champion, you will be impacted by those around you. This means that you might need to explain to others that you already have priorities and cannot always give them the time they are requesting. These could be commitments to people at work or people in your social and [support network](#support). Being able to prioritise these commitments is also important as you need to remember that you do rely on your [support network](#support).

As you get deeper into your PhD studies, you will get busier and your time will get tighter. You should know when these periods are coming up (from the [planning of your total study duration](#whole-period)), and this will allow you to get ready for them. As you become better known in your institution, you'll also find that people start asking you to do extra tasks. These could be favours for friends, or special assignments for your advisor. When you are asked to do these extra tasks you should take the time to look at your calendar and ask yourself where that extra time is going to come from. If you are in your busiest time and the deadline is before that ends, then you might have to say no. However, if your calendar is more open you can replace some of your activities with the task so that you can achieve it.

You do need to be clear to yourself and the person you are doing the task for exactly what the task is, when is the deadline and how much time it will take you, and what you won't be doing during that period in order to do the extra task. Many such tasks are worthwhile, and it may well be in your interest to find the time to do them. Your calendar should be helping you to commit time, not making your time allocation so rigid that you have to say no to everything. If you do need to say no, give an explanation about why you have to say no this time and suggest when you might be able to say yes in the future. If the task is time-limited and won't recur in the future, then be sure to be thankful for the opportunity and express your regret that you can't take part this time, but look forward to future opportunities. You do not want to close doors by saying no, you need to make it clear that there will be times in the future when you would welcome another opportunity to participate and that you want to say yes.

### Delegation

During your busiest times it may be that you will not be able to complete all of the tasks that you have listed. For example, you might need to attend a conference but your experimental plants still need to be watered and tended to. Use the [planning of your total study duration](#whole-period) to identify when these periods are likely to be, and alert your advisor well ahead of time (2 - 3 months). You should be able to come up with a solution so that you can go to the conference (or other task) while someone else looks after your experiment. However, that person may well need training, hence you will need to flag this and organise for cover well in advance.

## Don't stop managing your time

Once you have learned how to manage your time, you should find that this is a useful life-skill that you can take with you after you complete your PhD. Most importantly, it will empower you to have free time that you can devote to yourself and your family and friends. The [next chapter](#plan) takes these planning and managing ideas to planning ahead for your future.

### 2024-09-11 1.04-email.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

--- 2023-03-30 1.03-email.Rmd
@@
-# Communicating by email {#email}
+# Communicating by email and messaging apps {#email}
 
 Throughout your PhD, you will be expected to communicate with your advisor and other professionals inside and outside of academia using email.
-The chances are that you have already written a great deal of emails in the past.
-However, this does not mean that the style and content of those emails will be appropriate when communicating during your postgraduate studies.
+In some contexts, you may also be expected to use social media apps like WeChat and WhatsApp to communicate either directly or in groups.
+The chances are that you have already written a great deal of emails and instant messages in the past.
+However, this does not mean that the style and content of those emails and app posts will be appropriate when communicating during your postgraduate studies.
+
+In this chapter I will start by considering some issues around correct use of email, and then move onto direct messaging apps [below](#wechat).
+
+## Email
 
@@
 
-## The subject
+### The subject
 Do include a succinct subject for your email.
@@
 
-## Writing style
+### Writing style
 In general, it’s best to keep your style brief.
@@
  - read through what you've written before you press send
+ - try not to be overly familiar or informal [@lewin-jones2014understanding]
+ - be polite, as this may make a difference to you getting a positive response [@bolkan2012you]
  - let brevity be your watchword
 
-### Don't use slang or SMS language
+#### Don't use slang or SMS language
 I remember an email that a colleague received (many years ago) asking for a bursary. I don't remember the exact wording, but it went something like this:
@@
 
-### Formalities {#formal}
+#### Formalities {#formal}
 The level of formality that you use will largely depend on your existing relationship with your correspondent. The formal tone is set immediately with the greeting that you use.
@@
 
-## Why exactly are you communicating?
+### Why exactly are you communicating?
 The first two or three sentences of your email should state why you are communicating and whether or not you are expecting a reply.
@@
 
-### Using Structure
+#### Using Structure
 I like it when people use some structure in their email.
@@
 
-#### Replying inline
+##### Replying inline
 This simply means that the previous email is reproduced when you hit reply, often with some marker for what was written originally.
@@
 
-#### HTML emails or plain text
+##### HTML emails or plain text
 Most email software now defaults to emails written in HTML, which means that you can use **bold**, *italics*, bullet points and most other features that you would in a word processing package.
@@
 
-#### Email chain
+##### Email chain
 Most email software has an option to include the original message in the reply.
@@
 
-#### Bullet pointed or numbered lists
+##### Bullet pointed or numbered lists
 
@@
 
-### Use of all Capital Letters - DON’T DO THIS!!!!
+#### Use of all Capital Letters - DON’T DO THIS!!!!
 
@@
 
-## Signing off {#signoff}
+### Signing off {#signoff}
 
@@
 
-## Who else should your email be sent to?
+### Who else should your email be sent to?
 
@@
 
-## When to CC someone
+### When to CC someone
 
@@
 
-### Use of BCC
+#### Use of BCC
 
@@
 
-## Reply or Reply All? {#replyall}
+### Reply or Reply All? {#replyall}
 
@@
 
-### High Importance
+### Institutional or personal email addresses
+
+As a student at your institution, you will likely be allocated an institutional email address.
+You probably also have your own private email addresses (e.g.
+Gmail, Hotmail, Yahoo, etc.).
+Some institutions have great problems with their mail servers, and many academics will abandon their use of their institutional email address because it causes too much trouble for them.
+
+It is worth asking your advisor which address they would like you to use (if you know that they have several).
+However, I would suggest that you make use of the institutional address for all formal communication so that you have a record of [due diligence](#deliverables) should you need it.
+
+#### High Importance
 Some email software has the ability to flag an email as important.
@@
 
-### Sending emails outside office hours
+#### Urgent emails {#urgent}
+Sometimes it is necessary to communicate urgently.
+Deadlines are unforgiving and your advisor and other people are not always aware of when they are or what they mean to you.
+Deadlines are also very stressful, and you might be more prone to miscommunicating or making mistakes.
+
+If you need to communicate urgently by email, then here are some tips about how to get a swift response:
+
+1.
+**Urgent for whom?** I think it's worth starting by asking yourself why exactly this message is urgent, and for whom.
+It could be that this is only urgent for you and your deadline, or it could be that this is a shared urgency among you and the people you are corresponding with.
+Alternatively, the urgency might not be for you and only for the other person/people.
+Starting out knowing this information is important as it will help you set the tone for the message.
+2.
+**State that it is urgent in the subject**: Start the subject with "Urgent" and give a timeline by when you need an answer in the subject line.
+Most people will receive their emails on their phones and at some point they will see the subject line without reading the contents.
+If the subject line can inform them of what the topic is and when the urgency is for (e.g., Urgent: signature needed in the next 2 hours, the next 2 days, the next 2 weeks, etc.), this will really help them knowing exactly what is needed.
+3.
+**Within the message**, provide the explanation again together with a clear and concise account of exactly what needs to be done.
+If you need a signature, then remember the attachment that needs to be signed.
+The more concise you can be about what is needed, the faster it is likely to be done.
+4.
+**Follow up with an instant message or text**.
+If this is urgent and especially if it is outside office hours, then follow up with an instant message to check that they have received your message.
+See [section on texting below](#WeChat).
+
+Before you send the email, take a few moments to read it through and check that everything is correct.
+If it really is urgent, then the last thing you need is to have some to and fro over the actual meaning of what is written.
+
+Some deadlines are really important, and some correspondents are famously late with responses.
+In this case, consider providing a deadline one or two days earlier than the actual deadline.
+This only works when you have a couple of weeks to go, but it might help you build in more time and reduce the urgency.
+For example, you can use the [Send Later](#later) feature to send a couple of reminders building up the the date.
+But remember to cancel them if you get what you need before!
+
+#### Sending emails outside office hours
 Personally, I have no problem with people sending emails whenever they want, as long as they don't expect replies to come out of office hours.
@@
 
-### Send Later {#later}
+#### Send Later {#later}
 
@@
 
-Another scenario that I use for send later is when I have a deadline, but am awaiting feedback.
+Another scenario that I use for send later is [when I have a deadline](#urgent), but am awaiting feedback.
 I can make sure that I meet the deadline by sending the email with report or manuscript attachment or whatever it is that I have to send before the deadline.
@@
 
-### Forwarding emails
+#### Forwarding emails {#fwd}
 
@@
 
-## Out of Office autoreply
+### Out of Office autoreply
 I find the autoreply saying that someone is away very useful.
@@
 
-## Using email signatures and banners
+### Using email signatures and banners
 
@@
 
-## Using emails as a record of due diligence {#deliverables}
+### Using emails as a record of due diligence {#deliverables}
 
@@
 
-**If you are communicating with your advisor on a web based platform (like Slack), then do you need to send them emails as well?**
+**If you are communicating with your advisor on a web based platform (like Slack, WhatsAPp or WeChat), then do you need to send them emails as well?**
 
@@
 
-## Keeping emails
+### Keeping emails {#keepemails}
 
@@
 
-### Folders
+### Institutional Policies
+
+Your institution is very likely to have a policy on email use.
+If you are unsure and there is no summary available to you from your department, you can ask for a copy of the policy to see what is officially allowed and what is prohibited.
+
+#### Folders
 
@@
 I find that after 2 years, most correspondence is no longer useful. There is the odd occasion when I delve further back in time, but older institutional emails get archived after 2 years.
+
+## Use of WeChat or WhatsApp during your PhD studies {#wechat}
+
+[WeChat](https://wechat.com/) and [WhatsApp](https://whatsapp.com/) are two social media direct messaging apps that currently dominate the markets in Southeast Asia and the rest of the world, respectively.
+In terms of their potential to communicate during your PhD, they have some similarities which I will outline first, and some differences which I will treat later (see [here](#diffs)).
+Other platforms, such as [Slack](https://slack.com/) might be used in particular labs.
+These can be highly efficient ways of communicating with groups.
+
+Just like [emails](#keepemails), these social messaging apps can be kept and archived.
+This means that you can use them as a means of documenting [due diligence](#deliverables).
+Make sure that you have your advisor's consent to do this.
+They, or your institutional policies, may insist that you inform them in an email.
+
+It is important to bear in mind that these social media apps can have an adverse effect on your mental health [@montag2018multipurpose], and even become addictive [@hou2017excessive].
+If you think that you need help because you are spending too much time checking your social media apps, then please get in touch with your university's psotgraduate student support.
+
+### Direct messaging
+
+It may be appropriate and acceptable to use a messaging app to communicate directly with your advisor.
+For example, if you have an [urgent request](#urgent).
+Many of the points discussed [above](#email) also apply here, and some are even more relevant:
+
+ - Although a message may be received instantly, this does not mean that your advisor (or any other academic) may have the time to respond to you immediately.
+You need to give people time to respond.
+In particular this is especially important in evenings, weekends and holiday periods.
+ - As there is an expectation that direct messaging receives a quick answer, please try to avoid sending direct messages outside of office hours - unless this has been specifically sanctioned
+ - If you are new at communicating with your advisor, ask them whether they prefer to use direct messaging over email. 
+ - Having someone's direct messaging details means that you can call them and speak to them directly on their mobile device.
+Ask permission in the chat before you do this.
+If you are communicating with your advisor, ask them specifically about whether they are happy with this type of communication - don't just assume it's ok.
+ - Rather like email, it is possible to [forward content](#fwd) from WeChat and WhatsApp to other uses.
+Be sure that this is appropriate and sanctioned before doing so.
+Alternatively, assume that everything that is shared with you is confidential unless you are given specific permission to share.
+ - Do remember to be polite and thankful when people help you through direct messages.
+
+### Group messaging
+
+In general, I would advise to learn the unwritten rules of the group through observation over time. Here are some issues that I have come across with student messaging groups.
+
+Many students love to use gifs to display their emotions.
+I would suggest that this is inappropriate in most instances for two reasons: first they take up a lot of space and demand that anyone trying to catch up needs to scroll through a lot more content; second they are designed to be distracting and that is not the aim of communicating in these groups.
+
+Keep your messages on topic.
+Because every message that you (or anyone else) sends on this group platform will send a notification to all subscribers, it is important that you keep any messages that you send relevant to everyone in the group.
+Avoid sending messages that are about social events (e.g.
+party invites) unless this has already been sanctioned and (importantly) everyone is invited.
+
+#### In your postgraduate classes
+
+Joining a group for your class may be mandatory in some institutions, or at the very least the only way in which you will learn about tasks and deadlines for postgraduate classes.
+Some postgraduate classes are experimenting with platforms that are only mediated through groups, such as Problem-Based Learning [@luo2021wechat].
+In these instances you will have to use the app.
+
+Before you post a question on the group, make sure that someone hasn't just already asked the same question. Use the search functions and see if there's anything that you have missed.
+
+In your classes, it would be inappropriate to add someone who isn't in the class to the group. This might be seen as being nosey or snooping on class discussions.
+
+If you are in doubt about what can and cannot be done in a class group, consult your lecturer or ask your class rep to consult them.
+
+#### In your lab groups
+
+Your lab group may also have a group messaging.
+Generally, these are extremely efficient ways to disseminate information to a group.
+Be aware of the general issues with these platforms and make sure that if you want to try to do something new on your lab group, have the blessing of your advisor or senior lab member before posting.
+
+If your group message is intended for one person in particular, then rather don't bring it to the entire group.
+In the same way, if someone has what you believe to be inappropriate behaviour in the lab try to take this problem to the lab PI or bring it up in a lab meeting (as appropriate) rather than posting about it in the group.
+Calling out bad behaviour in a chat group is sometimes counter-productive and can be upsetting for many lab members.
+
+### Desktop versions
+
+Both WeChat and WhatsApp have desktop versions of their platforms, and these may replace the use of email in some places.
+For example, in China WeChat has almost entirely replaced email in many institutions such that sending an [email](#email) may elicit no response as the inbox is simply not checked.
+Note that this is not the case in Europe, North America and elsewhere.
+Institutions may well specifically forbid the official use of social media apps for official communications, so be aware of the policies at your institution.
+
+### Turning off notifications
+
+It is understandable that you (and others) in a group chat will want to turn off notifications if their mobile devices are pinging too often.
+If you do turn off your notifications for important class or lab groups, make sure that you look through the content at least once a day (excepting holidays).
+
+### Differences {#diffs}
+There are some important differences between WeChat and WhatsApp. At the time of writing, these differences are diminishing.
+
+#### WeChat
+
+WeChat has become a platform that offers far more than messaging and calls [@montag2018multipurpose].
+You can do a lot more on it, as it effectively works as a browser.
+Thus within your WeChat postgraduate class group, there may be scope to include parts of the learning onto the app.
+All of this makes it more likely that WeChat will be used as a part of your learning experience.
+Indeed, there is even evidence that using WeChat responsibly for knowledge sharing at work promotes creativity [@zhang2021excessive].
+
+Another important difference is that WeChat is often linked to someone's bank card and so can be used for transferring cash.
+Although this seems unlikely to be used in your PhD situation, always take care when doing this and find out before hand whether WeChat evidence for payments can be used for accounting purposes in your institution.
+
+#### WhatsApp
+
+At the moment, WhatsApp does not have the extended uses of the WeChat platform.
+
+Having someone's WhatsApp contact details means that you will have their mobile phone number (or at least the one used for their account which may not be the SIM in their phone).
+Use this information with respect and don't share it with third parties unless you have consent.
+
+## Other Social Media platforms
+
+During your PhD studies you may make use of other social media platforms to communicate such as [Atlmetrics](https://howtopublishscience.org/altmetrics.html) [e.g., see @measey2022how].
+As ever, use social media with caution and professionally.
+Respect the rights of you advisor and lab mates to have their own private social media, just as you may well have your own.

### 2024-09-11 1.05-useR.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

--- 2021-07-20 1.02-useR.Rmd
@@
-# Reconciling yourself to doing things you've been avoiding {#useR}
+# Upskilling {#useR}
 
@@
 And each machine had an odd white device attached called a mouse, and I had no idea what that did.
-All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC.
-This is practically where we are now with R (@rcoreteam2021r; Figure \@ref(fig:Rlogo)), and a wealth of other emerging and alternative programming platforms.
+All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a computer.
 
-(ref:Rlogo) **The ubiquitous R logo.** R (and the invaluable GUI RStudio) has become the go to platform for many statistics and figures in biological sciences
+It is obvious now that in order to conduct your work you are going to need skills in lots of tools.
+Some of these tools will be niche, but others will be universal and extend beyond your thesis into your world after graduating.
+The message of this chapter is that you may well find that there are skills that you need to learn during your PhD studies that you feel some initial aversion to.
+You may have a physical aversion to, for example, using a pipette in a lab or dissecting a euthanised animal.
+Or this could be some kind of skill such as programming in R or Python.
 
-```{r Rlogo, out.width = '33%', echo=FALSE, fig.align='centre', fig.cap="(ref:Rlogo)"}
- knitr::include_graphics("figures/Rlogo.png")
-```
+Reconciling yourself to learning new skills, even when you have an aversion, will be something that you come across time and again in life, and not just while you are doing a PhD.
+Because doing a PhD is a time-limited goal orientated process, you will need to quickly overcome your aversion and ulimtately this will help you in future life.
+
+## When to invest your time in learning a new skill {#upskilling}
+
+ Early adopters can either end up putting in a lot of effort and then finding that their effort is misplaced, or finding that they are ahead of the curve and that everyone else needs to catch up.
+
+Knowing whether or not to invest your time in learning a new skill is difficult.
+I suggest that you try to find out more about exactly how much time you might need to invest, and exactly how wide or narrow this new skill might be of use in the future.
+
+If your advisor tells you that you must learn this particular tool or skill, then you need to reconcile yourself to doing it.
+Remember that learning a new skill gives you more than that particular talent.
+Learning skills helps your brain become more flexible and adaptable to learning in the future [@chang2014reorganization].
+
+Whatever aversion you may have to learning this new skill, you will need to approach it with professionalism, and put aside any prejudice that you may have.
+Use your [time management](##Time-management) skills to carve a small piece out of every day to learning your new skill.
+This starts with being aware when in the day you learn best, and being aware of all the other commitments you have together with how this one will rank.
+
+### Example: Learning to write code in R
+In the early 2000s, R (@rcoreteam2024language) was relatively unknown, but a few pioneering individuals had spotted the advantage to using it in their own work (especially to help with repeatability).
+For those early adopters I am sure that their investment in learning R has paid off.
 

### 2024-09-11 1.06-healthy.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

--- 2021-07-20 1.04-healthy.Rmd
@@
 Anxiety and panic can be triggered.
-In this state, additional demands on your time may also push your life off balance, so that you start to neglect your personal wellbeing which can negatively impact on relationships, exercise regime, or even nutrition and personal hygiene.
-Some people can find that the additional stress can cause physical symptoms that may even need medical treatment.
+In this state, additional demands on your time may also push your life off balance, so that you start to neglect your personal well-being which can negatively impact on relationships, exercise regime, or even nutrition and personal hygiene.
+Some people can find that the additional stress can cause physical symptoms that may even need medical treatment [see @zhong2019depressive].
 Your sense of competence and mastery can be negatively impacted such that you may even suffer from feelings of inadequacy or imposter syndrome ([see part 2](#fear)).
+These types of pressure tend to be low at the start, and increase during your time as a postgraduate student [@cheng2020suicide].
 
 Although there are not many studies on mental health for PhD students, those that exist (as well as surveys: [Nature 2019](https://www.nature.com/articles/d41586-019-03459-7)) all suggest that there is a significant toll, which is proportionately higher than for others in society [@levecque2017work].
-Whatever your prior experience of stress in a working environment, academia is known to be particularly stressful, and as a PhD student, you are likely to absorb a significant amount of this stress into your own life [@stubb2011balancing].
+Whatever your prior experience of stress in a working environment, academia is known to be particularly stressful, and as a PhD student, you are likely to absorb a significant amount of this stress into your own life [@stubb2011balancing; @cheng2020suicide].
 
@@
 During the course of your PhD, if you feel that your scores may have changed, take the test again and compare them with your baseline scores.
-Although there are no hard rules, if three or more of your scores have moved by two or more points it could be worth discussing with your support network to help you decide whether or not to seek professional help.
+Although there are no hard rules, if three or more of your scores have moved by two or more points it could be worth discussing with your [support network](#support) to help you decide whether or not to seek professional help.
 
@@
 Additional stressors come from home and family situations.
-Your best means of coping will be to try and develop a support network and to understand where and with whom you can discuss any difficulties as they arise.
+Your best means of coping will be to try and develop a [support network](#support) and to understand where and with whom you can discuss any difficulties as they arise.
 Knowing who this is and how and when to approach them will put you in a stronger position if you need them in future.
+
+## Support network {#support}
+
+The people in your life make up your support network.
+Each person will have a different role, and not everyone in your support network will be able to support your mental wellbeing.
+For example, some of your colleagues at work might be in your support network for using equipment or doing statistics, but you may find that these sample people do not necessarily understand you as a person.
+People in your support network for your mental wellbeing are likely to consist of your close friends, family and those colleagues at work that you find a special connection with (i.e.
+new friends).
+
+You should identify your support network, and even place a tag next to their name in your contacts.
+Once you know who they are, you should also let them know that they are important to you, and that you may in future come to them to discuss issues around your mental wellbeing during your PhD studies in the future.
+It is important that they know that they have this role for you, and more often than not you can offer to fulfil this role for them in a reciprocal manor.
+
+During the course of your PhD, you should make sure that you take the time to check-in with your support network and let them know how things are going.
+This means sharing the good times as well as the bad times with them.
+If the only contact you make with them is when things get really bad, you are likely to find that they don't have a sufficient connection with you to help get you back on track.
+When you make contact with people in your support network, don't forget that they also require support from you.
+Inquire about how they are doing and be prepared to spend time sharing your respective highs and lows.
+Each time you invest time in your support network it will grow that relationship and deepen the level of support that you can hope to receive when you need it.
 
@@
 
-The positive relationship between the amount of physical activity and higher mental wellbeing is well established [e.g.
+The positive relationship between the amount of physical activity and higher mental well-being is well established [e.g.
 @grasdalsmoen2020physical; @gerber2014increased], but the kind of exercise required to achieve this improved result is varied.
@@
 
-There are other ways to achieve these benefits, including mindfulness or meditation.
+It is important to note that exercise is not about creating some perfect body image that you might see on social media.
+A useful tool to help think about the importance of exercise for your body came from from Stutz and Michels' [-@stutz2022tools] Life Force model in which they proposed that your life force is equivalent to a pyramid.
+The bottom of the pyramid represents your relationship with your body which is equivalent to 85% of the life force pyramid area (Figure \@ref(fig:life-force)).
+Working on this component consists of sleep, diet and exercise.
+The second part of this pyramid has your relationships with other people, both personal and in your work context.
+Lastly, your mental health (your relationship with yourself) sits atop these other components and is supported by them.
+Working on the lower components of the pyramid will support your mental health, but you can also work directly on your mental health through, for example, mindfulness and meditation.
+Stutz goes on to explain that writing is a way to enhance your relationship with yourself.
+This might explain why some people find writing their thesis a very positive experience, but also gives us another way in which we can think about writing and ourselves.
+
+(ref:life-force) **A model to understand the importance of your body in connection to your mental health.** In this model from Stutz and Michels [-@stutz2022tools] you see the body taking up 85% of the area of the pyramid, on which rests your relationships with other people and finally your mental health on top.
+Maintaining your mental health therefore rests on your body receiving enough sleep, a good diet and exercise.
+
+```{r life-force, echo=FALSE, out.width='50%', fig.cap="(ref:life-force)"}
+ knitr::include_graphics("figures/life-force.png")
+```
 
@@
 Now I have begun to realise that this represents some of my best thinking time.
-I use the time I spend trail running to turn ideas over in my head and especially to think through the logic in arguments, and potential flaws in experimental design.
+I use the time I spend trail running to turn ideas over in my head and especially to [think](#thinking) through the logic in arguments, and potential flaws in experimental design.
 This isn’t to say that I don’t need to put time in at the computer writing it all down, or talking through ideas with colleagues.
 But I find the meditative time I spend exercising to be especially productive.
-Don't take my word for it, see this great new book by Caroline Williams [-@williams2021move].
+Don't take my word for it, see this great book by Caroline Williams [-@williams2021move].
+In this book it is emphasised that exercise might not mean sport, but simply taking a walk or another aspect of self propelled movement.
+Actively seek out the activities that work best for you in order to [think](#thinking) or [reflect](#reflection).
+Once you know what they are, [plan](#time) to use them to optimise your thinking time.
 
@@
 If you don't have friends outside your academic life, then it would be a good idea to find some so that you can keep in contact with the 'real world'.
+
+Non-working time (e.g., holiday time) is also important in order to keep you productive during your PhD.
+You should not be working every day from beginning to end.
+Try to save at least one day every week work-free so that you can rest and spend time with friends and/or relatives.
+During busy times you might have to sacrifice this day, especially when you are working in the field or attending a conference.
+But you should make up this shortfall when times are less busy.
+You also need a real holiday when you take a break from your PhD work altogether at least twice every year.
 
@@
 
-Once you have a healthy body, you’ll find that there are all sorts of things that go along with this.
+Once you have a healthy body, I found that there are all sorts of things that go along with this.
 The desire to fill it with healthy food resulted, and this later tied in with my interest in the gut microbiome.
-I now have many fermented foods in my kitchen (much to the horror of my kids), and the improvement of my digestive system.
+I now have many fermented foods in my kitchen, and the consequent improvement of my digestive system.
 I could go on, but I’ll spare you.

### 2024-09-11 1.07-advisor.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

# Navigating the student-advisor relationship {#advisor}

It is widely acknowledged that a good relationship between a PhD student and their advisor is pivotal to the success of the student's research and to successfully obtain their PhD timeously [@gill2008student; @masek2017establishing; @abdullah2012relationships; @mainhard2009model]. Note that here and elsewhere I refer to this person as an advisor and not a supervisor ([see here](#advisor)). If this person were your supervisor, then it would be their project and you would be under their supervision. Instead, the reality is that [your PhD is your project](#Responsibilities) and they are [providing you with advice](#Take-advice).

This chapter aims to provide you with advice on navigating the student-advisor relationship. Over the course of your PhD, you will likely work closely with this person. You may well see them every day. I hope that you will share the [ups and downs](#celebrate) of your project together, and ultimately the success of getting your PhD, and [publishing your results](https://howtopublishscience.org/index.html). Because there is so much importance wrapped up in this relationship, you do need to approach it with some care and professionalism. Of particular note is the inherent [power imbalance](#power-imbalance) in your relationship, making it different from many other types of relationships. Hopefully, there will be no problems for you at all, but it is worth making provision for problems from the outset because problems between advisors and their PhD students are not uncommon. I hope that this chapter will provide you with enough information to start and continue your relationship with your advisor on the best terms possible.

In this chapter, I am going to assume that you have already chosen your advisor. Earlier, in [Part 1](#what-to-look-for-in-an-advisor), I explained what to look for in an advisor. Once you have started your PhD studies, it is unlikely that you will have the option to change advisors easily, so this chapter will focus on the dynamics of the relationship with this particular person.

## Expectations {#expectations}

One of the most important aspects of the student-advisor relationship is the expectations that each has about what the relationship should be like. For example, if you have already had an advisor for your MSc or undergraduate research project, then as a student, you likely have some experience of what this relationship can be like. However, it is important to remember that, just like elsewhere in life, the relationship you have with your new PhD advisor may not be like the relationship you had with your previous advisor. These expectations are particularly important at the outset of the relationship, as it is during this time that you will make the impressions that guide your advisor onwards. Similarly, your initial impression of your advisor will likely be important for your view of them.

From the outset, you need to understand what your expectations should be, and perhaps equally important what they should not be. These ideas are loosely based on several papers on this topic, each of which might be worth your reading if you are looking for different viewpoints and perspectives [@gill2008student; @masek2017establishing; @abdullah2012relationships; @mainhard2009model], and you can find them listed under responsibilities.

**Your advisor should:**

 - Support and advise you with your project 
 - Have time and space to meet with you to discuss your project. At the start, I would try to make such meetings regularly (e.g. every week or two weeks), so that you can both have some regular regime with clear deliverables
 - Provide a critical appraisal of your ideas and plans: these are both positive and negative, but more important is that they are reasoned feedback with explanations that allow you to understand each comment. At the start of your PhD, many of your meetings are going to be about planning and the practical side of doing the work ahead. Your advisor should have experience in these aspects, or be able to point you where to go, and you will be on a steep learning curve to catch up.
 - Read and comment on written work that you produce. Although this will likely come later in your relationship, you should expect that your advisor is willing to read and comment on what you have written and that this is done within an agreed time frame. If you have regular meetings, then the time between meetings is a logical period for this kind of feedback. There will be lots of caveats here because academics are busy people and some times of the year are especially busy. You will nearly always be the person who needs to remain flexible on timelines, but planning will go a long way toward managing your own expectations here. If you know that there is a deadline for comments a particular piece of written work, like a grant application, then be explicit about this before you hand it in and work with your advisor and their schedule. In my experience, the time taken to read a piece of written work, be it a literature review or a chapter draft, is a common reason for the upset of PhD students. Some difficulties can be ameliorated through meeting notes (see [below](#s-a-agreement)).

**As a student you should:**

 - [**Take ownership of your PhD**](#Responsibilities)
 - [Adhere to the scientific ethic](#Respect-and-Professionalism)
 - [Practice good time-management](#Time-management)
 - [Communicate effectively](#Communication) 
 - [Take advice seriously](#Take-advice)

Each of these ideas is discussed in more detail below.

### Keep your expectations real

Don't expect that you are going to be best friends with your advisor. It is known that student-advisor relationships that become too friendly are less productive [@hockey1995getting]. One of the reasons why advisors don't become best friends with their students is that often they simply don't have the time [@mccallin2012postgraduate]. The demands on the time of a typical academic are so great that they likely don't have much time or energy to take their relationship with you beyond anything professional, even if they wanted to. It may well be that you do form a great friendship, but this will likely only happen if you start with a professional relationship and mutual respect. Relationships take time, and it is important to start professionally. This does not mean that you should be cold, unkind or unfriendly. On the contrary, the best professional relationships include warmth, friendliness and kindness. If you have to have a single watchword for your relationship with your advisor be kind. Try reading the views of Howells et al. [-@howells2017role], the need-supportive relationships Of Janssen et al. [-@janssen2021sensemaking], and the impact of gratitude on the student-advisor relationship or that of Lundgren and Osika [-@lundgren2021cultivating] on gratitude. I prefer to ask people to concentrate on kindness, because all of us understand what it means to be kind, and we should all be able to bring kindness with us to our workplaces.

### Student-advisor agreements {#s-a-agreement}

Many departments or faculties have student-advisor agreements that are there to help both parties. They should be modified and signed at the beginning of your PhD time. Take this task seriously. Hopefully you will never need to go back and look at this agreement as your relationship with your advisor will be professional and productive. [But in the case of difficulties](#advisor-go-wrong), you can go back to this and know exactly what was agreed to by both parties, and this may help you judge on whose side the relationship has failed.

Having a student-advisor agreement at the start of your PhD can be very useful in future, and if this is not practised in your department, then you may want to suggest adopting a generic one. Even if they don't like your idea, it might be worth you finding one that you can use with your advisor. This becomes especially important if you think that there could be a potential [Conflict of Interest](#CoI) in future.

### Other students and postdocs in the group

While it is a very good idea for you to have good relationship with everyone in the group, I suggest that it is not a good idea to immediately start asking the opinion of different group members about your shared advisor(usually the group PI). The reason is that each member of the lab will have their own relationship with this person, and this may have little bearing on yours. You will likely hear if there are general observations that people share, but these may not be constructive.

Consider how this situation would be in any other relationship forming experience that you have in your life. When you first meet someone that you expect to spend constructive time with, discussing this person with other people that they know is less of a priority than making good impressions and spending time as constructively as possible with that person. For example, if I have a new doubles partner in a tennis competition, I should be most interested in spending time training with them to learn about their strengths, than going to other people in the tennis club and discussing their weaknesses. This is not to say that you should not form a relationship with other people in the group, this will also be an important part of your work dynamic. Just like other relationships in life, you will have individual relationships with others in the group, and then you will have a relationship with the group as a whole. Group dynamics are often very different from relationships you have with individuals in the group.

## Power imbalance {#power-imbalance}

Your advisor holds the power to influence your career progress. Navigating the power dynamic between PhD students and their advisors can be challenging, as supervisors typically hold more authority and experience in academia. Many academic institutions still have a hierarchical ethos that the most senior academics should never be questioned. This is clearly problematic from the outset, and if your advisor is a senior academic who has become habituated to this kind of high-power status, then the results for many of their students can be disastrous [see @malaga2004meyer;@abbott2004junior; but also see @vences2004meyer for alternative viewpoints]. Essentially, the situation is very similar to that of [bullying](https://howtopublishscience.org/bullying.html), which is quite prevalent in academia [see @measey2022how].

One particular outcome, that is not unique to this situation, is the divide and conquer tactics that are used by some academics whereby they have favoured and non-favoured lab members. My PhD advisor used this against all of the people in his lab so that individuals could be in favour one week, and condemned the next. It was a toxic workplace environment in which this PI was not only our PhD advisor, but also Head of Department and he used this additional power to quash any dissent. Suffice it to say that I am more than aware of what can happen when student-advisor relationships turn bad, and our purpose here is to try to ensure that it never gets that bad.

### Relationships between advisors and their students
Many institutions forbid any of their staff to have consensual relationships between their staff and students. This is especially true for advisors and their students due to the power imbalance inherent in this relationship. Where these do occur, they can lead directly to conflict within the relationship [see @gunnarsson2013experience] and the workplace [@secunda2004getting]. According to Gossett and Bellas [-@gossett2002you], when prohibitions on relationships do occur, there are three principle reasons given:

 1. The power asymmetry between professor and student
 1. The potential for favouritism or nepotism
 1. The negative effect on the scholarly environment

But the matter remains controversial as noncoercive relationships between professors and students do happen [@gossett2002you].

### Conflicts of interest {#CoI}

In certain situations, a conflict of interest (CoI) may arise between the student and advisor [see @macdonald2009supervisor]. Measey [-@measey2022how] has covered the issue of CoIs in academic publishing in some detail, and it would be worth reading these examples to get an idea of the potential problems. Several of these examples include conflicts of interest between advisors and students. This is a potentially very serious issue, and I have heard of some truly terrible examples. Sadly, because of the power imbalance inherent in the relationship, it is almost always the student that comes off worst in these situations. The arbitrators in any such disagreement are usually the institution, and they may also be within the same CoI - especially if money is involved.

Each situation is likely to be individual, and so it will be difficult to hand out any general advice. However, if you have a signed student-advisor agreement, which includes the issue of a CoI, then you may be able to use this to bolster your case with your institution in the occasion of a disagreement. If your advisor is involved in fraud, then there should be a mechanism within your institution whereby you can report them. You may need to seek advice both inside and outside of your institution before taking this step.

## Communication

Communication breakdowns are not unusual between individuals, especially if both parties are under a lot of pressure and stress. You have a responsibility to make your communication as effective as possible. If you are experiencing difficulties in communicating with your advisor, try for [more regular meetings](#advisor-meet) with more formal note taking and agreements on deliverables. Make notes at the meeting, deliver these afterwards in the form of an email with the agreed deliverables listed. Bring this with you at the start of the next meeting.

Be open and honest about the challenges you're facing, including difficulties with meeting agreed deliverables. Most advisors will be more than happy to accommodate you if you are clear about what you need in order to finish the task. Remember that your requests have to be realistic, and that you need to be considerate with other demands that your advisor may have on their time.

### Cultural differences

It is possible, and perhaps even likely, that you and your advisor come from different cultures. Academics are famously mobile and often work in countries and cultures that are not their own. Similarly, students often go overseas to study and so can end up with an advisor that is not from their own culture. Although it is difficult to provide specific guidance here, both parties need to be aware of potential cultural differences. For example, if your culture as a student means that it is inappropriate for you to say "no" to your advisor, if they are from another culture they may not be sensitive to this and think that you are agreeing to everything they suggest. The key will be communication. Communicate as honestly as you can and ask for clarification if you feel that your advisor has been unclear.

## Meetings {#advisor-meet}
I suggested above that you arrange to have regular meetings with your advisor, but it is important that you are proactive on what these meetings are about and follow up with notes. Always take your lab notebook (or a special meeting book) with you to the meeting so that you can make notes on the advice given, as well as being able to go back to look at what was agreed last time. As an advisor, I often get frustrated that students come with no notebook to a meeting and instantly forget all the advice they are given, only to return with the same set of issues the following meeting.

You should also think about having an agenda for your meeting (a list of points that you need to discuss). I like the idea of having a set meeting agenda where the topics inside it change, but the format is similar from meeting to meeting. You can see an example of this in Table \@ref(tab:advisor-agenda)

Table: (\#tab:advisor-agenda) **A typical agenda for a regular meeting with your advisor.** Although the subject of each topic may change for each meeting, the items are likely to follow a standard pattern.

| Agenda item | Examples of what this could consist of |
| - | ----- |
| Catch up | Provide a quick summary of what you have been up to since the last meeting, be sure to emphasise any key realisations (e.g. from your reading) or results (e.g. from experiments). By discussing these insights, it will help you both to align your current thinking on a topic, and quickly get to a more insightful level. This should also be the most lively part of the meeting where you can exchange ideas and get enthusiastic. |
| Tasks agreed during the last meeting | Using your notebook, go through the tasks that were agreed (on both sides) and report back on any deliverables. |
| Feedback on work submitted | Although feedback will often be given as comments on a piece of writing, it is helpful to get some verbal feedback that might give you a much better overall feeling. It will also allow your advisor to highlight points that they think are particularly important and gloss over others that are less significant. |
| Problems and possible solutions | Never put this item first. As an advisor, there's nothing worse than someone starting a meeting with all their problems. It sets the tone for the meeting and tends to drag it down. Remember that this is your project (and they are only advising) so the problems are yours (not theirs) and they are there to provide advice. Whenever you discuss a problem try to come up with a possible solution, or more than one. It is much easier to provide advice to someone who has already started to think about how to solve it than being expected to come up with a solution. For your part, it will be much more satisfying if your proposed solution turns out to coincide with the best advice. | 
| Tasks before the next meeting | Remember to make notes on any deliverables that are agreed upon during the meeting. At the end of the meeting, read these out and make sure that you have it right. You are looking for agreement on each of these. Also, ask your advisor if there's anything extra that they think you've missed. In the case that you are struggling with getting feedback on deliverables from your advisor, you may want to write them an email summary of the meeting with the agreed deliverables so that there is a record. You don't need to start doing this at the outset of your relationship, but after 4 to 6 months you should know if this is needed. |

Having regular meetings with your advisor will also help you to avoid any competition that there may be for their time. As we saw in the last chapter, there are only a certain number of hours in the day and your advisor is going to want to free themselves from most meeting commitments as possible. This means that you need to keep your meetings, short, productive and enjoyable.

## Responsibilities

Knowing who has what responsibilities in your student-advisor relationship is key to avoiding potential troubles. Certainly, in my experience the most misunderstood responsibility is that the student believes (wrongly) that the project that they are doing belongs to their advisor. This failure to take ownership of their work has a snowballing effect and tends to enter into every part of the student-advisor interactions. From the student side, they see every difficulty and problem with the research as something that they give to the advisor to solve. From the advisor's side, the student appears to take the role of a child and appears incapable of doing anything for themselves. Once this negative spiral starts, it will not end well for either party. No matter how many times the student is told that they need to take ownership of their work, they never appear to understand and, speaking for myself, this has resulted in the worst student-advisor interactions I have experienced. For this reason, the number one responsibility must be for the student to:

**Take ownership of your PhD**

Once this has happened, everything else follows without much difficulty.

Advisor responsibilities are:

 - To give advice and feedback in a timely manner
 - Provide the resources needed to complete the project (or at least guide towards resource provision in the case where students are raising their project funding)
 - See that all necessary administrative details are carried out. Nobody likes this aspect, and this does not mean that the advisor does all of this work - they may ask the student to help, or there may be an administrative assistant. In many institutions, timing is very important and forms must be completed and submitted by certain dates otherwise penalties may well be incurred. It is better if the student is also aware of these dates so that they can place them in their calendar. 
 - Not to burden the student with duties that are beyond their role. This can be complicated if the student is employed by the advisor (e.g., as a demonstrator or as a research assistant on a project), but I hear of nightmares where advisors effectively have students take over the running of their laboratories and the other personnel within them. Either party should be able to ask for help with things that are outside the purview of the project, but there are real examples of abuses that I find distressing. 
 - Help build a network for their students. Networks are extremely important for students when they move on from their PhDs, and even within this time [see @measey2022how](https://howtopublishscience.org/advisorless.html). The more your advisor understands your [plan](#Plan), the more likely they are to help you build an appropriate network. 
 - Professional development is very similar to building a network, but this may include training in particular skills. As above, your advisor can only do this if they understand your [plan](#Plan) and what your future needs will be.

Additional student responsibilities:

 - Be sufficiently well-organised and manage their own time (in meetings and other areas)
 - Communicate effectively
 - Be pro-active with [time-management](#time), finding solutions, and 
 - Carry out their project to their satisfaction. 
 - **The most important point will always be for you to: Take ownership of your PhD**

### Time management {#Time-management}

Time management is perhaps one of the most important skills that you will learn during the course of your PhD (see [Part I](#time)). For much of your PhD period, your time will be your own to do with as you see fit. But in order to complete your PhD project on time, you will need to work in an efficient manner. There is a separate chapter on [time management](#time) and a lot of advice on the internet, but the most important point will be to find a style that works for you. Different people will swear by what works for them, but time management is an individual relevant skill that is not necessarily transferable. Find what works for you best as soon as possible, and then plan [the total duration of your PhD studies](#whole-period) and [day-to-day calendar](#day-plan) accordingly.

With respect to your advisor, remember that they will also be managing their time and that they are likely to have a lot of additional commitments. They also only have a certain number of hours in the week to fulfil all of these commitments. This means that you will need to be flexible in accommodating them and fitting into their times. Then, when you do [meet with them](#advisor-meet), you need to make sure that you make the most of the time available.

## Respect and Professionalism

Your relationship with your advisor is a professional one. They are a professional, and they have many more duties than advising you and you should respect this. At the same time, they have a professional duty to your supervision. I suggest that you maintain a professional relationship with them whenever possible. This relationship should be built on the respect that you have for one another. The more that you display a professional relationship with your advisor, the more likely it is that they will treat you back with mutual respect and professionalism.

Both advisor and student should share the same goals which adhere to the four tenets of the scientific method: rigour, independence, transparency and reproducibility.

### Take advice {#Take-advice}

The key responsibility of your advisor is to advise you. They are in that role because they have experience, and they will bring this experience into play when they give you advice. If you are given advice, please consider it. Don't simply reject the advice immediately after it is given. If you think carefully about the advice and decide to act against it, bring this up as a discussion point in a meeting and be ready to defend your decision.

There are plenty of times when you are safe to not take advice that you are given by your advisor, but you should always carefully consider it. This is most common towards the end of your studies when you know far more about the project than your advisor does. When your advisor cannot accept a disagreement on their advice, this can lead to serious problems [see @gunnarsson2013experience] and you may need to go back to your student-advisor agreement. Most advisors should be sufficiently wise to allow you to not follow their advice. But there will be some occasions when you cannot go against the advice. For example, if the funds for your project have been raised by your advisor and they are responsible to the funder, then you do not have the freedom to decide how they will be spent. Similarly, if deadlines are in play for your advisor then you need to respect their professional responsibilities to others.

In general, your student-advisor relationship exists because your advisor is there to give you advice on how to achieve your project goals. If you never have any intention of following any of the advice that they give you, then you misunderstood the relationship.

### Personal problems

Everyone has their own personal problems, but it is not fair to bring all of them to the workplace unless they are likely to impact your work. I have started with this as a statement as students often find it hard to decide what personal problems they need to bring to their advisor. Simply ask yourself, is this problem likely to affect my work? If the answer is yes, then inform your advisor. When you inform them, be aware that it is your problem (and not theirs), and make sure to point out to them as precisely as you can how you think it will affect your work. Advisors are also human, and they are likely to understand. But you shouldn't be hoping for more than understanding. You can't look to your advisor to help solve your personal problem.

**Of course, this goes both ways.**

Advisors are people too. They make mistakes. They have limited hours in the day. They have lives outside of academia that can influence their lives inside (feeling sick, family issues, holidays, etc.). There will be days when your advisor might be in a bad mood; maybe it's a day when they didn't get the research funding they had applied for. There will be many external frustrations for them, and this is unlikely to be a reflection on you. Be kind and considerate. It is worth taking time to share their down times, if they are willing to, at the very least so that you can understand them a little better.

There will be times when the home life of your advisor may impinge on their responsibilities to you. When they tell you, they are looking for your understanding. Be kind to your advisor as you would want them to be kind to you.

## When things do go wrong {#advisor-go-wrong}

There are many possible courses of action open to you, but my best advice is to know what these are before anything bad happens. Just like protecting your [mental-health](#healthy), you can be proactive to know what your options are in the case of your student-advisor relationship going bad. The best is to have a signed [student-advisor agreement](#s-a-agreement). In addition, you can find out what provision your institution has made in such eventualities. If they have no formal schema (which seems unlikely these days), then you can try to have one instigated by taking a template from another institution.

Advice for when things go wrong with your advisor is very similar to that of [academic bullying](https://howtopublishscience.org/bullying.html) [@measey2022how; @mahmoudi2021filling; @mahmoudi2020survivor; @malaga2004meyer]:

1. Increase your level of professionalism.
1. Ask for your institutional rules.
1. Document your case.
1. Share your burden with a trusted friend or colleague.
1. Follow your university’s rules with your complaint.

I want to emphasise point 1 on the above list. While you may not be the cause of this problem, the best way of tackling it will be to become extra-professional. This means becoming meticulous about documenting [your interactions](#advisor-meet), and maintaining a high level of professionalism as you do so. The more you increase your level of professionalism, the less likely your badly behaved advisor will be able to get away with abusing their power.

## Having more than one advisor

This happens with increasing frequency, and disputes between advisors can leave the students in the middle with serious problems [see @gunnarsson2013experience]. I have experienced these kinds of issues with students who are co-supervised by another academic. When the academics disagree about the way forward, I tell the student that they have had advice from both advisors and that it is now up to the student to decide which advice (if either) to follow. I don't see this as a problem. But not all academics are the same!

In the case that you do find yourself in a position with two advisors then I suggest that you make sure that you have [an agreement](#s-a-agreement) with both of them that includes what to do in these situations.

## Celebrations

I hope that there will be lots of reasons for you to celebrate with your advisor and the rest of your team. Celebrating achievements is a really important part of working together, even if the achievement is simply that you have survived another year. If there is no existing tradition of celebrating in your lab, then it's time for you to start one.

### Share the down times

While I hope that there will be lots of reasons for you to celebrate, we should not pretend that there will not be down times when things don't work and people get frustrated, upset and disappointed. It is important to share these times too so that people understand why you might not be at your best on any particular day. Don't keep it to yourself. If you and your advisor can be kind to each other, and take the time to commiserate the down times as well as share the successes, then you will likely have a good productive relationship.

### 2024-09-11 1.09-regularwriting.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

# Regular writing {#regularwriting}

This book is about writing your PhD in Biological Sciences. This chapter is about using regular writing as a core skill to improve your writing skills throughout the duration of your PhD studies. Writing is a [core skill](#) that you need to work on, and a regular writing slot is the best way to develop this skill. This chapter will be short because, unlike the other core skills in this section of the book, the rest of the book will go into scientific writing in detail and provide you with all the advice you need to get started writing.

Just like any skill, you can't expect that scientific writing is going to come easily to you, and you need to invest time not only to learn how to do it (Parts [II](#partII) and [III](#partIII)), but also to practice writing so that you implement the skills that you learn on a daily basis. Regular writing and implementing what you learn in this book will be the best way to improve your scientific writing skills.

## Write every day

When you make your [day-to-day](#day-plan) calendar for your normal working week, I strongly suggest that you put a 40 minute block every day that you dedicate to scientific writing (i.e. thesis writing). Know what topic you are going to write about before you start, and make sure that you have already conducted the relevant [critical reading](#critical) to accomplish this. Start your 40 minute block by writing the [outline](#outline) of your work with a list of relevant papers to cite. Then concentrate for these 40 minutes on writing something that meets the goal you set.

You need to be realistic about what you can achieve in 40 minutes. Especially at the start of your PhD, you might not be able to write more than a few sentences or a paragraph in 40 minutes. At the very beginning you might even struggle to complete your [outline](#outline). However, the discipline of dedicating 40 minutes every day to undistracted writing will pay-off over time.

Don't expect perfection in these 40 minute sessions. Instead, concentrate on your plan and think about how to introduce ([topic sentence](#paragraph)), and [logically link](#argument) sentences that are backed up with [citations](#citations). The first paragraph that you write may not be a keeper, but making that start is important as it will show you what you need to change.

Build on your paragraph, but don't return to the same paragraph every day. It is important to change the writing topic often enough so that you don't get stuck on a particularly difficult task. By changing the topic several times during your week, you'll should find that it retains your interest. You can add plan in weeks to come to return to tasks that you are struggling with or which need more time to complete.

While I recommend that you write every day, I would stress that this means every working day. You should have a break for at least one day a week. You should also have some days every year when you have a holiday and others where you are doing other tasks, like fieldwork, that will preclude you from writing every day.

### Remember your critical reading

Daily writing will go hand-in-hand with daily [critical reading](#critical). Plan both of these activities in a complimentary way so that you have the reading foundation for what you plan to write. However, make sure that you mix it up so that you get some variety in what you are reading.

### Reflection {#reflection}

Once you start writing regularly, you will find that you have ideas about how to solve writing problems and ideas about what to cite at other times during your day. Keep a 'writing notes' file or a page in your notebook where you can jot down these ideas, and come back to them in your regular writing time-slot.

As stressed in the chapter on [health](#healthy), [time spent thinking](#time-to-think) and reflecting on many aspects of your PhD is very important. During your sleeping hours your brain will make connections between items and tasks that you have conducted during the day. Hence, it may be that it may be a only day or two later that connections become clear to you. These insights can be very valuable, so make sure that you keep a note of them and look into them further to see whether they are worthy or important. Remember that your brain can also make false connections, so not every idea will be a winner.

### Use one of the best time-slots

When planning your regular writing, use one of the time slots when you know that your [concentration will be at its best](#day-plan). Don't use a time when you know that you are likely to be tired or unfocussed. The idea here is to maximise your ability to write, during these time-slots. In time, you may well find that as your writing becomes better you will need to dedicate the best time-slots to other priority tasks. But when you start, I'd advise you to begin by keeping the best time-slots for daily writing tasks and [critical reading](#critical).

### When you haven't read enough

At first you will not have read enough papers to properly plan or cite all the relevant literature for the paragraph that you aim. In these first few weeks, concentrate on some of the writing techniques that you are reading about (in this book and elsewhere) so that you get some practical experience of writing some logical sentences, writing hypotheses, etc. The point of making this daily practice in writing is to get yourself used to writing in a scientific way. You can start this from day one, only having read a few of the chapters from this book, so give it a go.

## Be disciplined

Make sure that the 40 minutes you have allocated to writing are spent doing writing tasks. If you fill your 40 minutes with reading, arranging your chair, screen, table then you won't be helping yourself. Especially if you reinforce this by rewarding yourself it will actually cement these [displacement activities](#displacement) instead of rewarding being productive. It will be up to you to decide whether or not you have achieved your goals.

### 2024-09-11 1.10-lifescientific.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

--- 2021-07-20 1.03-lifescientific.Rmd
@@
 
-Throughout the book, I make reference to the ‘scientific project’ as a greater plan to which our studies in biological sciences is simply a part.
-I consider the scientific project as a path that humanity should continue to travel in order to reach a higher and more just way of living.
+Throughout the book, I refer to the ‘scientific project’ as a greater plan of which our studies in biological sciences are simply a part.
+I consider the scientific project as a path that humanity should continue to travel to reach a higher and more just way of living.
 Our societies have evolved from places where superstition and belief were the basis for societal rules that favoured the privileged classes.
-Science is a process in which we reach conclusions based on egalitarian reasoning and evidence, something that we can aspire for the larger community around us.
+Science is a process in which we reach conclusions based on egalitarian reasoning and evidence, something that we can aspire for in the larger community around us.
 Already, most societies are making the majority of their collective evidence-based decisions through the guidance of science.
 Therefore, if we consider our knowledge-based societies to be democratic, then all citizens of that society should understand the processes under which the foundation of that knowledge is acquired [@sagan2011demon].
-Yet globally, public skepticism in science has been rising, and has reached as high as 35%.
-To consolidate this growing position of science in society, and the decline in public acceptance, scientists will need to become more open, communicative and seen to be conducting their own profession in a way that reflects these ideals.
-There is evidence that increasing communication from scientists to the public results in a decrease of skepticism ([see here](https://multimedia.3m.com/mws/media/1898512O/3m-sosi-2020-pandemic-pulse-global-report-pdf.pdf)).
+Yet globally, public skepticism in science has been rising and has reached as high as 35% ([see here](https://multimedia.3m.com/mws/media/1898512O/3m-sosi-2020-pandemic-pulse-global-report-pdf.pdf)).
+To consolidate this growing position of science in society, and the decline in public acceptance, scientists will need to become more open, communicative and seen to be conducting their profession in a way that reflects these ideals.
+There is evidence that increasing communication from scientists to the public results in a decrease in scepticism [@goldenberg2023public].
 
@@
 Think of the views on the earth being round, or more recently, the general acceptance that the universe started with the ‘big bang’.
-The counterculture, also known as post-truth, faction in society that espouses views accepted by the mainstream and proved by science beyond reasonable doubt is not a novel phenomenon.
+The counterculture faction in society, also known as post-truth, that espouses views accepted by the mainstream and proved by science beyond a reasonable doubt is not a novel phenomenon.
 Indeed, in the scientific project, we need to accept that these views will persist and empower the general populace to understand why and how they can be shown to be false.
@@
 The scientific project is currently not the open virtuous eutopia that we would like.
-Indeed, as scientists are humans and part of the current society, they are prone to the same trappings of wielding power and influence in order to get ahead.
+Indeed, as scientists are humans and part of the current society, they are prone to the same trappings of wielding power and influence to get ahead.
 As you get toward the end of this book, you will see me calling out increasing parts of the scientific project that are currently riddled with bias, thereby weakening the entire process and disenfranchising the most vulnerable.
@@
 1.
-[Use a hypothetical deductive framework](howhypothesis.html).
+[Use a hypothetical deductive framework](#hypothesis).
 Instead of accepting the first (hypothesis) explanation, think of all the possible reasons why a result might have occurred and then consider how you could (or perhaps you already have in your experiment) disproved each one.
@@
 Citizen science contributions mean that the collection of data can be shared on a much broader scale, than would be traditionally possible.
-Of particular interest in the biological sciences is the identification and collection of locality data for biological species (on platforms such as iNaturalist).
+Of particular interest in the biological sciences is the identification and collection of locality data for biological species (on platforms such as [iNaturalist](www.inaturalist.org)).
 But increasing numbers of projects are demonstrating novel ways for the public to increasingly participate meaningfully in scientific endeavour [@bonney2014next; @silvertown2009new].

### 2024-09-11 2.03-NoHyp.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

@@
 Clearly, we are also able to make mistakes when we infer patterns from processes, but this does not mean that it should not be done.
-But we need to engage in a much more cautious manor when using this approach, always being aware of our bias towards certain explanations so that we encompass others.
+But we need to engage in a much more cautious manner when using this approach, always being aware of our bias towards certain explanations so that we encompass others.
 I would also argue here that this is instrumental in forming hypotheses to test experimentally or in models.

### 2024-09-11 2.05a-AI.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

@@
 The emergence of artificial intelligence (AI) platforms brings with it new issues around writing generally, including writing your PhD thesis.
-These bots are already widely available in early 2023, with both Microsoft (through their company OpenAI) and Google launching [ChatGPT](chat.openai.com) and [Bard](bard.google.com), respectively.
+These bots are already widely available in early 2023, with both Microsoft (through their company OpenAI) and Google launching [ChatGPT](chat.openai.com) and [Gemini](gemini.google.com), respectively.
+Note that Google originally launched as Bard in early 2023, but then later released as an update known as Gemini.
 There are already publications that list ChatGPT as an author [@oconnor2023open; @mijwil2023artificial] in acknowledgement that a large amount of the text therein had been generated by the bot.
@@
 
+## This field is moving very fast {.unlisted .unnumbered}
+
+Bear in mind when you read this that the original chapter was written in March 2023, with updates nearly every month since then.
+This current version was compiled on `r format(Sys.time(), '%d %B, %Y')`.
+If you think that the content here (or elsewhere in the book) is out of date, or needs additional text, please let me know (see [here](#contribute)).
+
 ## What is a Large Language Model (LLM)?
@@
 
-**ChatGPT**
+#### **ChatGPT** (March 2023) {.unlisted .unnumbered}
 
@@
 
-**Bard**
+#### **Bard** (March 2023) {.unlisted .unnumbered}
 
@@
 
-To summarise, LLMs are trained on an exceptionally large amount of data and use AI to first understand what you are asking it, and second pull out of their training set what they need to reply to your user input.
+#### **Gemini** (May 2024) - note the shift from using first person (Bard) to third person (Gemini). It's a deliberate change made by Google. {.unlisted .unnumbered}
+
+>Gemini is actually the new name for what was formerly known as Bard. It's like a more powerful and upgraded version. Here's the breakdown of the key differences:
+
+>Underlying Tech: Bard primarily relied on Recurrent Neural Networks (RNNs) for processing information. Gemini utilizes a transformer-based design, which is considered more powerful for various tasks.
+Training Data: Gemini is trained on a much larger and more diverse dataset compared to Bard.
+This includes text, code, and even images, allowing it to understand and respond to a wider range of prompts and questions.
+Capabilities: Bard was focused on text-based interactions.
+Gemini, however, has multimodal capabilities.
+This means it can not only process text but also understand and generate different formats of data, including images and potentially even video.
+In short, Gemini is designed to be more versatile and powerful than Bard,  with the ability to handle complex tasks and integrate with other Google tools more seamlessly.
+
+>To summarise, LLMs are trained on an exceptionally large amount of data and use AI to first understand what you are asking it, and second pull out of their training set what they need to reply to your user input.
 The key then is that whereas previous chat bots had (relatively) small training sets and limited ability to understand user input, these new LLMs have vast amount of data.
@@
 Some authors consider that even if the AI is programmed to avoid plagiarism, using this text in any way should be considered as plagiarism [@salvagno2023can].
-However, on balance, my own view is that any attempt to be overly restrictive on the use of AI generated text is bound to fail as AI itself becomes increasingly more sophisticated and widespread.
-Instead, I share the view [@lim2023generative] that it is better to embrace and learn how to use AI as a tool.
-Learn about what it can do, as well as its limitations.
+Certainly, if you use AI generated text and then claim that you have written this yourself, then this is clearly plagiarism.
+In this case, the plagiarism is due to the false claim that you wrote the text.
 
@@
 Clearly, the definition of plagiarism will need some work in the light of new abilities for AIs to write text.
+What I see happening is that both universities and publishers are asking students / authors to declare that they have not generated any text using AI.
+This is a catch-all and will mean that you simply cannot use AI generated text.
 
@@
 
-Plagiarism is the subject of another chapter, and you can read more about it [there](#plagiarise).
+My own view is that any attempt to be overly restrictive on the use of AI generated text is bound to fail as AI itself becomes increasingly more sophisticated and widespread.
+Instead, I share the view of Lim [-@lim2023generative] that it is better to embrace and learn how to use AI as a tool.
+Learn about what it can do, as well as its limitations.
+There is an entire chapter on plagiarism (in [Part II](#plagiarise)) where AI generated text is covered, and I'd advise you to read this to learn more.
 
@@
 Don't try to use it to replace your own writing.
-That will likely end in a poor product, and could possibly end with factually incorrect or biased text that you won't want to be associated with your name - and clearly not something that you should want your advisor to read.
+That will likely end in a poor product, and could possibly end with factually incorrect or biased text that you won't want to be associated with your name, and clearly not something that you should want your advisor to read.
 
 Getting started is often a struggle and I regularly suggest that students start by writing something/anything to fill out their outline, even if all of that text gets replaced at a later date.
-This is a way that you could use a tool like ChatGPT.
-Produce your own outline, because you do still need to plan your chapter, but then try using an AI like ChatGPT to flesh this out.
+This is a way that you could use AI as a tool.
+Produce your own outline, because you do still need to plan your chapter, but then try using an AI to flesh this out.
+Read through the output and get inspired.
 Hopefully, this will end by helping you to write your own text, replacing everything that the AI has produced.
 
-It is worth noting that the input for ChatGPT is dated to a portion of the internet from September 2021 [@azamfirei2023large].
+#### **But do not copy and paste it into your document.** {.unlisted .unnumbered}
+
+Use the AI to help you get stated but don't copy it.
+Just like you might read the text of an article and that helps you get some understanding, you can use AI generated text in the same way.
+But if you copy and paste it, you'll start on a slippery slope towards plagiarism and potential disqualification.
+
+#### **Be aware that AI can get it wrong.** {.unlisted .unnumbered}
+
+Back at the start, ChatGPT was dated to a portion of the internet from September 2021 [@azamfirei2023large].
 It cannot access the internet and so you will not get anything more recent.
 Whereas Bard is trained from Google search results, Wikipedia, [The Common Crawl](commoncrawl.org) and other publicly available datasets.
+This has now changed, but it may well be that the AI you use has not accessed the most recent research.
+Currently, you are a much better researcher and judge of the written work than any AI.
+So, don't believe what it tells you without checking everything!
 
@@
 Some people claim that they can detect the output from ChatGPT as it lacks the depth and insight that original authors usually have.
-In other words, the aim of ChatGPT is to produce the words (in a grammatically correct manor), while the aim of an author is to transmit an idea to the reader.
+In other words, the aim of ChatGPT is to produce the words (in a grammatically correct manner), while the aim of an author is to transmit an idea to the reader.
 As you might expect and hope, AI is not at the point where it can generate the same intent to communicate.

### 2024-09-11 2.16-foreign.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

@@
 Reading this book, you will learn of some biases in biological sciences and how you need to be aware of them in order to properly interpret your work.
-Be aware also that biological sciences is biased in where studies take place, and who publishes them [@culumber2019widespread; @nunez2021monolingual].
+Be aware also that biological sciences is biased in where studies take place [among [a lot of other biases](https://howtopublishscience.org/pubbias.html): @measey2022how], and who publishes them [@culumber2019widespread; @nunez2021monolingual; @martin2012mapping].
 What we really need then are more studies from less studied parts of the world, and we need these to be researched and published by people from those countries.
@@
 
+## Do I need to write in perfect English?
+
+The best news is that the plethora of translation tools from various languages to English is getting so good that although learning how to read and write good English is always going to be useful, it is not essential.
+I would suggest that instead, you put your efforts into learning scientific writing style (as taught in this book and others) instead of learning the rules of English grammar.
+What we are interested in is being able to communicate your science to a global (mostly non-English speaking) audience.
+There is a world of difference between being able to write good English and being able to write scientific English, and most English speakers struggle to make this change unless they are taught.
+The aim of this book is to teach people how to change their style into scientific English.
+
+One of the major difficulties for non-English speaking students learning to write scientific English is that they don't appreciate the difference between good English and scientific English.
+Actually, you don't need to learn to write English, you can learn to write in the scientific version of your own language.
+But that must encompass the same scientific style that is being taught in this book.
+If you want to learn what the scientific style reads like in your own language, then I suggest making a translation of a paper in a journal that forms part of your [core target reading](#critical).
+Then examine this translated paper using the [critical reading](#critical) tools you'll find later in [Part II](#critical).
+
+If you are not confident in writing in English, then write in your own language and have it translated. But you must make sure that the style used is the correct scientific style.
+
+As a reviewer, I read plenty of manuscripts submitted by people who are writing English as a second (third or more) language, or writing in their own language and having it translated.
+Those that have learned the scientific style generally do very well with minor issues in grammar.
+They will not get penalised for these minor grammatical issues, even though they could clear most of these up with a [good grammar editor](https://www.grammarly.com/).
+Those who don't produce manuscripts are not suffering with language as much as a failure to understand the scientific style.
+
+Specifically:
+
+ - Their [citations](#citations) are poor or incorrect
+ - Their studies often lack a [hypothesis](#hyp) or [clear aim](#nohyp)
+ - Their [arguments](#argument) are poorly made or inconsistent
+ - The study approach or study design is unclear
+ - The writing is not [concise](#concise)
+
+The good news about all of these deficiencies is that it is possible to learn how to do these well in your own language, and that none of them really have anything to do with English, but everything to do with writing in a good scientific style.
+Solving these problems can be done in combination.
+First, take a good guide to scientific writing ([like this one](#contribute)) and try to incorporate the lessons taught into your writing.
+Second is [critical reading](#critical) which should tell you all you need to know about how to write well.
+My suggestion is to do both.
+The book can help you to demystify the scientific style, and [critical reading](#critical) is a great skill to acquire and will take you forward with all you need to learn.
+Lastly, and perhaps most important, is to put into practice what you learn in your own writing.
+Don't let having to submit in English hold you back.
+Move forwards in your own language and apply all of the tools you learn and then translate once you've finished.
+As time goes by and your English improves, you'll learn that it's really not as hard as it seems.
+
+## Message from an editor
+
 One of my 'pet hates' is that many non-English speaking editors seem to insist that you have your work checked by a 'native English speaker' (see Figure \@ref(fig:editor)).
 In my experience, having advice from a 'native English speaker' is no guarantee to getting a well-written chapter or manuscript.
+I have had many native English speaking students who are not able to spot very obvious mistakes, and make many of their own.
 Most of the world’s scientists did not grow up speaking English.
@@
 So, if English isn’t your mother tongue, should you expect to receive help during peer review?
-
-## Message from an editor
 
@@
 
-Perhaps I should start from the outset by stating that English is my mother tongue, and that I have spent many hours correcting the language of colleagues for whom it wasn’t.
+Perhaps I should make it clear that English is my mother tongue, and that I have spent many hours correcting the language of colleagues for whom it wasn’t.
 However, most of these hours were spent when I was a postgraduate student or postdoc.
@@
 
+First, I would suggest that you check out the wide array of tools available online.
+Some may be behind paywalls, but provide you with a free trial.
+If the service that they offer is good, then consider subscribing so that they can help you in future.
+
+Your university may have a facility to help you with English language translations.
+
 Anyone you know who writes good English and is willing to help you.
@@
 
-Failing that, I’m afraid that the best route will be to pay for help.
+Failing that, the best route will likely be to pay for help. There are many services available to help you, and many scientific publishers now promote these if you are submitting a manuscript.
 

### 2024-09-11 2.17-plagiarise.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

--- 2021-09-05 2.17-plagiarise.Rmd
@@
 We're most familiar with plagiarism of writing and these days this is especially easy with the copy and paste function.
-Most people are not even aware that they have plagiarised because sometime in the past they copied and pasted the work of somebody else into a document and this later became incorporated into their text without them being aware of it.
+Whether the source is the internet, a published (or unpublished) paper, or a translation of something written in another language, if you copy someone else's work and paste it into your own this is blatant plagiarism.
+
+When finding that there is plagiarism, I find that most people are not even aware that they have plagiarised.
+I think that this is because at some point in the past they copied and pasted the work of somebody else into a document and this later became incorporated into their text without them being aware of it.
+If this is your style of writing, then I suggest that you move away from copy and paste mistakes as it can land you in a lot of trouble.
+Rather make your own notes.
 
 Plagiarism is a problem because, essentially, you are taking somebody else's work without attributing it to them.
+However, blatant plagiarism is only one of several kinds of plagiarism, and it is important that you know all of the different kinds of plagiarism so that you know how to avoid it.
+Studies of postgraduate students have found that although all are familiar with blatant plagiarism, more subtle plagiarism is not understood by all [@shen2021chinese].
+
+It is really easy to avoid plagiarism. Follow these three easy steps (Table \@ref(tab:plagiarise)).
+
+Table: (\#tab:plagiarise) **Three simple steps to avoid plagiarism.** Follow these three simple steps to avoid plagiarism in your work.
+
+|  Step 1 | Step 2 | Step 3 |
+| ----- | ---- | ---- |
+| Read the text written by someone else from which you want to draw inspiration from | Make basic (not detailed) notes on this written work | Put the original work out of sight, and using only your notes write out your own work |
+
+There is one final **Step 4** to the above Table \@ref(tab:plagiarise), and that is whether or not to [cite](#citations) the work that you have just used.
+Citations are the topic of another chapter (see [here](#citations)), and you should be familiar with what constitutes a citation and what doesn't.
+If you are taking an idea that is first provided in this specific text, and you are re-writing that concept, then you will need to cite the original work.
+If you are using someone else's writing as an inspiration for your own work, then it may or may not require a citation, and you will need to read the chapter on citations (see [here](#citations)) in order to judge.
+
+Let's try looking at an example:
+
+**Example: Darwin's survival of the fittest**
+
+> "It is not the strongest of the species that survives, nor the most intelligent that survives.
+It is the one that is the most adaptable to change, that lives within the means available and works co-operatively against common threats." ~ Charles Darwin [-@darwin1859origin]
+
+If you want to use Darwin's idea, then you could re-write the concept above, but then you would need to cite Darwin for this idea. As follows:
+
+**My text:** Darwin [-@darwin1859origin] stressed that it is individuals that are most adaptable to change who will survive, rather than the strongest or most intelligent of the species.
+
+But let us suppose that you came across this idea not from reading the original text of Darwin [-@darwin1859origin], but from another author who was writing about the same concept.
+
+> "Darwin proposed that the capacity of organisms to produce more offspring that can be supported by the environment would lead to a struggle for existence, and individuals that are most fit for survival and reproduction would be selected through natural selection."  ~ Samuel A.
+Cushman [-@cushman2023entropy]
+
+If this was the text that inspired you to use the concept, you should not cite Cushman [-@cushman2023entropy] for the idea, nor should you use any of his text.
+Instead, I'd suggest that you go back to Darwin [-@darwin1859origin], and take your inspiration from there.
+
+To take this to another step, I asked [ChatGPT](https://chat.openai.com/) to rewrite the text of Chushman [-@cushman2023entropy].
+
+**ChatGPT text:** Darwin suggested that when organisms produce more offspring than their environment can sustain, it triggers a competition for survival.
+Through natural selection, the individuals best adapted for survival and reproduction are favored.
+
+While the text provided by ChatGPT is not wholly incorrect, there are some subtleties that are worth inspecting.
+For example, the term "survival of the fittest" was coined by Herbert Spencer [-@spencer1882principles], not Charles Darwin and there is some consensus that Spencer's [-@spencer1882principles] use of this term was not completely aligned with that given by Darwin [@offer2014natural].
+It appears that Chushman [-@cushman2023entropy] linked Spencer's term to Darwin's [-@darwin1859origin] concept by using the word "fit", which is commonly the case.
+However, ChatGPT uses "best adapted for survival and reproduction" which is our modern definition for fitness, but not what Darwin [-@darwin1859origin] originally wrote: "the strongest...
+[and] ...the most intelligent".
+The point here is that when you cite the original text, you should go back to this text, and seek out what the author wrote and intended, not a later citation or interpretation.
+
+Next, when you are wanting to cite an original text, you are often interested in the variables that have been investigated in that written work (see [Part II](#hypothesis)).
+In the example of Darwin (above) this would be the variables strength and intelligence, and if your variable was reproductive fitness you would need to explain how this was derived from Darwin's [-@darwin1859origin] original concept.
+Of course, you could opt to use the original text in quotation marks and cite the text (and page) where this was from.
+However, this is not always appropriate, so you should get used to writing your own text about other people's findings, concepts and ideas.
+
+The other point about the ChatGPT text is that while it does not plagiarise Darwin [-@darwin1859origin] or Chushman [-@cushman2023entropy], we do not know if it has plagiarised other text from its input data: text from the internet which is mostly derived from original input by human authors (see chapter on AI [here](#AI)).
+Moreover, as increasing amounts of text on the internet will be derived from AI, we have no idea how skewed and misrepresentative AI generated text will become.
+
+There is more to plagiarism than meets the eye, and it is important that you have a good understanding of what plagiarism is as many postgraduate students only appear to understand the most blatant kind of plagarism when making a verbatim copy of someone else's text [@shen2021chinese].
+
+## The different kinds of plagiarism
+
+It is important to know what plagiarism is, so that you know how to avoid it.
+While almost everyone understands that they must not use other people's work verbatim, it is also important to know that other more subtle forms of plagiarism.
+
+### Blatant or verbatim plagiarism
+
+This is when you 'copy and paste' someone else's work and pretend that it is your own.
+Note that even if you didn't use the copy/paste functions, if you have incorporated someone else's written work into your own text, then this is plagiarism.
+This includes using a translated version of work in another language.
+
+### Paraphrasing
+
+Some students will copy an entire sentence and then change one or two words; this is known as paraphrasing and is also plagiarism.
+Subtle variations on this including switching the order of the clauses in a sentence and substituting some words with synonyms.
+The essence of paraphrasing is that you start with the verbatim form of plagiarism, and you are then attempting to engineer some differences by remixing this content.
+Instead you should take the three easy steps (Table \@ref(tab:plagiarise)) and write it yourself.
+Remember that you may need a citation (see above).
+
+**What follows are more subtle forms of plagiarism:**
+
+### Failure to attribute
+
+If someone else wrote pieces of your thesis and you do not attribute their input, this is plagiarism.
+There are many people that could help with your writing, such as friends, lab mates, or your advisor.
+But if you then pass this work off as your own, it is plagiarism.
+
+With failure to attribute plagiarism, it is not as simple as stating that some text was written by your friend or lab mate.
+This is not an option.
+When you submit your PhD, you will declare that the work is your own and that you have written all of it.
+
+There is another form of failure to attribute plagiarism when you employ a third party to write your thesis for you.
+This process is often referred to as a "[paper mill](https://howtopublishscience.org/authors.html#papermills)" [@measey2022how].
+Even if you paid this person and they gave you permission to use the text in your thesis, it is still plagiarism.
+
+#### Is it plagiarism if an AI bot writes it?
+I would also place the use of AI as a form of failure to attribute plagiarism.
+Currently, we are moving into the new territory of AI generated text.
+You can not use AI generated text (e.g.
+ChatGPT or Bard) in your PhD or in manuscripts submitted to journals as it is prohibited.
+Most institutions make you sign a declaration that you will never do this.
+You can read more on this topic in another chapter [here](#AI).
+
+#### Avoiding failure to attribute plagairism
+
+Avoiding failure to attribute plagiarism is just the same as avoiding blatant plagiarism (see [below](#anitplagiarise)).
+Don't copy and paste their work.
+Do read it, do make notes, but then put it out of sight and write your own work (see Table \@ref(tab:plagiarise)).
+
+#### What is co-authored work makes up part of your thesis?
+Many theses are made up of multi-authored papers.
+The reason for this is that different people in the author line contribute different aspects of the paper.
+If you are in the situation where one of your chapters is partly written by your co-author, what should you do?
+
+The reality is that when you submit such multi-authored work as part of your thesis, you should also have an opportunity to declare who did what in the manuscript (see the CREDIT Contributor Checklist [here](#credit)).
+You should have written the original draft, including all of the text.
+While other members of the author line have have read and reviewed your work, they should not have written any of the sections themselves.
+If they did, then you should take follow the three easy steps (Table \@ref(tab:plagiarise)) and turn it into your own writing.
+
+### Autoplagiarism
+
+I would maintain that there is not really any grey area in plagiarism.
+So what are people talking about when they refer to a grey zone?
+The nominal grey zone within plagiarism comes when you copy your own text, this is sometimes referred to as 'text-recycling', 'self-plagiarism' or 'autoplagiarism'.
+A new guideline from Hall et al.
+[-@hall2021understanding] sets out the different types of grey plagiarism, or text recycling:
+
+#### Developmental recycling {#develop}
+
+Developmental recycling is when you are reusing text that you have written for example between your proposal and something you intend for publication or in an ethics application that you also want to use in your thesis.
+All of this sort of developmental recycling is permitted and actually encouraged.
+I would further encourage you to use the opportunity of recycling this text to develop it and refine it further, condensing and improving where you can.
+
+#### Generative recycling {#generative}
+
+Generative recycling is where you take pieces of already published text for example from the methods when it does not make sense to change the text or actually makes it more obscure to reword it in order to avoid plagiarism.
+In my experience this doesn't amount to more than a few sentences describing technical settings on equipment.
+However this will depend strongly on your own subject area and may amount to larger chunks of text.
+I suggest that it is usually possible to reword most of the methods sections of papers.
+You should really only be generatively recycling material if you cannot avoid it: i.e.
+when the text becomes more obscure by your attempts to reword it.
+
+#### Adaptive recycling
+
+Adaptive recycling is where you are using published text as the basis for a different type of content (e.g.
+a popular article online, a magazine, or op-ed).
+I think that this kind of text recycling is quite unnecessary because you almost certainly need to reword your text for a different audience.
+There maybe times such as figure legends where you need to reuse text that was already published.
+If you do find yourself in such a position then check with the copyright owner of the material that you are able to reuse the text that you want without legal issues.
+
+#### Duplicate recycling
+
+Duplicate recycling (also known as 'manuscript recycling') is where large tracts of texts are essentially the same for the same message and audience.
+This is never likely to be sanctioned as it suggests that you are attempting to publish the same work twice.
+It will not be legal or ethical.
+Manuscript recycling is a bigger problem in other subject areas, notably economics [@geraldi2021self-plagiarism], publishing issues in economics appear to need a lot more professionalism.
 
@@
 
-## Free online plagiarism tools
+## Free online plagiarism detection tools
 
@@
 - Read the sentence that has been plagiarised several times to yourself. 
+- Make a few notes about the meaning of the text, so that you don't need to look back at it (avoid using the same words)
+- Move the original text away so that you cannot see it
 - Now without looking at that sentence, write another sentence that has the same meaning.
@@
 - This can now be added to your text, changed as appropriate to fit your existing text. And that should be the end of your plagiarism worries.
+
+This can be summarised into three easy steps (see Table \@ref(tab:plagiarise)).
 
@@
 
-## How can you make sure that you never plagiarise?
+## How can you make sure that you never plagiarise? {#anitplagiarise}
+
 Quite simply, if you never cut and paste, you won’t plagiarise.
@@
 Don’t forget that the penalty for plagiarism in your thesis might well be that you fail.
+More than that, you could lose your place at the university or even your job [@ruiperez2016plagiarism; @tatalovic2017croatia].
 
@@
 
-Want to read more about plagiarism to understand what it is: see Louw [-@louw2017defining]
+## In summary - just don’t plagiarise
 
-## What is autoplagiarism?
+Plagiarism is the unattributed use of other people's work.
+It is not simply the copying of other people's text verbatim.
+There are many forms, and you need to be aware of the extent of that which is considered to be plagiarism in order to avoid it.
 
-I would maintain that there is not really any grey area in plagiarism.
-So what are people talking about when they refer to a grey zone?
-The nominal grey zone within plagiarism comes when you copy your own text, this is sometimes referred to as 'text-recycling', 'self-plagiarism' or 'autoplagiarism'.
-A new guideline from Hall et al.
-[-@hall2021understanding] sets out the different types of grey plagiarism, or text recycling:
-
-### Developmental recycling {#develop}
-
-Developmental recycling is when you are reusing text that you have written for example between your proposal and something you intend for publication or in an ethics application that you also want to use in your thesis.
-All of this sort of developmental recycling is permitted and actually encouraged.
-I would further encourage you to use the opportunity of recycling this text to develop it and refine it further, condensing and improving where you can.
-
-### Generative recycling {#generative}
-
-Generative recycling is where you take pieces of already published text for example from the methods when it does not make sense to change the text or actually makes it more obscure to reword it in order to avoid plagiarism.
-In my experience this doesn't amount to more than a few sentences describing technical settings on equipment.
-However this will depend strongly on your own subject area and may amount to larger chunks of text.
-I suggest that it is usually possible to reword most of the methods sections of papers.
-You should really only be generatively recycling material if you cannot avoid it: i.e.
-when the text becomes more obscure by your attempts to reword it.
-
-### Adaptive recycling
-
-Adaptive recycling is where you are using published text as the basis for a different type of content (e.g.
-a popular article online, a magazine, or op-ed).
-I think that this kind of text recycling is quite unnecessary because you almost certainly need to reword your text for a different audience.
-There maybe times such as figure legends where you need to reuse text that was already published.
-If you do find yourself in such a position then check with the copyright owner of the material that you are able to reuse the text that you want without legal issues.
-
-### Duplicate recycling
-
-Duplicate recycling (also known as 'manuscript recycling') is where large tracts of texts are essentially the same for the same message and audience.
-This is never likely to be sanctioned as it suggests that you are attempting to publish the same work twice.
-It will not be legal or ethical.
-Manuscript recycling is a bigger problem in other subject areas, notably economics [@geraldi2021self-plagiarism], publishing issues in economics appear to need a lot more professionalism.
-
-## Overall - just don’t plagiarise
-
-The simple way around this problem is not to plagiarise any of your own text (self-plagiarism) or any of anybody else's.
-This is, as I've explained above, relatively easy because as long as you don't copy and paste any text you'll find it very hard to actually use somebody else's words.
-However it can get a little bit tricky when you are writing the materials and methods section of a paper, and especially when that paper uses exactly the same materials and methods as a previous paper.
-The temptation is great to go and copy the text that you've written before.
+The simple way around plagiarism is not to copy any of your own text (self-plagiarism) or any of anybody else's.
+This is, as I've explained above (see Table \@ref(tab:plagiarise)), relatively easy because as long as you use your own words, you should be plagiarism free.
+The temptation will always be great to go and copy the text that has been used before.
 But be aware that, however grey, this is plagiarism and you should not do it.
-
-The likelihood is that because you're using the same names and their relationship is staying fixed as shown in the example above in materials and methods you are more likely to come up with text that a piece of software like turn it in will show is plagiarised even when it isn't.

### 2024-09-11 2.17a-courses.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

@@
-# Choosing your courses
+# Choosing your postgraduate courses
 
-As a postgraduate student, your university should offer you a number of courses designed for postgraduate students.
+As a postgraduate student, your university should offer you a number of courses designed specifically for postgraduate students.
 In many countries, these postgraduate courses will be compulsory and admission to the graduate school (Admission to Candidacy) will only be made after you have completed sufficient courses with grades to qualify you.
-For example, in the USA there are normally two years of these postgraduate courses followed by three years of working on your thesis project.
+These are often made up in terms of credits that you gain for each course attended and/or pass.
+There are normally one or two years of these postgraduate courses followed by three years of working on your thesis project.
+
+Postgraduate courses are widely acknowledged as the reason why postgraduate students attain great advancement in their studies.
+As proof of this, know that countries that have historically never had such courses have now introduced them.
+Despite all PhD students already having a wide range of skills from their undergraduate and MSc studies, the skills attained from postgraduate courses are designed to augment their knowledge base to a higher level so that they attain systematic expertise within the discipline, and understand the dynamics of the frontier.
+Hence, these courses should not only provide the basic skills for you to understand the methods of scientific research, but they should enable you to fast-track to the vanguard of international academic frontiers.
 
@@
 1. Look back at your short and medium term goals (see [here](#Plan)) and consider which courses could be relevant in your future
+1. Some courses may have certification and be essential during your studies but also useful in your following career
+1. An important aim during this period is to build your network, so take note of who is teaching the courses and be cognisant how they might fit into your future network. 
+1.
+The courses that you attend as a postgraduate may well be listed on your certification.
+Some courses may well endear you to certain employers, so consider your medium to long term goals (see [here](#Plan))
 1.
@@
 Given that you have already done a lot of courses at university, you are going to need some that hold genuine interest for you.
-Remember that diversity is an important aspect of science, so think as widely as possible.
+Remember that diversity is an important aspect of science, so think as widely as possible
+
+### Mandatory courses
+
+Universities are usually very strict about their rules, but these should also available to you and so it is well worth asking for a copy.
+The rules for your institution may change from year to year and so it will be important to make sure that you have the relevant version, and if possible get these from the person responsible.
+These rules may also include other important information regarding the examination process (see [Part 1](#criteria)).
+
+The compulsory courses mainly refer to the participation of graduate students in academic activities, teaching practice and social practice.
+As the name suggests, these are mandatory for all students and so there is no need to consider any choices here.
+It may be incumbent on you to ensure that you have fulfilled all necessary credits for these courses.
+For this reason, I suggest that you are familiar with the current rules of your faculty.
+
+### Elective courses
+
+Although you have your mandatory courses decided for you, you also will likely have the opportunity to do some elective courses.
+If there are elective courses that are a close fit to your thesis topic, then clearly you should elect to take these courses.
+The advice here comes more for other points that you may want to consider if you have credits to make up and cannot decide what to do.
+
+1.
+**Consider your timetable:** You will have a timetable that you need to work to and make sure that you can physically get to a course as well as fit it in with your other mandatory and elective choices.
+1. **Where does the course happen?:** the courses may take place in another campus or online. This may suit you better, or prohibit you from attending.
+1.
+**How many credits does the course carry?:** You will need to make up a total amount of credits to gain admission to your postgraduate school.
+You might need to trade-off the time required for a course and the credits that it gives.
+1.
+**Consider something from another faculty:** If allowed in your institution, consider taking a course given in another faculty.
+Some might be very challenging if you have no undergraduate or MSc experience, but others could be very rewarding.
+    1. Social Sciences - very useful courses might include GIS courses in geography, 
+    1. Social Sciences - many biological systems have important human components that can really only be solved using questionnaires
+    1. Social Sciences - philosophy has some very important lessons for biologists, especially in relation to logic and hypothesis building
+    1.
+Art - There is a lot to be said for having a creative elective.
+A surprisingly large part of the process of doing science is actually creative.
+If you can conduct a course that will stimulate the creative part of your learning, so much the better.
+    1. Medicine - _Homo sapiens_ has become an excellent model organism, and might be much better studied in the same areas that you are working in. 
+1.
+**Something enjoyable** - the chances are that some of your mandatory courses are going to be less enjoyable, so it's fair to give yourself something more enjoyable to look forward to during your week.
+
+### Certified courses
+
+You may find that attending and passing some postgraduate courses will provide you with a certificate.
+For example, you may need to attend a course on laboratory animal management and welfare.
+Certificates from such courses may well be advantageous to you in your career following your postgraduate studies, making such certificated courses particularly attractive.
 
@@
 
-Some universities allow their students to attend postgraduate courses at other (usually nearby and affilitated) universities.
+Some universities allow their students to attend postgraduate courses at other (usually nearby and affiliated) universities.
 If you know of a course that it being conducted at another university, but that you would like to attend, it might be worth approaching your postgraduate teaching co-ordinator to ask them if you can go.
 
-There is also the opportunity to conduct a postgraduate course online.
+There may also be the opportunity to conduct a postgraduate course online.
 Again if you know of one, or put in the time to find an appropriate one, it may be worth approaching your postgraduate teaching co-ordinator to ask if you can do it.
-
 The most important consideration is whether or not your institution is able to use marks generated at another institution in their own system to assess your postgraduate courses.
@@
 You need to make sure that you get the most out of these courses.
+Your attitude toward each course will be an important aspect of how well you perform and enjoy the course.
+Hence, if you feel that one of your mandatory courses is a waste of time and you wish you didn't have to do it, you will be unlikely to get much out of this course.
+On the other hand, if you have a positive attitude towards all of your courses (mandatory and elective) you will likely find them much more interesting and useful to your time as a postgraduate student.
 
-Looking back at being an undergraduate, you may have been focussed on grades or shortcutting work on courses, but for postgraduates this must have changed.
-
-## Getting the most our of your courses
-
-Remember that it is not direct applicability that is of the only importance.
+Looking back at being an undergraduate, you may have been focussed on grades or short-cutting work on courses, but for postgraduates this must have changed.
+You should understand by now that learning is a two-way process and that engaging with your teachers is a vital part of improving your understanding of a topic.
+Your teachers are likely to have an incredible depth of information on the topic that they are teaching you, and probing this knowledge will help to to appreciate the subject and help you garner new interest.
+Therefore you should engage with your teachers on these courses.
+Remember that building your network is an important aim for you during this time.
 
@@
 Use your time with them to [increase your network](https://howtopublishscience.org/advisorless.html#networks).
+You should be using the courses to network among as many fellow postgraduate students as possible.
+There are likely to be people using your approaches and techniques in various other disciplines within your university, and making these connections can be very important in the future.
 
@@
 I you haven't already worked this out for yourself, learning is a real gift and the postgraduate courses you are offered at universities are probably the ultimate in their level of sophistication and knowledge transfer.
+Remember that your attitude is key in this process, so be sure that you are mentally prepared to having a positive attitude to all of your courses.
 Take this opportunity seriously and you should find it extremely rewarding.

### 2024-09-11 3.10-Figures.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

@@
 
-Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’s books is R ([see part 1](#useR)), and the many packages that have been developed for use within the R environment [@rcoreteam2021language].
+Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’s books is R ([see part 1](#useR)), and the many packages that have been developed for use within the R environment [@rcoreteam2024language].
 The distinct advantages that this produces are the almost limitless additions and details that you can make to these figures to produce exactly what you want, in a [vector format](#vector).

### 2024-09-11 4.035-history.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

# The history of the thesis

In this guide on how to write a PhD, I often refer to your thesis as a product that is more than the sum of its parts. The modern thesis is usually made up of a series of studies, each of which can be submitted for publication in a scientific journal, but that are related through a logical framework. Later, in Part III, I will discuss the importance of your thesis conceptualisation and how this can be used to pull your chapters together into a coherent argument.

This idea of a 'coherent argument' is at the heart of the history of the term thesis, which was defined by Aristotle in his treatise: Topica (384–322 BCE).

> **Thesis** comes from the Greek word "θέσις", meaning to put forwards, and referred to laying out an intellectual proposition. Aristotle being the first philosopher to define the word.

We have other words that we use for the way in which Aristotle used the word thesis, such as supposition, conjecture or hypothesis. But Aristotle's use of the word was specifically for the use of a contrary viewpoint to another that was held by the general populous or by another philosopher.

Although this historical context appears far from our use of the term 'thesis' today, you can still see how a modern thesis is used to test previous general assumptions or specific viewpoints. In the same way that the original thesis argued the case against these views, a modern thesis holds them up to greater scrutiny, and then through a set of inductive reasoning, discusses the veracity of their validity. Perhaps more importantly, the history of the term stresses how the thesis needs to come together into a conherent synthesis that can be presented to an audience. From the origin then, a thesis was a coherent concept, and not a set of related ideas or findings. In the same way, your thesis needs to be pulled together to produce a sum that is greater than the constituent parts.

Another word that is commonly used is dissertation, and today the terms thesis and dissertation are used interchangeably. However, the origin of the term dissertation is later than that of thesis:

> **Dissertation** comes from the Latin word "dissertātiō", meaning "discussion".

Hence, from the origin of the two terms we learn that the thesis puts forward an concept that may challenge a commonly held viewpoint, while a dissertation is more focussed on discussion of the concept. Perhaps this later term allows for a more balanced and less biased view from the author where the outcome of the study is logically considered against the background of previous findings and thought. Either way, the history of these terms is helpful for you as the author of a thesis in that it helps you see how what you are producing is a synergistic synthesis of studies which produce a coherent and discursive viewpoint gained from your years of scholarly work, and presented to an audience for consideration.

### The oldest theses

Although we should expect that the libraries at the oldest universities should hold theses from their students dating back to medieval times when the majority of oldest currently extant universities began, this is not the case.

For example, at the University of Oxford (established around 1200 but teaching since 1096), the library informs us that they have only been requiring PhD theses to be deposited since 1920, and this did not become compulsary for some subjects until 1953 (see [here](https://www.bodleian.ox.ac.uk/collections-and-resources/theses-and-dissertations/oxford-theses)). Hence, the tradition of keeping theses, and maintaining a resource of the knowledge gained during years of academic study, is relatively modern and certainly within the last 100 years.

Prior to these times, PhD theses were submitted to publishers for publication as books, a tradition that continues to this day in some subject areas. Hence, you might be able to find the published version of a university thesis that dates back beyond 100 years, but you won't know what students worked on what theses where and when.

The university thesis as we know it today, is a relatively new concept and does not have the ancient history that we might expect of our oldest scholarly institutions. I find this quite surprising, but perhaps this is just naivety on my part. After all, peer review has only been in existence for for a similar amount of time. Digital (and hence available online) theses have only become a requirement for less than 20 years, and so even my PhD only exists as a paper copy on a shelf in a university library (although I'll admit that I've never looked for it).

Perhaps the take home message here is that you should not be too caught up in the pomp that your institution (likely) holds around their great tradition of theses submission. And there is no great reason why we should not evolve and update the PhD thesis that is something more worthy of the 21st Century.

## A modern definition

The ISO (the International Organization for Standardization) defines a thesis / dissertation in ISO 7144 as:

> Document which presents the author's research and findings and submitted by him in support of his candidature for a degree or professional qualification. [@iso1986iso]

The document looks very interesting and has the potential to be of great use especially if you feel that your institution is not giving sufficient guidance in the preparation of your thesis. Most countries adhere to ISO standard (look for your country [here](https://www.iso.org/about/members)). ISO 7144 is, however, secured behind a paywall and therefore not free for you to access unless you want to pay for the privaledge. Yet another example of why paywalls are never a good idea (see Part II).

## Regional peculiarities

Wikipedia has a list of how theses differ depending on which country you are in, and even which region of which country (see [here](https://en.wikipedia.org/wiki/Thesis)). Although this may help you to see how different countries treat the thesis, the best place to look for information on your thesis is the faculty in which you are registered at your own university. Because universities confer the degree of PhD to those who pass their requirements, it is up to each university to provide their particular standards, which are likely to differ between faculties. You should be able to request a guidebook for your own institution from the faculty office or your library. This document should trump any other opinions given to you from others, and you should definitely be using it when it comes to the final preparation of the thesis that you are going to submit for your PhD.

## Final submission

It is important to note that your thesis is not finished until you have made the final submission to your institution. This process is normally handled by the library, but you can only receive permission to submit a final thesis after you have undergone the examination process and your faculty's dean is happy for your final submission.

Your university library is the final resting place for your thesis, where it can **rest in peace**.

### 2024-09-11 index.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

@@
 date: "`r Sys.Date()`"
-documentclass: krantz
+documentclass: book
 bibliography:
@@
 Whether you are a student just starting your studies, or a supervisor struggling to cope, this book will provide the insider information you need to get ahead."
-github-repo: yihui/bookdown-crc
+github-repo: https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences
 graphics: yes
@@
 **Cover illustration**: Saint George and the Dragon was adapted by the author from an original work by Paolo Uccello (~1470).
-While the student (St George), supported by their laboratory (White Horse), is attempting to bring about
-the end of their PhD (Dragon) using this Fool’s Guide (lance), the advisor (Princess) is leading the PhD by a leash from the darkness into the community towards the Scientific Project (town).
+While the student (St George), supported by their laboratory (White Horse), is attempting to bring about the end of their PhD (Dragon) using this Fool’s Guide (lance), the advisor (Princess) is leading the PhD by a leash from the darkness into the community towards the Scientific Project (town).
 
@@
 Instead of rewriting comments on manuscripts time after time, I decided to consolidate them and put them all in one place.
-Later, I started looking for particular problems that students were having, and then writing blog posts to try to preempt those problems, and to provide solutions before tasks were completed.
+Later, I started looking for particular problems that students were having, and then writing blog posts to try to pre-empt those problems, and to provide solutions before tasks were completed.
 Finally, I have tried to fill in the missing gaps in the blog posts.
@@
 Note that this is not a rulebook.
-There are times when the guide will be accurate and others when it will be vague, but providing some insight to point you in directions where you can explore more.
+Your university, faculty and department will all have rule books that you should obtain and become familiar with.
+This book does not replace these rule books.
+Instead, this guide tries to explain how to achieve your goals of getting a PhD.
+There are times when this guide will be accurate and others when it will be vague, but providing some insight to point you in directions where you can explore more.
 The intention then is to provide you with a starting point from which you can establish your confidence in the academic writing process, and build your own creativity.
@@
 This book is intended to be a guide.
+Originally, I called it the 'Fool's Guide to Writing a PhD in Biological Sciences', but that didn't fly.
 Don't sit down and read it from cover to cover.
 If you do, then I think you have likely misunderstood what this guide is for.
-Instead, I want you to think about this book as a guide much in the same way that you might pick up a guide to a foreign city, country or even like Douglas Adams' fabled electronic guide _The Hitchhiker's Guide to the Galaxy_ @adams2017ultimate (if you are not familiar [see here](https://en.wikipedia.org/wiki/The_Hitchhiker%27s_Guide_to_the_Galaxy)).
-However, instead of having hostels or restaurants recommended by name with a description of what makes them particularly praiseworthy, this guide provides a generic idea of what such places are for, what they tend to be like, the best ways to get to them, and pitfalls to avoid.
+Instead, I want you to think about this book as a guide much in the same way that you might pick up a guide to a foreign city or country.
+However, instead of having hotels or restaurants recommended by name with a description of what makes them particularly praiseworthy, this guide provides a generic idea of what such places are for, what they tend to be like, the best ways to get to them, and pitfalls to avoid.
 
@@
 [**Part 1**](#part1)
--	Part one is written as something that you should read early on in the process, once you’ve committed to doing your PhD.
-This section of the book deals with what a PhD is, how it’s examined, with whom, where and when to do it.
-Perversely, it asks you what you want to do after you’ve finished, pointing out that you will have opportunities to open doors during the process.
+-	Part one contains core skills that are relevant to all postgraduate students.
+These core skills are those that you will need to build upon from the very beginning of your studies.
+You are likely to already have some experience in some of them, but understanding how they fit together will provide a solid basis for your PhD process.
+Both critical reading and regular writing are core skills that you will need to develop from the outset of your studies.
+One of the most important aspects of doing a PhD is knowing what you want to use it for once you've completed it.
+Hence, this section of the book asks you to look not only at short term day-to-day planning, but also what you want to do after you’ve finished, pointing out that you will have opportunities to open doors during the duration of your PhD.
 At the start of your PhD is the time to be thinking of the bigger questions both in your study area, and your life, and this section provides some pointers and things to consider.
-Working within academia is particularly stressful environment, and students should be aware of the state of their mental health, how to watch become aware if it is deteriorating, and what you can do to help yourself.
-I also encourage you to participate in the **scientific project**, and become aware of your role as a scientist within society.
+Working within academia is particularly stressful, and students should be aware of the state of their mental health, how to measure it and determine whether it is deteriorating, and if so, what you can do to help yourself.
+I also encourage you to participate in the scientific project, and become aware of your role as a scientist within society.
 
@@
 I also provide some alternative approaches for those who have the luxury of more time to attack these sections in more detail.
-The very end of the book sets out some reasons for why you need to publish your thesis chapters.
+The very end of the book sets out some reasons why you need to publish your thesis chapters.
 
-Throughout this guide you will find links and references of things to read elsewhere.
+Throughout this guide you will find links and references of articles to read elsewhere.
 The reality is that there is a lot out there for you to read.
@@
 So, this guide is my practical attempt to help you to get the most that you can out of your PhD time.
-After all, it should be an amazing journey that culminates with you achieving academic maturity: the ability to conceive and carry out scientific investigations.
+You will doubtless hear many academics referring to your PhD studies as the best of times.
+Their retrospective view probably won't resonate with you as it's easy to forget all of the hard work involved.
+But you can only make the most out of your time if you know what you are trying to achieve, and that is the aim of this guide.
+After all, yoru PhD should be an amazing journey of discovery that culminates with you achieving academic maturity: the ability to conceive and carry out scientific investigations.
 When you have a PhD, you become a scientist.
-Too often, however, it's a nightmare journey of cul-de-sacs, and groping in the dark.
 
@@
 But finish it you must.
+
+## What's not in this book? {-}
+
+Depending on just how early you are in your career, there is a lot missing from this book that has been provided in another book: [How to publish in Biological Sciences](https://howtopublishscience.org/index.html).
+That book concentrates on publishing manuscripts, while this book concentrates on getting students started writing data chapters.
+Hence, if you want extra information about publishing in the biological sciences, I would point you to the other book.
+If you are still writing and generating content then this is the right book for you.
+
+If you want more help with publishing, and want help to demystify the publishing process, then [How to publish in Biological Sciences](https://howtopublishscience.org/index.html) has the content that you need:
+
+```{r fig.align='centre', out.width = '40%', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/'}
+knitr::include_graphics('figures/cover2.png', dpi = NA)
+```
+
+There is common ground in both books, and I will point to important chapters relevant to publishing in the other book that are not reproduced here.
 
@@
 A special mention must go to my brother, Richard, who has hosted my lab website for more than a decade, and especially for saving blog posts from hacking attacks.
-Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning.
-James Baxter-Gilbert, Ellie Brown, Jack Dougherty, Anthony Herrel, Andrea Hurst, Allan Ellis, Joe McArthur, Andrea Melotto, Max Mühlenhaupt, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, Ben Stevenson, James Vonesh, Carla Wagener all read or commented on different aspects of the book.
+Thanks also to Thalassa Matthews, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning.
+James Baxter-Gilbert, Ellie Brown, Sebastian Chekunov, Jack Dougherty, Anthony Herrel, Andrea Hurst, Allan Ellis, Joe McArthur, Andrea Melotto, Max Mühlenhaupt, Lisa and Mark O'Connell, Claire Riss, Pat Schloss, Ben Stevenson, James Vonesh, Carla Wagener all read or commented on different aspects of the book.
 Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and in the library at Stellenbosch University.
